### 2023-07-19 outline.Rmd
first draft

---
title: "Estimating differences in scale for high throughput sequencing analysis with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor, Michelle Pistner, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---
# scale simulation for the no-math crowd

Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis

# Introduction

HTS is pervasive in the life sciences. It is used to read out many different types of experimental designs; bulk and single-cell transcriptomics, shotgun and amplicon metagenomics, in vitro selection experiments, and more. There are a large number of analysis tools that are purpose-built for each experimental design with limited cross-over. In part this is because different tools make different assumptions about the underlying data that may or may not be met.

The process of sequencing a DNA fragment with high throughput sequencing (HTS) includes wet-lab methods that introduce unacknowledged bias that affects the downstream statistical analysis. In many cases in many datasets these biases may be minor and manageable. However, often the biases are unknown and ignored by both the experimentalist and the analyst.

HTS starts by making a 'library' which is a fixed size sample from the environment. The library is usually combined with other libraries through 'multiplexing' where again the goal is to combine the libraries such that each library contributes the the same number of molecules to the pool. Finally, a fixed number of molecules is sampled from the pool and loaded onto the instrument. Thus, the process of sequencing is akin to taking a poll (making a library), combining polls (multiplexing into a pool) and taking a poll of the pool. This process removes any relationship between actual numbers sampled from the environment and the number of molecules sequenced; that is, it removes the effect of scale on the output data.

All HTS data are normalized after sequencing to make the different libraries comparable. There are two main normalization methods. The first type of normalization uses an internal standard. This can be as simple as converting the counts to proportions; i.e., by dividing the count for each part in a sample by the total count of the sample. Derivatives of this method include the TPM and RPKM approaches where the proportion is multiplied by a constant (TPM by 10^6) or by a pair of constants (RPKM by 10^6 and the length of the part).  The other internal-only normalization is the CLR, where the count of each part is divided by the geometric mean of the counts of the parts in each sample and a logarithm is taken of the resulting ratio. In this case, a small pseudo-count or prior is first applied. The second class of normalization uses an external standard. Normalizations such as the RLE, TMM and CSS methods assume that the majority of the parts are invariant and can be identified and then used to determine a relative scale between samples. Each sample is then normalized by multiplying or dividing by this relative scale to produce normalized counts.

It should be clear that all normalizations are ratios where the only difference is in how the denominator is chosen. A further complication arises during analysis when logarithmic transformations may or may not be applied, and the CLR (or other log-ratio transform) is the only transformation that explicitly uses as log-transformed data as the starting point. Several groups have pointed out that apparent differences between normalizations [2Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020]. Indeed in many datasets a logarithm of one of the transforms exhibits behaviours that is similar to the logarithm of another transform [@GloorAJS2023].

These tranformations all have the effect of further reducing the variation between samples in the dataset such that the only variation remaining is relative variation. Nevertheless, there are many instances where the size of the measured system is an important confounding variable[@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than do non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria can have differnt gorwth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance [@Zozaya:2010], and the composition is dramatically different as well [@Ravel:2010,@Hummelen:2010]. Finally, antibiotic treatment can change both the total abundance and the composition of microbiomes grown in culture and in the human gut [???].

Clearly we need some way to account for the variation in, and effect of, scale of the underlying systems [@nixon2023scale] when determining differential abundance using high throughput sequencing. A proper model for DA estimates both the difference in abundance (location) and the difference in size (scale), but the differences in scale are unavailable after sequencing. However, the tools currently used to estimate measurement error (shot noise) use either the NB (or modified derivative) or Dirichlet models post sequencing, and these are used to estimate only location. No tool accounts for differences in the intrinsic or extrinsic scale. While we cannot measure scale directly,  we can estimate its effect post-hoc on results. By not accounting for scale we are usually committing Type I errors and presuming we have much more power than is actually available, but can also commit Type II errors [@nixon2023scale]. As one example, recent guidance suggests \emph{a very small number of samples} for RNA-sequencing because of overpowering [@Schurch:2016aa].

# Results - examples:

We use two example datasets that show common types of problems.

The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knockout. Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]. This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes because increasing sample size results in more false positive identifications of differential abundance [@Schurch:2016aa]. As shown in Figure 1 A,B,D,E the root cause of the false positives is the very large number of genes with  low or even negligible variance. We can see that almost all the transcripts that are differentially abundant with a FDR < 0.01 (red) have extremely low dispersion. In the most extreme cases transcripts with near 0 difference have a low FDR, leading to the common practice of choosing transcripts with a low FDR and at least a Difference of $2^{1.4}$, and these limits are shown by the dashed grey lines. A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts. This results begs the question, "why bother with the signficance test?".

It should be obvious that the root cause is the very low dispersion parameter which arises because  scale variation in the underlying data has been removed through sequencing. While we cannot determine the underlying scale of the data, we can estimate the effect on the output if scale could be included. In the case of simple underestimation of variance as in this dataset, we can use the `gamma=1` parameter when we call either `aldex()`, or `aldex.clr()` to add uncertainty around the differential scale of the data.

Applying this parameter we can see that the large number of transcripts with near 0 dispersion now have substantial dispersion (Figure 1C), and this results in many fewer transcripts being called significantly different as shown in the volcano plot in Figure 1F. Indeed, adding scale reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff.

```{r fig1, echo=T, warning=F, message=F,comment=F, result=F }
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

yst <- read.table('~/Documents/0_git/datasets/transcriptome.txt', header=T, 
  row.names=1)
# Gierlinski:2015aa
yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
conds <- c(rep('S', 44), rep('W', 42))
coldata <- data.frame(conds)

# DESeq2
dds <- DESeqDataSetFromMatrix(countData = yst,
          colData = coldata, design= ~ conds)
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="conds_W_vs_S")
set.seed(2023)
#ALDEx2
x <- aldex.clr(yst, conds, verbose=F)
x.e <- aldex.effect(x, verbose=F)
x.t <- aldex.ttest(x, verbose=F)
set.seed(2023)
x.s <- aldex.clr(yst, conds, gamma=1, verbose=F)
x.s.e <- aldex.effect(x.s, verbose=F)
x.s.t <- aldex.ttest(x.s, verbose=F)

sig.des <- which(res@listData$padj < 0.01)
sig.ald <- which(x.t$we.eBH < 0.01)
sig.s.ald <- which(x.s.t$we.eBH < 0.01)
```

In the unscaled situaiton, DESeq2 identifies `r length(sig.des)` and ALDEx2 identifies `r length(sig.ald)` of `r nrow(yst)` transcripts. Applying the rule of at least a $2^{1.4}$ fold change reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2. Applying only the `gamma=1` parameter results in `r length(sig.s.ald)` without resorting to the fold change cutoff. We can also see from the volcano plot that there is a substantially better relationship between fold change and the adjusted p-value.

```{r plot1, echo=F}
par(mfrow=c(2,3))
plot(res@listData$lfcSE, res@listData$log2FoldChange, xlim=c(0,1), 
  col=rgb(0,0,0,0.1), xlab='LFC SE', ylab='log2 Difference')
title('A: DESeq2 effect', adj=0, line= 0.8)
points(res@listData$lfcSE[sig.des],
  res@listData$log2FoldChange[sig.des], col=rgb(1,0,0,0.5), 
  pch=19, cex=0.5)
  abline(h=c(-1.4,1.4), lty=2, col='grey')

plot(x.e$diff.win, x.e$diff.btw, col=rgb(0,0,0,0.1), 
   xlab='Dispersion', ylab='log2 Difference')
title('B: ALDEx2 effect', adj=0, line= 0.8)
points(x.e$diff.win[sig.ald], x.e$diff.btw[sig.ald], 
  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.e$diff.win, x.s.e$diff.btw, col=rgb(0,0,0,0.1),
   xlab='Dispersion', ylab='log2 Difference', xlim=c(0.1,5))
title('C: ALDEx2 scaled effect', adj=0, line= 0.8)
points(x.s.e$diff.win[sig.s.ald], x.s.e$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

# volcano
plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('D: DESeq2 volcano', adj=0, line= 0.8)
points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,0,0,0.5), 
  pch=19, cex=0.5)
  abline(v=c(-1.4,1.4), lty=2, col='grey')

plot(x.e$diff.btw, -1*log10(x.t$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('E: ALDEx2 volcano', adj=0, line= 0.8)
points(x.e$diff.btw[sig.ald], -1*log10(x.t$we.eBH[sig.ald]  +1e-70), 
  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
    abline(v=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.e$diff.btw, -1*log10(x.s.t$we.eBH +1e-15), col=rgb(0,0,0,0.1),
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('F: ALDEx2 scaled volcano', adj=0, line= 0.8)
points(x.s.e$diff.btw[sig.s.ald], -1*log10(x.s.t$we.eBH[sig.s.ald]+1e-15),
  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
    abline(v=c(-1.4,1.4), lty=2, col='grey')

```

How is scale altering the dispersion? Figure 2 shows that the density of the clr values for the initial count data, for the unscaled posterior distribution and for the scaled posterior distribution for the gene 'NTS1-2'. In this example the distribution of expression values for the wild-type group is in blue and for the snf-2 knockout in red.

We can see that the major effect of  generating a posterior distribution is to broaden the tails of what is possible to observe for both categories. Adding in scale uncertainty produces an even broader distribution for the posterior for both distributions with the greatest effect, in this gene, found in the wild-type distribution. Broadening the distribution increases the pooled dispersion and so reduces the likelihood of the false-positive identifications seen in the unscaled analysis.

```{r uncertainty, echo=T, warning=F, message=F,comment=F, result=F, caption='Incorporating uncertainty in both measurement and in scale. Panel A shows the underlying point estimates and the density of the data. Panel B shows the distribution and density of the posterior estimate of the data after the Dirichlet Monte-Carlo sampling. Panel C shows the distribibution and density of the posterior estimate with scale uncertainty included; in this example gamma is set to 1. The distribution for group A is shown in red and for group B is shown in blue.'}

yst.clr <- apply(yst+0.5, 2, function(x) log2(x) - mean(log2(x)))

par(mfrow=c(1,3))
plot(density(yst.clr['NTS1-2',1:44]), col='red', main='Point Estimate',
  xlab='clr Value', xlim=c(-3,1))
points(density(yst.clr['NTS1-2',45:86]), col='blue', type='l')
segments(yst.clr['NTS1-2',1:44], 0.05, yst.clr['NTS1-2',1:44], 0, col='red')
segments(yst.clr['NTS1-2',45:86], 0.05, yst.clr['NTS1-2',45:86], 0, col='blue')

  aldex.plotFeature(x, 'NTS1-2', pooledOnly=T, densityOnly=T)

    aldex.plotFeature(x.s, 'NTS1-2', pooledOnly=T, densityOnly=T)
```

The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa,@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and BV-affected women.  In this environment, both the relative abundance of species between groups is different as is the gene expression levels within a species [@macklaim:2013]. We expect that the total bacterial load is about 10X more in BV than in H [@Zozaya:2010]. Thus, this is  an extremely challenging environment to determine differential abundance. Indeed, the accepted method to analyze vaginal metatranscriptomes is to conduct a taxon by taxon analysis [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa] because a pooled analysis falsely identifies many housekeeping genes as being differentially abundant between groups.

Figure 3A shows an effect plot of the data where reads are grouped into parts by function, corresponding approximately to orthologous proteins regardless of the organism of origin. Each point represents one of the 3728 functions, and we can see that there are many more functions represented in the BV group (bottom) than in the healthy group (top). This is because the Lactobacilli that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, and also because there is a greater diversity of organisms in BV than in H samples.

We can also see that there are a large number of functions that are shared between the two groups, this largely corresponds to core metabolic functions that would not be expected  not contribute to differences in pathogenicity. Overplotting the core translation (blue) or glycolysis functions (orange) shows anomalies in both their their location and scale in Figure 3A. The major group of these functions is located off the line of no difference (being approximately located at -1.5) and not surprisingly have among the lowest dispersion in the dataset. Nevertheless, they are identified as differentially abundant (red) along with many others. While changes in the abundance of housekeeping functions is a useful proxy for relative abundance in the environment, they tell us nothing about the functional capacity of the two groups because these are functions in common to every organism. Of more interest is determining the functions that are different between groups that are unique or over-expressed in one group relative to the other.

```{r meta, echo=T, warning=F, message=F,comment=F, result=F }

e.min <- read.table('~/Documents/0_git/Log-Ratio-Publication/data/twntyfr.txt', 
  header=T, row.names=1, check.names=F, sep="\t", comment.char="", quote="")

ribo <- c(grep("LSU", rownames(e.min)), grep("SSU",
   rownames(e.min)))
glycol <- c(2418,1392,1305,1306,2421,1049)
sparse.set <- names(which(apply(e.min, 1, min) == 0))

conds <-c("H","H","H","H","B","H","B","B","H","B","H","B","B","B","B",
  "B","B","B","H","B","H","H")

xt <- aldex.clr(e.min, conds)
xt.e <- aldex.effect(xt)
xt.t <- aldex.ttest(xt)
xt.all <- cbind(xt.e, xt.t)

## single gamma model
xg <- aldex.clr(e.min, conds, gamma=1)
xg.e <- aldex.effect(xg)
xg.t <- aldex.ttest(xg)
xg.all <- cbind(xg.e, xg.t)
## new scale model

mu.vec <- gsub('H', log2(1), conds)
mu.vec <- as.numeric(gsub('B', log2(1.05), mu.vec))
mu.mod <- sapply(mu.vec, FUN = function(mu) rlnorm(128, mu, 1))
xt.m <- aldex.clr(e.min, conds, gamma=t(mu.mod))
xt.m.e <- aldex.effect(xt.m)
#hist(xt.m.e$diff.btw, breaks=99)
#plot(density(xt.m.e$diff.btw))
#abline(v=0, lty=2, lwd=2, col='red')
xt.m.t <- aldex.ttest(xt.m)
xt.m.all <- cbind(xt.m.e, xt.m.t)

par(mfrow=c(1,3))
aldex.plot(xt.all)
title('A: ALDEx2 unscaled', adj=0, line= 0.8)
points(xt.all$diff.win[ribo], xt.all$diff.btw[ribo], col='blue', 
  cex=0.6, lwd=1.5)
points(xt.all$diff.win[glycol], xt.all$diff.btw[glycol], col='orange',
  cex=0.6, lwd=1.5)

aldex.plot(xg.all, xlim=c(0.3,9))
title('B: ALDEx2 gamma scaled', adj=0, line= 0.8)
points(xg.all$diff.win[ribo], xg.all$diff.btw[ribo], col='blue', cex=0.6,
  lwd=1.5)
points(xg.all$diff.win[glycol], xg.all$diff.btw[glycol], col='orange', 
  cex=0.6, lwd=1.5)

aldex.plot(xt.m.all, xlim=c(0.3,9))
title('C: ALDEx2 both scaled', adj=0, line= 0.8)
points(xt.m.all$diff.win[ribo], xt.m.all$diff.btw[ribo], col='blue',
  cex=0.6, lwd=1.5)
points(xt.m.all$diff.win[glycol], xt.m.all$diff.btw[glycol], col='orange',
  cex=0.6, lwd=1.5)

```

Simply adding in scale uncertainty with `gamma=1` as before has the expected effect of increasing the dispersion as seen in Figure 3B. However, there is little effect on the location of the housekeeping functions and many of those in common between the two groups are still identified as differentially abundant; a Type I error. Furthermore, there are many functions that are in the BV group that are not identified because of this location shift; a Type II error.

Figure 3C shows the effect of adding in differential scale to the two groups to account for the fact that there is a large difference in underlying abundance in the environment. Here we keep the `gamma` parameter near 1, but shift the location of the parameter in the BV samples. This has the effect of moving the location of the common housekeeping genes to the midline of no difference between groups, and provides us with a fair comparison between these two highly disparate microbiomes.

Discussion
  Scale is important and overlooked
    - solves some issues with HTS that are ignored
  Scale robustness should be an important part of data analysis
  Sca

# Methods or supplement
We can decompose the underlying environment data into two parts composed of the relative amount and the total amount. We will denote a sequence count dataset as an \(D \times N\) matrix \(Y\), with elements \(Y_{dn}\) denoting the number of sequenced DNA molecules mapping to the \(d^{th}\) entity (e.g., taxa or gene) in the \(n^{th}\) sample.  \(Y_{dn}\) is an estimate of the relative part in the  underlying environment.

A key element of scale reliant inference (SRI) is that the observed data is an imperfect measurement of the underlying biological system which we denote \(W\) and call a \textit{scaled system}. Like \(Y\), \(W\) is a \(D \times N\) matrix.
Unlike \(Y\), the elements \(W_{dn}\) denote the true (as opposed to measured) amount of entity \(d\) in the biological system from which the \(n\)-th sample was taken.

The counts \(W\) have two parts; first scale (i.e., total amounts) and second composition (i.e., relative amounts).
We denote the scale of the system as the \(N\)-vector \(W^{\perp}\) with elements \(W^{\perp}_{n}=\sum_{d=1}^{D}W_{dn}\).
We denote the composition of the system as a \(D\times N\) matrix \(W^{\parallel}\) with elements \(W^{\parallel}_{dn}=W_{dn}/W^{\perp}_{n}\).
It follows that the composition and scale of the system uniquely determine the system via: \(W_{dn}=W^{\parallel}_{dn}W^{\perp}_{n}\).
Throughout this paper, we use hat notation to denote an estimate of a quantity  (e.g., \(\hat{W}\) is an estimate of \(W\)).

  We can let \(W\) be the data matrix or table that describes the system we want to measure. This systems contains \(N\) samples and \(D\) parts where each part is a gene, species, etc and we denote the \(d\)-th part in the \(n\)-th sample as \(W_{dn}\). The actual data is composed of both proportional (relative) and scale (size) information, such that \(W = W^{\parallel}\) and \(W^{\perp}\). It is possible that the relative and size information can vary independently. For example, the total number of RNA molecules in a cell may vary by type, but the relative amounts of many of them may be constant; alternatively the total number of molecules may be the same, but the relative proportions may change. Of course, there could be a mixture of both possibilities as well.

We can directly estimate \(W^{\parallel}\) by taking a random sample and determining the counts of the parts in the sample. The properties of this measure \(Y\)  should be well known, and the accuracy with which \(Y\) is a good estimate of \(W^{\parallel}\) is determined by the number of samples.

   - \(Y\) is a point estimate of the data and under-samples the true underlying distribution because of sparse random sampling.

  - \(W^{\parallel}\) is a more accurate estimate of the sampling uncertainty derived from  Dirichlet Monte-Carlo sampling to 'fill-in'  and provide a posterior distribution of plausible values based on the point estimates

  - \(W\) with a given amount of scale uncertainty can be estimated by adding scale uncertainty \(W^{\perp}\) with \(\gamma\) = 1. This has the result of further smoothing and broadening the distribution to account for the uncertainty in the scale of the system from which the data was drawn.

  So in formal terms, the system we want to measure is a set of N samples with D parts (genes, species, etc) contained in a matrix or data table W. The observed data is in a matrix Y, which contains the same 
  We incorporate measurement uncertainty as uncertainty around the observed value
  We incorporate scale uncertainty as uncertainty around the correction used (Gx)
  	- contant uncertainty increases dispersion
  	- different amounts of uncertainty increases dispersion and moves the center

### 2023-08-21 go2.Rmd
third attempt

---
title: "Estimating differences in scale for high throughput sequencing analysis with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---
# scale simulation for the no-math crowd

Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis

# Introduction

High throughput sequencing (HTS) analysis is commonly used to read out many different types of experimental designs; bulk and single-cell transcriptomics, amplicon and shotgun metagenomics, in vitro selection experiments, and more. There are many analysis tools that are purpose-built for each experimental design and that work well when all assumptions are fulfilled. However, different tools make different assumptions about the underlying data and when the tools fail, they usually fail without error.

Despite the variety, all tools and analyses suffer from two common issues. First, that many features are statistically significantly different between groups even though the measured difference is very small. Second, that the results can be skewed if the underlying dataset has an asymmetric distribution between the groups. Here we show that both of these issues can be resolved by modelling the size, or scale, of the data in the underlying environment.

It is instructive to understand how the data are generated. HTS starts by making a 'library' which is a fixed-size sample from the environment. The library is usually combined with other libraries through 'multiplexing' where  the goal is to combine the libraries such that each library contributes equivalent numbers of molecules to the pool. Finally, a fixed number of molecules is sampled from the pooled samples and loaded onto the instrument. Thus, the process of sequencing is akin to taking a poll (making a library), combining polls (multiplexing into a pool) and taking a poll of the pool. It should be clear that this process removes any relationship between actual numbers in the environment and the actual number of molecules actually sequenced. Thus, the total number of molecules that are sequenced for a sample is simply a nuisance parameter.

All HTS data must be standardized and normalized after sequencing to make the different samples comparable because the total count per sample is non-informative. All of these normalizations are ratios but there are two main main approaches to identifying the denominator.  The first class of normalization uses a denominator that is internal to each sample. This can be as simple as converting the counts to proportions; i.e., by dividing the count for each feature in a sample by the total count of the sample. Derivatives of this method include the TPM (transcripts per million), FPKM( fragments per million), and RPKM (reads per kilobase per million) [@Mortazavi:2008]. In these approaches the proportion is multiplied by several constants that depend on the instrument, gene characteristics in the sample and the sequencing depth. These normalizations track closely the counts of genes determined using fluorescent hybridization methods, with typical correlations above 0.8 [@Moffitt:2016aa,@Taniguchi:2010aa].  The other internal-only normalization is the CLR (centred log-ratio) transform[@aitchison1982]. Here, the count of each feature in each sample is divided by the geometric mean of the counts in the sample with a logarithm being taken of the resulting ratio. In this case, a small pseudo-count or prior is first applied since a log(0) is undefined. The second class of normalization uses a denominator that is determined from data external to the sample being normalized, but that is still derived only from the sequencing data itself. Normalizations such as the RLE (relative log expression)[@Anders:2010], TMM (trimmed mean of M values)[@Robinson:2010a] and CSS (cumulative sum scaling)[@Paulson:2013aa] methods assume that a majority of the features are invariant and that a relative scale can be determined for each sample by dividing by a correction factor. These normalizations specifically normalize the read counts for each feature to a common total so that the counts can be compared between samples directly. In particular the TMM normalization was intended to account for asymmetry in occurrance or abundance in one group relative to the other, but assumes that the majority of features are invariant [@Robinson:2010a]. Note that the CLR normalization also assumes that a majority of the parts are invariant [@Wu2021].

A final complication arises during analysis when logarithmic transformations may or may not be applied, and the CLR is the only transformation that explicitly uses log-transformed data as the starting point. Several groups have pointed out that apparent differences between the behaviour of individual normalizations [@Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020]. Indeed in many datasets a logarithm of one of these transforms exhibits behaviours that is similar to the logarithm of one or more other transforms [@GloorAJS2023]. Finally, each normalization is a point estimate of the appropriate denominator and so an additional assumption is that the denominator is a good estimator of the measure.

These tranformations along with the process of sequencing itself remove all but the relative variation between samples in the dataset. Nevertheless, data that are generated by sequencing come from systems where scale is usually important and may be a confounding variable [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than do non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance [@Zozaya:2010], and the composition is dramatically different as well [@Ravel:2010,@Hummelen:2010]. Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but we can access only the composition.

It is enlightening to examine the dispersion or variance of the data as counts and as the logarithm of counts. The counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1 and this is why most tools model the counts with this distribution. However, the actual analysis is usually performed on the logarithm of the counts [Robinson:2010, Love:2014aa], or on the logarithm of the clr values[@fernandes:2013] and the mean-dispersion distribution of the logarithm of the counts is quite different as shown in panel B of Figure 1 and as noted elsewhere [@fernandes:2013, @Love:2014aa].

```{r dispersion, echo=F,warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as both counts and as logarithms of counts. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion, here the standard error (SE) of the mean, is very different when the data are log-transformed first. In addition, the amount of dispersion reaches a minimum at moderate count values. The red line in each panel shows the loess line of fit to the mid-point of the distributions. The dashed orange line in panel A is the line of equivalence, and in panel B is the minimum y value." }
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

yst <- read.table('~/Documents/0_git/datasets/transcriptome.txt', header=T, 
  row.names=1)
# remove the one gene with 0 reads

yst <- yst[rownames(yst) != "YOR072W-B",]

# Gierlinski:2015aa
yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL

conds <- c(rep('S', 44), rep('W', 42))
coldata <- data.frame(conds)

# DESeq2
dds <- DESeqDataSetFromMatrix(countData = yst,
          colData = coldata, design= ~ conds)
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="conds_W_vs_S")
set.seed(2023)
#ALDEx2
x <- aldex.clr(yst, conds, gamma=1e-3,verbose=F)
x.e <- aldex.effect(x, include.sample.summary=T, verbose=F)
x.t <- aldex.ttest(x, verbose=F)
set.seed(2023)
x.s <- aldex.clr(yst, conds, gamma=1, verbose=F)
x.s.e <- aldex.effect(x.s, include.sample.summary=T, verbose=F)
x.s.t <- aldex.ttest(x.s, verbose=F)

sig.des <- which(res@listData$padj < 0.01)
sig.ald <- which(x.t$we.eBH < 0.01)
sig.s.ald <- which(x.s.t$we.eBH < 0.01)

# get and plot mean/var or mean/SE for counts
# and log counts
yst.mn <- apply(yst, 1, mean)
yst.mn.log <- apply(yst, 1, function(x) mean(log2(x)))
yst.mn.log[is.infinite(yst.mn.log)] <- 0

yst.v <- apply(yst, 1, var)

par(mfrow=c(1,2))
plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean count', ylab='Variance')
def <- data.frame(log2(yst.mn),log2(yst.v))
lines(lowess(def, f=0.1), col=2, lwd=2)
title('A: Mean vs. Variance', adj=0, line= 0.8)
abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)

l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
lines(lowess(l.df), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
title('B: log2 Mean vs. log2 Variance', adj=0, line= 0.8)

```

By not accounting for scale we are usually identifying statistically significant changes that are biologically meaningless but are found because of the artificially low dispersion.  As one example, recent guidance suggests \emph{using only a  small number of samples} for bulk RNA-sequencing because  only a few samples is needed to identify the majority of statistically significant features found with a large sample size [@Schurch:2016aa]. This popularized the useage of both p-values and log2 fold change between groups as dual cutoffs when determining significance. However, it is also possible to fail to identify true positive differences because of scale [@nixon2023scale].

Clearly we need some way to account for the variation in, and effect of, the scale of the underlying systems [@nixon2023scale] when determining differential abundance (DA) using high throughput sequencing. While normalizaitons attempt to correct for differences in relative counts, no tool accounts for differences in the intrinsic or extrinsic scale of the underlying system. Moreover, while we cannot measure scale directly, we can estimate its effect post-hoc on results.

# Results - examples:

We can model the conundrum of scale by assuming that the true values from the system that we want to measure are counts  composed of both compositional and total values, and that their product is the full scaled system that fits the equation: \(W = W^{\parallel}W^{\perp}\). If the true data have $D$ features (genes, taxa, etc) and $N$ samples, then  $W$ is a $D \times N$ matrix of values and we can denote the true  count of a feature in a sample as \(W_{dn}\). However, sequencing only does not allow us to access \(W^{\parallel}\) directly, but instead returns only a single estimate of the composition for each feature \(Y^{\parallel}_{dn}\). This single value is not definitive since different technical replicates  return different numbers for \(Y^{\parallel}_{dn}\) because of sampling variation [@Marioni:2008,@fernandes:2013].

We can use Bayesian techniques to estimate both the sampling variation of \(W^{\parallel}\) and the scale uncertainty of \(W^{\perp}\) by Monte Carlo sampling with the ALDEx2 R package. ALDEx2 generates pseudo technical replicates of \(Y^{\parallel}\) by Monte Carlo sampling from a Dirichlet distribution to produce a posterior distribution for \(Y^{\parallel}\) that is  an estimate of \(W^{\parallel}\)[@fernandes:2013,@gloorAJS:2016]. Then a CLR transformation is applied to each Monte Carlo sample to produce a distribution of log-ratio transformed values. Crucially, the denominator used for the CLR can be seen as a point estimate of the scale of the system [@nixon2023scale]. Thus, modifying ALDEx2 by replacing the point estimate for the CLR denominator with Monte Carlo sampled values from a log-Normal distribution gives a distribution of credible denominators. Thus, the CLR values now are credible estimates for both  \(W^{\parallel}\) and \(W^{\perp}\). This allows the scaled CLR distribution to be used to estimate the effect of scale on the output and so make \(Y \sim W\) under the assumption of the scale uncertainty. Below we use two example datasets to show how including scale uncertainty can be used to make inferences about differentially abundant features more reliable.

The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knocK0ut[@Gierlinski:2015aa]. Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]. This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes and that different tools should be used for datasets with different sample sizes because the tools have different intrinsic statistical power [@Schurch:2016aa]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet is entrenched in all fields that use HTS as an experimental readout.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a FDR of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts. Applying the rule of thumb of at least a $2^{1.4}$ fold change reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2. Clearly, the necessary assumption of most features being invariant is not justified.

As shown in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2. We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups. In the most extreme cases transcripts with near 0 difference have a low FDR. This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines. A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts. This results begs the question, "why bother with the signficance test?".

The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization. The actual scale of the data is unaccessible post-sequencing but we can estimate the effect of scale on the output. To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples, and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013]. Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`. The `ALDEx2` package contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty. In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.

Applying `gamma=1` as a parameter we can see that the large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts (`r length(sig.s.ald)`) being called significantly different  as shown in the volcano plot in Figure 1F (red points). Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.  Indeed, adding scale uncertainty reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff. Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.

```{r plot1, echo=F, fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis. DESeq2 or ALDEx2 were used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion (effect plot) or difference and the Benjamini-Hochberg corrected p-values (volcano plot). Panels A,B,D,E are for the unscaled analysis, and Panels C,F are for the scaled analysis. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the scaled analysis and unscaled analysis (red) or in the unscaled analysis only (orange). Points in grey are not statistically signficantly different with any analysis. The horizontal dashed lines represent a log2(difference) of 1.4, which is a commonly applied cutoff when the majority of features are statistically significant."}
par(mfrow=c(2,3))
plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
title('A: DESeq2 effect', adj=0, line= 0.8)
points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
  pch=19, cex=0.5)
points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
  pch=19, cex=0.5)
abline(h=c(-1.4,1.4), lty=2, col='grey')

plot(x.e$diff.win, x.e$diff.btw, col=rgb(0,0,0,0.1), 
   xlab='Dispersion', ylab='log2 Difference')
title('B: ALDEx2 effect', adj=0, line= 0.8)
points(x.e$diff.win[sig.ald], x.e$diff.btw[sig.ald], 
  col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')
points(x.e$diff.win[sig.s.ald], x.e$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.6), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.e$diff.win, x.s.e$diff.btw, col=rgb(0,0,0,0.1),
   xlab='Dispersion', ylab='log2 Difference', xlim=c(0.1,5))
title('C: ALDEx2 scaled effect', adj=0, line= 0.8)
points(x.s.e$diff.win[sig.s.ald], x.s.e$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

# volcano
plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('D: DESeq2 volcano', adj=0, line= 0.8)
points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
  pch=19, cex=0.5)
points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
  pch=19, cex=0.5)
  abline(v=c(-1.4,1.4), lty=2, col='grey')

plot(x.e$diff.btw, -1*log10(x.t$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('E: ALDEx2 volcano', adj=0, line= 0.8)
points(x.e$diff.btw[sig.ald], -1*log10(x.t$we.eBH[sig.ald]  +1e-70), 
  col=rgb(1,.66,0,.3), cex=0.5, pch=19)
points(x.e$diff.btw[sig.s.ald], -1*log10(x.t$we.eBH[sig.s.ald]  +1e-70), 
  col=rgb(1,0,0,1), cex=0.5, pch=19)
abline(v=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.e$diff.btw, -1*log10(x.s.t$we.eBH +1e-15), col=rgb(0,0,0,0.1),
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('F: ALDEx2 scaled volcano', adj=0, line= 0.8)

points(x.s.e$diff.btw[sig.s.ald], -1*log10(x.s.t$we.eBH[sig.s.ald]+1e-15),
  col=rgb(1,0,0,1), cex=0.5, pch=19)
    abline(v=c(-1.4,1.4), lty=2, col='grey')

```

The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa,@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and BV-affected women.  In this environment, both the relative abundance of species between groups is different as is the gene expression levels within a species [@macklaim:2013]. We further expect that the total number of bacteria is about 10X more in BV than in H [@Zozaya:2010]. Thus, this is  an extremely challenging environment to determine differential abundance. Indeed, the accepted method to analyze vaginal metatranscriptomes is to conduct a taxon by taxon analysis rather than a system-wide analysis [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa] because a pooled analysis unexpectedly identifies many housekeeping genes as being differentially abundant between groups.

In this example we show how to specify the scale model explicitly and show that applying different scale models to each group can control for the very large difference in scale in the underlying data. When specifying the whole scale model we can pass a matrix of scale values instead of a single to `aldex.clr()`. This matrix should have the same number of rows as the of Monte-Carlo Dirichlet samples, and the same number of columns as the number of samples. This matrix encapsulates the additional uncertainty of the scale model on a per-sample basis.

Figure 3A shows an effect plot [@gloor:effect] of the data where reads are grouped by function, corresponding approximately to orthologous proteins regardless of the organism of origin. Each point represents one of the 3728 functions, and we can see that there are many more functions represented in the BV group (bottom) than in the healthy group (top). This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because  there is a greater diversity of organisms in BV than in H samples and because the BV condition has at least an order of magnitude more bacteria than does the H condition.

 We can  see that there are a large number of functions that are shared between the two groups (Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected  to contribute to differences in ecosystem behaviour. As a proxy for housekeeping functions the core ribosome functions (blue) shows that their mean location is not centred on 0.The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5) and not surprisingly have among the lowest dispersion in the dataset. Nevertheless, they are identified as differentially abundant (red) along with many others. While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups because these are functions in common to every organism. Of more interest is determining the functions that are different between groups that are unique or over-expressed in one group relative to the other.

```{r ribo, echo=F}
ribo.v <- c("K02863","K02864","K02865","K02866","K02867","K02871","K02872","K02873","K02874","K02876","K02877","K02878","K02879","K02880","K02881","K02883","K02884","K02885","K02886","K02887","K02888","K02890","K02892","K02895","K02897","K02899","K02902","K02904","K02905","K02906","K02907","K02909","K02911","K02913","K02914","K02916","K02920","K02922","K02926","K02931","K02932","K02933","K02935","K02936","K02938","K02939","K02941","K02943","K02945","K02946","K02948","K02949","K02950","K02952","K02954","K02955","K02956","K02959","K02961","K02963","K02964","K02965","K02967","K02968","K02969","K02970","K02975","K02981","K02982","K02984","K02986","K02987","K02988","K02990","K02991","K02992","K02994","K02995","K02996","K02997","K02998","K07590")
```

```{r makeScaleMatrix, echo=T}
aldex.makeScaleMatrix <- function(gamma, diff, conditions){
  ## new scale model
  # mu1 is the base relative variance
  # mu2 is the other relative variance
  # gamma is the dispersion paramenter
  gamma=gamma
  mu1 = 1 # set base to 1
  mu2 = 1 + diff # set other to adjust this until the housekeeping is centred

  # note: it is the log2 difference between mu1 and mu2 that is key here
  # eg; mu1=1, mu2=1.15 is equivalent to mu1=4, mu2=4.6
  # log2(1)=0, log2(1.15)~0.2; log2(4)=2, log2(4.6)~2.2

  mu.vec <- gsub(levels(factor(conditions))[1], log2(mu1), conditions)
  mu.vec <- as.numeric(gsub(levels(factor(conditions))[2], log2(mu2), mu.vec))
  return(t( sapply(mu.vec, FUN = function(mu) rlnorm(128, mu, gamma)) ))
}
```

```{r meta, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Analysis of vaginal transcriptome data aggregated at the Kegg Orthology (KO) functional level. Panel A shows the default analysis with samples from healthy individuals at the top and from BV individuals at the bottom. Highlighed in the box are highly abundant KOs that are almost exlusively housekeeping functions, with ribosomal KOs highlighted in blue,  statistically significant (FDR < 0.01) functions in red, and non-significant functions in black or orange. These housekeeping functions are off the midline of no difference. Panel B shows the same data scaled with `gamma = 1`, which increase the minimum dispersion approximatly by one unit. Here the housekeeping functions from Panel A are colored cyan or blue for reference. Panel C shows the same data scaled with `gamma = 0.75` and a 0.15 fold difference in dispersion applied to the BV samples relative to the H samples. The orange functions are now statistically significant. Note that this shifts the midpoint of the housekeeping functions towards the midline." }

set.seed(2023)
load('~/Documents/0_git/projects/metatranscriptome/Rdata/ko.both.all.Rda')

# make a vector of conditions
conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8))

xt <- aldex.clr(ko.both.all, conds.K0)
xt.e <- aldex.effect(xt, include.sample.summary=T)
xt.t <- aldex.ttest(xt)
xt.all <- cbind(xt.e, xt.t)

## single gamma model
xg <- aldex.clr(ko.both.all, conds.K0, gamma=1)
xg.e <- aldex.effect(xg)
xg.t <- aldex.ttest(xg)
xg.all <- cbind(xg.e, xg.t)

mu.mod <- aldex.makeScaleMatrix(0.75, .15, conds.K0)

xt.m <- aldex.clr(ko.both.all, conds.K0, gamma=mu.mod)
xt.m.e <- aldex.effect(xt.m, include.sample.summary=T)
xt.m.t <- aldex.ttest(xt.m)
xt.m.all <- cbind(xt.m.e, xt.m.t)

hk <- rownames(xt.e)[xt.e$diff.win < 2.5 & xt.e$diff.btw > 0 & xt.e$diff.btw < 3]
hk.off <- xt.e$diff.win < 2.5 & xt.e$diff.btw > 0 & xt.e$diff.btw < 3

fn <- rownames(xt.m.all)[xt.m.all$we.eBH < 0.01 & xt.m.all$diff.btw < 1 & xg.all$we.eBH > 0.01 & xt.all$we.eBH > 0.01]

#par(mfrow=c(1,1))
#plot(density(xt.m.e$diff.btw[hk.off]))
#abline(v=0, col='red')

par(mfrow=c(1,3))
aldex.plot(xt.all, cutoff.pval=0.01)
points(xt.all[ribo.v,'diff.win'], xt.all[ribo.v,'diff.btw'], col='blue', cex=0.5, pch=19)
points(xt.all[fn,'diff.win'], xt.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
rect(0.5,0,3,5, col=rgb(0,0,0,0.1), lty=2, lwd=3)
title('A: ALDEx2 unscaled', adj=0, line= 0.8)

aldex.plot(xg.all, xlim=c(0.3,9), cutoff.pval=0.01)
points(xg.all$diff.win[hk.off], xg.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
points(xg.all[ribo.v,'diff.win'], xg.all[ribo.v,'diff.btw'], col='blue', cex=0.5, pch=19)
points(xg.all[fn,'diff.win'], xg.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
title('B: ALDEx2 gamma scaled', adj=0, line= 0.8)

aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01)
points(xt.m.all$diff.win[hk.off], xt.m.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
points(xt.m.all[ribo.v,'diff.win'], xt.m.all[ribo.v,'diff.btw'], col='blue', cex=0.5, pch=19)
points(xt.m.all[fn,'diff.win'], xt.m.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
title('C: ALDEx2 both scaled', adj=0, line= 0.8)

```

Applying `gamma=1` as before increases the dispersion as expected, but does little to move the large number of housekeeping functions toward the midline of no difference. Nevertheless, about 50% of the housekeeping functions are no longer statistically significantly different.

Up to this point, scale uncertainty has been applied uniformly to both conditions, but the scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix. This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group. Applying a differential scale of 0.15, or 15% of the base scale now moves the housekeeping functions to the midline of no difference. Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because ... [HELP]. Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied. These former false negatives are noted in orange in each panel. Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant. Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.

Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant.

# Discussion

Biological count data can be decomposed into two parts the relative (compositional) and the absolute (scale), and the product of these generates a fully scaled biological system [@nixon2023scale]. Biological systems are inherently variable and stochastic and current measurement methods that rely on high throughput sequencing fail to capture all of that variation. In the absence of  information external to the sequencing run itself, no normalisation method can recapture any of the scale information, including scale variation [@Loven:2012aa].

While the underlying scale of the system cannot be measured easily, the effect on analysis can by including scale uncertainty in the analysis [@nixon2023scale]. Nixon et al, showed that this can be done by including uncertainty in the denominator used for the normalization. The ALDEx2 R package is ideally suited for this since this tool builds a Bayesian posterior of the compositional component of the dataset at the outset and then conducts the analysis on that posterior. Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty. For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a log-Normal distribution.

All normalizations attempt to make the samples in a dataset commensurate but cannot explicitly address the scale of the underlying system. However, the general lack of scale information has important consequences for the analysis of HTS datasets. One issue is that analysis tools  seem over-powered with even moderate sample sizes  [@Schurch:2016aa]. Using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reliable p-values (eg. [@Halsey:2015aa]). Thus, recommendations to use small sample sizes in multivariate datasets such as RNA-seq datasets are not supported by simple modelling in the univariate case. Another issue is that datasets are difficult to analyze when there they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a, @Wu2021].

In the case of overpowering, HTS analyses seem to be more robust when applying a dual cutoff of both p-value and difference between group means [@Schurch:2016aa]. Figure 2 shows one reason for this robustness could be that the dual cutoff is mimicking the effect of including scale uncertainty, since substantially similar transcripts are identified by the two approaches. However, while using the post-hoc difference cutoff is useful for differential abundance analysis it is not clear how this can be incorporated into other kinds of downstream analyses. Conversely data that include scale uncertainty are fully compatible with existing downstream analyses.

In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult to analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change. We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot be used. Incorporating differential scale on a per-group basis moves the mass of the data towards the midline of no difference and so affects both Type I and Type II error rates. Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant. In this analysis, transcripts that were previously not classed as differentially abundant are now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such. While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa,@SEQC/MAQC-III-Consortium:2014aa]. Thus, the use of this prior knowledge is not unique to our approach.

In summary, while the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled. Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on the analysis weighted accordingly. Additionally, the use of differential scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups. Thus, reportin scale uncertainty should become a standard practice in the analysis of HTS datasets as a way to identify which features are most robust to differences in the underlying system. Furhtermore, we supply a set of toosl that make incorporating scale simple even for datasets that come from highly asymmetrical environments.

# Methods or supplement

Count data can in the underlying system be decomposed into two parts; the relative amount (composition) and the total amount (scale). We can describe the full scaled system made up of the composition and the scale with the equation: \(W = W^{\parallel} \dot W^{\perp}\).  What we measure by sequencing is only the composition part  \(Y^{parallel}\) which is a single point estimate of the composition \(W^{parallel}\) of the underlying true system. We can denote the sequence count dataset as an \(D \times N\) matrix, with elements \(Y_{dn}\) indicating the number of sequenced DNA molecules mapping to the \(d^{th}\) entity (e.g., taxa or gene) in the \(n^{th}\) sample. We will use hat notation to denote an estimate of a quantity  (e.g., \(\hat{W}\) is an estimate of \(W\)).

It is possible that the relative and size information can vary independently. For example, the total number of RNA molecules in a cell may vary by type, but the relative amounts of many of them may be constant; alternatively the total number of molecules may be the same, but the relative proportions may change. Of course, there could be a mixture of both possibilities as well.

Making a library from the molecules in \(W\), sequencing and generating the count table from the output provides a single point estimate of \(W^{\parallel}\). The accuracy with which \(Y^{\parallel}\) is a good estimate of \(W^{\parallel}\) is determined by the number of samples, the number of DNA molecules sequenced relative to the size of the system and is affected by the bioinformatic pipelines. Nevertheless, it is assumed that \(Y^{\parallel} \sim W^{\parallel}\) under these constraints.

   - \(Y^{\parallel}\) is a point estimate of the data and under-samples the true underlying distribution because of sparse random sampling and because the number of samples sequenced is usually very small.

  - ALDEx2 uses \(Y^{\parallel}_{d.}\) as the basis to generate a posterior distribution by Monte-Carlo sampling from a Dirichlet distribution, where the posterior distribution contains a large number of credible values for \(W^{\parallel}\).

  - The Monte-Carlo instances are then CLR transformed. Here the denominator of the CLR is a single point estimate of the underlying scale.

  - ALDEx2 has been modified include Monte-Carlo sampling from a log-Normal distribution to generate a posterior distribution of credible scale values, given the estimated scale in \(W\). The amount of scale is controlled either by providing a single value \(\gamma\) which is the standard deviation (SD) of the scale, or by providing a matrix of SDs that vary between groups.

  - This has the result of further smoothing and broadening the distribution to account for the uncertainty in the scale of the system from which the data was drawn.

  - Applying uniform scale uncertainty increases the dispersion of the

  So in formal terms, the system we want to measure is a set of N samples with D parts (genes, species, etc) contained in a matrix or data table W. The observed data is in a matrix Y, which contains the same 
  We incorporate measurement uncertainty as uncertainty around the observed value
  We incorporate scale uncertainty as uncertainty around the correction used (Gx)
  	- contant uncertainty increases dispersion
  	- different amounts of uncertainty increases dispersion and moves the center

### 2023-08-21 go3.Rmd
third attempt

--- 2023-08-21 go2.Rmd
@@
 ---
-title: "Estimating differences in scale for high throughput sequencing analysis with ALDEx2"
+title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
+"
 shorttitle: "Scale ALDEx2"
@@
 ---
-# scale simulation for the no-math crowd
-
-Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis
 
@@
 
-High throughput sequencing (HTS) analysis is commonly used to read out many different types of experimental designs; bulk and single-cell transcriptomics, amplicon and shotgun metagenomics, in vitro selection experiments, and more.
-There are many analysis tools that are purpose-built for each experimental design and that work well when all assumptions are fulfilled.
-However, different tools make different assumptions about the underlying data and when the tools fail, they usually fail without error.
+High throughput sequencing (HTS) is a universal tool to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing).
+HTS proceeds by taking a fixed-number sample from the environment, making a library, multiplexing (merging) multiple libraries and applying a fixed number of molecules to the flow cell.
+In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken.
+It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment.
 
-Despite the variety, all tools and analyses suffer from two common issues.
-First, that many features are statistically significantly different between groups even though the measured difference is very small.
-Second, that the results can be skewed if the underlying dataset has an asymmetric distribution between the groups.
-Here we show that both of these issues can be resolved by modelling the size, or scale, of the data in the underlying environment.
-
-It is instructive to understand how the data are generated.
-HTS starts by making a 'library' which is a fixed-size sample from the environment.
-The library is usually combined with other libraries through 'multiplexing' where  the goal is to combine the libraries such that each library contributes equivalent numbers of molecules to the pool.
-Finally, a fixed number of molecules is sampled from the pooled samples and loaded onto the instrument.
-Thus, the process of sequencing is akin to taking a poll (making a library), combining polls (multiplexing into a pool) and taking a poll of the pool.
-It should be clear that this process removes any relationship between actual numbers in the environment and the actual number of molecules actually sequenced.
-Thus, the total number of molecules that are sequenced for a sample is simply a nuisance parameter.
-
-All HTS data must be standardized and normalized after sequencing to make the different samples comparable because the total count per sample is non-informative.
-All of these normalizations are ratios but there are two main main approaches to identifying the denominator.
-The first class of normalization uses a denominator that is internal to each sample.
-This can be as simple as converting the counts to proportions; i.e., by dividing the count for each feature in a sample by the total count of the sample.
-Derivatives of this method include the TPM (transcripts per million), FPKM( fragments per million), and RPKM (reads per kilobase per million) [@Mortazavi:2008].
-In these approaches the proportion is multiplied by several constants that depend on the instrument, gene characteristics in the sample and the sequencing depth.
-These normalizations track closely the counts of genes determined using fluorescent hybridization methods, with typical correlations above 0.8 [@Moffitt:2016aa,@Taniguchi:2010aa].
-The other internal-only normalization is the CLR (centred log-ratio) transform[@aitchison1982].
-Here, the count of each feature in each sample is divided by the geometric mean of the counts in the sample with a logarithm being taken of the resulting ratio.
-In this case, a small pseudo-count or prior is first applied since a log(0) is undefined.
-The second class of normalization uses a denominator that is determined from data external to the sample being normalized, but that is still derived only from the sequencing data itself.
-Normalizations such as the RLE (relative log expression)[@Anders:2010], TMM (trimmed mean of M values)[@Robinson:2010a] and CSS (cumulative sum scaling)[@Paulson:2013aa] methods assume that a majority of the features are invariant and that a relative scale can be determined for each sample by dividing by a correction factor.
-These normalizations specifically normalize the read counts for each feature to a common total so that the counts can be compared between samples directly.
-In particular the TMM normalization was intended to account for asymmetry in occurrance or abundance in one group relative to the other, but assumes that the majority of features are invariant [@Robinson:2010a].
-Note that the CLR normalization also assumes that a majority of the parts are invariant [@Wu2021].
-
-A final complication arises during analysis when logarithmic transformations may or may not be applied, and the CLR is the only transformation that explicitly uses log-transformed data as the starting point.
-Several groups have pointed out that apparent differences between the behaviour of individual normalizations [@Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020].
-Indeed in many datasets a logarithm of one of these transforms exhibits behaviours that is similar to the logarithm of one or more other transforms [@GloorAJS2023].
-Finally, each normalization is a point estimate of the appropriate denominator and so an additional assumption is that the denominator is a good estimator of the measure.
-
-These tranformations along with the process of sequencing itself remove all but the relative variation between samples in the dataset.
-Nevertheless, data that are generated by sequencing come from systems where scale is usually important and may be a confounding variable [@Lovell:2015].
+Data that are generated by sequencing come from systems where scale is usually important and may be a confounding variable [@Lovell:2015].
 For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than do non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa].
 In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa].
-As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance [@Zozaya:2010], and the composition is dramatically different as well [@Ravel:2010,@Hummelen:2010].
-Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but we can access only the composition.
+As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance [@Zozaya:2010], and the composition is dramatically different as well [@Ravel:2010;@Hummelen:2010].
+Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but we can access only the composition directly.
 
-It is enlightening to examine the dispersion or variance of the data as counts and as the logarithm of counts.
-The counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1 and this is why most tools model the counts with this distribution.
-However, the actual analysis is usually performed on the logarithm of the counts [Robinson:2010, Love:2014aa], or on the logarithm of the clr values[@fernandes:2013] and the mean-dispersion distribution of the logarithm of the counts is quite different as shown in panel B of Figure 1 and as noted elsewhere [@fernandes:2013, @Love:2014aa].
+More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix, then the elements of the matrix can be decomposed into its composition (relative information) and its scale (total sum).
+We can formally state this as \(W = W^{\parallel} \times W^{\perp}\).
+If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(W_{dn}\) as the \(d\)^{th} feature in sample \(n\).
+Any data that we collect by sequencing is thought to contains only proportional information and we can represent the corresponding sequenced matrix as \(Y\).
+Compositional data analysis makes the strong assumption  that \(Y_{n} = Y^{\parallel}_{n} = W^{\parallel}_{n}\); in other words that relative information is the only information available post-sequencing.
+However, the values after sequencing are only estimates of the true underlying values that we want to estimate and different technical and biological replicates vary.
+Note that under this strong assumption no corresponding estimate is obtained for the scale of the sample \(Y^{\perp}_{n}\).
 
-```{r dispersion, echo=F,warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as both counts and as logarithms of counts.
+One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data from different samples.
+A large number of normalizations were developed that depended on the data source.
+These include rarefaction[@Hughes:2005tu], proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a], and the centered log ratio (CLR)[@aitchison1982].
+With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.
+
+Recently, [,@nixon2023scale] showed that the denominator used to normalize \(Y^{\parallel}_{n}\) was an estimate of the scale of the system.
+In the context of compositional data analysis, this group proved that the geometric mean of \(Y^{\parallel}_{n}\) was a biased estimate of \(W^{\perp}_n\).
+This realization opened the possibility that \(Y^{\perp}_{n}\) could be made a less biased estimate by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated.
+In terms of information theory, the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(I\) (supplemental figure XX), suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.
+This makes intuitive sense because  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim I_n\).
+
+The realization that a strong assumption about scale is built into each denominator explains why HTS data are notoriously fickle as analysis of the same dataset with different workflows, tools and assumptions giving widely different results when benchmarked[@Thorsen:2016aa;@Quinn:2018aa;@Gardner:2019aa].
+This can be understood as the result of unacknowledged bias about the scale estimate which inevitably leads to large Type 1 error rates as sample sizes increase [@Gustafson2015].
+Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be robust to differential abundance analysis even when taking scale into account.
+
+The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets.
+ALDEx2 was designed originally to convert point estimates of \(Y^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution, where the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016].
+Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates.
+The CLR calculation uses as the denominator the geometric mean of each Monte-Carlo replicate of \(Y^{\parallel}_{dn}\) with the geometric mean representing a point estimate of scale, i.e.
+of \(Y^{\perp}_{n}\).
+We  now include a posterior distribution of the scale into this process by sampling \(Y^{\perp}_n\) from a log-Normal distribution and using these as the denominator.
+The scale estimates can be calculated on a per-experiment, per-group or even a per-sample basis.
+
+An advantage of incorporating scale is that the analysis can be made much more robust and that differences in scale can be accounted for explicitly.
+The examples below show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.
+
+# Results
+
+The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knockout[@Gierlinski:2015aa].
+Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
+This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes and that different tools should be used for datasets with different sample sizes because the tools have different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa].
+This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout.
+Increasing false precision with increasing sample size can be understood as the result of unacknowledged bias brought about because of false certainty as to the scale of the data.
+Thus, by adding scale uncertainty we can be more confident that the analysis is not converging on a precise but biased estimate of scale [@Gustafson2015].
+
+We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (clr).
+The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1.
+This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab].
+However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [Robinson:2010, Love:2014aa], or on  the clr values[@fernandes:2013] both of which are  log-ratios.
+The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013, @Love:2014aa].
+
+```{r dispersion, echo=F,warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values.
 Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale.
-Panel B shows that the relationship between the mean and the dispersion, here the standard error (SE) of the mean, is very different when the data are log-transformed first.
-In addition, the amount of dispersion reaches a minimum at moderate count values.
+Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first.
+Panel C shows the equivalent values calculated by ALDEx2.
+The expected clr value for each transcipt are plotted vs.
+the expected dispersion.
+In panels B and C the amount of dispersion reaches a minimum at moderate  values.
 The red line in each panel shows the loess line of fit to the mid-point of the distributions.
-The dashed orange line in panel A is the line of equivalence, and in panel B is the minimum y value." }
+The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value.
+The dashed grey line in panels B and C represent the maximum of the lowest 10% dispersion." }
 library(DESeq2)
@@
 
-par(mfrow=c(1,2))
+par(mfrow=c(1,3))
 plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
   xlab='Mean count', ylab='Variance')
+#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
 def <- data.frame(log2(yst.mn),log2(yst.v))
@@
   xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
-lines(lowess(l.df), col=2, lwd=2)
+lines(lowess(l.df, f=0.25), col=2, lwd=2)
 abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.0275, col='grey', lty=2)
 title('B: log2 Mean vs. log2 Variance', adj=0, line= 0.8)
 
+l.df <- data.frame(x.e$rab.all, x.e$diff.win)
+plot(x.e$rab.all, x.e$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='E(clr)', ylab='E(dispersion)')
+#points(x.e$rab.all[sig.ald], x.e$diff.win[sig.ald], col='orange', cex=0.4)
+lines(lowess(l.df, f=.1), col=2, lwd=2)
+abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.208, col='grey', lty=2)
+title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
 ```
 
-By not accounting for scale we are usually identifying statistically significant changes that are biologically meaningless but are found because of the artificially low dispersion.
-As one example, recent guidance suggests \emph{using only a  small number of samples} for bulk RNA-sequencing because  only a few samples is needed to identify the majority of statistically significant features found with a large sample size [@Schurch:2016aa].
-This popularized the useage of both p-values and log2 fold change between groups as dual cutoffs when determining significance.
-However, it is also possible to fail to identify true positive differences because of scale [@nixon2023scale].
-
-Clearly we need some way to account for the variation in, and effect of, the scale of the underlying systems [@nixon2023scale] when determining differential abundance (DA) using high throughput sequencing.
-While normalizaitons attempt to correct for differences in relative counts, no tool accounts for differences in the intrinsic or extrinsic scale of the underlying system.
-Moreover, while we cannot measure scale directly, we can estimate its effect post-hoc on results.
-
-# Results - examples:
-
-We can model the conundrum of scale by assuming that the true values from the system that we want to measure are counts  composed of both compositional and total values, and that their product is the full scaled system that fits the equation: \(W = W^{\parallel}W^{\perp}\).
-If the true data have $D$ features (genes, taxa, etc) and $N$ samples, then  $W$ is a $D \times N$ matrix of values and we can denote the true  count of a feature in a sample as \(W_{dn}\).
-However, sequencing only does not allow us to access \(W^{\parallel}\) directly, but instead returns only a single estimate of the composition for each feature \(Y^{\parallel}_{dn}\).
-This single value is not definitive since different technical replicates  return different numbers for \(Y^{\parallel}_{dn}\) because of sampling variation [@Marioni:2008,@fernandes:2013].
-
-We can use Bayesian techniques to estimate both the sampling variation of \(W^{\parallel}\) and the scale uncertainty of \(W^{\perp}\) by Monte Carlo sampling with the ALDEx2 R package.
-ALDEx2 generates pseudo technical replicates of \(Y^{\parallel}\) by Monte Carlo sampling from a Dirichlet distribution to produce a posterior distribution for \(Y^{\parallel}\) that is  an estimate of \(W^{\parallel}\)[@fernandes:2013,@gloorAJS:2016].
-Then a CLR transformation is applied to each Monte Carlo sample to produce a distribution of log-ratio transformed values.
-Crucially, the denominator used for the CLR can be seen as a point estimate of the scale of the system [@nixon2023scale].
-Thus, modifying ALDEx2 by replacing the point estimate for the CLR denominator with Monte Carlo sampled values from a log-Normal distribution gives a distribution of credible denominators.
-Thus, the CLR values now are credible estimates for both  \(W^{\parallel}\) and \(W^{\perp}\).
-This allows the scaled CLR distribution to be used to estimate the effect of scale on the output and so make \(Y \sim W\) under the assumption of the scale uncertainty.
-Below we use two example datasets to show how including scale uncertainty can be used to make inferences about differentially abundant features more reliable.
-
-The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knocK0ut[@Gierlinski:2015aa].
-Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
-This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes and that different tools should be used for datasets with different sample sizes because the tools have different intrinsic statistical power [@Schurch:2016aa].
-This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet is entrenched in all fields that use HTS as an experimental readout.
+Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data[@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution.
+This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes.
+This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation[@Taniguchi:2010aa].
+Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible.
+To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives.
+The work of [,@nixon2023scale] suggests that this is because the biased estimate of \(Y^{\perp}\) that is  calculated leads to a false estimate of the scale.
+Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.
 
 Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a FDR of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts.
-Applying the rule of thumb of at least a $2^{1.4}$ fold change reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2.
 Clearly, the necessary assumption of most features being invariant is not justified.
-
-As shown in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2.
-We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups.
-In the most extreme cases transcripts with near 0 difference have a low FDR.
-This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines.
-A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts.
-This results begs the question, "why bother with the signficance test?".
-
-The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization.
-The actual scale of the data is unaccessible post-sequencing but we can estimate the effect of scale on the output.
-To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples, and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013].
-Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`.
-The `ALDEx2` package contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty.
-In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.
-
-Applying `gamma=1` as a parameter we can see that the large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts (`r length(sig.s.ald)`) being called significantly different  as shown in the volcano plot in Figure 1F (red points).
-Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.
-Indeed, adding scale uncertainty reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff.
-Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.
+Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2.
 
@@
 DESeq2 or ALDEx2 were used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset.
-The results were plotted to show the relationship between difference and dispersion (effect plot) or difference and the Benjamini-Hochberg corrected p-values (volcano plot).
+The results were plotted to show the relationship between difference and dispersion (effect plot [@gloor:effect]) or difference and the Benjamini-Hochberg corrected p-values (volcano plot[@Cui:2003aa]).
 Panels A,B,D,E are for the unscaled analysis, and Panels C,F are for the scaled analysis.
@@
 
+As shown in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2.
+We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups.
+In the most extreme cases transcripts with near 0 difference have a low FDR.
+This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines.
+A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts.
+This results begs the question, "why bother with the signficance test?".
+
+The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization.
+While the actual scale of the data is inaccessible post-sequencing  we can estimate the effect of scale on the output.
+To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013].
+Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`.
+The `ALDEx2` package contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty.
+In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.
+
+Applying `gamma=1` as a parameter we can see that the large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts (`r length(sig.s.ald)`) being called significantly different  as shown in the volcano plot in Figure 1F (red points).
+Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.
+Indeed, adding scale uncertainty reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff.
+Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.
+
 The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa,@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and BV-affected women.
 In this environment, both the relative abundance of species between groups is different as is the gene expression levels within a species [@macklaim:2013].
-We further expect that the total number of bacteria is about 10X more in BV than in H [@Zozaya:2010].
+We further know that the total number of bacteria is about 10X more in BV than in H [@Zozaya:2010].
 Thus, this is  an extremely challenging environment to determine differential abundance.
@@
 
-Figure 3A shows an effect plot [@gloor:effect] of the data where reads are grouped by function, corresponding approximately to orthologous proteins regardless of the organism of origin.
+Figure 3A shows an effect plot [@gloor:effect] of the data where reads are grouped by function, corresponding approximately to grouping orthologous sequences regardless of the organism of origin.
 Each point represents one of the 3728 functions, and we can see that there are many more functions represented in the BV group (bottom) than in the healthy group (top).
@@
 
- We can  see that there are a large number of functions that are shared between the two groups (Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected  to contribute to differences in ecosystem behaviour.
-As a proxy for housekeeping functions the core ribosome functions (blue) shows that their mean location is not centred on 0.The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5) and not surprisingly have among the lowest dispersion in the dataset.
+ We can  see that there are a large number of functions that are shared between the two groups (boxed area in Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected  to contribute to differences in ecosystem behaviour.
+As a proxy for housekeeping functions the core ribosome functions (blue) shows that their mean location is not centred on 0.
+The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5) and not surprisingly have among the lowest dispersion in the dataset.
 Nevertheless, they are identified as differentially abundant (red) along with many others.
 While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups because these are functions in common to every organism.
-Of more interest is determining the functions that are different between groups that are unique or over-expressed in one group relative to the other.
+Of more interest is determining the functions that are different between groups because these are unique or over-expressed in one group relative to the other.
 
@@
 
-```{r makeScaleMatrix, echo=T}
+```{r makeScaleMatrix, echo=F}
 aldex.makeScaleMatrix <- function(gamma, diff, conditions){
@@
 This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group.
-Applying a differential scale of 0.15, or 15% of the base scale now moves the housekeeping functions to the midline of no difference.
-Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because ...
-[HELP].
+Applying a differential scale of 0.15, or 15% of the base scale now moves the housekeeping functions to the midline of no difference (Figure 3C).
+Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because the change in scael moves the mass of the data towards the group with the highest entropy (supplemental).
 Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied.
@@
 
-Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant.
-
 # Discussion
@@
 Biological systems are inherently variable and stochastic and current measurement methods that rely on high throughput sequencing fail to capture all of that variation.
-In the absence of  information external to the sequencing run itself, no normalisation method can recapture any of the scale information, including scale variation [@Loven:2012aa].
+In the absence of  information external to the sequencing run itself, no normalisation method can recapture any of the scale information [@Loven:2012aa].
 
@@
 However, the general lack of scale information has important consequences for the analysis of HTS datasets.
-One issue is that analysis tools  seem over-powered with even moderate sample sizes  [@Schurch:2016aa].
+One issue is that analysis tools  seem over-powered with even moderate sample sizes and different tools have different intrinsic power, Type 1 and Type 2 error rates  [@Schurch:2016aa].
 Using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reliable p-values (eg.
@@
 In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult to analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
-We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot be used.
+We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference (supplemental).
 Incorporating differential scale on a per-group basis moves the mass of the data towards the midline of no difference and so affects both Type I and Type II error rates.
@@
 
-In summary, while the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled.
+In summary, while the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and explained.
 Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on the analysis weighted accordingly.
 Additionally, the use of differential scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups.
-Thus, reportin scale uncertainty should become a standard practice in the analysis of HTS datasets as a way to identify which features are most robust to differences in the underlying system.
-Furhtermore, we supply a set of toosl that make incorporating scale simple even for datasets that come from highly asymmetrical environments.
-
-# Methods or supplement
-
-Count data can in the underlying system be decomposed into two parts; the relative amount (composition) and the total amount (scale).
-We can describe the full scaled system made up of the composition and the scale with the equation: \(W = W^{\parallel} \dot W^{\perp}\).
-What we measure by sequencing is only the composition part  \(Y^{parallel}\) which is a single point estimate of the composition \(W^{parallel}\) of the underlying true system.
-We can denote the sequence count dataset as an \(D \times N\) matrix, with elements \(Y_{dn}\) indicating the number of sequenced DNA molecules mapping to the \(d^{th}\) entity (e.g., taxa or gene) in the \(n^{th}\) sample.
-We will use hat notation to denote an estimate of a quantity  (e.g., \(\hat{W}\) is an estimate of \(W\)).
-
-It is possible that the relative and size information can vary independently.
-For example, the total number of RNA molecules in a cell may vary by type, but the relative amounts of many of them may be constant; alternatively the total number of molecules may be the same, but the relative proportions may change.
-Of course, there could be a mixture of both possibilities as well.
-
-Making a library from the molecules in \(W\), sequencing and generating the count table from the output provides a single point estimate of \(W^{\parallel}\).
-The accuracy with which \(Y^{\parallel}\) is a good estimate of \(W^{\parallel}\) is determined by the number of samples, the number of DNA molecules sequenced relative to the size of the system and is affected by the bioinformatic pipelines.
-Nevertheless, it is assumed that \(Y^{\parallel} \sim W^{\parallel}\) under these constraints.
-
-   - \(Y^{\parallel}\) is a point estimate of the data and under-samples the true underlying distribution because of sparse random sampling and because the number of samples sequenced is usually very small.
-
-  - ALDEx2 uses \(Y^{\parallel}_{d.}\) as the basis to generate a posterior distribution by Monte-Carlo sampling from a Dirichlet distribution, where the posterior distribution contains a large number of credible values for \(W^{\parallel}\).
-
-  - The Monte-Carlo instances are then CLR transformed. Here the denominator of the CLR is a single point estimate of the underlying scale.
-
-  - ALDEx2 has been modified include Monte-Carlo sampling from a log-Normal distribution to generate a posterior distribution of credible scale values, given the estimated scale in \(W\).
-The amount of scale is controlled either by providing a single value \(\gamma\) which is the standard deviation (SD) of the scale, or by providing a matrix of SDs that vary between groups.
-
-  - This has the result of further smoothing and broadening the distribution to account for the uncertainty in the scale of the system from which the data was drawn.
-
-  - Applying uniform scale uncertainty increases the dispersion of the
-
-  So in formal terms, the system we want to measure is a set of N samples with D parts (genes, species, etc) contained in a matrix or data table W.
-The observed data is in a matrix Y, which contains the same
-  We incorporate measurement uncertainty as uncertainty around the observed value
-  We incorporate scale uncertainty as uncertainty around the correction used (Gx)
-  	- contant uncertainty increases dispersion
-  	- different amounts of uncertainty increases dispersion and moves the center
+Thus, reporting scale uncertainty should become a standard practice in the analysis of HTS datasets as a way to identify which features are most robust to differences in the underlying system.
+Furthermore, we supply a set of toosl that make incorporating scale simple even for datasets that come from highly asymmetrical environments.

### 2023-08-21 outline.Rmd
third attempt

@@
 There are two main normalization methods.
-The first type of normalization uses an internal standard.
+The first type of normalization uses a standard that is internal to each sample.
 This can be as simple as converting the counts to proportions; i.e., by dividing the count for each part in a sample by the total count of the sample.
-Derivatives of this method include the TPM and RPKM approaches where the proportion is multiplied by a constant (TPM by 10^6) or by a pair of constants (RPKM by 10^6 and the length of the part).
-The other internal-only normalization is the CLR, where the count of each part is divided by the geometric mean of the counts of the parts in each sample and a logarithm is taken of the resulting ratio.
+Derivatives of this method include the TPM (transcripts per million) and RPKM (reads per kilobase per million [@Mortazavi:2008] approaches where the proportion is multiplied by several constants that depend on the instrument, gene characteristics in the sample and the sequencing depth.
+The other internal-only normalization is the CLR (centred log-ratio) transform[@aitchison1982], where the count of each part is divided by the geometric mean of the counts of the parts in each sample and a logarithm is taken of the resulting ratio.
 In this case, a small pseudo-count or prior is first applied.
-The second class of normalization uses an external standard.
-Normalizations such as the RLE, TMM and CSS methods assume that the majority of the parts are invariant and can be identified and then used to determine a relative scale between samples.
+The second class of normalization uses as a standard a measure that is determined externally to the sample being normalized.
+Normalizations such as the RLE (relative log expression)[@Anders:2010], TMM (trimmed mean of M values)[@Robinson:2010a] and CSS (cumulative sum scaling)[@Paulson:2013aa] methods assume that the majority of the parts are invariant and that they  can be identified and then used to determine a relative scale between samples.
 Each sample is then normalized by multiplying or dividing by this relative scale to produce normalized counts.
@@
 A further complication arises during analysis when logarithmic transformations may or may not be applied, and the CLR (or other log-ratio transform) is the only transformation that explicitly uses as log-transformed data as the starting point.
-Several groups have pointed out that apparent differences between normalizations [2Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020].
-Indeed in many datasets a logarithm of one of the transforms exhibits behaviours that is similar to the logarithm of another transform [@GloorAJS2023].
+Several groups have pointed out that apparent differences between normalizations [@Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020].
+Indeed in many datasets a logarithm of one of these transforms exhibits behaviours that is similar to the logarithm of another transform [@GloorAJS2023].
 

### 2023-08-21 checking-housekeeping-yeast.R

@@ -1,0 +1,28 @@
+# note: need to manually pull changed genes and then get the equivalent in KEGG search
+
+up.path <-  read.table('up.path.txt', row.names=1, sep='\t')
+dn.path <-  read.table('dn.path.txt', row.names=1, sep='\t')
+
+both.path <- intersect(rownames(up.path), rownames(dn.path))
+up.only.path <- setdiff(rownames(up.path), rownames(dn.path))
+dn.only.path <- setdiff(rownames(dn.path), rownames(up.path))
+all.path <- union(rownames(up.path), rownames(dn.path))
+
+all.path.df <- as.data.frame(matrix(data=NA, ncol=3, nrow=length(all.path))
+
+colnames(all.path.df) <- c('pathway', 'N.up', 'N.dn')
+ 
+rownames(all.path.df) <- all.path
+
+rownames(all.path.df) %in% up.only.path
+
+all.path.df[rownames(all.path.df) %in% up.only.path,1:2] <- up.path[up.only.path,1:2]
+all.path.df[rownames(all.path.df) %in% dn.only.path,c(1,3)] <- dn.path[dn.only.path,c(1,2)]
+all.path.df[rownames(all.path.df) %in% both.path,c(1,3)] <- dn.path[both.path,c(1,2)]
+all.path.df[rownames(all.path.df) %in% both.path,2] <- up.path[both.path,2]
+
+up.path[up.only.path,1]
+
+ribo <- c('YBL027W','YBL072C','YBR146W','YBR181C','YBR189W','YDL075W','YDL082W','YDL136W','YDR064W','YDR237W','YDR382W','YDR418W','YDR447C','YEL054C','YER102W','YER131W','YGL068W','YGL123W','YGR085C','YHL004W','YHL015W','YIL069C','YIL133C','YJL177W','YKL180W','YKR094C','YLL045C','YLR048W','YLR333C','YLR441C','YML026C','YML063W','YMR142C','YMR242C','YNL069C','YNL178W','YOL040C','YOR096W','YOR312C','YPL131W','YPR102C')
+
+ribo.unch <- c('YBR084C-A','YBR251W','YDL081C','YDL202W','YER050C','YER117W','YGL103W','YGR027C','YGR118W','YGR220C','YHR147C','YLR287C-A','YML024W','YML025C','YMR143W','YMR194W','YMR286W','YNL306W','YOR167C','YPL173W','YPL249C-A','YPR132W')

### 2023-08-23 go3.Rmd
makefile and Michelle's comments
Makefile now, all code external, all data loaded from gloor github data repository via URL begun supplemental

@@
 One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data from different samples.
-A large number of normalizations were developed that depended on the data source.
-These include rarefaction[@Hughes:2005tu], proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a], and the centered log ratio (CLR)[@aitchison1982].
+A large number of normalizations were developed.
+Rarefaction[@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a].
+The centered log ratio (CLR)[@aitchison1982] approach was proposed as a unifying normalization[@fernandes:2014].
 With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.
 
-Recently, [,@nixon2023scale] showed that the denominator used to normalize \(Y^{\parallel}_{n}\) was an estimate of the scale of the system.
-In the context of compositional data analysis, this group proved that the geometric mean of \(Y^{\parallel}_{n}\) was a biased estimate of \(W^{\perp}_n\).
-This realization opened the possibility that \(Y^{\perp}_{n}\) could be made a less biased estimate by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated.
+Recently, [,@nixon2023scale] showed that geometric mean of \(Y^{\parallel}_{n}\) was related to the geometric mean of \(W_n\) and so was related to \(W^{\perp}_n\).
+This realization opened the possibility that adding uncertainty to the geometric mean could provide a buffer of uncertainty on it that could be used to control for  the unacknowledged bias that the geometric mean was a perfect point estimate of the underlying system.
+The bias around  could be acknowledged by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated.
 In terms of information theory, the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(I\) (supplemental figure XX), suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.
@@
 This can be understood as the result of unacknowledged bias about the scale estimate which inevitably leads to large Type 1 error rates as sample sizes increase [@Gustafson2015].
-Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be robust to differential abundance analysis even when taking scale into account.
+Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be called as differentially abundant even when scale was taken into account, and those features that are differentially abundant only under strictly assumed normalizations.
 
@@
 
-```{r dispersion, echo=F,warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values.
+```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values.
 Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale.
@@
 The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value.
-The dashed grey line in panels B and C represent the maximum of the lowest 10% dispersion." }
+The dashed grey line in panels B and C represent the maximum of the lowest 10\\% dispersion." }
+
 library(DESeq2)
@@
 
-yst <- read.table('~/Documents/0_git/datasets/transcriptome.txt', header=T, 
-  row.names=1)
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
 # remove the one gene with 0 reads
@@
 # DESeq2
-dds <- DESeqDataSetFromMatrix(countData = yst,
-          colData = coldata, design= ~ conds)
-dds <- DESeq(dds)
-resultsNames(dds) # lists the coefficients
-res <- results(dds, name="conds_W_vs_S")
-set.seed(2023)
+load('analysis/res.Rda')
+
 #ALDEx2
-x <- aldex.clr(yst, conds, gamma=1e-3,verbose=F)
-x.e <- aldex.effect(x, include.sample.summary=T, verbose=F)
-x.t <- aldex.ttest(x, verbose=F)
-set.seed(2023)
-x.s <- aldex.clr(yst, conds, gamma=1, verbose=F)
-x.s.e <- aldex.effect(x.s, include.sample.summary=T, verbose=F)
-x.s.t <- aldex.ttest(x.s, verbose=F)
+load('analysis/x.all.Rda')
+load('analysis/x.s.all.Rda')
 
+# with gamma = 1
+load('analysis/x.s.all.Rda')
 sig.des <- which(res@listData$padj < 0.01)
-sig.ald <- which(x.t$we.eBH < 0.01)
-sig.s.ald <- which(x.s.t$we.eBH < 0.01)
+sig.ald <- which(x.all$we.eBH < 0.01)
+sig.s.ald <- which(x.s.all$we.eBH < 0.01)
 
@@
 
-l.df <- data.frame(x.e$rab.all, x.e$diff.win)
-plot(x.e$rab.all, x.e$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
+l.df <- data.frame(x.all$rab.all, x.all$diff.win)
+plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
   xlab='E(clr)', ylab='E(dispersion)')
-#points(x.e$rab.all[sig.ald], x.e$diff.win[sig.ald], col='orange', cex=0.4)
+#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
 lines(lowess(l.df, f=.1), col=2, lwd=2)
@@
 Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible.
-To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives.
+To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion.
 The work of [,@nixon2023scale] suggests that this is because the biased estimate of \(Y^{\perp}\) that is  calculated leads to a false estimate of the scale.
@@
 
-Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a FDR of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts.
+Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts.
 Clearly, the necessary assumption of most features being invariant is not justified.
-Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2.
+Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.all$we.eBH < 0.01 & abs(x.all$diff.btw) > 1.4))` for ALDEx2.
 
@@
 
-plot(x.e$diff.win, x.e$diff.btw, col=rgb(0,0,0,0.1), 
+plot(x.all$diff.win, x.all$diff.btw, col=rgb(0,0,0,0.1), 
    xlab='Dispersion', ylab='log2 Difference')
 title('B: ALDEx2 effect', adj=0, line= 0.8)
-points(x.e$diff.win[sig.ald], x.e$diff.btw[sig.ald], 
+points(x.all$diff.win[sig.ald], x.all$diff.btw[sig.ald], 
   col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
     abline(h=c(-1.4,1.4), lty=2, col='grey')
-points(x.e$diff.win[sig.s.ald], x.e$diff.btw[sig.s.ald], 
+points(x.all$diff.win[sig.s.ald], x.all$diff.btw[sig.s.ald], 
   col=rgb(1,0,0,0.6), cex=0.5, pch=19)
@@
 
-plot(x.s.e$diff.win, x.s.e$diff.btw, col=rgb(0,0,0,0.1),
+plot(x.s.all$diff.win, x.s.all$diff.btw, col=rgb(0,0,0,0.1),
    xlab='Dispersion', ylab='log2 Difference', xlim=c(0.1,5))
 title('C: ALDEx2 scaled effect', adj=0, line= 0.8)
-points(x.s.e$diff.win[sig.s.ald], x.s.e$diff.btw[sig.s.ald], 
+points(x.s.all$diff.win[sig.s.ald], x.s.all$diff.btw[sig.s.ald], 
   col=rgb(1,0,0,0.5), cex=0.5, pch=19)
@@
 
-plot(x.e$diff.btw, -1*log10(x.t$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
+plot(x.all$diff.btw, -1*log10(x.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
    xlab='log2 Difference', ylab='-1 log10(p.adjust)')
 title('E: ALDEx2 volcano', adj=0, line= 0.8)
-points(x.e$diff.btw[sig.ald], -1*log10(x.t$we.eBH[sig.ald]  +1e-70), 
+points(x.all$diff.btw[sig.ald], -1*log10(x.all$we.eBH[sig.ald]  +1e-70), 
   col=rgb(1,.66,0,.3), cex=0.5, pch=19)
-points(x.e$diff.btw[sig.s.ald], -1*log10(x.t$we.eBH[sig.s.ald]  +1e-70), 
+points(x.all$diff.btw[sig.s.ald], -1*log10(x.all$we.eBH[sig.s.ald]  +1e-70), 
   col=rgb(1,0,0,1), cex=0.5, pch=19)
@@
 
-plot(x.s.e$diff.btw, -1*log10(x.s.t$we.eBH +1e-15), col=rgb(0,0,0,0.1),
+plot(x.s.all$diff.btw, -1*log10(x.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
    xlab='log2 Difference', ylab='-1 log10(p.adjust)')
@@
 
-points(x.s.e$diff.btw[sig.s.ald], -1*log10(x.s.t$we.eBH[sig.s.ald]+1e-15),
+points(x.s.all$diff.btw[sig.s.ald], -1*log10(x.s.all$we.eBH[sig.s.ald]+1e-15),
   col=rgb(1,0,0,1), cex=0.5, pch=19)
@@
 This results begs the question, "why bother with the signficance test?".
+[XXXX CHECK THE DIRECTION OF SCALE IN THIS DATASET]
 
@@
 
-```{r makeScaleMatrix, echo=F}
-aldex.makeScaleMatrix <- function(gamma, diff, conditions){
-  ## new scale model
-  # mu1 is the base relative variance
-  # mu2 is the other relative variance
-  # gamma is the dispersion paramenter
-  gamma=gamma
-  mu1 = 1 # set base to 1
-  mu2 = 1 + diff # set other to adjust this until the housekeeping is centred
-
-  # note: it is the log2 difference between mu1 and mu2 that is key here
-  # eg; mu1=1, mu2=1.15 is equivalent to mu1=4, mu2=4.6
-  # log2(1)=0, log2(1.15)~0.2; log2(4)=2, log2(4.6)~2.2
-
-  mu.vec <- gsub(levels(factor(conditions))[1], log2(mu1), conditions)
-  mu.vec <- as.numeric(gsub(levels(factor(conditions))[2], log2(mu2), mu.vec))
-  return(t( sapply(mu.vec, FUN = function(mu) rlnorm(128, mu, gamma)) ))
-}
-```
-
 ```{r meta, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Analysis of vaginal transcriptome data aggregated at the Kegg Orthology (KO) functional level.
@@
 
-set.seed(2023)
-load('~/Documents/0_git/projects/metatranscriptome/Rdata/ko.both.all.Rda')
+# all code for ALDEx2 and DESeq is being moved to the code/directory
+# all variables recreated using the makefile
 
-# make a vector of conditions
-conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8))
+load('analysis/xt.all.Rda')
+load('analysis/xg.all.Rda')
+load('analysis/xt.m.all.Rda')
 
-xt <- aldex.clr(ko.both.all, conds.K0)
-xt.e <- aldex.effect(xt, include.sample.summary=T)
-xt.t <- aldex.ttest(xt)
-xt.all <- cbind(xt.e, xt.t)
-
-## single gamma model
-xg <- aldex.clr(ko.both.all, conds.K0, gamma=1)
-xg.e <- aldex.effect(xg)
-xg.t <- aldex.ttest(xg)
-xg.all <- cbind(xg.e, xg.t)
-
-mu.mod <- aldex.makeScaleMatrix(0.75, .15, conds.K0)
-
-xt.m <- aldex.clr(ko.both.all, conds.K0, gamma=mu.mod)
-xt.m.e <- aldex.effect(xt.m, include.sample.summary=T)
-xt.m.t <- aldex.ttest(xt.m)
-xt.m.all <- cbind(xt.m.e, xt.m.t)
-
-hk <- rownames(xt.e)[xt.e$diff.win < 2.5 & xt.e$diff.btw > 0 & xt.e$diff.btw < 3]
-hk.off <- xt.e$diff.win < 2.5 & xt.e$diff.btw > 0 & xt.e$diff.btw < 3
+hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3]
+hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
 
@@
 
-Applying `gamma=1` as before increases the dispersion as expected, but does little to move the large number of housekeeping functions toward the midline of no difference.
+Applying the default scale model of `gamma=1` as before increases the dispersion as expected, but does little to move the large number of housekeeping functions toward the midline of no difference.
 Nevertheless, about 50% of the housekeeping functions are no longer statistically significantly different.
 
-Up to this point, scale uncertainty has been applied uniformly to both conditions, but the scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix.
+Up to this point, scale uncertainty has been applied uniformly to both conditions, but a user-defined scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix.
 This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group.
-Applying a differential scale of 0.15, or 15% of the base scale now moves the housekeeping functions to the midline of no difference (Figure 3C).
-Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because the change in scael moves the mass of the data towards the group with the highest entropy (supplemental).
+Applying a per-group differential scale of 0.15, or 15% along with the base scale now moves the housekeeping functions to the midline of no difference (Figure 3C).
+Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because the change in scale moves the mass of the data towards the group with the highest entropy (supplemental).
 Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied.
@@
 Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.
+Thus, while the default scale model alone is useful and should be normally applied, using prior information about the underlying biology to develop a differential scale model can be very useful for many datasets.
 
@@
 
-While the underlying scale of the system cannot be measured easily, the effect on analysis can by including scale uncertainty in the analysis [@nixon2023scale].
+Since the underlying scale of the system cannot be measured easily --  methods such as flow cytometry [@Vandeputte:2017aa], spike-in probes and fluorescent in-situ hybridization [@Loven:2012aa;@yeast-absolute] have been used -- the effect on analysis can be modelled by including scale uncertainty in the analysis [@nixon2023scale].
 Nixon et al, showed that this can be done by including uncertainty in the denominator used for the normalization.

### 2023-08-23 supplement_figs.Rmd
makefile and Michelle's comments
Makefile now, all code external, all data loaded from gloor github data repository via URL begun supplemental

---
title: "Supplement: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2 Supp"
author:
- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2 1.33.1
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
date: 21 August 2023
link-citations: true
---

# GM is correlated with Shannon's entropy and includes

```{r info, echo=F,warning=F, message=F,comment=F, fig.cap="Plot of Shannon's entropy (I) vs geometric mean (G) for each sample in three different datasets. The groups that each sample belong to are highlighted as filled or open circles. Each group in each dataset has different entropy with the groups in the selex and metatranscriptome datasets being highly distinct."}
# Gm ~ Im
# plotting geometric mean vs Shannon's entropy
# shows differential information by group
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
data(selex)

load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
yst <- read.table(url, header=T, row.names=1)

# remove the one gene with 0 reads
yst <- yst[rownames(yst) != "YOR072W-B",]
# Gierlinski:2015aa
yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL

par(mfrow=c(1,1))
plot(apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),
  apply(selex+0.5, 2, function(x) mean(log2(x))), xlim=c(2.5,11), ylim=c(-0.5,10), 
  xlab="Shannon's entropy", ylab='Geometric mean', pch=c(rep(19,7),rep(1,7)),
  col=rgb(0,0,0,0.5))
points(apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),apply(yst+0.5, 2, function(x) mean(log2(x))), col=rgb(1,0.6,0,0.5), pch=c(rep(19,44), rep(1, 42)))
points(apply(ko.both.all+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),apply(ko.both.all+0.5, 2, function(x) mean(log2(x))), col=rgb(1,0,0,0.5),
 pch=c(rep(1,8), rep(19,28), rep(1,8)))

legend(3,10, legend=c('selex', 'transcriptome','metatranscriptome', 'low entropy group', 'high entropy group'),
  col=c('darkgrey', 'orange', 'red', 'grey','grey'), pch=c(19,19,19,1,19))
```

```{r move}
load('analysis/x.s.mu.all.Rda')
load('analysis/x.all.Rda')

par(mfrow=c(1,2))
aldex.plot(x.s.mu.all)
aldex.plot(x.all)

```

### 2023-08-23 code/meta-aldex.R

@@ -1,0 +1,32 @@
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+
+set.seed(2023)
+load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
+
+# make a vector of conditions
+conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8)) 
+
+xt <- aldex.clr(ko.both.all, conds.K0)
+xt.e <- aldex.effect(xt, include.sample.summary=T)
+xt.t <- aldex.ttest(xt)
+xt.all <- cbind(xt.e, xt.t)
+
+save(xt.all, file='analysis/xt.all.Rda')
+
+## single gamma model
+xg <- aldex.clr(ko.both.all, conds.K0, gamma=1)
+xg.e <- aldex.effect(xg)
+xg.t <- aldex.ttest(xg)
+xg.all <- cbind(xg.e, xg.t)
+
+save(xg.all, file='analysis/xg.all.Rda')
+
+mu.mod <- aldex.makeScaleMatrix(0.75, .15, conds.K0)
+ 
+xt.m <- aldex.clr(ko.both.all, conds.K0, gamma=mu.mod)
+xt.m.e <- aldex.effect(xt.m, include.sample.summary=T)
+xt.m.t <- aldex.ttest(xt.m)
+xt.m.all <- cbind(xt.m.e, xt.m.t)
+
+save(xt.m.all, file='analysis/xt.m.all.Rda')
+

### 2023-08-23 code/yst-DESeq.R

@@ -1,0 +1,23 @@
+library(DESeq2)
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+
+# remove the one gene with 0 reads
+
+yst <- yst[rownames(yst) != "YOR072W-B",]
+
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
+
+dds <- DESeqDataSetFromMatrix(countData = yst,
+          colData = coldata, design= ~ conds)
+dds <- DESeq(dds)
+
+res <- results(dds, name="conds_W_vs_S")
+
+save(res, file='analysis/res.Rda')

### 2023-08-23 code/yst-aldex.R

@@ -1,0 +1,32 @@
+library(DESeq2)
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+#library(ALDEx2)
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+
+yst <- yst[rownames(yst) != "YOR072W-B",]
+
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
+
+set.seed(2023)
+#ALDEx2
+x <- aldex.clr(yst, conds, gamma=1e-3,verbose=F)
+x.e <- aldex.effect(x, include.sample.summary=T, verbose=F)
+x.t <- aldex.ttest(x, verbose=F)
+x.all <- cbind(x.e, x.t)
+save(x.all, file='analysis/x.all.Rda')
+
+set.seed(2023)
+x.s <- aldex.clr(yst, conds, gamma=1, verbose=F)
+x.s.e <- aldex.effect(x.s, include.sample.summary=T, verbose=F)
+x.s.t <- aldex.ttest(x.s, verbose=F)
+x.s.all <- cbind(x.s.e, x.s.t)
+save(x.s.all, file='analysis/x.s.all.Rda')
+

### 2023-08-23 code/yst.aldex.mu.r

@@ -1,0 +1,21 @@
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+#library(ALDEx2)
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+
+yst <- yst[rownames(yst) != "YOR072W-B",]
+
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
+
+mu.dat <- aldex.makeScaleMatrix(1,0.25,conds,mc.samples=128)
+
+set.seed(2023)
+x.s.mu.all <- aldex(yst, conds, gamma=mu.dat, verbose=F)
+save(x.s.mu.all, file='analysis/x.s.mu.all.Rda')

### 2023-08-31 go3.Rmd
clarified H and G, added bold notation

@@
 author:
-- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
-  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
+- name: Greg Gloor
+  affiliation: Dep't of Biochemistry, University of Western Ontario
   email: ggloor@uwo.ca
+- name: Michelle Pistner Nixon
+  affiliation: College of Information Sciences and Technology, Pennsylvania  State University
+- name: Justin Silverman
+  affiliation: College of Information Sciences and Technology, Department of Statistics, Department of Medicine, Pennsylvania  State University
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
@@
 More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix, then the elements of the matrix can be decomposed into its composition (relative information) and its scale (total sum).
-We can formally state this as \(W = W^{\parallel} \times W^{\perp}\).
-If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(W_{dn}\) as the \(d\)^{th} feature in sample \(n\).
-Any data that we collect by sequencing is thought to contains only proportional information and we can represent the corresponding sequenced matrix as \(Y\).
-Compositional data analysis makes the strong assumption  that \(Y_{n} = Y^{\parallel}_{n} = W^{\parallel}_{n}\); in other words that relative information is the only information available post-sequencing.
+We can formally state this as \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\).
+If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).
+Any data that we collect by sequencing is thought to contain only proportional information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\).
+Compositional data analysis makes the strong assumption  that \(\mathbf{Y}_{n} = \mathbf{Y}^{\parallel}_{n} = \mathbf{W}^{\parallel}_{n}\); in other words that relative information is the only information available post-sequencing.
 However, the values after sequencing are only estimates of the true underlying values that we want to estimate and different technical and biological replicates vary.
-Note that under this strong assumption no corresponding estimate is obtained for the scale of the sample \(Y^{\perp}_{n}\).
+Note that under this strong assumption no corresponding estimate or information is obtained for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).
 
@@
 
-Recently, [,@nixon2023scale] showed that geometric mean of \(Y^{\parallel}_{n}\) was related to the geometric mean of \(W_n\) and so was related to \(W^{\perp}_n\).
-This realization opened the possibility that adding uncertainty to the geometric mean could provide a buffer of uncertainty on it that could be used to control for  the unacknowledged bias that the geometric mean was a perfect point estimate of the underlying system.
-The bias around  could be acknowledged by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated.
-In terms of information theory, the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(I\) (supplemental figure XX), suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.
-This makes intuitive sense because  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim I_n\).
-
-The realization that a strong assumption about scale is built into each denominator explains why HTS data are notoriously fickle as analysis of the same dataset with different workflows, tools and assumptions giving widely different results when benchmarked[@Thorsen:2016aa;@Quinn:2018aa;@Gardner:2019aa].
-This can be understood as the result of unacknowledged bias about the scale estimate which inevitably leads to large Type 1 error rates as sample sizes increase [@Gustafson2015].
-Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be called as differentially abundant even when scale was taken into account, and those features that are differentially abundant only under strictly assumed normalizations.
+Recently, [@nixon2023scale] showed that certain normalizations implied that geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\)  was related to the scale of \( \mathbf{W} \) (i.e., \(\mathbf{W}^{\perp}_n\)).
+From an information theoretic perspective, \(G_n\) is strongly correlated with Shannon's entropy, indicating that some knowledge of the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental).
+Recognition of the relationship between the underlying scale and the \(G_n\) led to the realization that this could introduce latent bias into analyses.
+The possibility of this bias has been unacknowledged until now because the geometric mean was assumed to be a perfect point estimate of the scale of the underlying system.
+This suggested that adding uncertainty to the geometric mean could provide a buffer of uncertainty and acknowledge the bias introduced by the normalization.
 
 The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets.
-ALDEx2 was designed originally to convert point estimates of \(Y^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution, where the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016].
+ALDEx2 was designed originally to convert point estimates of \(\mathbf{Y}^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution, where the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016].
 Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates.
-The CLR calculation uses as the denominator the geometric mean of each Monte-Carlo replicate of \(Y^{\parallel}_{dn}\) with the geometric mean representing a point estimate of scale, i.e.
-of \(Y^{\perp}_{n}\).
-We  now include a posterior distribution of the scale into this process by sampling \(Y^{\perp}_n\) from a log-Normal distribution and using these as the denominator.
+The CLR calculation uses as the denominator the geometric mean of each Monte-Carlo replicate of \(\mathbf{Y}^{\parallel}_{dn}\) with the geometric mean representing a point estimate of scale, i.e.
+of \(\mathbf{Y}^{\perp}_{n}\).
+We  now include a posterior distribution of the scale into this process by sampling \(\mathbf{Y}^{\perp}_n\) from a log-Normal distribution and using these as the denominator.
 The scale estimates can be calculated on a per-experiment, per-group or even a per-sample basis.
 
-An advantage of incorporating scale is that the analysis can be made much more robust and that differences in scale can be accounted for explicitly.
+An advantage of incorporating scale is that  analyses can be made much more robust and that differences in scale can be accounted for explicitly.
 The examples below show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.
@@
 This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab].
-However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [Robinson:2010, Love:2014aa], or on  the clr values[@fernandes:2013] both of which are  log-ratios.
-The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013, @Love:2014aa].
+However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa], or on  the clr values [@fernandes:2013] both of which are  log-ratios.
+The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa].
 
@@
 Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first.
-Panel C shows the equivalent values calculated by ALDEx2.
-The expected clr value for each transcipt are plotted vs.
+Panel C shows the equivalent values calculated by ALDEx2  in which the expected clr value for each transcipt are plotted vs.
 the expected dispersion.
+The red line in each panel shows the loess line of fit to the mid-point of the distributions.
 In panels B and C the amount of dispersion reaches a minimum at moderate  values.
-The red line in each panel shows the loess line of fit to the mid-point of the distributions.
 The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value.
-The dashed grey line in panels B and C represent the maximum of the lowest 10\\% dispersion." }
+The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }
 
@@
 
-Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data[@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution.
+Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data [@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution.
 This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes.
-This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation[@Taniguchi:2010aa].
+This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa].
 Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible.
 To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion.
-The work of [,@nixon2023scale] suggests that this is because the biased estimate of \(Y^{\perp}\) that is  calculated leads to a false estimate of the scale.
+The work of [@nixon2023scale] suggests that this is because the biased estimates of \(G_n\) that are  calculated leads to a false estimate of the scale.
 Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.
 
-Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts.
+Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e.
+`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts are statistically significant.
 Clearly, the necessary assumption of most features being invariant is not justified.
@@
 While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa,@SEQC/MAQC-III-Consortium:2014aa].
+Furthermore, the assumption that housekeeping genes should not generally be included in differential abundance analysis is implicit in the dual p-value, fold-change cutoff approach in widespread use.
 Thus, the use of this prior knowledge is not unique to our approach.

### 2023-08-31 supplement_figs.Rmd
clarified H and G, added bold notation

@@
 
-# GM is correlated with Shannon's entropy and includes
+# GM can correlate with Shannon's entropy
 
-```{r info, echo=F,warning=F, message=F,comment=F, fig.cap="Plot of Shannon's entropy (I) vs geometric mean (G) for each sample in three different datasets.
+We can also think about this relationship from an information theoretic point of view.
+Empirically in most datasets the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(H\) (supplemental figure XX) suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.
+Indeed, the logarithm of the geometric mean \( G \) and \( H \) can be understood from an information theoretic point of view to be an unweighted \( G \) or a weighted measure \( H \) of 'surprisal' in the dataset (supplement).
+Intuitively,  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
+
+Entropy \( H \) and the geometric mean \( G \) (or its logarithm \( \log_2 G \)) are not simple to relate algebraically, but they can be understood in terms of what they are measuring if their description is rephrased in a common language.
+Recall have a \(D \times N\) matrix of counts \(W\) decomposed into the proportions for the \(n^{th}\) sample \(W^{\parallel}_n\) (or the equivalent probability distribution \( p(w_n) \) ), and its scale \(W^{\perp}_n\).
+
+For notational simplicity assume a single discrete random variable \( X \) with a probability distribution \( p(x) \) over \(1 \dots d \) features. The entropy \( H(X) \) in bits is:
+
+\[
+H(X) = -\sum_{i=1}^{d} p_i \log_2 p_i
+\]
+
+and for the same distribution log2 of the geometric mean \( G \) is:
+
+\[
+\log_2 G = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
+\]
+
+As defined here \( H(X) \) is a weighted total of the uncertainty or 'surprisal' contained in \( p(x) \), and relates to the amount of information we would need to have in order to reproduce  \( p(x) \).
+Conversely, \( G \) is an unweighted average of the same distribution.
+However, \( G \) is used as the denominator to calculate the centred log-ratio normalization (clr):
+
+\[
+clr = \log_2(p_i) - \log_2 G 
+\]
+
+over \(i=1 \dots d \).
+This compares each \( p_i \) with the geometric mean \( G \).
+Thus \( G \) can be interpreted as a measure of difference for how far each \( p_i \) is from \( G \), or in information theoretic terms as an unweighted measure of the  mean surprisal for the distribution.
+
+Thus, we can thus understand \( H(X) \) as a measure of the total weighed surprisal and  \( G \)  as a measure of the average unweighted surprisal for \( p(x) \).
+The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.
+
+These two measures are expected to have different behaviours in different distributions of \( p_i \).
+In the case of a uniform distribution both \( H(X) \) and \( G \) are maximal since \( p(x) \) is equally and identically distributed.
+Thus, we expect that they are positively correlated here.
+In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values.
+In very sparse  datasets, the two measures could become uncoupled because \( H(X) \) could ascribe some uncertainty to the large number of low probability events, while \( G \) would tend to be very small.
+Here these two measures could be either uncorrelated or exhibit negative correlation.
+We can see this distributional behaviour in different datasets.
+
+```{r info, echo=F,warning=F, message=F,comment=F, fig.cap="Plot of Shannon's entropy (H) vs geometric mean (G) for each sample in different datasets.
 The groups that each sample belong to are highlighted as filled or open circles.
@@
 yst <- read.table(url, header=T, row.names=1)
-
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/meta16S.tsv"
+rRNA <- read.table(url, header=T, row.names=1, sep='\t')
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
+ss <- read.table(url, header=T, row.names=1, sep='\t')
+ss <- ss[,c(1:100,1501:1601)]
 # remove the one gene with 0 reads
@@
 
-par(mfrow=c(1,1))
-plot(apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),
-  apply(selex+0.5, 2, function(x) mean(log2(x))), xlim=c(2.5,11), ylim=c(-0.5,10), 
-  xlab="Shannon's entropy", ylab='Geometric mean', pch=c(rep(19,7),rep(1,7)),
-  col=rgb(0,0,0,0.5))
-points(apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),apply(yst+0.5, 2, function(x) mean(log2(x))), col=rgb(1,0.6,0,0.5), pch=c(rep(19,44), rep(1, 42)))
-points(apply(ko.both.all+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),apply(ko.both.all+0.5, 2, function(x) mean(log2(x))), col=rgb(1,0,0,0.5),
- pch=c(rep(1,8), rep(19,28), rep(1,8)))
+HS <- apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
+GS <- apply(selex+0.5, 2, function(x) mean(log2(x/sum(x))))
 
-legend(3,10, legend=c('selex', 'transcriptome','metatranscriptome', 'low entropy group', 'high entropy group'),
-  col=c('darkgrey', 'orange', 'red', 'grey','grey'), pch=c(19,19,19,1,19))
+HY <- apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
+GY <- apply(yst+0.5, 2, function(x) mean(log2(x/sum(x))))
+
+HM <- apply(ko.both.all+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
+GM <- apply(ko.both.all+0.5, 2, function(x) mean(log2(x/sum(x))))
+
+H16 <- apply(rRNA+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
+G16 <- apply(rRNA+0.5, 2, function(x) mean(log2(x/sum(x))))
+
+Hss <- apply(ss+0.1, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
+Gss <- apply(ss+0.1, 2, function(x) mean(log2(x/sum(x))))
+
+par(mfrow=c(2,3))
+plot(HS, GS, ylab="Geometric mean", xlab="Shannon's entropy",
+  pch=c(rep(19,7),rep(1,7)), col=rgb(0,0,0,0.5))
+title(main='selex')
+
+plot(HY,GY, 
+  ylab="Geometric mean", xlab="Shannon's entropy", pch=c(rep(19,7),rep(1,7)),
+  col=rgb(1,0.6,0,0.5))
+title(main='transcriptome')
+
+plot(HM, GM, col=rgb(1,0,0,0.5), pch=c(rep(1,8), rep(19,28), rep(1,8)), 
+  ylab="Geometric mean", xlab="Shannon's entropy")
+title(main='meta-transcriptome')
+
+plot(H16,G16, col=rgb(0,1,1,0.5),
+  pch=c(rep(1,198), rep(19,161)), 
+  ylab="Geometric mean", xlab="Shannon's entropy")
+title(main='16S rRNA')
+
+plot(Hss,Gss, col=rgb(0.5,1,0,0.5),
+ pch=c(rep(1,100), rep(19,100)), 
+  ylab="Geometric mean", xlab="Shannon's entropy")
+title(main="single cell")
+
+# empty plot for legend
+plot(selex[,1], selex[,2], xlab=NA, ylab=NA, axes=F,xlim=c(-10,-1), ylim=c(-10,-1))
+legend(-10,-1, legend=c('selex', 'transcriptome','metatranscriptome', '16S', 'SS', 'low entropy group', 'high entropy group'),
+  col=c('darkgrey', 'orange', 'red', 'cyan', 'greenyellow','grey','grey'), pch=c(19,19,19,19,19,1,19))
+
 ```
 
+```{r dist}
+rep.Inf <- function(x){ x[x == -Inf] <- min(x[x > -Inf]) -1 }
+
+# check correlation of H and G for different distributions
+sel.prop <- apply(selex, 2, function(x) x/sum(x))
+yst.prop <- apply(yst, 2, function(x) x/sum(x))
+
+L.yst.prop <- log2(yst.prop)
+L.sel.prop <- log2(sel.prop)
+
+L.yst <- apply(L.yst.prop, 2, rep.Inf)
+
+L.sel <- apply(L.sel.prop, 2, rep.Inf)
+
+```
 ```{r move}

### 2023-09-22 go3.Rmd
pre-comments

@@
 
-One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data from different samples.
+One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data.
 A large number of normalizations were developed.
 Rarefaction[@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a].
-The centered log ratio (CLR)[@aitchison1982] approach was proposed as a unifying normalization[@fernandes:2014].
+The centered log ratio (CLR)[@aitchison1982] the  approach was proposed as a unifying normalization[@fernandes:2014].
 With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.
 
-Recently, [@nixon2023scale] showed that certain normalizations implied that geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\)  was related to the scale of \( \mathbf{W} \) (i.e., \(\mathbf{W}^{\perp}_n\)).
+For the CLR, \(\mathbf{CLR}_{dn} = mathbf{Y}_{dn} / G_n \).
+Recently, [@nixon2023scale] showed that certain normalizations implied that the geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\)  was related to the scale of \( \mathbf{W} \); i.e., it was related to \(\mathbf{W}^{\perp}_n\).
 From an information theoretic perspective, \(G_n\) is strongly correlated with Shannon's entropy, indicating that some knowledge of the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental).
-Recognition of the relationship between the underlying scale and the \(G_n\) led to the realization that this could introduce latent bias into analyses.
-The possibility of this bias has been unacknowledged until now because the geometric mean was assumed to be a perfect point estimate of the scale of the underlying system.
-This suggested that adding uncertainty to the geometric mean could provide a buffer of uncertainty and acknowledge the bias introduced by the normalization.
+Recognition of the relationship between the underlying scale \(\mathbf{W}^{\perp}_n\) and  \(G_n\) led to the realization that the relationship could introduce latent bias into analyses that has been unacknowledged.
+The bias occurrs because the geometric mean, or other denominator, was assumed implicitly to be a perfect point estimate of the scale of the underlying system.
+This suggested that adding uncertainty to the geometric mean could provide a buffer of uncertainty and acknowledge the bias introduced by normalization.
 
 The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets.
-ALDEx2 was designed originally to convert point estimates of \(\mathbf{Y}^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution, where the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016].
+ALDEx2 was designed originally to convert point estimates of \(\mathbf{Y}^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution; the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016].
 Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates.
@@
 
-An advantage of incorporating scale is that  analyses can be made much more robust and that differences in scale can be accounted for explicitly.
+An advantage of incorporating scale is that  analyses can be made much more robust such that actual or potential differences in scale can be accounted for explicitly.
 The examples below show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.
@@
 
-We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (clr).
+We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (CLR).
 The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1.
@@
 Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible.
-To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion.
+To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion.
 The work of [@nixon2023scale] suggests that this is because the biased estimates of \(G_n\) that are  calculated leads to a false estimate of the scale.
@@
 Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e.
-`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts are statistically significant.
+`r length(sig.des)` or `r length(sig.ald)` of the `r nrow(yst)` transcripts are statistically significant.
 Clearly, the necessary assumption of most features being invariant is not justified.
@@
 In the most extreme cases transcripts with near 0 difference have a low FDR.
-This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines.
+This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold [@Schurch:2016aa] (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines.
 A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts.
@@
 The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization.
-While the actual scale of the data is inaccessible post-sequencing  we can estimate the effect of scale on the output.
+While the actual scale of the data is inaccessible post-sequencing  we can estimate the \emph{effect of scale on the output.
 To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013].
 Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`.
-The `ALDEx2` package contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty.
+The `ALDEx2` package further contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty.
 In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.
@@
 We further know that the total number of bacteria is about 10X more in BV than in H [@Zozaya:2010].
-Thus, this is  an extremely challenging environment to determine differential abundance.
+Thus, this is  an extremely challenging environment in which to determine differential abundance.
 Indeed, the accepted method to analyze vaginal metatranscriptomes is to conduct a taxon by taxon analysis rather than a system-wide analysis [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa] because a pooled analysis unexpectedly identifies many housekeeping genes as being differentially abundant between groups.
 
-In this example we show how to specify the scale model explicitly and show that applying different scale models to each group can control for the very large difference in scale in the underlying data.
+In this example we show how to specify the scale model explicitly and demonstrate that applying different scale models to each group independently can control for the very large difference in scale in the underlying data.
 When specifying the whole scale model we can pass a matrix of scale values instead of a single to `aldex.clr()`.
 This matrix should have the same number of rows as the of Monte-Carlo Dirichlet samples, and the same number of columns as the number of samples.
-This matrix encapsulates the additional uncertainty of the scale model on a per-sample basis.
+
+This scale model matrix encapsulates two aspects of scale uncertainty and explicitly illustrates the concept of .
+First, as in the example above, gamma indicates the uncertainty of the de of the scale model on a per-sample basis.
 
@@
 
-load('analysis/xt.all.Rda')
-load('analysis/xg.all.Rda')
-load('analysis/xt.m.all.Rda')
+load('analysis/xt.Rda')
+load('analysis/xg.Rda')
+load('analysis/xt.m.Rda')
 
@@
 Up to this point, scale uncertainty has been applied uniformly to both conditions, but a user-defined scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix.
-This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group.
-Applying a per-group differential scale of 0.15, or 15% along with the base scale now moves the housekeeping functions to the midline of no difference (Figure 3C).
-Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because the change in scale moves the mass of the data towards the group with the highest entropy (supplemental).
-Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied.
+For this we need to revisit the concept that all normalizations in widespread use are actually ratios with the denominator chosen by one of several methods.
+It is important to recognize that in some datasets the denominator itself is not crucial as all approaches give substantially similar results; the transcriptome example above is a good example of such an idealized dataset.
+This begs the question, "if it is not the denominator, what then is the important determinant?" As shown by [@nixon2023scale] it is the \emph{relationship} between the scale values that is important, not their actual value and this can be illustrated quite simply by starting with the mean ratio in \( G\) between groups for the yeast transcriptome dataset which are  \(6.05\times^{-5}\) for snf-w and \(5.08\times^{-5}\); their ratio being 1.17, or a 0.17-fold difference.
+Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 (supplemental).
+Altering this ratio moves the location of the data.
+
+With this background, we can understand how to apply a full scale to the metatranscriptome dataset   with the `aldex.makeScaleMatrix()` function.
+The full scale model option can be used ot make a matrix of relative scales that are distinct for each group (or even each sample) along with their uncertainties.
+Applying a per-group relative differential scale of 0.15 with a base scale of 1 moves the housekeeping functions to the midline of no difference (Figure 3C).
+Note that now a significant number of functions  are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied.
 These former false negatives are noted in orange in each panel.
@@
 Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.
-Thus, while the default scale model alone is useful and should be normally applied, using prior information about the underlying biology to develop a differential scale model can be very useful for many datasets.
+We suggest that the default scale model alone is useful and should be normally applied when the data are well centred.
+However, when datasets are not well centred or when the investigator has prior information about the underlying biology, we advocate developing and using a full differential scale model.
 
@@
 Biological count data can be decomposed into two parts the relative (compositional) and the absolute (scale), and the product of these generates a fully scaled biological system [@nixon2023scale].
-Biological systems are inherently variable and stochastic and current measurement methods that rely on high throughput sequencing fail to capture all of that variation.
+Biological systems are both predictably variable and stochastic and current measurement methods that rely on high throughput sequencing fail to capture all of that variation.
 In the absence of  information external to the sequencing run itself, no normalisation method can recapture any of the scale information [@Loven:2012aa].
@@
 
-All normalizations attempt to make the samples in a dataset commensurate but cannot explicitly address the scale of the underlying system.
+All normalizations attempt to make the samples in a dataset commensurate but do not explicitly address the scale of the underlying system.
 However, the general lack of scale information has important consequences for the analysis of HTS datasets.
 One issue is that analysis tools  seem over-powered with even moderate sample sizes and different tools have different intrinsic power, Type 1 and Type 2 error rates  [@Schurch:2016aa].
-Using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reliable p-values (eg.
+Using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reproducible p-values (eg.
 [@Halsey:2015aa]).
@@
 However, while using the post-hoc difference cutoff is useful for differential abundance analysis it is not clear how this can be incorporated into other kinds of downstream analyses.
-Conversely data that include scale uncertainty are fully compatible with existing downstream analyses.
+Conversely analyses that include scale uncertainty are fully compatible with existing downstream analyses.
 
 In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult to analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
-We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference (supplemental).
+We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference or exhibit other pathologies (supplemental).
 Incorporating differential scale on a per-group basis moves the mass of the data towards the midline of no difference and so affects both Type I and Type II error rates.
-Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant.
 In this analysis, transcripts that were previously not classed as differentially abundant are now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such.
-While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa,@SEQC/MAQC-III-Consortium:2014aa].
+While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa,@SEQC/MAQC-III-Consortium:2014aa].
 Furthermore, the assumption that housekeeping genes should not generally be included in differential abundance analysis is implicit in the dual p-value, fold-change cutoff approach in widespread use.
@@
 
-In summary, while the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and explained.
+In summary, while the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and can help explain some of the underlying biology.
 Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on the analysis weighted accordingly.
@@
 Thus, reporting scale uncertainty should become a standard practice in the analysis of HTS datasets as a way to identify which features are most robust to differences in the underlying system.
-Furthermore, we supply a set of toosl that make incorporating scale simple even for datasets that come from highly asymmetrical environments.
+Finally, we supply a toolkit that makes incorporating scale simple even for datasets that come from highly asymmetrical environments.

### 2023-09-22 supplement_figs.Rmd
pre-comments

@@
 
-# GM can correlate with Shannon's entropy
+# GM correlates with Shannon's entropy in HTS datasets
 
-We can also think about this relationship from an information theoretic point of view.
-Empirically in most datasets the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(H\) (supplemental figure XX) suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.
-Indeed, the logarithm of the geometric mean \( G \) and \( H \) can be understood from an information theoretic point of view to be an unweighted \( G \) or a weighted measure \( H \) of 'surprisal' in the dataset (supplement).
+We can think about the scale relationship from an information theoretic point of view.
+Empirically in most datasets the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Entropy \(H\) (supplemental table and supplemental Figure 1) suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.
 Intuitively,  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
 
+The logarithm of the geometric mean \( \log(G) \) and \( H \) can be understood from an information theoretic point of view to be an unweighted \( G \) or a weighted measure \( H \) of 'surprisal' in the dataset.
 Entropy \( H \) and the geometric mean \( G \) (or its logarithm \( \log_2 G \)) are not simple to relate algebraically, but they can be understood in terms of what they are measuring if their description is rephrased in a common language.
@@
 As defined here \( H(X) \) is a weighted total of the uncertainty or 'surprisal' contained in \( p(x) \), and relates to the amount of information we would need to have in order to reproduce  \( p(x) \).
-Conversely, \( G \) is an unweighted average of the same distribution.
-However, \( G \) is used as the denominator to calculate the centred log-ratio normalization (clr):
+Conversely, \( G \) is an unweighted mean of the same distribution.
+\( G \) is used as the denominator to calculate the centred log-ratio normalization (clr):
 
@@
 over \(i=1 \dots d \).
-This compares each \( p_i \) with the geometric mean \( G \).
+This form shows explicitly that  each \( p_i \) is compared with the geometric mean \( G \).
 Thus \( G \) can be interpreted as a measure of difference for how far each \( p_i \) is from \( G \), or in information theoretic terms as an unweighted measure of the  mean surprisal for the distribution.
@@
 
-load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
+load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
 url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
@@
 ss <- read.table(url, header=T, row.names=1, sep='\t')
-ss <- ss[,c(1:100,1501:1601)]
+ss <- ss[,c(1:100,1502:1601)]
 # remove the one gene with 0 reads
@@
 
-HM <- apply(ko.both.all+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GM <- apply(ko.both.all+0.5, 2, function(x) mean(log2(x/sum(x))))
+HM <- apply(ko.both+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
+GM <- apply(ko.both+0.5, 2, function(x) mean(log2(x/sum(x))))
 
@@
 plot(HY,GY, 
-  ylab="Geometric mean", xlab="Shannon's entropy", pch=c(rep(19,7),rep(1,7)),
+  ylab="Geometric mean", xlab="Shannon's entropy", pch=c(rep(19,44),rep(1,42)),
   col=rgb(1,0.6,0,0.5))
@@
 
-```{r dist}
-rep.Inf <- function(x){ x[x == -Inf] <- min(x[x > -Inf]) -1 }
+The table below summarizes the mean values for, and the correlation between, \( G \) and \( H \)  (cor)  and the sparsity defined as the proportion of features with less than 1 count per sample (spar) for each association in each group of samples:
 
-# check correlation of H and G for different distributions
-sel.prop <- apply(selex, 2, function(x) x/sum(x))
-yst.prop <- apply(yst, 2, function(x) x/sum(x))
+| Dataset | group | \(\overline{G}\) | \( \overline{H} \) | cor | spar |
+|---------|-------|---|---|-----|-------|
+| Selex   | control | -11.2 | 10.2 | 0.99 | 0 |
+| $~~~$"  | selected | -17.8 | 2.8 | -0.88 | 0.802 |
+| yeast   | snf2 ko  | -14.0 | 10.7 | 0.99 | 0.004 |
+| $~~~$"  | WT    | -14.2 | 10.4 | 0.99 | 0.007 |
+| Meta    | H     | -18.8 | 8.6 | 0.78 | 0.451 |
+| $~~~$"  | BV   | -18.2 | 8.9 | 0.79 | 0.238 |
+| 16S     | Pup   | -14.7 | 5.4 | 0.68 | 0.079 |
+| $~~~$"  | Cent | -15.2 | 5.4 | 0.53 | 0.251 |
+| SS      | A     | -13.0 | 8.2 | 0.83 | 0.978 |
+| $~~~$"  | B    | -12.9 | 8.3 | 0.80 | 0.977 |
 
-L.yst.prop <- log2(yst.prop)
-L.sel.prop <- log2(sel.prop)
+We can see that for most datasets the difference between the \(\overline{G}\) in each group is relatively small.
+Most significantly, the selex dataset has a very large difference of about 64 fold, and both the 16S and the metatranscriptome dataset have about a 1.5 fold difference.
+These datasets are candidates for a full scale model correction.
 
-L.yst <- apply(L.yst.prop, 2, rep.Inf)
+```{r full.scale, echo=F}
 
-L.sel <- apply(L.sel.prop, 2, rep.Inf)
+load('analysis/data.out.Rda' )
+
+plot(data.out[,6],data.out[,1], pch=c(rep(1,19),19), ylim=c(-11, 3),
+  xlab='full scale offset', ylab='mean difference between groups')
+points(data.out[,6],data.out[,2], pch=c(rep(1,19),19), col='red')
+points(data.out[,6],data.out[,3], pch=c(rep(1,19),19), col='orange')
+points(data.out[,6],data.out[,4], pch=c(rep(1,19),19), col='cyan')
+points(data.out[,6],data.out[,5], pch=c(rep(1,19),19), col='blue')
+abline(h=0, col='grey', lty=2)
+abline(v=0, col='grey', lty=2)
+legend(0.7, 0, legend=c('16S','selex','tscome','metatscome','ss'),
+  pch=19, col=c('black','red','orange','cyan','blue'))
 
 ```
-```{r move}
-load('analysis/x.s.mu.all.Rda')
-load('analysis/x.all.Rda')
 
-par(mfrow=c(1,2))
-aldex.plot(x.s.mu.all)
-aldex.plot(x.all)
+The full scale model option allows the investigator to set both the offset between the geometric means of the features and their dispersion.
+In the offset plot above we can see the cause and the effect of the full model with a fixed gamma of 0.5.
+We see that the mean location of well centred datasets (yeast transcriptome, single-cell transcriptome and 16S rRNA dataset) are relatively well centred, but could be centred better with small changes in scale ranging from 0 (single cell) to ~0.1 for the 16S and yeast datasets.
+In contrast, centring the metatranscriptome dataset would require about a 50\% change in relative scale between groups, but as shown in the main text, centring the housekeeping genes is more apt.
+
+The in vitro selection dataset is clearly an outlier in both the difference between the average group geometric mean, and in the offset plot.
+However, this dataset can also be used to illustrate the power of the full scale model.
+In Figure 3 we can see that the default output of ALDEx2 has a centered output.
+This occurs largely by chance, as the high sparsity of the selected (S) group is balanced almost exactly by the arbitrarily chosen sequencing depth.
+In this dataset the relative abundances of the majority of features are invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014].
+Neither DESeq nor edgeR are able to provide a reasonable analysis of this dataset [@gloorAJS:2016].
+
+A full scale model, where the strong assumption that the mean \(G\) is assumed to be the same between the two groups, dramatically skews the output and the large number of relatively invariant features are now identified as significantly different.
+While not wrong as long as the assumption that the scales are identical is stated, this is not a useful analysis outcome.
+Modifying the mean scale difference between groups to be 50-fold different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained without the full scale model.
+Note that we get exactly the same answer (within random sampling error) with a scale of 0.02 for group 1 and a scale of 1 for group 2, or using a scale of 1 for group 1 and a scale of 50 for group 2.
+This shows that it is the relative difference between scales that is important, not the absolute values.
+From this result we can conclude that, on average, the difference in underlying scale in the system is about 50-fold, and this is congruent with the circa 64-fold difference in mean \(G\) between groups.
+Thus, an advantage of a full scale model is that we have gained information and understanding about the underlying system.
+
+```{r selex, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters. "}
+
+conds <- c(rep('NS',7), rep('S',7))
+
+x.all <- aldex(selex, conditions=conds, gamma=0.5)
+
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,1), conds)
+x.1.all <- aldex(selex, conditions=conds, gamma=sel.mu)
+
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,120), conds)
+x.5.all <- aldex(selex, conditions=conds, gamma=sel.mu)
+
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(.00833,1), conds)
+x.2.all <- aldex(selex,conditions=conds, gamma=sel.mu)
+
+par(mfrow=c(1,4))
+aldex.plot(x.all, main=('gamma=0.5\nmu=G'))
+aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
+aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,100'))
+aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.01, 1'))
 
 ```
+
+# Issues with DESeq2 and edgeR
+
+DESeq2 and edgeR are two of the most commonly used tools for differential abundance analysis of bule RNA sequencing datasets.
+They both operate by finding a scaling factor that makes all the samples commesurate.
+DESeq2 does this by finding a midpoint feature that can be used as a reference in each sample; this can be different for different samples.
+The edgeR reference finds the midpoint of the 'typical' sample instead.
+In both cases the data are then scaled by dividing by a small factor that makes the read counts commesurate.
+Differential abundance analysis is then performed on the scaled values after taking their logarithm to base 2.
+In some ways this is similar to the log-ratio approach used by ALDEx2, but is more prone to dataset and sample effects than is the log-ratio method [@GloorAJS2023]
+
+```{r DESedg, echo=F, fig.cap="Shown here are the mean log2 fold change as a density plot, and a Volcano plot showing the location and adjusted p-value for each feature in the metatranscriptomic dataset.
+The DESeq2 approach does a good job of centring this data, while edgeR is less suitable.
+The volcano plots show dramatically different outcomes.
+The DESeq2 algorithm assigns very large fold changes to features that have only moderate change, and further identifies a very large proportion of features as significantly different.
+In contrast, edgeR exhibits a much smaller number of differentially abundant features.
+In both volcano plots, the housekeeping genes in the main Figure 3 are shown in orange.
+We can see that these are asymmetrically distributed in both plots.
+Additionally the location of the features taht DESeq2 identified as having a very large difference are shown in the edgeR volcano plot as blue circles."}
+
+# find housekeeping from ALDEx2
+
+load('analysis/xt.m.Rda')
+load('analysis/xt.Rda')
+hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
+
+load('analysis/meta.DES.res.Rda')
+load('analysis/meta.edge.qlf.Rda')
+
+# these are all very rare K0s
+zero.var <-  which(meta.DES.res@listData$lfcSE*sqrt(44) == 0)
+
+weird.des <- which(meta.DES.res@listData$log2FoldChange < -20 )
+
+# housekeeping genes centred with DESeq2 - nice
+
+sig.des <- which(meta.DES.res@listData$padj < 0.01)
+sig.edge <- which(p.adjust(meta.edg.qlf$PValue) < 0.01)
+par(mfrow=c(2,2))
+
+plot(density(meta.DES.res@listData$log2FoldChange[hk.off]), main='DESeq2 FC')
+abline(v=0, col='red', lty=2)
+
+plot(meta.DES.res@listData$log2FoldChange, -1*log10(meta.DES.res@listData$padj), 
+  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)', main='DESeq2 volcano')
+points(meta.DES.res@listData$log2FoldChange[sig.des],-1*log10(meta.DES.res@listData$padj[sig.des] + 1e-70), col='red')
+points(meta.DES.res@listData$log2FoldChange[hk.off],-1*log10(meta.DES.res@listData$padj[hk.off] + 1e-70), col='orange', cex=0.6)
+
+plot(density(meta.edg.qlf$logFC[hk.off]), main='edgeR FC')
+abline(v=0, col='red', lty=2)
+
+plot(meta.edg.qlf$logFC, -1*log10(p.adjust(meta.edg.qlf$PValue)), 
+  col=rgb(0,0,0,0.1), xlab='LFC', ylab='-1log10(p)', main='edgeR volcano')
+points(meta.edg.qlf$logFC[sig.edge], -1*log10(p.adjust(meta.edg.qlf$PValue)[sig.edge]), 
+  col='red')
+points(meta.edg.qlf$logFC[hk.off], -1*log10(p.adjust(meta.edg.qlf$PValue)[hk.off]), 
+  col='orange', cex=0.6)
+  points(meta.edg.qlf$logFC[weird.des], -1*log10(p.adjust(meta.edg.qlf$PValue)[weird.des]), 
+  col='blue', cex=0.6)
+
+```
+
+Examining the plots, edge R clearly not centred with the median housekeeping functions offset by `r median(meta.edg.qlf$logFC[hk.off==T])` and minimum FDR value  is `r min(p.adjust(meta.edg.qlf$PValue)[hk.off==T])`.
+Additionally, there is little range in the p-values relative to those seen for DESeq2, and the values are more in line with the range seen with ALDEx2.
+
+DESeq2 is better centred with median housekeeping functions offset by `r median(meta.DES.res@listData$log2FoldChange[hk.off==T])` but the minimum FDR value  is `r min(meta.DES.res@listData$padj[hk.off==T])`.
+In addition there are a large number of functions with 0 variance, these are very low count functions with very high sparsity in the dataset.
+These are not differential in the DESeq2 analysis.
+However, there are a number of functions in the DESeq2 analysis that are very differentially abundant, with a log2 fold change of < -20.
+Examination of the raw counts shows that these are uniformly 0 or 1 in the H dataset, and present at high counts in some, but not all of the BV dataset.
+That these stand out is odd, since inspection of the count table shows that there are many other functions that are missing from the H dataset, and have higher and more uniform counts in the BV dataset.
+Thus, the importance of the outlier functions seen in the DESeq2 volcano plot should be viewed with suspicion and may be an artefact of the normalization used.
+When overplotted on the edgeR volcano plot (blue), it is clear that these functions have non-signficant p-values.

### 2023-09-22 code/meta-aldex.R

@@ -1,32 +1,35 @@
 devtools::load_all('~/Documents/0_git/ALDEx_bioc')
 
+
 set.seed(2023)
-load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
+load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
 
 # make a vector of conditions
 conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8)) 
 
-xt <- aldex.clr(ko.both.all, conds.K0)
+xt <- aldex.clr(ko.both, conds.K0)
 xt.e <- aldex.effect(xt, include.sample.summary=T)
 xt.t <- aldex.ttest(xt)
 xt.all <- cbind(xt.e, xt.t)
 
-save(xt.all, file='analysis/xt.all.Rda')
+save(xt.all, file='analysis/xt.Rda')
 
 ## single gamma model
-xg <- aldex.clr(ko.both.all, conds.K0, gamma=1)
+xg <- aldex.clr(ko.both, conds.K0, gamma=0.5)
 xg.e <- aldex.effect(xg)
 xg.t <- aldex.ttest(xg)
 xg.all <- cbind(xg.e, xg.t)
 
-save(xg.all, file='analysis/xg.all.Rda')
-
-mu.mod <- aldex.makeScaleMatrix(0.75, .15, conds.K0)
+save(xg.all, file='analysis/xg.Rda')
+mu <- c(1,1.15) # 15% difference in 
+mu.mod <- aldex.makeScaleMatrix(gamma=0.5, mu=mu, conds.K0)
  
-xt.m <- aldex.clr(ko.both.all, conds.K0, gamma=mu.mod)
+xt.m <- aldex.clr(ko.both, conds.K0, gamma=mu.mod)
 xt.m.e <- aldex.effect(xt.m, include.sample.summary=T)
 xt.m.t <- aldex.ttest(xt.m)
 xt.m.all <- cbind(xt.m.e, xt.m.t)
 
-save(xt.m.all, file='analysis/xt.m.all.Rda')
+save(xt.m.all, file='analysis/xt.m.Rda')
+
+
 

### 2023-09-22 code/meta.DESeq.R

@@ -1,0 +1,32 @@
+
+# get the dataset
+load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
+
+conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8)) 
+
+# DESeq2
+library(DESeq2)
+
+# make a vector of conditions
+coldata <- data.frame(conds.K0)
+
+dds <- DESeqDataSetFromMatrix(countData = ko.both.all,
+          colData = coldata, design= ~ conds.K0)
+dds <- DESeq(dds)
+meta.DES.res <- results(dds, name="conds.K0_H_vs_B")
+save(meta.DES.res, file='analysis/meta.DES.res.Rda')
+
+# edgeR
+# qlf is recommended for bulk RNA seq 
+library(edgeR)
+group=factor(conds.K0)
+y <- DGEList(counts=ko.both.all,group=group)
+y <- normLibSizes(y)
+design <- model.matrix(~group)
+y <- estimateDisp(y,design)
+fit <- glmQLFit(y,design)
+qlf <- glmQLFTest(fit,coef=2)
+meta.edg.qlf <- qlf$table
+save(meta.edg.qlf, file='analysis/meta.edge.qlf.Rda')
+
+

### 2023-09-22 code/scale_model.R

@@ -1,0 +1,61 @@
+# mu.vec = c(log2(rep(1,7)), log2(rep(1.2,7)))
+# scale_samples <- t(sapply(mu.vec, FUN = function(mu) rlnorm(128, mu, 1e-3)))
+
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+data(selex)
+
+load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+yst <- yst[rownames(yst) != "YOR072W-B", ]
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/meta16S.tsv"
+rRNA <- read.table(url, header=T, row.names=1, sep='\t')
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
+ss <- read.table(url, header=T, row.names=1, sep='\t')
+ss <- ss[,c(1:100,1502:1601)]
+
+conds <- list()
+conds$rconds <- c(rep(1,198), rep(2,161))
+conds$sconds <- c(rep(1,7), rep(2,7))
+conds$yconds <- c(rep(1,44), rep(2,42))
+conds$mconds <- c(rep(1,8), rep(2,28), rep(1,8))
+conds$ssconds <- c(rep(1,100), rep(2,100))
+
+datasets <- list()
+datasets$rRNA <- rRNA
+datasets$selex <- selex
+datasets$yst <- yst
+datasets$meta <- ko.both
+datasets$ss <- ss
+
+scale <- c(-0.8,-0.7,-0.6,-0.5,-0.4,-0.3,-0.2,-0.1,0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
+
+# samples by column, offsets by row
+# using sweep '+'
+data.out <- matrix(data=NA, ncol=5, nrow=20)
+
+for(i in 1:5){
+  x <- aldex.clr(datasets[[i]], conds[[i]])
+  x.e <- aldex.effect(x, verbose=F)
[…]

### 2023-09-22 code/yst-DESeq.R

@@ -13,11 +13,9 @@ yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL
 
 conds <- c(rep('S', 44), rep('W', 42))
 coldata <- data.frame(conds)
-
 dds <- DESeqDataSetFromMatrix(countData = yst,
           colData = coldata, design= ~ conds)
 dds <- DESeq(dds)
-
 res <- results(dds, name="conds_W_vs_S")
-
 save(res, file='analysis/res.Rda')
+

### 2023-09-22 code/yst-aldex.R

@@ -24,7 +24,7 @@ x.all <- cbind(x.e, x.t)
 save(x.all, file='analysis/x.all.Rda')
 
 set.seed(2023)
-x.s <- aldex.clr(yst, conds, gamma=1, verbose=F)
+x.s <- aldex.clr(yst, conds, gamma=0.3, verbose=F)
 x.s.e <- aldex.effect(x.s, include.sample.summary=T, verbose=F)
 x.s.t <- aldex.ttest(x.s, verbose=F)
 x.s.all <- cbind(x.s.e, x.s.t)

### 2023-09-22 code/yst.aldex.mu.r

@@ -14,7 +14,7 @@ yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL
 conds <- c(rep('S', 44), rep('W', 42))
 coldata <- data.frame(conds)
 
-mu.dat <- aldex.makeScaleMatrix(1,0.25,conds,mc.samples=128)
+mu.dat <- aldex.makeScaleMatrix(1,c(1,1),conds,mc.samples=128)
 
 set.seed(2023)
 x.s.mu.all <- aldex(yst, conds, gamma=mu.dat, verbose=F)

### 2023-10-01 go3.Rmd
ready for comment

@@
 In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken.
-It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment.
+It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment[@Lovell:2015;@Vandeputte:2017aa;@Props:2017aa].
 
@@
 In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa].
-As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance [@Zozaya:2010], and the composition is dramatically different as well [@Ravel:2010;@Hummelen:2010].
-Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but we can access only the composition directly.
+As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010]; the composition is dramatically different as well [@Ravel:2010;@Hummelen:2010].
+Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but current methods access only the compositional information.
 
-More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix, then the elements of the matrix can be decomposed into its composition (relative information) and its scale (total sum).
-We can formally state this as \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\).
+More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix \( \mathbf{W} \), then the elements of the matrix can be decomposed into its composition (relative information, \( \mathbf{W}^{\parallel} \) ) and its scale (total sum, \( \mathbf{W}^{\perp} \) ); formally, \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\).
 If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).
-Any data that we collect by sequencing is thought to contain only proportional information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\).
-Compositional data analysis makes the strong assumption  that \(\mathbf{Y}_{n} = \mathbf{Y}^{\parallel}_{n} = \mathbf{W}^{\parallel}_{n}\); in other words that relative information is the only information available post-sequencing.
-However, the values after sequencing are only estimates of the true underlying values that we want to estimate and different technical and biological replicates vary.
-Note that under this strong assumption no corresponding estimate or information is obtained for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).
+Any data that we collect by sequencing is thought to contain only proportional [@lovell:2011] information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\).
+Compositional data analysis makes the strong assumption  that \(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\); in other words that what is measured by sequencing is relative information only, and that this measurement is a function of the relative information in the underlying environment.
+Note that under this  assumption no corresponding estimate or information is available for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).
 
@@
 Rarefaction[@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a].
-The centered log ratio (CLR)[@aitchison1982] the  approach was proposed as a unifying normalization[@fernandes:2014].
+The centered log ratio (CLR)[@aitchison1982]  approach was proposed as a unifying normalization[@fernandes:2014].
 With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.
 
-For the CLR, \(\mathbf{CLR}_{dn} = mathbf{Y}_{dn} / G_n \).
-Recently, [@nixon2023scale] showed that certain normalizations implied that the geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\)  was related to the scale of \( \mathbf{W} \); i.e., it was related to \(\mathbf{W}^{\perp}_n\).
-From an information theoretic perspective, \(G_n\) is strongly correlated with Shannon's entropy, indicating that some knowledge of the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental).
-Recognition of the relationship between the underlying scale \(\mathbf{W}^{\perp}_n\) and  \(G_n\) led to the realization that the relationship could introduce latent bias into analyses that has been unacknowledged.
-The bias occurrs because the geometric mean, or other denominator, was assumed implicitly to be a perfect point estimate of the scale of the underlying system.
-This suggested that adding uncertainty to the geometric mean could provide a buffer of uncertainty and acknowledge the bias introduced by normalization.
+Recently, Nixon et al.
+[-@nixon2023scale] showed that that *all* normalizations make some  identifying assumption about the scale of \(\mathbf{Y}\) but that the actual assumption being made was not always easy to interpret.
+The CLR which can be written as \( \mathbf{CLR}_{dn} = \mathbf{Y}_{dn} / G_n \) uses the geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the data.
+They discovered that \(G_n\) was related to the scale of the underlying environment \( \mathbf{W\perp} \).
+Interestingly, the information in a system is a fundamental property and \(G_n\) is strongly correlated with Shannon's entropy, indicating that information about the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental).
+The realization that all normalizations made assumptions of scale led to the understanding that all normalizations  introduced latent bias into analyses because the scaling factor has been assumed up to now to be a perfect point estimate of the scale of the underlying system [@nixon2023scale].
+
+With this new insight Nixon et al.
+[-@nixon2023scale] devised methods to set up a scale model that explicitly acknowledges and accounts for the uncertainty of the scale.
+In the case of the CLR which uses the  \(G_n\) as the denominator, this entails adding uncertainty into  \(G_n\) such that the latent bias can be exposed, acknowledged and accounted for.
 
 The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets.
-ALDEx2 was designed originally to convert point estimates of \(\mathbf{Y}^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution; the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016].
+ALDEx2 was designed originally to convert the discrete point estimates in the count matrix obtained into continuous posterior distributions through Monte-Carlo sampling from the Dirichlet distribution.
+The posterior more accurately represents variation on a per-sample basis[@gloorAJS:2016] and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate \(\mathbf{W}^{\parallel}_{dn}\).
 Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates.
-The CLR calculation uses as the denominator the geometric mean of each Monte-Carlo replicate of \(\mathbf{Y}^{\parallel}_{dn}\) with the geometric mean representing a point estimate of scale, i.e.
-of \(\mathbf{Y}^{\perp}_{n}\).
-We  now include a posterior distribution of the scale into this process by sampling \(\mathbf{Y}^{\perp}_n\) from a log-Normal distribution and using these as the denominator.
-The scale estimates can be calculated on a per-experiment, per-group or even a per-sample basis.
+The CLR calculation can be modified trivially to include a full scale model of the data by incorporating uncertainty into \(G\) and so produce a posterior estimate of the scale.
+The scale uncertainty  can be included on a per-experiment, per-group or even a per-sample basis.
+Thus, ALDEx2 can now be used to generate a full posterior model of both \(\mathbf{Y}^{\parallel}_{dn}\) and \(\mathbf{Y}^{\perp}_{dn}\).
 
-An advantage of incorporating scale is that  analyses can be made much more robust such that actual or potential differences in scale can be accounted for explicitly.
-The examples below show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.
+An advantage of incorporating scale is that  analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
+The examples below and in the supplement show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.
 
@@
 Increasing false precision with increasing sample size can be understood as the result of unacknowledged bias brought about because of false certainty as to the scale of the data.
-Thus, by adding scale uncertainty we can be more confident that the analysis is not converging on a precise but biased estimate of scale [@Gustafson2015].
+Thus, by adding scale uncertainty we can be more confident that the analysis is not converging on a precise but biased estimate of scale [@gustafson2015bayesian].
 
@@
 This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab].
-However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa], or on  the clr values [@fernandes:2013] both of which are log-ratios.
+However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa; @Andrews:2019aa], or on  the clr values [@fernandes:2013] both of which are log-ratios.
 The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa].
@@
 To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion.
-The work of [@nixon2023scale] suggests that this is because the biased estimates of \(G_n\) that are  calculated leads to a false estimate of the scale.
-Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.
+The work of Nixon et al.
+[-@nixon2023scale] suggests that this is because the biased scale estimates that are  calculated.
+Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias.
 
@@
 DESeq2 or ALDEx2 were used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset.
-The results were plotted to show the relationship between difference and dispersion (effect plot [@gloor:effect]) or difference and the Benjamini-Hochberg corrected p-values (volcano plot[@Cui:2003aa]).
+The results were plotted to show the relationship between difference and dispersion using effect plots ) or difference and the Benjamini-Hochberg corrected p-values (volcano plot).
 Panels A,B,D,E are for the unscaled analysis, and Panels C,F are for the scaled analysis.
@@
 
-As shown in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2.
+As shown using the effect  [@gloor:effect] and volcano plots [@Cui:2003aa] in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2.
 We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups.
@@
 This results begs the question, "why bother with the signficance test?".
-[XXXX CHECK THE DIRECTION OF SCALE IN THIS DATASET]
 
 The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization.
-While the actual scale of the data is inaccessible post-sequencing  we can estimate the \emph{effect of scale on the output.
+While the actual scale of the data is inaccessible post-sequencing  we can estimate the *effect* of scale on the output.
 To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013].
@@
 
-The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa,@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and BV-affected women.
+The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa;@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
 In this environment, both the relative abundance of species between groups is different as is the gene expression levels within a species [@macklaim:2013].
@@
 Thus, this is  an extremely challenging environment in which to determine differential abundance.
-Indeed, the accepted method to analyze vaginal metatranscriptomes is to conduct a taxon by taxon analysis rather than a system-wide analysis [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa] because a pooled analysis unexpectedly identifies many housekeeping genes as being differentially abundant between groups.
+Indeed, the accepted method to not analyze the vaginal metatranscriptome as a whole system but instead to do the simpler taxon-by-taxon [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa].
+The problem with the systemic analysis is that the scale confounding results in many housekeeping genes as being differentially abundant between groups.
 
 In this example we show how to specify the scale model explicitly and demonstrate that applying different scale models to each group independently can control for the very large difference in scale in the underlying data.
-When specifying the whole scale model we can pass a matrix of scale values instead of a single to `aldex.clr()`.
+When specifying the whole scale model we can pass a matrix of scale values instead of a single estimate of gamma to `aldex.clr()`.
 This matrix should have the same number of rows as the of Monte-Carlo Dirichlet samples, and the same number of columns as the number of samples.
+There is an `aldex.makeScaleModel()` function that can be used to simplify this step.
 
-This scale model matrix encapsulates two aspects of scale uncertainty and explicitly illustrates the concept of .
-First, as in the example above, gamma indicates the uncertainty of the de of the scale model on a per-sample basis.
+This scale model matrix encapsulates two aspects of scale uncertainty and explicitly illustrates the concept of scale.
+First, as in the first example, the gamma paramter indicates the scale uncertainty.
+Second, we need to explicitly indicate the difference in scale between the groups in the dataset.
 
@@
 ```{r meta, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Analysis of vaginal transcriptome data aggregated at the Kegg Orthology (KO) functional level.
-Panel A shows the default analysis with samples from healthy individuals at the top and from BV individuals at the bottom.
+Panel A shows an effect plot for the default analysis where the functions that are elevated in  the  healthy individuals have positive values and functions that are elevated in BV have negative values.
 Highlighed in the box are highly abundant KOs that are almost exlusively housekeeping functions, with ribosomal KOs highlighted in blue,  statistically significant (FDR < 0.01) functions in red, and non-significant functions in black or orange.
-These housekeeping functions are off the midline of no difference.
+These housekeeping functions should be located on the midline of no difference.
 Panel B shows the same data scaled with `gamma = 1`, which increase the minimum dispersion approximatly by one unit.
@@
 For this we need to revisit the concept that all normalizations in widespread use are actually ratios with the denominator chosen by one of several methods.
-It is important to recognize that in some datasets the denominator itself is not crucial as all approaches give substantially similar results; the transcriptome example above is a good example of such an idealized dataset.
-This begs the question, "if it is not the denominator, what then is the important determinant?" As shown by [@nixon2023scale] it is the \emph{relationship} between the scale values that is important, not their actual value and this can be illustrated quite simply by starting with the mean ratio in \( G\) between groups for the yeast transcriptome dataset which are  \(6.05\times^{-5}\) for snf-w and \(5.08\times^{-5}\); their ratio being 1.17, or a 0.17-fold difference.
-Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 (supplemental).
+It is important to recognize that in some datasets the denominator itself is not crucial as all approaches give substantially similar results; the transcriptome example used previously is a good example of such an idealized dataset.
+This begs the question, "if it is not the denominator, what then is the important determinant?" As shown by Nixon et al.
+[-@nixon2023scale]  it is usually the *relationship* between the group scale values (denominators) that is important, not their actual value.
+This can be illustrated quite simply by starting with the mean ratio in \( G\) between groups for the yeast transcriptome dataset which are  \(6.05\times^{-5}\) for snf-w and \(5.08\times^{-5}\); their ratio being 1.17, or a 0.17-fold difference.
+Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 1 as shown in Supplemental Figure 2.
 Altering this ratio moves the location of the data.
@@
 With this background, we can understand how to apply a full scale to the metatranscriptome dataset with the `aldex.makeScaleMatrix()` function.
+This function uses a logNormal distribution when building the model so that ratios that are chosen are symmetrical.
 The full scale model option can be used ot make a matrix of relative scales that are distinct for each group (or even each sample) along with their uncertainties.
@@
 
-Biological count data can be decomposed into two parts the relative (compositional) and the absolute (scale), and the product of these generates a fully scaled biological system [@nixon2023scale].
+Biological count data derived from HTS can be decomposed into two parts the relative (compositional) and the absolute (scale), and the product of these generates a fully scaled biological system [@nixon2023scale].
 Biological systems are both predictably variable and stochastic and current measurement methods that rely on high throughput sequencing fail to capture all of that variation.
@@
 The ALDEx2 R package is ideally suited for this since this tool builds a Bayesian posterior of the compositional component of the dataset at the outset and then conducts the analysis on that posterior.
-Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty.
-For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a log-Normal distribution.
+Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty, and this can be done without additional computational complexity.
+For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a Normal or logNormal distribution depending on preference.
 
@@
 Thus, recommendations to use small sample sizes in multivariate datasets such as RNA-seq datasets are not supported by simple modelling in the univariate case.
-Another issue is that datasets are difficult to analyze when there they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a, @Wu2021].
+Another issue is that datasets are difficult to analyze when there they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a; @Wu2021].
 
@@
 In this analysis, transcripts that were previously not classed as differentially abundant are now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such.
-While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa,@SEQC/MAQC-III-Consortium:2014aa].
+While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
 Furthermore, the assumption that housekeeping genes should not generally be included in differential abundance analysis is implicit in the dual p-value, fold-change cutoff approach in widespread use.

### 2023-10-01 supplement_figs.Rmd
ready for comment

@@
 
-# GM correlates with Shannon's entropy in HTS datasets
+# GM is related to Information and Shannon's entropy in HTS datasets
 
-We can think about the scale relationship from an information theoretic point of view.
-Empirically in most datasets the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Entropy \(H\) (supplemental table and supplemental Figure 1) suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.
-Intuitively,  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
+Information content is a fundamental property of all systems.
+We can think about the scale relationship from an information theoretic point of view which measures how much information is encoded in a particular sample.
+Intuitively,   systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
+Below, we show that the scale of a system as defined by Nixon et al.
+[-@nixon2023scale] is related to the information content and entropy of the data.
 
-The logarithm of the geometric mean \( \log(G) \) and \( H \) can be understood from an information theoretic point of view to be an unweighted \( G \) or a weighted measure \( H \) of 'surprisal' in the dataset.
-Entropy \( H \) and the geometric mean \( G \) (or its logarithm \( \log_2 G \)) are not simple to relate algebraically, but they can be understood in terms of what they are measuring if their description is rephrased in a common language.
-Recall have a \(D \times N\) matrix of counts \(W\) decomposed into the proportions for the \(n^{th}\) sample \(W^{\parallel}_n\) (or the equivalent probability distribution \( p(w_n) \) ), and its scale \(W^{\perp}_n\).
+Recall the underlying system is described by a \(D \times N\) matrix of counts \(\mathbf{W}\) decomposed into the proportions for the \(n^{th}\) sample \(\mathbf{W}^{\parallel}_n\) (or the equivalent probability distribution \( \mathbf{p}(w_n) \) ), and its scale \(\mathbf{W}^{\perp}_n\).
+Sequencing  measures which is related to the underlying proportion; i.e., \(\mathbf{Y}^{\parallel}_n \sim \mathbf{W}^{\parallel}_n\)
 
-For notational simplicity assume a single discrete random variable \( X \) with a probability distribution \( p(x) \) over \(1 \dots d \) features. The entropy \( H(X) \) in bits is:
+For notational simplicity assume we will use a single discrete random variable to represent the probability distribution; i.e.
+\( X =  \mathbf{p}(w_n) \)  over \(1 \dots d \) features.
+The geometric mean can be calculated as:
 
 \[
-H(X) = -\sum_{i=1}^{d} p_i \log_2 p_i
+\log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
+\];
+
+here every element in \(log_2 G(X)\) is the logarithm of the elemental probability \( log_2(p_i)\).
+
+In information theory, the elemental amount of information or uncertainty for \(p_i\) is the inverse \(I(X)_i = -log_2(p_i)\)[@rezaInfoIntro].
+The total entropy of the system  \( H(x) \) is the weighted sum of the elemental information
+
+\[
+H(x) = -\sum_{i=1}^{d} p_i \log_2 p_i
 \]
 
-and for the same distribution log2 of the geometric mean \( G \) is:
+In this formulation, \( \log_2(G(X)) \) and \( H(X) \) can be understood from an information theoretic point of view to be an unweighted and a weighted measure  of 'surprisal' in the dataset.
+The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.
+\( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \).
+Conversely, \( G \) is an unweighted mean of the same distribution and  \( G \) is the denominator to calculate the centred log-ratio normalization (CLR).
 
-\[
-\log_2 G = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
-\]
-
-As defined here \( H(X) \) is a weighted total of the uncertainty or 'surprisal' contained in \( p(x) \), and relates to the amount of information we would need to have in order to reproduce  \( p(x) \).
-Conversely, \( G \) is an unweighted mean of the same distribution.
-\( G \) is used as the denominator to calculate the centred log-ratio normalization (clr):
-
-\[
-clr = \log_2(p_i) - \log_2 G 
-\]
-
-over \(i=1 \dots d \).
-This form shows explicitly that  each \( p_i \) is compared with the geometric mean \( G \).
-Thus \( G \) can be interpreted as a measure of difference for how far each \( p_i \) is from \( G \), or in information theoretic terms as an unweighted measure of the  mean surprisal for the distribution.
-
-Thus, we can thus understand \( H(X) \) as a measure of the total weighed surprisal and  \( G \)  as a measure of the average unweighted surprisal for \( p(x) \).
-The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.
-
-These two measures are expected to have different behaviours in different distributions of \( p_i \).
-In the case of a uniform distribution both \( H(X) \) and \( G \) are maximal since \( p(x) \) is equally and identically distributed.
+Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(Y^{\parallel}_{n})\) (Supplemental Figure 1 and Table 1).
+These two measures are expected to have different behaviours with different distributions of \( p_i \).
+In the case of a uniform distribution both \( H(Y^{\parallel}_{n}) \) and \( G(Y^{\parallel}_{n}) \) are maximal since \( p(x) \) is equally and identically distributed.
 Thus, we expect that they are positively correlated here.
 In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values.
-In very sparse  datasets, the two measures could become uncoupled because \( H(X) \) could ascribe some uncertainty to the large number of low probability events, while \( G \) would tend to be very small.
+In very sparse  datasets, the two measures could become uncoupled because \( H(Y^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(Y^{\parallel}_{n}) \) would tend to be very small.
 Here these two measures could be either uncorrelated or exhibit negative correlation.
@@
 We can see that for most datasets the difference between the \(\overline{G}\) in each group is relatively small.
-Most significantly, the selex dataset has a very large difference of about 64 fold, and both the 16S and the metatranscriptome dataset have about a 1.5 fold difference.
-These datasets are candidates for a full scale model correction.
+Most significantly, the selex dataset has a very large difference of about 100-fold, and both the 16S and the metatranscriptome dataset have about a 1.5 fold difference.
+These three datasets are candidates for a full scale model correction.
 
-```{r full.scale, echo=F}
+```{r full.scale, echo=F, fig.cap="Plot of the offset of the mean differnece between groups as a function of scale ratio.
+For this, the default scale of 1:1 was altered in increments of 0.1 keeping the gamma parameter (dispersion) at 0.5.
+The filled circle shows the outcome when the calculation is done using the geometric mean and the same gamma parameter."}
 
@@
 
-plot(data.out[,6],data.out[,1], pch=c(rep(1,19),19), ylim=c(-11, 3),
+plot(data.out[,6],data.out[,1], pch=c(rep(1,19),19), ylim=c(-11, 4),
   xlab='full scale offset', ylab='mean difference between groups')
@@
 
-The full scale model option allows the investigator to set both the offset between the geometric means of the features and their dispersion.
-In the offset plot above we can see the cause and the effect of the full model with a fixed gamma of 0.5.
-We see that the mean location of well centred datasets (yeast transcriptome, single-cell transcriptome and 16S rRNA dataset) are relatively well centred, but could be centred better with small changes in scale ranging from 0 (single cell) to ~0.1 for the 16S and yeast datasets.
-In contrast, centring the metatranscriptome dataset would require about a 50\% change in relative scale between groups, but as shown in the main text, centring the housekeeping genes is more apt.
+Nixon et al.
+[-nixon2023scale] showed that many operations on HTS datasets relied on both the proportion and scale components of the data.
+Moreover, all normalizations impose a scale model on the data, but the appropriateness of these models has never been explicitly acknowledged or tested.
+Thus, the full scale model option allows the investigator to set both the offset between the geometric means of the features and their dispersion and observe how this affects the analysis outcome.
+
+In the offset plot above we can see the cause and the effect of the full model with a fixed gamma of 0.5 and a base scale of 1 in each group.
+We see that the mean location of well centred datasets (yeast transcriptome, single-cell transcriptome) are close to 0, but could be centred better with small changes in scale ranging from 0 (single cell) to 1:1.1 for the yeast transcriptome dataset.
+In contrast, centring the 16S dataset requires about a 1:1.3 fold change.
+The metatranscriptome dataset would require about a 0.5:1 change in relative scale between groups, but as shown in the main text, centring the housekeeping genes is more apt.
 
 The in vitro selection dataset is clearly an outlier in both the difference between the average group geometric mean, and in the offset plot.
-However, this dataset can also be used to illustrate the power of the full scale model.
+However, this dataset can  be used to illustrate the power of the full scale model.
 In Figure 3 we can see that the default output of ALDEx2 has a centered output.
-This occurs largely by chance, as the high sparsity of the selected (S) group is balanced almost exactly by the arbitrarily chosen sequencing depth.
-In this dataset the relative abundances of the majority of features are invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014].
-Neither DESeq nor edgeR are able to provide a reasonable analysis of this dataset [@gloorAJS:2016].
+This occurs largely by chance, as the high sparsity of the selected (S) group is balanced almost exactly by the arbitrarily chosen sequencing depth so the non-selected group (NS) appears to have a similar location as the S group.
+The difference in entropy between the two groups and the differences in geometric mean are very large, with the difference in \(\log_2{\overline{G}(S)}\) and \(log_2{\overline{G}(NS)}\) being about \(2^{6.6}\).
+Setting the scale of both the S and NS groups to 1 we find that the difference in location is approximately \(2^{7.7}\) in close agreement with the difference the geometric means.
+For this dataset to be centered we need to have a scale ratio \( \approx \) 1:50 or more.
+In fact, in this dataset the relative abundances of the majority of features are nearly invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014], thus changing the scale of the data.
+Neither DESeq nor edgeR are able to provide a reasonable analysis of this dataset because the normalizations used assume equivalent scales [@gloorAJS:2016].
 
-A full scale model, where the strong assumption that the mean \(G\) is assumed to be the same between the two groups, dramatically skews the output and the large number of relatively invariant features are now identified as significantly different.
+Figure 3 shows an effect plot of various scale models with this dataset.
+The full scale model, where the strong assumption that the mean \(G\) is assumed to be 1:1 between the two groups, dramatically skews the output and the large number of relatively invariant features are now identified as significantly different.
 While not wrong as long as the assumption that the scales are identical is stated, this is not a useful analysis outcome.
-Modifying the mean scale difference between groups to be 50-fold different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained without the full scale model.
+Modifying the mean scale difference between groups to be \( \approx \) 1:50 different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained using \(G_n\) as the scale estimate.
 Note that we get exactly the same answer (within random sampling error) with a scale of 0.02 for group 1 and a scale of 1 for group 2, or using a scale of 1 for group 1 and a scale of 50 for group 2.
-This shows that it is the relative difference between scales that is important, not the absolute values.
-From this result we can conclude that, on average, the difference in underlying scale in the system is about 50-fold, and this is congruent with the circa 64-fold difference in mean \(G\) between groups.
-Thus, an advantage of a full scale model is that we have gained information and understanding about the underlying system.
+This shows that it is the relative difference between scales that is important in this dataset, not the absolute values.
+From this result we can conclude that, on average, the difference in underlying scale in the system is about \(\approx 50\)-fold, and this is congruent with the circa 100-fold difference in  \(\overline{G}\) between groups; the discrepancy being explained because the default scale model is applied uniformly to all samples, whereas the different values of the within-group geometric means ranging over a \{ > 2.5 \) fold range.
+Thus, an advantage of a full scale model is that we have gained both information and understanding about the drivers of asymmetry underlying system.
 
-```{r selex, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters. "}
+```{r selex, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters.
+All scales are calculated with a logNormal distribution to ensure symmetry for the user.
+"}
 
@@
 
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,1), conds)
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
 x.1.all <- aldex(selex, conditions=conds, gamma=sel.mu)
 
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,120), conds)
+# setting ratio as 1:50 explicitly
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
 x.5.all <- aldex(selex, conditions=conds, gamma=sel.mu)
 
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(.00833,1), conds)
+# setting ratio in log space
+# 2^-5.64 : 2^0  = 0.02 : 1
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(-5.64,0), conds, log=TRUE)
 x.2.all <- aldex(selex,conditions=conds, gamma=sel.mu)
@@
 aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
-aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,100'))
-aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.01, 1'))
-
+aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,50'))
+aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.02, 1'))
 ```
@@
 Examining the plots, edge R clearly not centred with the median housekeeping functions offset by `r median(meta.edg.qlf$logFC[hk.off==T])` and minimum FDR value  is `r min(p.adjust(meta.edg.qlf$PValue)[hk.off==T])`.
-Additionally, there is little range in the p-values relative to those seen for DESeq2, and the values are more in line with the range seen with ALDEx2.
+Additionally, as shown in Figure 4there is little range in the p-values relative to those seen for DESeq2, and the values are more in line with the range seen with ALDEx2.
 

### 2023-10-01 code/scale_model.R

@@ -46,7 +46,7 @@ for(i in 1:5){
   data.out[20,i] <- mean(x.e$diff.btw)
 for(j in 1:19){
   mu <- c(1, 1+scale[j])
-  g.scale <- aldex.makeScaleMatrix(0.5, mu, conds[[i]])
+  g.scale <- aldex.makeScaleMatrix(0.5, mu, conds[[i]], log=FALSE)
   x <- aldex.clr(datasets[[i]], conds[[i]], gamma=g.scale, verbose=F)
   x.e <- aldex.effect(x)
   data.out[j,i] <- mean(x.e$diff.btw)

### 2023-10-01 code/yst-aldex.R

@@ -24,7 +24,7 @@ x.all <- cbind(x.e, x.t)
 save(x.all, file='analysis/x.all.Rda')
 
 set.seed(2023)
-x.s <- aldex.clr(yst, conds, gamma=0.3, verbose=F)
+x.s <- aldex.clr(yst, conds, gamma=0.5, verbose=F)
 x.s.e <- aldex.effect(x.s, include.sample.summary=T, verbose=F)
 x.s.t <- aldex.ttest(x.s, verbose=F)
 x.s.all <- cbind(x.s.e, x.s.t)

### 2023-10-01 code/yst.aldex.mu.r

@@ -14,7 +14,7 @@ yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL
 conds <- c(rep('S', 44), rep('W', 42))
 coldata <- data.frame(conds)
 
-mu.dat <- aldex.makeScaleMatrix(1,c(1,1),conds,mc.samples=128)
+mu.dat <- aldex.makeScaleMatrix(0.5,c(1,1.04),conds,mc.samples=128)
 
 set.seed(2023)
 x.s.mu.all <- aldex(yst, conds, gamma=mu.dat, verbose=F)

### 2023-10-17 go3-author.tex
sent for comment

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Beyond compositionally in high throughput sequencing; estimating
the importance of scale in data analysis with ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}
\date{\today}

\begin{document}
\maketitle

$^1$ Department of Biochemistry, University of Western Ontario
$^2$ College of Information Sciences and Technology, Pennsylvania  State University
$^3$ Department of Medicine, Pennsylvania  State University

\begin{abstract}
Despite scale being an important confounding variable, high throughput
sequencing generates data that are difficult to reconcile with the scale
of the underlying environment, but it has been shown that the scale of
the environment is related to the geometric mean of the parts of the
sample. The ALDEx2 R package builds a Bayesian model of the data and
here we report an update to the model that includes both random sampling
and scale estimates. We show that incorporating scale allows for robust
inference in a number of different experimental designs and we provide
guidance on how to incorporate scale inference into differential
abundance analysis. While all data normalizations in widespread use make
implicit assumptions about the scale of the system we suggest that
explicit rather than implicit scale models should be used.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

High throughput sequencing (HTS) is a universal tool to explore many
biological phenomenon such as gene expression (single-cell sequencing,
RNA-sequencing, meta-transcriptomics), microbial community composition
(16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme
activity (selex, CRISPR killing). HTS proceeds by taking a fixed-number
sample from the environment, making a library, multiplexing (merging)
multiple libraries and applying a fixed number of molecules to the flow
cell. In essence it is a poll of the environment that is mixed with
other polls and then a poll of the mixture is taken. It should be clear
that the total number of molecules sequenced is driven by the capacity
of the instrument and not by the number of molecules in the sampled
environment(Lovell et al. 2015; Vandeputte et al. 2017; Props et al.
2017) .

Data that are generated by sequencing come from systems where scale is
usually important and may be a confounding variable (Lovell et al.
2015). For example, cells transformed by the cMyc oncogene have about 3
times the amount of mRNA and about twice the rRNA content than do
non-transformed cells (Nie et al. 2012), and this dramatically skews
transcriptome analysis (Lovén et al. 2012). In addition, wild-type and
mutant strains of cell lines, yeast or bacteria often have different
growth rates, which would affect our ability to identify truly
differentially abundant genes (Yoshikawa et al. 2011). As another
example, the total bacterial load of the vaginal microbiome differs by
1-2 orders of magnitude in absolute abundance between the healthy and
bacterial vaginosis states (Zozaya-Hinchliffe et al. 2010); the
composition is dramatically different as well (Ravel et al. 2011;
Hummelen et al. 2010). Thus, the full description of these systems
includes both relative change (composition) and absolute abundance
(scale) but current methods access only the compositional information.

More formally, if the true information in the environment being sampled
can be summarized by counts of features (genes, taxa, etc) in a matrix
\(\mathbf{W}\), then the elements of the matrix can be decomposed into
its composition (relative information, \(\mathbf{W}^{\parallel}\) ) and
its scale (total sum, \(\mathbf{W}^{\perp}\) ); formally,
\(\mathbf{W} = \mathbf{W}^{\parallel} \mathbf{W}^{\perp}\). If the
matrix has \(N\) samples and \(D\) parts, then we can uniquely identify
\(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\). Any data
that we collect by sequencing is thought to contain only proportional
(Lovell et al. 2011) information and we can represent the corresponding
sequenced matrix as \(\mathbf{Y}\). Compositional data analysis makes
the strong assumption that
\(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\);
in other words that what is measured by sequencing is relative
information only, and that this measurement is a function of the
relative information in the underlying environment. Note that under this
assumption no corresponding estimate or information is available for the
scale of the sample \(\mathbf{Y}^{\perp}_{n}\).

One issue that was realized very early was that if the goal was to
compare one sample to another then the output from HTS needed to be
normalized in order to make the samples commensurate and to correct any
minor asymmetries in the data. A large number of normalizations were
developed. Rarefaction(Hughes and Hellmann 2005) was used in the
microbiome field, while transcriptome investigators developed and used
proportions and derivatives (reads per kilobase per million, and
transcripts per kilobase per million(Mortazavi et al. 2008; Wagner, Kin,
and Lynch 2012)), the relative log expression (RLE)(Anders and Huber
2010) and trimmed mean of M values (TMM)(Robinson and Oshlack 2010). The
centered log ratio (CLR)(Aitchison 1982) approach was proposed as a
unifying normalization(Fernandes et al. 2014). With the exception of
rarefaction, all of these normalizations are ratios with the major
differences between approaches being how the denominator is chosen and
whether the ratio is always log transformed or not.

Recently, Nixon et al. (2023) showed that that \emph{all} normalizations
make some identifying assumption about the scale of \(\mathbf{W}\) but
that the actual assumption being made was not always easy to interpret.
In essence, normalizations in widespread use assume that either one
sample can be chosen as a reference to which the others are scaled
(TMM), or that different sub-parts of each sample maintain a constant
scale across samples (RLE, CSS, CLR and derivatives).

The CLR, which can be written as
\(\mathbf{CLR}_{(1...d)n} = \mathbf{Y}_{(1...d)n} / G_n\), uses the
geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the
data. Nixon et al. (2023) discovered that using the CLR was equivalent
to making a statement that \(G_n\) was related to the scale of the
underlying environment \(\mathbf{W}^{\perp}\). Interestingly, the
information in a system is a fundamental property and \(G_n\) is
strongly correlated with Shannon's entropy, indicating that information
about the complexity of \(\mathbf{W}\) is available post-sequencing
(Supplemental). The realization that all normalizations made assumptions
of scale led to the understanding that all normalizations introduced
latent bias into analyses because the scaling factor has been assumed up
to now to be a perfect point estimate of the scale of the underlying
system (Nixon et al. 2023).

With this new insight Nixon et al. (2023) devised methods to set up a
scale model that explicitly acknowledges and accounts for the
uncertainty of the scale. In the case of the CLR which uses the \(G_n\)
as the denominator, this entails adding uncertainty into \(G_n\) such
that the latent bias can be exposed, acknowledged and accounted for.
Additionally it was realized that the actual values used in the
denominator to calculate the log-ratio were irrelevant, but instead that
the ratios between the denominators was the key parameter. These two
insights allow for a full scale model of scale to be built and tested
for its effect on analysis.

The ALDEx2 R package (Fernandes et al. 2013) represents a general
purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was
designed originally to convert the discrete point estimates in the count
matrix obtained into continuous posterior distributions through
Monte-Carlo sampling from the Dirichlet distribution. The posterior more
accurately represents variation on a per-sample basis(Gloor et al. 2016)
and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate
\(\mathbf{W}^{\parallel}_{dn}\). Each Monte-Carlo replicate is
normalized using the CLR and used to calculate summary and test
statistics, and finally expected values and confidence intervals are
reported across the Monte-Carlo replicates. The CLR calculation can be
modified trivially to include a full scale model of the data by
incorporating uncertainty into \(G\) and so produce a posterior estimate
of the scale. The scale uncertainty can be included on a per-experiment,
per-group or even a per-sample basis. Thus, ALDEx2 can now be used to
generate a full posterior model of both \(\mathbf{W}^{\parallel}_{dn}\)
and \(\mathbf{W}^{\perp}_{dn}\) from the observed data.

An advantage of incorporating scale is that analyses can be made much
more robust such that actual or potential differences in scale can be
tested and accounted for explicitly. The examples below and in the
supplement show how incorporating scale provides robust and
interpretable differential abundance estimates in several different
datasets.

\hypertarget{results}{%
\section{Results}\label{results}}

The first dataset is a highly replicated yeast transcriptome where one
condition is wild-type and the other has a snf-1 gene
knockout(Gierliński et al. 2015). Yeast deficient for snf-1 grow more
slowly and are sensitive to a variety of common agents that cause cell
stress (Yoshikawa et al. 2011). This dataset has been used to argue that
only a small number of replicates need to be used to identity
differentially abundant genes and that different tools should be used
for datasets with different sample sizes because the tools have
different intrinsic statistical power and Type 1 and Type 2 errors
(Schurch et al. 2016). This guidance runs counter to standard
statistical practice where power is intrinsically linked to sample size
(Halsey et al. 2015), yet the concept of sample-size independent power
is entrenched in all fields that use HTS as an experimental readout.
Increasing false precision with increasing sample size can be understood
as the result of unacknowledged bias brought about because of false
certainty as to the scale of the data. Thus, by adding scale uncertainty
we can be more confident that the analysis is not converging on a
precise but biased estimate of scale (Gustafson 2015).

We start by examining the dispersion or variance of the data as counts
and as the logarithm of the normalized counts as calculated by DESeq2
(log(RLE)) and by ALDEx2 (CLR). The actual counts of data derived from
sequencing are overdispersed with the mean value being less than the
variance (Robinson, McCarthy, and Smyth 2010) as seen in Panel A of
Figure 1. This relationship is why most tools model the counts with the
Negative Binomial distribution(Frazee et al. 2023), and use this
distribution as the basis for batch correction(Zhang, Parmigiani, and
Johnson 2020). However, the actual analysis of differential abundance is
performed on the logarithm of the normalized counts (Robinson, McCarthy,
and Smyth 2010; Love, Huber, and Anders 2014; Andrews and Hemberg 2019),
or on the clr values (Fernandes et al. 2013) both of which are
log-ratios. The mean-dispersion distribution of these log-ratio
transformed data is quite different as shown in panels B and C of Figure
1 and as noted elsewhere (Fernandes et al. 2013; Love, Huber, and Anders
2014).

\begin{figure}
\centering
\includegraphics{go3_files/figure-latex/dispersion-1.pdf}
\caption{Plot of abundance v dispersion for a typical transcriptome
dataset as counts, as logarithms of counts, and as clr values. Panel A
shows that the data are over-dispersed relative to a Poisson
distribution which is represented by the dashed line when plotted on a
log-log scale. Panel B shows that the relationship between the mean and
the dispersion calculated in DESeq2, here the standard error (SE) of the
mean, is very different when the data are log-transformed first. Panel C
shows the equivalent values calculated by ALDEx2 in which the expected
clr value for each transcipt are plotted vs.~the expected dispersion.
The red line in each panel shows the loess line of fit to the mid-point
of the distributions. In panels B and C the amount of dispersion reaches
a minimum at moderate values. The dashed orange line in panel A is the
line of equivalence, and in panel B and C is the minimum y value. The
values below the dashed grey line in panels B and C represent those
below the first decile of dispersion.}
\end{figure}

Here we can see that while the variance or dispersion always increases
with increasing raw read count, it actually decreases when measured on
the log-ratio transformed data (Fernandes et al. 2013; Love, Huber, and
Anders 2014) and reaches a minimum at some mid-point of the
distribution. This makes the counter-intuitive suggestion that genes
with moderate expression have more predictable expression than genes
with very high expression such as housekeeping genes. This is at odds
with the known biology of cells where single cell counting of
housekeeping transcripts shows that they are both highly expressed and
have little intrinsic variation (Taniguchi et al. 2010). Furthermore,
the actual amount of dispersion is very small and for many transcripts
is almost negligible. To show this point more clearly, the majority of
the transcripts in the lowest decile of dispersion indicated below the
dashed grey line are statistically significantly different (75\% with
DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead
to many false positives because of the underestimation of dispersion.
The work of Nixon et al. (2023) suggests that this is because the biased
scale estimates that are calculated. Thus, the actual variation of
highly expressed genes is not captured accurately by current approaches
to analysis and supports the idea that HTS have unacknowledged bias.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts
are statistically significantly different between groups even with a
Benjamini-Hochberg (Benjamini and Hochberg 1995) false discovery rate
(FDR) of 0.01; i.e.~4264 or 3790 of the 5891 transcripts are
statistically significant. Clearly, the necessary assumption of most
features being invariant is not justified. Applying the rule of thumb of
at least a \(2^{1.4}\) fold change (Schurch et al. 2016) reduces these
outputs to 193 for DESeq2 and to 186 for ALDEx2.

\begin{figure}
\centering
\includegraphics{go3_files/figure-latex/plot1-1.pdf}
\caption{Effect and volcano plots for unscaled and scaled transcriptome
analysis. DESeq2 or ALDEx2 were used to conduct a differential abundance
(DA) analysis on the yeast transcriptome dataset. The results were
plotted to show the relationship between difference and dispersion using
effect plots ) or difference and the Benjamini-Hochberg corrected
p-values (volcano plot). Panels A,B,D,E are for the unscaled analysis,
and Panels C,F are for the scaled analysis. Each point represents the
values for one transcript, with the color indicating if that transcript
was significant in the scaled analysis and unscaled analysis (red) or in
the unscaled analysis only (orange). Points in grey are not
statistically signficantly different with any analysis. The horizontal
dashed lines represent a log2(difference) of 1.4, which is a commonly
applied cutoff when the majority of features are statistically
significant.}
\end{figure}

As shown using the effect (Gloor, Macklaim, and Fernandes 2016) and
volcano plots (Cui and Churchill 2003) in Figure 2 A,B,D,E the root
cause of the many statistically significant positive transcripts is the
very large number of transcripts with negligible variance with both
DESeq2 and ALDEx2. We can see that almost all the transcripts that are
differentially abundant with an FDR \textless{} 0.01 (orange and red
points) have extremely low dispersion and a very low difference between
groups. In the most extreme cases transcripts with near 0 difference
have a low FDR. This issue lead to the common practice of choosing
transcripts with a low FDR and a fold-change threshold (Schurch et al.
2016) (commonly set at \(\pm 2^{1.4}\)), and these limits are shown by
the dashed grey lines. A similar sitation arises when using the ALDEx2
package, and indeed the two methods identify substantially similar
transcripts. This results begs the question, ``why bother with the
signficance test?''.

The very low dispersion estimate for most of the features arises because
scale variation in the underlying dataset has been removed through
sequencing and normalization. While the actual scale of the data is
inaccessible post-sequencing we can estimate the \emph{effect} of scale
on the output. To do this, we add uncertainty to the denominator that is
used to calculate the log-ratio of the samples (Nixon et al. 2023), and
combine this with the posterior probabilities that ALDEx2 calculates on
a per-feature basis (Fernandes et al. 2013). Scale uncertainty is
incorporated using the \texttt{gamma} parameter that controls the amount
uncertainty being included when we call either \texttt{aldex()}, or
\texttt{aldex.clr()}. The \texttt{ALDEx2} package further contains a
sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be
used to explore the effect of different amounts of scale uncertainty. In
practice we suggest that a \texttt{gamma} parameter between 0.5 and 1 is
realistic for most experimental designs.

Applying \texttt{gamma=1} as a parameter we can see that the large
number of transcripts with near 0 dispersion have had their dispersion
increased (Figure 2C), and this results in many fewer transcripts (100)
being called significantly different as shown in the volcano plot in
Figure 1F (red points). Furthermore, overplotting the significant
transcripts identified after adding scale uncertainty on the un-scaled
analysis shows that adding scale uncertainty removes the need for the
dual cutoff. Indeed, adding scale uncertainty reduces the significant
transcripts to approximately the number observed with the somewhat
arbitrary difference cutoff. Thus, incorporating scale uncertainty
through the default scale model allows us to determine which variables
are likely to be significant due to sequencing and normalization, and
which are significantly different even with scale uncertainty included.

The second example dataset is a vaginal metatranscriptome dataset used
in (Macklaim and Gloor 2018; Wu et al. 2021), where we are comparing
gene expression in bacteria collected from healthy (H) and bacterial
vaginosis (BV) affected women. In this environment, both the relative
abundance of species between groups is different as is the gene
expression levels within a species (Macklaim et al. 2013). We further
know that the total number of bacteria is about 10X more in BV than in H
(Zozaya-Hinchliffe et al. 2010). Thus, this is an extremely challenging
environment in which to determine differential abundance. Indeed, the
accepted method to not analyze the vaginal metatranscriptome as a whole
system but instead to do the simpler taxon-by-taxon (Macklaim et al.
2013; Deng et al. 2018; Fettweis et al. 2019). The problem with the
systemic analysis is that the scale confounding results in many
housekeeping genes as being differentially abundant between groups.

In this example we show how to specify the scale model explicitly and
demonstrate that applying different scale models to each group
independently can control for the very large difference in scale in the
underlying data. When specifying the whole scale model we can pass a
matrix of scale values instead of a single estimate of gamma to
\texttt{aldex.clr()}. This matrix should have the same number of rows as
the of Monte-Carlo Dirichlet samples, and the same number of columns as
the number of samples. There is an \texttt{aldex.makeScaleModel()}
function that can be used to simplify this step.

This scale model matrix encapsulates two aspects of scale uncertainty
and explicitly illustrates the concept of scale. First, as in the first
example, the gamma paramter indicates the scale uncertainty. Second, we
need to explicitly indicate the difference in scale between the groups
in the dataset.

Figure 3A shows an effect plot (Gloor, Macklaim, and Fernandes 2016) of
the data where reads are grouped by function, corresponding
approximately to grouping orthologous sequences regardless of the
organism of origin. Each point represents one of the 3728 functions, and
we can see that there are many more functions represented in the BV
group (bottom) than in the healthy group (top). This is because the
\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
reduced genome content relative to the anaerobic organisms that dominate
in BV, because there is a greater diversity of organisms in BV than in H
samples and because the BV condition has at least an order of magnitude
more bacteria than does the H condition.

We can see that there are a large number of functions that are shared
between the two groups (boxed area in Figure 3A), and inspection shows
that these largely correspond to core metabolic functions that would not
be expected to contribute to differences in ecosystem behaviour. As a
proxy for housekeeping functions the core ribosome functions (blue)
shows that their mean location is not centred on 0. The major group of
these housekeeping functions is located off the line of no difference
(being approximately located at +1.5) and not surprisingly have among
the lowest dispersion in the dataset. Nevertheless, they are identified
as differentially abundant (red) along with many others. While changes
in the abundance of housekeeping functions is a useful proxy for
relative abundance of species in the environment, they tell us nothing
about the functional capacity of the two groups because these are
functions in common to every organism. Of more interest is determining
the functions that are different between groups because these are unique
or over-expressed in one group relative to the other.

\begin{figure}
\centering
\includegraphics{go3_files/figure-latex/meta-1.pdf}
\caption{Analysis of vaginal transcriptome data aggregated at the Kegg
Orthology (KO) functional level. Panel A shows an effect plot for the
default analysis where the functions that are elevated in the healthy
individuals have positive values and functions that are elevated in BV
have negative values. Highlighed in the box are highly abundant KOs that
are almost exlusively housekeeping functions, with ribosomal KOs
highlighted in blue, statistically significant (FDR \textless{} 0.01)
functions in red, and non-significant functions in black or orange.
These housekeeping functions should be located on the midline of no
difference. Panel B shows the same data scaled with
\texttt{gamma\ =\ 1}, which increase the minimum dispersion approximatly
by one unit. Here the housekeeping functions from Panel A are colored
cyan or blue for reference. Panel C shows the same data scaled with
\texttt{gamma\ =\ 0.75} and a 0.15 fold difference in dispersion applied
to the BV samples relative to the H samples. The orange functions are
now statistically significant. Note that this shifts the midpoint of the
housekeeping functions towards the midline.}
\end{figure}

Applying the default scale model of \texttt{gamma=1} as before increases
the dispersion as expected, but does little to move the large number of
housekeeping functions toward the midline of no difference.
Nevertheless, about 50\% of the housekeeping functions are no longer
statistically significantly different.

Up to this point, scale uncertainty has been applied uniformly to both
conditions, but a user-defined scale adjustment can be applied to each
condition, or even each sample independently through a custom scale
matrix. For this we need to revisit the concept that all normalizations
in widespread use are actually ratios with the denominator chosen by one
of several methods. It is important to recognize that in some datasets
the denominator itself is not crucial as all approaches give
substantially similar results; the transcriptome example used previously
is a good example of such an idealized dataset. This begs the question,
``if it is not the denominator, what then is the important
determinant?'' As shown by Nixon et al. (2023) it is usually the
\emph{relationship} between the group scale values (denominators) that
is important, not their actual value. This can be illustrated quite
simply by starting with the mean ratio in \(G\) between groups for the
yeast transcriptome dataset which are \(6.05\times^{-5}\) for snf-w and
\(5.08\times^{-5}\); their ratio being 1.17, or a 0.17-fold difference.
Using this information, we can recapitulate the differential abundance
analysis in Figure 2B and 2C exactly by using setting the mean
denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 1 as
shown in Supplemental Figure 2. Altering this ratio moves the location
of the data.

With this background, we can understand how to apply a full scale to the
metatranscriptome dataset with the \texttt{aldex.makeScaleMatrix()}
function. This function uses a logNormal distribution when building the
model so that ratios that are chosen are symmetrical. The full scale
model option can be used to make a matrix of relative scales that are
distinct for each group (or even each sample) along with their
uncertainties. Applying a per-group relative differential scale of 0.15
with a base scale of 1 moves the housekeeping functions to the midline
of no difference (Figure 3C). Note that now a significant number of
functions are differentially up in BV that were formerly classed as not
different without scale, or when only a uniform scale was applied. These
former false negatives are noted in orange in each panel. Inspection of
the functions shows that these are largely missing from the
Lactobacillus species and so should actually be captured as
differentially abundant. Thus, applying differential scale allows us to
distinguish between both false positives (housekeeping functions in
cyan) and false negatives (orange functions) even in a very difficult to
analyze dataset. We suggest that the default scale model alone is useful
and should be normally applied when the data are well centred. However,
when datasets are not well centred or when the investigator has prior
information about the underlying biology, we advocate developing and
using a full differential scale model.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Biological count data derived from HTS can be decomposed into two parts
the relative (compositional) and the absolute (scale), and the product
of these generates a fully scaled biological system (Nixon et al. 2023).
Biological systems are both predictably variable and stochastic and
current measurement methods that rely on high throughput sequencing fail
to capture all of that variation. In the absence of information external
to the sequencing run itself, no normalisation method can recapture any
of the scale information (Lovén et al. 2012).

Since the underlying scale of the system cannot be measured easily --
methods such as flow cytometry (Vandeputte et al. 2017), spike-in probes
and fluorescent in-situ hybridization (Lovén et al. 2012; Marguerat et
al. 2012) have been used -- the effect on analysis can be modelled by
including scale uncertainty in the analysis (Nixon et al. 2023). Nixon
et al, showed that this can be done by including uncertainty in the
denominator used for the normalization. The ALDEx2 R package is ideally
suited for this since this tool builds a Bayesian posterior of the
compositional component of the dataset at the outset and then conducts
the analysis on that posterior. Adding scale uncertainty can be done at
the same time thus producing a posterior model that incorporates both
compositional and scale uncertainty, and this can be done without
additional computational complexity. For this, the compositional
uncertainty is sampled from a Dirichlet distribution, and the scale
uncertainty can be sampled from any distribution depending on prior
knowledge or preference, but ALDEx2 samples from a logNormal
distribution by default.

All normalizations attempt to make the samples in a dataset commensurate
but do not explicitly address the scale of the underlying system. The
two supplementary tables show that these normalizations can be
interpreted as scale assumptions. However, the general lack of scale
information has important consequences for the analysis of HTS datasets.
One issue is that analysis tools seem over-powered with even moderate
sample sizes and different tools have different intrinsic power, Type 1
and Type 2 error rates (Schurch et al. 2016). Using small sample sizes
in analysis leads to less reliability and reproducibility in analyses
since surprisingly large sample sizes are needed to determine
reproducible p-values (eg. (Halsey et al. 2015)). Thus, recommendations
to use small sample sizes in multivariate datasets such as RNA-seq
datasets are not supported by simple modelling in the univariate case.
Another issue is that datasets are difficult to analyze when there they
contain systematic asymmetry, with different tools exhibiting differing
pathologies with these datasets (Robinson and Oshlack 2010; Wu et al.
2021).

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as transcriptomes
(Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
Quinn, Crowley, and Richardson 2018) and microbiomes(Thorsen et al.
2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
observed that the actual agreement between methods can be modest, and
there is usually a recommendation that some methods are better suited
for individual datasets or for specific types of analysis;
e.g.~`Normalization and microbial differential abundance strategies
depend upon data characteristics' (Weiss et al. 2017). However, another
interpretation of these studies is that the specific assumptions of the
tools and the way they are used by different groups can lead to results
that include both false negative and false positive results, skewed
towards the latter. There are now published examples of multiple large
replication efforts in psychology (Open Science Collaboration 2015),
cancer biology (Rodgers and Collings 2021), ecology and evolution (Gould
and et.al, n.d.). In a nutshell, experiences from these studies indicate
that published data typically have higher effect sizes and appear more
significant than replication efforts with the same data, or with newly
generated replication data (Rodgers and Collings 2021). Thus, most
published data should be viewed with suspicion (Ioannidis 2005) unless
efforts have been made to ensure that the results are robust

In the case of overpowering, HTS analyses seem to be more robust when
applying a dual cutoff of both p-value and difference between group
means (Schurch et al. 2016). Figure 2 shows one reason for this
robustness could be that the dual cutoff is mimicking the effect of
including scale uncertainty, since substantially similar transcripts are
identified by the two approaches. However, while using the post-hoc
difference cutoff is useful for differential abundance analysis it is
not clear how this can be incorporated into other kinds of downstream
analyses. Conversely analyses that include scale uncertainty are fully
compatible with existing downstream analyses.

In the case of asymmetry, the use of a user-specified scale model can be
very useful for otherwise difficult to analyze datasets such as
meta-transcriptomes and in-vitro selection datasets where the majority
of features can change. We showed one such example in Figure 3 where the
dataset was highly asymmetrical, and the TMM and RLE normalizations
cannot fully move all the housekeeping genes to the midline of no
difference or exhibit other pathologies (supplemental). Incorporating
differential scale on a per-group basis moves the mass of the data
towards the midline of no difference and so affects both Type I and Type
II error rates. In this analysis, transcripts that were previously not
classed as differentially abundant are now called as significantly
different, and the housekeeping transcripts move from being
significantly different to not being identified as such. While we
acknowledge that some prior information on which housekeeping
transcripts should not be classed as DA is needed, we suggest that this
information is widely available and is already used when performing the
gold-standard quantitative PCR test of differential abundance (Thellin
et al. 1999; SEQC/MAQC-III Consortium 2014). Furthermore, the assumption
that housekeeping genes should not generally be included in differential
abundance analysis is implicit in the dual p-value, fold-change cutoff
approach in widespread use. Thus, the use of this prior knowledge is not
unique to our approach.

In summary, while the underlying scale of the system is generally
inaccessible, the effect of scale on the analysis outcomes can be
modelled and can help explain some of the underlying biology. Adding
scale information to the analysis allows for more robust inference
because the features that are sensitive to scale can be identified and
their impact on the analysis weighted accordingly. Additionally, the use
of differential scale models permits difficult to analyze datasets to be
examined in a robust and principled manner even when the majority of
features are asymmetrically distributed or expressed (or both) in the
groups. Thus, reporting scale uncertainty should become a standard
practice in the analysis of HTS datasets as a way to identify which
features are most robust to differences in the underlying system.
Finally, we supply a toolkit that makes incorporating scale simple even
for datasets that come from highly asymmetrical environments.

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-aitchison1982}{}}%
Aitchison, John. 1982. {``The Statistical Analysis of Compositional
Data.''} \emph{Journal of the Royal Statistical Society: Series B
(Methodological)} 44 (2): 139--60.

\leavevmode\vadjust pre{\hypertarget{ref-Anders:2010}{}}%
Anders, Simon, and Wolfgang Huber. 2010. {``Differential Expression
Analysis for Sequence Count Data.''} \emph{Genome Biol} 11 (10): R106.
\url{https://doi.org/10.1186/gb-2010-11-10-r106}.

\leavevmode\vadjust pre{\hypertarget{ref-Andrews:2019aa}{}}%
Andrews, Tallulah S, and Martin Hemberg. 2019. {``M3Drop: Dropout-Based
Feature Selection for scRNASeq.''} \emph{Bioinformatics} 35 (16):
2865--67. \url{https://doi.org/10.1093/bioinformatics/bty1044}.

\leavevmode\vadjust pre{\hypertarget{ref-benjamini:1995}{}}%
Benjamini, Yoav, and Yosef Hochberg. 1995. {``Controlling the False
Discovery Rate: A Practical and Powerful Approach to Multiple
Testing.''} \emph{Journal of the Royal Statistical Society. Series B
(Methodological)} 57 (1): 289--300.

\leavevmode\vadjust pre{\hypertarget{ref-Bullard:2010}{}}%
Bullard, James H, Elizabeth Purdom, Kasper D Hansen, and Sandrine
Dudoit. 2010. {``Evaluation of Statistical Methods for Normalization and
Differential Expression in m{RNA-Seq} Experiments.''} \emph{BMC
Bioinformatics} 11: 94. \url{https://doi.org/10.1186/1471-2105-11-94}.

\leavevmode\vadjust pre{\hypertarget{ref-Cui:2003aa}{}}%
Cui, Xiangqin, and Gary A Churchill. 2003.
{``\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical Tests
for Differential Expression in cDNA Microarray Experiments}.''}
\emph{Genome Biol} 4 (4): 210.1--10.

\leavevmode\vadjust pre{\hypertarget{ref-Denge00262-18}{}}%
Deng, Zhi-Luo, Cornelia Gottschick, Sabin Bhuju, Clarissa Masur,
Christoph Abels, and Irene Wagner-Döbler. 2018. {``Metatranscriptome
Analysis of the Vaginal Microbiota Reveals Potential Mechanisms for
Protection Against Metronidazole in Bacterial Vaginosis.''} Edited by
Craig D. Ellermeier, Janet Hill, and Andrew Onderdonk. \emph{mSphere} 3
(3). \url{https://doi.org/10.1128/mSphereDirect.00262-18}.

\leavevmode\vadjust pre{\hypertarget{ref-fernandes:2013}{}}%
Fernandes, Andrew D, Jean M Macklaim, Thomas G Linn, Gregor Reid, and
Gregory B Gloor. 2013. {``ANOVA-Like Differential Expression (ALDEx)
Analysis for Mixed Population RNA-Seq.''} \emph{PLoS One} 8 (7): e67019.
\url{https://doi.org/10.1371/journal.pone.0067019}.

\leavevmode\vadjust pre{\hypertarget{ref-fernandes:2014}{}}%
Fernandes, Andrew D, Jennifer Ns Reid, Jean M Macklaim, Thomas A
McMurrough, David R Edgell, and Gregory B Gloor. 2014. {``Unifying the
Analysis of High-Throughput Sequencing Datasets: Characterizing
{RNA}-Seq, 16{S} r{RNA} Gene Sequencing and Selective Growth Experiments
by Compositional Data Analysis.''} \emph{Microbiome} 2: 15.1--13.
\url{https://doi.org/10.1186/2049-2618-2-15}.

\leavevmode\vadjust pre{\hypertarget{ref-Fettweis:2019aa}{}}%
Fettweis, Jennifer M, Myrna G Serrano, J Paul Brooks, David J Edwards,
Philippe H Girerd, Hardik I Parikh, Bernice Huang, et al. 2019. {``The
Vaginal Microbiome and Preterm Birth.''} \emph{Nat Med} 25 (6):
1012--21. \url{https://doi.org/10.1038/s41591-019-0450-2}.

\leavevmode\vadjust pre{\hypertarget{ref-polyester:2016}{}}%
Frazee, Alyssa C., Andrew E. Jaffe, Rory Kirchner, and Jeffrey T. Leek.
2023. {``{polyester: Simulate RNA-seq reads}.''}
\url{https://doi.org/10.18129/B9.bioc.polyester}.

\leavevmode\vadjust pre{\hypertarget{ref-Gierlinski:2015aa}{}}%
Gierliński, Marek, Christian Cole, Pietà Schofield, Nicholas J Schurch,
Alexander Sherstnev, Vijender Singh, Nicola Wrobel, et al. 2015.
{``Statistical Models for RNA-Seq Data Derived from a Two-Condition
48-Replicate Experiment.''} \emph{Bioinformatics} 31 (22): 3625--30.
\url{https://doi.org/10.1093/bioinformatics/btv425}.

\leavevmode\vadjust pre{\hypertarget{ref-gloor:effect}{}}%
Gloor, GB, JM Macklaim, and AD Fernandes. 2016. {``Displaying Variation
in Large Datasets: Plotting a Visual Summary of Effect Sizes.''}
\emph{Journal of Computational and Graphical Statistics} 25 (3C):
971--79. \url{https://doi.org/10.1080/10618600.2015.1131161}.

\leavevmode\vadjust pre{\hypertarget{ref-gloorAJS:2016}{}}%
Gloor, GB, JM Macklaim, M Vu, and AD Fernandes. 2016. {``Compositional
Uncertainty Should Not Be Ignored in High-Throughput Sequencing Data
Analysis.''} \emph{Austrian Journal of Statistics} 45: 73--87.
\url{https://doi.org/doi:10.17713/ajs.v45i4.122}.

\leavevmode\vadjust pre{\hypertarget{ref-evoRepr-pre}{}}%
Gould, E, and et.al. n.d. {``Same Data, Different Analysts: Variation in
Effect Sizes Due to Analytical Decisions in Ecology and Evolutionary
Biology.''} \emph{EcoEvoRxiv}.
https://doi.org/\url{https://doi.org/10.32942/X2GG62}.

\leavevmode\vadjust pre{\hypertarget{ref-gustafson2015bayesian}{}}%
Gustafson, Paul. 2015. \emph{Bayesian Inference for Partially Identified
Models: Exploring the Limits of Limited Data}. Vol. 140. CRC Press.

\leavevmode\vadjust pre{\hypertarget{ref-Halsey:2015aa}{}}%
Halsey, Lewis G, Douglas Curran-Everett, Sarah L Vowler, and Gordon B
Drummond. 2015. {``The Fickle p Value Generates Irreproducible
Results.''} \emph{Nat Methods} 12 (3): 179--85.
\url{https://doi.org/10.1038/nmeth.3288}.

\leavevmode\vadjust pre{\hypertarget{ref-hawinkel2017}{}}%
Hawinkel, Stijn, Federico Mattiello, Luc Bijnens, and Olivier Thas.
2018. {``A Broken Promise : Microbiome Differential Abundance Methods Do
Not Control the False Discovery Rate.''} \emph{BRIEFINGS IN
BIOINFORMATICS}. \url{http://dx.doi.org/10.1093/bib/bbx104}.

\leavevmode\vadjust pre{\hypertarget{ref-Hughes:2005tu}{}}%
Hughes, Jennifer B, and Jessica J Hellmann. 2005. {``The Application of
Rarefaction Techniques to Molecular Inventories of Microbial
Diversity.''} \emph{Methods Enzymol} 397: 292--308.
\url{https://doi.org/10.1016/S0076-6879(05)97017-1}.

\leavevmode\vadjust pre{\hypertarget{ref-Hummelen:2010}{}}%
Hummelen, Ruben, Andrew D Fernandes, Jean M Macklaim, Russell J Dickson,
John Changalucha, Gregory B Gloor, and Gregor Reid. 2010. {``Deep
Sequencing of the Vaginal Microbiota of Women with {HIV}.''} \emph{PLoS
One} 5 (8): e12078. \url{https://doi.org/10.1371/journal.pone.0012078}.

\leavevmode\vadjust pre{\hypertarget{ref-Ioannidis:2005aa}{}}%
Ioannidis, John P A. 2005. {``Why Most Published Research Findings Are
False.''} \emph{PLoS Med} 2 (8): e124.
\url{https://doi.org/10.1371/journal.pmed.0020124}.

\leavevmode\vadjust pre{\hypertarget{ref-Love:2014aa}{}}%
Love, Michael I, Wolfgang Huber, and Simon Anders. 2014. {``Moderated
Estimation of Fold Change and Dispersion for RNA-Seq Data with
DESeq2.''} \emph{Genome Biol} 15 (12): 550.1--21.
\url{https://doi.org/10.1186/s13059-014-0550-8}.

\leavevmode\vadjust pre{\hypertarget{ref-lovell:2011}{}}%
Lovell, David, Warren Müller, Jen Taylor, Alec Zwart, and Chris
Helliwell. 2011. {``Proportions, Percentages, Ppm: Do the Molecular
Biosciences Treat Compositional Data Right?''} In \emph{Compositional
Data Analysis: Theory and Applications}, edited by Vera Pawlowsky-Glahn
and Antonella Buccianti, 193--207. London: John Wiley; Sons New York,
NY.

\leavevmode\vadjust pre{\hypertarget{ref-Lovell:2015}{}}%
Lovell, David, Vera Pawlowsky-Glahn, Juan José Egozcue, Samuel
Marguerat, and Jürg Bähler. 2015. {``Proportionality: A Valid
Alternative to Correlation for Relative Data.''} \emph{PLoS Comput Biol}
11 (3): e1004075.
https://doi.org/\url{https://doi.org/10.1371/journal.pcbi.1004075}.

\leavevmode\vadjust pre{\hypertarget{ref-Loven:2012aa}{}}%
Lovén, Jakob, David A Orlando, Alla A Sigova, Charles Y Lin, Peter B
Rahl, Christopher B Burge, David L Levens, Tong Ihn Lee, and Richard A
Young. 2012. {``Revisiting Global Gene Expression Analysis.''}
\emph{Cell} 151 (3): 476--82.
\url{https://doi.org/10.1016/j.cell.2012.10.012}.

\leavevmode\vadjust pre{\hypertarget{ref-macklaim:2013}{}}%
Macklaim, Jean M, Andrew D Fernandes, Julia M Di Bella, Jo-Anne Hammond,
Gregor Reid, and Gregory B Gloor. 2013. {``Comparative Meta-{RNA}-Seq of
the Vaginal Microbiota and Differential Expression by Lactobacillus
Iners in Health and Dysbiosis.''} \emph{Microbiome} 1 (1): 12.
\url{https://doi.org/10.1186/2049-2618-1-12}.

\leavevmode\vadjust pre{\hypertarget{ref-Macklaim:2018aa}{}}%
Macklaim, Jean M, and Gregory B Gloor. 2018. {``From {RNA}-Seq to
Biological Inference: Using Compositional Data Analysis in
Meta-Transcriptomics.''} \emph{Methods Mol Biol} 1849: 193--213.
\url{https://doi.org/10.1007/978-1-4939-8728-3/_13}.

\leavevmode\vadjust pre{\hypertarget{ref-yeast-absolute}{}}%
Marguerat, Samuel, Alexander Schmidt, Sandra Codlin, Wei Chen, Ruedi
Aebersold, and Jürg Bähler. 2012. {``Quantitative Analysis of Fission
Yeast Transcriptomes and Proteomes in Proliferating and Quiescent
Cells.''} \emph{Cell} 151 (3): 671--83.
\url{https://doi.org/10.1016/j.cell.2012.09.019}.

\leavevmode\vadjust pre{\hypertarget{ref-Mortazavi:2008}{}}%
Mortazavi, Ali, Brian A Williams, Kenneth McCue, Lorian Schaeffer, and
Barbara Wold. 2008. {``Mapping and Quantifying Mammalian Transcriptomes
by {RNA-seq}.''} \emph{Nat Methods} 5 (7): 621--28.
\url{https://doi.org/10.1038/nmeth.1226}.

\leavevmode\vadjust pre{\hypertarget{ref-Nie:2012aa}{}}%
Nie, Zuqin, Gangqing Hu, Gang Wei, Kairong Cui, Arito Yamane, Wolfgang
Resch, Ruoning Wang, et al. 2012. {``C-{M}yc Is a Universal Amplifier of
Expressed Genes in Lymphocytes and Embryonic Stem Cells.''} \emph{Cell}
151 (1): 68--79. \url{https://doi.org/10.1016/j.cell.2012.08.033}.

\leavevmode\vadjust pre{\hypertarget{ref-nixon2023scale}{}}%
Nixon, Michelle Pistner, Jeffrey Letourneau, Lawrence A. David, Nicole
A. Lazar, Sayan Mukherjee, and Justin D. Silverman. 2023. {``Scale
Reliant Inference.''} \url{https://arxiv.org/abs/2201.03616}.

\leavevmode\vadjust pre{\hypertarget{ref-Open-Science-Collaboration:2015aa}{}}%
Open Science Collaboration. 2015. {``Estimating the Reproducibility of
Psychological Science.''} \emph{Science} 349 (6251): aac4716.
\url{https://doi.org/10.1126/science.aac4716}.

\leavevmode\vadjust pre{\hypertarget{ref-Props:2017aa}{}}%
Props, Ruben, Frederiek-Maarten Kerckhof, Peter Rubbens, Jo De Vrieze,
Emma Hernandez Sanabria, Willem Waegeman, Pieter Monsieurs, Frederik
Hammes, and Nico Boon. 2017. {``Absolute Quantification of Microbial
Taxon Abundances.''} \emph{ISME J} 11 (2): 584--87.
\url{https://doi.org/10.1038/ismej.2016.117}.

\leavevmode\vadjust pre{\hypertarget{ref-Quinn:2018aa}{}}%
Quinn, Thomas P, Tamsyn M Crowley, and Mark F Richardson. 2018.
{``Benchmarking Differential Expression Analysis Tools for RNA-Seq:
Normalization-Based Vs. Log-Ratio Transformation-Based Methods.''}
\emph{BMC Bioinformatics} 19 (1): 274.
\url{https://doi.org/10.1186/s12859-018-2261-8}.

\leavevmode\vadjust pre{\hypertarget{ref-Ravel:2010}{}}%
Ravel, Jacques, Pawel Gajer, Zaid Abdo, G Maria Schneider, Sara S K
Koenig, Stacey L McCulle, Shara Karlebach, et al. 2011. {``Vaginal
Microbiome of Reproductive-Age Women.''} \emph{Proc Natl Acad Sci U S
A}, no. 108: 4680--87. \url{https://doi.org/doi/10.1073/pnas.100611107}.

\leavevmode\vadjust pre{\hypertarget{ref-Robinson:2010}{}}%
Robinson, Mark D, Davis J McCarthy, and Gordon K Smyth. 2010. {``edgeR:
A Bioconductor Package for Differential Expression Analysis of Digital
Gene Expression Data.''} \emph{Bioinformatics} 26 (1): 139--40.
\url{https://doi.org/10.1093/bioinformatics/btp616}.

\leavevmode\vadjust pre{\hypertarget{ref-Robinson:2010a}{}}%
Robinson, Mark D, and Alicia Oshlack. 2010. {``A Scaling Normalization
Method for Differential Expression Analysis of {RNA-seq} Data.''}
\emph{Genome Biol} 11 (3): R25.1--9.
\url{https://doi.org/10.1186/gb-2010-11-3-r25}.

\leavevmode\vadjust pre{\hypertarget{ref-Rodgers:2021aa}{}}%
Rodgers, Peter, and Andy Collings. 2021. {``What Have We Learned?''}
\emph{Elife} 10 (December). \url{https://doi.org/10.7554/eLife.75830}.

\leavevmode\vadjust pre{\hypertarget{ref-Schurch:2016aa}{}}%
Schurch, Nicholas J, Pietá Schofield, Marek Gierliński, Christian Cole,
Alexander Sherstnev, Vijender Singh, Nicola Wrobel, et al. 2016. {``How
Many Biological Replicates Are Needed in an RNA-Seq Experiment and Which
Differential Expression Tool Should You Use?''} \emph{RNA} 22 (6):
839--51. \url{https://doi.org/10.1261/rna.053959.115}.

\leavevmode\vadjust pre{\hypertarget{ref-SEQCux2fMAQC-III-Consortium:2014aa}{}}%
SEQC/MAQC-III Consortium. 2014. {``A Comprehensive Assessment of RNA-Seq
Accuracy, Reproducibility and Information Content by the Sequencing
Quality Control Consortium.''} \emph{Nat Biotechnol} 32 (9): 903--14.
\url{https://doi.org/10.1038/nbt.2957}.

\leavevmode\vadjust pre{\hypertarget{ref-Soneson:2013}{}}%
Soneson, Charlotte, and Mauro Delorenzi. 2013. {``A Comparison of
Methods for Differential Expression Analysis of {RNA-seq} Data.''}
\emph{BMC Bioinformatics} 14: 91.
\url{https://doi.org/10.1186/1471-2105-14-91}.

\leavevmode\vadjust pre{\hypertarget{ref-Taniguchi:2010aa}{}}%
Taniguchi, Yuichi, Paul J Choi, Gene-Wei Li, Huiyi Chen, Mohan Babu,
Jeremy Hearn, Andrew Emili, and X Sunney Xie. 2010. {``Quantifying e.
Coli Proteome and Transcriptome with Single-Molecule Sensitivity in
Single Cells.''} \emph{Science} 329 (5991): 533--38.
\url{https://doi.org/10.1126/science.1188308}.

\leavevmode\vadjust pre{\hypertarget{ref-Thellin:1999aa}{}}%
Thellin, O, W Zorzi, B Lakaye, B De Borman, B Coumans, G Hennen, T
Grisar, A Igout, and E Heinen. 1999.
{``\href{https://www.ncbi.nlm.nih.gov/pubmed/10617337}{Housekeeping
Genes as Internal Standards: Use and Limits}.''} \emph{J Biotechnol} 75
(2-3): 291--95.

\leavevmode\vadjust pre{\hypertarget{ref-Thorsen:2016aa}{}}%
Thorsen, Jonathan, Asker Brejnrod, Martin Mortensen, Morten A Rasmussen,
Jakob Stokholm, Waleed Abu Al-Soud, Søren Sørensen, Hans Bisgaard, and
Johannes Waage. 2016. {``Large-Scale Benchmarking Reveals False
Discoveries and Count Transformation Sensitivity in 16{S} r{RNA} Gene
Amplicon Data Analysis Methods Used in Microbiome Studies.''}
\emph{Microbiome} 4 (1): 62.
\url{https://doi.org/10.1186/s40168-016-0208-8}.

\leavevmode\vadjust pre{\hypertarget{ref-Vandeputte:2017aa}{}}%
Vandeputte, Doris, Gunter Kathagen, Kevin D'hoe, Sara Vieira-Silva,
Mireia Valles-Colomer, João Sabino, Jun Wang, et al. 2017.
{``Quantitative Microbiome Profiling Links Gut Community Variation to
Microbial Load.''} \emph{Nature} 551 (7681): 507--11.
\url{https://doi.org/10.1038/nature24460}.

\leavevmode\vadjust pre{\hypertarget{ref-wagner:tpm}{}}%
Wagner, Günter P, Koryu Kin, and Vincent J Lynch. 2012. {``Measurement
of mRNA Abundance Using RNA-Seq Data: RPKM Measure Is Inconsistent Among
Samples.''} \emph{Theory Biosci} 131 (4): 281--85.
\url{https://doi.org/10.1007/s12064-012-0162-3}.

\leavevmode\vadjust pre{\hypertarget{ref-Weiss:2017aa}{}}%
Weiss, Sophie, Zhenjiang Zech Xu, Shyamal Peddada, Amnon Amir, Kyle
Bittinger, Antonio Gonzalez, Catherine Lozupone, et al. 2017.
{``Normalization and Microbial Differential Abundance Strategies Depend
Upon Data Characteristics.''} \emph{Microbiome} 5 (1): 27.
\url{https://doi.org/10.1186/s40168-017-0237-y}.

\leavevmode\vadjust pre{\hypertarget{ref-Wu2021}{}}%
Wu, Jia R., Jean M. Macklaim, Briana L. Genge, and Gregory B. Gloor.
2021. {``Finding the Centre: Compositional Asymmetry in High-Throughput
Sequencing Datasets.''} In \emph{Advances in Compositional Data
Analysis: Festschrift in Honour of Vera Pawlowsky-Glahn}, edited by
Peter Filzmoser, Karel Hron, Josep Antoni Martìn-Fernàndez, and Javier
Palarea-Albaladejo, 329--46. Cham: Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-71175-7_17}.

\leavevmode\vadjust pre{\hypertarget{ref-Yoshikawa:2011aa}{}}%
Yoshikawa, Katsunori, Tadamasa Tanaka, Yoshihiro Ida, Chikara Furusawa,
Takashi Hirasawa, and Hiroshi Shimizu. 2011. {``Comprehensive Phenotypic
Analysis of Single-Gene Deletion and Overexpression Strains of
Saccharomyces Cerevisiae.''} \emph{Yeast} 28 (5): 349--61.
\url{https://doi.org/10.1002/yea.1843}.

\leavevmode\vadjust pre{\hypertarget{ref-Zhang:2020ab}{}}%
Zhang, Yuqing, Giovanni Parmigiani, and W Evan Johnson. 2020.
{``ComBat-Seq: Batch Effect Adjustment for RNA-Seq Count Data.''}
\emph{NAR Genom Bioinform} 2 (3): lqaa078.
\url{https://doi.org/10.1093/nargab/lqaa078}.

\leavevmode\vadjust pre{\hypertarget{ref-Zozaya:2010}{}}%
Zozaya-Hinchliffe, Marcela, Rebecca Lillis, David H Martin, and Michael
J Ferris. 2010. {``Quantitative PCR Assessments of Bacterial Species in
Women with and Without Bacterial Vaginosis.''} \emph{J Clin Microbiol}
48 (5): 1812--19. \url{https://doi.org/10.1128/JCM.00851-09}.

\end{CSLReferences}

\end{document}

### 2023-10-17 go3-toMPN.tex
sent for comment

--- 2023-10-17 go3-author.tex
@@
 \author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}
+
 \date{\today}
@@
 significant than replication efforts with the same data, or with newly
-generated replication data (Rodgers and Collings 2021). Thus, most
-published data should be viewed with suspicion (Ioannidis 2005) unless
-efforts have been made to ensure that the results are robust
+generated replication data (Rodgers and Collings 2021). This is if the
+data can be replicated de-novo at all {[}Errington:2021aa{]}.
 
@@
 One} 5 (8): e12078. \url{https://doi.org/10.1371/journal.pone.0012078}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ioannidis:2005aa}{}}%
-Ioannidis, John P A. 2005. {``Why Most Published Research Findings Are
-False.''} \emph{PLoS Med} 2 (8): e124.
-\url{https://doi.org/10.1371/journal.pmed.0020124}.
 

### 2023-10-17 go3.Rmd
sent for comment

@@
   - name: Greg Gloor
-  affiliation: Dep't of Biochemistry, University of Western Ontario
+    affiliation: Department of Biochemistry, University of Western Ontario
     email: ggloor@uwo.ca
@@
   - name: Justin Silverman
-  affiliation: College of Information Sciences and Technology, Department of Statistics, Department of Medicine, Pennsylvania  State University
+    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
 output:
-  BiocStyle::pdf_document: default
+  nucleic-acids-research.csl: pdf_document
 package: ALDEx2
-abstract: |
-    Introduction to scale simulation and FDR correction with ALDEx2.
+abstract: 
+  Despite scale being an important confounding variable, high throughput sequencing generates data that are difficult to reconcile with the scale of the underlying environment, but it has been shown that the scale of the environment is related to the geometric mean of the parts of the sample.
+The ALDEx2 R package builds a Bayesian model of the data and here we report an update to the model that includes both random sampling and scale estimates.
+We show that incorporating scale allows for robust inference in a number of different experimental designs and we provide guidance on how to incorporate scale inference into differential abundance analysis.
+While all data normalizations in widespread use make implicit assumptions about the scale of the system we suggest that explicit rather than implicit scale models should be used.
 ---
@@
 In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken.
-It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment[@Lovell:2015;@Vandeputte:2017aa;@Props:2017aa].
+It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment[@Lovell:2015;@Vandeputte:2017aa;@Props:2017aa] .
 
@@
 
-More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix \( \mathbf{W} \), then the elements of the matrix can be decomposed into its composition (relative information, \( \mathbf{W}^{\parallel} \) ) and its scale (total sum, \( \mathbf{W}^{\perp} \) ); formally, \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\).
+More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix \( \mathbf{W} \), then the elements of the matrix can be decomposed into its composition (relative information, \( \mathbf{W}^{\parallel} \) ) and its scale (total sum, \( \mathbf{W}^{\perp} \) ); formally, \(\mathbf{W} = \mathbf{W}^{\parallel}  \mathbf{W}^{\perp}\).
 If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).
@@
 Recently, Nixon et al.
-[-@nixon2023scale] showed that that *all* normalizations make some  identifying assumption about the scale of \(\mathbf{Y}\) but that the actual assumption being made was not always easy to interpret.
-The CLR which can be written as \( \mathbf{CLR}_{dn} = \mathbf{Y}_{dn} / G_n \) uses the geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the data.
-They discovered that \(G_n\) was related to the scale of the underlying environment \( \mathbf{W\perp} \).
+[-@nixon2023scale] showed that that *all* normalizations make some  identifying assumption about the scale of \(\mathbf{W}\) but that the actual assumption being made was not always easy to interpret.
+In essence, normalizations in widespread use assume that either one sample can be chosen as a reference to which the others are scaled (TMM), or that different sub-parts of each sample maintain a constant scale across samples (RLE, CSS, CLR and derivatives).
+
+The CLR, which can be written as \( \mathbf{CLR}_{(1...d)n} = \mathbf{Y}_{(1...d)n} / G_n \),  uses the geometric mean \( G_n \) of \( \mathbf{Y}^{\parallel}_{n} \) to scale the data.
+Nixon et al.
+[-@nixon2023scale] discovered that using the CLR was equivalent to making a statement that \( G_n \) was related to the scale of the underlying environment \( \mathbf{W}^{\perp} \).
 Interestingly, the information in a system is a fundamental property and \(G_n\) is strongly correlated with Shannon's entropy, indicating that information about the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental).
@@
 In the case of the CLR which uses the	 \(G_n\) as the denominator, this entails adding uncertainty into  \(G_n\) such that the latent bias can be exposed, acknowledged and accounted for.
+Additionally it was realized that the actual values used in the denominator to calculate the log-ratio were irrelevant, but instead that the ratios between the denominators was the key parameter.
+These two insights allow for a full scale model of scale to be built and tested for its effect on analysis.
 
@@
 ALDEx2 was designed originally to convert the discrete point estimates in the count matrix obtained into continuous posterior distributions through Monte-Carlo sampling from the Dirichlet distribution.
-The posterior more accurately represents variation on a per-sample basis[@gloorAJS:2016] and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate \(\mathbf{W}^{\parallel}_{dn}\).
-Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates.
+The posterior more accurately represents variation on a per-sample basis[@gloorAJS:2016]	 and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate \(\mathbf{W}^{\parallel}_{dn}\).	 Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally	expected values and confidence intervals are reported across the Monte-Carlo replicates.
 The CLR calculation can be modified trivially to include a full scale model of the data by incorporating uncertainty into \(G\) and so produce a posterior estimate of the scale.
 The scale uncertainty  can be included on a per-experiment, per-group or even a per-sample basis.
-Thus, ALDEx2 can now be used to generate a full posterior model of both \(\mathbf{Y}^{\parallel}_{dn}\) and \(\mathbf{Y}^{\perp}_{dn}\).
+Thus, ALDEx2 can now be used to generate a full posterior model of both \(\mathbf{W}^{\parallel}_{dn}\) and \(\mathbf{W}^{\perp}_{dn}\) from the observed data.
 
@@
 The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa].
-
-```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values.
+```{r, echo=F, warning=F, message=F,comment=F, result=F} 
+library(DESeq2)
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+#library(ALDEx2)
+```
+```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as clr values.
 Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale.
@@
 The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }
-
-library(DESeq2)
-devtools::load_all('~/Documents/0_git/ALDEx_bioc')
-#library(ALDEx2)
 
@@
 
-The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization.
+The very low dispersion estimate for most of the features  arises because  scale variation in the underlying dataset has been removed through sequencing and normalization.
 While the actual scale of the data is inaccessible post-sequencing	we can estimate the *effect* of scale on the output.
@@
 Applying `gamma=1` as a parameter we can see that the large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts (`r length(sig.s.ald)`) being called significantly different  as shown in the volcano plot in Figure 1F (red points).
-Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.
-Indeed, adding scale uncertainty reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff.
+Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.	Indeed, adding scale uncertainty reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff.
 Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.
@@
 This function uses a logNormal distribution when building the model so that ratios that are chosen are symmetrical.
-The full scale model option can be used ot make a matrix of relative scales that are distinct for each group (or even each sample) along with their uncertainties.
+The full scale model option can be used to make a matrix of relative scales that are distinct for each group (or even each sample) along with their uncertainties.
 Applying a per-group relative differential scale of 0.15 with a base scale of 1 moves the housekeeping functions to the midline of no difference (Figure 3C).
@@
 Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty, and this can be done without additional computational complexity.
-For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a Normal or logNormal distribution depending on preference.
+For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty can be sampled from any distribution depending on prior knowledge or preference, but ALDEx2 samples from a logNormal distribution by default.
 
 All normalizations attempt to make the samples in a dataset commensurate but do not explicitly address the scale of the underlying system.
+The two supplementary tables show that these normalizations can be interpreted as scale assumptions.
 However, the general lack of scale information has important consequences for the analysis of HTS datasets.
@@
 Another issue is that datasets are difficult to analyze when there they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a; @Wu2021].
+
+Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes [@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa] and microbiomes[@Thorsen:2016aa;@Weiss:2017aa;@hawinkel2017].
+Generically, it is observed  that the actual agreement between methods can be modest, and there is usually a recommendation that some methods are better suited for individual datasets or for specific types of analysis; e.g.
+'Normalization and microbial differential abundance strategies depend upon data characteristics' [@Weiss:2017aa].
+However, another interpretation of these studies is that the specific assumptions of the tools and the way they are used by different groups can lead to results that include both false negative and false positive results, skewed towards the latter.
+There are now published examples of multiple large replication efforts in psychology [@Open-Science-Collaboration:2015aa], cancer biology [@Rodgers:2021aa], ecology and evolution [@evoRepr-pre].
+In a nutshell, experiences from these studies indicate that published data typically have higher effect sizes and appear more significant than replication efforts with the same data, or with newly generated replication data [@Rodgers:2021aa].
+Thus, most published data should be viewed with suspicion [@Ioannidis:2005aa] unless efforts have been made to ensure that the results are robust
 

### 2023-10-17 supplement_figs.Rmd
sent for comment

@@
 Information content is a fundamental property of all systems.
-We can think about the scale relationship from an information theoretic point of view which measures how much information is encoded in a particular sample.
+We can think about  scale  from an information theoretic point of view as a measure of how much information is encoded in a particular sample.
 Intuitively,   systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
 Below, we show that the scale of a system as defined by Nixon et al.
-[-@nixon2023scale] is related to the information content and entropy of the data.
+[-@nixon2023scale] is inversely related to the information content and entropy of the data.
 
-Recall the underlying system is described by a \(D \times N\) matrix of counts \(\mathbf{W}\) decomposed into the proportions for the \(n^{th}\) sample \(\mathbf{W}^{\parallel}_n\) (or the equivalent probability distribution \( \mathbf{p}(w_n) \) ), and its scale \(\mathbf{W}^{\perp}_n\).
-Sequencing  measures which is related to the underlying proportion; i.e., \(\mathbf{Y}^{\parallel}_n \sim \mathbf{W}^{\parallel}_n\)
+Recall the underlying system is described by a \(D \times N\) matrix of counts \(\mathbf{W}\) decomposed into the proportions for the \(n^{th}\) sample \(\mathbf{W}^{\parallel}_n\) (or the equivalent probability distribution \( \mathbf{p}(w_n) \) ), and its scale \(\mathbf{W}^{\perp}_n\), such that \(\mathbf{W}=\mathbf{W}^{\parallel}\mathbf{W}^{\perp} \).
+Sequencing  returns counts which are related to the underlying proportion; i.e., \(\mathbf{Y}^{\parallel}_n \sim \mathbf{W}^{\parallel}_n\)
 
 For notational simplicity assume we will use a single discrete random variable to represent the probability distribution; i.e.
-\( X =  \mathbf{p}(w_n) \)  over \(1 \dots d \) features.
+\( X =  \mathbf{p}(w_n) \)  over \(1 \dots d \) elements.
 The geometric mean can be calculated as:
@@
 In information theory, the elemental amount of information or uncertainty for \(p_i\) is the inverse \(I(X)_i = -log_2(p_i)\)[@rezaInfoIntro].
+
 The total entropy of the system  \( H(x) \) is the weighted sum of the elemental information
@@
 
-In this formulation, \( \log_2(G(X)) \) and \( H(X) \) can be understood from an information theoretic point of view to be an unweighted and a weighted measure  of 'surprisal' in the dataset.
+In this formulation, \( \log_2(G(X)) \) and \( H(X) \) can be understood  to be an unweighted and a weighted measure  of 'surprisal' in the dataset.
 The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.
@@
 Conversely, \( G \) is an unweighted mean of the same distribution and  \( G \) is the denominator to calculate the centred log-ratio normalization (CLR).
+Note that when calculating the CLR we are taking a ratio, and similarly for calculating entropy, but that the scale formula takes the product.
+Thus, the scale of a system is inversely related to the other two measures.
 
-Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(Y^{\parallel}_{n})\) (Supplemental Figure 1 and Table 1).
+Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(\mathbf{Y}^{\parallel}_{n})\) (Supplemental Figure 1 and Table 1).
 These two measures are expected to have different behaviours with different distributions of \( p_i \).
-In the case of a uniform distribution both \( H(Y^{\parallel}_{n}) \) and \( G(Y^{\parallel}_{n}) \) are maximal since \( p(x) \) is equally and identically distributed.
+In the case of a uniform distribution both \( H(\mathbf{Y}^{\parallel}_{n}) \) and \( G(\mathbf{Y}^{\parallel}_{n}) \) are maximal since \( p(x) \) is equally and identically distributed.
 Thus, we expect that they are positively correlated here.
 In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values.
-In very sparse  datasets, the two measures could become uncoupled because \( H(Y^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(Y^{\parallel}_{n}) \) would tend to be very small.
+In very sparse  datasets, the two measures could become uncoupled because \( H(\mathbf{Y}^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(\mathbf{Y}^{\parallel}_{n}) \) would tend to be very small.
 Here these two measures could be either uncorrelated or exhibit negative correlation.
@@
 Nixon et al.
-[-nixon2023scale] showed that many operations on HTS datasets relied on both the proportion and scale components of the data.
+[-@nixon2023scale] showed that many operations on HTS datasets relied on both the proportion and scale components of the data.
 Moreover, all normalizations impose a scale model on the data, but the appropriateness of these models has never been explicitly acknowledged or tested.
@@
 The in vitro selection dataset is clearly an outlier in both the difference between the average group geometric mean, and in the offset plot.
-However, this dataset can  be used to illustrate the power of the full scale model.
+However, this dataset can  be used to illustrate the power of the full scale model and the relationship between \( G_n \) and scale.
 In Figure 3 we can see that the default output of ALDEx2 has a centered output.
@@
 For this dataset to be centered we need to have a scale ratio \( \approx \) 1:50 or more.
+Note that the scale ratio is inverse to the ratio of geometric means as described above.
 In fact, in this dataset the relative abundances of the majority of features are nearly invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014], thus changing the scale of the data.
@@
 While not wrong as long as the assumption that the scales are identical is stated, this is not a useful analysis outcome.
-Modifying the mean scale difference between groups to be \( \approx \) 1:50 different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained using \(G_n\) as the scale estimate.
-Note that we get exactly the same answer (within random sampling error) with a scale of 0.02 for group 1 and a scale of 1 for group 2, or using a scale of 1 for group 1 and a scale of 50 for group 2.
+Modifying the mean scale difference between the NS:S groups to be \( \approx \) 1:50 different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained using \(G_n\) as the scale estimate where the ratio is \( \sim 100:1 \).
+Note that we get exactly the same answer (within random sampling error) with a scale of .02 for group NS and a scale of 1 for group S, or using a scale of 1 for group NS and a scale of 50 for group S.
 This shows that it is the relative difference between scales that is important in this dataset, not the absolute values.
@@
 
-x.all <- aldex(selex, conditions=conds, gamma=0.5)
+x.all <- aldex(selex, conditions=conds, CI=T, gamma=0.5)
 
 sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
-x.1.all <- aldex(selex, conditions=conds, gamma=sel.mu)
+x.1.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
 
 # setting ratio as 1:50 explicitly
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
-x.5.all <- aldex(selex, conditions=conds, gamma=sel.mu)
+#sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
+scale <- c(rep(1,7), rep(50,7)) 
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=FALSE)
+x.5.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
 
@@
 # 2^-5.64 : 2^0  = 0.02 : 1
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(-5.64,0), conds, log=TRUE)
-x.2.all <- aldex(selex,conditions=conds, gamma=sel.mu)
+# this also works, but reversed because of factor()  
+# sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(0,-5.64), conds, log=TRUE)
+scale <- c(rep(-5.6,7), rep(0,7))
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=TRUE)
+
+x.2.all <- aldex(selex,conditions=conds, CI=T, gamma=sel.mu)
 
@@
 When overplotted on the edgeR volcano plot (blue), it is clear that these functions have non-signficant p-values.
+
+# Checking the scale assumptions of the RLE and TMM normalizations
+
+We can show that the normalizations built into the edgeR Bioconductor package (RLE, TMM, TMMwsp, upperquantile) are scale assumptions by using the normalization factor as an input to `aldex.makeScaleMatrix()` and then measure the mean location of the data as above.
+The ideal behavior of a normalization is that the mean location of the data should be close to 0 or unchanged.
+This behavior will ensure that Type 1 and Type 2 errors due to scale assumptions are minimized.
+
+The results, shown in the tables below compare these normalizations to the no scale assumption (iso) and the geometric mean normalization (GM) assuming that the normalizations are either on a log scale or a linear scale.
+That these affect scale can be observed by their effect on the mean location of the data.
+For the most part assuming no scale differences between groups centres the date less well than does the geometric mean.
+In most cases the other normalizations perform poorer than assuming no scale at all for the rRNA dataset and about as well as the no scale assumption in the metatranscriptome datasets and the single-cell transcriptome dataset.
+All normalizations perform poorly in the selex dataset.
+Overall, these results may not be surprising because the TMM and RLE (and related normalizations) were developed specifically to scale and normalize transcriptome datasets [@Robinson:2010a;@Anders:2010], but have become widely used in the analysis of other data modalities; e.g.
+[@McMurdie:2014a].
+
+```{r getNorm, echo=F}
+# output from code/scale_norms.R
+
+load('analysis/normalizations.Rda')
+knitr::kable(norm.data.out, 'simple', caption='Log scale models')
+```
+
+```{r getNormNL, echo=F}
+# output from code/scale_norms.R
+
+load('analysis/normalizations.nl.Rda')
+knitr::kable(norm.nl.data.out, 'simple', caption='Linear scale models')
+```
+
+# References

### 2023-10-17 code/markdown.R

@@ -1,0 +1,2 @@
+library(rmarkdown)
+rmarkdown::render('go3.Rmd', output_format=pdf_document(keep_tex = T))

### 2023-10-17 code/scale_norms.R

@@ -1,0 +1,103 @@
+# mu.vec = c(log2(rep(1,7)), log2(rep(1.2,7)))
+# scale_samples <- t(sapply(mu.vec, FUN = function(mu) rlnorm(128, mu, 1e-3)))
+library(edgeR)
+
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+data(selex)
+
+load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+yst <- yst[rownames(yst) != "YOR072W-B", ]
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/meta16S.tsv"
+rRNA <- read.table(url, header=T, row.names=1, sep='\t')
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
+ss <- read.table(url, header=T, row.names=1, sep='\t')
+ss <- ss[,c(1:100,1502:1601)]
+
+conds <- list()
+conds$rconds <- c(rep(1,198), rep(2,161))
+conds$sconds <- c(rep(1,7), rep(2,7))
+conds$yconds <- c(rep(1,44), rep(2,42))
+conds$mconds <- c(rep(1,8), rep(2,28), rep(1,8))
+conds$ssconds <- c(rep(1,100), rep(2,100))
+
+datasets <- list()
+datasets$rRNA <- rRNA
+datasets$selex <- selex
+datasets$yst <- yst
+datasets$meta <- ko.both
+datasets$ss <- ss
+
+norm.method <- c('RLE', 'TMM', 'TMMwsp', 'upperquartile')
+
+# samples by column, offsets by row
+# using sweep '+'
+norm.data.out <- matrix(data=NA, ncol=6, nrow=5)
+norm.nl.data.out <- matrix(data=NA, ncol=6, nrow=5)
+for(i in 1:5){
+	x <- aldex.clr(datasets[[5]], conds[[5]], gamma=0.3)
[…]

### 2023-10-17 fix_authors.sh

@@ -1,0 +1,14 @@
+#!/bin/zsh
+
+awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin Silverman$^{2,3}$")};1' go3.tex > jnk.tex
+
+awk '{sub(/\\date{}/,"\\date{\\today}")};1' jnk.tex > jnk2.tex
+
+awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$ Department of Biochemistry, University of Western Ontario\n\$^2\$ College of Information Sciences and Technology, Pennsylvania  State University\n\$^3\$ Department of Medicine, Pennsylvania  State University\n")};1' jnk2.tex > go3-author.tex
+
+rm jnk.tex
+rm jnk2.tex
+rm go3.aux
+rm go3.pdf
+rm go3.out
+rm go3.toc

### 2023-10-21 go3-author.tex
submitted version

@@
 \hypersetup{
-  pdftitle={Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2},
+  pdftitle={Normalizations are not what you think; Explicit Scale Simulation in ALDEx2},
   hidelinks,
@@
 
-\title{Beyond compositionally in high throughput sequencing; estimating
-the importance of scale in data analysis with ALDEx2}
-\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}
+\title{Normalizations are not what you think; Explicit Scale Simulation
+in ALDEx2}
+\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4}$}
 \date{\today}
@@
 $^3$ Department of Medicine, Pennsylvania  State University
+$^4$ Institute for Computational and Data Science, Pennsylvania State University
 
 \begin{abstract}
-Despite scale being an important confounding variable, high throughput
-sequencing generates data that are difficult to reconcile with the scale
-of the underlying environment, but it has been shown that the scale of
-the environment is related to the geometric mean of the parts of the
-sample. The ALDEx2 R package builds a Bayesian model of the data and
-here we report an update to the model that includes both random sampling
-and scale estimates. We show that incorporating scale allows for robust
-inference in a number of different experimental designs and we provide
-guidance on how to incorporate scale inference into differential
-abundance analysis. While all data normalizations in widespread use make
-implicit assumptions about the scale of the system we suggest that
-explicit rather than implicit scale models should be used.
+In many high-throughput sequencing (HTS) studies, sample-to-sample
+variation in sequencing depth is not driven by variation in the scale
+(e.g., total size, microbial load, or total gene expression) of the
+underlying biological systems being measured but rather is driven by
+technical factors. Typically the technical variation is addressed using
+some form of statistical normalization. Normalizations are data or
+parameter transformations that remove unwanted technical variation in
+the hopes of facilitating analyses sensitive to scale; e.g.,
+differential abundance and differential expression analyses. Recently we
+showed that any normalization makes implicit assumptions about the
+unmeasured system scale and that errors in these assumptions can lead to
+dramatic increases in false positive and false negative rates. Here we
+describe updates to the ALDEx2 R package that mitigate these problems by
+directly modeling uncertainty in the unmeasured system scale through the
+use of a \textit{scale model}. Scale models generalize the idea of
+normalizations and can be thought of as explicitly modeling potential
+error in the chosen normalization. Beyond enhancing the robustness of
+HTS analyses, the use of scale models within ALDEx2 enhances the
+transparency and reproducibility of analyses by making implicit
+normalizing assumptions an explicit part of the model building process.
 \end{abstract}
@@
 
-High throughput sequencing (HTS) is a universal tool to explore many
-biological phenomenon such as gene expression (single-cell sequencing,
-RNA-sequencing, meta-transcriptomics), microbial community composition
-(16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme
-activity (selex, CRISPR killing). HTS proceeds by taking a fixed-number
-sample from the environment, making a library, multiplexing (merging)
-multiple libraries and applying a fixed number of molecules to the flow
-cell. In essence it is a poll of the environment that is mixed with
-other polls and then a poll of the mixture is taken. It should be clear
-that the total number of molecules sequenced is driven by the capacity
-of the instrument and not by the number of molecules in the sampled
-environment(Lovell et al. 2015; Vandeputte et al. 2017; Props et al.
-2017) .
-
-Data that are generated by sequencing come from systems where scale is
-usually important and may be a confounding variable (Lovell et al.
-2015). For example, cells transformed by the cMyc oncogene have about 3
-times the amount of mRNA and about twice the rRNA content than do
-non-transformed cells (Nie et al. 2012), and this dramatically skews
-transcriptome analysis (Lovén et al. 2012). In addition, wild-type and
-mutant strains of cell lines, yeast or bacteria often have different
-growth rates, which would affect our ability to identify truly
-differentially abundant genes (Yoshikawa et al. 2011). As another
-example, the total bacterial load of the vaginal microbiome differs by
-1-2 orders of magnitude in absolute abundance between the healthy and
-bacterial vaginosis states (Zozaya-Hinchliffe et al. 2010); the
-composition is dramatically different as well (Ravel et al. 2011;
-Hummelen et al. 2010). Thus, the full description of these systems
-includes both relative change (composition) and absolute abundance
-(scale) but current methods access only the compositional information.
-
-More formally, if the true information in the environment being sampled
-can be summarized by counts of features (genes, taxa, etc) in a matrix
-\(\mathbf{W}\), then the elements of the matrix can be decomposed into
-its composition (relative information, \(\mathbf{W}^{\parallel}\) ) and
-its scale (total sum, \(\mathbf{W}^{\perp}\) ); formally,
-\(\mathbf{W} = \mathbf{W}^{\parallel} \mathbf{W}^{\perp}\). If the
-matrix has \(N\) samples and \(D\) parts, then we can uniquely identify
-\(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\). Any data
-that we collect by sequencing is thought to contain only proportional
-(Lovell et al. 2011) information and we can represent the corresponding
-sequenced matrix as \(\mathbf{Y}\). Compositional data analysis makes
-the strong assumption that
-\(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\);
-in other words that what is measured by sequencing is relative
-information only, and that this measurement is a function of the
-relative information in the underlying environment. Note that under this
-assumption no corresponding estimate or information is available for the
-scale of the sample \(\mathbf{Y}^{\perp}_{n}\).
-
-One issue that was realized very early was that if the goal was to
-compare one sample to another then the output from HTS needed to be
-normalized in order to make the samples commensurate and to correct any
-minor asymmetries in the data. A large number of normalizations were
-developed. Rarefaction(Hughes and Hellmann 2005) was used in the
-microbiome field, while transcriptome investigators developed and used
-proportions and derivatives (reads per kilobase per million, and
-transcripts per kilobase per million(Mortazavi et al. 2008; Wagner, Kin,
-and Lynch 2012)), the relative log expression (RLE)(Anders and Huber
-2010) and trimmed mean of M values (TMM)(Robinson and Oshlack 2010). The
-centered log ratio (CLR)(Aitchison 1982) approach was proposed as a
-unifying normalization(Fernandes et al. 2014). With the exception of
-rarefaction, all of these normalizations are ratios with the major
-differences between approaches being how the denominator is chosen and
-whether the ratio is always log transformed or not.
-
-Recently, Nixon et al. (2023) showed that that \emph{all} normalizations
-make some identifying assumption about the scale of \(\mathbf{W}\) but
-that the actual assumption being made was not always easy to interpret.
-In essence, normalizations in widespread use assume that either one
-sample can be chosen as a reference to which the others are scaled
-(TMM), or that different sub-parts of each sample maintain a constant
-scale across samples (RLE, CSS, CLR and derivatives).
-
-The CLR, which can be written as
-\(\mathbf{CLR}_{(1...d)n} = \mathbf{Y}_{(1...d)n} / G_n\), uses the
-geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the
-data. Nixon et al. (2023) discovered that using the CLR was equivalent
-to making a statement that \(G_n\) was related to the scale of the
-underlying environment \(\mathbf{W}^{\perp}\). Interestingly, the
-information in a system is a fundamental property and \(G_n\) is
-strongly correlated with Shannon's entropy, indicating that information
-about the complexity of \(\mathbf{W}\) is available post-sequencing
-(Supplemental). The realization that all normalizations made assumptions
-of scale led to the understanding that all normalizations introduced
-latent bias into analyses because the scaling factor has been assumed up
-to now to be a perfect point estimate of the scale of the underlying
-system (Nixon et al. 2023).
-
-With this new insight Nixon et al. (2023) devised methods to set up a
-scale model that explicitly acknowledges and accounts for the
-uncertainty of the scale. In the case of the CLR which uses the \(G_n\)
-as the denominator, this entails adding uncertainty into \(G_n\) such
-that the latent bias can be exposed, acknowledged and accounted for.
-Additionally it was realized that the actual values used in the
-denominator to calculate the log-ratio were irrelevant, but instead that
-the ratios between the denominators was the key parameter. These two
-insights allow for a full scale model of scale to be built and tested
-for its effect on analysis.
-
-The ALDEx2 R package (Fernandes et al. 2013) represents a general
-purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was
-designed originally to convert the discrete point estimates in the count
-matrix obtained into continuous posterior distributions through
-Monte-Carlo sampling from the Dirichlet distribution. The posterior more
-accurately represents variation on a per-sample basis(Gloor et al. 2016)
-and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate
-\(\mathbf{W}^{\parallel}_{dn}\). Each Monte-Carlo replicate is
-normalized using the CLR and used to calculate summary and test
-statistics, and finally expected values and confidence intervals are
-reported across the Monte-Carlo replicates. The CLR calculation can be
-modified trivially to include a full scale model of the data by
-incorporating uncertainty into \(G\) and so produce a posterior estimate
-of the scale. The scale uncertainty can be included on a per-experiment,
-per-group or even a per-sample basis. Thus, ALDEx2 can now be used to
-generate a full posterior model of both \(\mathbf{W}^{\parallel}_{dn}\)
-and \(\mathbf{W}^{\perp}_{dn}\) from the observed data.
+High-throughput sequencing (HTS) is a ubiquitous tool used to explore
+many biological phenomenon such as gene expression (single-cell
+sequencing, RNA-sequencing, meta-transcriptomics), microbial community
+composition (16S rRNA gene sequencing, shotgun metagenomics) and
+differential enzyme activity (selex, CRISPR killing). HTS proceeds by
+taking a sample from the environment, making a library, multiplexing
+(merging) multiple libraries together, and then applying a sample of the
+multiplexed library to the flow cell. Each of these steps can be though
+of as a sampling step: a pool of DNA that could be measured is sampled
+and only that subsampled DNA is carried over to subsequent steps. Within
+each sampling step, the connection between the size of the sampled DNA
+pool and the scale (e.g., size, microbial load, or total gene
+expression) of the measured biological system is degraded or lost.
+Increasingly, researchers are turning to modified experimental protocols
+(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
+biological variation in scale; see for example (Vandeputte et al. 2017;
+Props et al. 2017). However, these protocols often do not recover
+meaningful biological variation but instead recover scale variation at
+some intermediate step in the sample preparation protocol (e.g., at the
+stage where the DNA spike-in was added). In short, the disconnect
+between sample-to-sample variation in sequencing depth and biological
+variation in scale remains an outstanding challenge.
+
+Biological variation in scale can represent an important unmeasured
+confounder in HTS analyses (Lovell et al. 2015). For example, cells
+transformed by the cMyc oncogene have about 3 times the amount of mRNA
+and about twice the rRNA content than non-transformed cells (Nie et al.
+2012), and this dramatically skews transcriptome analysis (Lovén et al.
+2012). In addition, wild-type and mutant strains of cell lines, yeast or
+bacteria often have different growth rates, which would affect our
+ability to identify truly differentially abundant genes (Yoshikawa et
+al. 2011). As another example, the total bacterial load of the vaginal
+microbiome differs by 1-2 orders of magnitude in absolute abundance
+between the healthy and bacterial vaginosis states (Zozaya-Hinchliffe et
+al. 2010), and the composition between these states is dramatically
+different (Ravel et al. 2011; Hummelen et al. 2010). Thus, a full
+description of these systems includes both relative change (composition)
+and absolute abundance (scale), yet current methods access only the
+compositional information.
+
+To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix
+of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number
+of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or
+gene) in sample \(n\). Let \(\mathbf{W}\) also represent a
+\(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing
+the true (as opposed to measured) amount of class \(d\) in the
+biological system from which sample \(n\) was obtained. Succinctly,
+\(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\)
+denotes the true counts in the system under study. We can think of
+\(\mathbf{W}\) as consisting of two parts, the scale
+\(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
+\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
+\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
+\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
+\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
+\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
+Note that with these definitions \(\mathbf{W}\) can be written as the
+element-wise combination of scale and composition:
+\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
+The technical variation in sequencing depth
+(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
+observed data \(\mathbf{Y}\) provides us with information about the
+system composition \(\mathbf{W}^{\parallel}\) but little to no
+information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
+al.~2011).
+
+Early on, researchers recognized the potential confounding effects of
+the non-biological variation in sequencing depth and approached the
+problem using various forms of \textit{normalization}. The idea of
+normalization is to remove this variation thus making the samples
+commensurate and to correct any minor asymmetries in the data. A large
+number of normalizations were developed. Rarefaction (Hughes and
+Hellmann 2005) was used in the microbiome field, while transcriptome
+investigators developed and used proportions and derivatives (reads per
+kilobase per million, and transcripts per kilobase per million)
+(Mortazavi et al. 2008; Wagner, Kin, and Lynch 2012), the relative log
+expression (RLE) (Anders and Huber 2010) and trimmed mean of M values
+(TMM) (Robinson and Oshlack 2010). The Centered Log Ratio (CLR)
+(Aitchison 1982) was proposed as a unifying normalization (Fernandes et
+al. 2014). With the exception of rarefaction, all of these
+normalizations are data transformations that can be stated as ratios of
+the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the
+major differences being how the denominator is chosen and whether the
+ratio is subsequently always log-transformed or not.
+
+Recently, Nixon et al. (2023) suggested that the challenge of
+non-biological variation in sequencing depth be viewed as a problem of
+partially-identified models. Through this lens, they showed that
+\emph{all} normalizations make some assumption about the system scale
+\(\mathbf{W}^{\perp}\) in order to identify the chosen model. Moreover,
+these assumptions are often implicit, difficult to interpret, and
+provided different outputs when applied to the same dataset (Dillies et
+al. 2013; Weiss et al. 2017). Intuitively, normalizations in widespread
+use assume that either one sample can be chosen as a reference to which
+the others are scaled (TMM), or that different sub-parts of each sample
+maintain a constant scale across samples (RLE, CLR and derivatives).
+Further, Nixon et al. (2023) showed that the original ALDEx2 model made
+a strict assumption about scale through the CLR normalization. They
+showed that this assumption could be remedied since ALDEx2 permits scale
+to be easily incorporated into its framework. In this paper, we discuss
+these modifications and show how including scale uncertainty can greatly
+improve modeling and lead to more robust and reproducible results.
+
+\hypertarget{adding-scale-uncertainty-in-aldex2}{%
+\subsection{Adding Scale Uncertainty in
+ALDEx2}\label{adding-scale-uncertainty-in-aldex2}}
+
+The ALDEx2 R package (Fernandes et al. 2013) is a general purpose
+toolbox for Bayesian modeling of HTS data. For brevity and concreteness,
+we discuss ALDEx2 in its simplest form as a tool for estimating
+Log-Fold-Changes (LFC) (e.g., differential abundance or differential
+expression analysis), but note that it can be used to fit more complex
+linear models. At a high-level, ALDEx2 involves three steps: 1) a
+Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given
+observations \(\mathbf{Y}\); 2) a normalization is used to estimate
+\(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the
+\(\mathbf{W}\) estimates are used to estimate LFC as part of
+differential abundance (or expression) analyses. For more details on
+ALDEx2 see Fernandes et al. (2013).
+
+By default, ALDEx2 uses the CLR normalization which can be written as
+\(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where
+\(G_{n}\) denotes the geometric mean of the vector
+\((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental
+Materials for how the CLR connects to Shannon entropy). Nixon et
+al.~(2023) showed that in this context the CLR normalization equates to
+an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\). They
+showed that even slight errors in this assumption introduced bias into
+LFC estimate that was not accounted for in estimates of uncertainty
+(e.g., confidence intervals or credible sets). In fact, they showed that
+the only way in which the ALDEx2 model, or any model that uses
+normalizations, could ever be calibrated (e.g., control Type-I Error
+rates) was if this assumption was exactly true. While this assumption
+may be approximately correct in some instances it is surely never
+exactly true. In support of this assertion it is known that false
+discovery rates vary widely by analysis and normalization method
+(Hawinkel et al. 2018; Li et al. 2022). As a result, Nixon et al.~(2023)
+suggested that models should consider potential error in the assumptions
+implied by normalizations.
+
+Nixon et al.~(2023) proposed the concept of a \textit{scale model} as a
+generalization of normalizations that account for potential error. They
+showed how scale models can be incorporated into ALDEx2, turning the
+ALDEx2 model into a specialized type of statistical model which they
+called a \textit{Scale Simulation Random   Variable} (SSRV). Recall that
+\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
+As the HTS data only informs directly on
+\(\mathbf{W}^{\parallel}_{dn}\), they proposed to augment analyses with
+a model for \(\mathbf{W}^{\perp}_{n}\) (the scale model). Since the CLR
+normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\),
+the CLR normalization can be generalized by considering probability
+models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean
+\(1/G_{n}\). For example, the following scale model generalizes the CLR:
+
+\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
+
+where \(\gamma\) is a tunable parameter that controls the degree of
+uncertainty in the CLR assumption and \(x_{n}\) denotes a binary
+condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\)
+denotes control). Beyond theoretical studies, they demonstrated through
+the analysis of simulated and real data the remarkable performance
+improvements that could result from incorporating this type of scale
+uncertainty into analyses. Subsequently, we have made those
+modifications a permanent fixture of ALDEx2 which now represents the
+first software package designed for SSRV-based inference.
+
+Beyond concerns of fidelity and rigor, scale models also enhance the
+reproducibility and transparency of HTS analyses. This occurs because
+rather than using an implicit assumption (e.g.,
+\(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale
+models make this an explicit part of the model building process. While
+it is beyond the scope of the present article, we note that there are
+many ways of building scale models that enhance the interpretability of
+the parameters and assumptions. Moreover, the scale model listed above,
+which is now the default in ALDEx2, also has an appealing
+interpretation. A detailed description of these points as well as the
+mathematical underpinnings of the updated ALDEx2 R package is
+forthcoming and will be made available as a companion publication. Here,
+we simply treat the parameter \(\gamma\) in the default scale model as
+an abstract term that represents uncertainty in the CLR assumption and
+note that larger values of \(\gamma\) correspond to more uncertainty in
+this assumption. We also show that the ratio of scales between
+conditions can be used to build a full scale model when the conditions
+have dramatically different underlying scales.
 
@@
 
-The first dataset is a highly replicated yeast transcriptome where one
-condition is wild-type and the other has a snf-1 gene
-knockout(Gierliński et al. 2015). Yeast deficient for snf-1 grow more
-slowly and are sensitive to a variety of common agents that cause cell
-stress (Yoshikawa et al. 2011). This dataset has been used to argue that
-only a small number of replicates need to be used to identity
-differentially abundant genes and that different tools should be used
-for datasets with different sample sizes because the tools have
-different intrinsic statistical power and Type 1 and Type 2 errors
-(Schurch et al. 2016). This guidance runs counter to standard
-statistical practice where power is intrinsically linked to sample size
-(Halsey et al. 2015), yet the concept of sample-size independent power
-is entrenched in all fields that use HTS as an experimental readout.
-Increasing false precision with increasing sample size can be understood
-as the result of unacknowledged bias brought about because of false
-certainty as to the scale of the data. Thus, by adding scale uncertainty
-we can be more confident that the analysis is not converging on a
-precise but biased estimate of scale (Gustafson 2015).
-
-We start by examining the dispersion or variance of the data as counts
-and as the logarithm of the normalized counts as calculated by DESeq2
-(log(RLE)) and by ALDEx2 (CLR). The actual counts of data derived from
+\hypertarget{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}{%
+\subsection{Adding scale uncertainty replaces the need for dual
+significance
+cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}}
+
+Gierliński et al. (2015) conducted a highly replicated yeast
+transcriptome experiment comparing a wild-type strain with a snf1 gene
+knockout. This dataset has been used to argue that only a small number
+of replicates are needed to identity differentially abundant genes and
+makes the recommendation that some tools should be used in some
+conditions because each tool has different intrinsic statistical power
+and Type 1 and Type 2 errors (Schurch et al. 2016){]}. This guidance
+runs counter to standard statistical practice where power is
+intrinsically linked to sample size (Halsey et al. 2015), yet the
+concept of sample-size independent power is entrenched in all fields
+that use HTS as an experimental readout. Through the lens of this paper,
+this is a hallmark of unacknowledged bias: a false certainty in the
+scale of the data brings about an increasing false precision with
+increasing sample sizes. Furthermore, the yeast transcriptome experiment
+has an unacknowledged difference in scale between conditions: yeast
+deficient for snf1 are smaller, grow more slowly and are sensitive to a
+variety of common agents that cause cell stress (Yoshikawa et al.
+2011){]}. Thus, by adding scale uncertainty, we can be more confident
+that the results of the analysis reflect true biological differences and
+not differences due to the assumptions made by the normalization method.
+
+We start by examining the dispersion (variance) of the data as counts
+and as the logarithm of the normalized counts when calculated by DESeq2
+(log(RLE)) or by ALDEx2 (CLR). The raw data counts derived from
 sequencing are overdispersed with the mean value being less than the
-variance (Robinson, McCarthy, and Smyth 2010) as seen in Panel A of
-Figure 1. This relationship is why most tools model the counts with the
-Negative Binomial distribution(Frazee et al. 2023), and use this
-distribution as the basis for batch correction(Zhang, Parmigiani, and
-Johnson 2020). However, the actual analysis of differential abundance is
-performed on the logarithm of the normalized counts (Robinson, McCarthy,
-and Smyth 2010; Love, Huber, and Anders 2014; Andrews and Hemberg 2019),
-or on the clr values (Fernandes et al. 2013) both of which are
-log-ratios. The mean-dispersion distribution of these log-ratio
+variance (Robinson and Oshlack 2010; Robinson, McCarthy, and Smyth 2010)
+as seen in Panel A of Figure 1. However, the actual analysis of
+differential abundance is performed on the logarithm of the normalized
+counts (Robinson, McCarthy, and Smyth 2010; Love, Huber, and Anders
+2014; Andrews and Hemberg 2019), which includes the CLR (Fernandes et
+al. 2013) yet the mean-dispersion distribution of these log-ratio
 transformed data is quite different as shown in panels B and C of Figure
@@
 
-Here we can see that while the variance or dispersion always increases
-with increasing raw read count, it actually decreases when measured on
-the log-ratio transformed data (Fernandes et al. 2013; Love, Huber, and
-Anders 2014) and reaches a minimum at some mid-point of the
-distribution. This makes the counter-intuitive suggestion that genes
-with moderate expression have more predictable expression than genes
-with very high expression such as housekeeping genes. This is at odds
-with the known biology of cells where single cell counting of
-housekeeping transcripts shows that they are both highly expressed and
-have little intrinsic variation (Taniguchi et al. 2010). Furthermore,
-the actual amount of dispersion is very small and for many transcripts
-is almost negligible. To show this point more clearly, the majority of
-the transcripts in the lowest decile of dispersion indicated below the
-dashed grey line are statistically significantly different (75\% with
-DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead
-to many false positives because of the underestimation of dispersion.
-The work of Nixon et al. (2023) suggests that this is because the biased
-scale estimates that are calculated. Thus, the actual variation of
-highly expressed genes is not captured accurately by current approaches
-to analysis and supports the idea that HTS have unacknowledged bias.
-
-Using either DESeq2 or ALDEx2, we observe that a majority of transcripts
-are statistically significantly different between groups even with a
+These relationships depend on what is being analyzed: the variance or
+dispersion always increases with increasing raw (or normalzied) read
+count but decreases when measured on the log-ratio transformed data
+(Fernandes et al. 2013; Love, Huber, and Anders 2014), reaching a
+minimum at some mid-point of the distribution. This makes the
+counter-intuitive suggestion that genes with moderate expression have
+more predictable expression than genes with very high expression such as
+housekeeping genes. This is at odds with the known biology of cells
+where single cell counting of housekeeping transcripts shows that they
+are both highly expressed and have little intrinsic variation (Taniguchi
+et al. 2010). Furthermore, the dispersion is exceeding small being, for
+many transcripts, almost negligible. To show this point more clearly,
+the majority of the transcripts in the lowest decile of dispersion
+indicated below the dashed grey line are statistically significantly
+different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low
+dispersion estimates lead to many false positives. From the perspective
+of Nixon et al. (2023), it is reasonable to conclude that some of these
+results may be due to the choice of normalization. Indeed, benchmarking
+and comparison studies have repeatedly shown that the choice of
+normalization plays a role in which results are returned as significant;
+see for example (Maza et al. 2013), (Dillies et al. 2013) and (Weiss et
+al. 2017).
+
+Using either DESeq2 or ALDEx2, a majority of transcripts are
+statistically significantly different between groups even with a
 Benjamini-Hochberg (Benjamini and Hochberg 1995) false discovery rate
 (FDR) of 0.01; i.e.~4264 or 3790 of the 5891 transcripts are
-statistically significant. Clearly, the necessary assumption of most
-features being invariant is not justified. Applying the rule of thumb of
-at least a \(2^{1.4}\) fold change (Schurch et al. 2016) reduces these
-outputs to 193 for DESeq2 and to 186 for ALDEx2.
+statistically significant. Clearly, a 70\% positive rate seems
+biologically unjustified, and furthermore this breaks the necessary
+assumption made by the normalization methods that most features must be
+invariant. Finally, that 90 transcripts are identified by ALDEx2 and not
+DESeq2, while DESeq2 identifies 564 transcripts that ALDEx2 does not
+suggests that the choice of normalization plays a role in which results
+are returned as significant.
 
@@
 
-As shown using the effect (Gloor, Macklaim, and Fernandes 2016) and
-volcano plots (Cui and Churchill 2003) in Figure 2 A,B,D,E the root
-cause of the many statistically significant positive transcripts is the
-very large number of transcripts with negligible variance with both
-DESeq2 and ALDEx2. We can see that almost all the transcripts that are
-differentially abundant with an FDR \textless{} 0.01 (orange and red
+As shown using effect (Gloor, Macklaim, and Fernandes 2016) and volcano
+plots (Cui and Churchill 2003) in Figure 2 A,B,D,E, the root cause of
+the many statistically significant positive transcripts is the very
+large number of transcripts with negligible variance. This phenomenon is
+shared between methods: almost all the transcripts that are
+differentially abundant with an FDR \textless 0.01 (orange and red
 points) have extremely low dispersion and a very low difference between
-groups. In the most extreme cases transcripts with near 0 difference
-have a low FDR. This issue lead to the common practice of choosing
-transcripts with a low FDR and a fold-change threshold (Schurch et al.
-2016) (commonly set at \(\pm 2^{1.4}\)), and these limits are shown by
-the dashed grey lines. A similar sitation arises when using the ALDEx2
-package, and indeed the two methods identify substantially similar
-transcripts. This results begs the question, ``why bother with the
-signficance test?''.
-
-The very low dispersion estimate for most of the features arises because
-scale variation in the underlying dataset has been removed through
+groups. In the most extreme cases, some transcripts have a near zero
+difference between groups. This issue is not unique to this data set and
+it is common practice to use a dual-cutoff by choosing transcripts based
+on a thresholds for both corrected p-values and fold-changes (Schurch et
+al. 2016) (commonly set at \(\pm 2^{1.4}\) for the latter). These limits
+are shown by the dashed grey lines. In this case, applying a dual-cutoff
+using a heuristic of at least a \(2^{1.4}\) fold change reduces the
+number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
+This result begs the question: why bother with significance tests at
+all?
+
+The very low dispersion estimate for most features is a by-product of
 sequencing and normalization. While the actual scale of the data is
-inaccessible post-sequencing we can estimate the \emph{effect} of scale
-on the output. To do this, we add uncertainty to the denominator that is
-used to calculate the log-ratio of the samples (Nixon et al. 2023), and
-combine this with the posterior probabilities that ALDEx2 calculates on
-a per-feature basis (Fernandes et al. 2013). Scale uncertainty is
-incorporated using the \texttt{gamma} parameter that controls the amount
-uncertainty being included when we call either \texttt{aldex()}, or
-\texttt{aldex.clr()}. The \texttt{ALDEx2} package further contains a
-sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be
-used to explore the effect of different amounts of scale uncertainty. In
-practice we suggest that a \texttt{gamma} parameter between 0.5 and 1 is
-realistic for most experimental designs.
-
-Applying \texttt{gamma=1} as a parameter we can see that the large
-number of transcripts with near 0 dispersion have had their dispersion
-increased (Figure 2C), and this results in many fewer transcripts (100)
+inaccessible post-sequencing, we can study the sensitivity of the
+results to the choice of normalization and thus scale. To do this we
+incorporate scale models (Nixon et al. 2023) in lieu of normalizations
+through the default scale model that is now built into ALDEx2. Scale
+uncertainty is incorporated using the \texttt{gamma} parameter that
+controls the amount uncertainty added to the CLR mean assumption when we
+call either \texttt{aldex()}, or \texttt{aldex.clr()}. The
+\texttt{ALDEx2} package further contains a sensitivity analysis
+function, \texttt{aldex.senAnalysis()}, that can be used to explore the
+effect of different amounts of scale uncertainty. In practice, we
+suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic
+for most experimental designs.
+
+Applying the default scale model with \texttt{gamma=1}, a large number
+of transcripts with near 0 dispersion have had their dispersion
+increased (Figure 2C), and this results in many fewer transcripts 100
 being called significantly different as shown in the volcano plot in
@@
 dual cutoff. Indeed, adding scale uncertainty reduces the significant
-transcripts to approximately the number observed with the somewhat
-arbitrary difference cutoff. Thus, incorporating scale uncertainty
+transcripts to a subset of those identified with the dual cutoff that
+have the largest effect size. Thus, incorporating scale uncertainty
 through the default scale model allows us to determine which variables
@@
 
-The second example dataset is a vaginal metatranscriptome dataset used
-in (Macklaim and Gloor 2018; Wu et al. 2021), where we are comparing
-gene expression in bacteria collected from healthy (H) and bacterial
-vaginosis (BV) affected women. In this environment, both the relative
-abundance of species between groups is different as is the gene
-expression levels within a species (Macklaim et al. 2013). We further
-know that the total number of bacteria is about 10X more in BV than in H
-(Zozaya-Hinchliffe et al. 2010). Thus, this is an extremely challenging
-environment in which to determine differential abundance. Indeed, the
-accepted method to not analyze the vaginal metatranscriptome as a whole
-system but instead to do the simpler taxon-by-taxon (Macklaim et al.
-2013; Deng et al. 2018; Fettweis et al. 2019). The problem with the
-systemic analysis is that the scale confounding results in many
-housekeeping genes as being differentially abundant between groups.
-
-In this example we show how to specify the scale model explicitly and
-demonstrate that applying different scale models to each group
-independently can control for the very large difference in scale in the
-underlying data. When specifying the whole scale model we can pass a
-matrix of scale values instead of a single estimate of gamma to
-\texttt{aldex.clr()}. This matrix should have the same number of rows as
-the of Monte-Carlo Dirichlet samples, and the same number of columns as
-the number of samples. There is an \texttt{aldex.makeScaleModel()}
-function that can be used to simplify this step.
-
-This scale model matrix encapsulates two aspects of scale uncertainty
-and explicitly illustrates the concept of scale. First, as in the first
-example, the gamma paramter indicates the scale uncertainty. Second, we
-need to explicitly indicate the difference in scale between the groups
-in the dataset.
+\hypertarget{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}{%
+\subsection{Housekeeping genes can be used to guide scale model
+choices.}\label{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}}
+
+Macklaim and Gloor (2018) and Wu et al. (2021) used a vaginal
+metatranscriptome dataset used to compare the gene expression in
+bacteria collected from healthy (H) and bacterial vaginosis (BV)
+affected women. In this environment, both the relative abundance of
+species between groups and the gene expression level within a species is
+different (Macklaim et al. 2013). Additionally, prior research suggests
+that the total number of bacteria is about 10 times more in the BV than
+in the H condition (Zozaya-Hinchliffe et al. 2010). Thus, this is an
+extremely challenging environment in which to determine differential
+abundance as there are both compositional and scale changes between
+conditions. The accepted method to analyze vaginal metratranscriptome
+data is on a taxon-by-taxon basis (Macklaim et al. 2013; Deng et al.
+2018; Fettweis et al. 2019) because the scale confounding can be
+ignored. The problem with a systemic analysis is that many housekeeping
+genes are returned as differentially abundant between groups, a result
+likely due to a disconnect between the scale assumptions of the
+normalization used (Wu et al. 2021) and the actual change in scale
+between conditions.
+
+In this example, we show how to specify the scale model explicitly and
+demonstrate that applying a user-defined scale model can account for
+some of these modeling difficulties. A user-defined scale model can
+control for both the mean difference of scale between groups (e.g.,
+directly incorporate information on the differences in total number of
+bacteria between the BV and H conditions) as well as the uncertainty
+assumed in that difference. To specify a user-defined scale model, we
+can pass a matrix of scale values instead of a single estimate of gamma
+to \texttt{aldex.clr()}. This matrix should have the same number of rows
+as the of Monte-Carlo Dirichlet samples, and the same number of columns
+as the number of samples. While this matrix can be computed from scratch
+by the analyst, there is an \texttt{aldex.makeScaleModel()} function
+that can be used to simplify this step in most cases.
 
@@
 approximately to grouping orthologous sequences regardless of the
-organism of origin. Each point represents one of the 3728 functions, and
-we can see that there are many more functions represented in the BV
-group (bottom) than in the healthy group (top). This is because the
-\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
-reduced genome content relative to the anaerobic organisms that dominate
-in BV, because there is a greater diversity of organisms in BV than in H
-samples and because the BV condition has at least an order of magnitude
-more bacteria than does the H condition.
-
-We can see that there are a large number of functions that are shared
-between the two groups (boxed area in Figure 3A), and inspection shows
-that these largely correspond to core metabolic functions that would not
-be expected to contribute to differences in ecosystem behaviour. As a
-proxy for housekeeping functions the core ribosome functions (blue)
-shows that their mean location is not centred on 0. The major group of
-these housekeeping functions is located off the line of no difference
-(being approximately located at +1.5) and not surprisingly have among
-the lowest dispersion in the dataset. Nevertheless, they are identified
-as differentially abundant (red) along with many others. While changes
-in the abundance of housekeeping functions is a useful proxy for
-relative abundance of species in the environment, they tell us nothing
-about the functional capacity of the two groups because these are
-functions in common to every organism. Of more interest is determining
-the functions that are different between groups because these are unique
-or over-expressed in one group relative to the other.
+organism of origin. Each point represents one of the 3728 functions.
+There are many more functions represented in the BV group (bottom) than
+in the healthy group (top). This is because the \textit{Lactobacilli}
+that dominate a healthy vaginal microbiome have reduced genome content
+relative to the anaerobic organisms that dominate in BV, because there
+is a greater diversity of organisms in BV than in H samples and because
+the BV condition has at least an order of magnitude more bacteria than
+does the H condition.
+
+We can see that there are a large number of core metabolic and
+information processing functions that are shared between the two groups
+(boxed area in Figure 3A), and inspection shows that these largely
+correspond to core metabolic functions that would not be expected to
+contribute to differences in ecosystem behaviour and so should not be
+differentially abundant. As a proxy for housekeeping functions the core
+ribosome functions (blue) shows that their mean location is not centred
+on 0. The major group of these housekeeping functions is located off the
+line of no difference (being approximately located at +1.5). Since they
+have very low dispersions, these functions are identified as
+differentially abundant (red) along with many others. While changes in
+the abundance of housekeeping functions is a useful proxy for relative
+abundance of species in the environment, they tell us nothing about the
+functional capacity of the two groups because these are functions in
+common to every organism. Of more interest is determining the functions
+that are different between groups because these are unique or
+over-expressed in one group relative to the other.
 
@@
 difference. Panel B shows the same data scaled with
-\texttt{gamma\ =\ 1}, which increase the minimum dispersion approximatly
-by one unit. Here the housekeeping functions from Panel A are colored
-cyan or blue for reference. Panel C shows the same data scaled with
-\texttt{gamma\ =\ 0.75} and a 0.15 fold difference in dispersion applied
-to the BV samples relative to the H samples. The orange functions are
-now statistically significant. Note that this shifts the midpoint of the
-housekeeping functions towards the midline.}
+\texttt{gamma\ =\ 0.5}, which increase the minimum dispersion
+approximatly by one unit. Here the housekeeping functions from Panel A
+are colored cyan or blue for reference. Panel C shows the same data
+scaled with \texttt{gamma\ =\ 0.5} and a 0.15 fold difference in
+dispersion applied to the BV samples relative to the H samples. The
+orange functions are now statistically significant. Note that this
+shifts the midpoint of the housekeeping functions towards the midline.}
 \end{figure}
 
-Applying the default scale model of \texttt{gamma=1} as before increases
-the dispersion as expected, but does little to move the large number of
-housekeeping functions toward the midline of no difference.
-Nevertheless, about 50\% of the housekeeping functions are no longer
-statistically significantly different.
-
-Up to this point, scale uncertainty has been applied uniformly to both
-conditions, but a user-defined scale adjustment can be applied to each
-condition, or even each sample independently through a custom scale
-matrix. For this we need to revisit the concept that all normalizations
-in widespread use are actually ratios with the denominator chosen by one
-of several methods. It is important to recognize that in some datasets
-the denominator itself is not crucial as all approaches give
-substantially similar results; the transcriptome example used previously
-is a good example of such an idealized dataset. This begs the question,
-``if it is not the denominator, what then is the important
-determinant?'' As shown by Nixon et al. (2023) it is usually the
-\emph{relationship} between the group scale values (denominators) that
-is important, not their actual value. This can be illustrated quite
-simply by starting with the mean ratio in \(G\) between groups for the
-yeast transcriptome dataset which are \(6.05\times^{-5}\) for snf-w and
-\(5.08\times^{-5}\); their ratio being 1.17, or a 0.17-fold difference.
-Using this information, we can recapitulate the differential abundance
-analysis in Figure 2B and 2C exactly by using setting the mean
-denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 1 as
-shown in Supplemental Figure 2. Altering this ratio moves the location
-of the data.
-
-With this background, we can understand how to apply a full scale to the
-metatranscriptome dataset with the \texttt{aldex.makeScaleMatrix()}
-function. This function uses a logNormal distribution when building the
-model so that ratios that are chosen are symmetrical. The full scale
-model option can be used to make a matrix of relative scales that are
-distinct for each group (or even each sample) along with their
-uncertainties. Applying a per-group relative differential scale of 0.15
-with a base scale of 1 moves the housekeeping functions to the midline
-of no difference (Figure 3C). Note that now a significant number of
-functions are differentially up in BV that were formerly classed as not
-different without scale, or when only a uniform scale was applied. These
-former false negatives are noted in orange in each panel. Inspection of
-the functions shows that these are largely missing from the
-Lactobacillus species and so should actually be captured as
-differentially abundant. Thus, applying differential scale allows us to
-distinguish between both false positives (housekeeping functions in
-cyan) and false negatives (orange functions) even in a very difficult to
-analyze dataset. We suggest that the default scale model alone is useful
-and should be normally applied when the data are well centred. However,
-when datasets are not well centred or when the investigator has prior
-information about the underlying biology, we advocate developing and
-using a full differential scale model.
+Applying the default scale model of \texttt{gamma=0.5} increases the
+dispersion as expected but does little to move the large number of
+housekeeping functions toward the midline of no difference. This is as
+expected: the mean of the default scale model is based on the CLR
+normalization so no shift would be expected over the original ALDEx2
+model. Nevertheless, about 50\% of the housekeeping functions are no
+longer statistically significantly different. Note that this change is
+simple to conduct, has no additional computational complexity and
+requires only a slight modification from the analyst.
+
+Up to this point, scale uncertainty has been applied as an extension of
+the CLR normalization via the default scale model, but a user-defined
+scale adjustment can be applied to each condition, or even each sample
+independently through a custom scale matrix. While not necessary in all
+datasets (e.g., the transcriptome example discussed in the previous
+section and see the Supplementary material), it can greatly improve
+modeling in certain cases, especially when the scale is highly
+asymmetric between conditions. In order to specify a scale model from
+scratch, we need to revisit the concept that all normalizations in
+widespread use are actually ratios with the denominator implied by the
+normalization. Therefore, we can easily deviate from a certain
+normalization (e.g., the geometric mean assumption implied by the CLR)
+by specifying a total model based on the mean difference between
+conditions. While knowing the mean difference between conditions may
+seem cumbersome in practice, it is the \emph{relationship} between the
+group scale values that is important, not their raw values (Nixon et
+al.~2023). This can be illustrated quite simply by starting with the
+mean ratio in \(G\) between groups for the yeast transcriptome dataset
+which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for
+WT; their ratio being 1.17, or a 0.17-fold difference. Using this
+information, we can recapitulate the differential abundance analysis in
+Figure 2B and 2C exactly by using setting the mean denominator of group
+1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental
+Figure 2. This ratio can be adjusted to alter the mean assumption placed
+on the group scale values.
+
+With this background, we can understand how to apply a user-defined
+scale model to the metratranscriptome dataset. While a user-defined
+scale model can be quite flexible with relative scales that are distinct
+for each group (or even each sample) along with their uncertainties,
+here we focus on using the \texttt{aldex.makeScaleMatrix()} function.
+This function uses a logNormal distribution to build a scale matrix
+given a user-specified mean difference between groups and uncertainty
+level. Applying a per-group relative differential scale of 0.15 moves
+the housekeeping functions to the midline of no difference, and applying
+a gamma of 0.5 provides the same dispersion as using a gamma of 0.5 with
+the CLR (Figure 3C). Note that now a significant number of functions are
+differentially up in BV that were formerly classed as not different
+without scale, or when only a default scale was applied. These former
+false negatives are noted in orange in each panel. Inspection of the
+functions shows that these are largely missing from the Lactobacillus
+species and so should actually be captured as differentially abundant.
+Thus, applying a differential scale allows us to distinguish between
+both false positives (housekeeping functions in cyan) and false
+negatives (orange functions) even in a very difficult to analyze
+dataset. We suggest that the default scale model is sufficient when the
+data are approximately centred. However, when datasets are not well
+centred or when the investigator has prior information about the
+underlying biology, we advocate developing and using a user-specified
+scale model.
 
@@
 
-Biological count data derived from HTS can be decomposed into two parts
+Biological count data derived from HTS can be decomposed into two parts:
 the relative (compositional) and the absolute (scale), and the product
-of these generates a fully scaled biological system (Nixon et al. 2023).
-Biological systems are both predictably variable and stochastic and
+of these generates a fully scaled biological system (Nixon et al.~2023).
+Biological systems are both predictably variable and stochastic, and
 current measurement methods that rely on high throughput sequencing fail
-to capture all of that variation. In the absence of information external
-to the sequencing run itself, no normalisation method can recapture any
-of the scale information (Lovén et al. 2012).
-
-Since the underlying scale of the system cannot be measured easily --
-methods such as flow cytometry (Vandeputte et al. 2017), spike-in probes
-and fluorescent in-situ hybridization (Lovén et al. 2012; Marguerat et
-al. 2012) have been used -- the effect on analysis can be modelled by
-including scale uncertainty in the analysis (Nixon et al. 2023). Nixon
-et al, showed that this can be done by including uncertainty in the
-denominator used for the normalization. The ALDEx2 R package is ideally
-suited for this since this tool builds a Bayesian posterior of the
-compositional component of the dataset at the outset and then conducts
-the analysis on that posterior. Adding scale uncertainty can be done at
-the same time thus producing a posterior model that incorporates both
-compositional and scale uncertainty, and this can be done without
-additional computational complexity. For this, the compositional
-uncertainty is sampled from a Dirichlet distribution, and the scale
-uncertainty can be sampled from any distribution depending on prior
-knowledge or preference, but ALDEx2 samples from a logNormal
-distribution by default.
+to capture all of that variation, particularly variation due to scale.
+In the absence of information external to the sequencing run itself, no
+normalisation method can recapture any of the scale information (Lovén
+et al.~2012). Even so, many methods such as flow cytometry (Vandeputte
+et al.~2017), spike-in probes and fluorescent in-situ hybridization
+(Lovén et al.~2012; Marguerat et al.~2012) have been proposed as a means
+to do this. In the context of the updated ALDEx2 software suite, these
+methods can be useful if they contain information on the particular
+scale the researcher is interested in.
+
+The ALDEx2 R package is readily amenable for incorporating scale
+uncertainty. Originally, this tool used the only the Dirichlet
+distribution to sample compositional uncertainty. Rather than applying
+the CLR normalization as in the original ALDEx2, scale uncertainty can
+be added through the use of a scale model with no additional
+computational complexity. Scale uncertainty can be sampled from any
+distribution depending on prior knowledge or preference. By default,
+ALDEx2 samples scale from a logNormal distribution inspired by the CLR
+normalization. However, there is the option to introduce a full scale
+model that encapsulates both uncertainty and asymmetry in underlying
+scale.
 
@@
 One issue is that analysis tools seem over-powered with even moderate
-sample sizes and different tools have different intrinsic power, Type 1
-and Type 2 error rates (Schurch et al. 2016). Using small sample sizes
+sample sizes, and different tools have different intrinsic power, Type 1
+and Type 2 error rates (Schurch et al.~2016) and fail to account for
+false discovery rates properly (Hawinkel et al. 2018; Li et al. 2022).
+Traditional recommendations would suggest that using small sample sizes
 in analysis leads to less reliability and reproducibility in analyses
 since surprisingly large sample sizes are needed to determine
-reproducible p-values (eg. (Halsey et al. 2015)). Thus, recommendations
-to use small sample sizes in multivariate datasets such as RNA-seq
-datasets are not supported by simple modelling in the univariate case.
-Another issue is that datasets are difficult to analyze when there they
-contain systematic asymmetry, with different tools exhibiting differing
-pathologies with these datasets (Robinson and Oshlack 2010; Wu et al.
-2021).
+reproducible p-values (e.g., Halsey et al. (2015)). However,
+recommendations to use small sample sizes are widespread in multivariate
+datasets such as RNA-seq datasets. We argue that this is a hallmark of
+unacknowledged bias: since normalizations are typically assumed to be
+correct with no uncertainty, smaller samples offer an unintentional
+buffer against this misspecification. Another issue is that datasets are
+difficult to analyze when they contain systematic asymmetry, with
+different tools exhibiting differing pathologies with these datasets
+(Robinson and Oshlack 2010; Wu et al. 2021).
 
@@
 (Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
-Quinn, Crowley, and Richardson 2018) and microbiomes(Thorsen et al.
+Quinn, Crowley, and Richardson 2018) and microbiomes (Thorsen et al.
 2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
@@
 for individual datasets or for specific types of analysis;
-e.g.~`Normalization and microbial differential abundance strategies
-depend upon data characteristics' (Weiss et al. 2017). However, another
-interpretation of these studies is that the specific assumptions of the
-tools and the way they are used by different groups can lead to results
-that include both false negative and false positive results, skewed
-towards the latter. There are now published examples of multiple large
-replication efforts in psychology (Open Science Collaboration 2015),
-cancer biology (Rodgers and Collings 2021), ecology and evolution (Gould
-and et.al, n.d.). In a nutshell, experiences from these studies indicate
-that published data typically have higher effect sizes and appear more
-significant than replication efforts with the same data, or with newly
-generated replication data (Rodgers and Collings 2021). Thus, most
-published data should be viewed with suspicion (Ioannidis 2005) unless
-efforts have been made to ensure that the results are robust
+e.g.,\textasciitilde`Normalization and microbial differential abundance
+strategies depend upon data characteristics' (Weiss et al. 2017).{]}
+However, another interpretation of such studies is that different
+normalizations can be better suited to different datasets by
+happenstance, but it is not obvious how the use of specific
+normalizations could be justified on a per dataset basis. There are now
+published examples of multiple large replication efforts in psychology
+(Open Science Collaboration 2015), cancer biology (Rodgers and Collings
+2021), ecology and evolution (Gould and et.al 2023). In a nutshell,
+experiences from these studies indicate that published data typically
+have higher effect sizes and appear more significant than replication
+efforts with the same data, or with newly generated replication data
+(Rodgers and Collings 2021). Thus, most published data should be viewed
+with suspicion (Ioannidis 2005) unless efforts have been made to ensure
+that the results are robust.
 
@@
 including scale uncertainty, since substantially similar transcripts are
-identified by the two approaches. However, while using the post-hoc
-difference cutoff is useful for differential abundance analysis it is
-not clear how this can be incorporated into other kinds of downstream
-analyses. Conversely analyses that include scale uncertainty are fully
-compatible with existing downstream analyses.
+identified by the two approaches. We believe that adding scale
+uncertainty leads to a more transparent analysis compared to the cutoff
+as it is easy to describe the conditions under which certain conclusions
+hold. Furthermore, while using the post-hoc difference cutoff is useful
+for differential abundance analysis, it is not clear how this can be
+incorporated into other kinds of downstream analyses. Conversely,
+analyses that include scale uncertainty are fully compatible with
+existing downstream analyses.
 
 In the case of asymmetry, the use of a user-specified scale model can be
-very useful for otherwise difficult to analyze datasets such as
+very useful for otherwise difficult-to-analyze datasets such as
 meta-transcriptomes and in-vitro selection datasets where the majority
@@
 cannot fully move all the housekeeping genes to the midline of no
-difference or exhibit other pathologies (supplemental). Incorporating
+difference or exhibit other pathologies (Supplemental). Incorporating
 differential scale on a per-group basis moves the mass of the data
@@
 their impact on the analysis weighted accordingly. Additionally, the use
-of differential scale models permits difficult to analyze datasets to be
+of user-defined scale models permits difficult to analyze datasets to be
 examined in a robust and principled manner even when the majority of
@@
 
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
 \hypertarget{refs}{}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Dillies:2013}{}}%
+Dillies, Marie-Agnès, Andrea Rau, Julie Aubert, Christelle
+Hennequet-Antier, Marine Jeanmougin, Nicolas Servant, Céline Keime, et
+al. 2013. {``A Comprehensive Evaluation of Normalization Methods for
+{Illumina} High-Throughput {RNA} Sequencing Data Analysis.''}
+\emph{Brief Bioinform} 14 (6): 671--83.
+\url{https://doi.org/10.1093/bib/bbs046}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-fernandes:2013}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-polyester:2016}{}}%
-Frazee, Alyssa C., Andrew E. Jaffe, Rory Kirchner, and Jeffrey T. Leek.
-2023. {``{polyester: Simulate RNA-seq reads}.''}
-\url{https://doi.org/10.18129/B9.bioc.polyester}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-Gierlinski:2015aa}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gloorAJS:2016}{}}%
-Gloor, GB, JM Macklaim, M Vu, and AD Fernandes. 2016. {``Compositional
-Uncertainty Should Not Be Ignored in High-Throughput Sequencing Data
-Analysis.''} \emph{Austrian Journal of Statistics} 45: 73--87.
-\url{https://doi.org/doi:10.17713/ajs.v45i4.122}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-evoRepr-pre}{}}%
-Gould, E, and et.al. n.d. {``Same Data, Different Analysts: Variation in
-Effect Sizes Due to Analytical Decisions in Ecology and Evolutionary
+Gould, E, and et.al. 2023. {``Same Data, Different Analysts: Variation
+in Effect Sizes Due to Analytical Decisions in Ecology and Evolutionary
 Biology.''} \emph{EcoEvoRxiv}.
 https://doi.org/\url{https://doi.org/10.32942/X2GG62}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gustafson2015bayesian}{}}%
-Gustafson, Paul. 2015. \emph{Bayesian Inference for Partially Identified
-Models: Exploring the Limits of Limited Data}. Vol. 140. CRC Press.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Li:2022aa}{}}%
+Li, Yumei, Xinzhou Ge, Fanglue Peng, Wei Li, and Jingyi Jessica Li.
+2022. {``Exaggerated False Positives by Popular Differential Expression
+Methods When Analyzing Human Population Samples.''} \emph{Genome Biol}
+23 (1): 79. \url{https://doi.org/10.1186/s13059-022-02648-4}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Love:2014aa}{}}%
@@
 \url{https://doi.org/10.1186/s13059-014-0550-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lovell:2011}{}}%
-Lovell, David, Warren Müller, Jen Taylor, Alec Zwart, and Chris
-Helliwell. 2011. {``Proportions, Percentages, Ppm: Do the Molecular
-Biosciences Treat Compositional Data Right?''} In \emph{Compositional
-Data Analysis: Theory and Applications}, edited by Vera Pawlowsky-Glahn
-and Antonella Buccianti, 193--207. London: John Wiley; Sons New York,
-NY.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-yeast-absolute}{}}%
-Marguerat, Samuel, Alexander Schmidt, Sandra Codlin, Wei Chen, Ruedi
-Aebersold, and Jürg Bähler. 2012. {``Quantitative Analysis of Fission
-Yeast Transcriptomes and Proteomes in Proliferating and Quiescent
-Cells.''} \emph{Cell} 151 (3): 671--83.
-\url{https://doi.org/10.1016/j.cell.2012.09.019}.
+\leavevmode\vadjust pre{\hypertarget{ref-maza2013}{}}%
+Maza, Elie, Pierre Frasse, Pavel Senin, Mondher Bouzayen, and Mohamed
+Zouine. 2013. {``Comparison of Normalization Methods for Differential
+Gene Expression Analysis in RNA-Seq Experiments: A Matter of Relative
+Size of Studied Transcriptomes.''} \emph{Commun Integr Biol} 6 (6):
+e25849. \url{https://doi.org/10.4161/cib.25849}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Zhang:2020ab}{}}%
-Zhang, Yuqing, Giovanni Parmigiani, and W Evan Johnson. 2020.
-{``ComBat-Seq: Batch Effect Adjustment for RNA-Seq Count Data.''}
-\emph{NAR Genom Bioinform} 2 (3): lqaa078.
-\url{https://doi.org/10.1093/nargab/lqaa078}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-Zozaya:2010}{}}%

### 2023-10-21 go3.Rmd
submitted version

--- 2023-10-01 go3.Rmd
@@
 ---
-title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
+title: "Normalizations are not what you think; Explicit Scale Simulation in ALDEx2
 "
@@
   - name: Justin Silverman
-  affiliation: College of Information Sciences and Technology, Department of Statistics, Department of Medicine, Pennsylvania  State University
+    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
 output:
-  BiocStyle::pdf_document: default
+  nucleic-acids-research.csl: pdf_document
 package: ALDEx2
-abstract: |
-    Introduction to scale simulation and FDR correction with ALDEx2.
+abstract: In many high-throughput sequencing (HTS) studies, sample-to-sample variation  in sequencing depth is  not driven by  variation in the scale (e.g., total size, microbial load, or total gene  expression) of the underlying biological systems being measured but rather is driven by technical factors.
+Typically the technical variation  is addressed using some form of statistical normalization.
+Normalizations are data or parameter transformations that remove unwanted  technical variation in the hopes of facilitating analyses sensitive to scale;  e.g., differential abundance and differential expression analyses.
+Recently we  showed that any normalization makes implicit assumptions about the unmeasured  system scale and that errors in these assumptions can lead to dramatic increases in  false positive and false negative rates.
+Here we describe updates to the ALDEx2 R  package that mitigate these problems by directly modeling uncertainty in the  unmeasured system scale through the use of a \textit{scale model}.
+Scale  models generalize the idea of normalizations and can be thought of as  explicitly modeling potential error in the chosen normalization.
+Beyond  enhancing the robustness of HTS analyses, the use of scale models within ALDEx2   enhances the transparency and reproducibility of analyses by making implicit  normalizing assumptions an explicit part of the model building process.
 ---
@@
 
-High throughput sequencing (HTS) is a universal tool to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing).
-HTS proceeds by taking a fixed-number sample from the environment, making a library, multiplexing (merging) multiple libraries and applying a fixed number of molecules to the flow cell.
-In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken.
-It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment[@Lovell:2015;@Vandeputte:2017aa;@Props:2017aa].
+High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing).
+HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to the flow cell.
+Each of these steps can be though of as a sampling step: a pool of DNA that could be measured is sampled and only that subsampled DNA is carried over to subsequent steps.
+Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., size, microbial load, or total gene expression) of the measured biological system is degraded or lost.
+Increasingly, researchers are turning to modified experimental protocols (e.g., DNA spike-ins or cell counting) in an attempt to recover the lost biological variation in scale; see for example [@Vandeputte:2017aa;@Props:2017aa].
+However, these protocols often do not recover meaningful biological variation but instead recover scale variation at some intermediate step in the sample preparation protocol (e.g., at the stage where the DNA spike-in was added).
+In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an outstanding challenge.
 
-Data that are generated by sequencing come from systems where scale is usually important and may be a confounding variable [@Lovell:2015].
-For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than do non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa].
+Biological variation in scale can represent an important unmeasured confounder in HTS analyses [@Lovell:2015].
+For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa].
 In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa].
-As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010]; the composition is dramatically different as well [@Ravel:2010;@Hummelen:2010].
-Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but current methods access only the compositional information.
+As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010].
+Thus, a full description of these systems includes both relative change (composition) and absolute abundance (scale), yet current methods access only the compositional information.
 
-More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix \( \mathbf{W} \), then the elements of the matrix can be decomposed into its composition (relative information, \( \mathbf{W}^{\parallel} \) ) and its scale (total sum, \( \mathbf{W}^{\perp} \) ); formally, \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\).
-If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).
-Any data that we collect by sequencing is thought to contain only proportional [@lovell:2011] information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\).
-Compositional data analysis makes the strong assumption  that \(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\); in other words that what is measured by sequencing is relative information only, and that this measurement is a function of the relative information in the underlying environment.
-Note that under this  assumption no corresponding estimate or information is available for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).
+To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
+Let \(\mathbf{W}\) also represent a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true (as opposed to measured) amount of class \(d\) in the biological system from which sample \(n\) was obtained.
+Succinctly, \(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\) denotes the true counts in the system under study.
+We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions).
+That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
+Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
+The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al.
+2011).
 
-One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data.
+Early on, researchers recognized the potential confounding effects of the non-biological variation in sequencing depth and approached the problem using various forms of \textit{normalization}.
+The idea of normalization is to remove this variation thus making the samples commensurate and to correct any minor asymmetries in the data.
 A large number of normalizations were developed.
-Rarefaction[@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a].
-The centered log ratio (CLR)[@aitchison1982]  approach was proposed as a unifying normalization[@fernandes:2014].
-With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.
+Rarefaction [@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million) [@Mortazavi:2008; @wagner:tpm], the relative log expression (RLE) [@Anders:2010] and trimmed mean of M values (TMM) [@Robinson:2010a].
+The Centered Log Ratio (CLR) [@aitchison1982]  was proposed as a unifying normalization [@fernandes:2014].
+With the exception of rarefaction, all of these normalizations are data transformations that can be stated as ratios of the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the major differences being how the denominator is chosen and whether the ratio is subsequently always log-transformed or not.
 
 Recently, Nixon et al.
-[-@nixon2023scale] showed that that *all* normalizations make some  identifying assumption about the scale of \(\mathbf{Y}\) but that the actual assumption being made was not always easy to interpret.
-The CLR which can be written as \( \mathbf{CLR}_{dn} = \mathbf{Y}_{dn} / G_n \) uses the geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the data.
-They discovered that \(G_n\) was related to the scale of the underlying environment \( \mathbf{W\perp} \).
-Interestingly, the information in a system is a fundamental property and \(G_n\) is strongly correlated with Shannon's entropy, indicating that information about the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental).
-The realization that all normalizations made assumptions of scale led to the understanding that all normalizations  introduced latent bias into analyses because the scaling factor has been assumed up to now to be a perfect point estimate of the scale of the underlying system [@nixon2023scale].
+[-@nixon2023scale] suggested that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models.
+Through this lens, they showed that \emph{all} normalizations make some assumption about the system scale \(\mathbf{W}^{\perp}\) in order to identify the chosen model.
+Moreover, these assumptions are often implicit, difficult to interpret, and provided different outputs when applied to the same dataset [@Dillies:2013;@Weiss:2017aa].
+Intuitively, normalizations in widespread use assume that either one sample can be chosen as a reference to which the others are scaled (TMM), or that different sub-parts of each sample maintain a constant scale across samples (RLE, CLR and derivatives).
+Further, Nixon et al.
+[-@nixon2023scale] showed  that the original ALDEx2 model made a strict assumption about scale through the CLR normalization.
+They showed that this assumption could be remedied since ALDEx2 permits scale to be easily incorporated into its framework.
+In this paper, we discuss these modifications and show how including scale uncertainty can greatly improve modeling and lead to more robust and reproducible results.
 
-With this new insight Nixon et al.
-[-@nixon2023scale] devised methods to set up a scale model that explicitly acknowledges and accounts for the uncertainty of the scale.
-In the case of the CLR which uses the  \(G_n\) as the denominator, this entails adding uncertainty into  \(G_n\) such that the latent bias can be exposed, acknowledged and accounted for.
+## Adding Scale Uncertainty in ALDEx2
 
-The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets.
-ALDEx2 was designed originally to convert the discrete point estimates in the count matrix obtained into continuous posterior distributions through Monte-Carlo sampling from the Dirichlet distribution.
-The posterior more accurately represents variation on a per-sample basis[@gloorAJS:2016] and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate \(\mathbf{W}^{\parallel}_{dn}\).
-Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates.
-The CLR calculation can be modified trivially to include a full scale model of the data by incorporating uncertainty into \(G\) and so produce a posterior estimate of the scale.
-The scale uncertainty  can be included on a per-experiment, per-group or even a per-sample basis.
-Thus, ALDEx2 can now be used to generate a full posterior model of both \(\mathbf{Y}^{\parallel}_{dn}\) and \(\mathbf{Y}^{\perp}_{dn}\).
+The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data.
+For brevity and concreteness, we discuss ALDEx2 in its simplest form as a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models.
+At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the \(\mathbf{W}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses.
+For more details on ALDEx2 see Fernandes et al.
+[-@fernandes:2013].
+
+By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where \(G_{n}\) denotes the geometric mean of the vector \((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental Materials for how the CLR connects to Shannon entropy).
+Nixon et al.
+(2023) showed that in this context the CLR normalization equates to an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\).
+They showed that even slight errors in this assumption introduced bias into LFC estimate that was not accounted for in estimates of uncertainty (e.g., confidence intervals or credible sets).
+In fact, they showed that the only way in which the ALDEx2 model, or any model that uses  normalizations, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.
+While this assumption may be approximately correct in some instances it is surely never exactly true.
+In support of this assertion it is known that false discovery rates vary widely by analysis and normalization method [@hawinkel2017;@Li:2022aa].
+As a result, Nixon et al.
+(2023) suggested that models should consider potential error in the assumptions implied by normalizations.
+
+Nixon et al.
+(2023) proposed the concept of a \textit{scale model} as a generalization of normalizations that account for potential error.
+They showed how scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV).
+Recall that \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
+As the HTS data only informs directly on $\mathbf{W}^{\parallel}_{dn}$, they proposed to augment analyses with a model for $\mathbf{W}^{\perp}_{n}$ (the scale model).
+Since the CLR normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean \(1/G_{n}\).
+For example, the following scale model generalizes the CLR:
+
+\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
+
+where \(\gamma\) is a tunable parameter that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control).
+Beyond theoretical studies, they demonstrated through the analysis of simulated and real data the remarkable performance improvements that could result from incorporating this type of scale uncertainty into analyses.
+Subsequently, we have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.
+
+Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses.
+This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models make this an explicit part of the model building process.
+While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions.
+Moreover, the scale model listed above, which is now the default in ALDEx2, also has an appealing interpretation.
+A detailed description of these points as well as the mathematical underpinnings of the updated ALDEx2 R package is forthcoming and will be made available as a companion publication.
+Here, we simply treat the parameter \(\gamma\) in the default scale model as an abstract term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption.
+We also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.
 
@@
 
-The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knockout[@Gierlinski:2015aa].
-Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
-This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes and that different tools should be used for datasets with different sample sizes because the tools have different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa].
+## Adding scale uncertainty replaces the need for dual significance cutoffs.
+
+Gierliński et al.
+[-@Gierlinski:2015aa] conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf1 gene knockout.
+This dataset has been used to argue that only a small number of replicates are needed to identity differentially abundant genes and makes the recommendation that some tools should be used in some conditions because each tool has different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]].
 This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout.
-Increasing false precision with increasing sample size can be understood as the result of unacknowledged bias brought about because of false certainty as to the scale of the data.
-Thus, by adding scale uncertainty we can be more confident that the analysis is not converging on a precise but biased estimate of scale [@gustafson2015bayesian].
+Through the lens of this paper, this is a hallmark of unacknowledged bias: a false certainty in the scale of the data brings about an increasing false precision with increasing sample sizes.
+Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]].
+Thus, by adding scale uncertainty, we can be more confident that the results of the analysis reflect true biological differences and not differences due to the assumptions made by the normalization method.
 
-We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (CLR).
-The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1.
-This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab].
-However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa; @Andrews:2019aa], or on  the clr values [@fernandes:2013] both of which are log-ratios.
-The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa].
+We start by examining the dispersion (variance) of the data as counts and as the logarithm of the normalized counts when calculated by DESeq2 (log(RLE)) or by ALDEx2 (CLR).
+The raw data counts derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010a;@Robinson:2010] as seen in Panel A of Figure 1.
+However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010;@Love:2014aa;@Andrews:2019aa], which includes the CLR [@fernandes:2013] yet the mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013;@Love:2014aa].
 
-```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values.
+```{r, echo=F, warning=F, message=F,comment=F, result=F} 
+library(DESeq2)
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+#library(ALDEx2)
+```
+```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as clr values.
 Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale.
@@
 lines(lowess(def, f=0.1), col=2, lwd=2)
-title('A: Mean vs. Variance', adj=0, line= 0.8)
+title('A: M v. V', adj=0, line= 0.8)
 abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
@@
  abline(h=0.0275, col='grey', lty=2)
-title('B: log2 Mean vs. log2 Variance', adj=0, line= 0.8)
+title('B: log2(M) v. log2(V)', adj=0, line= 0.8)
 
@@
 plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
-  xlab='E(clr)', ylab='E(dispersion)')
+  xlab='E(clr)', ylab='E(disp)')
 #points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
@@
 
-Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data [@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution.
+These relationships depend on what is being analyzed: the variance or dispersion always increases with increasing raw (or normalzied) read count but decreases when measured on the log-ratio transformed data [@fernandes:2013;@Love:2014aa], reaching a minimum at some mid-point of the distribution.
 This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes.
 This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa].
-Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible.
-To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion.
-The work of Nixon et al.
-[-@nixon2023scale] suggests that this is because the biased scale estimates that are  calculated.
-Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias.
+Furthermore, the dispersion is exceeding small being, for many transcripts, almost negligible.
+To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead to many false positives.
+From the perspective of Nixon et al.
+[-@nixon2023scale], it is reasonable to conclude that some of these results may be due to the choice of normalization.
+Indeed, benchmarking and comparison studies have repeatedly shown that the choice of normalization plays a role in which results are returned as significant; see for example [@maza2013], [@Dillies:2013] and [@Weiss:2017aa].
 
-Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e.
+Using either DESeq2 or ALDEx2,  a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e.
 `r length(sig.des)` or `r length(sig.ald)` of the `r nrow(yst)` transcripts are statistically significant.
-Clearly, the necessary assumption of most features being invariant is not justified.
-Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.all$we.eBH < 0.01 & abs(x.all$diff.btw) > 1.4))` for ALDEx2.
+Clearly, a 70\% positive rate seems biologically unjustified, and furthermore this breaks the necessary assumption made by the normalization methods that most features must be invariant.
+Finally, that `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not suggests that the choice of normalization plays a role in which results are returned as significant.
 
@@
 
-As shown using the effect  [@gloor:effect] and volcano plots [@Cui:2003aa] in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2.
-We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups.
-In the most extreme cases transcripts with near 0 difference have a low FDR.
-This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold [@Schurch:2016aa] (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines.
-A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts.
-This results begs the question, "why bother with the signficance test?".
+As shown using effect [@gloor:effect] and volcano plots [@Cui:2003aa] in Figure 2 A,B,D,E, the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance.
+This phenomenon is shared between methods: almost all the transcripts that are differentially abundant with an FDR \textless 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups.
+In the most extreme cases, some transcripts have a near zero difference between groups.
+This issue is not unique to this data set and it is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both corrected p-values and fold-changes  [@Schurch:2016aa] (commonly set at \(\pm 2^{1.4}\) for the latter).
+These limits are shown by the dashed grey lines.
+In this case, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
+This result begs the question: why bother with significance tests at all?
 
-The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization.
-While the actual scale of the data is inaccessible post-sequencing  we can estimate the *effect* of scale on the output.
-To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013].
-Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`.
-The `ALDEx2` package further contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty.
-In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.
+The very low dispersion estimate for most features is a by-product of sequencing and normalization.
+While the actual scale of the data is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale.
+To do this we incorporate scale models [@nixon2023scale] in lieu of normalizations through the default scale model that is now built into ALDEx2.
+Scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
+The \texttt{ALDEx2} package further contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty.
+In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs.
 
-Applying `gamma=1` as a parameter we can see that the large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts (`r length(sig.s.ald)`) being called significantly different  as shown in the volcano plot in Figure 1F (red points).
+Applying the default scale model with \texttt{gamma=1}, a large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts `r length(sig.s.ald)` being called significantly different as shown in the volcano plot in Figure 1F (red points).
 Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.
-Indeed, adding scale uncertainty reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff.
+Indeed, adding scale uncertainty reduces the significant transcripts to a subset of those identified with the dual cutoff that have the largest effect size.
 Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.
 
-The second example dataset is a vaginal metatranscriptome dataset used in [@Macklaim:2018aa;@Wu2021], where we are comparing gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
-In this environment, both the relative abundance of species between groups is different as is the gene expression levels within a species [@macklaim:2013].
-We further know that the total number of bacteria is about 10X more in BV than in H [@Zozaya:2010].
-Thus, this is  an extremely challenging environment in which to determine differential abundance.
-Indeed, the accepted method to not analyze the vaginal metatranscriptome as a whole system but instead to do the simpler taxon-by-taxon [@macklaim:2013;@Denge00262-18;@Fettweis:2019aa].
-The problem with the systemic analysis is that the scale confounding results in many housekeeping genes as being differentially abundant between groups.
+## Housekeeping genes can be used to guide scale model choices.
 
-In this example we show how to specify the scale model explicitly and demonstrate that applying different scale models to each group independently can control for the very large difference in scale in the underlying data.
-When specifying the whole scale model we can pass a matrix of scale values instead of a single estimate of gamma to `aldex.clr()`.
+Macklaim and Gloor [-@Macklaim:2018aa] and Wu et al.
+[-@Wu2021] used a vaginal metatranscriptome dataset used to compare the gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
+In this environment, both the relative abundance of species between groups and the gene expression level within a species is different [@macklaim:2013].
+Additionally, prior research suggests that the total number of bacteria is about 10 times more in the BV than in the H condition [@Zozaya:2010].
+Thus, this is an extremely challenging environment in which to determine differential abundance as there are both compositional and scale changes between conditions.
+The accepted method to analyze vaginal metratranscriptome data is on a taxon-by-taxon basis [@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding can be ignored.
+The problem with a systemic analysis is that many housekeeping genes are returned as differentially abundant between groups, a result likely due to a disconnect between the scale assumptions of the normalization used [@Wu2021] and the actual change in scale between conditions.
+
+In this example, we show how to specify the scale model explicitly and demonstrate that applying a user-defined scale model can account for some of these modeling difficulties.
+A user-defined scale model can control for both the mean difference of scale between groups (e.g., directly incorporate information on the differences in total number of bacteria between the BV and H conditions) as well as the uncertainty assumed in that difference.
+To specify a user-defined scale model, we can pass a matrix of scale values instead of a single estimate of gamma to \texttt{aldex.clr()}.
 This matrix should have the same number of rows as the of Monte-Carlo Dirichlet samples, and the same number of columns as the number of samples.
-There is an `aldex.makeScaleModel()` function that can be used to simplify this step.
-
-This scale model matrix encapsulates two aspects of scale uncertainty and explicitly illustrates the concept of scale.
-First, as in the first example, the gamma paramter indicates the scale uncertainty.
-Second, we need to explicitly indicate the difference in scale between the groups in the dataset.
+While this matrix can be computed from scratch by the analyst, there is an \texttt{aldex.makeScaleModel()} function that can be used to simplify this step in most cases.
 
 Figure 3A shows an effect plot [@gloor:effect] of the data where reads are grouped by function, corresponding approximately to grouping orthologous sequences regardless of the organism of origin.
-Each point represents one of the 3728 functions, and we can see that there are many more functions represented in the BV group (bottom) than in the healthy group (top).
+Each point represents one of the 3728 functions.
+There are many more functions represented in the BV group (bottom) than in the healthy group (top).
 This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because	there is a greater diversity of organisms in BV than in H samples and because the BV condition has at least an order of magnitude more bacteria than does the H condition.
 
- We can  see that there are a large number of functions that are shared between the two groups (boxed area in Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected  to contribute to differences in ecosystem behaviour.
+ We can	 see that there are a large number of core metabolic and information processing functions that are shared between the two groups (boxed area in Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected	 to contribute to differences in ecosystem behaviour and so should not be differentially abundant.
 As a proxy for housekeeping functions the core ribosome functions (blue) shows that their mean location is not centred on 0.
-The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5) and not surprisingly have among the lowest dispersion in the dataset.
-Nevertheless, they are identified as differentially abundant (red) along with many others.
+The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5).
+Since they have very low dispersions, these functions are identified as differentially abundant (red) along with many others.
 While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups because these are functions in common to every organism.
@@
 These housekeeping functions should be located on the midline of no difference.
-Panel B shows the same data scaled with `gamma = 1`, which increase the minimum dispersion approximatly by one unit.
+Panel B shows the same data scaled with `gamma = 0.5`, which increase the minimum dispersion approximatly by one unit.
 Here the housekeeping functions from Panel A are colored cyan or blue for reference.
-Panel C shows the same data scaled with `gamma = 0.75` and a 0.15 fold difference in dispersion applied to the BV samples relative to the H samples.
+Panel C shows the same data scaled with `gamma = 0.5` and a 0.15 fold difference in dispersion applied to the BV samples relative to the H samples.
 The orange functions are now statistically significant.
@@
 points(xg.all[fn,'diff.win'], xg.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
-title('B: ALDEx2 gamma scaled', adj=0, line= 0.8)
+title('B: ALDEx2 gamma', adj=0, line= 0.8)
 
@@
 points(xt.m.all[fn,'diff.win'], xt.m.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
-title('C: ALDEx2 both scaled', adj=0, line= 0.8)
+title('C: ALDEx2 both', adj=0, line= 0.8)
 
@@
 
-Applying the default scale model of `gamma=1` as before increases the dispersion as expected, but does little to move the large number of housekeeping functions toward the midline of no difference.
-Nevertheless, about 50% of the housekeeping functions are no longer statistically significantly different.
+Applying the default scale model of \texttt{gamma=0.5} increases the dispersion as expected but does little to move the large number of housekeeping functions toward the midline of no difference.
+This is as expected: the mean of the default scale model is based on the CLR normalization so no shift would be expected over the original ALDEx2 model.
+Nevertheless, about 50\% of the housekeeping functions are no longer statistically significantly different.
+Note that this change is simple to conduct, has no additional computational complexity and requires only a slight modification from the analyst.
 
-Up to this point, scale uncertainty has been applied uniformly to both conditions, but a user-defined scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix.
-For this we need to revisit the concept that all normalizations in widespread use are actually ratios with the denominator chosen by one of several methods.
-It is important to recognize that in some datasets the denominator itself is not crucial as all approaches give substantially similar results; the transcriptome example used previously is a good example of such an idealized dataset.
-This begs the question, "if it is not the denominator, what then is the important determinant?" As shown by Nixon et al.
-[-@nixon2023scale]  it is usually the *relationship* between the group scale values (denominators) that is important, not their actual value.
-This can be illustrated quite simply by starting with the mean ratio in \( G\) between groups for the yeast transcriptome dataset which are  \(6.05\times^{-5}\) for snf-w and \(5.08\times^{-5}\); their ratio being 1.17, or a 0.17-fold difference.
-Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 1 as shown in Supplemental Figure 2.
-Altering this ratio moves the location of the data.
+Up to this point, scale uncertainty has been applied as an extension of the CLR normalization via the default scale model, but a user-defined scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix.
+While not necessary in all datasets (e.g., the transcriptome example discussed in the previous section and see the Supplementary material), it can greatly improve modeling in certain cases, especially when the scale is highly asymmetric between conditions.
+In order to specify a scale model from scratch, we need to revisit the concept that all normalizations in widespread use are actually ratios with the denominator implied by the normalization.
+Therefore, we can easily deviate from a certain normalization (e.g., the geometric mean assumption implied by the CLR) by specifying a total model based on the mean difference between conditions.
+While knowing the mean difference between conditions may seem cumbersome in practice, it is the \emph{relationship} between the group scale values that is important, not their raw values (Nixon et al.
+2023).
+This can be illustrated quite simply by starting with the mean ratio in \(G\) between groups for the yeast transcriptome dataset which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for WT; their ratio being 1.17, or a 0.17-fold difference.
+Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental Figure 2.
+This ratio can be adjusted to alter the mean assumption placed on the group scale values.
 
-With this background, we can understand how to apply a full scale to the metatranscriptome dataset with the `aldex.makeScaleMatrix()` function.
-This function uses a logNormal distribution when building the model so that ratios that are chosen are symmetrical.
-The full scale model option can be used ot make a matrix of relative scales that are distinct for each group (or even each sample) along with their uncertainties.
-Applying a per-group relative differential scale of 0.15 with a base scale of 1 moves the housekeeping functions to the midline of no difference (Figure 3C).
-Note that now a significant number of functions  are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied.
+With this background, we can understand how to apply a user-defined scale model to the metratranscriptome dataset.
+While a user-defined scale model can be quite flexible with relative scales that are distinct for each group (or even each sample) along with their uncertainties, here we focus on using the  \texttt{aldex.makeScaleMatrix()} function.
+This function uses a logNormal distribution to build a scale matrix given a user-specified mean difference between groups and uncertainty level.
+Applying a per-group relative differential scale of 0.15 moves the housekeeping functions to the midline of no difference, and applying a gamma of 0.5 provides the same dispersion as using a gamma of 0.5 with the CLR (Figure 3C).
+Note that now a significant number of functions are differentially up in BV that were formerly classed as not different without scale, or when only a default scale was applied.
 These former false negatives are noted in orange in each panel.
 Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant.
-Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.
-We suggest that the default scale model alone is useful and should be normally applied when the data are well centred.
-However, when datasets are not well centred or when the investigator has prior information about the underlying biology, we advocate developing and using a full differential scale model.
+Thus, applying a differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.
+We suggest that the default scale model  is sufficient when the data are approximately  centred.
+However, when datasets are not well centred or when the investigator has prior information about the underlying biology, we advocate developing and using a user-specified scale model.
 
@@
 
-Biological count data derived from HTS can be decomposed into two parts the relative (compositional) and the absolute (scale), and the product of these generates a fully scaled biological system [@nixon2023scale].
-Biological systems are both predictably variable and stochastic and current measurement methods that rely on high throughput sequencing fail to capture all of that variation.
-In the absence of  information external to the sequencing run itself, no normalisation method can recapture any of the scale information [@Loven:2012aa].
+Biological count data derived from HTS can be decomposed into two parts: the relative (compositional) and the absolute (scale), and the product of these generates a fully scaled biological system (Nixon et al.
+2023).
+Biological systems are both predictably variable and stochastic, and current measurement methods that rely on high throughput sequencing fail to capture all of that variation, particularly variation due to scale.
+In the absence of information external to the sequencing run itself, no normalisation method can recapture any of the scale information (Lovén et al.
+2012).
+Even so, many methods such as flow cytometry (Vandeputte et al.
+2017), spike-in probes and fluorescent in-situ hybridization (Lovén et al.
+2012; Marguerat et al.
+2012) have been proposed as a means to do this.
+In the context of the updated ALDEx2 software suite, these methods can be useful if they contain information on the particular scale the researcher is interested in.
 
-Since the underlying scale of the system cannot be measured easily -- methods such as flow cytometry [@Vandeputte:2017aa], spike-in probes and fluorescent in-situ hybridization [@Loven:2012aa;@yeast-absolute] have been used -- the effect on analysis can be modelled by including scale uncertainty in the analysis [@nixon2023scale].
-Nixon et al, showed that this can be done by including uncertainty in the denominator used for the normalization.
-The ALDEx2 R package is ideally suited for this since this tool builds a Bayesian posterior of the compositional component of the dataset at the outset and then conducts the analysis on that posterior.
-Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty, and this can be done without additional computational complexity.
-For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a Normal or logNormal distribution depending on preference.
+The ALDEx2 R package is readily amenable for incorporating scale uncertainty.
+Originally, this tool used the only the Dirichlet distribution to sample compositional uncertainty.
+Rather than applying the CLR normalization as in the original ALDEx2, scale uncertainty can be added through the use of a scale model with no additional computational complexity.
+Scale uncertainty can be sampled from any distribution depending on prior knowledge or preference.
+By default, ALDEx2 samples scale from a logNormal distribution inspired by the CLR normalization.
+However, there is the option to introduce a full scale model that  encapsulates both uncertainty and asymmetry in underlying scale.
 
 All normalizations attempt to make the samples in a dataset commensurate but do not explicitly address the scale of the underlying system.
+The two supplementary tables show that these normalizations can be interpreted as scale assumptions.
 However, the general lack of scale information has important consequences for the analysis of HTS datasets.
-One issue is that analysis tools  seem over-powered with even moderate sample sizes and different tools have different intrinsic power, Type 1 and Type 2 error rates  [@Schurch:2016aa].
-Using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reproducible p-values (eg.
-[@Halsey:2015aa]).
-Thus, recommendations to use small sample sizes in multivariate datasets such as RNA-seq datasets are not supported by simple modelling in the univariate case.
-Another issue is that datasets are difficult to analyze when there they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a; @Wu2021].
+One issue is that analysis tools seem over-powered with even moderate sample sizes, and different tools have different intrinsic power, Type 1 and Type 2 error rates (Schurch et al.
+2016) and fail to account for false discovery rates properly [@hawinkel2017;@Li:2022aa].
+Traditional recommendations would suggest that using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reproducible p-values (e.g., Halsey et al.
+[-@Halsey:2015aa]).
+However, recommendations to use small sample sizes are widespread in multivariate datasets such as RNA-seq datasets.
+We argue that this is a hallmark of unacknowledged bias: since normalizations are typically assumed to be correct with no uncertainty, smaller samples offer an unintentional buffer against this misspecification.
+Another issue is that datasets are difficult to analyze when  they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a;@Wu2021].
+
+Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes [@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa] and microbiomes [@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017].
+Generically, it is observed that the actual agreement between methods can be modest, and there is usually a recommendation that some methods are better suited for individual datasets or for specific types of analysis; e.g.,~`Normalization and microbial differential abundance strategies depend upon data characteristics' [@Weiss:2017aa].] However, another interpretation of such studies is that different normalizations can be better suited to different datasets by happenstance, but it is not obvious how the use of specific normalizations could be justified on a per dataset basis.
+There are now published examples of multiple large replication efforts in psychology [@Open-Science-Collaboration:2015aa], cancer biology [@Rodgers:2021aa], ecology and evolution [@evoRepr-pre].
+In a nutshell, experiences from these studies indicate that published data typically have higher effect sizes and appear more significant than replication efforts with the same data, or with newly generated replication data [@Rodgers:2021aa].
+Thus, most published data should be viewed with suspicion [@Ioannidis:2005aa] unless efforts have been made to ensure that the results are robust.
 
@@
 Figure 2 shows one reason for this robustness could be that the dual cutoff is mimicking the effect of including scale uncertainty, since substantially similar transcripts are identified by the two approaches.
-However, while using the post-hoc difference cutoff is useful for differential abundance analysis it is not clear how this can be incorporated into other kinds of downstream analyses.
-Conversely analyses that include scale uncertainty are fully compatible with existing downstream analyses.
+We believe that adding scale uncertainty leads to a more transparent analysis compared to the cutoff as it is easy to describe the conditions under which certain conclusions hold.
+Furthermore, while using the post-hoc difference cutoff is useful for differential abundance analysis, it is not clear how this can be incorporated into other kinds of downstream analyses.
+Conversely, analyses that include scale uncertainty are fully compatible with existing downstream analyses.
 
-In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult to analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
-We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference or exhibit other pathologies (supplemental).
+In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult-to-analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
+We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference or exhibit other pathologies (Supplemental).
 Incorporating differential scale on a per-group basis moves the mass of the data towards the midline of no difference and so affects both Type I and Type II error rates.
@@
 Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on the analysis weighted accordingly.
-Additionally, the use of differential scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups.
+Additionally, the use of user-defined  scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups.
 Thus, reporting scale uncertainty should become a standard practice in the analysis of HTS datasets as a way to identify which features are most robust to differences in the underlying system.
 Finally, we supply a toolkit that makes incorporating scale simple even for datasets that come from highly asymmetrical environments.
+
+# References

### 2023-10-21 go3.tex
sent for comment
submitted version

--- 2023-10-17 go3-author.tex
@@
 
-\title{Beyond compositionally in high throughput sequencing; estimating
-the importance of scale in data analysis with ALDEx2}
-\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}
-\date{\today}
+\title{Normalizations are not what you think; Explicit Scale Simulation
+in ALDEx2}
+\author{true \and true \and true}
+\date{}
 
@@
 \maketitle
-
-$^1$ Department of Biochemistry, University of Western Ontario
-$^2$ College of Information Sciences and Technology, Pennsylvania  State University
-$^3$ Department of Medicine, Pennsylvania  State University
-
 \begin{abstract}
-Despite scale being an important confounding variable, high throughput
-sequencing generates data that are difficult to reconcile with the scale
-of the underlying environment, but it has been shown that the scale of
-the environment is related to the geometric mean of the parts of the
-sample. The ALDEx2 R package builds a Bayesian model of the data and
-here we report an update to the model that includes both random sampling
-and scale estimates. We show that incorporating scale allows for robust
-inference in a number of different experimental designs and we provide
-guidance on how to incorporate scale inference into differential
-abundance analysis. While all data normalizations in widespread use make
-implicit assumptions about the scale of the system we suggest that
-explicit rather than implicit scale models should be used.
+In many high-throughput sequencing (HTS) studies, sample-to-sample
+variation in sequencing depth is not driven by variation in the scale
+(e.g., total size, microbial load, or total gene expression) of the
+underlying biological systems being measured but rather is driven by
+technical factors. Typically the technical variation is addressed using
+some form of statistical normalization. Normalizations are data or
+parameter transformations that remove unwanted technical variation in
+the hopes of facilitating analyses sensitive to scale; e.g.,
+differential abundance and differential expression analyses. Recently we
+showed that any normalization makes implicit assumptions about the
+unmeasured system scale and that errors in these assumptions can lead to
+dramatic increases in false positive and false negative rates. Here we
+describe updates to the ALDEx2 R package that mitigate these problems by
+directly modeling uncertainty in the unmeasured system scale through the
+use of a \textit{scale model}. Scale models generalize the idea of
+normalizations and can be thought of as explicitly modeling potential
+error in the chosen normalization. Beyond enhancing the robustness of
+HTS analyses, the use of scale models within ALDEx2 enhances the
+transparency and reproducibility of analyses by making implicit
+normalizing assumptions an explicit part of the model building process.
 \end{abstract}

### 2023-10-21 fix_authors.sh

@@ -1,10 +1,10 @@
 #!/bin/zsh
 
-awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin Silverman$^{2,3}$")};1' go3.tex > jnk.tex
+awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin D. Silverman$^{2,3,4}$")};1' go3.tex > jnk.tex
 
 awk '{sub(/\\date{}/,"\\date{\\today}")};1' jnk.tex > jnk2.tex
 
-awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$ Department of Biochemistry, University of Western Ontario\n\$^2\$ College of Information Sciences and Technology, Pennsylvania  State University\n\$^3\$ Department of Medicine, Pennsylvania  State University\n")};1' jnk2.tex > go3-author.tex
+awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$ Department of Biochemistry, University of Western Ontario\n\$^2\$ College of Information Sciences and Technology, Pennsylvania  State University\n\$^3\$ Department of Medicine, Pennsylvania  State University\n\$^4\$ Institute for Computational and Data Science, Pennsylvania State University\n")};1' jnk2.tex > go3-author.tex
 
 rm jnk.tex
 rm jnk2.tex

### 2024-05-16 App-note.Rmd
go3-edits

--- 2023-10-21 go3.Rmd
@@
 
-High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing).
+High-throughput sequencing (HTS) is a ubiquitous tool used to explore  biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing).
 HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to the flow cell.
 Each of these steps can be though of as a sampling step: a pool of DNA that could be measured is sampled and only that subsampled DNA is carried over to subsequent steps.
-Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., size, microbial load, or total gene expression) of the measured biological system is degraded or lost.
-Increasingly, researchers are turning to modified experimental protocols (e.g., DNA spike-ins or cell counting) in an attempt to recover the lost biological variation in scale; see for example [@Vandeputte:2017aa;@Props:2017aa].
-However, these protocols often do not recover meaningful biological variation but instead recover scale variation at some intermediate step in the sample preparation protocol (e.g., at the stage where the DNA spike-in was added).
+Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., complexity, size, microbial load, or total gene expression) of the measured biological system is degraded or lost.
+One strategy is to include DNA spike-ins or cell counting) to include information on the lost biological variation [@Vandeputte:2017aa;@Props:2017aa].
+However, this only recovers variation at the stage where the DNA spike-in was added.
 In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an outstanding challenge.
@@
 As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010].
-Thus, a full description of these systems includes both relative change (composition) and absolute abundance (scale), yet current methods access only the compositional information.
+Thus, a full description of these systems includes both relative change (composition) and complexity or size (scale), yet current methods access only the compositional information.
 
-To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
-Let \(\mathbf{W}\) also represent a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true (as opposed to measured) amount of class \(d\) in the biological system from which sample \(n\) was obtained.
-Succinctly, \(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\) denotes the true counts in the system under study.
-We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions).
+A sequence count observation is summarized in a  \(D \times N\) count matrix \(\mathbf{Y}\) where \(\mathbf{Y}_{dn}\) indicates the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
+We want this observation to tell us about the true environment \(\mathbf{W}\) also represented by a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true as opposed to the measured values in the biological system from which sample \(n\) was obtained.
+
+We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., complexity or total) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions).
 That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
@@
 The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data.
-For brevity and concreteness, we discuss ALDEx2 in its simplest form as a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models.
+In its simplest form it is a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but can be used to fit more complex linear models.
 At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the \(\mathbf{W}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses.
@@
 Through the lens of this paper, this is a hallmark of unacknowledged bias: a false certainty in the scale of the data brings about an increasing false precision with increasing sample sizes.
-Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]].
+Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
 Thus, by adding scale uncertainty, we can be more confident that the results of the analysis reflect true biological differences and not differences due to the assumptions made by the normalization method.

### 2024-05-16 go3-author.tex
go3-edits

@@
 
-\begin{abstract}
+\section{Abstract}\label{abstract}
 In many high-throughput sequencing (HTS) studies, sample-to-sample
@@
 normalizations and can be thought of as explicitly modeling potential
-error in the chosen normalization. Beyond enhancing the robustness of
-HTS analyses, the use of scale models within ALDEx2 enhances the
-transparency and reproducibility of analyses by making implicit
-normalizing assumptions an explicit part of the model building process.
-\end{abstract}
+error in normalization by examining a distribution over all possible 
+normalizations.  Beyond enhancing the robustness of HTS analyses, 
+the use of scale models within ALDEx2 enhances the transparency and 
+reproducibility of analyses by making implicit normalizing assumptions an 
+explicit part of the model building process.
+%\end{abstract}
 
@@
 deficient for snf1 are smaller, grow more slowly and are sensitive to a
-variety of common agents that cause cell stress (Yoshikawa et al.
-2011){]}. Thus, by adding scale uncertainty, we can be more confident
-that the results of the analysis reflect true biological differences and
-not differences due to the assumptions made by the normalization method.
+variety of common agents that cause cell stress (Yoshikawa et al. 2011).
+Thus, by adding scale uncertainty, we can be more confident that the
+results of the analysis reflect true biological differences and not
+differences due to the assumptions made by the normalization method.
 

### 2024-05-16 supplement_figs.Rmd
go3-edits

@@
 
+```{r HvG_function, echo=F}
+H <- function(x.vec){
+  x.p <- x.vec/sum(x.vec)
+  return(-1 * sum(x.p * log2(x.p)))
+}
+myfun.H <- function(x){ y <- c(x, 1-x); H(y)}
+mf.H <- Vectorize(myfun.H)
+
+G <- function(x.vec){
+  x.p <- x.vec/sum(x.vec)
+  return(mean(log2(x.p)))
+}
+
+myfun.G <- function(x){ y <- c(x, 1-x); G(y)}
+mf.G <- Vectorize(myfun.G)
+
+n.random = 20
+n.parts = 500
+test='N' # or U, B for Normal or Uniform or Beta
+
+# note mean is only used when 
+HvG <- function(n.random, n.parts, test='U', mean=30){
+
+  mat.log <- matrix(data=NA, nrow=n.parts, ncol=n.random)
+  mat.h <- matrix(data=NA, nrow=n.parts, ncol=n.random)
+
+  # 2000 entries
+  val <- seq(2,(n.parts*2), by=2)
+
+  # this will generate data of the 
+  # correspondence between H and G in a runif
+  # for random number vectors of length
+  # 2 to 10000. Does this 5 times
+  # N approaches 0 difference as the mean of the N approaches infinity
+  for(i in 1:n.parts){
+    for(j in 1:n.random){	 
+     if(test=='N'){ d <- rnorm(val[i], mean=mean) }
+     else if (test == 'U'){ d <- runif(val[i]) * mean }
+     else if (test == 'B'){ d <- rbeta(val[i], 0.1, 10) * mean }
+     d[d < 0] <- 0
+     d <- d/sum(d)
+     # mean of log(d), exp = G
+     mat.log[i,j] <- G(d)
+     # entropy
+     mat.h[i,j] <- H(d)
+    }
+}
+
+jnk <- list(mat.log, mat.h)
+names(jnk) <- c("mat.log","mat.h")
+return(jnk)
+}
+
+HvG.plot <- function(HvG.out){
+
+#plot(rep(val,n.random),mat.log- mat.h, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
+#  log='x', xlab='length vector', ylab='(-1* G) - H')
+plot(HvG.out$mat.log, HvG.out$mat.h, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
+   xlab='G', ylab='H')
+
+#abline(lm((HvG.out$mat.log[,2]-HvG.out$mat.h[,2]) ~ val), col='red')
+abline(lm((HvG.out$mat.h[,2]~HvG.out$mat.log[,2])), col='red')
+
+#lm((mat.log[,2]-mat.h[,2]) ~ val)
+lm((HvG.out$mat.h[,2]~HvG.out$mat.log[,2]))
+}
+
+```
+
 # GM is related to Information and Shannon's entropy in HTS datasets
 
-Information content is a fundamental property of all systems.
-We can think about  scale  from an information theoretic point of view as a measure of how much information is encoded in a particular sample.
+## Shannon's entropy has a volume or size
+
+Information  is a fundamental property of all measured systems.
+For discrete probability vectors, Shannon defined their information properties and launched the field of information theory for communications.
+Famously, for any probability vector, Shannon set the total entropy as the inverse of the sum of the probability weighted logarithm of the probabilities.
+Less well known is that an unweighted average entropy was also defined, and that this turns out to be the inverse of the logarithm of the geometric mean of the probability vector.
+Thus, information theory and compositional data analysis intersect through the geometric mean of a probability vector.
+Here I show that the information theoretic interpretation can help us understand the scale of a system as defined by Nixon and Silverman, and that this interpretation can help in the interpretation of highly asymmetric datasets.
+I will show the Asymptotic Equipartition Property of information can be defined as a volume for discrete distributions, and how that volume relates to the scale of a system.
+Finally, I will show how Shannon's entropy can substitute for the geometric mean in common CoDa operations and how that alters the interpretation of the results.
+
+We can think about  scale  from an information theoretic point of view as a measure of how much information, or total uncertainty, is encoded in a particular sample [@Shannon.48;@Jaynes:2003].
+In the geometric interpretation of information theory used in quantum information theory, formally described as  the  Asymptotic Equipartition Property of information [@Cover:1991;@wilde_2017], entropy can be interpreted as the volume occupied by a probability distribution.
+See chapters 4 of the PhD thesis of Lecamwasam [-@ruvi.blog] for a nice explanation of this.
+
+For notational simplicity assume we have a single discrete random variable to represent a probability distribution with d elements; i.e.
+\( X =  \mathbf{p}_{i=(1 \dots d )} \).
+In information theory, the elemental amount of information or surprisal for \(p_i\) is the inverse of the logarithm of the elemental probability, \( -log_2(p_i)\)[@rezaInfoIntro].
+This measure is often called self-information.
+
+The total entropy of the system  \( H(X) \) is the weighted sum of the elemental information;
+
+\[
+H(X) = - \sum_{i=1}^{d} p_i \log_2 p_i
+\]
+
+\( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \).
+We can also calculate the expected amount of information for each observation in the random variable, and this is the mean of the elemental probability.
+This measure is also called the sample average of the information [@wilde_2017];
+
+\[
+ h(X) = - \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
+\]
+
+The linkage between compositional analysis, scale inference and information theory comes when we realize that the logarithm of the geometric mean calculated in base(2) is:
+
+\[
+ l2G(X) = \log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
+\];
+
+We see that \( h(X) = - l2G(x) \). Furthermore, \(l2G\) is used as the basis for the centred log-ratio transform and is the starting point for scale-based inference:
+
+ \[CLR(X) = log2(p_i) - l2G(X) = log2(p_i) + h(X)  \]
+
+Thus we see that the geometric mean used in the centred log ratio (CLR), often used for Compositional Data Analysis (CoDa) [@aitchison1982] is directly related to entropy or \(H\).
+Indeed, we can rearrange the CLR formula to show that it can be interpreted as computing the difference between the elemental information and the mean information content:
+
+  \[CLR(X) = -log2(p_i) - (-l2G(X)) = -(-log2(p_i) - h(X))\]
+
+As defined in [@nixon2023scale], the scale is the inverse of \( l2G \), which is \(h(X)\).
+Thus, one way we can understand scale is that it is measuring the total complexity of the system, and this is expected to increase with absolute size in most cases.
+Moreover, \(H(X)\) and \(G(X)\) share similar shapes in the continuum between 0-1 for a bivariate distribution as shown below:
+
+```{r} 
+par(mfrow=c(1,1))
+curve(mf.G, from=1e-2, to =.99,  col='red', lty=2, ylim=c(-3,1),
+   xlab="x in [0,1]", ylab="value")
+curve(mf.H, from=1e-2, to =.99,  col='black', lty=2, add=T )
+legend(.4,-1.5, legend=c('H','G'), col=c('black','red'), pch="-")
+```
+
+The difference being that entropy is constructed to have a value of 0 at the margins because Shannon defined \( 0log(0) = 0\) while the geometric mean approaches negative infinity.
+
+Now let's think about the idea of entropy as a volume which allows us to identify what amount of the available entropy space a given observation fills.
+The following is taken and modified from [@ruvi.blog], to which you should refer for a more fulsome discussion, and it is covered in Chapter 2 of Wilde[-@wilde_2017] and Chapter 3 of Cover and Thomas [-@Cover:1991].
+If we start with a four part system \(X1 = [A,C,G,T]\) where the frequencies are equally and identically distributed, then \(p_A = p_C = p_G = p_T = \frac{1}{4} \).
+\( H(X1) = -1 * 4 * (\frac{1}{4} * log2(\frac{1}{4}) ) = 2\).
+This is the maximum entropy possible.
+We can obtain the "volume" of \(X1\) by exponentiating \(H1\) using the same base as was used to calculate the entropy; \( V1 = 2^{H1} = 4 \).
+This is the same as the number of letters in the system; so the volume needed to explain the system is 4 units (in this case bits).
+But what happens in another system, \(X2\) where A occurs with a much higher probability, say 0.7, and the other three are distributed equiprobably amongst the remainder with a probability of 0.1; i.e.
+\(p_C = p_G = p_T = 0.1 \).
+In this case \( H2 = -1 * ((\frac{7}{10} * log2(\frac{7}{10}) ) + (3 * (\frac{1}{10} * log2(\frac{1}{10}) ))) = 1.358 \).
+Here the volume of \(X2 = 2^{H2} = 2.56\); meaning that less than the maximum volume is taken up by the information.
+Here \(X2\) consumes about 64% of the volume of system \(X1\).
+Thus, the volume is a measure of the total complexity or the scales of the two systems.
+In this way we can understand that scale is related to the information volume of a system.
+
+Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(\mathbf{Y}^{\parallel}_{n})\) as expected from the discussion above, and that this difference converges to a constant as the number of entries in the probability vector increases regardless of the distribution, although different distributions converge at different rates.
+For example, if we plot the relationship between H and G as a function of the length of the probability vector we can see a direct inverse relationship.
+
+```{r, echo=F, fig.cap="Association between entropy (H) and geometric mean (G) as a function of vector length.
+Twenty random vectors were constructed for each length between  2 and 1000 in increments of 2 for each of the random distributions in the legend; N = Normal, U = Uniform, B = Beta.
+The bottom right of each plot represents vectors of length 2, and the top left represents the vector of lenght 500.
+The maximum value of H increases as the vector length increases, and the maximum value of G decreases in lock-step.
+Each random distribution has an obviously distinct relationship between the two measures.
+For the purposes of high throughput sequencing the Beta distribution is most similar to that seen in the majority of instances."}
+set.seed=24
+par(mfrow=c(1,3))
+U <- HvG(20, 500, test='U', mean=1)
+U.plot <- HvG.plot(U)
+text(-4,0.5,labels='vec len = 2')
+text(-7.4,9.8,labels='vec len = 1000')
+title(main='random uniform')
+B <- HvG(20, 500, test='B', mean=1)
+B.plot <- HvG.plot(B)
+title(main='Extreme skew')
+N <- HvG(20, 500, test='N', mean=30)
+N.plot <- HvG.plot(N)
+title(main='N, mean=300')
+# intercept
+U.plot$coefficients[1]
+B.plot$coefficients[1]
+N.plot$coefficients[1]
+par(mfrow=c(1,1))
+
+```
+
+When we plot the relationship for any individual probability vector, we  see that there is an direct relationship between the entropy and the log of the geometric mean, but that this relationship strongly depends on the underlying distribution of the probability distribution \( X \).
+
+```{r, echo=F, fig.cap="Plot of the association between H and G at a vector length of 30.
+The relationship between H and G is inverse, and the strength of that association depends on the distribution.
+The N distribution shows a very strong assocation, while the Beta distribution is less well defined.
+Associations are shown for a vector length of 30."}
+par(mfrow=c(1,3))
+plot(N$mat.log[15,], N$mat.h[15,], main="Normal", xlab="G", ylab="H")
+abline(lm(N$mat.h[15,]~N$mat.log[15,]), lty=2, col='grey', lwd=2)
+plot(U$mat.log[15,], U$mat.h[15,], main="Uniform", xlab="G", ylab="H")
+abline(lm(U$mat.h[15,]~U$mat.log[15,]), lty=2, col='grey', lwd=2)
+plot(B$mat.log[15,], B$mat.h[15,], main="Beta", xlab="G", ylab="H")
+abline(lm(B$mat.h[15,]~B$mat.log[15,]), lty=2, col='grey', lwd=2)
+par(mfrow=c(1,1))
+```
+
+In real data, shown in Supplemental Figure 3 and Table 1 the correspondence is not as predictable, likely because the real data is a more complex distribution than any of the idealized distributions.
+Thus, these two measures have different behaviours with different distributions of \( p_i \).
+In the case of a uniform distribution both \( H(\mathbf{Y}^{\parallel}_{n}) \) and \( G(\mathbf{Y}^{\parallel}_{n}) \) are maximal when \( p(x) \) is equally and identically distributed.
+Thus, we expect that they are positively correlated here.
+In a Normal or a skewed distribution, we also observe a positive correlation because both are affected in the same direction by outlier values.
+In very sparse  datasets, the two measures could become uncoupled because \( H(\mathbf{Y}^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(\mathbf{Y}^{\parallel}_{n}) \) would tend to be very small.
+Here these two measures could be either uncorrelated or exhibit negative correlation.
+We can see this distributional behaviour in different datasets.
+
 Intuitively,  systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
-Below, we show that the scale of a system as defined by Nixon et al.
-[-@nixon2023scale] is inversely related to the information content and entropy of the data.
+As the scale of a system as defined by Nixon et al.
+[-@nixon2023scale] is inversely related to \(G\), this means that scale is directly proportional to the information content and entropy of the data.
+
+Below I show that we can replace \( G \) with \(H \) in the calculations performed by ALDEx2 without loss of utility.
 
@@
 Sequencing  returns counts which are related to the underlying proportion; i.e., \(\mathbf{Y}^{\parallel}_n \sim \mathbf{W}^{\parallel}_n\)
-
-For notational simplicity assume we will use a single discrete random variable to represent the probability distribution; i.e.
-\( X =  \mathbf{p}(w_n) \)  over \(1 \dots d \) elements.
-The geometric mean can be calculated as:
-
-\[
-\log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
-\];
-
-here every element in \(log_2 G(X)\) is the logarithm of the elemental probability \( log_2(p_i)\).
-
-In information theory, the elemental amount of information or uncertainty for \(p_i\) is the inverse \(I(X)_i = -log_2(p_i)\)[@rezaInfoIntro].
-
-The total entropy of the system  \( H(x) \) is the weighted sum of the elemental information
-
-\[
-H(x) = -\sum_{i=1}^{d} p_i \log_2 p_i
-\]
-
-In this formulation, \( \log_2(G(X)) \) and \( H(X) \) can be understood  to be an unweighted and a weighted measure  of 'surprisal' in the dataset.
-The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.
-\( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \).
-Conversely, \( G \) is an unweighted mean of the same distribution and  \( G \) is the denominator to calculate the centred log-ratio normalization (CLR).
-Note that when calculating the CLR we are taking a ratio, and similarly for calculating entropy, but that the scale formula takes the product.
-Thus, the scale of a system is inversely related to the other two measures.
-
-Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(\mathbf{Y}^{\parallel}_{n})\) (Supplemental Figure 1 and Table 1).
-These two measures are expected to have different behaviours with different distributions of \( p_i \).
-In the case of a uniform distribution both \( H(\mathbf{Y}^{\parallel}_{n}) \) and \( G(\mathbf{Y}^{\parallel}_{n}) \) are maximal since \( p(x) \) is equally and identically distributed.
-Thus, we expect that they are positively correlated here.
-In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values.
-In very sparse  datasets, the two measures could become uncoupled because \( H(\mathbf{Y}^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(\mathbf{Y}^{\parallel}_{n}) \) would tend to be very small.
-Here these two measures could be either uncorrelated or exhibit negative correlation.
-We can see this distributional behaviour in different datasets.
 
@@
 
-HS <- apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GS <- apply(selex+0.5, 2, function(x) mean(log2(x/sum(x))))
+HS <- apply(selex+0.5, 2, function(x)  H(x) )
+GS <- apply(selex+0.5, 2, function(x) G(x) )
 
-HY <- apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GY <- apply(yst+0.5, 2, function(x) mean(log2(x/sum(x))))
+HY <- apply(yst+0.5, 2, function(x) H(x) )
+GY <- apply(yst+0.5, 2, function(x) G(x) )
 
-HM <- apply(ko.both+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GM <- apply(ko.both+0.5, 2, function(x) mean(log2(x/sum(x))))
+HM <- apply(ko.both+0.5, 2, function(x) H(x) )
+GM <- apply(ko.both+0.5, 2, function(x) G(x) )
 
-H16 <- apply(rRNA+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-G16 <- apply(rRNA+0.5, 2, function(x) mean(log2(x/sum(x))))
+H16 <- apply(rRNA+0.5, 2, function(x)  H(x) )
+G16 <- apply(rRNA+0.5, 2, function(x) G(x) )
 
-Hss <- apply(ss+0.1, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-Gss <- apply(ss+0.1, 2, function(x) mean(log2(x/sum(x))))
+Hss <- apply(ss+0.1, 2, function(x) H(x) )
+Gss <- apply(ss+0.1, 2, function(x) G(x) )
 
@@
 
-```{r selex, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters.
+```{r selex, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters.
 All scales are calculated with a logNormal distribution to ensure symmetry for the user.
@@
 
+# How scaling affects dispersion
+
+Thus far we have observed that adding scale uncertainty increases the minimum dispersion value with little effect on the difference between values as in panel A.
+Panel B shows the change in dispersion relative to the rAbundance of the features.
+Here we can see that the minimum dispersion is increased as gamma increases.
+The horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5.
+Panel C shows that this increase is non-linear, being more pronounced among those features that had minimal dispersion when gamma=0.
+Panel D shows that increasing gamma actually shrinks the dispersion estimates towards the mean dispersion, while increasing the total dispersion as shown in panel C.
+
+Thus, adding scale uncertainty has two effects both on dispersion.
+The first is to increase the dispersion estimate for each part and this has its primary effect to reduce the  p-values and standardized effect sizes returned by ALDEx2.
+The second is to reduce the range of dispersions such that the parts with the lowest dispersions have the greatest increase.
+This increase in dispersion comes about because the underlying distributions are widened because of the additional uncertainty added by the inclusion of scale uncertainty.
+
+```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="jnk"}
+yst.conds <- c(rep('S',44),rep('W',42))
+x <- aldex.clr(yst, yst.conds, gamma=NULL)
+xs <- aldex.clr(yst, yst.conds, gamma=0.5)
+x1 <- aldex.clr(yst, yst.conds, gamma=1)
+
+x.e <- aldex.effect(x)
+xs.e <- aldex.effect(xs)
+x1.e <- aldex.effect(x1)
+
+par(mfrow=c(2,2))
+# we imperceptibly change difference
+plot(x.e$diff.btw, x1.e$diff.btw, xlab='diff g=0', ylab='diff g=1')
+abline(0,1, col='grey', lwd=2, lty=2)
+title(main='A', line=-1.2, adj=0.5)
+#=-1
+cuts <- x.e$rab.all > -0.5 & x.e$rab.all < 0.5
+# the residual of the 
+mn.mid <- median(x.e$diff.win[cuts])
+mn.mid.s <- median(xs.e$diff.win[cuts])
+mn.mid.1 <- median(x1.e$diff.win[cuts])
+
+# we change dispersion a lot for the bulk of the features
+plot(x.e$rab.all, x.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  xlab='rAbundance', ylab='dispersion')
+points(xs.e$rab.all, xs.e$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+points(x1.e$rab.all, x1.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3))
+abline(h=mn.mid, lty=2)
+abline(h=mn.mid.s, lty=2, col='red')
+abline(h=mn.mid.1, lty=2, col='blue')
+title(main='B', line=-1.2, adj=0.5)
+
+plot(x.e$diff.win, x1.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3), ylim=c(0,4.5), 
+  xlab='g=0 dispersion', ylab='g=0.5 | g=1 dispersion')
+points(x.e$diff.win, xs.e$diff.win, pch=19, cex=0.5, col=rgb(1,.4,.1,0.3))
+abline(0,1, lty=2)
+title(main='C', line=-1.2, adj=0.5)
+
+strip.data <- list(x.e$diff.win[cuts] - mn.mid, xs.e$diff.win[cuts] - mn.mid.s, x1.e$diff.win[cuts] - mn.mid.1)
+
+stripchart( strip.data, col=c(rgb(0,0,0,0.3), rgb(1,0,0,0.3), rgb(0,0,1,0.3)), vertical=T, method='jitter', pch=19, group.names=c('g=0', 'g=0.5', 'g=1'), ylab='residual dispersion')
+title(main='D', line=-1.2, adj=0.5)
+
+```
+
 # Issues with DESeq2 and edgeR
@@
 
-```{r DESedg, echo=F, fig.cap="Shown here are the mean log2 fold change as a density plot, and a Volcano plot showing the location and adjusted p-value for each feature in the metatranscriptomic dataset.
+```{r DESedg, eval=T, echo=F, fig.cap="Shown here are the mean log2 fold change as a density plot, and a Volcano plot showing the location and adjusted p-value for each feature in the metatranscriptomic dataset.
 The DESeq2 approach does a good job of centring this data, while edgeR is less suitable.
@@
 load('analysis/xt.Rda')
-hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
+hk.off <- xt.all$diff.win < 5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
 
@@
 sig.edge <- which(p.adjust(meta.edg.qlf$PValue) < 0.01)
-par(mfrow=c(2,2))
+sig.ald <- which(xt.m.all$we.eBH < 0.01)
 
-plot(density(meta.DES.res@listData$log2FoldChange[hk.off]), main='DESeq2 FC')
+par(mfrow=c(2,3))
+
+plot(density(xt.all$diff.btw), xlim=c(-5,5), ylim=c(0,1), main='ALDEx2 FC')
+points(density(xt.all$diff.btw[hk.off]), type='l', col='orange')
+abline(v=0, col='orange', lty=2)
+
+plot(density(meta.DES.res@listData$log2FoldChange), xlim=c(-5,5), ylim=c(0,0.8), main='DESeq2 FC')
+points(density(meta.DES.res@listData$log2FoldChange[hk.off]), type='l', col='orange')
+abline(v=0, col='orange', lty=2)
+
+plot(density(meta.edg.qlf$logFC), xlim=c(-5,5), ylim=c(0,0.5), main='edgeR FC')
+points(density(meta.edg.qlf$logFC[hk.off]), type='l', col='orange')
 abline(v=0, col='red', lty=2)
+
+plot(xt.m.all$diff.btw, -1*log10(xt.m.all$we.eBH), 
+  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)', main='ALDEx2 volcano')
+points(xt.m.all$diff.btw[sig.ald],-1*log10(xt.m.all$we.eBH)[sig.ald] + 1e-8, col='red')
+points(xt.m.all$diff.btw[hk.off],-1*log10(xt.m.all$we.eBH)[hk.off] + 1e-8, col='orange', cex=0.6)
 
@@
 points(meta.DES.res@listData$log2FoldChange[hk.off],-1*log10(meta.DES.res@listData$padj[hk.off] + 1e-70), col='orange', cex=0.6)
-
-plot(density(meta.edg.qlf$logFC[hk.off]), main='edgeR FC')
-abline(v=0, col='red', lty=2)
 

### 2024-05-16 code/meta.DESeq.R

@@ -1,6 +1,6 @@
 
 # get the dataset
-load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
+load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
 
 conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8)) 
 
@@ -10,7 +10,7 @@ library(DESeq2)
 # make a vector of conditions
 coldata <- data.frame(conds.K0)
 
-dds <- DESeqDataSetFromMatrix(countData = ko.both.all,
+dds <- DESeqDataSetFromMatrix(countData = ko.both,
           colData = coldata, design= ~ conds.K0)
 dds <- DESeq(dds)
 meta.DES.res <- results(dds, name="conds.K0_H_vs_B")
@@ -20,7 +20,7 @@ save(meta.DES.res, file='analysis/meta.DES.res.Rda')
 # qlf is recommended for bulk RNA seq 
 library(edgeR)
 group=factor(conds.K0)
-y <- DGEList(counts=ko.both.all,group=group)
+y <- DGEList(counts=ko.both,group=group)
 y <- normLibSizes(y)
 design <- model.matrix(~group)
 y <- estimateDisp(y,design)

### 2024-07-22 go3-author.tex
focussed on dispersion version

--- 2023-10-21 go3-author.tex
@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{setspace}
+\usepackage{bookmark}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 \begin{abstract}
-In many high-throughput sequencing (HTS) studies, sample-to-sample
-variation in sequencing depth is not driven by variation in the scale
-(e.g., total size, microbial load, or total gene expression) of the
-underlying biological systems being measured but rather is driven by
-technical factors. Typically the technical variation is addressed using
-some form of statistical normalization. Normalizations are data or
-parameter transformations that remove unwanted technical variation in
-the hopes of facilitating analyses sensitive to scale; e.g.,
-differential abundance and differential expression analyses. Recently we
-showed that any normalization makes implicit assumptions about the
-unmeasured system scale and that errors in these assumptions can lead to
-dramatic increases in false positive and false negative rates. Here we
-describe updates to the ALDEx2 R package that mitigate these problems by
-directly modeling uncertainty in the unmeasured system scale through the
-use of a \textit{scale model}. Scale models generalize the idea of
-normalizations and can be thought of as explicitly modeling potential
-error in the chosen normalization. Beyond enhancing the robustness of
-HTS analyses, the use of scale models within ALDEx2 enhances the
-transparency and reproducibility of analyses by making implicit
-normalizing assumptions an explicit part of the model building process.
+In high-throughput sequencing (HTS) studies, sample-to-sample variation
+in sequencing depth is driven by technical factors, and not by variation
+in the scale (e.g., total size, microbial load, or total gene
+expression) of the underlying biological systems. Typically a
+statistical normalization is used to remove unwanted technical variation
+in the data or the parameters of the model to enable analyses that are
+sensitive to scale; e.g., differential abundance and differential
+expression analyses. Recently we showed that all normalizations make
+implicit assumptions about the unmeasured system scale and that errors
+in these assumptions can lead to dramatic increases in false positive
+and false negative rates. Here we describe updates to the ALDEx2 R
+package that mitigate these problems by directly modeling uncertainty in
+the unmeasured system scale through the use of a \textit{scale model}.
+Scale models generalize the idea of normalizations and can be thought of
+as explicitly modeling the error in normalization by examining a
+distribution over all possible normalizations. Beyond enhancing the
+robustness of HTS analyses, the use of scale models within ALDEx2
+enhances the transparency and reproducibility of analyses by making
+implicit normalizing assumptions an explicit part of the model building
+process.
 \end{abstract}
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 (merging) multiple libraries together, and then applying a sample of the
-multiplexed library to the flow cell. Each of these steps can be though
-of as a sampling step: a pool of DNA that could be measured is sampled
-and only that subsampled DNA is carried over to subsequent steps. Within
-each sampling step, the connection between the size of the sampled DNA
-pool and the scale (e.g., size, microbial load, or total gene
-expression) of the measured biological system is degraded or lost.
-Increasingly, researchers are turning to modified experimental protocols
-(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
-biological variation in scale; see for example (Vandeputte et al. 2017;
-Props et al. 2017). However, these protocols often do not recover
-meaningful biological variation but instead recover scale variation at
-some intermediate step in the sample preparation protocol (e.g., at the
-stage where the DNA spike-in was added). In short, the disconnect
-between sample-to-sample variation in sequencing depth and biological
-variation in scale remains an outstanding challenge.
-
-Biological variation in scale can represent an important unmeasured
-confounder in HTS analyses (Lovell et al. 2015). For example, cells
+multiplexed library to the flow cell. Each of these steps is a
+compositional sampling step as only a fixed-size subsample of nucleic
+acid is carried over to subsequent steps. Thus, with each sampling step
+the connection between the size of the sampled DNA pool and the scale
+(e.g., size, microbial load, or total gene expression) of the measured
+biological system is degraded or lost. Increasingly, researchers are
+turning to modified experimental protocols (e.g., DNA spike-ins or cell
+counting) in an attempt to recover the lost biological variation in
+scale; see for example (1, 2). However, these protocols often do not
+recover meaningful biological variation but instead recover scale
+variation at the intermediate step in the sample preparation protocol
+where the DNA spike-in was added. In short, the disconnect between
+sample-to-sample variation in sequencing depth and biological variation
+in scale remains an outstanding challenge.
+
+The analysis of HTS data suffers from several known problems that can be
+traced, in whole or in part, to misspecification of scale. The first
+issue is poor control of the false discovery rate (FDR) (3, 4),
+exhibited as dataset-dependent FDR control. The FDR problem is connected
+to the double-filtering approach that is often used, but which is known
+not to have appropriate FDR control (5, 6). The second issue is poor
+performance when analyzing asymmetric data; data that arises because of
+a bias in the direction of change. This type of data frequently arises
+in in-vitro selection experiments (SELEX), transcriptome analysis, and
+microbiome analysis (7). The third issue is that these problems become
+more pronounced as more samples are collected; that is, more information
+results in a worsening of the accuracy of the analysis (9). The final
+problem is that these data are relative data, i.e.~compositional, and
+substantial effort has been put into removing this constraint.
+
+The first three problems were recently shown by Nixon and Silverman (8)
+to be a result of a mismatch between the underlying size or scale of the
+system and the assumptions of the normalizations used for the analysis
+of HTS. Biological variation in scale often represents an important
+unmeasured confounder in HTS analyses (10). For example, cells
 transformed by the cMyc oncogene have about 3 times the amount of mRNA
-and about twice the rRNA content than non-transformed cells (Nie et al.
-2012), and this dramatically skews transcriptome analysis (Lovén et al.
-2012). In addition, wild-type and mutant strains of cell lines, yeast or
-bacteria often have different growth rates, which would affect our
-ability to identify truly differentially abundant genes (Yoshikawa et
-al. 2011). As another example, the total bacterial load of the vaginal
+and about twice the rRNA content than non-transformed cells (11), and
+this dramatically skews transcriptome analysis (12). In addition,
+wild-type and mutant strains of cell lines, yeast or bacteria have
+different growth rates and RNA contents under different conditions,
+which affect our ability to identify truly differentially abundant genes
+(13--15). As another example, the total bacterial load of the vaginal
 microbiome differs by 1-2 orders of magnitude in absolute abundance
-between the healthy and bacterial vaginosis states (Zozaya-Hinchliffe et
-al. 2010), and the composition between these states is dramatically
-different (Ravel et al. 2011; Hummelen et al. 2010). Thus, a full
-description of these systems includes both relative change (composition)
-and absolute abundance (scale), yet current methods access only the
-compositional information.
-
-To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix
-of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number
-of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or
-gene) in sample \(n\). Let \(\mathbf{W}\) also represent a
-\(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing
-the true (as opposed to measured) amount of class \(d\) in the
-biological system from which sample \(n\) was obtained. Succinctly,
-\(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\)
-denotes the true counts in the system under study. We can think of
-\(\mathbf{W}\) as consisting of two parts, the scale
-\(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
+between the healthy and bacterial vaginosis states (16), and the
+composition between these states is dramatically different (17, 18).
+Thus, a full description of any of these systems includes both relative
+change (composition) and absolute abundance (scale). Current methods
+access only the compositional information yet make implicit assumptions
+about the scale.
+
+Recently, Nixon et al. (8) showed that the challenge of non-biological
+variation in sequencing depth be viewed as a problem of
+partially-identified models. They showed that \emph{all} normalizations
+make some assumption about scale. These implicit assumptions are often
+difficult to interpret, and different normalizations provide different
+outputs when applied to the same dataset (3, 19--21). Intuitively,
+normalizations in widespread use assume that either all samples have the
+same scale, e.g.~proportions, rarefaction (22), RPKM (23, 24), etc; or
+that a subset of features in one sample can be chosen as a reference to
+which the others are scaled e.g.~the TMM,(25), or LVHA (7); or that
+different sub-parts of each sample maintain a constant scale across
+samples e.g.~the RLE (26); or that the geometric mean of the parts is
+appropriate e.g the CLR (27) and its derivatives.
+
+The original ALDEx2 (28) model made a strict assumption about scale
+through the CLR normalization (8). While this assumption could be useful
+in many cases it could never be exactly true, and others have shown that
+it is not always the best option (29). Nixon et al. (8) showed that
+better scale assumptions resulted in more reproducible data analysis
+including better control of both false positive and false negative
+results. In essence ALDEx2 has been modified to explicitly model the
+scale over a range of reasonable normalization parameters. Here, we
+briefly introduce these modifications and then show how scale
+uncertainty can greatly improve modeling in transcriptome and
+meta-transcriptome datasets to provide more robust and reproducible
+results.
+
+\section{Implementation}\label{implementation}
+
+To be concrete, we let \(\mathbf{Y}\) denote the \emph{measured}
+\(D \times N\) matrix of sequence counts with elements
+\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
+mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
+Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount of
+class \(d\) in the biological system from which sample \(n\) was
+obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
+scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
 \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
@@
 \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
-The technical variation in sequencing depth
+
+All the normalizations in current use are data transformations that can
+be stated as ratios of the form
+\(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
+denominator is determined by some function of the observation and the
+\(\hat{{\mathbf{W}}}\) notation indicates that the normalization is
+attempting to provide an estimate of the true value. The technical
+variation in sequencing depth
 (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
@@
 
-Early on, researchers recognized the potential confounding effects of
-the non-biological variation in sequencing depth and approached the
-problem using various forms of \textit{normalization}. The idea of
-normalization is to remove this variation thus making the samples
-commensurate and to correct any minor asymmetries in the data. A large
-number of normalizations were developed. Rarefaction (Hughes and
-Hellmann 2005) was used in the microbiome field, while transcriptome
-investigators developed and used proportions and derivatives (reads per
-kilobase per million, and transcripts per kilobase per million)
-(Mortazavi et al. 2008; Wagner, Kin, and Lynch 2012), the relative log
-expression (RLE) (Anders and Huber 2010) and trimmed mean of M values
-(TMM) (Robinson and Oshlack 2010). The Centered Log Ratio (CLR)
-(Aitchison 1982) was proposed as a unifying normalization (Fernandes et
-al. 2014). With the exception of rarefaction, all of these
-normalizations are data transformations that can be stated as ratios of
-the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the
-major differences being how the denominator is chosen and whether the
-ratio is subsequently always log-transformed or not.
-
-Recently, Nixon et al. (2023) suggested that the challenge of
-non-biological variation in sequencing depth be viewed as a problem of
-partially-identified models. Through this lens, they showed that
-\emph{all} normalizations make some assumption about the system scale
-\(\mathbf{W}^{\perp}\) in order to identify the chosen model. Moreover,
-these assumptions are often implicit, difficult to interpret, and
-provided different outputs when applied to the same dataset (Dillies et
-al. 2013; Weiss et al. 2017). Intuitively, normalizations in widespread
-use assume that either one sample can be chosen as a reference to which
-the others are scaled (TMM), or that different sub-parts of each sample
-maintain a constant scale across samples (RLE, CLR and derivatives).
-Further, Nixon et al. (2023) showed that the original ALDEx2 model made
-a strict assumption about scale through the CLR normalization. They
-showed that this assumption could be remedied since ALDEx2 permits scale
-to be easily incorporated into its framework. In this paper, we discuss
-these modifications and show how including scale uncertainty can greatly
-improve modeling and lead to more robust and reproducible results.
-
-\hypertarget{adding-scale-uncertainty-in-aldex2}{%
 \subsection{Adding Scale Uncertainty in
-ALDEx2}\label{adding-scale-uncertainty-in-aldex2}}
-
-The ALDEx2 R package (Fernandes et al. 2013) is a general purpose
-toolbox for Bayesian modeling of HTS data. For brevity and concreteness,
-we discuss ALDEx2 in its simplest form as a tool for estimating
-Log-Fold-Changes (LFC) (e.g., differential abundance or differential
+ALDEx2}\label{adding-scale-uncertainty-in-aldex2}
+
+The ALDEx2 R package (28) is a general purpose toolbox for Bayesian
+modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
+form as a tool for estimating the magnitude and statistical significance
+of Log-Fold-Changes (LFC) (e.g., differential abundance or differential
 expression analysis), but note that it can be used to fit more complex
 linear models. At a high-level, ALDEx2 involves three steps: 1) a
-Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given
-observations \(\mathbf{Y}\); 2) a normalization is used to estimate
-\(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the
-\(\mathbf{W}\) estimates are used to estimate LFC as part of
-differential abundance (or expression) analyses. For more details on
-ALDEx2 see Fernandes et al. (2013).
+Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\)
+given observations \(\mathbf{Y}\); 2) a normalization is used to
+estimate \(\mathbf{\hat{W}}\) given the estimate
+\(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates
+are used to estimate LFC as part of differential abundance (or
+expression) analyses. For more details on ALDEx2 see (8, 28).
 
 By default, ALDEx2 uses the CLR normalization which can be written as
-\(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where
-\(G_{n}\) denotes the geometric mean of the vector
-\((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental
-Materials for how the CLR connects to Shannon entropy). Nixon et
-al.~(2023) showed that in this context the CLR normalization equates to
-an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\). They
-showed that even slight errors in this assumption introduced bias into
-LFC estimate that was not accounted for in estimates of uncertainty
-(e.g., confidence intervals or credible sets). In fact, they showed that
-the only way in which the ALDEx2 model, or any model that uses
-normalizations, could ever be calibrated (e.g., control Type-I Error
-rates) was if this assumption was exactly true. While this assumption
-may be approximately correct in some instances it is surely never
+\(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\)
+where \(G_{n}\) denotes the geometric mean of the vector
+\((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR
+normalization equates to an implicit assumption that
+\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)(8) and even slight errors in
+this assumption introduces bias into the LFC estimate that is not
+accounted for when estimating uncertainty (e.g., confidence intervals or
+credible sets). In fact, Nixon et al. (8) showed that the only way in
+which the ALDEx2 model, or any normalization-based model, could ever be
+calibrated (e.g., control Type-I Error rates) was if this assumption was
 exactly true. In support of this assertion it is known that false
-discovery rates vary widely by analysis and normalization method
-(Hawinkel et al. 2018; Li et al. 2022). As a result, Nixon et al.~(2023)
-suggested that models should consider potential error in the assumptions
-implied by normalizations.
-
-Nixon et al.~(2023) proposed the concept of a \textit{scale model} as a
-generalization of normalizations that account for potential error. They
-showed how scale models can be incorporated into ALDEx2, turning the
-ALDEx2 model into a specialized type of statistical model which they
-called a \textit{Scale Simulation Random   Variable} (SSRV). Recall that
-\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
-As the HTS data only informs directly on
-\(\mathbf{W}^{\parallel}_{dn}\), they proposed to augment analyses with
-a model for \(\mathbf{W}^{\perp}_{n}\) (the scale model). Since the CLR
-normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\),
-the CLR normalization can be generalized by considering probability
-models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean
-\(1/G_{n}\). For example, the following scale model generalizes the CLR:
-
-\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
-
-where \(\gamma\) is a tunable parameter that controls the degree of
-uncertainty in the CLR assumption and \(x_{n}\) denotes a binary
-condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\)
-denotes control). Beyond theoretical studies, they demonstrated through
-the analysis of simulated and real data the remarkable performance
-improvements that could result from incorporating this type of scale
-uncertainty into analyses. Subsequently, we have made those
-modifications a permanent fixture of ALDEx2 which now represents the
-first software package designed for SSRV-based inference.
-
-Beyond concerns of fidelity and rigor, scale models also enhance the
-reproducibility and transparency of HTS analyses. This occurs because
-rather than using an implicit assumption (e.g.,
-\(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale
-models make this an explicit part of the model building process. While
-it is beyond the scope of the present article, we note that there are
-many ways of building scale models that enhance the interpretability of
-the parameters and assumptions. Moreover, the scale model listed above,
-which is now the default in ALDEx2, also has an appealing
-interpretation. A detailed description of these points as well as the
-mathematical underpinnings of the updated ALDEx2 R package is
-forthcoming and will be made available as a companion publication. Here,
-we simply treat the parameter \(\gamma\) in the default scale model as
-an abstract term that represents uncertainty in the CLR assumption and
-note that larger values of \(\gamma\) correspond to more uncertainty in
-this assumption. We also show that the ratio of scales between
-conditions can be used to build a full scale model when the conditions
-have dramatically different underlying scales.
-
-An advantage of incorporating scale is that analyses can be made much
-more robust such that actual or potential differences in scale can be
-tested and accounted for explicitly. The examples below and in the
-supplement show how incorporating scale provides robust and
-interpretable differential abundance estimates in several different
-datasets.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}{%
+discovery rates vary widely by analysis, dataset, and normalization
+method (30, 31). To fix this, Nixon et al. (8) showed that models should
+incorporate potential error in the assumptions implied by
+normalizations.
+
+Nixon et al.~(2023) generalized the concept of normalizations by
+introducing the concept of a \textit{scale model} to account for
+potential error. Scale models can be incorporated into ALDEx2, turning
+the ALDEx2 model into a specialized type of statistical model which they
+called a \textit{Scale Simulation Random   Variable} (SSRV). They did
+this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). Since
+the CLR normalization makes the assumption
+\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be
+generalized by considering probability models for the scale
+\(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
+example, the following scale model generalizes the CLR:
+
+\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
+
+where \(\gamma\) is a tunable parameter drawn from a Gaussian
+distribution that controls the degree of uncertainty in the CLR
+assumption and \(x_{n}\) denotes a binary condition indicator (e.g.,
+\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). We have made
+those modifications a permanent fixture of ALDEx2 which now represents
+the first software package designed for SSRV-based inference.
+
+\section{Results}\label{results}
+
 \subsection{Adding scale uncertainty replaces the need for dual
 significance
-cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}}
-
-Gierliński et al. (2015) conducted a highly replicated yeast
-transcriptome experiment comparing a wild-type strain with a snf1 gene
-knockout. This dataset has been used to argue that only a small number
-of replicates are needed to identity differentially abundant genes and
-makes the recommendation that some tools should be used in some
-conditions because each tool has different intrinsic statistical power
-and Type 1 and Type 2 errors (Schurch et al. 2016){]}. This guidance
-runs counter to standard statistical practice where power is
-intrinsically linked to sample size (Halsey et al. 2015), yet the
-concept of sample-size independent power is entrenched in all fields
-that use HTS as an experimental readout. Through the lens of this paper,
-this is a hallmark of unacknowledged bias: a false certainty in the
-scale of the data brings about an increasing false precision with
-increasing sample sizes. Furthermore, the yeast transcriptome experiment
-has an unacknowledged difference in scale between conditions: yeast
-deficient for snf1 are smaller, grow more slowly and are sensitive to a
-variety of common agents that cause cell stress (Yoshikawa et al.
-2011){]}. Thus, by adding scale uncertainty, we can be more confident
-that the results of the analysis reflect true biological differences and
-not differences due to the assumptions made by the normalization method.
-
-We start by examining the dispersion (variance) of the data as counts
-and as the logarithm of the normalized counts when calculated by DESeq2
-(log(RLE)) or by ALDEx2 (CLR). The raw data counts derived from
-sequencing are overdispersed with the mean value being less than the
-variance (Robinson and Oshlack 2010; Robinson, McCarthy, and Smyth 2010)
-as seen in Panel A of Figure 1. However, the actual analysis of
-differential abundance is performed on the logarithm of the normalized
-counts (Robinson, McCarthy, and Smyth 2010; Love, Huber, and Anders
-2014; Andrews and Hemberg 2019), which includes the CLR (Fernandes et
-al. 2013) yet the mean-dispersion distribution of these log-ratio
-transformed data is quite different as shown in panels B and C of Figure
-1 and as noted elsewhere (Fernandes et al. 2013; Love, Huber, and Anders
-2014).
+cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}
+
+It is standard practice in many fields of HTS, but especially
+transcriptomics, to use the dual cutoff approach graphically exemplified
+by volcano plots (32, 33) because not all statistically significant
+differences are biologically relevant. Paraphrasing this we can say that
+in in some types of datasets more samples leads to a majority of
+features being statistically significant. A detailed reasoning for this
+using modelled data is explained in two reports by Nixon et al. (8, 9),
+which shows that scaled models completely address this analytic issue.
+
+We use the data of Gierliński et al. (34) who conducted a highly
+replicated yeast transcriptome experiment comparing a wild-type strain
+with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to
+argue that a dual cutoff approach is appropriate to limit the number of
+significant parts and for purposes of reproducibility (33). However, in
+this benchmarking study, the number of significantly different
+transcripts varied between 65\% to \textgreater80\% of all transcripts
+depending on the tool; in essence the desired behavior was that almost
+all transcripts were significantly different. The guidance on
+reproducibility runs counter to standard statistical practice where
+power is intrinsically linked to sample size and very large sample sizes
+are indeed desired (35). Furthermore, the dual-cutoff approach is known
+not to provide appropriate FDR control (5, 6). Schurch et al. (33)
+further suggested that different tools might be better in some
+conditions or datasets than others because each tool has different
+intrinsic statistical power and Type 1 and Type 2 errors. Through the
+lens of scale uncertainty, the behaviour of the tools in this study show
+unacknowledged bias; false confidence in the precision of the estimate
+as sample size increases because o mis-specification in the scale of the
+data. This certainty is driven by the assumptions of the tool not the
+actual experiment being investigated (9).
+
+Using either DESeq2 or ALDEx2, a majority of transcripts are
+statistically significantly different between groups with a
+Benjamini-Hochberg (36) false discovery rate (FDR) of 0.05; i.e.~4636
+(79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
+large numbers of significant transcripts seems biologically unrealistic
+and furthermore breaks the necessary assumption made by the
+normalization methods that the majority of the features must be
+invariant. That 118 transcripts are identified by ALDEx2 and not DESeq2,
+while DESeq2 identifies 582 transcripts that ALDEx2 does not, suggests
+that the choice of normalization plays a role in which results are
+returned as significant and that some if not the majority are driven by
+technical differences in the analysis (20, 21, 33, 37, 38).
 
@@
 \centering
-\includegraphics{go3_files/figure-latex/dispersion-1.pdf}
-\caption{Plot of abundance v dispersion for a typical transcriptome
-dataset as counts, as logarithms of counts, and as clr values. Panel A
-shows that the data are over-dispersed relative to a Poisson
-distribution which is represented by the dashed line when plotted on a
-log-log scale. Panel B shows that the relationship between the mean and
-the dispersion calculated in DESeq2, here the standard error (SE) of the
-mean, is very different when the data are log-transformed first. Panel C
-shows the equivalent values calculated by ALDEx2 in which the expected
-clr value for each transcipt are plotted vs.~the expected dispersion.
-The red line in each panel shows the loess line of fit to the mid-point
-of the distributions. In panels B and C the amount of dispersion reaches
-a minimum at moderate values. The dashed orange line in panel A is the
-line of equivalence, and in panel B and C is the minimum y value. The
-values below the dashed grey line in panels B and C represent those
-below the first decile of dispersion.}
+\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/plot1-1.pdf}}
+\caption{Effect and volcano plots for unscaled and scaled transcriptome
+analysis. ALDEx2 was used to conduct a differential abundance (DA)
+analysis on the yeast transcriptome dataset. The results were plotted to
+show the relationship between difference and dispersion using effect
+plots ) or difference and the Benjamini-Hochberg corrected p-values
+(volcano plot). Panels A,C are for the unscaled analysis, and Panels B,D
+are for the scaled analysis. Each point represents the values for one
+transcript, with the color indicating if that transcript was significant
+in the scaled analysis and unscaled analysis (red) or in the unscaled
+analysis only (orange). Points in grey are not statistically
+signficantly different with any analysis. The horizontal dashed lines
+represent a log2(difference) of \(\pm 1.4\).}
 \end{figure}
 
-These relationships depend on what is being analyzed: the variance or
-dispersion always increases with increasing raw (or normalzied) read
-count but decreases when measured on the log-ratio transformed data
-(Fernandes et al. 2013; Love, Huber, and Anders 2014), reaching a
-minimum at some mid-point of the distribution. This makes the
-counter-intuitive suggestion that genes with moderate expression have
-more predictable expression than genes with very high expression such as
-housekeeping genes. This is at odds with the known biology of cells
-where single cell counting of housekeeping transcripts shows that they
-are both highly expressed and have little intrinsic variation (Taniguchi
-et al. 2010). Furthermore, the dispersion is exceeding small being, for
-many transcripts, almost negligible. To show this point more clearly,
-the majority of the transcripts in the lowest decile of dispersion
-indicated below the dashed grey line are statistically significantly
-different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low
-dispersion estimates lead to many false positives. From the perspective
-of Nixon et al. (2023), it is reasonable to conclude that some of these
-results may be due to the choice of normalization. Indeed, benchmarking
-and comparison studies have repeatedly shown that the choice of
-normalization plays a role in which results are returned as significant;
-see for example (Maza et al. 2013), (Dillies et al. 2013) and (Weiss et
-al. 2017).
-
-Using either DESeq2 or ALDEx2, a majority of transcripts are
-statistically significantly different between groups even with a
-Benjamini-Hochberg (Benjamini and Hochberg 1995) false discovery rate
-(FDR) of 0.01; i.e.~4264 or 3790 of the 5891 transcripts are
-statistically significant. Clearly, a 70\% positive rate seems
-biologically unjustified, and furthermore this breaks the necessary
-assumption made by the normalization methods that most features must be
-invariant. Finally, that 90 transcripts are identified by ALDEx2 and not
-DESeq2, while DESeq2 identifies 564 transcripts that ALDEx2 does not
-suggests that the choice of normalization plays a role in which results
-are returned as significant.
+The effect plots (39) in Figure 1A (ALDEx2) and Supplementary Figure 1
+(DESeq2) shows that the majority of significant transcripts (red,
+orange) have negligible differences between groups and very low
+dispersion and we suggest that this is driven by the experimental design
+(33). Scale uncertainty can be incorporated using the \texttt{gamma}
+parameter that controls the amount uncertainty added to the CLR mean
+assumption when we call either \texttt{aldex()}, or
+\texttt{aldex.clr()}. Figure 1B shows that setting \(\gamma=0.5\)
+results in far fewer transcripts being significant (205) and we observe
+that the minimum dispersion increases from 0.12 (unscaled) to 0.67
+(scaled). The Volcano plots in Figure 1 C and D show a similar story.
+Here we can see that adding scale increases the minimum FDR value and
+increases the concordance between the FDR value and the difference
+between groups (compare panels C and D).
 
@@
 \centering
-\includegraphics{go3_files/figure-latex/plot1-1.pdf}
-\caption{Effect and volcano plots for unscaled and scaled transcriptome
-analysis. DESeq2 or ALDEx2 were used to conduct a differential abundance
-(DA) analysis on the yeast transcriptome dataset. The results were
-plotted to show the relationship between difference and dispersion using
-effect plots ) or difference and the Benjamini-Hochberg corrected
-p-values (volcano plot). Panels A,B,D,E are for the unscaled analysis,
-and Panels C,F are for the scaled analysis. Each point represents the
-values for one transcript, with the color indicating if that transcript
-was significant in the scaled analysis and unscaled analysis (red) or in
-the unscaled analysis only (orange). Points in grey are not
-statistically signficantly different with any analysis. The horizontal
-dashed lines represent a log2(difference) of 1.4, which is a commonly
-applied cutoff when the majority of features are statistically
-significant.}
+\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/disp-1.pdf}}
+\caption{Adding scale uncertainty changes the dispersion distribution.
+Panel A shows a plot of the expected value for relative abundance vs the
+expected value for the pooled dispersion as output by
+\texttt{aldex.effect}. The dashed horizontal lines show the median value
+for the features with a rAbundance between -0.5 and 0.5, and the light
+colored lines are lowess lines of fit through the center of mass of the
+data. Panel B plots the dispersion difference between \(\gamma = 1\) and
+\(\gamma = 0\); note the non-linear relationship that highlights the
+rotation that is evident in Panel A. The colored lines indicate the
+lowess line of fit through the centre of mass of the plot for the
+various populations of points. The grey line is the total population and
+shows the difference \(\Delta\), the red line is the population of
+significant transcripts (*) with no scale, the orange line is the
+population of significant transcripts with a difference threshold (T) of
+about \(\pm 2^{1.4}\)-fold change, the blue line is the population of
+significant transcripts with \(\gamma = 0.5\), and the cyan line is the
+significant population with \(\gamma = 1\). \(\Delta\): Difference, *:
+significant, T: thresholded.}
 \end{figure}
 
-As shown using effect (Gloor, Macklaim, and Fernandes 2016) and volcano
-plots (Cui and Churchill 2003) in Figure 2 A,B,D,E, the root cause of
-the many statistically significant positive transcripts is the very
-large number of transcripts with negligible variance. This phenomenon is
-shared between methods: almost all the transcripts that are
-differentially abundant with an FDR \textless 0.01 (orange and red
-points) have extremely low dispersion and a very low difference between
-groups. In the most extreme cases, some transcripts have a near zero
-difference between groups. This issue is not unique to this data set and
-it is common practice to use a dual-cutoff by choosing transcripts based
-on a thresholds for both corrected p-values and fold-changes (Schurch et
-al. 2016) (commonly set at \(\pm 2^{1.4}\) for the latter). These limits
-are shown by the dashed grey lines. In this case, applying a dual-cutoff
-using a heuristic of at least a \(2^{1.4}\) fold change reduces the
-number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
-This result begs the question: why bother with significance tests at
-all?
-
-The very low dispersion estimate for most features is a by-product of
-sequencing and normalization. While the actual scale of the data is
-inaccessible post-sequencing, we can study the sensitivity of the
-results to the choice of normalization and thus scale. To do this we
-incorporate scale models (Nixon et al. 2023) in lieu of normalizations
-through the default scale model that is now built into ALDEx2. Scale
-uncertainty is incorporated using the \texttt{gamma} parameter that
-controls the amount uncertainty added to the CLR mean assumption when we
-call either \texttt{aldex()}, or \texttt{aldex.clr()}. The
-\texttt{ALDEx2} package further contains a sensitivity analysis
-function, \texttt{aldex.senAnalysis()}, that can be used to explore the
-effect of different amounts of scale uncertainty. In practice, we
-suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic
-for most experimental designs.
-
-Applying the default scale model with \texttt{gamma=1}, a large number
-of transcripts with near 0 dispersion have had their dispersion
-increased (Figure 2C), and this results in many fewer transcripts 100
-being called significantly different as shown in the volcano plot in
-Figure 1F (red points). Furthermore, overplotting the significant
+As shown by the effect plot in Figure 1 the root cause of the many
+statistically significant positive transcripts is the very large number
+of transcripts with negligible variance. This phenomenon is not unique
+to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon
+occurs in DESeq2 (and presumably other methods although the relevant
+parameters are not exposed). This issue is not unique to this dataset
+and it is common practice to use a dual-cutoff by choosing transcripts
+based on a thresholds for both corrected p-values and fold-changes (33)
+(here set at \(\pm 2^{1.4}\) for the latter), although considerable
+variation in cutoff values is observed. These limits are shown by the
+dashed grey lines. Here, applying a dual-cutoff using a heuristic of at
+least a \(2^{1.4}\) fold change reduces the number of significant
+outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with
+that observed with that found by ALDEx2 with \(\gamma = 0.5\) which
+identifies 205. Indeed, Supplementary Figure 2, shows that even adding a
+very small amount of scale \(\gamma = 0.1\) reduces the number of
+significant transcripts by more than half and the
+\texttt{aldex.scaleSim()} function can be used to identify those
+transcripts that are significant only because of an absence of scale.
+These results begs the question: why bother with significance tests at
+all if all transcripts with \(> 1.4\)-fold expression change are
+statistically significant?
+
+The effect on dispersion with increasing amounts of scale are shown in
+Figure 2A, where we can see that the dispersion increases as scale is
+added. Note that the dispersion in the unscaled data in Figure 2A
+reaches a minimum near the mid-point of the distribution, and also does
+so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
+This shows more clearly that dispersion of many transcripts is almost
+negligible in the absence of scale and makes the counter-intuitive
+suggestion that the variance in expression of the majority of genes with
+moderate expression is more predictable than highly-expressed genes or
+of housekeeping genes (40). This is at odds with the known biology of
+cells where single cell counting of highly-expressed transcripts shows
+that they have little intrinsic variation (41).
+
+Adding scale by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases
+the minimum dispersion as shown in Figure 2A by the red and blue data
+points, and by the colored lines of fit through the centre of mass of
+the data. Less obvious is that the scaled dispersion estimates are
+rotated. Figure 2B shows a plot of the difference between the
+\(\gamma= 0\) and \(\gamma= 1\) data to show this more clearly and here
+we can see that the scale is preferentially increasing the dispersion of
+the mid-expressed transcripts that formerly had negligible dispersion;
+examine the grey line of best fit (overlaid by the red line) for the
+trend. Panel B also shows the trend of the expression-dispersion
+relationship for transcripts that are classed as statistically
+significant. The red line shows the trendline with no added scale, and
+this trendline exactly overlays with the grey trendline of the bulk of
+transcripts. The orange trendline indicates those transcripts that are
+both statistically significant and that have a thresholded expression
+level of \(\pm 1.4\), and the dark blue and cyan lines show the
+statistically significant trendline for \(\gamma=0.5\) , or \(1.0\).
+Note that this has the effect of changing the distribution of parts
+identified as significant and that the substantially fewer significantly
+different genes are in the very high abundance but low dispersion
+category.
+
+\subsection{Housekeeping genes can be used to guide scale model
+choices.}\label{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}
+
+Dos Santos et al. (42) used a vaginal metatranscriptome dataset to
+compare the gene expression in bacteria collected from healthy (H) and
+bacterial vaginosis (BV) affected women. In this environment, both the
+relative abundance of species between groups and the gene expression
+level within a species is different (43). Additionally, prior research
+suggests that the total number of bacteria is about 10 times more in the
+BV than in the H condition (16). Thus, this is an extremely challenging
+environment in which to determine differential abundance as there are
+both compositional and scale changes between conditions. The usual
+method to analyze vaginal metratranscriptome data is on a taxon-by-taxon
+basis (43--45) because the scale confounding can be ignored. Attempts at
+system-wide analysis show that many housekeeping functions are returned
+as differentially abundant between groups; a result likely due to a
+disconnect between the scale assumptions of the normalization used (7).
+
+In this example, we show how to specify a user-defined, or
+\emph{informed}, scale model explicitly that can account for some of
+these modeling difficulties. An informed scale model can control for
+both the mean difference of scale between groups (e.g., directly
+incorporate information on the differences in total number of bacteria
+between the BV and H conditions) as well as the uncertainty assumed in
+that difference. To specify a user-defined scale model, we can pass a
+matrix of scale values instead of a single estimate of gamma to
+\texttt{aldex.clr()}. This matrix should have the same number of rows as
+the of Monte-Carlo Dirichlet samples, and the same number of columns as
+the number of samples. While this matrix can be computed from scratch by
+the analyst, there is an \texttt{aldex.makeScaleModel()} function that
+can be used to simplify this step in most cases. This encodes the scale
+model as \(\Lambda \sim N(log2 \mu_n, \gamma^{2})\), where \(\mu_n\)
+represents the actual scale value for each sample. This can be a
+measured value (cell count, nucleic acid input, etc), or an imputed
+value. Nixon et al. (9) showed that only the ratio between the scale
+values for each sample was important; see Supplementary Figures 4 and 5
+for a demonstration of this point.
+
+Figure 3A shows an effect plot of the data where reads are grouped by
+function, corresponding to grouping homologous sequences regardless of
+the organism of origin. Each point represents one of 3728 KEGG functions
+(46). There are many more functions represented in the BV group (bottom)
+than in the healthy group (top). This is because the
+\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
+reduced genome content relative to the anaerobic organisms that dominate
+in BV, because there is a greater diversity of organisms in BV than in H
+samples, and because the BV condition has at least an order of magnitude
+more bacteria than does the H condition.
+
+There are 101 functions with low dispersion that appear to be share by
+both groups (boxed area in Figure 3A, and colored in cyan). Inspection
+shows that these largely correspond to core metabolic functions such as
+transcription, translation, ribosomal proteins, glycolysis, replication,
+chaperones, etc (Supplementary file housekeeping.txt). The transcripts
+of many of these are commonly used as invariant reference sequences (40)
+and so would not be expected to contribute to differences in ecosystem
+behaviour and should be centred on 0 difference. These should not be
+scored as among the most differentially abundant. The major group of
+these housekeeping functions is located off the line of no difference
+(being approximately located at +1.5). While changes in the abundance of
+housekeeping functions is a useful proxy for relative abundance of
+species in the environment, they tell us nothing about the functional
+capacity of the two groups as these are common to every organism. Of
+more interest is determining the functions that are different between
+groups because these are unique or over-expressed in one group relative
+to the other.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/meta-1.pdf}}
+\caption{Analysis of vaginal transcriptome data aggregated at the Kegg
+Orthology (KO) functional level. Panel A shows an effect plot for the
+default analysis where the functions that are elevated in the healthy
+individuals have positive values and functions that are elevated in BV
+have negative values. Highlighed in the box are KOs that are almost
+exlusively housekeeping functions and are colored cyan. These
+housekeeping functions should be located on the midline of no
+difference. Panel B shows the same data scaled with \(\gamma = 0.5\),
+which increase the minimum dispersion as before. Panel C shows the same
+data scaled with \(\gamma = 0.5\) and a 0.15 fold difference in
+dispersion applied to the BV samples relative to the H samples. In these
+plots statistically significant (FDR \textless{} 0.01) functions in the
+informed model are in red, false positive functions are in blue,
+non-significant functions in black and false negative functions are in
+orange.}
+\end{figure}
+
+Applying the default scale model of \(\gamma=0.5\) increases the
+dispersion as expected but does little to move the large number of
+housekeeping functions toward the midline of no difference. This is as
+expected; the mean of the default scale model is based on the CLR
+normalization so no shift in location would be expected over the
+original ALDEx2 model. Nevertheless, about 30\% of the housekeeping
+functions are no longer statistically significantly different. Note that
+this change is simple to conduct, has no additional computational
+complexity and requires only a slight modification from the analyst.
+
+Up to this point, scale uncertainty has been applied as an extension of
+the CLR normalization via the default scale model, but a user-defined
+scale adjustment can be applied to each condition, or even each sample
+independently through a custom scale matrix. Our starting point for this
+is to identify the naive scale estimate from the data, and this can be
+done by calculating the mean scale value for each group. The scale
+estimate for samples can be accessed in the \texttt{@scaleSamps} slot
+from the \texttt{aldex.clr} output. In this datset the scale estimate
+for the healthy group is 17.41 and for the BV group is 14.59 for a
+difference of 2.82. This is interpreted as the scale of the H group of
+samples being 7.06 greater than the BV group and this precise but
+incorrect estimate places the location of the housekeeping genes off the
+midline of no difference.
+
+We desire a scale model that approximately centres the housekeeping
+functions. We anticipate that housekeeping functions should be nearly
+invariant and so an appropriate scale in this dataset is likely closer
+to 0 than the naive estimate. While a user-defined scale model can be
+quite flexible with relative scales that are distinct for each group (or
+even each sample) along with their uncertainties, here we focus on using
+the \texttt{aldex.makeScaleMatrix()} function. This function uses a
+logNormal distribution to build a scale matrix given a user-specified
+mean difference between groups and uncertainty level. Applying a
+per-group relative differential scale of 0.15 moves the housekeeping
+functions to the midline of no difference (Figure 3C), and applying a
+gamma of 0.5 provides the same dispersion as in panel B). Note that now
+a significant number of functions are differentially up in BV that were
+formerly classed as not different without the full scale model (orange),
+or when only a default scale was applied. Inspection of the functions
+shows that these are largely missing from the \emph{Lactobacillus}
+species and so should actually be captured as differentially abundant in
+the BV group. Thus, applying a differential scale allows us to
+distinguish between both false positives (housekeeping functions in
+cyan, and others in blue) and false negatives (orange functions) even in
+a very difficult to analyze dataset. The remarkable improvements in
+biological interpretation afforded by this full scale model, and the
+transferrability of it between sample cohorts of the same condition is
+outlined elsewhere (42). We suggest that the default scale model is
+sufficient when the data are approximately centred. However, an informed
+model is more appropriate with datasets are not well centred or when the
+investigator has prior information about the underlying biology.
+
+\section{Discussion}\label{discussion}
+
+\doublespacing
+\singlespacing
+
+Biological systems are both predictably variable and stochastic (41) and
+current measurement methods that rely on high throughput sequencing fail
+to capture all of that variation, particularly variation due to scale
+(8, 9). In the absence of external information (such as spike-in probes
+(12), cell counting (1), FISH (47) etc) sequencing depth normalisation
+methods cannot recapture the scale information (12), and can only
+normalize for the technical variation due to sequencing depth.
+
+Many groups have conducted benchmarking studies on different tools and
+normalizations used for the analysis of datasets such as
+transcriptomes(4, 19, 33, 37) and microbiomes (3, 21, 29, 30, 48).
+Generically, it is observed that the actual agreement between analysis
+methods can be modest, and no single method outperforms all others in
+every dataset or type of experiment. That different tools appear to work
+more reliably in different datasets from different sources can be
+explained as different normalizations being a better fit to the scale a
+particular dataset by chance.
+
+From our perspective, the disagreement between tools can be explained by
+the partially identifiable nature of the data; that is, the observed
+data are consistent with multiple ways of estimating the parameters (8).
+Partially identifiable data exhibit multiple pathologies, chief among
+them being that that different analytic approaches can produce different
+parameter estimates and that more data produces worse estimates because
+the additional data increases the precision of a flawed estimate (49).
+In the analysis of HTS data it is often observed that larger datasets
+converge on the majority of parts being significantly different (8, 33),
+and that different analytic approaches result in different parts being
+chosen as significantly different (4, 33, 37).
+
+Scale simulation is now build into ALDEx2 to address the issue of
+partially identifiability, and through this lens the two main root
+causes to common HTS data pathologies can be proposed. The first
+contributing factor is the observed very low dispersion estimate for
+many features that is a by-product of experimental design, sequencing
+and normalization. In the Schurch et al. (33) dataset, the data were
+from single colonies derived from a single culture. Thus, it is more
+accurate to describe the 96 samples as wet-lab technical replicates
+rather than independent samples. However, this type of replicate is
+standard in the molecular literature, and would be expected to result in
+the very low dispersion that is observed.
+
+In the yeast transcriptome dataset, applying the default scale model
+with \(\gamma=0.5\) a large number of transcripts with near 0 dispersion
+have had their dispersion increased (Figure 1D), and this results in
+modest number of transcripts, 205, being called significantly different
+as shown in the volcano plot in Figure 1D (red points). In addition,
+there was now a strong concordance between the difference and p-values
+(Figure 1D), this is not surprising. In hindsight, what is not obvious
+is why the unscaled volcano plot shows such poor correspondence. We
+suggest that this can be explained by random fluctuations in the many
+variance estimates with very low values, and this is supported by the
+plot shown in Figure 2B. Furthermore, overplotting the significant
 transcripts identified after adding scale uncertainty on the un-scaled
@@
 
-\hypertarget{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}{%
-\subsection{Housekeeping genes can be used to guide scale model
-choices.}\label{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}}
-
-Macklaim and Gloor (2018) and Wu et al. (2021) used a vaginal
-metatranscriptome dataset used to compare the gene expression in
-bacteria collected from healthy (H) and bacterial vaginosis (BV)
-affected women. In this environment, both the relative abundance of
-species between groups and the gene expression level within a species is
-different (Macklaim et al. 2013). Additionally, prior research suggests
-that the total number of bacteria is about 10 times more in the BV than
-in the H condition (Zozaya-Hinchliffe et al. 2010). Thus, this is an
-extremely challenging environment in which to determine differential
-abundance as there are both compositional and scale changes between
-conditions. The accepted method to analyze vaginal metratranscriptome
-data is on a taxon-by-taxon basis (Macklaim et al. 2013; Deng et al.
-2018; Fettweis et al. 2019) because the scale confounding can be
-ignored. The problem with a systemic analysis is that many housekeeping
-genes are returned as differentially abundant between groups, a result
-likely due to a disconnect between the scale assumptions of the
-normalization used (Wu et al. 2021) and the actual change in scale
-between conditions.
-
-In this example, we show how to specify the scale model explicitly and
-demonstrate that applying a user-defined scale model can account for
-some of these modeling difficulties. A user-defined scale model can
-control for both the mean difference of scale between groups (e.g.,
-directly incorporate information on the differences in total number of
-bacteria between the BV and H conditions) as well as the uncertainty
-assumed in that difference. To specify a user-defined scale model, we
-can pass a matrix of scale values instead of a single estimate of gamma
-to \texttt{aldex.clr()}. This matrix should have the same number of rows
-as the of Monte-Carlo Dirichlet samples, and the same number of columns
-as the number of samples. While this matrix can be computed from scratch
-by the analyst, there is an \texttt{aldex.makeScaleModel()} function
-that can be used to simplify this step in most cases.
-
-Figure 3A shows an effect plot (Gloor, Macklaim, and Fernandes 2016) of
-the data where reads are grouped by function, corresponding
-approximately to grouping orthologous sequences regardless of the
-organism of origin. Each point represents one of the 3728 functions.
-There are many more functions represented in the BV group (bottom) than
-in the healthy group (top). This is because the \textit{Lactobacilli}
-that dominate a healthy vaginal microbiome have reduced genome content
-relative to the anaerobic organisms that dominate in BV, because there
-is a greater diversity of organisms in BV than in H samples and because
-the BV condition has at least an order of magnitude more bacteria than
-does the H condition.
-
-We can see that there are a large number of core metabolic and
-information processing functions that are shared between the two groups
-(boxed area in Figure 3A), and inspection shows that these largely
-correspond to core metabolic functions that would not be expected to
-contribute to differences in ecosystem behaviour and so should not be
-differentially abundant. As a proxy for housekeeping functions the core
-ribosome functions (blue) shows that their mean location is not centred
-on 0. The major group of these housekeeping functions is located off the
-line of no difference (being approximately located at +1.5). Since they
-have very low dispersions, these functions are identified as
-differentially abundant (red) along with many others. While changes in
-the abundance of housekeeping functions is a useful proxy for relative
-abundance of species in the environment, they tell us nothing about the
-functional capacity of the two groups because these are functions in
-common to every organism. Of more interest is determining the functions
-that are different between groups because these are unique or
-over-expressed in one group relative to the other.
-
-\begin{figure}
-\centering
-\includegraphics{go3_files/figure-latex/meta-1.pdf}
-\caption{Analysis of vaginal transcriptome data aggregated at the Kegg
-Orthology (KO) functional level. Panel A shows an effect plot for the
-default analysis where the functions that are elevated in the healthy
-individuals have positive values and functions that are elevated in BV
-have negative values. Highlighed in the box are highly abundant KOs that
-are almost exlusively housekeeping functions, with ribosomal KOs
-highlighted in blue, statistically significant (FDR \textless{} 0.01)
-functions in red, and non-significant functions in black or orange.
-These housekeeping functions should be located on the midline of no
-difference. Panel B shows the same data scaled with
-\texttt{gamma\ =\ 0.5}, which increase the minimum dispersion
-approximatly by one unit. Here the housekeeping functions from Panel A
-are colored cyan or blue for reference. Panel C shows the same data
-scaled with \texttt{gamma\ =\ 0.5} and a 0.15 fold difference in
-dispersion applied to the BV samples relative to the H samples. The
-orange functions are now statistically significant. Note that this
-shifts the midpoint of the housekeeping functions towards the midline.}
-\end{figure}
-
-Applying the default scale model of \texttt{gamma=0.5} increases the
-dispersion as expected but does little to move the large number of
-housekeeping functions toward the midline of no difference. This is as
-expected: the mean of the default scale model is based on the CLR
-normalization so no shift would be expected over the original ALDEx2
-model. Nevertheless, about 50\% of the housekeeping functions are no
-longer statistically significantly different. Note that this change is
-simple to conduct, has no additional computational complexity and
-requires only a slight modification from the analyst.
-
-Up to this point, scale uncertainty has been applied as an extension of
-the CLR normalization via the default scale model, but a user-defined
-scale adjustment can be applied to each condition, or even each sample
-independently through a custom scale matrix. While not necessary in all
-datasets (e.g., the transcriptome example discussed in the previous
-section and see the Supplementary material), it can greatly improve
-modeling in certain cases, especially when the scale is highly
-asymmetric between conditions. In order to specify a scale model from
-scratch, we need to revisit the concept that all normalizations in
-widespread use are actually ratios with the denominator implied by the
-normalization. Therefore, we can easily deviate from a certain
-normalization (e.g., the geometric mean assumption implied by the CLR)
-by specifying a total model based on the mean difference between
-conditions. While knowing the mean difference between conditions may
-seem cumbersome in practice, it is the \emph{relationship} between the
-group scale values that is important, not their raw values (Nixon et
-al.~2023). This can be illustrated quite simply by starting with the
-mean ratio in \(G\) between groups for the yeast transcriptome dataset
-which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for
-WT; their ratio being 1.17, or a 0.17-fold difference. Using this
-information, we can recapitulate the differential abundance analysis in
-Figure 2B and 2C exactly by using setting the mean denominator of group
-1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental
-Figure 2. This ratio can be adjusted to alter the mean assumption placed
-on the group scale values.
-
-With this background, we can understand how to apply a user-defined
-scale model to the metratranscriptome dataset. While a user-defined
-scale model can be quite flexible with relative scales that are distinct
-for each group (or even each sample) along with their uncertainties,
-here we focus on using the \texttt{aldex.makeScaleMatrix()} function.
-This function uses a logNormal distribution to build a scale matrix
-given a user-specified mean difference between groups and uncertainty
-level. Applying a per-group relative differential scale of 0.15 moves
-the housekeeping functions to the midline of no difference, and applying
-a gamma of 0.5 provides the same dispersion as using a gamma of 0.5 with
-the CLR (Figure 3C). Note that now a significant number of functions are
-differentially up in BV that were formerly classed as not different
-without scale, or when only a default scale was applied. These former
-false negatives are noted in orange in each panel. Inspection of the
-functions shows that these are largely missing from the Lactobacillus
-species and so should actually be captured as differentially abundant.
-Thus, applying a differential scale allows us to distinguish between
-both false positives (housekeeping functions in cyan) and false
-negatives (orange functions) even in a very difficult to analyze
-dataset. We suggest that the default scale model is sufficient when the
-data are approximately centred. However, when datasets are not well
-centred or when the investigator has prior information about the
-underlying biology, we advocate developing and using a user-specified
-scale model.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Biological count data derived from HTS can be decomposed into two parts:
-the relative (compositional) and the absolute (scale), and the product
-of these generates a fully scaled biological system (Nixon et al.~2023).
-Biological systems are both predictably variable and stochastic, and
-current measurement methods that rely on high throughput sequencing fail
-to capture all of that variation, particularly variation due to scale.
-In the absence of information external to the sequencing run itself, no
-normalisation method can recapture any of the scale information (Lovén
-et al.~2012). Even so, many methods such as flow cytometry (Vandeputte
-et al.~2017), spike-in probes and fluorescent in-situ hybridization
-(Lovén et al.~2012; Marguerat et al.~2012) have been proposed as a means
-to do this. In the context of the updated ALDEx2 software suite, these
-methods can be useful if they contain information on the particular
-scale the researcher is interested in.
+While the actual scale of the underlying environment is inaccessible
+post-sequencing, we can study the sensitivity of the results to the
+choice of normalization and thus scale. To do this we incorporate scale
+models (8) in lieu of normalizations through the default scale model
+that is now built into ALDEx2. Scale uncertainty is incorporated using
+the \texttt{gamma} parameter that controls the amount uncertainty added
+to the CLR mean assumption when we call either \texttt{aldex()}, or
+\texttt{aldex.clr()}. The ALDEx2 package contains a sensitivity analysis
+function, \texttt{aldex.senAnalysis()}, that can be used to explore the
+effect of different amounts of scale uncertainty, and an example is
+shown in Supplementary Figure 2 for the yeast transcriptome dataset.
+Here it is clear that even tiny amounts of scale preclude the majority
+of transcripts from being considered statistically significant. In
+practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
+is realistic for most experimental designs, but further work is needed.
+
+The second contributing factor is unacknowledged asymmetry in many
+datasets. In the case of asymmetry, the use of a user-specified scale
+model can be very useful for otherwise difficult-to-analyze datasets
+such as meta-transcriptomes and in-vitro selection datasets where the
+majority of features can change. We showed one such example for the
+metatranscriptome dataset in Figure 3. Here the dataset was highly
+asymmetrical, and the TMM and RLE normalizations could not fully move
+all the housekeeping genes to the midline of no difference or the tools
+exhibited other pathologies (Supplemental Figure 6). Incorporating
+differential scale on a per-group basis moves the mass of the
+housekeeping functions towards the midline of no difference and so
+affects both Type I and Type II error rates. It is also of note that in
+the case of true biological replicates (different individuals) that
+adding a modest amount of scale \(\gamma=0.5\) had little effect on the
+concordance between the difference between groups and the p-values as
+shown in Supplementary Figure 7. Thus, in this dataset the scale
+mis-specification was affecting mainly the location of the difference
+between groups.
+
+In the metatranscriptome analysis, transcripts that were previously not
+classed as differentially abundant were now called as significantly
+different, and the housekeeping transcripts move from being
+significantly different to not being identified as such. Note the
+assumption that housekeeping genes should not generally be included in
+differential abundance analysis is implicit in the dual
+p-value:fold-change cutoff approach in widespread use. While we
+acknowledge that some prior information on which housekeeping
+transcripts should not be classed as differentially abundant is needed,
+we suggest that this information is widely available and is already used
+when performing the gold-standard quantitative PCR test of differential
+abundance (50, 51). Thus, the use of this prior knowledge is not unique
+to our approach.
+
+Beyond concerns of fidelity and rigor, scale models also enhance the
+reproducibility and transparency of HTS analyses. An advantage of
+incorporating scale is that analyses can be made much more robust such
+that actual or potential differences in scale can be tested and
+accounted for explicitly. This occurs because rather than using an
+implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the
+CLR normalization), scale models can make this an explicit part of the
+model building process. While it is beyond the scope of the present
+article, we note that there are many ways of building scale models that
+enhance the interpretability of the parameters and assumptions and a
+detailed description of these points is describe elsewhere (8). In this
+work, we simply use the parameter \(\gamma\) in the default scale model
+as a term that represents uncertainty in the CLR assumption and note
+that larger values of \(\gamma\) correspond to more uncertainty in this
+assumption. In the metatranscriptome example and elsewhere (9), we also
+show that the ratio of scales between conditions can be used to build a
+full scale model when the conditions have dramatically different
+underlying scales.
 
@@
 uncertainty. Originally, this tool used the only the Dirichlet
-distribution to sample compositional uncertainty. Rather than applying
-the CLR normalization as in the original ALDEx2, scale uncertainty can
-be added through the use of a scale model with no additional
-computational complexity. Scale uncertainty can be sampled from any
-distribution depending on prior knowledge or preference. By default,
-ALDEx2 samples scale from a logNormal distribution inspired by the CLR
-normalization. However, there is the option to introduce a full scale
-model that encapsulates both uncertainty and asymmetry in underlying
-scale.
-
-All normalizations attempt to make the samples in a dataset commensurate
-but do not explicitly address the scale of the underlying system. The
-two supplementary tables show that these normalizations can be
-interpreted as scale assumptions. However, the general lack of scale
-information has important consequences for the analysis of HTS datasets.
-One issue is that analysis tools seem over-powered with even moderate
-sample sizes, and different tools have different intrinsic power, Type 1
-and Type 2 error rates (Schurch et al.~2016) and fail to account for
-false discovery rates properly (Hawinkel et al. 2018; Li et al. 2022).
-Traditional recommendations would suggest that using small sample sizes
-in analysis leads to less reliability and reproducibility in analyses
-since surprisingly large sample sizes are needed to determine
-reproducible p-values (e.g., Halsey et al. (2015)). However,
-recommendations to use small sample sizes are widespread in multivariate
-datasets such as RNA-seq datasets. We argue that this is a hallmark of
-unacknowledged bias: since normalizations are typically assumed to be
-correct with no uncertainty, smaller samples offer an unintentional
-buffer against this misspecification. Another issue is that datasets are
-difficult to analyze when they contain systematic asymmetry, with
-different tools exhibiting differing pathologies with these datasets
-(Robinson and Oshlack 2010; Wu et al. 2021).
-
-Many groups have conducted benchmarking studies on different tools and
-normalizations used for the analysis of datasets such as transcriptomes
-(Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
-Quinn, Crowley, and Richardson 2018) and microbiomes (Thorsen et al.
-2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
-observed that the actual agreement between methods can be modest, and
-there is usually a recommendation that some methods are better suited
-for individual datasets or for specific types of analysis;
-e.g.,\textasciitilde`Normalization and microbial differential abundance
-strategies depend upon data characteristics' (Weiss et al. 2017).{]}
-However, another interpretation of such studies is that different
-normalizations can be better suited to different datasets by
-happenstance, but it is not obvious how the use of specific
-normalizations could be justified on a per dataset basis. There are now
-published examples of multiple large replication efforts in psychology
-(Open Science Collaboration 2015), cancer biology (Rodgers and Collings
-2021), ecology and evolution (Gould and et.al 2023). In a nutshell,
-experiences from these studies indicate that published data typically
-have higher effect sizes and appear more significant than replication
-efforts with the same data, or with newly generated replication data
-(Rodgers and Collings 2021). Thus, most published data should be viewed
-with suspicion (Ioannidis 2005) unless efforts have been made to ensure
-that the results are robust.
-
-In the case of overpowering, HTS analyses seem to be more robust when
-applying a dual cutoff of both p-value and difference between group
-means (Schurch et al. 2016). Figure 2 shows one reason for this
-robustness could be that the dual cutoff is mimicking the effect of
-including scale uncertainty, since substantially similar transcripts are
-identified by the two approaches. We believe that adding scale
-uncertainty leads to a more transparent analysis compared to the cutoff
-as it is easy to describe the conditions under which certain conclusions
-hold. Furthermore, while using the post-hoc difference cutoff is useful
-for differential abundance analysis, it is not clear how this can be
-incorporated into other kinds of downstream analyses. Conversely,
-analyses that include scale uncertainty are fully compatible with
-existing downstream analyses.
-
-In the case of asymmetry, the use of a user-specified scale model can be
-very useful for otherwise difficult-to-analyze datasets such as
-meta-transcriptomes and in-vitro selection datasets where the majority
-of features can change. We showed one such example in Figure 3 where the
-dataset was highly asymmetrical, and the TMM and RLE normalizations
-cannot fully move all the housekeeping genes to the midline of no
-difference or exhibit other pathologies (Supplemental). Incorporating
-differential scale on a per-group basis moves the mass of the data
-towards the midline of no difference and so affects both Type I and Type
-II error rates. In this analysis, transcripts that were previously not
-classed as differentially abundant are now called as significantly
-different, and the housekeeping transcripts move from being
-significantly different to not being identified as such. While we
-acknowledge that some prior information on which housekeeping
-transcripts should not be classed as DA is needed, we suggest that this
-information is widely available and is already used when performing the
-gold-standard quantitative PCR test of differential abundance (Thellin
-et al. 1999; SEQC/MAQC-III Consortium 2014). Furthermore, the assumption
-that housekeeping genes should not generally be included in differential
-abundance analysis is implicit in the dual p-value, fold-change cutoff
-approach in widespread use. Thus, the use of this prior knowledge is not
-unique to our approach.
-
-In summary, while the underlying scale of the system is generally
-inaccessible, the effect of scale on the analysis outcomes can be
-modelled and can help explain some of the underlying biology. Adding
-scale information to the analysis allows for more robust inference
-because the features that are sensitive to scale can be identified and
-their impact on the analysis weighted accordingly. Additionally, the use
-of user-defined scale models permits difficult to analyze datasets to be
-examined in a robust and principled manner even when the majority of
-features are asymmetrically distributed or expressed (or both) in the
-groups. Thus, reporting scale uncertainty should become a standard
-practice in the analysis of HTS datasets as a way to identify which
-features are most robust to differences in the underlying system.
-Finally, we supply a toolkit that makes incorporating scale simple even
-for datasets that come from highly asymmetrical environments.
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
+distribution to sample compositional uncertainty, but scale uncertainty
+can be added through the use of a scale model with no additional
+computational complexity. By default, ALDEx2 samples scale from a
+logNormal distribution inspired by the CLR normalization. However, scale
+uncertainty can be sampled from any distribution depending on prior
+knowledge or preference, using the option to introduce a full informed
+scale model that encapsulates both uncertainty and asymmetry in
+underlying scale.
+
+In summary, we supply a toolkit that makes incorporating scale simple to
+incorporate for any type of HTS dataset. While the underlying scale of
+the system is generally inaccessible, the effect of scale on the
+analysis outcomes can be modelled and can help explain some of the
+underlying biology, and known issues with the analysis of HTS data.
+Adding scale information to the analysis allows for more robust
+inference because the features that are sensitive to scale can be
+identified and their impact on conclusions weighted accordingly.
+Additionally, the use of user-defined scale models permits difficult to
+analyze datasets to be examined in a robust and principled manner even
+when the majority of features are asymmetrically distributed or
+expressed (or both) in the groups (42). Thus, reporting scale
+uncertainty should become a standard practice in the analysis of HTS
+datasets as a way to identify which features are most robust to
+differences in the underlying system.
+
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-aitchison1982}{}}%
-Aitchison, John. 1982. {``The Statistical Analysis of Compositional
-Data.''} \emph{Journal of the Royal Statistical Society: Series B
-(Methodological)} 44 (2): 139--60.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Anders:2010}{}}%
-Anders, Simon, and Wolfgang Huber. 2010. {``Differential Expression
-Analysis for Sequence Count Data.''} \emph{Genome Biol} 11 (10): R106.
-\url{https://doi.org/10.1186/gb-2010-11-10-r106}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Andrews:2019aa}{}}%
-Andrews, Tallulah S, and Martin Hemberg. 2019. {``M3Drop: Dropout-Based
-Feature Selection for scRNASeq.''} \emph{Bioinformatics} 35 (16):
-2865--67. \url{https://doi.org/10.1093/bioinformatics/bty1044}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-benjamini:1995}{}}%
-Benjamini, Yoav, and Yosef Hochberg. 1995. {``Controlling the False
-Discovery Rate: A Practical and Powerful Approach to Multiple
-Testing.''} \emph{Journal of the Royal Statistical Society. Series B
-(Methodological)} 57 (1): 289--300.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bullard:2010}{}}%
-Bullard, James H, Elizabeth Purdom, Kasper D Hansen, and Sandrine
-Dudoit. 2010. {``Evaluation of Statistical Methods for Normalization and
-Differential Expression in m{RNA-Seq} Experiments.''} \emph{BMC
-Bioinformatics} 11: 94. \url{https://doi.org/10.1186/1471-2105-11-94}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Cui:2003aa}{}}%
-Cui, Xiangqin, and Gary A Churchill. 2003.
-{``\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical Tests
-for Differential Expression in cDNA Microarray Experiments}.''}
-\emph{Genome Biol} 4 (4): 210.1--10.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Denge00262-18}{}}%
-Deng, Zhi-Luo, Cornelia Gottschick, Sabin Bhuju, Clarissa Masur,
-Christoph Abels, and Irene Wagner-Döbler. 2018. {``Metatranscriptome
-Analysis of the Vaginal Microbiota Reveals Potential Mechanisms for
-Protection Against Metronidazole in Bacterial Vaginosis.''} Edited by
-Craig D. Ellermeier, Janet Hill, and Andrew Onderdonk. \emph{mSphere} 3
-(3). \url{https://doi.org/10.1128/mSphereDirect.00262-18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Dillies:2013}{}}%
-Dillies, Marie-Agnès, Andrea Rau, Julie Aubert, Christelle
-Hennequet-Antier, Marine Jeanmougin, Nicolas Servant, Céline Keime, et
-al. 2013. {``A Comprehensive Evaluation of Normalization Methods for
-{Illumina} High-Throughput {RNA} Sequencing Data Analysis.''}
-\emph{Brief Bioinform} 14 (6): 671--83.
-\url{https://doi.org/10.1093/bib/bbs046}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fernandes:2013}{}}%
-Fernandes, Andrew D, Jean M Macklaim, Thomas G Linn, Gregor Reid, and
-Gregory B Gloor. 2013. {``ANOVA-Like Differential Expression (ALDEx)
-Analysis for Mixed Population RNA-Seq.''} \emph{PLoS One} 8 (7): e67019.
-\url{https://doi.org/10.1371/journal.pone.0067019}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fernandes:2014}{}}%
-Fernandes, Andrew D, Jennifer Ns Reid, Jean M Macklaim, Thomas A
-McMurrough, David R Edgell, and Gregory B Gloor. 2014. {``Unifying the
-Analysis of High-Throughput Sequencing Datasets: Characterizing
-{RNA}-Seq, 16{S} r{RNA} Gene Sequencing and Selective Growth Experiments
-by Compositional Data Analysis.''} \emph{Microbiome} 2: 15.1--13.
-\url{https://doi.org/10.1186/2049-2618-2-15}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fettweis:2019aa}{}}%
-Fettweis, Jennifer M, Myrna G Serrano, J Paul Brooks, David J Edwards,
-Philippe H Girerd, Hardik I Parikh, Bernice Huang, et al. 2019. {``The
-Vaginal Microbiome and Preterm Birth.''} \emph{Nat Med} 25 (6):
-1012--21. \url{https://doi.org/10.1038/s41591-019-0450-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Gierlinski:2015aa}{}}%
-Gierliński, Marek, Christian Cole, Pietà Schofield, Nicholas J Schurch,
-Alexander Sherstnev, Vijender Singh, Nicola Wrobel, et al. 2015.
-{``Statistical Models for RNA-Seq Data Derived from a Two-Condition
-48-Replicate Experiment.''} \emph{Bioinformatics} 31 (22): 3625--30.
-\url{https://doi.org/10.1093/bioinformatics/btv425}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gloor:effect}{}}%
-Gloor, GB, JM Macklaim, and AD Fernandes. 2016. {``Displaying Variation
-in Large Datasets: Plotting a Visual Summary of Effect Sizes.''}
-\emph{Journal of Computational and Graphical Statistics} 25 (3C):
-971--79. \url{https://doi.org/10.1080/10618600.2015.1131161}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-evoRepr-pre}{}}%
-Gould, E, and et.al. 2023. {``Same Data, Different Analysts: Variation
-in Effect Sizes Due to Analytical Decisions in Ecology and Evolutionary
-Biology.''} \emph{EcoEvoRxiv}.
-https://doi.org/\url{https://doi.org/10.32942/X2GG62}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Halsey:2015aa}{}}%
-Halsey, Lewis G, Douglas Curran-Everett, Sarah L Vowler, and Gordon B
-Drummond. 2015. {``The Fickle p Value Generates Irreproducible
-Results.''} \emph{Nat Methods} 12 (3): 179--85.
-\url{https://doi.org/10.1038/nmeth.3288}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hawinkel2017}{}}%
-Hawinkel, Stijn, Federico Mattiello, Luc Bijnens, and Olivier Thas.
-2018. {``A Broken Promise : Microbiome Differential Abundance Methods Do
-Not Control the False Discovery Rate.''} \emph{BRIEFINGS IN
-BIOINFORMATICS}. \url{http://dx.doi.org/10.1093/bib/bbx104}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hughes:2005tu}{}}%
-Hughes, Jennifer B, and Jessica J Hellmann. 2005. {``The Application of
-Rarefaction Techniques to Molecular Inventories of Microbial
-Diversity.''} \emph{Methods Enzymol} 397: 292--308.
-\url{https://doi.org/10.1016/S0076-6879(05)97017-1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hummelen:2010}{}}%
-Hummelen, Ruben, Andrew D Fernandes, Jean M Macklaim, Russell J Dickson,
-John Changalucha, Gregory B Gloor, and Gregor Reid. 2010. {``Deep
-Sequencing of the Vaginal Microbiota of Women with {HIV}.''} \emph{PLoS
-One} 5 (8): e12078. \url{https://doi.org/10.1371/journal.pone.0012078}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ioannidis:2005aa}{}}%
-Ioannidis, John P A. 2005. {``Why Most Published Research Findings Are
-False.''} \emph{PLoS Med} 2 (8): e124.
-\url{https://doi.org/10.1371/journal.pmed.0020124}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Li:2022aa}{}}%
-Li, Yumei, Xinzhou Ge, Fanglue Peng, Wei Li, and Jingyi Jessica Li.
-2022. {``Exaggerated False Positives by Popular Differential Expression
-Methods When Analyzing Human Population Samples.''} \emph{Genome Biol}
-23 (1): 79. \url{https://doi.org/10.1186/s13059-022-02648-4}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Love:2014aa}{}}%
-Love, Michael I, Wolfgang Huber, and Simon Anders. 2014. {``Moderated
-Estimation of Fold Change and Dispersion for RNA-Seq Data with
-DESeq2.''} \emph{Genome Biol} 15 (12): 550.1--21.
-\url{https://doi.org/10.1186/s13059-014-0550-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lovell:2015}{}}%
-Lovell, David, Vera Pawlowsky-Glahn, Juan José Egozcue, Samuel
-Marguerat, and Jürg Bähler. 2015. {``Proportionality: A Valid
-Alternative to Correlation for Relative Data.''} \emph{PLoS Comput Biol}
-11 (3): e1004075.
-https://doi.org/\url{https://doi.org/10.1371/journal.pcbi.1004075}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Loven:2012aa}{}}%
-Lovén, Jakob, David A Orlando, Alla A Sigova, Charles Y Lin, Peter B
-Rahl, Christopher B Burge, David L Levens, Tong Ihn Lee, and Richard A
-Young. 2012. {``Revisiting Global Gene Expression Analysis.''}
-\emph{Cell} 151 (3): 476--82.
-\url{https://doi.org/10.1016/j.cell.2012.10.012}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-macklaim:2013}{}}%
-Macklaim, Jean M, Andrew D Fernandes, Julia M Di Bella, Jo-Anne Hammond,
-Gregor Reid, and Gregory B Gloor. 2013. {``Comparative Meta-{RNA}-Seq of
-the Vaginal Microbiota and Differential Expression by Lactobacillus
-Iners in Health and Dysbiosis.''} \emph{Microbiome} 1 (1): 12.
-\url{https://doi.org/10.1186/2049-2618-1-12}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Macklaim:2018aa}{}}%
-Macklaim, Jean M, and Gregory B Gloor. 2018. {``From {RNA}-Seq to
-Biological Inference: Using Compositional Data Analysis in
-Meta-Transcriptomics.''} \emph{Methods Mol Biol} 1849: 193--213.
-\url{https://doi.org/10.1007/978-1-4939-8728-3/_13}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-maza2013}{}}%
-Maza, Elie, Pierre Frasse, Pavel Senin, Mondher Bouzayen, and Mohamed
-Zouine. 2013. {``Comparison of Normalization Methods for Differential
-Gene Expression Analysis in RNA-Seq Experiments: A Matter of Relative
-Size of Studied Transcriptomes.''} \emph{Commun Integr Biol} 6 (6):
-e25849. \url{https://doi.org/10.4161/cib.25849}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Mortazavi:2008}{}}%
-Mortazavi, Ali, Brian A Williams, Kenneth McCue, Lorian Schaeffer, and
-Barbara Wold. 2008. {``Mapping and Quantifying Mammalian Transcriptomes
-by {RNA-seq}.''} \emph{Nat Methods} 5 (7): 621--28.
-\url{https://doi.org/10.1038/nmeth.1226}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Nie:2012aa}{}}%
-Nie, Zuqin, Gangqing Hu, Gang Wei, Kairong Cui, Arito Yamane, Wolfgang
-Resch, Ruoning Wang, et al. 2012. {``C-{M}yc Is a Universal Amplifier of
-Expressed Genes in Lymphocytes and Embryonic Stem Cells.''} \emph{Cell}
-151 (1): 68--79. \url{https://doi.org/10.1016/j.cell.2012.08.033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nixon2023scale}{}}%
-Nixon, Michelle Pistner, Jeffrey Letourneau, Lawrence A. David, Nicole
-A. Lazar, Sayan Mukherjee, and Justin D. Silverman. 2023. {``Scale
-Reliant Inference.''} \url{https://arxiv.org/abs/2201.03616}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Open-Science-Collaboration:2015aa}{}}%
-Open Science Collaboration. 2015. {``Estimating the Reproducibility of
-Psychological Science.''} \emph{Science} 349 (6251): aac4716.
-\url{https://doi.org/10.1126/science.aac4716}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Props:2017aa}{}}%
-Props, Ruben, Frederiek-Maarten Kerckhof, Peter Rubbens, Jo De Vrieze,
-Emma Hernandez Sanabria, Willem Waegeman, Pieter Monsieurs, Frederik
-Hammes, and Nico Boon. 2017. {``Absolute Quantification of Microbial
-Taxon Abundances.''} \emph{ISME J} 11 (2): 584--87.
-\url{https://doi.org/10.1038/ismej.2016.117}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Quinn:2018aa}{}}%
-Quinn, Thomas P, Tamsyn M Crowley, and Mark F Richardson. 2018.
-{``Benchmarking Differential Expression Analysis Tools for RNA-Seq:
-Normalization-Based Vs. Log-Ratio Transformation-Based Methods.''}
-\emph{BMC Bioinformatics} 19 (1): 274.
-\url{https://doi.org/10.1186/s12859-018-2261-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ravel:2010}{}}%
-Ravel, Jacques, Pawel Gajer, Zaid Abdo, G Maria Schneider, Sara S K
-Koenig, Stacey L McCulle, Shara Karlebach, et al. 2011. {``Vaginal
-Microbiome of Reproductive-Age Women.''} \emph{Proc Natl Acad Sci U S
-A}, no. 108: 4680--87. \url{https://doi.org/doi/10.1073/pnas.100611107}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Robinson:2010}{}}%
-Robinson, Mark D, Davis J McCarthy, and Gordon K Smyth. 2010. {``edgeR:
-A Bioconductor Package for Differential Expression Analysis of Digital
-Gene Expression Data.''} \emph{Bioinformatics} 26 (1): 139--40.
-\url{https://doi.org/10.1093/bioinformatics/btp616}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Robinson:2010a}{}}%
-Robinson, Mark D, and Alicia Oshlack. 2010. {``A Scaling Normalization
-Method for Differential Expression Analysis of {RNA-seq} Data.''}
-\emph{Genome Biol} 11 (3): R25.1--9.
-\url{https://doi.org/10.1186/gb-2010-11-3-r25}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Rodgers:2021aa}{}}%
-Rodgers, Peter, and Andy Collings. 2021. {``What Have We Learned?''}
-\emph{Elife} 10 (December). \url{https://doi.org/10.7554/eLife.75830}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Schurch:2016aa}{}}%
-Schurch, Nicholas J, Pietá Schofield, Marek Gierliński, Christian Cole,
-Alexander Sherstnev, Vijender Singh, Nicola Wrobel, et al. 2016. {``How
-Many Biological Replicates Are Needed in an RNA-Seq Experiment and Which
-Differential Expression Tool Should You Use?''} \emph{RNA} 22 (6):
-839--51. \url{https://doi.org/10.1261/rna.053959.115}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SEQCux2fMAQC-III-Consortium:2014aa}{}}%
-SEQC/MAQC-III Consortium. 2014. {``A Comprehensive Assessment of RNA-Seq
-Accuracy, Reproducibility and Information Content by the Sequencing
-Quality Control Consortium.''} \emph{Nat Biotechnol} 32 (9): 903--14.
-\url{https://doi.org/10.1038/nbt.2957}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Soneson:2013}{}}%
-Soneson, Charlotte, and Mauro Delorenzi. 2013. {``A Comparison of
-Methods for Differential Expression Analysis of {RNA-seq} Data.''}
-\emph{BMC Bioinformatics} 14: 91.
-\url{https://doi.org/10.1186/1471-2105-14-91}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Taniguchi:2010aa}{}}%
-Taniguchi, Yuichi, Paul J Choi, Gene-Wei Li, Huiyi Chen, Mohan Babu,
-Jeremy Hearn, Andrew Emili, and X Sunney Xie. 2010. {``Quantifying e.
-Coli Proteome and Transcriptome with Single-Molecule Sensitivity in
-Single Cells.''} \emph{Science} 329 (5991): 533--38.
-\url{https://doi.org/10.1126/science.1188308}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Thellin:1999aa}{}}%
-Thellin, O, W Zorzi, B Lakaye, B De Borman, B Coumans, G Hennen, T
-Grisar, A Igout, and E Heinen. 1999.
-{``\href{https://www.ncbi.nlm.nih.gov/pubmed/10617337}{Housekeeping
-Genes as Internal Standards: Use and Limits}.''} \emph{J Biotechnol} 75
-(2-3): 291--95.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Thorsen:2016aa}{}}%
-Thorsen, Jonathan, Asker Brejnrod, Martin Mortensen, Morten A Rasmussen,
-Jakob Stokholm, Waleed Abu Al-Soud, Søren Sørensen, Hans Bisgaard, and
-Johannes Waage. 2016. {``Large-Scale Benchmarking Reveals False
-Discoveries and Count Transformation Sensitivity in 16{S} r{RNA} Gene
-Amplicon Data Analysis Methods Used in Microbiome Studies.''}
-\emph{Microbiome} 4 (1): 62.
-\url{https://doi.org/10.1186/s40168-016-0208-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Vandeputte:2017aa}{}}%
-Vandeputte, Doris, Gunter Kathagen, Kevin D'hoe, Sara Vieira-Silva,
-Mireia Valles-Colomer, João Sabino, Jun Wang, et al. 2017.
-{``Quantitative Microbiome Profiling Links Gut Community Variation to
-Microbial Load.''} \emph{Nature} 551 (7681): 507--11.
-\url{https://doi.org/10.1038/nature24460}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wagner:tpm}{}}%
-Wagner, Günter P, Koryu Kin, and Vincent J Lynch. 2012. {``Measurement
-of mRNA Abundance Using RNA-Seq Data: RPKM Measure Is Inconsistent Among
-Samples.''} \emph{Theory Biosci} 131 (4): 281--85.
-\url{https://doi.org/10.1007/s12064-012-0162-3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Weiss:2017aa}{}}%
-Weiss, Sophie, Zhenjiang Zech Xu, Shyamal Peddada, Amnon Amir, Kyle
-Bittinger, Antonio Gonzalez, Catherine Lozupone, et al. 2017.
-{``Normalization and Microbial Differential Abundance Strategies Depend
-Upon Data Characteristics.''} \emph{Microbiome} 5 (1): 27.
-\url{https://doi.org/10.1186/s40168-017-0237-y}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Wu2021}{}}%
-Wu, Jia R., Jean M. Macklaim, Briana L. Genge, and Gregory B. Gloor.
-2021. {``Finding the Centre: Compositional Asymmetry in High-Throughput
-Sequencing Datasets.''} In \emph{Advances in Compositional Data
-Analysis: Festschrift in Honour of Vera Pawlowsky-Glahn}, edited by
-Peter Filzmoser, Karel Hron, Josep Antoni Martìn-Fernàndez, and Javier
-Palarea-Albaladejo, 329--46. Cham: Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-71175-7_17}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Yoshikawa:2011aa}{}}%
-Yoshikawa, Katsunori, Tadamasa Tanaka, Yoshihiro Ida, Chikara Furusawa,
-Takashi Hirasawa, and Hiroshi Shimizu. 2011. {``Comprehensive Phenotypic
-Analysis of Single-Gene Deletion and Overexpression Strains of
-Saccharomyces Cerevisiae.''} \emph{Yeast} 28 (5): 349--61.
-\url{https://doi.org/10.1002/yea.1843}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Zozaya:2010}{}}%
-Zozaya-Hinchliffe, Marcela, Rebecca Lillis, David H Martin, and Michael
-J Ferris. 2010. {``Quantitative PCR Assessments of Bacterial Species in
-Women with and Without Bacterial Vaginosis.''} \emph{J Clin Microbiol}
-48 (5): 1812--19. \url{https://doi.org/10.1128/JCM.00851-09}.
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{1}
+\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
+1. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
+Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
+Darzi,Y., \emph{et al.} (2017)
+\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
+profiling links gut community variation to microbial load}.
+\emph{Nature}, \textbf{551}, 507--511.
+
+\bibitem[\citeproctext]{ref-Props:2017aa}
+2. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
+Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
+\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
+microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.
+
+\bibitem[\citeproctext]{ref-Thorsen:2016aa}
+3. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
+Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
+\href{https://doi.org/10.1186/s40168-016-0208-8}{Large-scale
+benchmarking reveals false discoveries and count transformation
+sensitivity in 16{S} r{RNA} gene amplicon data analysis methods used in
+microbiome studies}. \emph{Microbiome}, \textbf{4}, 62.
+
+\bibitem[\citeproctext]{ref-Quinn:2018aa}
+4. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
+\href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
+differential expression analysis tools for RNA-seq: Normalization-based
+vs. Log-ratio transformation-based methods}. \emph{BMC Bioinformatics},
+\textbf{19}, 274.
+
+\bibitem[\citeproctext]{ref-Zhang:2009aa}
+5. Zhang,S. and Cao,J. (2009)
+\href{https://doi.org/10.1186/1471-2105-10-402}{A close examination of
+double filtering with fold change and t test in microarray analysis}.
+\emph{BMC Bioinformatics}, \textbf{10}, 402.
+
+\bibitem[\citeproctext]{ref-Ebrahimpoor:2021aa}
+6. Ebrahimpoor,M. and Goeman,J.J. (2021)
+\href{https://doi.org/10.1093/bib/bbab053}{Inflated false discovery rate
+due to volcano plots: Problem and solutions}. \emph{Brief Bioinform},
+\textbf{22}.
+
+\bibitem[\citeproctext]{ref-Wu2021}
+7. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
+\href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
+Compositional asymmetry in high-throughput sequencing datasets}. In
+Filzmoser,P., Hron,K., Martìn-Fernàndez,J.A., Palarea-Albaladejo,J.
+(eds), \emph{Advances in compositional data analysis: Festschrift in
+honour of vera pawlowsky-glahn}. Springer International Publishing,
+Cham, pp. 329--346.
+
+\bibitem[\citeproctext]{ref-nixon2023scale}
+8. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
+Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
+reliant inference}.
+
+\bibitem[\citeproctext]{ref-Nixon2024B}
+9. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
+normalization: Incorporating scale uncertainty in microbiome and gene
+expression analysis. \emph{bioRxiv},
+\href{https://doi.org/10.1101/2024.04.01.587602}{10.1101/2024.04.01.587602}.
+
+\bibitem[\citeproctext]{ref-Lovell:2015}
+10. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
+Bähler,J. (2015)
+\href{https://doi.org/10.1371/journal.pcbi.1004075}{Proportionality: A
+valid alternative to correlation for relative data}. \emph{PLoS Comput
+Biol}, \textbf{11}, e1004075.
+
+\bibitem[\citeproctext]{ref-Nie:2012aa}
+11. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
+Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
+\href{https://doi.org/10.1016/j.cell.2012.08.033}{C-{M}yc is a universal
+amplifier of expressed genes in lymphocytes and embryonic stem cells}.
+\emph{Cell}, \textbf{151}, 68--79.
+
+\bibitem[\citeproctext]{ref-Loven:2012aa}
+12. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
+Burge,C.B., Levens,D.L., Lee,T.I. and Young,R.A. (2012)
+\href{https://doi.org/10.1016/j.cell.2012.10.012}{Revisiting global gene
+expression analysis}. \emph{Cell}, \textbf{151}, 476--82.
+
+\bibitem[\citeproctext]{ref-Scott:2010}
+13. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
+\href{https://doi.org/10.1126/science.1192588}{Interdependence of cell
+growth and gene expression: Origins and consequences}. \emph{Science},
+\textbf{330}, 1099--102.
+
+\bibitem[\citeproctext]{ref-Yoshikawa:2011aa}
+14. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
+Shimizu,H. (2011) \href{https://doi.org/10.1002/yea.1843}{Comprehensive
+phenotypic analysis of single-gene deletion and overexpression strains
+of saccharomyces cerevisiae}. \emph{Yeast}, \textbf{28}, 349--61.
+
+\bibitem[\citeproctext]{ref-Lin:2018aa}
+15. Lin,J. and Amir,A. (2018)
+\href{https://doi.org/10.1038/s41467-018-06714-z}{Homeostasis of protein
+and mRNA concentrations in growing cells}. \emph{Nat Commun},
+\textbf{9}, 4496.
+
+\bibitem[\citeproctext]{ref-Zozaya:2010}
+16. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
+\href{https://doi.org/10.1128/JCM.00851-09}{Quantitative PCR assessments
+of bacterial species in women with and without bacterial vaginosis}.
+\emph{J Clin Microbiol}, \textbf{48}, 1812--9.
+
+\bibitem[\citeproctext]{ref-Ravel:2010}
+17. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
+McCulle,S.L., Karlebach,S., Gorle,R., Russell,J., Tacket,C.O., \emph{et
+al.} (2011) Vaginal microbiome of reproductive-age women. \emph{Proc
+Natl Acad Sci U S A},
+\href{https://doi.org/doi/10.1073/pnas.100611107}{doi/10.1073/pnas.100611107}.
+
+\bibitem[\citeproctext]{ref-Hummelen:2010}
+18. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
+Changalucha,J., Gloor,G.B. and Reid,G. (2010)
+\href{https://doi.org/10.1371/journal.pone.0012078}{Deep sequencing of
+the vaginal microbiota of women with {HIV}}. \emph{PLoS One},
+\textbf{5}, e12078.
+
+\bibitem[\citeproctext]{ref-Bullard:2010}
+19. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
+\href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
+methods for normalization and differential expression in m{RNA-seq}
+experiments}. \emph{BMC Bioinformatics}, \textbf{11}, 94.
+
+\bibitem[\citeproctext]{ref-Dillies:2013}
+20. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
+Jeanmougin,M., Servant,N., Keime,C., Marot,G., Castel,D., Estelle,J.,
+\emph{et al.} (2013) \href{https://doi.org/10.1093/bib/bbs046}{A
+comprehensive evaluation of normalization methods for {Illumina}
+high-throughput {RNA} sequencing data analysis}. \emph{Brief Bioinform},
+\textbf{14}, 671--83.
+
+\bibitem[\citeproctext]{ref-Weiss:2017aa}
+21. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
+Lozupone,C., Zaneveld,J.R., Vázquez-Baeza,Y., Birmingham,A., \emph{et
+al.} (2017)
+\href{https://doi.org/10.1186/s40168-017-0237-y}{Normalization and
+microbial differential abundance strategies depend upon data
+characteristics}. \emph{Microbiome}, \textbf{5}, 27.
+
+\bibitem[\citeproctext]{ref-Hughes:2005tu}
+22. Hughes,J.B. and Hellmann,J.J. (2005)
+\href{https://doi.org/10.1016/S0076-6879(05)97017-1}{The application of
+rarefaction techniques to molecular inventories of microbial diversity}.
+\emph{Methods Enzymol}, \textbf{397}, 292--308.
+
+\bibitem[\citeproctext]{ref-Mortazavi:2008}
+23. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
+(2008) \href{https://doi.org/10.1038/nmeth.1226}{Mapping and quantifying
+mammalian transcriptomes by {RNA-seq}}. \emph{Nat Methods}, \textbf{5},
+621--8.
+
+\bibitem[\citeproctext]{ref-wagner:tpm}
+24. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
+\href{https://doi.org/10.1007/s12064-012-0162-3}{Measurement of mRNA
+abundance using RNA-seq data: RPKM measure is inconsistent among
+samples}. \emph{Theory Biosci}, \textbf{131}, 281--5.
+
+\bibitem[\citeproctext]{ref-Robinson:2010a}
+25. Robinson,M.D. and Oshlack,A. (2010)
+\href{https://doi.org/10.1186/gb-2010-11-3-r25}{A scaling normalization
+method for differential expression analysis of {RNA-seq} data}.
+\emph{Genome Biol}, \textbf{11}, R25.1--R25.9.
+
+\bibitem[\citeproctext]{ref-Anders:2010}
+26. Anders,S. and Huber,W. (2010)
+\href{https://doi.org/10.1186/gb-2010-11-10-r106}{Differential
+expression analysis for sequence count data}. \emph{Genome Biol},
+\textbf{11}, R106.
+
+\bibitem[\citeproctext]{ref-aitchison1982}
+27. Aitchison,J. (1982) The statistical analysis of compositional data.
+\emph{Journal of the Royal Statistical Society: Series B
+(Methodological)}, \textbf{44}, 139--160.
+
+\bibitem[\citeproctext]{ref-fernandes:2013}
+28. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
+(2013) \href{https://doi.org/10.1371/journal.pone.0067019}{ANOVA-like
+differential expression (ALDEx) analysis for mixed population RNA-seq}.
+\emph{PLoS One}, \textbf{8}, e67019.
+
+\bibitem[\citeproctext]{ref-Yerke:2024aa}
+29. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
+\href{https://doi.org/10.1186/s40168-023-01747-z}{Proportion-based
+normalizations outperform compositional data transformations in machine
+learning applications}. \emph{Microbiome}, \textbf{12}, 45.
+
+\bibitem[\citeproctext]{ref-hawinkel2017}
+30. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
+\href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
+Microbiome differential abundance methods do not control the false
+discovery rate}. \emph{BRIEFINGS IN BIOINFORMATICS}.
+
+\bibitem[\citeproctext]{ref-Li:2022aa}
+31. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+\href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
+positives by popular differential expression methods when analyzing
+human population samples}. \emph{Genome Biol}, \textbf{23}, 79.
+
+\bibitem[\citeproctext]{ref-Cui:2003aa}
+32. Cui,X. and Churchill,G.A. (2003)
+\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
+for differential expression in cDNA microarray experiments}.
+\emph{Genome Biol}, \textbf{4}, 210.1--210.10.
+
+\bibitem[\citeproctext]{ref-Schurch:2016aa}
+33. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
+Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
+al.} (2016) \href{https://doi.org/10.1261/rna.053959.115}{How many
+biological replicates are needed in an RNA-seq experiment and which
+differential expression tool should you use?} \emph{RNA}, \textbf{22},
+839--51.
+
+\bibitem[\citeproctext]{ref-Gierlinski:2015aa}
+34. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
+Singh,V., Wrobel,N., Gharbi,K., Simpson,G., Owen-Hughes,T., \emph{et
+al.} (2015)
+\href{https://doi.org/10.1093/bioinformatics/btv425}{Statistical models
+for RNA-seq data derived from a two-condition 48-replicate experiment}.
+\emph{Bioinformatics}, \textbf{31}, 3625--3630.
+
+\bibitem[\citeproctext]{ref-Halsey:2015aa}
+35. Halsey,L.G., Curran-Everett,D., Vowler,S.L. and Drummond,G.B. (2015)
+\href{https://doi.org/10.1038/nmeth.3288}{The fickle p value generates
+irreproducible results}. \emph{Nat Methods}, \textbf{12}, 179--85.
+
+\bibitem[\citeproctext]{ref-benjamini:1995}
+36. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
+rate: A practical and powerful approach to multiple testing.
+\emph{Journal of the Royal Statistical Society. Series B
+(Methodological)}, \textbf{57}, 289--300.
+
+\bibitem[\citeproctext]{ref-Soneson:2013}
+37. Soneson,C. and Delorenzi,M. (2013)
+\href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
+for differential expression analysis of {RNA-seq} data}. \emph{BMC
+Bioinformatics}, \textbf{14}, 91.
+
+\bibitem[\citeproctext]{ref-maza2013}
+38. Maza,E., Frasse,P., Senin,P., Bouzayen,M. and Zouine,M. (2013)
+\href{https://doi.org/10.4161/cib.25849}{Comparison of normalization
+methods for differential gene expression analysis in RNA-seq
+experiments: A matter of relative size of studied transcriptomes}.
+\emph{Commun Integr Biol}, \textbf{6}, e25849.
+
+\bibitem[\citeproctext]{ref-gloor:effect}
+39. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
+\href{https://doi.org/10.1080/10618600.2015.1131161}{Displaying
+variation in large datasets: Plotting a visual summary of effect sizes}.
+\emph{Journal of Computational and Graphical Statistics}, \textbf{25},
+971--979.
+
+\bibitem[\citeproctext]{ref-Rocha:2020aa}
+40. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
+(2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
+expression analysis in bacteria by RT-qPCR}. \emph{Methods Mol Biol},
+\textbf{2065}, 119--137.
+
+\bibitem[\citeproctext]{ref-Taniguchi:2010aa}
+41. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
+Emili,A. and Xie,X.S. (2010)
+\href{https://doi.org/10.1126/science.1188308}{Quantifying e. Coli
+proteome and transcriptome with single-molecule sensitivity in single
+cells}. \emph{Science}, \textbf{329}, 533--8.
+
+\bibitem[\citeproctext]{ref-dosSantos:2024}
+42. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
+Gloor,G.B. (2024) Vaginal metatranscriptome meta-analysis reveals
+functional BV subgroups and novel colonisation strategies.
+\emph{bioRxiv},
+\href{https://doi.org/10.1101/2024.04.24.590967}{10.1101/2024.04.24.590967}.
+
+\bibitem[\citeproctext]{ref-macklaim:2013}
+43. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
+and Gloor,G.B. (2013)
+\href{https://doi.org/10.1186/2049-2618-1-12}{Comparative meta-{RNA}-seq
+of the vaginal microbiota and differential expression by lactobacillus
+iners in health and dysbiosis}. \emph{Microbiome}, \textbf{1}, 12.
+
+\bibitem[\citeproctext]{ref-Denge00262-18}
+44. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
+Wagner-Döbler,I. (2018)
+\href{https://doi.org/10.1128/mSphereDirect.00262-18}{Metatranscriptome
+analysis of the vaginal microbiota reveals potential mechanisms for
+protection against metronidazole in bacterial vaginosis}.
+\emph{mSphere}, \textbf{3}.
+
+\bibitem[\citeproctext]{ref-Fettweis:2019aa}
+45. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
+Parikh,H.I., Huang,B., Arodz,T.J., Edupuganti,L., Glascock,A.L.,
+\emph{et al.} (2019)
+\href{https://doi.org/10.1038/s41591-019-0450-2}{The vaginal microbiome
+and preterm birth}. \emph{Nat Med}, \textbf{25}, 1012--1021.
+
+\bibitem[\citeproctext]{ref-Okuda:2008}
+46. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
+Goto,S. and Kanehisa,M. (2008)
+\href{https://doi.org/10.1093/nar/gkn282}{KEGG atlas mapping for global
+analysis of metabolic pathways}. \emph{Nucleic Acids Res}, \textbf{36},
+W423--6.
+
+\bibitem[\citeproctext]{ref-yeast-absolute}
+47. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
+Bähler,J. (2012)
+\href{https://doi.org/10.1016/j.cell.2012.09.019}{Quantitative analysis
+of fission yeast transcriptomes and proteomes in proliferating and
+quiescent cells}. \emph{Cell}, \textbf{151}, 671--83.
+
+\bibitem[\citeproctext]{ref-McMurdie:2014a}
+48. McMurdie,P.J. and Holmes,S. (2014)
+\href{https://doi.org/10.1371/journal.pcbi.1003531}{Waste not, want not:
+Why rarefying microbiome data is inadmissible}. \emph{PLoS Comput Biol},
+\textbf{10}, e1003531.
+
+\bibitem[\citeproctext]{ref-gustafson2015bayesian}
+49. Gustafson,P. (2015) Bayesian inference for partially identified
+models: Exploring the limits of limited data CRC Press.
+
+\bibitem[\citeproctext]{ref-Thellin:1999aa}
+50. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
+Hennen,G., Grisar,T., Igout,A. and Heinen,E. (1999)
+\href{https://www.ncbi.nlm.nih.gov/pubmed/10617337}{Housekeeping genes
+as internal standards: Use and limits}. \emph{J Biotechnol},
+\textbf{75}, 291--5.
+
+\bibitem[\citeproctext]{ref-SEQCux2fMAQC-III-Consortium:2014aa}
+51. SEQC/MAQC-III Consortium (2014)
+\href{https://doi.org/10.1038/nbt.2957}{A comprehensive assessment of
+RNA-seq accuracy, reproducibility and information content by the
+sequencing quality control consortium}. \emph{Nat Biotechnol},
+\textbf{32}, 903--14.
 

### 2024-07-22 go3.Rmd
focussed on dispersion version

@@
     affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
+citeproc: TRUE
+cite-method: citeproc
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
+csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
+pdf-engine: latex
 output:
-  nucleic-acids-research.csl: pdf_document
+  pdf_document 
+header-includes:
+  - \usepackage{setspace}
 package: ALDEx2
-abstract: In many high-throughput sequencing (HTS) studies, sample-to-sample variation  in sequencing depth is  not driven by  variation in the scale (e.g., total size, microbial load, or total gene  expression) of the underlying biological systems being measured but rather is driven by technical factors.
-Typically the technical variation  is addressed using some form of statistical normalization.
-Normalizations are data or parameter transformations that remove unwanted  technical variation in the hopes of facilitating analyses sensitive to scale;  e.g., differential abundance and differential expression analyses.
-Recently we  showed that any normalization makes implicit assumptions about the unmeasured  system scale and that errors in these assumptions can lead to dramatic increases in  false positive and false negative rates.
+abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total gene expression) of the underlying biological systems.
+Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses.
+Recently we showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can lead to dramatic increases in false positive and false negative rates.
 Here we describe updates to the ALDEx2 R package that mitigate these problems by directly modeling uncertainty in the unmeasured system scale through the use of a \textit{scale model}.
-Scale  models generalize the idea of normalizations and can be thought of as  explicitly modeling potential error in the chosen normalization.
+Scale models generalize the idea of normalizations and can be thought of as explicitly modeling the error in normalization by examining a distribution over all possible  normalizations.
 Beyond enhancing the robustness of HTS analyses,  the use of scale models within ALDEx2 enhances the transparency and  reproducibility of analyses by making implicit normalizing assumptions an explicit part of the model building process.
+
 ---
@@
 
-High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing).
-HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to the flow cell.
-Each of these steps can be though of as a sampling step: a pool of DNA that could be measured is sampled and only that subsampled DNA is carried over to subsequent steps.
-Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., size, microbial load, or total gene expression) of the measured biological system is degraded or lost.
-Increasingly, researchers are turning to modified experimental protocols (e.g., DNA spike-ins or cell counting) in an attempt to recover the lost biological variation in scale; see for example [@Vandeputte:2017aa;@Props:2017aa].
-However, these protocols often do not recover meaningful biological variation but instead recover scale variation at some intermediate step in the sample preparation protocol (e.g., at the stage where the DNA spike-in was added).
-In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an outstanding challenge.
+High-throughput sequencing (HTS) is a ubiquitous tool used to explore
+many biological phenomenon such as gene expression (single-cell
+sequencing, RNA-sequencing, meta-transcriptomics), microbial community
+composition (16S rRNA gene sequencing, shotgun metagenomics) and
+differential enzyme activity (selex, CRISPR killing). HTS proceeds by
+taking a sample from the environment, making a library, multiplexing
+(merging) multiple libraries together, and then applying a sample of the
+multiplexed library to the flow cell. Each of these steps is a 
+compositional sampling step as only a fixed-size subsample of nucleic acid is 
+carried over to subsequent steps. Thus, with
+each sampling step the connection between the size of the sampled DNA
+pool and the scale (e.g., size, microbial load, or total gene
+expression) of the measured biological system is degraded or lost.
+Increasingly, researchers are turning to modified experimental protocols
+(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
+biological variation in scale; see for example  [@Vandeputte:2017aa;@Props:2017aa]. 
+However, these protocols often do not recover
+meaningful biological variation but instead recover scale variation at
+the intermediate step in the sample preparation protocol where the DNA spike-in was added. 
+In short, the disconnect
+between sample-to-sample variation in sequencing depth and biological
+variation in scale remains an outstanding challenge.
 
-Biological variation in scale can represent an important unmeasured confounder in HTS analyses [@Lovell:2015].
+The analysis of  HTS data suffers from several known problems that can be
+traced, in whole or in part, to  misspecification of scale. The first issue 
+is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa], 
+  exhibited as dataset-dependent 
+FDR control.  The FDR problem is connected to the double-filtering approach
+that is often used, but which is known not to have appropriate FDR control 
+[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
+analyzing asymmetric data; data that arises because of a bias in the direction of change.
+This type of data frequently arises in in-vitro selection experiments (SELEX), 
+transcriptome analysis, and microbiome analysis [@Wu2021]. The third issue is that these
+problems become more pronounced as more samples are collected; that is,
+more information results in a worsening of the accuracy of the analysis [@nixon2023scale,@Nixon2024B]. 
+The final problem is that these data are relative data, i.e. compositional, 
+and substantial effort has been put into removing this constraint.
+
+The first three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.
+Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Lovell:2015].
 For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa].
-In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa].
+In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa].
 As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010].
-Thus, a full description of these systems includes both relative change (composition) and absolute abundance (scale), yet current methods access only the compositional information.
+Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).
+Current methods access only the compositional information yet make implicit assumptions about the scale.
 
-To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
-Let \(\mathbf{W}\) also represent a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true (as opposed to measured) amount of class \(d\) in the biological system from which sample \(n\) was obtained.
-Succinctly, \(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\) denotes the true counts in the system under study.
+Recently, Nixon et al.
+[-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models.
+They showed that \emph{all} normalizations make some assumption about  scale.
+These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa].
+Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g.
+proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g.
+the TMM,[@Robinson:2010a], or LVHA [@Wu2021]; or that different sub-parts of each sample maintain a constant scale across samples e.g.
+the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@aitchison1982] and its derivatives.
+
+The original ALDEx2 [@fernandes:2013] model made a strict assumption about scale through the CLR normalization [@nixon2023scale].
+While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa].
+Nixon et al.
+[-@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results.
+In essence ALDEx2 has been modified to explicitly model the scale over a range of reasonable normalization parameters.
+Here, we briefly introduce these modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.
+
+# Implementation
+
+To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
+Likewise, we can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained.
 We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions).
@@
 Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
+
+All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation and the \(\hat{{\mathbf{W}}}\) notation indicates that the normalization is attempting to provide an estimate of the true value.
 The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al.
 2011).
-
-Early on, researchers recognized the potential confounding effects of the non-biological variation in sequencing depth and approached the problem using various forms of \textit{normalization}.
-The idea of normalization is to remove this variation thus making the samples commensurate and to correct any minor asymmetries in the data.
-A large number of normalizations were developed.
-Rarefaction [@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million) [@Mortazavi:2008; @wagner:tpm], the relative log expression (RLE) [@Anders:2010] and trimmed mean of M values (TMM) [@Robinson:2010a].
-The Centered Log Ratio (CLR) [@aitchison1982]  was proposed as a unifying normalization [@fernandes:2014].
-With the exception of rarefaction, all of these normalizations are data transformations that can be stated as ratios of the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the major differences being how the denominator is chosen and whether the ratio is subsequently always log-transformed or not.
-
-Recently, Nixon et al.
-[-@nixon2023scale] suggested that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models.
-Through this lens, they showed that \emph{all} normalizations make some assumption about the system scale \(\mathbf{W}^{\perp}\) in order to identify the chosen model.
-Moreover, these assumptions are often implicit, difficult to interpret, and provided different outputs when applied to the same dataset [@Dillies:2013;@Weiss:2017aa].
-Intuitively, normalizations in widespread use assume that either one sample can be chosen as a reference to which the others are scaled (TMM), or that different sub-parts of each sample maintain a constant scale across samples (RLE, CLR and derivatives).
-Further, Nixon et al.
-[-@nixon2023scale] showed  that the original ALDEx2 model made a strict assumption about scale through the CLR normalization.
-They showed that this assumption could be remedied since ALDEx2 permits scale to be easily incorporated into its framework.
-In this paper, we discuss these modifications and show how including scale uncertainty can greatly improve modeling and lead to more robust and reproducible results.
 
@@
 The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data.
-For brevity and concreteness, we discuss ALDEx2 in its simplest form as a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models.
-At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the \(\mathbf{W}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses.
-For more details on ALDEx2 see Fernandes et al.
-[-@fernandes:2013].
+For brevity, we discuss ALDEx2 in its simplest form as a tool for estimating the magnitude and statistical significance of Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models.
+At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{\hat{W}}\) given the estimate \(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses.
+For more details on ALDEx2 see [@fernandes:2013;@nixon2023scale].
 
-By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where \(G_{n}\) denotes the geometric mean of the vector \((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental Materials for how the CLR connects to Shannon entropy).
-Nixon et al.
-(2023) showed that in this context the CLR normalization equates to an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\).
-They showed that even slight errors in this assumption introduced bias into LFC estimate that was not accounted for in estimates of uncertainty (e.g., confidence intervals or credible sets).
-In fact, they showed that the only way in which the ALDEx2 model, or any model that uses  normalizations, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.
-While this assumption may be approximately correct in some instances it is surely never exactly true.
-In support of this assertion it is known that false discovery rates vary widely by analysis and normalization method [@hawinkel2017;@Li:2022aa].
-As a result, Nixon et al.
-(2023) suggested that models should consider potential error in the assumptions implied by normalizations.
+By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\) where \(G_{n}\) denotes the geometric mean of the vector \((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\).
+The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale,Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets).
+In fact, Nixon et al.
+[-@nixon2023scale] showed that the only way in which the ALDEx2 model, or any normalization-based model, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.
+In support of this assertion it is known that false discovery rates vary widely by analysis, dataset, and normalization method [@hawinkel2017;@Li:2022aa,Nixon2024B].
+To fix this, Nixon et al.
+[-@nixon2023scale] showed that models should incorporate potential error in the assumptions implied by normalizations.
 
 Nixon et al.
-(2023) proposed the concept of a \textit{scale model} as a generalization of normalizations that account for potential error.
-They showed how scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV).
-Recall that \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
-As the HTS data only informs directly on $\mathbf{W}^{\parallel}_{dn}$, they proposed to augment analyses with a model for $\mathbf{W}^{\perp}_{n}$ (the scale model).
-Since the CLR normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean \(1/G_{n}\).
+(2023) generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error.
+Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV).
+They did this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\).
+Since the CLR normalization makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\).
 For example, the following scale model generalizes the CLR:
 
-\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
+\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
 
-where \(\gamma\) is a tunable parameter that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control).
-Beyond theoretical studies, they demonstrated through the analysis of simulated and real data the remarkable performance improvements that could result from incorporating this type of scale uncertainty into analyses.
-Subsequently, we have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.
-
-Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses.
-This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models make this an explicit part of the model building process.
-While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions.
-Moreover, the scale model listed above, which is now the default in ALDEx2, also has an appealing interpretation.
-A detailed description of these points as well as the mathematical underpinnings of the updated ALDEx2 R package is forthcoming and will be made available as a companion publication.
-Here, we simply treat the parameter \(\gamma\) in the default scale model as an abstract term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption.
-We also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.
-
-An advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
-The examples below and in the supplement show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.
+where \(\gamma\) is a tunable parameter drawn from a Gaussian distribution that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control).
+We have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.
 
@@
 
-Gierliński et al.
-[-@Gierlinski:2015aa] conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf1 gene knockout.
-This dataset has been used to argue that only a small number of replicates are needed to identity differentially abundant genes and makes the recommendation that some tools should be used in some conditions because each tool has different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]].
-This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout.
-Through the lens of this paper, this is a hallmark of unacknowledged bias: a false certainty in the scale of the data brings about an increasing false precision with increasing sample sizes.
-Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
-Thus, by adding scale uncertainty, we can be more confident that the results of the analysis reflect true biological differences and not differences due to the assumptions made by the normalization method.
+It is standard practice in many fields of HTS, but especially transcriptomics, to use the dual cutoff approach graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa] because not all statistically significant differences are biologically relevant.
+Paraphrasing this we can say that in in some types of datasets more samples leads to a majority of features being statistically significant.
+A detailed reasoning for this using modelled data is explained in two reports by Nixon et al.
+[-@nixon2023scale;-@Nixon2024B], which shows that scaled models completely address this analytic issue.
 
-We start by examining the dispersion (variance) of the data as counts and as the logarithm of the normalized counts when calculated by DESeq2 (log(RLE)) or by ALDEx2 (CLR).
-The raw data counts derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010a;@Robinson:2010] as seen in Panel A of Figure 1.
-However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010;@Love:2014aa;@Andrews:2019aa], which includes the CLR [@fernandes:2013] yet the mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013;@Love:2014aa].
+We use the data of Gierliński et al.
+[-@Gierlinski:2015aa] who conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf2 gene knockout, \(\Delta\)snf2.
+This dataset has been used to argue that a dual cutoff approach is appropriate to limit the number of significant parts and for purposes of reproducibility [@Schurch:2016aa].
+However, in this benchmarking study,  the number of significantly different transcripts varied between 65% to >80% of all transcripts depending on the tool; in essence the desired behavior was that almost all transcripts were significantly different.
+The guidance on reproducibility runs counter to standard statistical practice where power is intrinsically linked to sample size and very large sample sizes are indeed desired [@Halsey:2015aa].
+Furthermore, the dual-cutoff approach is  known not to provide appropriate FDR control [@Zhang:2009aa; @Ebrahimpoor:2021aa].
+Schurch et al.
+-@Schurch:2016aa further suggested that different tools might be better in some conditions or datasets than others because each tool has different intrinsic statistical power and Type 1 and Type 2 errors.
+Through the lens of scale uncertainty, the behaviour of the tools in this study show unacknowledged bias; false confidence in the precision of the estimate as sample size increases because o mis-specification  in the scale of the data.
+This certainty is driven by the assumptions of the tool not the actual experiment being investigated [@Nixon2024B].
 
-```{r, echo=F, warning=F, message=F,comment=F, result=F} 
-library(DESeq2)
-devtools::load_all('~/Documents/0_git/ALDEx_bioc')
-#library(ALDEx2)
-```
-```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as clr values.
-Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale.
-Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first.
-Panel C shows the equivalent values calculated by ALDEx2	 in which the expected clr value for each transcipt are plotted vs.
-the expected dispersion.
-The red line in each panel shows the loess line of fit to the mid-point of the distributions.
-In panels B and C the amount of dispersion reaches a minimum at moderate	 values.
-The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value.
-The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }
-
-url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
-yst <- read.table(url, header=T, row.names=1)
-# remove the one gene with 0 reads
-
-yst <- yst[rownames(yst) != "YOR072W-B",]
-
-# Gierlinski:2015aa
-yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
-yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL
-
-conds <- c(rep('S', 44), rep('W', 42))
-coldata <- data.frame(conds)
-
+```{r yst-res, echo=F, warning=F, message=F,comment=F,}
+#ALDEx2
+library(ALDEx2)
+load(file="analysis/yst.all.Rda")
+load(file="analysis/yst.s.all.Rda")
+load(file="analysis/yst.1.all.Rda")
 # DESeq2
@@
 
-#ALDEx2
-load('analysis/x.all.Rda')
+# with gamma = 0.5
 load('analysis/x.s.all.Rda')
+sig.des <- which(res@listData$padj < 0.05)
+sig.ald <- which(yst.all$we.eBH < 0.05)
+sig.s.ald <- which(yst.s.all$we.eBH < 0.05)
 
-# with gamma = 1
-load('analysis/x.s.all.Rda')
-sig.des <- which(res@listData$padj < 0.01)
-sig.ald <- which(x.all$we.eBH < 0.01)
-sig.s.ald <- which(x.s.all$we.eBH < 0.01)
-
-# get and plot mean/var or mean/SE for counts
-# and log counts
-yst.mn <- apply(yst, 1, mean)
-yst.mn.log <- apply(yst, 1, function(x) mean(log2(x)))
-yst.mn.log[is.infinite(yst.mn.log)] <- 0
-
-yst.v <- apply(yst, 1, var)
-
-par(mfrow=c(1,3))
-plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
-  xlab='Mean count', ylab='Variance')
-#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
-def <- data.frame(log2(yst.mn),log2(yst.v))
-lines(lowess(def, f=0.1), col=2, lwd=2)
-title('A: M v. V', adj=0, line= 0.8)
-abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
-
-l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
-plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
-  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
-lines(lowess(l.df, f=0.25), col=2, lwd=2)
-abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
- abline(h=0.0275, col='grey', lty=2)
-title('B: log2(M) v. log2(V)', adj=0, line= 0.8)
-
-l.df <- data.frame(x.all$rab.all, x.all$diff.win)
-plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
-  xlab='E(clr)', ylab='E(disp)')
-#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
-lines(lowess(l.df, f=.1), col=2, lwd=2)
-abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
- abline(h=0.208, col='grey', lty=2)
-title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
+sig.all <- yst.all$we.eBH < 0.05
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
+sig.s <- yst.s.all$we.eBH < 0.05
+sig.1 <- yst.1.all$we.eBH < 0.05
 ```
 
-These relationships depend on what is being analyzed: the variance or dispersion always increases with increasing raw (or normalzied) read count but decreases when measured on the log-ratio transformed data [@fernandes:2013;@Love:2014aa], reaching a minimum at some mid-point of the distribution.
-This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes.
-This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa].
-Furthermore, the dispersion is exceeding small being, for many transcripts, almost negligible.
-To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead to many false positives.
-From the perspective of Nixon et al.
-[-@nixon2023scale], it is reasonable to conclude that some of these results may be due to the choice of normalization.
-Indeed, benchmarking and comparison studies have repeatedly shown that the choice of normalization plays a role in which results are returned as significant; see for example [@maza2013], [@Dillies:2013] and [@Weiss:2017aa].
+Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.05; i.e.
+`r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or `r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2) of the `r nrow(yst.all)` transcripts.
+Such  large numbers of significant transcripts seems biologically unrealistic  and furthermore  breaks the necessary assumption made by the normalization methods that the majority of the features must be invariant.
+That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some if not the majority are driven by technical differences in the analysis [@Soneson:2013;@maza2013;@Dillies:2013;@Weiss:2017aa; @Schurch:2016aa].
 
-Using either DESeq2 or ALDEx2,  a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e.
-`r length(sig.des)` or `r length(sig.ald)` of the `r nrow(yst)` transcripts are statistically significant.
-Clearly, a 70\% positive rate seems biologically unjustified, and furthermore this breaks the necessary assumption made by the normalization methods that most features must be invariant.
-Finally, that `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not suggests that the choice of normalization plays a role in which results are returned as significant.
-
-```{r plot1, echo=F, fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis.
-DESeq2 or ALDEx2 were used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset.
+```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis.
+ALDEx2 was used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset.
 The results were plotted to show the relationship between difference and dispersion using effect plots ) or difference and the Benjamini-Hochberg corrected p-values (volcano plot).
-Panels A,B,D,E are for the unscaled analysis, and Panels C,F are for the scaled analysis.
+Panels A,C are for the unscaled analysis, and Panels B,D are for the scaled analysis.
 Each point represents the values for one transcript, with the color indicating if that transcript was significant in the scaled analysis and unscaled analysis (red) or in the unscaled analysis only (orange).
 Points in grey are not statistically signficantly different with any analysis.
-The horizontal dashed lines represent a log2(difference) of 1.4, which is a commonly applied cutoff when the majority of features are statistically significant."}
-par(mfrow=c(2,3))
-plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
-  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
-title('A: DESeq2 effect', adj=0, line= 0.8)
-points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
-  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
-  pch=19, cex=0.5)
-points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
-  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
-  pch=19, cex=0.5)
-abline(h=c(-1.4,1.4), lty=2, col='grey')
+The horizontal dashed lines represent a log2(difference) of \\(\\pm 1.4\\)."}
 
-plot(x.all$diff.win, x.all$diff.btw, col=rgb(0,0,0,0.1), 
+par(mfrow=c(2,2))
+
+#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
+#  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
+#title('A: DESeq2 effect', adj=0, line= 0.8)
+#points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
+#  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
+#  pch=19, cex=0.5)
+#points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
+#  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
+#  pch=19, cex=0.5)
+#abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+
+plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), 
    xlab='Dispersion', ylab='log2 Difference')
-title('B: ALDEx2 effect', adj=0, line= 0.8)
-points(x.all$diff.win[sig.ald], x.all$diff.btw[sig.ald], 
+title(expression(paste("A: Effect: ", gamma, " = 0")), adj=0, line= 0.8)
+points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald], 
   col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
-	abline(h=c(-1.4,1.4), lty=2, col='grey')
-points(x.all$diff.win[sig.s.ald], x.all$diff.btw[sig.s.ald], 
+	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+points(yst.all$diff.win[sig.s.ald], yst.all$diff.btw[sig.s.ald], 
   col=rgb(1,0,0,0.6), cex=0.5, pch=19)
-	abline(h=c(-1.4,1.4), lty=2, col='grey')
+	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-plot(x.s.all$diff.win, x.s.all$diff.btw, col=rgb(0,0,0,0.1),
+plot(yst.s.all$diff.win, yst.s.all$diff.btw, col=rgb(0,0,0,0.1),
    xlab='Dispersion', ylab='log2 Difference', xlim=c(0.1,5))
-title('C: ALDEx2 scaled effect', adj=0, line= 0.8)
-points(x.s.all$diff.win[sig.s.ald], x.s.all$diff.btw[sig.s.ald], 
+title(expression(paste("B: Effect: ", gamma, " = 0.5")), adj=0, line= 0.8)
+points(yst.s.all$diff.win[sig.s.ald], yst.s.all$diff.btw[sig.s.ald], 
   col=rgb(1,0,0,0.5), cex=0.5, pch=19)
-	abline(h=c(-1.4,1.4), lty=2, col='grey')
+	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
 # volcano
-plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
-  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
-title('D: DESeq2 volcano', adj=0, line= 0.8)
-points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
-  pch=19, cex=0.5)
-points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
-  pch=19, cex=0.5)
-  abline(v=c(-1.4,1.4), lty=2, col='grey')
+#plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
+#  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
+#title('D: DESeq2 volcano', adj=0, line= 0.8)
+#points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
+#  pch=19, cex=0.5)
+#points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
+#  pch=19, cex=0.5)
+#  abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-plot(x.all$diff.btw, -1*log10(x.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
+plot(yst.all$diff.btw, -1*log10(yst.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
    xlab='log2 Difference', ylab='-1 log10(p.adjust)')
-title('E: ALDEx2 volcano', adj=0, line= 0.8)
-points(x.all$diff.btw[sig.ald], -1*log10(x.all$we.eBH[sig.ald]	+1e-70), 
+title(expression(paste("C: Volcano: ", gamma, " = 0")), adj=0, line= 0.8)
+points(yst.all$diff.btw[sig.ald], -1*log10(yst.all$we.eBH[sig.ald]	+1e-70), 
   col=rgb(1,.66,0,.3), cex=0.5, pch=19)
-points(x.all$diff.btw[sig.s.ald], -1*log10(x.all$we.eBH[sig.s.ald]	+1e-70), 
+points(yst.all$diff.btw[sig.s.ald], -1*log10(yst.all$we.eBH[sig.s.ald]	+1e-70), 
   col=rgb(1,0,0,1), cex=0.5, pch=19)
-abline(v=c(-1.4,1.4), lty=2, col='grey')
+abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-plot(x.s.all$diff.btw, -1*log10(x.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
+plot(yst.s.all$diff.btw, -1*log10(yst.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
    xlab='log2 Difference', ylab='-1 log10(p.adjust)')
-title('F: ALDEx2 scaled volcano', adj=0, line= 0.8)
-
-points(x.s.all$diff.btw[sig.s.ald], -1*log10(x.s.all$we.eBH[sig.s.ald]+1e-15),
+title(expression(paste("D: Volcano: ", gamma, " = 0.5")), adj=0, line= 0.8)
+points(yst.s.all$diff.btw[sig.s.ald], -1*log10(yst.s.all$we.eBH[sig.s.ald]+1e-15),
   col=rgb(1,0,0,1), cex=0.5, pch=19)
-	abline(v=c(-1.4,1.4), lty=2, col='grey')
+	abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
@@
 
-As shown using effect [@gloor:effect] and volcano plots [@Cui:2003aa] in Figure 2 A,B,D,E, the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance.
-This phenomenon is shared between methods: almost all the transcripts that are differentially abundant with an FDR \textless 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups.
-In the most extreme cases, some transcripts have a near zero difference between groups.
-This issue is not unique to this data set and it is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both corrected p-values and fold-changes  [@Schurch:2016aa] (commonly set at \(\pm 2^{1.4}\) for the latter).
+The effect plots  [@gloor:effect] in Figure 1A (ALDEx2) and Supplementary Figure 1 (DESeq2) shows that the majority of significant transcripts (red, orange) have negligible differences between groups and  very low dispersion and we suggest that this is driven by the experimental design [@Schurch:2016aa].
+Scale uncertainty can be incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
+Figure 1B shows that  setting  \( \gamma=0.5 \) results in far fewer transcripts being significant (`r length(sig.s.ald)`) and we observe that  the minimum dispersion increases from  `r round(min(yst.all$diff.win), 2)` (unscaled) to  `r round(min(yst.s.all$diff.win), 2)` (scaled).
+The Volcano plots in Figure 1 C and D show a similar story.
+Here we can see that adding scale increases the minimum FDR value and increases the concordance between the FDR value and the difference between groups (compare panels C and D).
+
+```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="Adding scale uncertainty changes the dispersion distribution.
+Panel A shows a plot of the expected value for relative abundance vs the expected value for the pooled dispersion as output by \\texttt{aldex.effect}.
+The dashed  horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5, and the light colored lines are lowess lines of fit through the center of mass of the data.
+Panel B plots the dispersion difference between  \\(\\gamma = 1\\) and \\(\\gamma = 0\\); note the non-linear relationship that highlights  the rotation that is evident in Panel A.
+The colored lines indicate the lowess line of fit through the centre of mass of the plot for the various  populations of points.
+The grey line is the total population and shows the difference \\(\\Delta\\), the red line is the population of significant transcripts (\\*) with no scale, the orange line is the population of significant transcripts with a difference threshold (T) of about \\(\\pm 2^{1.4}\\)-fold change, the blue line is the population of significant transcripts with \\(\\gamma = 0.5\\), and the cyan line is the significant population with \\(\\gamma = 1\\).
+\\(\\Delta\\): Difference, *: significant, T: thresholded.
+"}
+
+cuts <- yst.all$rab.all > -0.5 & yst.all$rab.all < 5
+par(mfrow=c(1,2))
+mn.mid <- median(yst.all$diff.win[cuts])
+mn.mid.s <- median(yst.s.all$diff.win[cuts])
+mn.mid.1 <- median(yst.1.all$diff.win[cuts])
+
+no.scale <- data.frame(yst.all$rab.all, yst.all$diff.win)
+half.scale <- data.frame(yst.s.all$rab.all, yst.s.all$diff.win)
+full.scale <- data.frame(yst.1.all$rab.all, yst.1.all$diff.win)
+
+# we change dispersion a lot for the bulk of the features
+plot(yst.all$rab.all, yst.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  xlab='rAbundance', ylab='dispersion')
+lines(lowess(no.scale, f=0.1), col='grey', lwd=3)
+points(yst.s.all$rab.all, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+lines(lowess(half.scale, f=0.1), col='pink', lwd=3)
+points(yst.1.all$rab.all, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3))
+lines(lowess(full.scale, f=0.1), col='royalblue1', lwd=3)
+abline(h=mn.mid, lty=2)
+abline(h=mn.mid.s, lty=2, col='red')
+abline(h=mn.mid.1, lty=2, col='blue')
+title(main='A', line=-1.2, adj=0.9)
+
+g0 <- expression(paste(gamma, " = 0"))
+g5 <- expression(paste(gamma, " = 0.5"))
+g1 <- expression(paste(gamma, " = 1"))
+legend(1,4, legend=c(g0, g5, g1), 
+  pch=19, col=c("black", "red", "blue"))
+
+# plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3), ylim=c(0,4.5), 
+#   xlab='g=0 dispersion', ylab='g=0.5 | g=1 dispersion')
+# points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+# abline(0,1, lty=2)
+# title(main='B', line=-1.2, adj=0.5)
+# legend(0,4, legend=c( "0.5", "1"), 
+#   pch=19, col=c("red", "blue"))
+
+diff.df <- data.frame(yst.all$rab.all, 
+  yst.1.all$diff.win- yst.all$diff.win)
+sig.df <- data.frame(yst.all$rab.all[sig.all],
+   yst.1.all$diff.win[sig.all]- yst.all$diff.win[sig.all])
+sig.t.df <- data.frame(yst.all$rab.al[sig.t], 
+  yst.1.all$diff.win[sig.t] - yst.all$diff.win[sig.t])
+sig.s.df <- data.frame(yst.all$rab.all[sig.s], 
+  yst.1.all$diff.win[sig.s] - yst.all$diff.win[sig.s])
+sig.1.df <- data.frame(yst.all$rab.all[sig.1], 
+  yst.1.all$diff.win[sig.1]- yst.all$diff.win[sig.1])
+
+ylab1 <- expression(paste("dispersion: ", gamma, "1 - ", gamma, "0"))
+
+plot(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win, xlab='rAbundance', 
+  ylab=ylab1, pch=19, cex=0.5, xlim=c(-11,10), ylim=c(.1,1.5))
+title(main='B', line=-1.2, adj=0.9)
+lines(lowess(diff.df, f=0.1), col='grey60', lwd=3)
+lines(lowess(sig.df, f=0.1), col='red', lwd=3)
+lines(lowess(sig.t.df, f=0.5), col='orange', lwd=3)
+lines(lowess(sig.s.df, f=0.5), col='royalblue1', lwd=3)
+lines(lowess(sig.1.df, f=0.7), col='cyan', lwd=3)
+
+d0 <- expression(paste(Delta))
+dstar <- expression(paste(Delta,"*"))
+dstarT <- expression(paste(Delta,"*T"))
+dg5 <- expression(paste(Delta,"*", gamma, "0.5"))
+dg1 <- expression(paste(Delta,"*", gamma, "1"))
+
+legend(-12,1.6, legend=c( d0, dstar, dstarT,dg5, dg1), 
+  pch="-", col=c("grey", "red", "orange", "royalblue1", "cyan"))
+
+```
+
+As shown by the effect plot in Figure 1 the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance.
+This phenomenon is not unique to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon occurs in DESeq2 (and presumably other methods although the relevant parameters are not exposed).
+This issue is not unique to this dataset and it is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both corrected p-values and fold-changes  [@Schurch:2016aa] (here set at \(\pm 2^{1.4}\) for the latter), although considerable variation in cutoff values is observed.
 These limits are shown by the dashed grey lines.
-In this case, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
-This result begs the question: why bother with significance tests at all?
+Here, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with that observed with that found by ALDEx2 with \( \gamma = 0.5 \) which identifies `r length(sig.s.ald)`.
+Indeed, Supplementary Figure 2, shows that even adding a very small amount of scale \( \gamma = 0.1 \) reduces the number of significant transcripts by more than half and the \texttt{aldex.scaleSim()} function can be used to identify those transcripts that are significant only because of an absence of scale.
+These results begs the question: why bother with significance tests at all if all transcripts with \( > 1.4\)-fold expression change are statistically significant?
 
-The very low dispersion estimate for most features is a by-product of sequencing and normalization.
-While the actual scale of the data is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale.
-To do this we incorporate scale models [@nixon2023scale] in lieu of normalizations through the default scale model that is now built into ALDEx2.
-Scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
-The \texttt{ALDEx2} package further contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty.
-In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs.
+The effect on dispersion with increasing  amounts of scale are shown in Figure 2A, where we can see that the  dispersion increases as scale is added.
+Note that the dispersion in the unscaled data in Figure 2A reaches a minimum near the mid-point of the distribution, and also does so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
+This shows more clearly that dispersion of many transcripts is  almost negligible in the absence of scale and makes the counter-intuitive suggestion that the variance in expression of the majority of genes with moderate expression is more predictable  than highly-expressed genes or of housekeeping genes [@Rocha:2020aa].
+This is at odds with the known biology of cells where single cell counting of highly-expressed transcripts shows that they have little intrinsic variation [@Taniguchi:2010aa].
 
-Applying the default scale model with \texttt{gamma=1}, a large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts `r length(sig.s.ald)` being called significantly different as shown in the volcano plot in Figure 1F (red points).
-Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.
-Indeed, adding scale uncertainty reduces the significant transcripts to a subset of those identified with the dual cutoff that have the largest effect size.
-Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.
+Adding scale by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 2A by the red and blue data points, and by the colored lines of fit through the centre of mass of the data.
+Less obvious is that the scaled dispersion estimates are rotated.
+Figure 2B shows a plot of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data to show this more clearly and here we can see that the scale is preferentially increasing the dispersion of the mid-expressed transcripts that formerly had negligible dispersion; examine the grey line of best fit (overlaid by the red line) for the trend.
+Panel B also shows the trend of the expression-dispersion relationship for transcripts that are classed as statistically significant.
+The red line shows the trendline with no added scale, and this trendline exactly overlays with the grey trendline of the bulk of transcripts.
+The orange trendline indicates those transcripts that are both statistically significant and that have a thresholded expression level of \( \pm 1.4\), and the dark blue and cyan lines show the statistically significant trendline for \(\gamma=0.5\) , or \(1.0\).
+Note that this has the effect of changing the distribution of parts identified as significant and that the substantially fewer significantly different genes are  in the very high abundance but low dispersion category.
+
+<!-- 
+Interestingly, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions; yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
+The CLR normalization assumes that the scale of the wild-type is exactly 98.4\% of the snf1 strain, and so is \( \theta = 0.016\).
+Given the per-generation time difference of about 15\%, this seems implausible.
+Setting \(\gamma=0.5\) can be interpreted as acknowledging uncertainty in the difference in the underlaying scale between groups.
+Nixon et al.
+-@Nixon2024B showed that the 95\% bounds on the scale uncertainty for a given \(\gamma\) value, say 0.5, can be calculated as \(2^{-(2 \gamma + \theta)}, 2^{(2 \gamma + \theta)}\), or \(2^{-(2*0.5 + 0.016)}, 2^{(2*0.5 + 0.016)} = 0.494,2.022\).
+Thus, by adding scale uncertainty, we are acknowledging that the data are uncertain and that any results are robust to that uncertainty, within the bounds of our model.
+-->
 
@@
 
-Macklaim and Gloor [-@Macklaim:2018aa] and Wu et al.
-[-@Wu2021] used a vaginal metatranscriptome dataset used to compare the gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
+Dos Santos et al.
+-@dosSantos:2024  used a vaginal metatranscriptome dataset  to compare the gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
 In this environment, both the relative abundance of species between groups and the gene expression level within a species is different [@macklaim:2013].
@@
 Thus, this is an extremely challenging environment in which to determine differential abundance as there are both compositional and scale changes between conditions.
-The accepted method to analyze vaginal metratranscriptome data is on a taxon-by-taxon basis [@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding can be ignored.
-The problem with a systemic analysis is that many housekeeping genes are returned as differentially abundant between groups, a result likely due to a disconnect between the scale assumptions of the normalization used [@Wu2021] and the actual change in scale between conditions.
+The usual method to analyze vaginal metratranscriptome data is on a taxon-by-taxon basis [@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding can be ignored.
+Attempts at system-wide analysis show that many housekeeping functions are returned as differentially abundant between groups; a result likely due to a disconnect between the scale assumptions of the normalization used [@Wu2021].
 
-In this example, we show how to specify the scale model explicitly and demonstrate that applying a user-defined scale model can account for some of these modeling difficulties.
-A user-defined scale model can control for both the mean difference of scale between groups (e.g., directly incorporate information on the differences in total number of bacteria between the BV and H conditions) as well as the uncertainty assumed in that difference.
+In this example, we show how to specify a user-defined, or \emph{informed}, scale model explicitly that can account for some of these modeling difficulties.
+An informed scale model can control for both the mean difference of scale between groups (e.g., directly incorporate information on the differences in total number of bacteria between the BV and H conditions) as well as the uncertainty assumed in that difference.
 To specify a user-defined scale model, we can pass a matrix of scale values instead of a single estimate of gamma to \texttt{aldex.clr()}.
@@
 While this matrix can be computed from scratch by the analyst, there is an \texttt{aldex.makeScaleModel()} function that can be used to simplify this step in most cases.
+This encodes the scale model as
+ \(\Lambda \sim N(log2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents the actual scale value for each sample.
+This can be a measured value (cell count, nucleic acid input, etc), or an imputed value.
+Nixon et al.
+-@Nixon2024B showed that only the ratio between the scale values for each sample was important; see Supplementary Figures 4 and 5 for a demonstration of this point.
 
-Figure 3A shows an effect plot [@gloor:effect] of the data where reads are grouped by function, corresponding approximately to grouping orthologous sequences regardless of the organism of origin.
-Each point represents one of the 3728 functions.
+Figure 3A shows an effect plot of the data where reads are grouped by function, corresponding  to grouping homologous sequences regardless of the organism of origin.
+Each point represents one of  3728 KEGG  functions [@Okuda:2008].
 There are many more functions represented in the BV group (bottom) than in the healthy group (top).
-This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because	there is a greater diversity of organisms in BV than in H samples and because the BV condition has at least an order of magnitude more bacteria than does the H condition.
+This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because	there is a greater diversity of organisms in BV than in H samples, and because the BV condition has at least an order of magnitude more bacteria than does the H condition.
 
- We can	 see that there are a large number of core metabolic and information processing functions that are shared between the two groups (boxed area in Figure 3A), and inspection shows that these largely correspond to core metabolic functions that would not be expected	 to contribute to differences in ecosystem behaviour and so should not be differentially abundant.
-As a proxy for housekeeping functions the core ribosome functions (blue) shows that their mean location is not centred on 0.
+There are 101  functions with low dispersion that appear to be share by both groups (boxed area in Figure 3A, and colored in cyan).
+Inspection shows that these   largely correspond to core metabolic functions such as transcription, translation, ribosomal proteins, glycolysis, replication, chaperones, etc (Supplementary file housekeeping.txt).
+The transcripts of many of these are commonly used as invariant reference sequences [@Rocha:2020aa] and so would not be expected	 to contribute to differences in ecosystem behaviour and should be centred on 0 difference.
+These should not be scored as among the most differentially abundant.
 The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5).
-Since they have very low dispersions, these functions are identified as differentially abundant (red) along with many others.
-While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups because these are functions in common to every organism.
+While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups as these are common to every organism.
 Of more interest is determining the functions that are different between groups because these are unique or over-expressed in one group relative to the other.
@@
 ```{r ribo, echo=F}
-ribo.v <- c("K02863","K02864","K02865","K02866","K02867","K02871","K02872","K02873","K02874","K02876","K02877","K02878","K02879","K02880","K02881","K02883","K02884","K02885","K02886","K02887","K02888","K02890","K02892","K02895","K02897","K02899","K02902","K02904","K02905","K02906","K02907","K02909","K02911","K02913","K02914","K02916","K02920","K02922","K02926","K02931","K02932","K02933","K02935","K02936","K02938","K02939","K02941","K02943","K02945","K02946","K02948","K02949","K02950","K02952","K02954","K02955","K02956","K02959","K02961","K02963","K02964","K02965","K02967","K02968","K02969","K02970","K02975","K02981","K02982","K02984","K02986","K02987","K02988","K02990","K02991","K02992","K02994","K02995","K02996","K02997","K02998","K07590")
+# list of ribosomal assocaited functions in the datset
+# TIL about complete.cases()
+ribo.v <- c("K02863","K02864","K02867","K02871","K02874","K02876","K02878","K02879","K02881","K02884","K02886","K02887","K02888","K02890","K02892","K02895","K02897","K02899","K02902","K02904","K02906","K02907","K02909","K02911","K02913","K02914","K02916","K02926","K02931","K02933","K02935","K02939","K02945","K02946","K02948","K02950","K02952","K02954","K02956","K02959","K02961","K02963","K02965","K02967","K02968","K02970","K02982","K02986","K02988","K02990","K02992","K02994","K02996")
 ```
@@
 Panel A shows an effect plot for the default analysis where the functions that are elevated in  the  healthy individuals have positive values and functions that are elevated in BV have negative values.
-Highlighed in the box are highly abundant KOs that are almost exlusively housekeeping functions, with ribosomal KOs highlighted in blue,  statistically significant (FDR < 0.01) functions in red, and non-significant functions in black or orange.
+Highlighed in the box are KOs that are almost exlusively housekeeping functions and are  colored cyan.
 These housekeeping functions should be located on the midline of no difference.
-Panel B shows the same data scaled with `gamma = 0.5`, which increase the minimum dispersion approximatly by one unit.
-Here the housekeeping functions from Panel A are colored cyan or blue for reference.
-Panel C shows the same data scaled with `gamma = 0.5` and a 0.15 fold difference in dispersion applied to the BV samples relative to the H samples.
-The orange functions are now statistically significant.
-Note that this shifts the midpoint of the housekeeping functions towards the midline." }
+Panel B shows the same data scaled with \\(\\gamma = 0.5\\), which increase the minimum dispersion as before.
+Panel C shows the same data scaled with \\(\\gamma = 0.5\\) and a 0.15 fold difference in dispersion applied to the BV samples relative to the H samples.
+In these plots statistically significant (FDR < 0.01) functions in the informed model are  in red, false positive functions are in blue,  non-significant functions in black and false negative functions are in orange.
+" }
 
@@
 # all variables recreated using the makefile
-
+# from code/meta_aldex.R
+load('analysis/xg.clr.Rda')
 load('analysis/xt.Rda')
@@
 
-hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3]
-hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
+hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3]
+hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3
 
-fn <- rownames(xt.m.all)[xt.m.all$we.eBH < 0.01 & xt.m.all$diff.btw < 1 & xg.all$we.eBH > 0.01 & xt.all$we.eBH > 0.01]
+fn <- rownames(xt.m.all)[xt.m.all$we.eBH < 0.01 & xt.m.all$diff.btw < 1 &  xt.all$we.eBH > 0.01]
+fp <- rownames(xt.all)[xt.all$we.eBH < 0.01 & xt.m.all$we.eBH > 0.01 & !xt.m.all$we.eBH < 0.01]
 
@@
 aldex.plot(xt.all, cutoff.pval=0.01)
-points(xt.all[ribo.v,'diff.win'], xt.all[ribo.v,'diff.btw'], col='blue', cex=0.5, pch=19)
-points(xt.all[fn,'diff.win'], xt.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
-rect(0.5,0,3,5, col=rgb(0,0,0,0.1), lty=2, lwd=3)
+points(xt.all[fp,'diff.win'], xt.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
+points(xt.all$diff.win[hk.off], xt.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
+points(xt.all[fn,'diff.win'], xt.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19)
+rect(0.5,1,2.5,3, col=rgb(0,0,0,0.1), lty=2, lwd=3)
 title('A: ALDEx2 unscaled', adj=0, line= 0.8)
+legend(0,12, legend=c("HK", "TP", "FP", "FN"), pch=19, cex=0.5, col=c("cyan", "red", "orange", "blue"))
 
 aldex.plot(xg.all, xlim=c(0.3,9), cutoff.pval=0.01)
+points(xg.all[fp,'diff.win'], xg.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
 points(xg.all$diff.win[hk.off], xg.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
-points(xg.all[ribo.v,'diff.win'], xg.all[ribo.v,'diff.btw'], col='blue', cex=0.5, pch=19)
-points(xg.all[fn,'diff.win'], xg.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
+points(xg.all[fn,'diff.win'], xg.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19)
 title('B: ALDEx2 gamma', adj=0, line= 0.8)
@@
 aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01)
+points(xt.m.all[fp,'diff.win'], xt.m.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
 points(xt.m.all$diff.win[hk.off], xt.m.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
-points(xt.m.all[ribo.v,'diff.win'], xt.m.all[ribo.v,'diff.btw'], col='blue', cex=0.5, pch=19)
-points(xt.m.all[fn,'diff.win'], xt.m.all[fn,'diff.btw'], col='orange', cex=0.5, pch=19)
+points(xt.m.all[fn,'diff.win'], xt.m.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19)
 title('C: ALDEx2 both', adj=0, line= 0.8)
@@
 
-Applying the default scale model of \texttt{gamma=0.5} increases the dispersion as expected but does little to move the large number of housekeeping functions toward the midline of no difference.
-This is as expected: the mean of the default scale model is based on the CLR normalization so no shift would be expected over the original ALDEx2 model.
-Nevertheless, about 50\% of the housekeeping functions are no longer statistically significantly different.
+Applying the default scale model of \(\gamma=0.5\) increases the dispersion as expected but does little to move the large number of housekeeping functions toward the midline of no difference.
+This is as expected; the mean of the default scale model is based on the CLR normalization so no shift in location would be expected over the original ALDEx2 model.
+Nevertheless, about 30\% of the housekeeping functions are no longer statistically significantly different.
 Note that this change is simple to conduct, has no additional computational complexity and requires only a slight modification from the analyst.
@@
 Up to this point, scale uncertainty has been applied as an extension of the CLR normalization via the default scale model, but a user-defined scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix.
-While not necessary in all datasets (e.g., the transcriptome example discussed in the previous section and see the Supplementary material), it can greatly improve modeling in certain cases, especially when the scale is highly asymmetric between conditions.
-In order to specify a scale model from scratch, we need to revisit the concept that all normalizations in widespread use are actually ratios with the denominator implied by the normalization.
-Therefore, we can easily deviate from a certain normalization (e.g., the geometric mean assumption implied by the CLR) by specifying a total model based on the mean difference between conditions.
-While knowing the mean difference between conditions may seem cumbersome in practice, it is the \emph{relationship} between the group scale values that is important, not their raw values (Nixon et al.
-2023).
-This can be illustrated quite simply by starting with the mean ratio in \(G\) between groups for the yeast transcriptome dataset which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for WT; their ratio being 1.17, or a 0.17-fold difference.
-Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental Figure 2.
-This ratio can be adjusted to alter the mean assumption placed on the group scale values.
+Our starting point for this is to identify the naive scale estimate from the data, and this can be done by calculating the mean scale value for each group.
+The scale estimate for samples can be accessed in the \texttt{@scaleSamps} slot from the \texttt{aldex.clr} output.
+In this datset the scale estimate for the healthy group is `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])),2)` and for the BV group is  `r round(mean(rowMeans(xg@scaleSamps[c(9:36),])),2)` for a difference of `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),])),2)`.
+This is interpreted as the scale of the H group of samples being `r round(2^(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),]))),2)` greater than the BV group and this precise but incorrect estimate places the location of the housekeeping genes off the midline of no difference.
 
-With this background, we can understand how to apply a user-defined scale model to the metratranscriptome dataset.
+We desire a scale model that approximately centres the housekeeping functions.
+We anticipate that housekeeping functions should be nearly invariant and so an appropriate scale in this dataset is likely closer to 0 than the naive estimate.
 While a user-defined scale model can be quite flexible with relative scales that are distinct for each group (or even each sample) along with their uncertainties, here we focus on using the  \texttt{aldex.makeScaleMatrix()} function.
 This function uses a logNormal distribution to build a scale matrix given a user-specified mean difference between groups and uncertainty level.
-Applying a per-group relative differential scale of 0.15 moves the housekeeping functions to the midline of no difference, and applying a gamma of 0.5 provides the same dispersion as using a gamma of 0.5 with the CLR (Figure 3C).
-Note that now a significant number of functions are differentially up in BV that were formerly classed as not different without scale, or when only a default scale was applied.
-These former false negatives are noted in orange in each panel.
-Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant.
-Thus, applying a differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.
+Applying a per-group relative differential scale of 0.15 moves the housekeeping functions to the midline of no difference (Figure 3C), and applying a gamma of 0.5 provides the same dispersion as in panel B).
+Note that now a significant number of functions are differentially up in BV that were formerly classed as not different without the full scale model (orange), or when only a default scale was applied.
+Inspection of the functions shows that these are largely missing from the \emph{Lactobacillus} species and so should actually be captured as differentially abundant in the BV group.
+Thus, applying a differential scale allows us to distinguish between both false positives (housekeeping functions in cyan, and others in blue) and false negatives (orange functions) even in a very difficult to analyze dataset.
+The remarkable improvements in biological interpretation afforded by this full scale model, and the transferrability of it between sample cohorts of the same condition is outlined elsewhere [@dosSantos:2024].
 We suggest that the default scale model  is sufficient when the data are approximately  centred.
-However, when datasets are not well centred or when the investigator has prior information about the underlying biology, we advocate developing and using a user-specified scale model.
+However, an informed model is more appropriate with datasets are not well centred or when the investigator has prior information about the underlying biology.
 
 # Discussion
+\doublespacing
+\singlespacing
+Biological systems are both predictably variable and stochastic [@Taniguchi:2010aa] and current measurement methods that rely on high throughput sequencing fail to capture all of that variation, particularly variation due to scale [@nixon2023scale;@Nixon2024B].
+In the absence of external information (such as spike-in probes [@Loven:2012aa], cell counting [@Vandeputte:2017aa], FISH [@yeast-absolute] etc) sequencing depth normalisation methods cannot recapture the scale information [@Loven:2012aa], and can only normalize for the technical variation due to sequencing depth.
 
-Biological count data derived from HTS can be decomposed into two parts: the relative (compositional) and the absolute (scale), and the product of these generates a fully scaled biological system (Nixon et al.
-2023).
-Biological systems are both predictably variable and stochastic, and current measurement methods that rely on high throughput sequencing fail to capture all of that variation, particularly variation due to scale.
-In the absence of information external to the sequencing run itself, no normalisation method can recapture any of the scale information (Lovén et al.
-2012).
-Even so, many methods such as flow cytometry (Vandeputte et al.
-2017), spike-in probes and fluorescent in-situ hybridization (Lovén et al.
-2012; Marguerat et al.
-2012) have been proposed as a means to do this.
-In the context of the updated ALDEx2 software suite, these methods can be useful if they contain information on the particular scale the researcher is interested in.
+Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Yerke:2024aa].
+Generically, it is observed that the actual agreement between analysis methods can be modest, and  no single method outperforms all others in every dataset or type of experiment.
+That different tools appear to work more reliably in different datasets from different sources can be explained as different  normalizations being a better fit to the scale  a particular dataset by chance.
+
+From our perspective, the disagreement between tools can be explained by the partially identifiable nature of the data; that is,  the  observed data are consistent with multiple ways of estimating the parameters [@nixon2023scale].
+Partially identifiable data exhibit multiple pathologies, chief among them being that that different analytic approaches can produce  different parameter estimates and that more data produces worse estimates because the additional data increases the precision of a flawed estimate [@gustafson2015bayesian].
+In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@nixon2023scale], and that different analytic approaches result in different parts being chosen as significantly different [@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa].
+
+Scale simulation  is now build into ALDEx2 to address the issue of partially identifiability, and through this lens the two main root causes to common HTS data pathologies can be proposed.
+The first contributing factor is the observed very low dispersion estimate for many features that is a by-product of experimental design, sequencing and normalization.
+In the Schurch et al.
+@Schurch:2016aa dataset, the data were  from single colonies derived from a single culture.
+Thus, it is more accurate to describe the 96 samples as wet-lab technical replicates rather than independent samples.
+However, this type of replicate is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.
+
+In the yeast transcriptome dataset, applying the default scale model with \( \gamma=0.5 \) a large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 1D), and this results in modest number of transcripts, `r length(sig.s.ald)`, being called significantly different as shown in the volcano plot in Figure 1D (red points).
+In addition, there was now a strong concordance between the difference and p-values (Figure 1D), this is not surprising.
+In hindsight, what is not obvious is why the unscaled volcano plot shows such poor correspondence.
+We suggest that this can be explained by random fluctuations in the many variance estimates with  very low values, and this is supported by the  plot shown in Figure 2B.
+Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.
+Indeed, adding scale uncertainty reduces the significant transcripts to a subset of those identified with the dual cutoff that have the largest effect size.
+Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.
+
+While the actual scale of the underlying environment is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale.
+To do this we incorporate scale models [@nixon2023scale] in lieu of normalizations through the default scale model that is now built into ALDEx2.
+Scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
+The ALDEx2 package  contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty, and an example is shown in  Supplementary Figure 2 for the yeast transcriptome dataset.
+Here it is clear that even tiny amounts of scale preclude the majority of transcripts from being considered statistically significant.
+In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs, but further work is needed.
+
+The second contributing factor is unacknowledged asymmetry in many datasets.
+In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult-to-analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
+We showed one such example for the metatranscriptome dataset in Figure 3.
+Here the dataset was highly asymmetrical, and the TMM and RLE normalizations could not fully move all the housekeeping genes to the midline of no difference or the tools exhibited other pathologies (Supplemental Figure 6).
+Incorporating differential scale on a per-group basis moves the mass of the housekeeping functions towards the midline of no difference and so affects both Type I and Type II error rates.
+It is also of note that in the case of true biological replicates (different individuals) that adding a modest amount of scale \(\gamma=0.5\) had little effect on the concordance between the difference between groups and the p-values as shown in Supplementary Figure 7.
+Thus, in this dataset the scale mis-specification was affecting mainly the location of the difference between groups.
+
+In the metatranscriptome analysis, transcripts that were previously not classed as differentially abundant were now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such.
+Note  the assumption that housekeeping genes should not generally be included in differential abundance analysis is implicit in the dual p-value:fold-change cutoff approach in widespread use.
+While we acknowledge that some prior information on which housekeeping transcripts should not be classed as differentially abundant is needed, we suggest that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
+Thus, the use of this prior knowledge is not unique to our approach.
+
+Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses.
+An advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
+This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models can make this an explicit part of the model building process.
+While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions and a detailed description of these points is describe elsewhere [@nixon2023scale].
+In this work, we simply use the parameter \(\gamma\) in the default scale model as a term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption.
+In the metatranscriptome example and elsewhere [@Nixon2024B], we also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.
 
 The ALDEx2 R package is readily amenable for incorporating scale uncertainty.
-Originally, this tool used the only the Dirichlet distribution to sample compositional uncertainty.
-Rather than applying the CLR normalization as in the original ALDEx2, scale uncertainty can be added through the use of a scale model with no additional computational complexity.
-Scale uncertainty can be sampled from any distribution depending on prior knowledge or preference.
+Originally, this tool used the only the Dirichlet distribution to sample compositional uncertainty, but scale uncertainty can be added through the use of a scale model with no additional computational complexity.
 By default, ALDEx2 samples scale from a logNormal distribution inspired by the CLR normalization. 
-However, there is the option to introduce a full scale model that  encapsulates both uncertainty and asymmetry in underlying scale.
+However, scale uncertainty can be sampled from any distribution depending on prior knowledge or preference, using the option to introduce a full informed scale model that  encapsulates both uncertainty and asymmetry in underlying scale.
 
-All normalizations attempt to make the samples in a dataset commensurate but do not explicitly address the scale of the underlying system.
-The two supplementary tables show that these normalizations can be interpreted as scale assumptions.
-However, the general lack of scale information has important consequences for the analysis of HTS datasets.
-One issue is that analysis tools seem over-powered with even moderate sample sizes, and different tools have different intrinsic power, Type 1 and Type 2 error rates (Schurch et al.
-2016) and fail to account for false discovery rates properly [@hawinkel2017;@Li:2022aa].
-Traditional recommendations would suggest that using small sample sizes in analysis leads to less reliability and reproducibility in analyses since surprisingly large sample sizes are needed to determine reproducible p-values (e.g., Halsey et al.
-[-@Halsey:2015aa]).
-However, recommendations to use small sample sizes are widespread in multivariate datasets such as RNA-seq datasets.
-We argue that this is a hallmark of unacknowledged bias: since normalizations are typically assumed to be correct with no uncertainty, smaller samples offer an unintentional buffer against this misspecification.
-Another issue is that datasets are difficult to analyze when  they contain systematic asymmetry, with different tools exhibiting differing pathologies with these datasets [@Robinson:2010a;@Wu2021].
-
-Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes [@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa] and microbiomes [@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017].
-Generically, it is observed that the actual agreement between methods can be modest, and there is usually a recommendation that some methods are better suited for individual datasets or for specific types of analysis; e.g.,~`Normalization and microbial differential abundance strategies depend upon data characteristics' [@Weiss:2017aa].] However, another interpretation of such studies is that different normalizations can be better suited to different datasets by happenstance, but it is not obvious how the use of specific normalizations could be justified on a per dataset basis.
-There are now published examples of multiple large replication efforts in psychology [@Open-Science-Collaboration:2015aa], cancer biology [@Rodgers:2021aa], ecology and evolution [@evoRepr-pre].
-In a nutshell, experiences from these studies indicate that published data typically have higher effect sizes and appear more significant than replication efforts with the same data, or with newly generated replication data [@Rodgers:2021aa].
-Thus, most published data should be viewed with suspicion [@Ioannidis:2005aa] unless efforts have been made to ensure that the results are robust.
-
-In the case of overpowering, HTS analyses seem to be more robust when applying a dual cutoff of both p-value and difference between group means [@Schurch:2016aa].
-Figure 2 shows one reason for this robustness could be that the dual cutoff is mimicking the effect of including scale uncertainty, since substantially similar transcripts are identified by the two approaches.
-We believe that adding scale uncertainty leads to a more transparent analysis compared to the cutoff as it is easy to describe the conditions under which certain conclusions hold.
-Furthermore, while using the post-hoc difference cutoff is useful for differential abundance analysis, it is not clear how this can be incorporated into other kinds of downstream analyses.
-Conversely, analyses that include scale uncertainty are fully compatible with existing downstream analyses.
-
-In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult-to-analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
-We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference or exhibit other pathologies (Supplemental).
-Incorporating differential scale on a per-group basis moves the mass of the data towards the midline of no difference and so affects both Type I and Type II error rates.
-In this analysis, transcripts that were previously not classed as differentially abundant are now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such.
-While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
-Furthermore, the assumption that housekeeping genes should not generally be included in differential abundance analysis is implicit in the dual p-value, fold-change cutoff approach in widespread use.
-Thus, the use of this prior knowledge is not unique to our approach.
-
-In summary, while the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and can help explain some of the underlying biology.
-Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on the analysis weighted accordingly.
-Additionally, the use of user-defined  scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups.
+In summary, we supply a toolkit that makes incorporating scale simple to incorporate for any type of HTS dataset.
+While the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and can help explain some of the underlying biology, and known issues with the analysis of HTS data.
+Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on conclusions weighted accordingly.
+Additionally, the use of user-defined  scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups [@dosSantos:2024].
 Thus, reporting scale uncertainty should become a standard practice in the analysis of HTS datasets as a way to identify which features are most robust to differences in the underlying system.
-Finally, we supply a toolkit that makes incorporating scale simple even for datasets that come from highly asymmetrical environments.
 

### 2024-07-22 go4.Rmd
focussed on dispersion version

---
title: "Controlling for false precision using Scale Simulation in ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  nucleic-acids-research.csl: pdf_document
package: ALDEx2
abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total gene expression) of the underlying biological systems. Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses. Recently we showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can lead to dramatic increases in false positive and false negative rates. Here we describe updates to the ALDEx2 R package that mitigate these problems by directly modeling uncertainty in the unmeasured system scale through the use of a \textit{scale model}. Scale models generalize the idea of normalizations and can be thought of as explicitly modeling the error in normalization by examining a distribution over all possible  normalizations.  Beyond enhancing the robustness of HTS analyses,  the use of scale models within ALDEx2 enhances the transparency and  reproducibility of analyses by making implicit normalizing assumptions an explicit part of the model building process.

---

# Results

- HTS has a dispersion problem since the log(var(x)) \(\ne\) var(log(x)) (Fig 1)

- this causes false positive inference and the double filtering approach is commonly used to weed out FP, or nuisance positives. However this causes specificity problems since BH corrections are useful only in the total dataset and not when subset

- HTS has a location problem, and common normalizations are assumed to correct for this. However, there is no universal normalization and no standard way to identify when the assumptions of the normalization are broken

```{r, echo=F, warning=F, message=F,comment=F, result=F} 
library(DESeq2)
library(ALDEx2)
```
```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T,fig.dim=c(7,3), fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as CLR values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2	 in which the expected CLR value for each transcipt are plotted vs. the expected dispersion. Panel D shows the output for ALDEx2 with \\( \\gamma=0.5 \\). The red line in each panel shows the LOESS line of fit to the mid-point of the distributions. In panels B and C the amount of dispersion reaches a minimum at moderate	 values.  The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }

url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
yst <- read.table(url, header=T, row.names=1)
# remove the one gene with 0 reads

yst <- yst[rownames(yst) != "YOR072W-B",]

# Gierlinski:2015aa
yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL

conds <- c(rep('S', 44), rep('W', 42))
coldata <- data.frame(conds)

# DESeq2
load('analysis/res.Rda')

#ALDEx2
load('analysis/x.all.Rda')
load('analysis/x.s.all.Rda')

# with gamma = 1
load('analysis/x.s.all.Rda')
sig.des <- which(res@listData$padj < 0.05)
sig.ald <- which(x.all$we.eBH < 0.05)
sig.s.ald <- which(x.s.all$we.eBH < 0.05)

# get and plot mean/var or mean/SE for counts
# and log counts
yst.mn <- apply(yst, 1, mean)
yst.mn.log <- apply(yst, 1, function(x) mean(log2(x)))
yst.mn.log[is.infinite(yst.mn.log)] <- 0

yst.v <- apply(yst, 1, var)

par(mfrow=c(1,3))
#plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
#  xlab='Mean count', ylab='Variance')
#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
#points(log2(yst.mn[sig.s.ald]), log2(yst.v[sig.s.ald]), pch=19, cex=0.5, col='red')
#def <- data.frame(log2(yst.mn),log2(yst.v))
#lines(lowess(def, f=0.1), col=2, lwd=2)
#title('A', adj=0, line= 0.8)
#abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)

l.df <- data.frame(yst.mn.log, res@listData$lfcSE)

plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
points(yst.mn.log[sig.des], res@listData$lfcSE[sig.des], pch=19, 
  cex=0.5, col='orange')
points(yst.mn.log[sig.s.ald], res@listData$lfcSE[sig.s.ald], pch=19, 
  cex=0.5, col='red')
lines(lowess(l.df, f=0.25), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.0275, col='grey', lty=2)
title('B', adj=0, line= 0.8)

l.df <- data.frame(x.all$rab.all, x.all$diff.win)
plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(disp)')
points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
points(x.all$rab.all[sig.s.ald], x.all$diff.win[sig.s.ald], col='red', cex=0.4)
lines(lowess(l.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('C', adj=0, line= 0.8)

l.s.df <- data.frame(x.s.all$rab.all, x.s.all$diff.win)
plot(x.s.all$rab.all, x.s.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(disp)', ylim=c(0,4.5))
points(x.s.all$rab.all[sig.ald], x.s.all$diff.win[sig.ald], col='orange', cex=0.4)
points(x.s.all$rab.all[sig.s.ald], x.s.all$diff.win[sig.s.ald], col='red', cex=0.4)
lines(lowess(l.s.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.s.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('D', adj=0, line= 0.8)
```
There is an anomaly in the dispersion curve, whereby mid-range parts are less dispersed than high-range parts [@Anders:2010, Fig 7].

```{r }
# flattening the dispersion anomaly curve
par(mfrow=c(1,2))
 plot(x.all$rab.all, x.all$diff.win)
 points(x.s.all$rab.all,x.s.all$diff.win, col='red', pch=19, cex=0.5)

ls.diff <- data.frame(x.s.all$rab.all, x.s.all$diff.win -  x.all$diff.win)

 plot(x.s.all$rab.all, x.s.all$diff.win -  x.all$diff.win, pch=19, col=rgb(0,0,0,0.2), cex=0.5)
lines(lowess(ls.diff, f=.1), col=2, lwd=2)

```
- scale is a universal way of dealing with both of these common problems.

- all normalizations can be written as the denominator of a ratio, and all normalizations make an assumption about the scale of the system. 
  - Proportions, and any other constant denominator normalisation implicitly assume that the scale is equal between every sample is constant. 
  - The CLR or other log-ratios where the denominator is calculated from the data implicitly assume that the value of the denominator is related to the scale of the data
  - by analogy, the TMM assumes that a subset of parts in one sample can be chosen as the scale reference
  - by analogy, the RLE assumes that a different part in each sample can be chosen as the scale reference

```{r fig-filter}
# code/yeast-aldex.R
load("analysis/x.all.Rda") # gamma=1e-3
load("analysis/x.s.all.Rda") # gamma=0.5

sig.ald <- x.all$we.eBH < 0.05
sig.s.ald <- x.s.all$we.eBH < 0.05

par(mfrow=c(2,3))
# effect plot
plot(x.all$diff.win, x.all$diff.btw, pch=19, cex=0.2, col="grey30",
  xlab="dispersion", ylab="log2FC", xlim=c(0, 4.5))
points(x.all$diff.win[sig.ald], x.all$diff.btw[sig.ald], 
  pch=19, cex=0.2, col="orange")
points(x.all$diff.win[sig.s.ald], x.all$diff.btw[sig.s.ald], 
  pch=19, cex=0.3, col="red")
title(main="effect g=1e-3")
abline(v=0, lty=2)
abline(h=0, lty=2)

# volcano plot
plot(x.all$diff.btw, -log10(x.all$we.eBH+1e-80), pch=19, cex=0.2, col="grey30",
  xlab="log2FC", ylab="-log10(q)")
points(x.all$diff.btw[sig.ald], -log10(x.all$we.eBH+1e-80)[sig.ald], 
  pch=19, cex=0.2, col="orange")
points(x.all$diff.btw[sig.s.ald], -log10(x.all$we.eBH+1e-80)[sig.s.ald], 
  pch=19, cex=0.3, col="red")
title(main="VQ g=1e-3")
abline(h=0, lty=3)

# Dispersion (variance) Q plot
plot(x.all$diff.win, -log10(x.all$we.eBH+1e-80), pch=19, cex=0.2, col="grey30",
  xlab="dispersion", ylab="-log10(q)", xlim=c(0, 4.5))
points(x.all$diff.win[sig.ald], -log10(x.all$we.eBH+1e-80)[sig.ald], 
  pch=19, cex=0.2, col="orange")
points(x.all$diff.win[sig.s.ald], -log10(x.all$we.eBH+1e-80)[sig.s.ald], 
  pch=19, cex=0.3, col="red")
title(main="VQ g=1e-3")
abline(v=0, lty=2)
abline(h=-log10(0.05), lty=3)

## MA plot
#plot(x.all$rab.all, x.all$diff.btw, pch=19, cex=0.2, col="grey30",
#  xlab="log2rAB", ylab="log2FC")
#points(x.all$rab.all[sig.ald], x.all$diff.btw[sig.ald], 
#  pch=19, cex=0.2, col="orange")
#points(x.all$rab.all[sig.s.ald], x.all$diff.btw[sig.s.ald], 
#  pch=19, cex=0.3, col="red")
#title(main="MA g=1e-3")
#abline(h=0, lty=2)

# effect plot
plot(x.s.all$diff.win, x.s.all$diff.btw, pch=19, cex=0.2, col="grey30",
  xlab="dispersion", ylab="log2FC", xlim=c(0, 4.5))
points(x.s.all$diff.win[sig.ald], x.s.all$diff.btw[sig.ald], 
  pch=19, cex=0.2, col="orange")
points(x.s.all$diff.win[sig.s.ald], x.s.all$diff.btw[sig.s.ald], 
  pch=19, cex=0.3, col="red")
title(main="effect g=0.5")
abline(v=0, lty=2)
abline(h=0, lty=2)

# volcano plot
plot(x.s.all$diff.btw, -log10(x.s.all$we.eBH+1e-45), pch=19, cex=0.2, col="grey30",
  xlab="log2FC", ylab="-log10(q)")
points(x.s.all$diff.btw[sig.ald], -log10(x.s.all$we.eBH+1e-45)[sig.ald], 
  pch=19, cex=0.2, col="orange")
points(x.s.all$diff.btw[sig.s.ald], -log10(x.s.all$we.eBH+1e-45)[sig.s.ald], 
  pch=19, cex=0.3, col="red")
title(main="volcano g=0.5")
abline(v=0, lty=2)
abline(h=-log10(0.05), lty=3)

# Dispersion (variance) Q plot
plot(x.s.all$diff.win, -log10(x.s.all$we.eBH+1e-45), pch=19, cex=0.2, col="grey30",
  xlab="dispersion", ylab="-log10(q)", xlim=c(0, 4.5))
points(x.s.all$diff.win[sig.ald], -log10(x.s.all$we.eBH+1e-45)[sig.ald], 
  pch=19, cex=0.2, col="orange")
points(x.s.all$diff.win[sig.s.ald], -log10(x.s.all$we.eBH+1e-45)[sig.s.ald], 
  pch=19, cex=0.3, col="red")
title(main="VQ g=0.5")
abline(v=0, lty=2)
abline(h=-log10(0.05), lty=3)

## MA plot
#plot(x.s.all$rab.all, x.s.all$diff.btw, pch=19, cex=0.2, col="grey30",
#  xlab="log2rAB", ylab="log2FC")
#points(x.s.all$rab.all[sig.ald], x.s.all$diff.btw[sig.ald], 
#  pch=19, cex=0.2, col="orange")
#points(x.s.all$rab.all[sig.s.ald], x.s.all$diff.btw[sig.s.ald], 
#  pch=19, cex=0.3, col="red")
#title(main="MA g=0.5")
#abline(h=0, lty=2)

```

### 2024-07-22 information-supplement.Rmd
focussed on dispersion version

--- 2024-05-16 supplement_figs.Rmd
@@
 ---
-title: "Supplement: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
+title: "Supplement-information: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
 "
-shorttitle: "Scale ALDEx2 Supp"
+shorttitle: "Information ALDEx2 Supp"
 author:
@@
   email: ggloor@uwo.ca
+citeproc: TRUE
+cite-method: citeproc
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
+csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
+pdf-engine: latex
 output:
-  BiocStyle::pdf_document: default
-package: ALDEx2 1.33.1
-abstract: |
-    Introduction to scale simulation and FDR correction with ALDEx2.
-date: 21 August 2023
-link-citations: true
+  pdf_document
 ---
@@
 mf.H <- Vectorize(myfun.H)
+myfun.2H <- function(x){ y <- c(x, 1-x); 2^H(y)}
+mf.2H <- Vectorize(myfun.2H)
 
@@
 myfun.G <- function(x){ y <- c(x, 1-x); G(y)}
+myfun.2G <- function(x){ y <- c(x, 1-x); 2^G(y)}
 mf.G <- Vectorize(myfun.G)
+mf.2G <- Vectorize(myfun.2G)
 
@@
 Information  is a fundamental property of all measured systems.
-For discrete probability vectors, Shannon defined their information properties and launched the field of information theory for communications.
+Shannon defined the information properties of discrete probability vectors and launched the field of information theory for communications.
 Famously, for any probability vector, Shannon set the total entropy as the inverse of the sum of the probability weighted logarithm of the probabilities.
-Less well known is that an unweighted average entropy was also defined, and that this turns out to be the inverse of the logarithm of the geometric mean of the probability vector.
+Less well known is that an unweighted average entropy was also defined.
+In the context of compositional data (CoDa) this measure is the inverse of the logarithm of the geometric mean of the probability vector.
 Thus, information theory and compositional data analysis intersect through the geometric mean of a probability vector.
@@
 We can think about  scale  from an information theoretic point of view as a measure of how much information, or total uncertainty, is encoded in a particular sample [@Shannon.48;@Jaynes:2003].
-In the geometric interpretation of information theory used in quantum information theory, formally described as  the  Asymptotic Equipartition Property of information [@Cover:1991;@wilde_2017], entropy can be interpreted as the volume occupied by a probability distribution.
-See chapters 4 of the PhD thesis of Lecamwasam [-@ruvi.blog] for a nice explanation of this.
+In the geometric interpretation of information theory used in quantum information theory, formally described as  the  Asymptotic Equipartition Property of information [@Cover:1991;@wilde_2017], entropy can be interpreted as the volume occupied by a probability distribution relative to the maximum total entropy.
+See chapters 4 of the PhD thesis of Lecamwasam [-@ruvi.blog] for a more complete explanation of this.
 
@@
 
-Thus we see that the geometric mean used in the centred log ratio (CLR), often used for Compositional Data Analysis (CoDa) [@aitchison1982] is directly related to entropy or \(H\).
+Thus we see that the geometric mean used in the centred log ratio (CLR), often used for Compositional Data Analysis (CoDa) [@aitchison1982] is  related to entropy or \(H\).
 Indeed, we can rearrange the CLR formula to show that it can be interpreted as computing the difference between the elemental information and the mean information content:
@@
 ```{r} 
-par(mfrow=c(1,1))
+par(mfrow=c(1,2))
 curve(mf.G, from=1e-2, to =.99,  col='red', lty=2, ylim=c(-3,1),
-   xlab="x in [0,1]", ylab="value")
+   xlab="x in [0,1]", ylab="value", main="H and G")
 curve(mf.H, from=1e-2, to =.99,  col='black', lty=2, add=T )
+
 legend(.4,-1.5, legend=c('H','G'), col=c('black','red'), pch="-")
+
+vals <- seq(from=.001, to=0.99, by=0.001)
+
+plot(mf.2G(vals),mf.2H(vals), xlim=c(0,0.5), ylim=c(1,2), pch=19, cex=0.1,
+  xlab="2^G", ylab="2^H", main="2^G vs 2^H")
+
 ```
@@
 Here \(X2\) consumes about 64% of the volume of system \(X1\).
-Thus, the volume is a measure of the total complexity or the scales of the two systems.
-In this way we can understand that scale is related to the information volume of a system.
+Thus, the entropic volume is a measure of the total complexity or the scales of the two systems.
+
+But what happens if we consider the geometric mean instead of the entropy? In the example above, 
+\( l2G(X1) = (4 * log2(0.25))/ 4 \) = -2, and \( l2G(X2) = (log2(\frac{7}{10}) + 3 * log2(\frac{1}{10}))/4 = -2.62 \).
+Exponentiating gives us values of 0.25 and 0.16 suggesting that the \( l2G \) measure includes some estimate of size.
+Comparing the size of \( H \) vs \(2^(l2G) \) suggests that both measures contain related information.
+Thus we can understand that scale is related to the information volume of a system, which in turn is related to the size of the system.
 
@@
 # shows differential information by group
-devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+library(ALDEx2)
 data(selex)
@@
 | $~~~$"  | B    | -12.9 | 8.3 | 0.80 | 0.977 |
-
-We can see that for most datasets the difference between the \(\overline{G}\) in each group is relatively small.
-Most significantly, the selex dataset has a very large difference of about 100-fold, and both the 16S and the metatranscriptome dataset have about a 1.5 fold difference.
-These three datasets are candidates for a full scale model correction.
-
-```{r full.scale, echo=F, fig.cap="Plot of the offset of the mean differnece between groups as a function of scale ratio.
-For this, the default scale of 1:1 was altered in increments of 0.1 keeping the gamma parameter (dispersion) at 0.5.
-The filled circle shows the outcome when the calculation is done using the geometric mean and the same gamma parameter."}
-
-load('analysis/data.out.Rda' )
-
-plot(data.out[,6],data.out[,1], pch=c(rep(1,19),19), ylim=c(-11, 4),
-  xlab='full scale offset', ylab='mean difference between groups')
-points(data.out[,6],data.out[,2], pch=c(rep(1,19),19), col='red')
-points(data.out[,6],data.out[,3], pch=c(rep(1,19),19), col='orange')
-points(data.out[,6],data.out[,4], pch=c(rep(1,19),19), col='cyan')
-points(data.out[,6],data.out[,5], pch=c(rep(1,19),19), col='blue')
-abline(h=0, col='grey', lty=2)
-abline(v=0, col='grey', lty=2)
-legend(0.7, 0, legend=c('16S','selex','tscome','metatscome','ss'),
-  pch=19, col=c('black','red','orange','cyan','blue'))
-
-```
-
-Nixon et al.
-[-@nixon2023scale] showed that many operations on HTS datasets relied on both the proportion and scale components of the data.
-Moreover, all normalizations impose a scale model on the data, but the appropriateness of these models has never been explicitly acknowledged or tested.
-Thus, the full scale model option allows the investigator to set both the offset between the geometric means of the features and their dispersion and observe how this affects the analysis outcome.
-
-In the offset plot above we can see the cause and the effect of the full model with a fixed gamma of 0.5 and a base scale of 1 in each group.
-We see that the mean location of well centred datasets (yeast transcriptome, single-cell transcriptome) are close to 0, but could be centred better with small changes in scale ranging from 0 (single cell) to 1:1.1 for the yeast transcriptome dataset.
-In contrast, centring the 16S dataset requires about a 1:1.3 fold change.
-The metatranscriptome dataset would require about a 0.5:1 change in relative scale between groups, but as shown in the main text, centring the housekeeping genes is more apt.
-
-The in vitro selection dataset is clearly an outlier in both the difference between the average group geometric mean, and in the offset plot.
-However, this dataset can  be used to illustrate the power of the full scale model and the relationship between \( G_n \) and scale.
-In Figure 3 we can see that the default output of ALDEx2 has a centered output.
-This occurs largely by chance, as the high sparsity of the selected (S) group is balanced almost exactly by the arbitrarily chosen sequencing depth so the non-selected group (NS) appears to have a similar location as the S group.
-The difference in entropy between the two groups and the differences in geometric mean are very large, with the difference in \(\log_2{\overline{G}(S)}\) and \(log_2{\overline{G}(NS)}\) being about \(2^{6.6}\).
-Setting the scale of both the S and NS groups to 1 we find that the difference in location is approximately \(2^{7.7}\) in close agreement with the difference the geometric means.
-For this dataset to be centered we need to have a scale ratio \( \approx \) 1:50 or more.
-Note that the scale ratio is inverse to the ratio of geometric means as described above.
-In fact, in this dataset the relative abundances of the majority of features are nearly invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014], thus changing the scale of the data.
-Neither DESeq nor edgeR are able to provide a reasonable analysis of this dataset because the normalizations used assume equivalent scales [@gloorAJS:2016].
-
-Figure 3 shows an effect plot of various scale models with this dataset.
-The full scale model, where the strong assumption that the mean \(G\) is assumed to be 1:1 between the two groups, dramatically skews the output and the large number of relatively invariant features are now identified as significantly different.
-While not wrong as long as the assumption that the scales are identical is stated, this is not a useful analysis outcome.
-Modifying the mean scale difference between the NS:S groups to be \( \approx \) 1:50 different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained using \(G_n\) as the scale estimate where the ratio is \( \sim 100:1 \).
-Note that we get exactly the same answer (within random sampling error) with a scale of .02 for group NS and a scale of 1 for group S, or using a scale of 1 for group NS and a scale of 50 for group S.
-This shows that it is the relative difference between scales that is important in this dataset, not the absolute values.
-From this result we can conclude that, on average, the difference in underlying scale in the system is about \(\approx 50\)-fold, and this is congruent with the circa 100-fold difference in  \(\overline{G}\) between groups; the discrepancy being explained because the default scale model is applied uniformly to all samples, whereas the different values of the within-group geometric means ranging over a \{ > 2.5 \) fold range.
-Thus, an advantage of a full scale model is that we have gained both information and understanding about the drivers of asymmetry underlying system.
-
-```{r selex, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters.
-All scales are calculated with a logNormal distribution to ensure symmetry for the user.
-"}
-
-conds <- c(rep('NS',7), rep('S',7))
-
-x.all <- aldex(selex, conditions=conds, CI=T, gamma=0.5)
-
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
-x.1.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio as 1:50 explicitly
-#sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
-scale <- c(rep(1,7), rep(50,7)) 
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=FALSE)
-x.5.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio in log space
-# 2^-5.64 : 2^0  = 0.02 : 1
-# this also works, but reversed because of factor()  
-# sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(0,-5.64), conds, log=TRUE)
-scale <- c(rep(-5.6,7), rep(0,7))
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=TRUE)
-
-x.2.all <- aldex(selex,conditions=conds, CI=T, gamma=sel.mu)
-
-par(mfrow=c(1,4))
-aldex.plot(x.all, main=('gamma=0.5\nmu=G'))
-aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
-aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,50'))
-aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.02, 1'))
-```
-
-# How scaling affects dispersion
-
-Thus far we have observed that adding scale uncertainty increases the minimum dispersion value with little effect on the difference between values as in panel A.
-Panel B shows the change in dispersion relative to the rAbundance of the features.
-Here we can see that the minimum dispersion is increased as gamma increases.
-The horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5.
-Panel C shows that this increase is non-linear, being more pronounced among those features that had minimal dispersion when gamma=0.
-Panel D shows that increasing gamma actually shrinks the dispersion estimates towards the mean dispersion, while increasing the total dispersion as shown in panel C.
-
-Thus, adding scale uncertainty has two effects both on dispersion.
-The first is to increase the dispersion estimate for each part and this has its primary effect to reduce the  p-values and standardized effect sizes returned by ALDEx2.
-The second is to reduce the range of dispersions such that the parts with the lowest dispersions have the greatest increase.
-This increase in dispersion comes about because the underlying distributions are widened because of the additional uncertainty added by the inclusion of scale uncertainty.
-
-```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="jnk"}
-yst.conds <- c(rep('S',44),rep('W',42))
-x <- aldex.clr(yst, yst.conds, gamma=NULL)
-xs <- aldex.clr(yst, yst.conds, gamma=0.5)
-x1 <- aldex.clr(yst, yst.conds, gamma=1)
-
-x.e <- aldex.effect(x)
-xs.e <- aldex.effect(xs)
-x1.e <- aldex.effect(x1)
-
-par(mfrow=c(2,2))
-# we imperceptibly change difference
-plot(x.e$diff.btw, x1.e$diff.btw, xlab='diff g=0', ylab='diff g=1')
-abline(0,1, col='grey', lwd=2, lty=2)
-title(main='A', line=-1.2, adj=0.5)
-#=-1
-cuts <- x.e$rab.all > -0.5 & x.e$rab.all < 0.5
-# the residual of the 
-mn.mid <- median(x.e$diff.win[cuts])
-mn.mid.s <- median(xs.e$diff.win[cuts])
-mn.mid.1 <- median(x1.e$diff.win[cuts])
-
-# we change dispersion a lot for the bulk of the features
-plot(x.e$rab.all, x.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
-  xlab='rAbundance', ylab='dispersion')
-points(xs.e$rab.all, xs.e$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
-points(x1.e$rab.all, x1.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3))
-abline(h=mn.mid, lty=2)
-abline(h=mn.mid.s, lty=2, col='red')
-abline(h=mn.mid.1, lty=2, col='blue')
-title(main='B', line=-1.2, adj=0.5)
-
-plot(x.e$diff.win, x1.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3), ylim=c(0,4.5), 
-  xlab='g=0 dispersion', ylab='g=0.5 | g=1 dispersion')
-points(x.e$diff.win, xs.e$diff.win, pch=19, cex=0.5, col=rgb(1,.4,.1,0.3))
-abline(0,1, lty=2)
-title(main='C', line=-1.2, adj=0.5)
-
-strip.data <- list(x.e$diff.win[cuts] - mn.mid, xs.e$diff.win[cuts] - mn.mid.s, x1.e$diff.win[cuts] - mn.mid.1)
-
-stripchart( strip.data, col=c(rgb(0,0,0,0.3), rgb(1,0,0,0.3), rgb(0,0,1,0.3)), vertical=T, method='jitter', pch=19, group.names=c('g=0', 'g=0.5', 'g=1'), ylab='residual dispersion')
-title(main='D', line=-1.2, adj=0.5)
-
-```
-
-# Issues with DESeq2 and edgeR
-
-DESeq2 and edgeR are two of the most commonly used tools for differential abundance analysis of bule RNA sequencing datasets.
-They both operate by finding a scaling factor that makes all the samples commesurate.
-DESeq2 does this by finding a midpoint feature that can be used as a reference in each sample; this can be different for different samples.
-The edgeR reference finds the midpoint of the 'typical' sample instead.
-In both cases the data are then scaled by dividing by a small factor that makes the read counts commesurate.
-Differential abundance analysis is then performed on the scaled values after taking their logarithm to base 2.
-In some ways this is similar to the log-ratio approach used by ALDEx2, but is more prone to dataset and sample effects than is the log-ratio method [@GloorAJS2023]
-
-```{r DESedg, eval=T, echo=F, fig.cap="Shown here are the mean log2 fold change as a density plot, and a Volcano plot showing the location and adjusted p-value for each feature in the metatranscriptomic dataset.
-The DESeq2 approach does a good job of centring this data, while edgeR is less suitable.
-The volcano plots show dramatically different outcomes.
-The DESeq2 algorithm assigns very large fold changes to features that have only moderate change, and further identifies a very large proportion of features as significantly different.
-In contrast, edgeR exhibits a much smaller number of differentially abundant features.
-In both volcano plots, the housekeeping genes in the main Figure 3 are shown in orange.
-We can see that these are asymmetrically distributed in both plots.
-Additionally the location of the features taht DESeq2 identified as having a very large difference are shown in the edgeR volcano plot as blue circles."}
-
-# find housekeeping from ALDEx2
-
-load('analysis/xt.m.Rda')
-load('analysis/xt.Rda')
-hk.off <- xt.all$diff.win < 5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
-
-load('analysis/meta.DES.res.Rda')
-load('analysis/meta.edge.qlf.Rda')
-
-# these are all very rare K0s
-zero.var <-  which(meta.DES.res@listData$lfcSE*sqrt(44) == 0)
-
-weird.des <- which(meta.DES.res@listData$log2FoldChange < -20 )
-
-# housekeeping genes centred with DESeq2 - nice
-
-sig.des <- which(meta.DES.res@listData$padj < 0.01)
-sig.edge <- which(p.adjust(meta.edg.qlf$PValue) < 0.01)
-sig.ald <- which(xt.m.all$we.eBH < 0.01)
-
-par(mfrow=c(2,3))
-
-plot(density(xt.all$diff.btw), xlim=c(-5,5), ylim=c(0,1), main='ALDEx2 FC')
-points(density(xt.all$diff.btw[hk.off]), type='l', col='orange')
-abline(v=0, col='orange', lty=2)
-
-plot(density(meta.DES.res@listData$log2FoldChange), xlim=c(-5,5), ylim=c(0,0.8), main='DESeq2 FC')
-points(density(meta.DES.res@listData$log2FoldChange[hk.off]), type='l', col='orange')
-abline(v=0, col='orange', lty=2)
-
-plot(density(meta.edg.qlf$logFC), xlim=c(-5,5), ylim=c(0,0.5), main='edgeR FC')
-points(density(meta.edg.qlf$logFC[hk.off]), type='l', col='orange')
-abline(v=0, col='red', lty=2)
-
-plot(xt.m.all$diff.btw, -1*log10(xt.m.all$we.eBH), 
-  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)', main='ALDEx2 volcano')
-points(xt.m.all$diff.btw[sig.ald],-1*log10(xt.m.all$we.eBH)[sig.ald] + 1e-8, col='red')
-points(xt.m.all$diff.btw[hk.off],-1*log10(xt.m.all$we.eBH)[hk.off] + 1e-8, col='orange', cex=0.6)
-
-plot(meta.DES.res@listData$log2FoldChange, -1*log10(meta.DES.res@listData$padj), 
-  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)', main='DESeq2 volcano')
-points(meta.DES.res@listData$log2FoldChange[sig.des],-1*log10(meta.DES.res@listData$padj[sig.des] + 1e-70), col='red')
-points(meta.DES.res@listData$log2FoldChange[hk.off],-1*log10(meta.DES.res@listData$padj[hk.off] + 1e-70), col='orange', cex=0.6)
-
-plot(meta.edg.qlf$logFC, -1*log10(p.adjust(meta.edg.qlf$PValue)), 
-  col=rgb(0,0,0,0.1), xlab='LFC', ylab='-1log10(p)', main='edgeR volcano')
-points(meta.edg.qlf$logFC[sig.edge], -1*log10(p.adjust(meta.edg.qlf$PValue)[sig.edge]), 
-  col='red')
-points(meta.edg.qlf$logFC[hk.off], -1*log10(p.adjust(meta.edg.qlf$PValue)[hk.off]), 
-  col='orange', cex=0.6)
-  points(meta.edg.qlf$logFC[weird.des], -1*log10(p.adjust(meta.edg.qlf$PValue)[weird.des]), 
-  col='blue', cex=0.6)
-
-```
-
-Examining the plots, edge R clearly not centred with the median housekeeping functions offset by `r median(meta.edg.qlf$logFC[hk.off==T])` and minimum FDR value  is `r min(p.adjust(meta.edg.qlf$PValue)[hk.off==T])`.
-Additionally, as shown in Figure 4there is little range in the p-values relative to those seen for DESeq2, and the values are more in line with the range seen with ALDEx2.
-
-DESeq2 is better centred with median housekeeping functions offset by `r median(meta.DES.res@listData$log2FoldChange[hk.off==T])` but the minimum FDR value  is `r min(meta.DES.res@listData$padj[hk.off==T])`.
-In addition there are a large number of functions with 0 variance, these are very low count functions with very high sparsity in the dataset.
-These are not differential in the DESeq2 analysis.
-However, there are a number of functions in the DESeq2 analysis that are very differentially abundant, with a log2 fold change of < -20.
-Examination of the raw counts shows that these are uniformly 0 or 1 in the H dataset, and present at high counts in some, but not all of the BV dataset.
-That these stand out is odd, since inspection of the count table shows that there are many other functions that are missing from the H dataset, and have higher and more uniform counts in the BV dataset.
-Thus, the importance of the outlier functions seen in the DESeq2 volcano plot should be viewed with suspicion and may be an artefact of the normalization used.
-When overplotted on the edgeR volcano plot (blue), it is clear that these functions have non-signficant p-values.
-
-# Checking the scale assumptions of the RLE and TMM normalizations
-
-We can show that the normalizations built into the edgeR Bioconductor package (RLE, TMM, TMMwsp, upperquantile) are scale assumptions by using the normalization factor as an input to `aldex.makeScaleMatrix()` and then measure the mean location of the data as above.
-The ideal behavior of a normalization is that the mean location of the data should be close to 0 or unchanged.
-This behavior will ensure that Type 1 and Type 2 errors due to scale assumptions are minimized.
-
-The results, shown in the tables below compare these normalizations to the no scale assumption (iso) and the geometric mean normalization (GM) assuming that the normalizations are either on a log scale or a linear scale.
-That these affect scale can be observed by their effect on the mean location of the data.
-For the most part assuming no scale differences between groups centres the date less well than does the geometric mean.
-In most cases the other normalizations perform poorer than assuming no scale at all for the rRNA dataset and about as well as the no scale assumption in the metatranscriptome datasets and the single-cell transcriptome dataset.
-All normalizations perform poorly in the selex dataset.
-Overall, these results may not be surprising because the TMM and RLE (and related normalizations) were developed specifically to scale and normalize transcriptome datasets [@Robinson:2010a;@Anders:2010], but have become widely used in the analysis of other data modalities; e.g.
-[@McMurdie:2014a].
-
-```{r getNorm, echo=F}
-# output from code/scale_norms.R
-
-load('analysis/normalizations.Rda')
-knitr::kable(norm.data.out, 'simple', caption='Log scale models')
-```
-
-```{r getNormNL, echo=F}
-# output from code/scale_norms.R
-
-load('analysis/normalizations.nl.Rda')
-knitr::kable(norm.nl.data.out, 'simple', caption='Linear scale models')
-```
-
-# References

### 2024-07-22 supplement_figs.Rmd
focussed on dispersion version

--- 2023-10-17 supplement_figs.Rmd
@@
   email: ggloor@uwo.ca
+citeproc: TRUE
+cite-method: citeproc
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
+csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
+pdf-engine: latex
 output:
-  BiocStyle::pdf_document: default
-package: ALDEx2 1.33.1
-abstract: |
-    Introduction to scale simulation and FDR correction with ALDEx2.
-date: 21 August 2023
-link-citations: true
+  pdf_document
 ---
 
-# GM is related to Information and Shannon's entropy in HTS datasets
+```{r echo=F, warning=F, message=F}
+library(ALDEx2)
+data(selex)
+```
 
-Information content is a fundamental property of all systems.
-We can think about  scale  from an information theoretic point of view as a measure of how much information is encoded in a particular sample.
-Intuitively,   systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
-Below, we show that the scale of a system as defined by Nixon et al.
-[-@nixon2023scale] is inversely related to the information content and entropy of the data.
+One root cause of the large number of significant parts is the very low dispersion when the data are normalized and log-transformed.
+The raw data counts derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010a;@Robinson:2010] as seen in Panel A of Figure 1.
+However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010;@Love:2014aa;@Andrews:2019aa] which has a very different abundance-dispersion relationship.
+Similar relationships are found in both DESeq2 (Panel B) and ALDEx2 (Panel C), and have been  noted before [@fernandes:2013;@Love:2014aa].
 
-Recall the underlying system is described by a \(D \times N\) matrix of counts \(\mathbf{W}\) decomposed into the proportions for the \(n^{th}\) sample \(\mathbf{W}^{\parallel}_n\) (or the equivalent probability distribution \( \mathbf{p}(w_n) \) ), and its scale \(\mathbf{W}^{\perp}_n\), such that \(\mathbf{W}=\mathbf{W}^{\parallel}\mathbf{W}^{\perp} \).
-Sequencing  returns counts which are related to the underlying proportion; i.e., \(\mathbf{Y}^{\parallel}_n \sim \mathbf{W}^{\parallel}_n\)
+```{r, echo=F, warning=F, message=F,comment=F, result=F} 
+library(DESeq2)
+library(ALDEx2)
+```
 
-For notational simplicity assume we will use a single discrete random variable to represent the probability distribution; i.e.
-\( X =  \mathbf{p}(w_n) \)  over \(1 \dots d \) elements.
-The geometric mean can be calculated as:
+```{r echo=F, warning=F, message=F,comment=F,, fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis.
+DESeq2 was used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset.
+The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the Benjamini-Hochberg corrected p-values (volcano plot).
+Each point represents the values for one transcript, with the color indicating if that transcript was significant in the ALDEx2 scaled analysis and the DESeq2 analysis (red) or in the DESeq2 unscaled analysis only (orange).
+Points in grey are not statistically signficantly different with any analysis.
+The horizontal dashed lines represent a log2(difference) of \\(\\pm 1.4\\)"}
+# DESeq2
+load('analysis/res.Rda')
 
-\[
-\log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
-\];
+#ALDEx2
+load(file="analysis/yst.all.Rda")
+load(file="analysis/yst.s.all.Rda")
+load(file="analysis/yst.1.all.Rda")
+# with gamma = 0.5
+load('analysis/x.s.all.Rda')
 
-here every element in \(log_2 G(X)\) is the logarithm of the elemental probability \( log_2(p_i)\).
+sig.des <- which(res@listData$padj < 0.05)
+sig.ald <- which(yst.all$we.eBH < 0.05)
+sig.s.ald <- which(yst.s.all$we.eBH < 0.05)
 
-In information theory, the elemental amount of information or uncertainty for \(p_i\) is the inverse \(I(X)_i = -log_2(p_i)\)[@rezaInfoIntro].
+sig.all <- yst.all$we.eBH < 0.05
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
+sig.s <- yst.s.all$we.eBH < 0.05
+sig.1 <- yst.1.all$we.eBH < 0.05
 
-The total entropy of the system  \( H(x) \) is the weighted sum of the elemental information
+par(mfrow=c(1,2))
+plot(res@listData$lfcSE, res@listData$log2FoldChange, xlim=c(0,1), 
+  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
+title('A: DESeq2 effect', adj=0, line= 0.8)
+points(res@listData$lfcSE[sig.des],
+  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
+  pch=19, cex=0.5)
+points(res@listData$lfcSE[sig.s.ald],
+  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
+  pch=19, cex=0.5)
+abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-\[
-H(x) = -\sum_{i=1}^{d} p_i \log_2 p_i
-\]
+# volcano
+plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
+  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
+title('B: DESeq2 volcano', adj=0, line= 0.8)
+points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
+  pch=19, cex=0.5)
+points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
+  pch=19, cex=0.5)
+  abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-In this formulation, \( \log_2(G(X)) \) and \( H(X) \) can be understood  to be an unweighted and a weighted measure  of 'surprisal' in the dataset.
-The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.
-\( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \).
-Conversely, \( G \) is an unweighted mean of the same distribution and  \( G \) is the denominator to calculate the centred log-ratio normalization (CLR).
-Note that when calculating the CLR we are taking a ratio, and similarly for calculating entropy, but that the scale formula takes the product.
-Thus, the scale of a system is inversely related to the other two measures.
+```
 
-Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(\mathbf{Y}^{\parallel}_{n})\) (Supplemental Figure 1 and Table 1).
-These two measures are expected to have different behaviours with different distributions of \( p_i \).
-In the case of a uniform distribution both \( H(\mathbf{Y}^{\parallel}_{n}) \) and \( G(\mathbf{Y}^{\parallel}_{n}) \) are maximal since \( p(x) \) is equally and identically distributed.
-Thus, we expect that they are positively correlated here.
-In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values.
-In very sparse  datasets, the two measures could become uncoupled because \( H(\mathbf{Y}^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(\mathbf{Y}^{\parallel}_{n}) \) would tend to be very small.
-Here these two measures could be either uncorrelated or exhibit negative correlation.
-We can see this distributional behaviour in different datasets.
+```{r echo=FALSE, include=TRUE, out.width="0.8\\linewidth",  fig.align="center", fig.cap=c("the aldex.senAnalysis() function was used to generate a dataset with \\(\\gamma\\) values of 1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7 and 1.
+The plotGamma() function was used to plot the result.
+Transcripts that are statistically significant are shown in black, and if not significant are in grey.
+The \\(\\gamma\\) values, standardized effect size and number of significant transcripts at each value are given on the axes.
+")}
+knitr::include_graphics("analysis/scaleSim-yeast.pdf")
+## plotGamma needs to be cleaned up for production
+## so this figure was done by hand with the example
+## snippet below - ugly sorry
+#load('analysis/yst.sens.Rda')
+#plotGamma(x.sens, blackWhite=T)
+#text(6.2,345, label=length(which(x.sens[[8]]$we.eBH < 0.05)))
+#text(-2,345, label=length(which(x.sens[[7]]$we.eBH < 0.05)))
+#text(-6.9,345, label=length(which(x.sens[[6]]$we.eBH < 0.05)))
+#text(-9.6,345, label=length(which(x.sens[[5]]$we.eBH < 0.05)))
+#text(-12.2,345, label=length(which(x.sens[[4]]$we.eBH < 0.05)))
+#text(-14.8,345, label=length(which(x.sens[[3]]$we.eBH < 0.05)))
+#text(-17.4,345, label=length(which(x.sens[[2]]$we.eBH < 0.05)))
+#text(-20,345, label=length(which(x.sens[[1]]$we.eBH < 0.05)))
+#text(-6.9,355, label="N significant")```
 
-```{r info, echo=F,warning=F, message=F,comment=F, fig.cap="Plot of Shannon's entropy (H) vs geometric mean (G) for each sample in different datasets.
-The groups that each sample belong to are highlighted as filled or open circles.
-Each group in each dataset has different entropy with the groups in the selex and metatranscriptome datasets being highly distinct."}
-# Gm ~ Im
-# plotting geometric mean vs Shannon's entropy
-# shows differential information by group
-devtools::load_all('~/Documents/0_git/ALDEx_bioc')
-data(selex)
+```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T,fig.dim=c(7,3), fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as CLR values.
+Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale.
+Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first.
+Panel C shows the equivalent values calculated by ALDEx2	 in which the expected CLR value for each transcipt are plotted vs.
+the expected dispersion.
+Panel D shows the output for ALDEx2 with \\( \\gamma=0.5 \\).
+The red line in each panel shows the LOESS line of fit to the mid-point of the distributions.
+In panels B and C the amount of dispersion reaches a minimum at moderate	 values.
+The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value.
+The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }
 
-load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
 url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
 yst <- read.table(url, header=T, row.names=1)
-url <- "https://raw.githubusercontent.com/ggloor/datasets/main/meta16S.tsv"
-rRNA <- read.table(url, header=T, row.names=1, sep='\t')
-url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
-ss <- read.table(url, header=T, row.names=1, sep='\t')
-ss <- ss[,c(1:100,1502:1601)]
 # remove the one gene with 0 reads
+
 yst <- yst[rownames(yst) != "YOR072W-B",]
+
 # Gierlinski:2015aa
@@
 
-HS <- apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GS <- apply(selex+0.5, 2, function(x) mean(log2(x/sum(x))))
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
 
-HY <- apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GY <- apply(yst+0.5, 2, function(x) mean(log2(x/sum(x))))
+# DESeq2
+load('analysis/res.Rda')
 
-HM <- apply(ko.both+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GM <- apply(ko.both+0.5, 2, function(x) mean(log2(x/sum(x))))
+#ALDEx2
+load('analysis/x.all.Rda')
+load('analysis/x.s.all.Rda')
 
-H16 <- apply(rRNA+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-G16 <- apply(rRNA+0.5, 2, function(x) mean(log2(x/sum(x))))
+# with gamma = 1
+load('analysis/x.s.all.Rda')
+sig.des <- which(res@listData$padj < 0.05)
+sig.ald <- which(x.all$we.eBH < 0.05)
+sig.s.ald <- which(x.s.all$we.eBH < 0.05)
 
-Hss <- apply(ss+0.1, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-Gss <- apply(ss+0.1, 2, function(x) mean(log2(x/sum(x))))
+# get and plot mean/var or mean/SE for counts
+# and log counts
+yst.mn <- apply(yst, 1, mean)
+yst.mn.log <- apply(yst, 1, function(x) mean(log2(x)))
+yst.mn.log[is.infinite(yst.mn.log)] <- 0
 
-par(mfrow=c(2,3))
-plot(HS, GS, ylab="Geometric mean", xlab="Shannon's entropy",
-  pch=c(rep(19,7),rep(1,7)), col=rgb(0,0,0,0.5))
-title(main='selex')
+yst.v <- apply(yst, 1, var)
 
-plot(HY,GY, 
-  ylab="Geometric mean", xlab="Shannon's entropy", pch=c(rep(19,44),rep(1,42)),
-  col=rgb(1,0.6,0,0.5))
-title(main='transcriptome')
+par(mfrow=c(1,4))
+plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='Mean count', ylab='Variance')
+#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
+def <- data.frame(log2(yst.mn),log2(yst.v))
+lines(lowess(def, f=0.1), col=2, lwd=2)
+title('A', adj=0, line= 0.8)
+abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 
-plot(HM, GM, col=rgb(1,0,0,0.5), pch=c(rep(1,8), rep(19,28), rep(1,8)), 
-  ylab="Geometric mean", xlab="Shannon's entropy")
-title(main='meta-transcriptome')
+l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
+plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
+lines(lowess(l.df, f=0.25), col=2, lwd=2)
+abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.0275, col='grey', lty=2)
+title('B', adj=0, line= 0.8)
 
-plot(H16,G16, col=rgb(0,1,1,0.5),
-  pch=c(rep(1,198), rep(19,161)), 
-  ylab="Geometric mean", xlab="Shannon's entropy")
-title(main='16S rRNA')
+l.df <- data.frame(x.all$rab.all, x.all$diff.win)
+plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='E(clr)', ylab='E(disp)')
+#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
+lines(lowess(l.df, f=.1), col=2, lwd=2)
+abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.208, col='grey', lty=2)
+title('C', adj=0, line= 0.8)
 
-plot(Hss,Gss, col=rgb(0.5,1,0,0.5),
- pch=c(rep(1,100), rep(19,100)), 
-  ylab="Geometric mean", xlab="Shannon's entropy")
-title(main="single cell")
-
-# empty plot for legend
-plot(selex[,1], selex[,2], xlab=NA, ylab=NA, axes=F,xlim=c(-10,-1), ylim=c(-10,-1))
-legend(-10,-1, legend=c('selex', 'transcriptome','metatranscriptome', '16S', 'SS', 'low entropy group', 'high entropy group'),
-  col=c('darkgrey', 'orange', 'red', 'cyan', 'greenyellow','grey','grey'), pch=c(19,19,19,19,19,1,19))
-
+l.s.df <- data.frame(x.s.all$rab.all, x.s.all$diff.win)
+plot(x.s.all$rab.all, x.s.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='E(clr)', ylab='E(disp)', ylim=c(0,4.5))
+#points(x.s.all$rab.all[sig.ald], x.s.all$diff.win[sig.ald], col='orange', cex=0.4)
+lines(lowess(l.s.df, f=.1), col=2, lwd=2)
+abline(h=min(lowess(l.s.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.208, col='grey', lty=2)
+title('D', adj=0, line= 0.8)
 ```
 
-The table below summarizes the mean values for, and the correlation between, \( G \) and \( H \)  (cor)  and the sparsity defined as the proportion of features with less than 1 count per sample (spar) for each association in each group of samples:
+The variance or dispersion always increases with increasing raw (or normalized) read count but decreases when measured on the log-ratio transformed data [@fernandes:2013;@Love:2014aa], reaching a minimum at some mid-point of the distribution.
+This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes.
+This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa].
+Furthermore, the dispersion is exceeding small being, for many transcripts, almost negligible.
+To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead to many false positives.
+Indeed, benchmarking and comparison studies  repeatedly show that the choice of normalization plays a role in which results are returned as significant [@maza2013;@Dillies:2013;@Weiss:2017aa].
+From the perspective of Nixon et al.
+[-@nixon2023scale], it is reasonable to conclude that some of these results may be due to the choice of normalization.
 
-| Dataset | group | \(\overline{G}\) | \( \overline{H} \) | cor | spar |
-|---------|-------|---|---|-----|-------|
-| Selex   | control | -11.2 | 10.2 | 0.99 | 0 |
-| $~~~$"  | selected | -17.8 | 2.8 | -0.88 | 0.802 |
-| yeast   | snf2 ko  | -14.0 | 10.7 | 0.99 | 0.004 |
-| $~~~$"  | WT    | -14.2 | 10.4 | 0.99 | 0.007 |
-| Meta    | H     | -18.8 | 8.6 | 0.78 | 0.451 |
-| $~~~$"  | BV   | -18.2 | 8.9 | 0.79 | 0.238 |
-| 16S     | Pup   | -14.7 | 5.4 | 0.68 | 0.079 |
-| $~~~$"  | Cent | -15.2 | 5.4 | 0.53 | 0.251 |
-| SS      | A     | -13.0 | 8.2 | 0.83 | 0.978 |
-| $~~~$"  | B    | -12.9 | 8.3 | 0.80 | 0.977 |
+While not necessary in all datasets (e.g., the transcriptome example discussed in the previous section and see the Supplementary material), it can greatly improve modeling in certain cases, especially when the scale is highly asymmetric between conditions.
+In order to specify a scale model from scratch, we need to revisit the concept that all normalizations in widespread use are actually ratios with the denominator implied by the normalization.
+Therefore, we can easily deviate from a certain normalization (e.g., the geometric mean assumption implied by the CLR) by specifying a total model based on the mean difference between conditions.
+While knowing the mean difference between conditions may seem cumbersome in practice, it is the \emph{relationship} between the group scale values that is important, not their raw values (Nixon et al.
+2023).
+This can be illustrated quite simply by starting with the mean ratio in \(G\) between groups for the yeast transcriptome dataset which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for WT; their ratio being 1.17, or a 0.17-fold difference.
+Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental Figure 2.
+This ratio can be adjusted to alter the mean assumption placed on the group scale values.
 
@@
 
-conds <- c(rep('NS',7), rep('S',7))
-
-x.all <- aldex(selex, conditions=conds, CI=T, gamma=0.5)
-
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
-x.1.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio as 1:50 explicitly
-#sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
-scale <- c(rep(1,7), rep(50,7)) 
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=FALSE)
-x.5.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio in log space
-# 2^-5.64 : 2^0  = 0.02 : 1
-# this also works, but reversed because of factor()  
-# sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(0,-5.64), conds, log=TRUE)
-scale <- c(rep(-5.6,7), rep(0,7))
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=TRUE)
-
-x.2.all <- aldex(selex,conditions=conds, CI=T, gamma=sel.mu)
+# from selex.aldex.mu.R
+load(file="analysis/sel.2.all.Rda")
+load(file="analysis/sel.5.all.Rda")
+load(file="analysis/sel.1.all.Rda")
+load(file="analysis/sel.all.Rda")
 
 par(mfrow=c(1,4))
-aldex.plot(x.all, main=('gamma=0.5\nmu=G'))
-aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
-aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,50'))
-aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.02, 1'))
+aldex.plot(sel.all, main=('gamma=0.5\nmu=G'))
+aldex.plot(sel.1.all, main=('gamma=0.5\nmu=1,1'))
+aldex.plot(sel.5.all, main=('gamma=0.5\nmu=1,50'))
+aldex.plot(sel.2.all, main=('gamma=0.5\nmu=0.02, 1'))
+```
+
+# How scaling affects dispersion
+
+```{r, echo=F}
+load(file="analysis/yst.all.Rda")
+load(file="analysis/yst.s.all.Rda")
+load(file="analysis/yst.1.all.Rda")
 ```
@@
 We can see that these are asymmetrically distributed in both plots.
-Additionally the location of the features taht DESeq2 identified as having a very large difference are shown in the edgeR volcano plot as blue circles."}
+Additionally the location of the features that DESeq2 identified as having a very large difference are shown in the edgeR volcano plot as blue circles."}
 
@@
 load('analysis/xt.Rda')
-hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
+load('analysis/xg.Rda')
+hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3]
+hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3
 
@@
 When overplotted on the edgeR volcano plot (blue), it is clear that these functions have non-signficant p-values.
+
+```{r eval=T, echo=F, fig.cap="Shown here are two volcano plots that plot the mean log2 fold change plotted vs the log of the FDR for the yeast transcriptome dataset and for the metatranscriptome dataset.
+Data were generated in ALDEx2 with \\(\\gamma = [0, 0.5] \\) in black or in red."}
+
+par(mfrow=c(1,2))
+plot(yst.all$diff.btw, -log10(yst.all$we.eBH +1e-70), pch=19, cex=0.5, col=rgb(0,0,0,0.3), 
+  ylab='-log10(FDR)', xlab="log2(diff)", main='yeast')
+points(yst.s.all$diff.btw, -log10(yst.s.all$we.eBH + 1e-20), pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+legend(4,60, legend=c("gamma=0", "gamma=0.5"), pch=19, cex=0.5, col=c("black", "red"))
+
+plot(xt.all$diff.btw, -log10(xt.all$we.eBH), pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  ylab='-log10(FDR)', xlab="log2(diff)", main='meta')
+points(xg.all$diff.btw, -log10(xg.all$we.eBH), pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+
+```
+
+Adding a small amount of scale, \(\gamma=0.5\), has a major effect on the volcano plot for the yeast transcriptome dataset, but mimimal effect on the vaginal metatranscriptome dataset.
+In both datasets the FDR values are raised, but no effect is observed on the difference between values.
+However, the concordance between the FDR values and the difference between become very tight for the yeast transcriptome dataset, but the concordance is not visibly unchanged for the other dataset.
+This is likely because both the difference between and the dispersion were both very small in the former and substantially larger in the latter.
 
@@
 
+Thus far we have observed that adding scale uncertainty has a negligible effect on either the differences between groups or the relative abundance measure with a correlation of `r round(cor(yst.all$diff.btw, yst.1.all$diff.btw), 3)` and `r round(cor(yst.all$rab.all, yst.1.all$rab.all), 3)`.
+The effect is entirely restricted to the dispersion estimats as shown below.
+Panel A shows the change in dispersion relative to the rAbundance of the features.
+Here we can see that the minimum dispersion is increased as gamma increases.
+The horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5.
+Panel B shows that this increase is non-linear, being more pronounced among those features that had minimal dispersion when gamma=0.
+Panel C shows that increasing gamma rotates the dispersion estimates at high relative abundances, such that housekeeping genes no longer have larger mean dispersions (rAB > 5) than do regulated genes (rAB >0, < 5).
+Overplotted are the lowess lines of fit through the densities of transcripts that are judged as significantly different using either scaled or unscaled data, and with or without thresholding.
+Statistical significance was determined with a FDR < 0.05.
+
+Thus, adding scale uncertainty has two effects both on dispersion.
+The first is to increase the dispersion estimate for each part and this has its primary effect to reduce the  p-values and standardized effect sizes returned by ALDEx2.
+The second is to reduce the range of dispersions, and rotate them such that the parts with the lowest dispersions have the greatest increase.
+This increase in dispersion comes about because the underlying distributions are widened with the inclusion of scale uncertainty.
+
+```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="Adding scale uncertainty changes the dispersion distribution.
+Panel A shows a plot of the expected value for relative abundance vs the expected value for the pooled dispersion as output by aldex.effect.
+The dashed  horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5.
+Panel B plots the unscaled vs scaled dispersion, note the non-linear relationship.
+Panel C plots relative abundance vs the  difference between dispersion with gamma set to 0 and to 1 to highlight the rotation that is evident in Panel A.
+The colored lines indicate the lowess line of fit through the centre of mass of the plot.
+Line colors represent different populations of points.
+The grey line is the total population, the red line is the population of significant transcripts with no scale, the orange line is the population of significant transcripts with a difference threshold of about 2.5-fold change, the blue line is the population of significant transcripts with gamma =0.5, and the cyan line is the significant population with gamma = 1.
+"}
+
+cuts <- yst.all$rab.all > -0.5 & yst.all$rab.all < 5
+par(mfrow=c(1,3))
+mn.mid <- median(yst.all$diff.win[cuts])
+mn.mid.s <- median(yst.s.all$diff.win[cuts])
+mn.mid.1 <- median(yst.1.all$diff.win[cuts])
+
+# we change dispersion a lot for the bulk of the features
+plot(yst.all$rab.all, yst.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  xlab='rAbundance', ylab='dispersion')
+points(yst.s.all$rab.all, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+points(yst.1.all$rab.all, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3))
+abline(h=mn.mid, lty=2)
+abline(h=mn.mid.s, lty=2, col='red')
+abline(h=mn.mid.1, lty=2, col='blue')
+title(main='A', line=-1.2, adj=0.5)
+legend(0,4, legend=c("0", "0.5", "1"), 
+  pch=19, col=c("black", "red", "blue"))
+
+plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3), ylim=c(0,4.5), 
+  xlab='g=0 dispersion', ylab='g=0.5 | g=1 dispersion')
+points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+abline(0,1, lty=2)
+title(main='B', line=-1.2, adj=0.5)
+legend(0,4, legend=c( "0.5", "1"), 
+  pch=19, col=c("red", "blue"))
+
+sig.all <- yst.all$we.eBH < 0.05
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
+sig.s <- yst.s.all$we.eBH < 0.05
+sig.1 <- yst.1.all$we.eBH < 0.05
+
+diff.df <- data.frame(yst.all$rab.all, 
+  yst.1.all$diff.win- yst.all$diff.win)
+sig.df <- data.frame(yst.all$rab.all[sig.all],
+   yst.1.all$diff.win[sig.all]- yst.all$diff.win[sig.all])
+sig.t.df <- data.frame(yst.all$rab.al[sig.t], 
+  yst.1.all$diff.win[sig.t] - yst.all$diff.win[sig.t])
+sig.s.df <- data.frame(yst.all$rab.all[sig.s], 
+  yst.1.all$diff.win[sig.s] - yst.all$diff.win[sig.s])
+sig.1.df <- data.frame(yst.all$rab.all[sig.1], 
+  yst.1.all$diff.win[sig.1]- yst.all$diff.win[sig.1])
+
+plot(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win, xlab='rAbundance', ylab="disp:1 - disp:0", pch=19, cex=0.5)
+title(main='C', line=-1.2, adj=0.5)
+lines(lowess(diff.df, f=0.1), col='grey60', lwd=3)
+lines(lowess(sig.df, f=0.1), col='red', lwd=3)
+lines(lowess(sig.t.df, f=0.5), col='orange', lwd=3)
+lines(lowess(sig.s.df, f=0.5), col='royalblue', lwd=3)
+lines(lowess(sig.1.df, f=0.7), col='cyan', lwd=3)
+
+legend(-10,1.3, legend=c( "diff", "sig diff", "t + sig","0.5 sig", "1 sig"), 
+  pch="-", col=c("grey", "red", "orange", "royalblue", "cyan"))
+
+```
+
 # References

### 2024-07-22 code/fix_authors.sh

@@ -1,0 +1,25 @@
+#!/bin/zsh
+
+Rscript code/markdown.R
+
+awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin D. Silverman$^{2,3,4}$")};1' go3.tex > jnk.tex
+
+awk '{sub(/\\date{}/,"\\date{\\today}")};1' jnk.tex > jnk2.tex
+
+awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$ Department of Biochemistry, University of Western Ontario\n\$^2\$ College of Information Sciences and Technology, Pennsylvania  State University\n\$^3\$ Department of Medicine, Pennsylvania  State University\n\$^4\$ Institute for Computational and Data Science, Pennsylvania State University\n")};1' jnk2.tex > go3-author.tex
+
+pdflatex go3-author.tex -o go3-author.pdf
+bibtex go3-author
+pdflatex go3-author.tex -o go3-author.pdf
+
+rm jnk.tex
+rm jnk2.tex
+rm go3.aux
+rm go3.out
+rm go3.toc
+rm go3.log
+rm go3.tex
+rm go3-author.aux
+rm go3-author.log
+rm go3-author.blg
+

### 2024-07-22 code/meta-aldex.R

@@ -1,4 +1,4 @@
-devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+library(ALDEx2)
 
 
 set.seed(2023)
@@ -19,6 +19,11 @@ xg <- aldex.clr(ko.both, conds.K0, gamma=0.5)
 xg.e <- aldex.effect(xg)
 xg.t <- aldex.ttest(xg)
 xg.all <- cbind(xg.e, xg.t)
+
+for(i in 1:length(xt@dirichletData)){xt@dirichletData[[1]] <- NULL}
+save(xt, file='analysis/xt.clr.Rda')
+for(i in 1:length(xg@dirichletData)){xg@dirichletData[[1]] <- NULL}
+save(xg, file='analysis/xg.clr.Rda')
 
 save(xg.all, file='analysis/xg.Rda')
 mu <- c(1,1.15) # 15% difference in 

### 2024-07-22 code/scale-repro-aldex.R

@@ -1,0 +1,45 @@
+library(ALDEx2)
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+
+yst <- yst[rownames(yst) != "YOR072W-B",]
+
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+
+# what I want is an analysis of the reproducibility with sample size of scaled and unscaled data
+nreps=25
+ss.vec <- c(0,3,5,7,10,14,20)
+
+out.mat.scale <- matrix(data=NA, ncol=nreps, nrow=length(ss.vec))
+out.mat.raw <- matrix(data=NA, ncol=nreps, nrow=length(ss.vec))
+for(j in 2:length(ss.vec)){
+  for(i in 1:nreps){
+    sample.size=ss.vec[j]
+	yst.sub <- cbind(yst[,sample(1:44,sample.size)],
+	   yst[,sample(45:86,sample.size)])
+	sub.conds <-  c(rep('S', sample.size), rep('W', sample.size))
+	
+	x.sub <- aldex.clr(yst.sub, sub.conds)
+	x.sub.t <- aldex.ttest(x.sub)
+	
+	xs.sub <- aldex.clr(yst.sub, sub.conds, gamma=0.5)
+	xs.sub.t <- aldex.ttest(xs.sub)
+	
+	sig.all <- rownames(x.all)[x.all$we.eBH < 0.05]
+	sig.s.all <- rownames(x.s.all)[x.s.all$we.eBH < 0.05]
+	sig.sub.t <- rownames(x.sub.t)[x.sub.t$we.eBH < 0.05]
+	sig.s.sub.t <- rownames(xs.sub.t)[xs.sub.t$we.eBH < 0.05]
+	out.mat.scale[1,i] <- i
+	out.mat.scale[j,i] <- round(length(intersect(sig.s.sub.t, sig.s.all))/length(sig.s.all),2)
+	out.mat.raw[1,i] <- i
+	out.mat.raw[j,i] <- round(length(intersect(sig.sub.t, sig.all))/length(sig.all),2)
+  }
+} 
+}
+
[…]

### 2024-07-22 code/selex.aldex.mu.R

@@ -1,0 +1,29 @@
+library(ALDEx2)
+data(selex)
+
+conds <- c(rep('NS',7), rep('S',7))
+
+sel.all <- aldex(selex, conditions=conds, CI=T, gamma=0.5)
+
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
+sel.1.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
+
+# setting ratio as 1:50 explicitly
+#sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
+scale <- c(rep(1,7), rep(50,7)) 
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=FALSE)
+sel.5.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
+
+# setting ratio in log space
+# 2^-5.64 : 2^0  = 0.02 : 1
+# this also works, but reversed because of factor()  
+# sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(0,-5.64), conds, log=TRUE)
+scale <- c(rep(-5.6,7), rep(0,7))
+sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=TRUE)
+
+sel.2.all <- aldex(selex,conditions=conds, CI=T, gamma=sel.mu)
+
+save(sel.2.all, file="analysis/sel.2.all.Rda")
+save(sel.5.all, file="analysis/sel.5.all.Rda")
+save(sel.1.all, file="analysis/sel.1.all.Rda")
+save(sel.all, file="analysis/sel.all.Rda")

### 2024-07-22 code/yst-FP.R

@@ -1,0 +1,111 @@
+#devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+library(ALDEx2)
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+
+yst <- yst[rownames(yst) != "YOR072W-B",]
+
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
+
+# what I want is an analysis of the FP with a threshold
+# subset vs remainder of set
+nreps=25
+ss.vec <- c(3,5,7,10,14,20)
+counter <- 0
+
+load(file='analysis/x.all.Rda')
+load(file='analysis/x.s.all.Rda')
+
+mat <- matrix(data=NA, ncol=20, nrow=nreps * length(ss.vec))
+
+# note ref, sub  are the numbers in the reference and subset
+# the reference is over-powered
+# FP is the number in the subset not in the reference
+FP.df <- data.frame(mat)
+names(FP.df) <- c('count', 'ssize','ref.unscale', 'sub.unscale', 'ref.scale',
+  'sub.scale', 'sig.s.t.ref1', 'sig.t.ref1','sig.s.t.ref1.4', 'sig.t.ref1.4','sig.s.t.ref2', 'sig.t.ref2','FP.scale', 'FP.unscale','FP.scale1', 'FP.unscale1', 'FP.scale1.4', 
+  'FP.unscale1.4', 'FP.scale2', 'FP.unscale2')
+# keep
+# ssvec[j], nrep, sig.ref, sig.sub, sig.sc, sig.sc.sub, FP.sig.sub, FP.sig.sub1, FP.sig.sub2,
+#   FP.sig.sc.sub, FP.sig.sc.sub1, FP.sig.sc.sub2
+for(j in 1:length(ss.vec)){
+  for(i in 1:nreps){
[…]

### 2024-07-22 code/yst-reproducibility.R

@@ -1,0 +1,51 @@
+devtools::load_all('~/Documents/0_git/ALDEx_bioc')
+#library(ALDEx2)
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+
+yst <- yst[rownames(yst) != "YOR072W-B",]
+
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
+
+# what I want is an analysis of the reproducibility with sample size of scaled and unscaled data
+nreps=25
+ss.vec <- c(0,3,5,7,10,14,20)
+
+load(file='analysis/x.all.Rda')
+load(file='analysis/x.s.all.Rda')
+
+out.mat.scale <- matrix(data=NA, ncol=nreps, nrow=length(ss.vec))
+out.mat.raw <- matrix(data=NA, ncol=nreps, nrow=length(ss.vec))
+for(j in 2:length(ss.vec)){
+  for(i in 1:nreps){
+    sample.size=ss.vec[j]
+	yst.sub <- cbind(yst[,sample(1:44,sample.size)],
+	   yst[,sample(45:86,sample.size)])
+	sub.conds <-  c(rep('S', sample.size), rep('W', sample.size))
+	
+	x.sub <- aldex.clr(yst.sub, sub.conds)
+	x.sub.t <- aldex.ttest(x.sub)
+	
+	xs.sub <- aldex.clr(yst.sub, sub.conds, gamma=0.5)
+	xs.sub.t <- aldex.ttest(xs.sub)
+	
+	sig.all <- rownames(x.all)[x.all$we.eBH < 0.05]
+	sig.s.all <- rownames(x.s.all)[x.s.all$we.eBH < 0.05]
+	sig.sub.t <- rownames(x.sub.t)[x.sub.t$we.eBH < 0.05]
+	sig.s.sub.t <- rownames(xs.sub.t)[xs.sub.t$we.eBH < 0.05]
+	out.mat.scale[1,i] <- i
[…]

### 2024-07-22 code/yst-sens.R

@@ -1,0 +1,24 @@
+# this gives a sensitivity analysis for transcriptome
+# maybe supplement?
+# x.sens <- aldex.senAnalysis(x, gamma=c(1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7,1))
+
+library(ALDEx2)
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
+yst <- read.table(url, header=T, row.names=1)
+# remove the one gene with 0 reads
+
+yst <- yst[rownames(yst) != "YOR072W-B",]
+
+# Gierlinski:2015aa
+yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
+yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
+
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
+
+set.seed(2023)
+#ALDEx2
+x <- aldex.clr(yst, conds, gamma=1e-3, verbose=F)
+x.sens <- aldex.senAnalysis(x, gamma=c(1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7,1))
+save(x.sens, file="analysis/yst.sens.Rda")

### 2024-07-22 code/yst.aldex.mu.r

@@ -11,11 +11,19 @@ yst <- yst[rownames(yst) != "YOR072W-B",]
 yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
 yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL  
 
-conds <- c(rep('S', 44), rep('W', 42))
-coldata <- data.frame(conds)
+yst.conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(yst.conds)
 
-mu.dat <- aldex.makeScaleMatrix(0.5,c(1,1.04),conds,mc.samples=128)
+mu.dat <- aldex.makeScaleMatrix(0.5,c(1,1.04),yst.conds,mc.samples=128)
 
 set.seed(2023)
-x.s.mu.all <- aldex(yst, conds, gamma=mu.dat, verbose=F)
+x.s.mu.all <- aldex(yst, yst.conds, gamma=mu.dat, verbose=F)
 save(x.s.mu.all, file='analysis/x.s.mu.all.Rda')
+
+yst.all <- aldex(yst, yst.conds, gamma=NULL)
+yst.s.all <- aldex(yst, yst.conds, gamma=0.5)
+yst.1.all <-aldex(yst, yst.conds, gamma=1)
+
+save(yst.all, file="analysis/yst.all.Rda")
+save(yst.s.all, file="analysis/yst.s.all.Rda")
+save(yst.1.all, file="analysis/yst.1.all.Rda")

### 2024-07-22 fix_authors.sh

@@ -1,14 +1,0 @@
-#!/bin/zsh
-
-awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin D. Silverman$^{2,3,4}$")};1' go3.tex > jnk.tex
-
-awk '{sub(/\\date{}/,"\\date{\\today}")};1' jnk.tex > jnk2.tex
-
-awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$ Department of Biochemistry, University of Western Ontario\n\$^2\$ College of Information Sciences and Technology, Pennsylvania  State University\n\$^3\$ Department of Medicine, Pennsylvania  State University\n\$^4\$ Institute for Computational and Data Science, Pennsylvania State University\n")};1' jnk2.tex > go3-author.tex
-
-rm jnk.tex
-rm jnk2.tex
-rm go3.aux
-rm go3.pdf
-rm go3.out
-rm go3.toc

### 2024-08-13 go3-author.tex
final go3 before Justin

@@
 traced, in whole or in part, to misspecification of scale. The first
-issue is poor control of the false discovery rate (FDR) (3, 4),
+issue is poor control of the false discovery rate (FDR)
+{[}(3);(4);Nearing:\href{mailto:2022aa;@Li}{\nolinkurl{2022aa;@Li}}:2022aa{]},
 exhibited as dataset-dependent FDR control. The FDR problem is connected
@@
 results in a worsening of the accuracy of the analysis (9). The final
-problem is that these data are relative data, i.e.~compositional, and
-substantial effort has been put into removing this constraint.
+problem is that these data are relative data, i.e.~compositional (10,
+11), and substantial effort has been put into removing this constraint.
 
@@
 of HTS. Biological variation in scale often represents an important
-unmeasured confounder in HTS analyses (10). For example, cells
+unmeasured confounder in HTS analyses (12). For example, cells
 transformed by the cMyc oncogene have about 3 times the amount of mRNA
-and about twice the rRNA content than non-transformed cells (11), and
-this dramatically skews transcriptome analysis (12). In addition,
+and about twice the rRNA content than non-transformed cells (13), and
+this dramatically skews transcriptome analysis (14). In addition,
 wild-type and mutant strains of cell lines, yeast or bacteria have
@@
 which affect our ability to identify truly differentially abundant genes
-(13--15). As another example, the total bacterial load of the vaginal
+(15--17). As another example, the total bacterial load of the vaginal
 microbiome differs by 1-2 orders of magnitude in absolute abundance
-between the healthy and bacterial vaginosis states (16), and the
-composition between these states is dramatically different (17, 18).
+between the healthy and bacterial vaginosis states (18), and the
+composition between these states is dramatically different (19, 20).
 Thus, a full description of any of these systems includes both relative
@@
 difficult to interpret, and different normalizations provide different
-outputs when applied to the same dataset (3, 19--21). Intuitively,
+outputs when applied to the same dataset (3, 21--24). Intuitively,
 normalizations in widespread use assume that either all samples have the
-same scale, e.g.~proportions, rarefaction (22), RPKM (23, 24), etc; or
+same scale, e.g.~proportions, rarefaction (25), RPKM (26, 27), etc; or
 that a subset of features in one sample can be chosen as a reference to
-which the others are scaled e.g.~the TMM,(25), or LVHA (7); or that
-different sub-parts of each sample maintain a constant scale across
-samples e.g.~the RLE (26); or that the geometric mean of the parts is
-appropriate e.g the CLR (27) and its derivatives.
-
-The original ALDEx2 (28) model made a strict assumption about scale
-through the CLR normalization (8). While this assumption could be useful
-in many cases it could never be exactly true, and others have shown that
-it is not always the best option (29). Nixon et al. (8) showed that
-better scale assumptions resulted in more reproducible data analysis
-including better control of both false positive and false negative
-results. In essence ALDEx2 has been modified to explicitly model the
-scale over a range of reasonable normalization parameters. Here, we
-briefly introduce these modifications and then show how scale
+which the others are scaled e.g.~the TMM (28), or LVHA (7) or the
+additive log-ratio (29); or that different sub-parts of each sample
+maintain a constant scale across samples e.g.~the RLE (30); or that the
+geometric mean of the parts is appropriate e.g the CLR and its
+derivatives.
+
+The original ALDEx2 (31) model unwittingly made a strict assumption
+about scale through the CLR normalization (8). While this assumption
+could be useful in many cases it could never be exactly true, and others
+have shown that it is not always the best option (32). Nixon et al. (8)
+showed that better scale assumptions resulted in more reproducible data
+analysis including better control of both false positive and false
+negative results. In essence ALDEx2 has been modified to explicitly
+model the scale over a range of reasonable normalization parameters.
+Here, we briefly introduce these modifications and then show how scale
 uncertainty can greatly improve modeling in transcriptome and
@@
 \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
-denominator is determined by some function of the observation and the
-\(\hat{{\mathbf{W}}}\) notation indicates that the normalization is
-attempting to provide an estimate of the true value. The technical
+denominator is determined by some function of the observation. We use
+the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization
+is attempting to provide an estimate of the true value. The technical
 variation in sequencing depth
@@
 
-The ALDEx2 R package (28) is a general purpose toolbox for Bayesian
+The ALDEx2 R package (31, 33) is a general purpose toolbox for Bayesian
 modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
@@
 are used to estimate LFC as part of differential abundance (or
-expression) analyses. For more details on ALDEx2 see (8, 28).
+expression) analyses. For more details on ALDEx2 see (8, 31).
 
@@
 discovery rates vary widely by analysis, dataset, and normalization
-method (30, 31). To fix this, Nixon et al. (8) showed that models should
+method (34, 35). To fix this, Nixon et al. (8) showed that models should
 incorporate potential error in the assumptions implied by
@@
 transcriptomics, to use the dual cutoff approach graphically exemplified
-by volcano plots (32, 33) because not all statistically significant
+by volcano plots (36, 37) because not all statistically significant
 differences are biologically relevant. Paraphrasing this we can say that
@@
 
-We use the data of Gierliński et al. (34) who conducted a highly
+We use the data of Gierliński et al. (38) who conducted a highly
 replicated yeast transcriptome experiment comparing a wild-type strain
@@
 argue that a dual cutoff approach is appropriate to limit the number of
-significant parts and for purposes of reproducibility (33). However, in
+significant parts and for purposes of reproducibility (37). However, in
 this benchmarking study, the number of significantly different
@@
 power is intrinsically linked to sample size and very large sample sizes
-are indeed desired (35). Furthermore, the dual-cutoff approach is known
-not to provide appropriate FDR control (5, 6). Schurch et al. (33)
+are indeed desired (39). Furthermore, the dual-cutoff approach is known
+not to provide appropriate FDR control (5, 6). Schurch et al. (37)
 further suggested that different tools might be better in some
@@
 statistically significantly different between groups with a
-Benjamini-Hochberg (36) false discovery rate (FDR) of 0.05; i.e.~4636
+Benjamini-Hochberg (40) false discovery rate (FDR) of 0.05; i.e.~4636
 (79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
@@
 returned as significant and that some if not the majority are driven by
-technical differences in the analysis (20, 21, 33, 37, 38).
+technical differences in the analysis (22, 23, 37, 41, 42).
 
@@
 
-The effect plots (39) in Figure 1A (ALDEx2) and Supplementary Figure 1
+The effect plots (43) in Figure 1A (ALDEx2) and Supplementary Figure 1
 (DESeq2) shows that the majority of significant transcripts (red,
@@
 dispersion and we suggest that this is driven by the experimental design
-(33). Scale uncertainty can be incorporated using the \texttt{gamma}
+(37). Scale uncertainty can be incorporated using the \texttt{gamma}
 parameter that controls the amount uncertainty added to the CLR mean
@@
 and it is common practice to use a dual-cutoff by choosing transcripts
-based on a thresholds for both corrected p-values and fold-changes (33)
+based on a thresholds for both corrected p-values and fold-changes (37)
 (here set at \(\pm 2^{1.4}\) for the latter), although considerable
@@
 moderate expression is more predictable than highly-expressed genes or
-of housekeeping genes (40). This is at odds with the known biology of
+of housekeeping genes (44). This is at odds with the known biology of
 cells where single cell counting of highly-expressed transcripts shows
-that they have little intrinsic variation (41).
+that they have little intrinsic variation (45).
 
@@
 
-Dos Santos et al. (42) used a vaginal metatranscriptome dataset to
+Dos Santos et al. (46) used a vaginal metatranscriptome dataset to
 compare the gene expression in bacteria collected from healthy (H) and
@@
 relative abundance of species between groups and the gene expression
-level within a species is different (43). Additionally, prior research
+level within a species is different (47). Additionally, prior research
 suggests that the total number of bacteria is about 10 times more in the
-BV than in the H condition (16). Thus, this is an extremely challenging
+BV than in the H condition (18). Thus, this is an extremely challenging
 environment in which to determine differential abundance as there are
@@
 method to analyze vaginal metratranscriptome data is on a taxon-by-taxon
-basis (43--45) because the scale confounding can be ignored. Attempts at
+basis (47--49) because the scale confounding can be ignored. Attempts at
 system-wide analysis show that many housekeeping functions are returned
@@
 the organism of origin. Each point represents one of 3728 KEGG functions
-(46). There are many more functions represented in the BV group (bottom)
+(50). There are many more functions represented in the BV group (bottom)
 than in the healthy group (top). This is because the
@@
 chaperones, etc (Supplementary file housekeeping.txt). The transcripts
-of many of these are commonly used as invariant reference sequences (40)
+of many of these are commonly used as invariant reference sequences (44)
 and so would not be expected to contribute to differences in ecosystem
@@
 transferrability of it between sample cohorts of the same condition is
-outlined elsewhere (42). We suggest that the default scale model is
+outlined elsewhere (46). We suggest that the default scale model is
 sufficient when the data are approximately centred. However, an informed
@@
 
-Biological systems are both predictably variable and stochastic (41) and
+Biological systems are both predictably variable and stochastic (45) and
 current measurement methods that rely on high throughput sequencing fail
@@
 (8, 9). In the absence of external information (such as spike-in probes
-(12), cell counting (1), FISH (47) etc) sequencing depth normalisation
-methods cannot recapture the scale information (12), and can only
+(14), cell counting (1), FISH (51) etc) sequencing depth normalisation
+methods cannot recapture the scale information (14), and can only
 normalize for the technical variation due to sequencing depth.
@@
 normalizations used for the analysis of datasets such as
-transcriptomes(4, 19, 33, 37) and microbiomes (3, 21, 29, 30, 48).
-Generically, it is observed that the actual agreement between analysis
-methods can be modest, and no single method outperforms all others in
-every dataset or type of experiment. That different tools appear to work
-more reliably in different datasets from different sources can be
-explained as different normalizations being a better fit to the scale a
-particular dataset by chance.
-
-From our perspective, the disagreement between tools can be explained by
-the partially identifiable nature of the data; that is, the observed
-data are consistent with multiple ways of estimating the parameters (8).
-Partially identifiable data exhibit multiple pathologies, chief among
-them being that that different analytic approaches can produce different
-parameter estimates and that more data produces worse estimates because
-the additional data increases the precision of a flawed estimate (49).
+transcriptomes(4, 21, 35, 37, 41) and microbiomes (3, 23, 24, 32, 34,
+52). Generically, it is observed that the actual agreement between
+analysis methods can be modest, and no single method outperforms all
+others in every dataset or type of experiment. That different tools
+appear to work more reliably in different datasets from different
+sources can be explained as different normalizations being a better fit
+to the scale of a particular dataset by chance.
+
 In the analysis of HTS data it is often observed that larger datasets
-converge on the majority of parts being significantly different (8, 33),
-and that different analytic approaches result in different parts being
-chosen as significantly different (4, 33, 37).
+converge on the majority of parts being significantly different (8, 35,
+37), and that different analytic approaches result in different parts
+being chosen as significantly different (4, 24, 37, 41). Li et al. (35)
+conducted a permutation-based benchmarking study and found that popular
+tools performed worse than simple Wilcoxon rank-sum tests in controlling
+the FDR when sample sizes became large. They suggested that the presence
+of outliers were one of the factors driving this observation. Brooks et
+al. (53) suggested that inappropriate choice of benchmarking methods are
+also a major contributing factor. From our perspective, the disagreement
+between tools can be explained by the partially identifiable nature of
+the data; that is, the observed data are consistent with multiple ways
+of estimating the parameters (8). Partially identifiable data exhibit
+multiple pathologies, chief among them being that that different
+analytic approaches can produce different parameter estimates and that
+more data produces worse estimates because the additional data increases
+the precision of a flawed estimate (54).
 
@@
 many features that is a by-product of experimental design, sequencing
-and normalization. In the Schurch et al. (33) dataset, the data were
+and normalization. In the Schurch et al. (37) dataset, the data were
 from single colonies derived from a single culture. Thus, it is more
 accurate to describe the 96 samples as wet-lab technical replicates
-rather than independent samples. However, this type of replicate is
-standard in the molecular literature, and would be expected to result in
-the very low dispersion that is observed.
+rather than independent samples. However, this type of replication
+approach is standard in the molecular literature, and would be expected
+to result in the very low dispersion that is observed.
 
@@
 post-sequencing, we can study the sensitivity of the results to the
-choice of normalization and thus scale. To do this we incorporate scale
-models (8) in lieu of normalizations through the default scale model
-that is now built into ALDEx2. Scale uncertainty is incorporated using
-the \texttt{gamma} parameter that controls the amount uncertainty added
-to the CLR mean assumption when we call either \texttt{aldex()}, or
-\texttt{aldex.clr()}. The ALDEx2 package contains a sensitivity analysis
-function, \texttt{aldex.senAnalysis()}, that can be used to explore the
-effect of different amounts of scale uncertainty, and an example is
-shown in Supplementary Figure 2 for the yeast transcriptome dataset.
-Here it is clear that even tiny amounts of scale preclude the majority
-of transcripts from being considered statistically significant. In
+choice of normalization and thus scale. Varying amounts of scale
+uncertainty is incorporated using the \texttt{gamma} parameter that
+controls the amount uncertainty added to the CLR mean assumption when we
+call either \texttt{aldex()}, or \texttt{aldex.clr()}. The ALDEx2
+package contains a sensitivity analysis function,
+\texttt{aldex.senAnalysis()}, that can be used to explore the effect of
+different amounts of scale uncertainty, and an example is shown in
+Supplementary Figure 2 for the yeast transcriptome dataset. Here it is
+clear that even tiny amounts of scale preclude the majority of
+transcripts from being considered statistically significant. In
 practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
@@
 when performing the gold-standard quantitative PCR test of differential
-abundance (50, 51). Thus, the use of this prior knowledge is not unique
+abundance (55, 56). Thus, the use of this prior knowledge is not unique
 to our approach.
@@
 Beyond concerns of fidelity and rigor, scale models also enhance the
-reproducibility and transparency of HTS analyses. An advantage of
-incorporating scale is that analyses can be made much more robust such
-that actual or potential differences in scale can be tested and
-accounted for explicitly. This occurs because rather than using an
+reproducibility and transparency of HTS analyses. The addition of scale
+uncertainty essentially tests the model over a range of normalizations
+(8) and so can replace the consensus approach that has been proposed by
+some groups (24, 57) with no additional computational overhead. Thus, an
+advantage of incorporating scale is that analyses can be made much more
+robust such that actual or potential differences in scale can be tested
+and accounted for explicitly. This occurs because rather than using an
 implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the
@@
 when the majority of features are asymmetrically distributed or
-expressed (or both) in the groups (42). Thus, reporting scale
+expressed (or both) in the groups (46). Thus, reporting scale
 uncertainty should become a standard practice in the analysis of HTS
@@
 
+\bibitem[\citeproctext]{ref-lovell:2011}
+10. Lovell,D., Müller,W., Taylor,J., Zwart,A. and Helliwell,C. (2011)
+Proportions, percentages, ppm: Do the molecular biosciences treat
+compositional data right? In Pawlowsky-Glahn,V., Buccianti,A. (eds),
+\emph{Compositional Data Analysis: Theory and Applications}. John Wiley;
+Sons New York, NY, London, pp. 193--207.
+
+\bibitem[\citeproctext]{ref-Quinn:2019aa}
+11. Quinn,T.P., Erb,I., Gloor,G., Notredame,C., Richardson,M.F. and
+Crowley,T.M. (2019) \href{https://doi.org/10.1093/gigascience/giz107}{A
+field guide for the compositional analysis of any-omics data}.
+\emph{Gigascience}, \textbf{8}.
+
 \bibitem[\citeproctext]{ref-Lovell:2015}
-10. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
+12. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
 Bähler,J. (2015)
@@
 \bibitem[\citeproctext]{ref-Nie:2012aa}
-11. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
+13. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
 Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
@@
 \bibitem[\citeproctext]{ref-Loven:2012aa}
-12. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
+14. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
 Burge,C.B., Levens,D.L., Lee,T.I. and Young,R.A. (2012)
@@
 \bibitem[\citeproctext]{ref-Scott:2010}
-13. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
+15. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
 \href{https://doi.org/10.1126/science.1192588}{Interdependence of cell
@@
 \bibitem[\citeproctext]{ref-Yoshikawa:2011aa}
-14. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
+16. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
 Shimizu,H. (2011) \href{https://doi.org/10.1002/yea.1843}{Comprehensive
@@
 \bibitem[\citeproctext]{ref-Lin:2018aa}
-15. Lin,J. and Amir,A. (2018)
+17. Lin,J. and Amir,A. (2018)
 \href{https://doi.org/10.1038/s41467-018-06714-z}{Homeostasis of protein
@@
 \bibitem[\citeproctext]{ref-Zozaya:2010}
-16. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
+18. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
 \href{https://doi.org/10.1128/JCM.00851-09}{Quantitative PCR assessments
@@
 \bibitem[\citeproctext]{ref-Ravel:2010}
-17. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
+19. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
 McCulle,S.L., Karlebach,S., Gorle,R., Russell,J., Tacket,C.O., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hummelen:2010}
-18. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
+20. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
 Changalucha,J., Gloor,G.B. and Reid,G. (2010)
@@
 \bibitem[\citeproctext]{ref-Bullard:2010}
-19. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
+21. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
 \href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
@@
 \bibitem[\citeproctext]{ref-Dillies:2013}
-20. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
+22. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
 Jeanmougin,M., Servant,N., Keime,C., Marot,G., Castel,D., Estelle,J.,
@@
 \bibitem[\citeproctext]{ref-Weiss:2017aa}
-21. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
+23. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
 Lozupone,C., Zaneveld,J.R., Vázquez-Baeza,Y., Birmingham,A., \emph{et
@@
 
+\bibitem[\citeproctext]{ref-Nearing:2022aa}
+24. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
+Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
+\emph{et al.} (2022)
+\href{https://doi.org/10.1038/s41467-022-28034-z}{Microbiome
+differential abundance methods produce different results across 38
+datasets}. \emph{Nat Commun}, \textbf{13}, 342.
+
 \bibitem[\citeproctext]{ref-Hughes:2005tu}
-22. Hughes,J.B. and Hellmann,J.J. (2005)
+25. Hughes,J.B. and Hellmann,J.J. (2005)
 \href{https://doi.org/10.1016/S0076-6879(05)97017-1}{The application of
@@
 \bibitem[\citeproctext]{ref-Mortazavi:2008}
-23. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
+26. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
 (2008) \href{https://doi.org/10.1038/nmeth.1226}{Mapping and quantifying
@@
 \bibitem[\citeproctext]{ref-wagner:tpm}
-24. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
+27. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
 \href{https://doi.org/10.1007/s12064-012-0162-3}{Measurement of mRNA
@@
 \bibitem[\citeproctext]{ref-Robinson:2010a}
-25. Robinson,M.D. and Oshlack,A. (2010)
+28. Robinson,M.D. and Oshlack,A. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-3-r25}{A scaling normalization
@@
 
+\bibitem[\citeproctext]{ref-aitchison1982}
+29. Aitchison,J. (1982) The statistical analysis of compositional data.
+\emph{Journal of the Royal Statistical Society: Series B
+(Methodological)}, \textbf{44}, 139--160.
+
 \bibitem[\citeproctext]{ref-Anders:2010}
-26. Anders,S. and Huber,W. (2010)
+30. Anders,S. and Huber,W. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-10-r106}{Differential
@@
 
-\bibitem[\citeproctext]{ref-aitchison1982}
-27. Aitchison,J. (1982) The statistical analysis of compositional data.
-\emph{Journal of the Royal Statistical Society: Series B
-(Methodological)}, \textbf{44}, 139--160.
-
 \bibitem[\citeproctext]{ref-fernandes:2013}
-28. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
+31. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
 (2013) \href{https://doi.org/10.1371/journal.pone.0067019}{ANOVA-like
@@
 \bibitem[\citeproctext]{ref-Yerke:2024aa}
-29. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
+32. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
 \href{https://doi.org/10.1186/s40168-023-01747-z}{Proportion-based
@@
 
+\bibitem[\citeproctext]{ref-fernandes:2014}
+33. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
+Edgell,D.R. and Gloor,G.B. (2014)
+\href{https://doi.org/10.1186/2049-2618-2-15}{Unifying the analysis of
+high-throughput sequencing datasets: Characterizing {RNA}-seq, 16{S}
+r{RNA} gene sequencing and selective growth experiments by compositional
+data analysis}. \emph{Microbiome}, \textbf{2}, 15.1--15.13.
+
 \bibitem[\citeproctext]{ref-hawinkel2017}
-30. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
+34. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
 \href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
@@
 \bibitem[\citeproctext]{ref-Li:2022aa}
-31. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+35. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
 \href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
@@
 \bibitem[\citeproctext]{ref-Cui:2003aa}
-32. Cui,X. and Churchill,G.A. (2003)
+36. Cui,X. and Churchill,G.A. (2003)
 \href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
@@
 \bibitem[\citeproctext]{ref-Schurch:2016aa}
-33. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
+37. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-Gierlinski:2015aa}
-34. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
+38. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-Halsey:2015aa}
-35. Halsey,L.G., Curran-Everett,D., Vowler,S.L. and Drummond,G.B. (2015)
+39. Halsey,L.G., Curran-Everett,D., Vowler,S.L. and Drummond,G.B. (2015)
 \href{https://doi.org/10.1038/nmeth.3288}{The fickle p value generates
@@
 \bibitem[\citeproctext]{ref-benjamini:1995}
-36. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
+40. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
 rate: A practical and powerful approach to multiple testing.
@@
 \bibitem[\citeproctext]{ref-Soneson:2013}
-37. Soneson,C. and Delorenzi,M. (2013)
+41. Soneson,C. and Delorenzi,M. (2013)
 \href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
@@
 \bibitem[\citeproctext]{ref-maza2013}
-38. Maza,E., Frasse,P., Senin,P., Bouzayen,M. and Zouine,M. (2013)
+42. Maza,E., Frasse,P., Senin,P., Bouzayen,M. and Zouine,M. (2013)
 \href{https://doi.org/10.4161/cib.25849}{Comparison of normalization
@@
 \bibitem[\citeproctext]{ref-gloor:effect}
-39. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
+43. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
 \href{https://doi.org/10.1080/10618600.2015.1131161}{Displaying
@@
 \bibitem[\citeproctext]{ref-Rocha:2020aa}
-40. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
+44. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
 (2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
@@
 \bibitem[\citeproctext]{ref-Taniguchi:2010aa}
-41. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
+45. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
 Emili,A. and Xie,X.S. (2010)
@@
 \bibitem[\citeproctext]{ref-dosSantos:2024}
-42. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
+46. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
 Gloor,G.B. (2024) Vaginal metatranscriptome meta-analysis reveals
@@
 \bibitem[\citeproctext]{ref-macklaim:2013}
-43. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
+47. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
 and Gloor,G.B. (2013)
@@
 \bibitem[\citeproctext]{ref-Denge00262-18}
-44. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
+48. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
 Wagner-Döbler,I. (2018)
@@
 \bibitem[\citeproctext]{ref-Fettweis:2019aa}
-45. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
+49. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
 Parikh,H.I., Huang,B., Arodz,T.J., Edupuganti,L., Glascock,A.L.,
@@
 \bibitem[\citeproctext]{ref-Okuda:2008}
-46. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
+50. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
 Goto,S. and Kanehisa,M. (2008)
@@
 \bibitem[\citeproctext]{ref-yeast-absolute}
-47. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
+51. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
 Bähler,J. (2012)
@@
 \bibitem[\citeproctext]{ref-McMurdie:2014a}
-48. McMurdie,P.J. and Holmes,S. (2014)
+52. McMurdie,P.J. and Holmes,S. (2014)
 \href{https://doi.org/10.1371/journal.pcbi.1003531}{Waste not, want not:
@@
 
+\bibitem[\citeproctext]{ref-Brooks:2024aa}
+53. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
+\href{https://doi.org/10.1038/s41576-023-00679-6}{Challenges and best
+practices in omics benchmarking}. \emph{Nat Rev Genet}, \textbf{25},
+326--339.
+
 \bibitem[\citeproctext]{ref-gustafson2015bayesian}
-49. Gustafson,P. (2015) Bayesian inference for partially identified
+54. Gustafson,P. (2015) Bayesian inference for partially identified
 models: Exploring the limits of limited data CRC Press.
@@
 \bibitem[\citeproctext]{ref-Thellin:1999aa}
-50. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
+55. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
 Hennen,G., Grisar,T., Igout,A. and Heinen,E. (1999)
@@
 \bibitem[\citeproctext]{ref-SEQCux2fMAQC-III-Consortium:2014aa}
-51. SEQC/MAQC-III Consortium (2014)
+56. SEQC/MAQC-III Consortium (2014)
 \href{https://doi.org/10.1038/nbt.2957}{A comprehensive assessment of
@@
 
+\bibitem[\citeproctext]{ref-Song:2023aa}
+57. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
+Klatt,N.R., Hensley-McBain,T. and Wu,M.C. (2023)
+\href{https://doi.org/10.1186/s12859-023-05147-w}{Accommodating multiple
+potential normalizations in microbiome associations studies}. \emph{BMC
+Bioinformatics}, \textbf{24}, 22.
+
 \end{CSLReferences}

### 2024-08-13 go3.Rmd
final go3 before Justin

@@
 traced, in whole or in part, to  misspecification of scale. The first issue 
-is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa], 
+is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;Nearing:2022aa;@Li:2022aa], 
   exhibited as dataset-dependent 
@@
 more information results in a worsening of the accuracy of the analysis [@nixon2023scale,@Nixon2024B]. 
-The final problem is that these data are relative data, i.e. compositional, 
+The final problem is that these data are relative data, i.e. compositional [@lovell:2011;@Quinn:2019aa], 
 and substantial effort has been put into removing this constraint.
@@
 They showed that \emph{all} normalizations make some assumption about  scale.
-These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa].
+These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
 Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g.
 proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g.
-the TMM,[@Robinson:2010a], or LVHA [@Wu2021]; or that different sub-parts of each sample maintain a constant scale across samples e.g.
-the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@aitchison1982] and its derivatives.
+the TMM [@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g.
+the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR  and its derivatives.
 
-The original ALDEx2 [@fernandes:2013] model made a strict assumption about scale through the CLR normalization [@nixon2023scale].
+The original ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale].
 While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa].
@@
 
-All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation and the \(\hat{{\mathbf{W}}}\) notation indicates that the normalization is attempting to provide an estimate of the true value.
+All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation.
+We use the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization is attempting to provide an estimate of the true value.
 The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al.
@@
 
-The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data.
+The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox for Bayesian modeling of HTS data.
 For brevity, we discuss ALDEx2 in its simplest form as a tool for estimating the magnitude and statistical significance of Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models.
@@
 By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\) where \(G_{n}\) denotes the geometric mean of the vector \((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\).
-The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale,Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets).
+The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale;@Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets).
 In fact, Nixon et al.
@@
 
-Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Yerke:2024aa].
+Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa].
 Generically, it is observed that the actual agreement between analysis methods can be modest, and  no single method outperforms all others in every dataset or type of experiment.
-That different tools appear to work more reliably in different datasets from different sources can be explained as different  normalizations being a better fit to the scale  a particular dataset by chance.
+That different tools appear to work more reliably in different datasets from different sources can be explained as different  normalizations being a better fit to the scale of a particular dataset by chance.
 
+In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@Li:2022aa;@nixon2023scale], and that different analytic approaches result in different parts being chosen as significantly different [@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Nearing:2022aa].
+Li et al.
+-@Li:2022aa conducted a permutation-based benchmarking study and found that popular tools performed worse than simple Wilcoxon rank-sum tests in controlling the FDR when sample sizes became large.
+They suggested that the presence of outliers were one of the factors driving this observation.
+Brooks et al.
+-@Brooks:2024aa suggested that inappropriate choice of benchmarking methods are also a major contributing factor.
 From our perspective, the disagreement between tools can be explained by the partially identifiable nature of the data; that is,  the  observed data are consistent with multiple ways of estimating the parameters [@nixon2023scale].
 Partially identifiable data exhibit multiple pathologies, chief among them being that that different analytic approaches can produce  different parameter estimates and that more data produces worse estimates because the additional data increases the precision of a flawed estimate [@gustafson2015bayesian].
-In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@nixon2023scale], and that different analytic approaches result in different parts being chosen as significantly different [@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa].
 
@@
 Thus, it is more accurate to describe the 96 samples as wet-lab technical replicates rather than independent samples.
-However, this type of replicate is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.
+However, this type of replication approach is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.
 
@@
 While the actual scale of the underlying environment is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale.
-To do this we incorporate scale models [@nixon2023scale] in lieu of normalizations through the default scale model that is now built into ALDEx2.
-Scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
+Varying amounts of scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
 The ALDEx2 package  contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty, and an example is shown in  Supplementary Figure 2 for the yeast transcriptome dataset.
@@
 Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses.
-An advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
+The addition of scale uncertainty essentially tests the model over a range of normalizations [@nixon2023scale] and so can replace the consensus approach that has been proposed by some groups [@Nearing:2022aa;@Song:2023aa] with no additional computational overhead.
+Thus, an advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
 This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models can make this an explicit part of the model building process.

### 2024-08-13 go3.tex
go3-edits
final go3 before Justin

@@
 (merging) multiple libraries together, and then applying a sample of the
-multiplexed library to the flow cell. Each of these steps can be though
-of as a sampling step: a pool of DNA that could be measured is sampled
-and only that subsampled DNA is carried over to subsequent steps. Within
-each sampling step, the connection between the size of the sampled DNA
-pool and the scale (e.g., size, microbial load, or total gene
-expression) of the measured biological system is degraded or lost.
-Increasingly, researchers are turning to modified experimental protocols
-(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
-biological variation in scale; see for example (Vandeputte et al. 2017;
-Props et al. 2017). However, these protocols often do not recover
-meaningful biological variation but instead recover scale variation at
-some intermediate step in the sample preparation protocol (e.g., at the
-stage where the DNA spike-in was added). In short, the disconnect
-between sample-to-sample variation in sequencing depth and biological
-variation in scale remains an outstanding challenge.
-
-Biological variation in scale can represent an important unmeasured
-confounder in HTS analyses (Lovell et al. 2015). For example, cells
+multiplexed library to the flow cell. Each of these steps is a
+compositional sampling step as only a fixed-size subsample of nucleic
+acid is carried over to subsequent steps. Thus, with each sampling step
+the connection between the size of the sampled DNA pool and the scale
+(e.g., size, microbial load, or total gene expression) of the measured
+biological system is degraded or lost. Increasingly, researchers are
+turning to modified experimental protocols (e.g., DNA spike-ins or cell
+counting) in an attempt to recover the lost biological variation in
+scale; see for example (1, 2). However, these protocols often do not
+recover meaningful biological variation but instead recover scale
+variation at the intermediate step in the sample preparation protocol
+where the DNA spike-in was added. In short, the disconnect between
+sample-to-sample variation in sequencing depth and biological variation
+in scale remains an outstanding challenge.
+
+The analysis of HTS data suffers from several known problems that can be
+traced, in whole or in part, to misspecification of scale. The first
+issue is poor control of the false discovery rate (FDR)
+{[}(3);(4);Nearing:\href{mailto:2022aa;@Li}{\nolinkurl{2022aa;@Li}}:2022aa{]},
+exhibited as dataset-dependent FDR control. The FDR problem is connected
+to the double-filtering approach that is often used, but which is known
+not to have appropriate FDR control (5, 6). The second issue is poor
+performance when analyzing asymmetric data; data that arises because of
+a bias in the direction of change. This type of data frequently arises
+in in-vitro selection experiments (SELEX), transcriptome analysis, and
+microbiome analysis (7). The third issue is that these problems become
+more pronounced as more samples are collected; that is, more information
+results in a worsening of the accuracy of the analysis (9). The final
+problem is that these data are relative data, i.e.~compositional (10,
+11), and substantial effort has been put into removing this constraint.
+
+The first three problems were recently shown by Nixon and Silverman (8)
+to be a result of a mismatch between the underlying size or scale of the
+system and the assumptions of the normalizations used for the analysis
+of HTS. Biological variation in scale often represents an important
+unmeasured confounder in HTS analyses (12). For example, cells
 transformed by the cMyc oncogene have about 3 times the amount of mRNA
-and about twice the rRNA content than non-transformed cells (Nie et al.
-2012), and this dramatically skews transcriptome analysis (Lovén et al.
-2012). In addition, wild-type and mutant strains of cell lines, yeast or
-bacteria often have different growth rates, which would affect our
-ability to identify truly differentially abundant genes (Yoshikawa et
-al. 2011). As another example, the total bacterial load of the vaginal
+and about twice the rRNA content than non-transformed cells (13), and
+this dramatically skews transcriptome analysis (14). In addition,
+wild-type and mutant strains of cell lines, yeast or bacteria have
+different growth rates and RNA contents under different conditions,
+which affect our ability to identify truly differentially abundant genes
+(15--17). As another example, the total bacterial load of the vaginal
 microbiome differs by 1-2 orders of magnitude in absolute abundance
-between the healthy and bacterial vaginosis states (Zozaya-Hinchliffe et
-al. 2010), and the composition between these states is dramatically
-different (Ravel et al. 2011; Hummelen et al. 2010). Thus, a full
-description of these systems includes both relative change (composition)
-and absolute abundance (scale), yet current methods access only the
-compositional information.
-
-To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix
-of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number
-of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or
-gene) in sample \(n\). Let \(\mathbf{W}\) also represent a
-\(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing
-the true (as opposed to measured) amount of class \(d\) in the
-biological system from which sample \(n\) was obtained. Succinctly,
-\(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\)
-denotes the true counts in the system under study. We can think of
-\(\mathbf{W}\) as consisting of two parts, the scale
-\(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
+between the healthy and bacterial vaginosis states (18), and the
+composition between these states is dramatically different (19, 20).
+Thus, a full description of any of these systems includes both relative
+change (composition) and absolute abundance (scale). Current methods
+access only the compositional information yet make implicit assumptions
+about the scale.
+
+Recently, Nixon et al. (8) showed that the challenge of non-biological
+variation in sequencing depth be viewed as a problem of
+partially-identified models. They showed that \emph{all} normalizations
+make some assumption about scale. These implicit assumptions are often
+difficult to interpret, and different normalizations provide different
+outputs when applied to the same dataset (3, 21--24). Intuitively,
+normalizations in widespread use assume that either all samples have the
+same scale, e.g.~proportions, rarefaction (25), RPKM (26, 27), etc; or
+that a subset of features in one sample can be chosen as a reference to
+which the others are scaled e.g.~the TMM (28), or LVHA (7) or the
+additive log-ratio (29); or that different sub-parts of each sample
+maintain a constant scale across samples e.g.~the RLE (30); or that the
+geometric mean of the parts is appropriate e.g the CLR and its
+derivatives.
+
+The original ALDEx2 (31) model unwittingly made a strict assumption
+about scale through the CLR normalization (8). While this assumption
+could be useful in many cases it could never be exactly true, and others
+have shown that it is not always the best option (32). Nixon et al. (8)
+showed that better scale assumptions resulted in more reproducible data
+analysis including better control of both false positive and false
+negative results. In essence ALDEx2 has been modified to explicitly
+model the scale over a range of reasonable normalization parameters.
+Here, we briefly introduce these modifications and then show how scale
+uncertainty can greatly improve modeling in transcriptome and
+meta-transcriptome datasets to provide more robust and reproducible
+results.
+
+\section{Implementation}\label{implementation}
+
+To be concrete, we let \(\mathbf{Y}\) denote the \emph{measured}
+\(D \times N\) matrix of sequence counts with elements
+\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
+mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
+Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount of
+class \(d\) in the biological system from which sample \(n\) was
+obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
+scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
 \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
@@
 \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
-The technical variation in sequencing depth
+
+All the normalizations in current use are data transformations that can
+be stated as ratios of the form
+\(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
+denominator is determined by some function of the observation. We use
+the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization
+is attempting to provide an estimate of the true value. The technical
+variation in sequencing depth
 (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
@@
 
-Early on, researchers recognized the potential confounding effects of
-the non-biological variation in sequencing depth and approached the
-problem using various forms of \textit{normalization}. The idea of
-normalization is to remove this variation thus making the samples
-commensurate and to correct any minor asymmetries in the data. A large
-number of normalizations were developed. Rarefaction (Hughes and
-Hellmann 2005) was used in the microbiome field, while transcriptome
-investigators developed and used proportions and derivatives (reads per
-kilobase per million, and transcripts per kilobase per million)
-(Mortazavi et al. 2008; Wagner, Kin, and Lynch 2012), the relative log
-expression (RLE) (Anders and Huber 2010) and trimmed mean of M values
-(TMM) (Robinson and Oshlack 2010). The Centered Log Ratio (CLR)
-(Aitchison 1982) was proposed as a unifying normalization (Fernandes et
-al. 2014). With the exception of rarefaction, all of these
-normalizations are data transformations that can be stated as ratios of
-the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the
-major differences being how the denominator is chosen and whether the
-ratio is subsequently always log-transformed or not.
-
-Recently, Nixon et al. (2023) suggested that the challenge of
-non-biological variation in sequencing depth be viewed as a problem of
-partially-identified models. Through this lens, they showed that
-\emph{all} normalizations make some assumption about the system scale
-\(\mathbf{W}^{\perp}\) in order to identify the chosen model. Moreover,
-these assumptions are often implicit, difficult to interpret, and
-provided different outputs when applied to the same dataset (Dillies et
-al. 2013; Weiss et al. 2017). Intuitively, normalizations in widespread
-use assume that either one sample can be chosen as a reference to which
-the others are scaled (TMM), or that different sub-parts of each sample
-maintain a constant scale across samples (RLE, CLR and derivatives).
-Further, Nixon et al. (2023) showed that the original ALDEx2 model made
-a strict assumption about scale through the CLR normalization. They
-showed that this assumption could be remedied since ALDEx2 permits scale
-to be easily incorporated into its framework. In this paper, we discuss
-these modifications and show how including scale uncertainty can greatly
-improve modeling and lead to more robust and reproducible results.
-
-\hypertarget{adding-scale-uncertainty-in-aldex2}{%
 \subsection{Adding Scale Uncertainty in
-ALDEx2}\label{adding-scale-uncertainty-in-aldex2}}
-
-The ALDEx2 R package (Fernandes et al. 2013) is a general purpose
-toolbox for Bayesian modeling of HTS data. For brevity and concreteness,
-we discuss ALDEx2 in its simplest form as a tool for estimating
-Log-Fold-Changes (LFC) (e.g., differential abundance or differential
+ALDEx2}\label{adding-scale-uncertainty-in-aldex2}
+
+The ALDEx2 R package (31, 33) is a general purpose toolbox for Bayesian
+modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
+form as a tool for estimating the magnitude and statistical significance
+of Log-Fold-Changes (LFC) (e.g., differential abundance or differential
 expression analysis), but note that it can be used to fit more complex
 linear models. At a high-level, ALDEx2 involves three steps: 1) a
-Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given
-observations \(\mathbf{Y}\); 2) a normalization is used to estimate
-\(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the
-\(\mathbf{W}\) estimates are used to estimate LFC as part of
-differential abundance (or expression) analyses. For more details on
-ALDEx2 see Fernandes et al. (2013).
+Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\)
+given observations \(\mathbf{Y}\); 2) a normalization is used to
+estimate \(\mathbf{\hat{W}}\) given the estimate
+\(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates
+are used to estimate LFC as part of differential abundance (or
+expression) analyses. For more details on ALDEx2 see (8, 31).
 
 By default, ALDEx2 uses the CLR normalization which can be written as
-\(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where
-\(G_{n}\) denotes the geometric mean of the vector
-\((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental
-Materials for how the CLR connects to Shannon entropy). Nixon et
-al.~(2023) showed that in this context the CLR normalization equates to
-an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\). They
-showed that even slight errors in this assumption introduced bias into
-LFC estimate that was not accounted for in estimates of uncertainty
-(e.g., confidence intervals or credible sets). In fact, they showed that
-the only way in which the ALDEx2 model, or any model that uses
-normalizations, could ever be calibrated (e.g., control Type-I Error
-rates) was if this assumption was exactly true. While this assumption
-may be approximately correct in some instances it is surely never
+\(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\)
+where \(G_{n}\) denotes the geometric mean of the vector
+\((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR
+normalization equates to an implicit assumption that
+\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)(8) and even slight errors in
+this assumption introduces bias into the LFC estimate that is not
+accounted for when estimating uncertainty (e.g., confidence intervals or
+credible sets). In fact, Nixon et al. (8) showed that the only way in
+which the ALDEx2 model, or any normalization-based model, could ever be
+calibrated (e.g., control Type-I Error rates) was if this assumption was
 exactly true. In support of this assertion it is known that false
-discovery rates vary widely by analysis and normalization method
-(Hawinkel et al. 2018; Li et al. 2022). As a result, Nixon et al.~(2023)
-suggested that models should consider potential error in the assumptions
-implied by normalizations.
-
-Nixon et al.~(2023) proposed the concept of a \textit{scale model} as a
-generalization of normalizations that account for potential error. They
-showed how scale models can be incorporated into ALDEx2, turning the
-ALDEx2 model into a specialized type of statistical model which they
-called a \textit{Scale Simulation Random   Variable} (SSRV). Recall that
-\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
-As the HTS data only informs directly on
-\(\mathbf{W}^{\parallel}_{dn}\), they proposed to augment analyses with
-a model for \(\mathbf{W}^{\perp}_{n}\) (the scale model). Since the CLR
-normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\),
-the CLR normalization can be generalized by considering probability
-models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean
-\(1/G_{n}\). For example, the following scale model generalizes the CLR:
-
-\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
-
-where \(\gamma\) is a tunable parameter that controls the degree of
-uncertainty in the CLR assumption and \(x_{n}\) denotes a binary
-condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\)
-denotes control). Beyond theoretical studies, they demonstrated through
-the analysis of simulated and real data the remarkable performance
-improvements that could result from incorporating this type of scale
-uncertainty into analyses. Subsequently, we have made those
-modifications a permanent fixture of ALDEx2 which now represents the
-first software package designed for SSRV-based inference.
-
-Beyond concerns of fidelity and rigor, scale models also enhance the
-reproducibility and transparency of HTS analyses. This occurs because
-rather than using an implicit assumption (e.g.,
-\(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale
-models make this an explicit part of the model building process. While
-it is beyond the scope of the present article, we note that there are
-many ways of building scale models that enhance the interpretability of
-the parameters and assumptions. Moreover, the scale model listed above,
-which is now the default in ALDEx2, also has an appealing
-interpretation. A detailed description of these points as well as the
-mathematical underpinnings of the updated ALDEx2 R package is
-forthcoming and will be made available as a companion publication. Here,
-we simply treat the parameter \(\gamma\) in the default scale model as
-an abstract term that represents uncertainty in the CLR assumption and
-note that larger values of \(\gamma\) correspond to more uncertainty in
-this assumption. We also show that the ratio of scales between
-conditions can be used to build a full scale model when the conditions
-have dramatically different underlying scales.
-
-An advantage of incorporating scale is that analyses can be made much
-more robust such that actual or potential differences in scale can be
-tested and accounted for explicitly. The examples below and in the
-supplement show how incorporating scale provides robust and
-interpretable differential abundance estimates in several different
-datasets.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}{%
+discovery rates vary widely by analysis, dataset, and normalization
+method (34, 35). To fix this, Nixon et al. (8) showed that models should
+incorporate potential error in the assumptions implied by
+normalizations.
+
+Nixon et al.~(2023) generalized the concept of normalizations by
+introducing the concept of a \textit{scale model} to account for
+potential error. Scale models can be incorporated into ALDEx2, turning
+the ALDEx2 model into a specialized type of statistical model which they
+called a \textit{Scale Simulation Random   Variable} (SSRV). They did
+this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). Since
+the CLR normalization makes the assumption
+\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be
+generalized by considering probability models for the scale
+\(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
+example, the following scale model generalizes the CLR:
+
+\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
+
+where \(\gamma\) is a tunable parameter drawn from a Gaussian
+distribution that controls the degree of uncertainty in the CLR
+assumption and \(x_{n}\) denotes a binary condition indicator (e.g.,
+\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). We have made
+those modifications a permanent fixture of ALDEx2 which now represents
+the first software package designed for SSRV-based inference.
+
+\section{Results}\label{results}
+
 \subsection{Adding scale uncertainty replaces the need for dual
 significance
-cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}}
-
-Gierliński et al. (2015) conducted a highly replicated yeast
-transcriptome experiment comparing a wild-type strain with a snf1 gene
-knockout. This dataset has been used to argue that only a small number
-of replicates are needed to identity differentially abundant genes and
-makes the recommendation that some tools should be used in some
-conditions because each tool has different intrinsic statistical power
-and Type 1 and Type 2 errors (Schurch et al. 2016){]}. This guidance
-runs counter to standard statistical practice where power is
-intrinsically linked to sample size (Halsey et al. 2015), yet the
-concept of sample-size independent power is entrenched in all fields
-that use HTS as an experimental readout. Through the lens of this paper,
-this is a hallmark of unacknowledged bias: a false certainty in the
-scale of the data brings about an increasing false precision with
-increasing sample sizes. Furthermore, the yeast transcriptome experiment
-has an unacknowledged difference in scale between conditions: yeast
-deficient for snf1 are smaller, grow more slowly and are sensitive to a
-variety of common agents that cause cell stress (Yoshikawa et al.
-2011){]}. Thus, by adding scale uncertainty, we can be more confident
-that the results of the analysis reflect true biological differences and
-not differences due to the assumptions made by the normalization method.
-
-We start by examining the dispersion (variance) of the data as counts
-and as the logarithm of the normalized counts when calculated by DESeq2
-(log(RLE)) or by ALDEx2 (CLR). The raw data counts derived from
-sequencing are overdispersed with the mean value being less than the
-variance (Robinson and Oshlack 2010; Robinson, McCarthy, and Smyth 2010)
-as seen in Panel A of Figure 1. However, the actual analysis of
-differential abundance is performed on the logarithm of the normalized
-counts (Robinson, McCarthy, and Smyth 2010; Love, Huber, and Anders
-2014; Andrews and Hemberg 2019), which includes the CLR (Fernandes et
-al. 2013) yet the mean-dispersion distribution of these log-ratio
-transformed data is quite different as shown in panels B and C of Figure
-1 and as noted elsewhere (Fernandes et al. 2013; Love, Huber, and Anders
-2014).
+cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}
+
+It is standard practice in many fields of HTS, but especially
+transcriptomics, to use the dual cutoff approach graphically exemplified
+by volcano plots (36, 37) because not all statistically significant
+differences are biologically relevant. Paraphrasing this we can say that
+in in some types of datasets more samples leads to a majority of
+features being statistically significant. A detailed reasoning for this
+using modelled data is explained in two reports by Nixon et al. (8, 9),
+which shows that scaled models completely address this analytic issue.
+
+We use the data of Gierliński et al. (38) who conducted a highly
+replicated yeast transcriptome experiment comparing a wild-type strain
+with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to
+argue that a dual cutoff approach is appropriate to limit the number of
+significant parts and for purposes of reproducibility (37). However, in
+this benchmarking study, the number of significantly different
+transcripts varied between 65\% to \textgreater80\% of all transcripts
+depending on the tool; in essence the desired behavior was that almost
+all transcripts were significantly different. The guidance on
+reproducibility runs counter to standard statistical practice where
+power is intrinsically linked to sample size and very large sample sizes
+are indeed desired (39). Furthermore, the dual-cutoff approach is known
+not to provide appropriate FDR control (5, 6). Schurch et al. (37)
+further suggested that different tools might be better in some
+conditions or datasets than others because each tool has different
+intrinsic statistical power and Type 1 and Type 2 errors. Through the
+lens of scale uncertainty, the behaviour of the tools in this study show
+unacknowledged bias; false confidence in the precision of the estimate
+as sample size increases because o mis-specification in the scale of the
+data. This certainty is driven by the assumptions of the tool not the
+actual experiment being investigated (9).
+
+Using either DESeq2 or ALDEx2, a majority of transcripts are
+statistically significantly different between groups with a
+Benjamini-Hochberg (40) false discovery rate (FDR) of 0.05; i.e.~4636
+(79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
+large numbers of significant transcripts seems biologically unrealistic
+and furthermore breaks the necessary assumption made by the
+normalization methods that the majority of the features must be
+invariant. That 118 transcripts are identified by ALDEx2 and not DESeq2,
+while DESeq2 identifies 582 transcripts that ALDEx2 does not, suggests
+that the choice of normalization plays a role in which results are
+returned as significant and that some if not the majority are driven by
+technical differences in the analysis (22, 23, 37, 41, 42).
 
@@
 \centering
-\includegraphics{go3_files/figure-latex/dispersion-1.pdf}
-\caption{Plot of abundance v dispersion for a typical transcriptome
-dataset as counts, as logarithms of counts, and as clr values. Panel A
-shows that the data are over-dispersed relative to a Poisson
-distribution which is represented by the dashed line when plotted on a
-log-log scale. Panel B shows that the relationship between the mean and
-the dispersion calculated in DESeq2, here the standard error (SE) of the
-mean, is very different when the data are log-transformed first. Panel C
-shows the equivalent values calculated by ALDEx2 in which the expected
-clr value for each transcipt are plotted vs.~the expected dispersion.
-The red line in each panel shows the loess line of fit to the mid-point
-of the distributions. In panels B and C the amount of dispersion reaches
-a minimum at moderate values. The dashed orange line in panel A is the
-line of equivalence, and in panel B and C is the minimum y value. The
-values below the dashed grey line in panels B and C represent those
-below the first decile of dispersion.}
+\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/plot1-1.pdf}}
+\caption{Effect and volcano plots for unscaled and scaled transcriptome
+analysis. ALDEx2 was used to conduct a differential abundance (DA)
+analysis on the yeast transcriptome dataset. The results were plotted to
+show the relationship between difference and dispersion using effect
+plots ) or difference and the Benjamini-Hochberg corrected p-values
+(volcano plot). Panels A,C are for the unscaled analysis, and Panels B,D
+are for the scaled analysis. Each point represents the values for one
+transcript, with the color indicating if that transcript was significant
+in the scaled analysis and unscaled analysis (red) or in the unscaled
+analysis only (orange). Points in grey are not statistically
+signficantly different with any analysis. The horizontal dashed lines
+represent a log2(difference) of \(\pm 1.4\).}
 \end{figure}
 
-These relationships depend on what is being analyzed: the variance or
-dispersion always increases with increasing raw (or normalzied) read
-count but decreases when measured on the log-ratio transformed data
-(Fernandes et al. 2013; Love, Huber, and Anders 2014), reaching a
-minimum at some mid-point of the distribution. This makes the
-counter-intuitive suggestion that genes with moderate expression have
-more predictable expression than genes with very high expression such as
-housekeeping genes. This is at odds with the known biology of cells
-where single cell counting of housekeeping transcripts shows that they
-are both highly expressed and have little intrinsic variation (Taniguchi
-et al. 2010). Furthermore, the dispersion is exceeding small being, for
-many transcripts, almost negligible. To show this point more clearly,
-the majority of the transcripts in the lowest decile of dispersion
-indicated below the dashed grey line are statistically significantly
-different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low
-dispersion estimates lead to many false positives. From the perspective
-of Nixon et al. (2023), it is reasonable to conclude that some of these
-results may be due to the choice of normalization. Indeed, benchmarking
-and comparison studies have repeatedly shown that the choice of
-normalization plays a role in which results are returned as significant;
-see for example (Maza et al. 2013), (Dillies et al. 2013) and (Weiss et
-al. 2017).
-
-Using either DESeq2 or ALDEx2, a majority of transcripts are
-statistically significantly different between groups even with a
-Benjamini-Hochberg (Benjamini and Hochberg 1995) false discovery rate
-(FDR) of 0.01; i.e.~4264 or 3790 of the 5891 transcripts are
-statistically significant. Clearly, a 70\% positive rate seems
-biologically unjustified, and furthermore this breaks the necessary
-assumption made by the normalization methods that most features must be
-invariant. Finally, that 90 transcripts are identified by ALDEx2 and not
-DESeq2, while DESeq2 identifies 564 transcripts that ALDEx2 does not
-suggests that the choice of normalization plays a role in which results
-are returned as significant.
+The effect plots (43) in Figure 1A (ALDEx2) and Supplementary Figure 1
+(DESeq2) shows that the majority of significant transcripts (red,
+orange) have negligible differences between groups and very low
+dispersion and we suggest that this is driven by the experimental design
+(37). Scale uncertainty can be incorporated using the \texttt{gamma}
+parameter that controls the amount uncertainty added to the CLR mean
+assumption when we call either \texttt{aldex()}, or
+\texttt{aldex.clr()}. Figure 1B shows that setting \(\gamma=0.5\)
+results in far fewer transcripts being significant (205) and we observe
+that the minimum dispersion increases from 0.12 (unscaled) to 0.67
+(scaled). The Volcano plots in Figure 1 C and D show a similar story.
+Here we can see that adding scale increases the minimum FDR value and
+increases the concordance between the FDR value and the difference
+between groups (compare panels C and D).
 
@@
 \centering
-\includegraphics{go3_files/figure-latex/plot1-1.pdf}
-\caption{Effect and volcano plots for unscaled and scaled transcriptome
-analysis. DESeq2 or ALDEx2 were used to conduct a differential abundance
-(DA) analysis on the yeast transcriptome dataset. The results were
-plotted to show the relationship between difference and dispersion using
-effect plots ) or difference and the Benjamini-Hochberg corrected
-p-values (volcano plot). Panels A,B,D,E are for the unscaled analysis,
-and Panels C,F are for the scaled analysis. Each point represents the
-values for one transcript, with the color indicating if that transcript
-was significant in the scaled analysis and unscaled analysis (red) or in
-the unscaled analysis only (orange). Points in grey are not
-statistically signficantly different with any analysis. The horizontal
-dashed lines represent a log2(difference) of 1.4, which is a commonly
-applied cutoff when the majority of features are statistically
-significant.}
+\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/disp-1.pdf}}
+\caption{Adding scale uncertainty changes the dispersion distribution.
+Panel A shows a plot of the expected value for relative abundance vs the
+expected value for the pooled dispersion as output by
+\texttt{aldex.effect}. The dashed horizontal lines show the median value
+for the features with a rAbundance between -0.5 and 0.5, and the light
+colored lines are lowess lines of fit through the center of mass of the
+data. Panel B plots the dispersion difference between \(\gamma = 1\) and
+\(\gamma = 0\); note the non-linear relationship that highlights the
+rotation that is evident in Panel A. The colored lines indicate the
+lowess line of fit through the centre of mass of the plot for the
+various populations of points. The grey line is the total population and
+shows the difference \(\Delta\), the red line is the population of
+significant transcripts (*) with no scale, the orange line is the
+population of significant transcripts with a difference threshold (T) of
+about \(\pm 2^{1.4}\)-fold change, the blue line is the population of
+significant transcripts with \(\gamma = 0.5\), and the cyan line is the
+significant population with \(\gamma = 1\). \(\Delta\): Difference, *:
+significant, T: thresholded.}
 \end{figure}
 
-As shown using effect (Gloor, Macklaim, and Fernandes 2016) and volcano
-plots (Cui and Churchill 2003) in Figure 2 A,B,D,E, the root cause of
-the many statistically significant positive transcripts is the very
-large number of transcripts with negligible variance. This phenomenon is
-shared between methods: almost all the transcripts that are
-differentially abundant with an FDR \textless 0.01 (orange and red
-points) have extremely low dispersion and a very low difference between
-groups. In the most extreme cases, some transcripts have a near zero
-difference between groups. This issue is not unique to this data set and
-it is common practice to use a dual-cutoff by choosing transcripts based
-on a thresholds for both corrected p-values and fold-changes (Schurch et
-al. 2016) (commonly set at \(\pm 2^{1.4}\) for the latter). These limits
-are shown by the dashed grey lines. In this case, applying a dual-cutoff
-using a heuristic of at least a \(2^{1.4}\) fold change reduces the
-number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
-This result begs the question: why bother with significance tests at
-all?
-
-The very low dispersion estimate for most features is a by-product of
-sequencing and normalization. While the actual scale of the data is
-inaccessible post-sequencing, we can study the sensitivity of the
-results to the choice of normalization and thus scale. To do this we
-incorporate scale models (Nixon et al. 2023) in lieu of normalizations
-through the default scale model that is now built into ALDEx2. Scale
-uncertainty is incorporated using the \texttt{gamma} parameter that
-controls the amount uncertainty added to the CLR mean assumption when we
-call either \texttt{aldex()}, or \texttt{aldex.clr()}. The
-\texttt{ALDEx2} package further contains a sensitivity analysis
-function, \texttt{aldex.senAnalysis()}, that can be used to explore the
-effect of different amounts of scale uncertainty. In practice, we
-suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic
-for most experimental designs.
-
-Applying the default scale model with \texttt{gamma=1}, a large number
-of transcripts with near 0 dispersion have had their dispersion
-increased (Figure 2C), and this results in many fewer transcripts 100
-being called significantly different as shown in the volcano plot in
-Figure 1F (red points). Furthermore, overplotting the significant
+As shown by the effect plot in Figure 1 the root cause of the many
+statistically significant positive transcripts is the very large number
+of transcripts with negligible variance. This phenomenon is not unique
+to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon
+occurs in DESeq2 (and presumably other methods although the relevant
+parameters are not exposed). This issue is not unique to this dataset
+and it is common practice to use a dual-cutoff by choosing transcripts
+based on a thresholds for both corrected p-values and fold-changes (37)
+(here set at \(\pm 2^{1.4}\) for the latter), although considerable
+variation in cutoff values is observed. These limits are shown by the
+dashed grey lines. Here, applying a dual-cutoff using a heuristic of at
+least a \(2^{1.4}\) fold change reduces the number of significant
+outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with
+that observed with that found by ALDEx2 with \(\gamma = 0.5\) which
+identifies 205. Indeed, Supplementary Figure 2, shows that even adding a
+very small amount of scale \(\gamma = 0.1\) reduces the number of
+significant transcripts by more than half and the
+\texttt{aldex.scaleSim()} function can be used to identify those
+transcripts that are significant only because of an absence of scale.
+These results begs the question: why bother with significance tests at
+all if all transcripts with \(> 1.4\)-fold expression change are
+statistically significant?
+
+The effect on dispersion with increasing amounts of scale are shown in
+Figure 2A, where we can see that the dispersion increases as scale is
+added. Note that the dispersion in the unscaled data in Figure 2A
+reaches a minimum near the mid-point of the distribution, and also does
+so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
+This shows more clearly that dispersion of many transcripts is almost
+negligible in the absence of scale and makes the counter-intuitive
+suggestion that the variance in expression of the majority of genes with
+moderate expression is more predictable than highly-expressed genes or
+of housekeeping genes (44). This is at odds with the known biology of
+cells where single cell counting of highly-expressed transcripts shows
+that they have little intrinsic variation (45).
+
+Adding scale by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases
+the minimum dispersion as shown in Figure 2A by the red and blue data
+points, and by the colored lines of fit through the centre of mass of
+the data. Less obvious is that the scaled dispersion estimates are
+rotated. Figure 2B shows a plot of the difference between the
+\(\gamma= 0\) and \(\gamma= 1\) data to show this more clearly and here
+we can see that the scale is preferentially increasing the dispersion of
+the mid-expressed transcripts that formerly had negligible dispersion;
+examine the grey line of best fit (overlaid by the red line) for the
+trend. Panel B also shows the trend of the expression-dispersion
+relationship for transcripts that are classed as statistically
+significant. The red line shows the trendline with no added scale, and
+this trendline exactly overlays with the grey trendline of the bulk of
+transcripts. The orange trendline indicates those transcripts that are
+both statistically significant and that have a thresholded expression
+level of \(\pm 1.4\), and the dark blue and cyan lines show the
+statistically significant trendline for \(\gamma=0.5\) , or \(1.0\).
+Note that this has the effect of changing the distribution of parts
+identified as significant and that the substantially fewer significantly
+different genes are in the very high abundance but low dispersion
+category.
+
+\subsection{Housekeeping genes can be used to guide scale model
+choices.}\label{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}
+
+Dos Santos et al. (46) used a vaginal metatranscriptome dataset to
+compare the gene expression in bacteria collected from healthy (H) and
+bacterial vaginosis (BV) affected women. In this environment, both the
+relative abundance of species between groups and the gene expression
+level within a species is different (47). Additionally, prior research
+suggests that the total number of bacteria is about 10 times more in the
+BV than in the H condition (18). Thus, this is an extremely challenging
+environment in which to determine differential abundance as there are
+both compositional and scale changes between conditions. The usual
+method to analyze vaginal metratranscriptome data is on a taxon-by-taxon
+basis (47--49) because the scale confounding can be ignored. Attempts at
+system-wide analysis show that many housekeeping functions are returned
+as differentially abundant between groups; a result likely due to a
+disconnect between the scale assumptions of the normalization used (7).
+
+In this example, we show how to specify a user-defined, or
+\emph{informed}, scale model explicitly that can account for some of
+these modeling difficulties. An informed scale model can control for
+both the mean difference of scale between groups (e.g., directly
+incorporate information on the differences in total number of bacteria
+between the BV and H conditions) as well as the uncertainty assumed in
+that difference. To specify a user-defined scale model, we can pass a
+matrix of scale values instead of a single estimate of gamma to
+\texttt{aldex.clr()}. This matrix should have the same number of rows as
+the of Monte-Carlo Dirichlet samples, and the same number of columns as
+the number of samples. While this matrix can be computed from scratch by
+the analyst, there is an \texttt{aldex.makeScaleModel()} function that
+can be used to simplify this step in most cases. This encodes the scale
+model as \(\Lambda \sim N(log2 \mu_n, \gamma^{2})\), where \(\mu_n\)
+represents the actual scale value for each sample. This can be a
+measured value (cell count, nucleic acid input, etc), or an imputed
+value. Nixon et al. (9) showed that only the ratio between the scale
+values for each sample was important; see Supplementary Figures 4 and 5
+for a demonstration of this point.
+
+Figure 3A shows an effect plot of the data where reads are grouped by
+function, corresponding to grouping homologous sequences regardless of
+the organism of origin. Each point represents one of 3728 KEGG functions
+(50). There are many more functions represented in the BV group (bottom)
+than in the healthy group (top). This is because the
+\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
+reduced genome content relative to the anaerobic organisms that dominate
+in BV, because there is a greater diversity of organisms in BV than in H
+samples, and because the BV condition has at least an order of magnitude
+more bacteria than does the H condition.
+
+There are 101 functions with low dispersion that appear to be share by
+both groups (boxed area in Figure 3A, and colored in cyan). Inspection
+shows that these largely correspond to core metabolic functions such as
+transcription, translation, ribosomal proteins, glycolysis, replication,
+chaperones, etc (Supplementary file housekeeping.txt). The transcripts
+of many of these are commonly used as invariant reference sequences (44)
+and so would not be expected to contribute to differences in ecosystem
+behaviour and should be centred on 0 difference. These should not be
+scored as among the most differentially abundant. The major group of
+these housekeeping functions is located off the line of no difference
+(being approximately located at +1.5). While changes in the abundance of
+housekeeping functions is a useful proxy for relative abundance of
+species in the environment, they tell us nothing about the functional
+capacity of the two groups as these are common to every organism. Of
+more interest is determining the functions that are different between
+groups because these are unique or over-expressed in one group relative
+to the other.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/meta-1.pdf}}
+\caption{Analysis of vaginal transcriptome data aggregated at the Kegg
+Orthology (KO) functional level. Panel A shows an effect plot for the
+default analysis where the functions that are elevated in the healthy
+individuals have positive values and functions that are elevated in BV
+have negative values. Highlighed in the box are KOs that are almost
+exlusively housekeeping functions and are colored cyan. These
+housekeeping functions should be located on the midline of no
+difference. Panel B shows the same data scaled with \(\gamma = 0.5\),
+which increase the minimum dispersion as before. Panel C shows the same
+data scaled with \(\gamma = 0.5\) and a 0.15 fold difference in
+dispersion applied to the BV samples relative to the H samples. In these
+plots statistically significant (FDR \textless{} 0.01) functions in the
+informed model are in red, false positive functions are in blue,
+non-significant functions in black and false negative functions are in
+orange.}
+\end{figure}
+
+Applying the default scale model of \(\gamma=0.5\) increases the
+dispersion as expected but does little to move the large number of
+housekeeping functions toward the midline of no difference. This is as
+expected; the mean of the default scale model is based on the CLR
+normalization so no shift in location would be expected over the
+original ALDEx2 model. Nevertheless, about 30\% of the housekeeping
+functions are no longer statistically significantly different. Note that
+this change is simple to conduct, has no additional computational
+complexity and requires only a slight modification from the analyst.
+
+Up to this point, scale uncertainty has been applied as an extension of
+the CLR normalization via the default scale model, but a user-defined
+scale adjustment can be applied to each condition, or even each sample
+independently through a custom scale matrix. Our starting point for this
+is to identify the naive scale estimate from the data, and this can be
+done by calculating the mean scale value for each group. The scale
+estimate for samples can be accessed in the \texttt{@scaleSamps} slot
+from the \texttt{aldex.clr} output. In this datset the scale estimate
+for the healthy group is 17.41 and for the BV group is 14.59 for a
+difference of 2.82. This is interpreted as the scale of the H group of
+samples being 7.06 greater than the BV group and this precise but
+incorrect estimate places the location of the housekeeping genes off the
+midline of no difference.
+
+We desire a scale model that approximately centres the housekeeping
+functions. We anticipate that housekeeping functions should be nearly
+invariant and so an appropriate scale in this dataset is likely closer
+to 0 than the naive estimate. While a user-defined scale model can be
+quite flexible with relative scales that are distinct for each group (or
+even each sample) along with their uncertainties, here we focus on using
+the \texttt{aldex.makeScaleMatrix()} function. This function uses a
+logNormal distribution to build a scale matrix given a user-specified
+mean difference between groups and uncertainty level. Applying a
+per-group relative differential scale of 0.15 moves the housekeeping
+functions to the midline of no difference (Figure 3C), and applying a
+gamma of 0.5 provides the same dispersion as in panel B). Note that now
+a significant number of functions are differentially up in BV that were
+formerly classed as not different without the full scale model (orange),
+or when only a default scale was applied. Inspection of the functions
+shows that these are largely missing from the \emph{Lactobacillus}
+species and so should actually be captured as differentially abundant in
+the BV group. Thus, applying a differential scale allows us to
+distinguish between both false positives (housekeeping functions in
+cyan, and others in blue) and false negatives (orange functions) even in
+a very difficult to analyze dataset. The remarkable improvements in
+biological interpretation afforded by this full scale model, and the
+transferrability of it between sample cohorts of the same condition is
+outlined elsewhere (46). We suggest that the default scale model is
+sufficient when the data are approximately centred. However, an informed
+model is more appropriate with datasets are not well centred or when the
+investigator has prior information about the underlying biology.
+
+\section{Discussion}\label{discussion}
+
+\doublespacing
+\singlespacing
+
+Biological systems are both predictably variable and stochastic (45) and
+current measurement methods that rely on high throughput sequencing fail
+to capture all of that variation, particularly variation due to scale
+(8, 9). In the absence of external information (such as spike-in probes
+(14), cell counting (1), FISH (51) etc) sequencing depth normalisation
+methods cannot recapture the scale information (14), and can only
+normalize for the technical variation due to sequencing depth.
+
+Many groups have conducted benchmarking studies on different tools and
+normalizations used for the analysis of datasets such as
+transcriptomes(4, 21, 35, 37, 41) and microbiomes (3, 23, 24, 32, 34,
+52). Generically, it is observed that the actual agreement between
+analysis methods can be modest, and no single method outperforms all
+others in every dataset or type of experiment. That different tools
+appear to work more reliably in different datasets from different
+sources can be explained as different normalizations being a better fit
+to the scale of a particular dataset by chance.
+
+In the analysis of HTS data it is often observed that larger datasets
+converge on the majority of parts being significantly different (8, 35,
+37), and that different analytic approaches result in different parts
+being chosen as significantly different (4, 24, 37, 41). Li et al. (35)
+conducted a permutation-based benchmarking study and found that popular
+tools performed worse than simple Wilcoxon rank-sum tests in controlling
+the FDR when sample sizes became large. They suggested that the presence
+of outliers were one of the factors driving this observation. Brooks et
+al. (53) suggested that inappropriate choice of benchmarking methods are
+also a major contributing factor. From our perspective, the disagreement
+between tools can be explained by the partially identifiable nature of
+the data; that is, the observed data are consistent with multiple ways
+of estimating the parameters (8). Partially identifiable data exhibit
+multiple pathologies, chief among them being that that different
+analytic approaches can produce different parameter estimates and that
+more data produces worse estimates because the additional data increases
+the precision of a flawed estimate (54).
+
+Scale simulation is now build into ALDEx2 to address the issue of
+partially identifiability, and through this lens the two main root
+causes to common HTS data pathologies can be proposed. The first
+contributing factor is the observed very low dispersion estimate for
+many features that is a by-product of experimental design, sequencing
+and normalization. In the Schurch et al. (37) dataset, the data were
+from single colonies derived from a single culture. Thus, it is more
+accurate to describe the 96 samples as wet-lab technical replicates
+rather than independent samples. However, this type of replication
+approach is standard in the molecular literature, and would be expected
+to result in the very low dispersion that is observed.
+
+In the yeast transcriptome dataset, applying the default scale model
+with \(\gamma=0.5\) a large number of transcripts with near 0 dispersion
+have had their dispersion increased (Figure 1D), and this results in
+modest number of transcripts, 205, being called significantly different
+as shown in the volcano plot in Figure 1D (red points). In addition,
+there was now a strong concordance between the difference and p-values
+(Figure 1D), this is not surprising. In hindsight, what is not obvious
+is why the unscaled volcano plot shows such poor correspondence. We
+suggest that this can be explained by random fluctuations in the many
+variance estimates with very low values, and this is supported by the
+plot shown in Figure 2B. Furthermore, overplotting the significant
 transcripts identified after adding scale uncertainty on the un-scaled
@@
 
-\hypertarget{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}{%
-\subsection{Housekeeping genes can be used to guide scale model
-choices.}\label{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}}
-
-Macklaim and Gloor (2018) and Wu et al. (2021) used a vaginal
-metatranscriptome dataset used to compare the gene expression in
-bacteria collected from healthy (H) and bacterial vaginosis (BV)
-affected women. In this environment, both the relative abundance of
-species between groups and the gene expression level within a species is
-different (Macklaim et al. 2013). Additionally, prior research suggests
-that the total number of bacteria is about 10 times more in the BV than
-in the H condition (Zozaya-Hinchliffe et al. 2010). Thus, this is an
-extremely challenging environment in which to determine differential
-abundance as there are both compositional and scale changes between
-conditions. The accepted method to analyze vaginal metratranscriptome
-data is on a taxon-by-taxon basis (Macklaim et al. 2013; Deng et al.
-2018; Fettweis et al. 2019) because the scale confounding can be
-ignored. The problem with a systemic analysis is that many housekeeping
-genes are returned as differentially abundant between groups, a result
-likely due to a disconnect between the scale assumptions of the
-normalization used (Wu et al. 2021) and the actual change in scale
-between conditions.
-
-In this example, we show how to specify the scale model explicitly and
-demonstrate that applying a user-defined scale model can account for
-some of these modeling difficulties. A user-defined scale model can
-control for both the mean difference of scale between groups (e.g.,
-directly incorporate information on the differences in total number of
-bacteria between the BV and H conditions) as well as the uncertainty
-assumed in that difference. To specify a user-defined scale model, we
-can pass a matrix of scale values instead of a single estimate of gamma
-to \texttt{aldex.clr()}. This matrix should have the same number of rows
-as the of Monte-Carlo Dirichlet samples, and the same number of columns
-as the number of samples. While this matrix can be computed from scratch
-by the analyst, there is an \texttt{aldex.makeScaleModel()} function
-that can be used to simplify this step in most cases.
-
-Figure 3A shows an effect plot (Gloor, Macklaim, and Fernandes 2016) of
-the data where reads are grouped by function, corresponding
-approximately to grouping orthologous sequences regardless of the
-organism of origin. Each point represents one of the 3728 functions.
-There are many more functions represented in the BV group (bottom) than
-in the healthy group (top). This is because the \textit{Lactobacilli}
-that dominate a healthy vaginal microbiome have reduced genome content
-relative to the anaerobic organisms that dominate in BV, because there
-is a greater diversity of organisms in BV than in H samples and because
-the BV condition has at least an order of magnitude more bacteria than
-does the H condition.
-
-We can see that there are a large number of core metabolic and
-information processing functions that are shared between the two groups
-(boxed area in Figure 3A), and inspection shows that these largely
-correspond to core metabolic functions that would not be expected to
-contribute to differences in ecosystem behaviour and so should not be
-differentially abundant. As a proxy for housekeeping functions the core
-ribosome functions (blue) shows that their mean location is not centred
-on 0. The major group of these housekeeping functions is located off the
-line of no difference (being approximately located at +1.5). Since they
-have very low dispersions, these functions are identified as
-differentially abundant (red) along with many others. While changes in
-the abundance of housekeeping functions is a useful proxy for relative
-abundance of species in the environment, they tell us nothing about the
-functional capacity of the two groups because these are functions in
-common to every organism. Of more interest is determining the functions
-that are different between groups because these are unique or
-over-expressed in one group relative to the other.
-
-\begin{figure}
-\centering
-\includegraphics{go3_files/figure-latex/meta-1.pdf}
-\caption{Analysis of vaginal transcriptome data aggregated at the Kegg
-Orthology (KO) functional level. Panel A shows an effect plot for the
-default analysis where the functions that are elevated in the healthy
-individuals have positive values and functions that are elevated in BV
-have negative values. Highlighed in the box are highly abundant KOs that
-are almost exlusively housekeeping functions, with ribosomal KOs
-highlighted in blue, statistically significant (FDR \textless{} 0.01)
-functions in red, and non-significant functions in black or orange.
-These housekeeping functions should be located on the midline of no
-difference. Panel B shows the same data scaled with
-\texttt{gamma\ =\ 0.5}, which increase the minimum dispersion
-approximatly by one unit. Here the housekeeping functions from Panel A
-are colored cyan or blue for reference. Panel C shows the same data
-scaled with \texttt{gamma\ =\ 0.5} and a 0.15 fold difference in
-dispersion applied to the BV samples relative to the H samples. The
-orange functions are now statistically significant. Note that this
-shifts the midpoint of the housekeeping functions towards the midline.}
-\end{figure}
-
-Applying the default scale model of \texttt{gamma=0.5} increases the
-dispersion as expected but does little to move the large number of
-housekeeping functions toward the midline of no difference. This is as
-expected: the mean of the default scale model is based on the CLR
-normalization so no shift would be expected over the original ALDEx2
-model. Nevertheless, about 50\% of the housekeeping functions are no
-longer statistically significantly different. Note that this change is
-simple to conduct, has no additional computational complexity and
-requires only a slight modification from the analyst.
-
-Up to this point, scale uncertainty has been applied as an extension of
-the CLR normalization via the default scale model, but a user-defined
-scale adjustment can be applied to each condition, or even each sample
-independently through a custom scale matrix. While not necessary in all
-datasets (e.g., the transcriptome example discussed in the previous
-section and see the Supplementary material), it can greatly improve
-modeling in certain cases, especially when the scale is highly
-asymmetric between conditions. In order to specify a scale model from
-scratch, we need to revisit the concept that all normalizations in
-widespread use are actually ratios with the denominator implied by the
-normalization. Therefore, we can easily deviate from a certain
-normalization (e.g., the geometric mean assumption implied by the CLR)
-by specifying a total model based on the mean difference between
-conditions. While knowing the mean difference between conditions may
-seem cumbersome in practice, it is the \emph{relationship} between the
-group scale values that is important, not their raw values (Nixon et
-al.~2023). This can be illustrated quite simply by starting with the
-mean ratio in \(G\) between groups for the yeast transcriptome dataset
-which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for
-WT; their ratio being 1.17, or a 0.17-fold difference. Using this
-information, we can recapitulate the differential abundance analysis in
-Figure 2B and 2C exactly by using setting the mean denominator of group
-1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental
-Figure 2. This ratio can be adjusted to alter the mean assumption placed
-on the group scale values.
-
-With this background, we can understand how to apply a user-defined
-scale model to the metratranscriptome dataset. While a user-defined
-scale model can be quite flexible with relative scales that are distinct
-for each group (or even each sample) along with their uncertainties,
-here we focus on using the \texttt{aldex.makeScaleMatrix()} function.
-This function uses a logNormal distribution to build a scale matrix
-given a user-specified mean difference between groups and uncertainty
-level. Applying a per-group relative differential scale of 0.15 moves
-the housekeeping functions to the midline of no difference, and applying
-a gamma of 0.5 provides the same dispersion as using a gamma of 0.5 with
-the CLR (Figure 3C). Note that now a significant number of functions are
-differentially up in BV that were formerly classed as not different
-without scale, or when only a default scale was applied. These former
-false negatives are noted in orange in each panel. Inspection of the
-functions shows that these are largely missing from the Lactobacillus
-species and so should actually be captured as differentially abundant.
-Thus, applying a differential scale allows us to distinguish between
-both false positives (housekeeping functions in cyan) and false
-negatives (orange functions) even in a very difficult to analyze
-dataset. We suggest that the default scale model is sufficient when the
-data are approximately centred. However, when datasets are not well
-centred or when the investigator has prior information about the
-underlying biology, we advocate developing and using a user-specified
-scale model.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Biological count data derived from HTS can be decomposed into two parts:
-the relative (compositional) and the absolute (scale), and the product
-of these generates a fully scaled biological system (Nixon et al.~2023).
-Biological systems are both predictably variable and stochastic, and
-current measurement methods that rely on high throughput sequencing fail
-to capture all of that variation, particularly variation due to scale.
-In the absence of information external to the sequencing run itself, no
-normalisation method can recapture any of the scale information (Lovén
-et al.~2012). Even so, many methods such as flow cytometry (Vandeputte
-et al.~2017), spike-in probes and fluorescent in-situ hybridization
-(Lovén et al.~2012; Marguerat et al.~2012) have been proposed as a means
-to do this. In the context of the updated ALDEx2 software suite, these
-methods can be useful if they contain information on the particular
-scale the researcher is interested in.
+While the actual scale of the underlying environment is inaccessible
+post-sequencing, we can study the sensitivity of the results to the
+choice of normalization and thus scale. Varying amounts of scale
+uncertainty is incorporated using the \texttt{gamma} parameter that
+controls the amount uncertainty added to the CLR mean assumption when we
+call either \texttt{aldex()}, or \texttt{aldex.clr()}. The ALDEx2
+package contains a sensitivity analysis function,
+\texttt{aldex.senAnalysis()}, that can be used to explore the effect of
+different amounts of scale uncertainty, and an example is shown in
+Supplementary Figure 2 for the yeast transcriptome dataset. Here it is
+clear that even tiny amounts of scale preclude the majority of
+transcripts from being considered statistically significant. In
+practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
+is realistic for most experimental designs, but further work is needed.
+
+The second contributing factor is unacknowledged asymmetry in many
+datasets. In the case of asymmetry, the use of a user-specified scale
+model can be very useful for otherwise difficult-to-analyze datasets
+such as meta-transcriptomes and in-vitro selection datasets where the
+majority of features can change. We showed one such example for the
+metatranscriptome dataset in Figure 3. Here the dataset was highly
+asymmetrical, and the TMM and RLE normalizations could not fully move
+all the housekeeping genes to the midline of no difference or the tools
+exhibited other pathologies (Supplemental Figure 6). Incorporating
+differential scale on a per-group basis moves the mass of the
+housekeeping functions towards the midline of no difference and so
+affects both Type I and Type II error rates. It is also of note that in
+the case of true biological replicates (different individuals) that
+adding a modest amount of scale \(\gamma=0.5\) had little effect on the
+concordance between the difference between groups and the p-values as
+shown in Supplementary Figure 7. Thus, in this dataset the scale
+mis-specification was affecting mainly the location of the difference
+between groups.
+
+In the metatranscriptome analysis, transcripts that were previously not
+classed as differentially abundant were now called as significantly
+different, and the housekeeping transcripts move from being
+significantly different to not being identified as such. Note the
+assumption that housekeeping genes should not generally be included in
+differential abundance analysis is implicit in the dual
+p-value:fold-change cutoff approach in widespread use. While we
+acknowledge that some prior information on which housekeeping
+transcripts should not be classed as differentially abundant is needed,
+we suggest that this information is widely available and is already used
+when performing the gold-standard quantitative PCR test of differential
+abundance (55, 56). Thus, the use of this prior knowledge is not unique
+to our approach.
+
+Beyond concerns of fidelity and rigor, scale models also enhance the
+reproducibility and transparency of HTS analyses. The addition of scale
+uncertainty essentially tests the model over a range of normalizations
+(8) and so can replace the consensus approach that has been proposed by
+some groups (24, 57) with no additional computational overhead. Thus, an
+advantage of incorporating scale is that analyses can be made much more
+robust such that actual or potential differences in scale can be tested
+and accounted for explicitly. This occurs because rather than using an
+implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the
+CLR normalization), scale models can make this an explicit part of the
+model building process. While it is beyond the scope of the present
+article, we note that there are many ways of building scale models that
+enhance the interpretability of the parameters and assumptions and a
+detailed description of these points is describe elsewhere (8). In this
+work, we simply use the parameter \(\gamma\) in the default scale model
+as a term that represents uncertainty in the CLR assumption and note
+that larger values of \(\gamma\) correspond to more uncertainty in this
+assumption. In the metatranscriptome example and elsewhere (9), we also
+show that the ratio of scales between conditions can be used to build a
+full scale model when the conditions have dramatically different
+underlying scales.
 
@@
 uncertainty. Originally, this tool used the only the Dirichlet
-distribution to sample compositional uncertainty. Rather than applying
-the CLR normalization as in the original ALDEx2, scale uncertainty can
-be added through the use of a scale model with no additional
-computational complexity. Scale uncertainty can be sampled from any
-distribution depending on prior knowledge or preference. By default,
-ALDEx2 samples scale from a logNormal distribution inspired by the CLR
-normalization. However, there is the option to introduce a full scale
-model that encapsulates both uncertainty and asymmetry in underlying
-scale.
-
-All normalizations attempt to make the samples in a dataset commensurate
-but do not explicitly address the scale of the underlying system. The
-two supplementary tables show that these normalizations can be
-interpreted as scale assumptions. However, the general lack of scale
-information has important consequences for the analysis of HTS datasets.
-One issue is that analysis tools seem over-powered with even moderate
-sample sizes, and different tools have different intrinsic power, Type 1
-and Type 2 error rates (Schurch et al.~2016) and fail to account for
-false discovery rates properly (Hawinkel et al. 2018; Li et al. 2022).
-Traditional recommendations would suggest that using small sample sizes
-in analysis leads to less reliability and reproducibility in analyses
-since surprisingly large sample sizes are needed to determine
-reproducible p-values (e.g., Halsey et al. (2015)). However,
-recommendations to use small sample sizes are widespread in multivariate
-datasets such as RNA-seq datasets. We argue that this is a hallmark of
-unacknowledged bias: since normalizations are typically assumed to be
-correct with no uncertainty, smaller samples offer an unintentional
-buffer against this misspecification. Another issue is that datasets are
-difficult to analyze when they contain systematic asymmetry, with
-different tools exhibiting differing pathologies with these datasets
-(Robinson and Oshlack 2010; Wu et al. 2021).
-
-Many groups have conducted benchmarking studies on different tools and
-normalizations used for the analysis of datasets such as transcriptomes
-(Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
-Quinn, Crowley, and Richardson 2018) and microbiomes (Thorsen et al.
-2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
-observed that the actual agreement between methods can be modest, and
-there is usually a recommendation that some methods are better suited
-for individual datasets or for specific types of analysis;
-e.g.,\textasciitilde`Normalization and microbial differential abundance
-strategies depend upon data characteristics' (Weiss et al. 2017).{]}
-However, another interpretation of such studies is that different
-normalizations can be better suited to different datasets by
-happenstance, but it is not obvious how the use of specific
-normalizations could be justified on a per dataset basis. There are now
-published examples of multiple large replication efforts in psychology
-(Open Science Collaboration 2015), cancer biology (Rodgers and Collings
-2021), ecology and evolution (Gould and et.al 2023). In a nutshell,
-experiences from these studies indicate that published data typically
-have higher effect sizes and appear more significant than replication
-efforts with the same data, or with newly generated replication data
-(Rodgers and Collings 2021). Thus, most published data should be viewed
-with suspicion (Ioannidis 2005) unless efforts have been made to ensure
-that the results are robust.
-
-In the case of overpowering, HTS analyses seem to be more robust when
-applying a dual cutoff of both p-value and difference between group
-means (Schurch et al. 2016). Figure 2 shows one reason for this
-robustness could be that the dual cutoff is mimicking the effect of
-including scale uncertainty, since substantially similar transcripts are
-identified by the two approaches. We believe that adding scale
-uncertainty leads to a more transparent analysis compared to the cutoff
-as it is easy to describe the conditions under which certain conclusions
-hold. Furthermore, while using the post-hoc difference cutoff is useful
-for differential abundance analysis, it is not clear how this can be
-incorporated into other kinds of downstream analyses. Conversely,
-analyses that include scale uncertainty are fully compatible with
-existing downstream analyses.
-
-In the case of asymmetry, the use of a user-specified scale model can be
-very useful for otherwise difficult-to-analyze datasets such as
-meta-transcriptomes and in-vitro selection datasets where the majority
-of features can change. We showed one such example in Figure 3 where the
-dataset was highly asymmetrical, and the TMM and RLE normalizations
-cannot fully move all the housekeeping genes to the midline of no
-difference or exhibit other pathologies (Supplemental). Incorporating
-differential scale on a per-group basis moves the mass of the data
-towards the midline of no difference and so affects both Type I and Type
-II error rates. In this analysis, transcripts that were previously not
-classed as differentially abundant are now called as significantly
-different, and the housekeeping transcripts move from being
-significantly different to not being identified as such. While we
-acknowledge that some prior information on which housekeeping
-transcripts should not be classed as DA is needed, we suggest that this
-information is widely available and is already used when performing the
-gold-standard quantitative PCR test of differential abundance (Thellin
-et al. 1999; SEQC/MAQC-III Consortium 2014). Furthermore, the assumption
-that housekeeping genes should not generally be included in differential
-abundance analysis is implicit in the dual p-value, fold-change cutoff
-approach in widespread use. Thus, the use of this prior knowledge is not
-unique to our approach.
-
-In summary, while the underlying scale of the system is generally
-inaccessible, the effect of scale on the analysis outcomes can be
-modelled and can help explain some of the underlying biology. Adding
-scale information to the analysis allows for more robust inference
-because the features that are sensitive to scale can be identified and
-their impact on the analysis weighted accordingly. Additionally, the use
-of user-defined scale models permits difficult to analyze datasets to be
-examined in a robust and principled manner even when the majority of
-features are asymmetrically distributed or expressed (or both) in the
-groups. Thus, reporting scale uncertainty should become a standard
-practice in the analysis of HTS datasets as a way to identify which
-features are most robust to differences in the underlying system.
-Finally, we supply a toolkit that makes incorporating scale simple even
-for datasets that come from highly asymmetrical environments.
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
+distribution to sample compositional uncertainty, but scale uncertainty
+can be added through the use of a scale model with no additional
+computational complexity. By default, ALDEx2 samples scale from a
+logNormal distribution inspired by the CLR normalization. However, scale
+uncertainty can be sampled from any distribution depending on prior
+knowledge or preference, using the option to introduce a full informed
+scale model that encapsulates both uncertainty and asymmetry in
+underlying scale.
+
+In summary, we supply a toolkit that makes incorporating scale simple to
+incorporate for any type of HTS dataset. While the underlying scale of
+the system is generally inaccessible, the effect of scale on the
+analysis outcomes can be modelled and can help explain some of the
+underlying biology, and known issues with the analysis of HTS data.
+Adding scale information to the analysis allows for more robust
+inference because the features that are sensitive to scale can be
+identified and their impact on conclusions weighted accordingly.
+Additionally, the use of user-defined scale models permits difficult to
+analyze datasets to be examined in a robust and principled manner even
+when the majority of features are asymmetrically distributed or
+expressed (or both) in the groups (46). Thus, reporting scale
+uncertainty should become a standard practice in the analysis of HTS
+datasets as a way to identify which features are most robust to
+differences in the underlying system.
+
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-aitchison1982}{}}%
-Aitchison, John. 1982. {``The Statistical Analysis of Compositional
-Data.''} \emph{Journal of the Royal Statistical Society: Series B
-(Methodological)} 44 (2): 139--60.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Anders:2010}{}}%
-Anders, Simon, and Wolfgang Huber. 2010. {``Differential Expression
-Analysis for Sequence Count Data.''} \emph{Genome Biol} 11 (10): R106.
-\url{https://doi.org/10.1186/gb-2010-11-10-r106}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Andrews:2019aa}{}}%
-Andrews, Tallulah S, and Martin Hemberg. 2019. {``M3Drop: Dropout-Based
-Feature Selection for scRNASeq.''} \emph{Bioinformatics} 35 (16):
-2865--67. \url{https://doi.org/10.1093/bioinformatics/bty1044}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-benjamini:1995}{}}%
-Benjamini, Yoav, and Yosef Hochberg. 1995. {``Controlling the False
-Discovery Rate: A Practical and Powerful Approach to Multiple
-Testing.''} \emph{Journal of the Royal Statistical Society. Series B
-(Methodological)} 57 (1): 289--300.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bullard:2010}{}}%
-Bullard, James H, Elizabeth Purdom, Kasper D Hansen, and Sandrine
-Dudoit. 2010. {``Evaluation of Statistical Methods for Normalization and
-Differential Expression in m{RNA-Seq} Experiments.''} \emph{BMC
-Bioinformatics} 11: 94. \url{https://doi.org/10.1186/1471-2105-11-94}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Cui:2003aa}{}}%
-Cui, Xiangqin, and Gary A Churchill. 2003.
-{``\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical Tests
-for Differential Expression in cDNA Microarray Experiments}.''}
-\emph{Genome Biol} 4 (4): 210.1--10.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Denge00262-18}{}}%
-Deng, Zhi-Luo, Cornelia Gottschick, Sabin Bhuju, Clarissa Masur,
-Christoph Abels, and Irene Wagner-Döbler. 2018. {``Metatranscriptome
-Analysis of the Vaginal Microbiota Reveals Potential Mechanisms for
-Protection Against Metronidazole in Bacterial Vaginosis.''} Edited by
-Craig D. Ellermeier, Janet Hill, and Andrew Onderdonk. \emph{mSphere} 3
-(3). \url{https://doi.org/10.1128/mSphereDirect.00262-18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Dillies:2013}{}}%
-Dillies, Marie-Agnès, Andrea Rau, Julie Aubert, Christelle
-Hennequet-Antier, Marine Jeanmougin, Nicolas Servant, Céline Keime, et
-al. 2013. {``A Comprehensive Evaluation of Normalization Methods for
-{Illumina} High-Throughput {RNA} Sequencing Data Analysis.''}
-\emph{Brief Bioinform} 14 (6): 671--83.
-\url{https://doi.org/10.1093/bib/bbs046}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fernandes:2013}{}}%
-Fernandes, Andrew D, Jean M Macklaim, Thomas G Linn, Gregor Reid, and
-Gregory B Gloor. 2013. {``ANOVA-Like Differential Expression (ALDEx)
-Analysis for Mixed Population RNA-Seq.''} \emph{PLoS One} 8 (7): e67019.
-\url{https://doi.org/10.1371/journal.pone.0067019}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fernandes:2014}{}}%
-Fernandes, Andrew D, Jennifer Ns Reid, Jean M Macklaim, Thomas A
-McMurrough, David R Edgell, and Gregory B Gloor. 2014. {``Unifying the
-Analysis of High-Throughput Sequencing Datasets: Characterizing
-{RNA}-Seq, 16{S} r{RNA} Gene Sequencing and Selective Growth Experiments
-by Compositional Data Analysis.''} \emph{Microbiome} 2: 15.1--13.
-\url{https://doi.org/10.1186/2049-2618-2-15}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fettweis:2019aa}{}}%
-Fettweis, Jennifer M, Myrna G Serrano, J Paul Brooks, David J Edwards,
-Philippe H Girerd, Hardik I Parikh, Bernice Huang, et al. 2019. {``The
-Vaginal Microbiome and Preterm Birth.''} \emph{Nat Med} 25 (6):
-1012--21. \url{https://doi.org/10.1038/s41591-019-0450-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Gierlinski:2015aa}{}}%
-Gierliński, Marek, Christian Cole, Pietà Schofield, Nicholas J Schurch,
-Alexander Sherstnev, Vijender Singh, Nicola Wrobel, et al. 2015.
-{``Statistical Models for RNA-Seq Data Derived from a Two-Condition
-48-Replicate Experiment.''} \emph{Bioinformatics} 31 (22): 3625--30.
-\url{https://doi.org/10.1093/bioinformatics/btv425}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gloor:effect}{}}%
-Gloor, GB, JM Macklaim, and AD Fernandes. 2016. {``Displaying Variation
-in Large Datasets: Plotting a Visual Summary of Effect Sizes.''}
-\emph{Journal of Computational and Graphical Statistics} 25 (3C):
-971--79. \url{https://doi.org/10.1080/10618600.2015.1131161}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-evoRepr-pre}{}}%
-Gould, E, and et.al. 2023. {``Same Data, Different Analysts: Variation
-in Effect Sizes Due to Analytical Decisions in Ecology and Evolutionary
-Biology.''} \emph{EcoEvoRxiv}.
-https://doi.org/\url{https://doi.org/10.32942/X2GG62}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Halsey:2015aa}{}}%
-Halsey, Lewis G, Douglas Curran-Everett, Sarah L Vowler, and Gordon B
-Drummond. 2015. {``The Fickle p Value Generates Irreproducible
-Results.''} \emph{Nat Methods} 12 (3): 179--85.
-\url{https://doi.org/10.1038/nmeth.3288}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hawinkel2017}{}}%
-Hawinkel, Stijn, Federico Mattiello, Luc Bijnens, and Olivier Thas.
-2018. {``A Broken Promise : Microbiome Differential Abundance Methods Do
-Not Control the False Discovery Rate.''} \emph{BRIEFINGS IN
-BIOINFORMATICS}. \url{http://dx.doi.org/10.1093/bib/bbx104}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hughes:2005tu}{}}%
-Hughes, Jennifer B, and Jessica J Hellmann. 2005. {``The Application of
-Rarefaction Techniques to Molecular Inventories of Microbial
-Diversity.''} \emph{Methods Enzymol} 397: 292--308.
-\url{https://doi.org/10.1016/S0076-6879(05)97017-1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hummelen:2010}{}}%
-Hummelen, Ruben, Andrew D Fernandes, Jean M Macklaim, Russell J Dickson,
-John Changalucha, Gregory B Gloor, and Gregor Reid. 2010. {``Deep
-Sequencing of the Vaginal Microbiota of Women with {HIV}.''} \emph{PLoS
-One} 5 (8): e12078. \url{https://doi.org/10.1371/journal.pone.0012078}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ioannidis:2005aa}{}}%
-Ioannidis, John P A. 2005. {``Why Most Published Research Findings Are
-False.''} \emph{PLoS Med} 2 (8): e124.
-\url{https://doi.org/10.1371/journal.pmed.0020124}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Li:2022aa}{}}%
-Li, Yumei, Xinzhou Ge, Fanglue Peng, Wei Li, and Jingyi Jessica Li.
-2022. {``Exaggerated False Positives by Popular Differential Expression
-Methods When Analyzing Human Population Samples.''} \emph{Genome Biol}
-23 (1): 79. \url{https://doi.org/10.1186/s13059-022-02648-4}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Love:2014aa}{}}%
-Love, Michael I, Wolfgang Huber, and Simon Anders. 2014. {``Moderated
-Estimation of Fold Change and Dispersion for RNA-Seq Data with
-DESeq2.''} \emph{Genome Biol} 15 (12): 550.1--21.
-\url{https://doi.org/10.1186/s13059-014-0550-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lovell:2015}{}}%
-Lovell, David, Vera Pawlowsky-Glahn, Juan José Egozcue, Samuel
-Marguerat, and Jürg Bähler. 2015. {``Proportionality: A Valid
-Alternative to Correlation for Relative Data.''} \emph{PLoS Comput Biol}
-11 (3): e1004075.
-https://doi.org/\url{https://doi.org/10.1371/journal.pcbi.1004075}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Loven:2012aa}{}}%
-Lovén, Jakob, David A Orlando, Alla A Sigova, Charles Y Lin, Peter B
-Rahl, Christopher B Burge, David L Levens, Tong Ihn Lee, and Richard A
-Young. 2012. {``Revisiting Global Gene Expression Analysis.''}
-\emph{Cell} 151 (3): 476--82.
-\url{https://doi.org/10.1016/j.cell.2012.10.012}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-macklaim:2013}{}}%
-Macklaim, Jean M, Andrew D Fernandes, Julia M Di Bella, Jo-Anne Hammond,
-Gregor Reid, and Gregory B Gloor. 2013. {``Comparative Meta-{RNA}-Seq of
-the Vaginal Microbiota and Differential Expression by Lactobacillus
-Iners in Health and Dysbiosis.''} \emph{Microbiome} 1 (1): 12.
-\url{https://doi.org/10.1186/2049-2618-1-12}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Macklaim:2018aa}{}}%
-Macklaim, Jean M, and Gregory B Gloor. 2018. {``From {RNA}-Seq to
-Biological Inference: Using Compositional Data Analysis in
-Meta-Transcriptomics.''} \emph{Methods Mol Biol} 1849: 193--213.
-\url{https://doi.org/10.1007/978-1-4939-8728-3/_13}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-maza2013}{}}%
-Maza, Elie, Pierre Frasse, Pavel Senin, Mondher Bouzayen, and Mohamed
-Zouine. 2013. {``Comparison of Normalization Methods for Differential
-Gene Expression Analysis in RNA-Seq Experiments: A Matter of Relative
-Size of Studied Transcriptomes.''} \emph{Commun Integr Biol} 6 (6):
-e25849. \url{https://doi.org/10.4161/cib.25849}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Mortazavi:2008}{}}%
-Mortazavi, Ali, Brian A Williams, Kenneth McCue, Lorian Schaeffer, and
-Barbara Wold. 2008. {``Mapping and Quantifying Mammalian Transcriptomes
-by {RNA-seq}.''} \emph{Nat Methods} 5 (7): 621--28.
-\url{https://doi.org/10.1038/nmeth.1226}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Nie:2012aa}{}}%
-Nie, Zuqin, Gangqing Hu, Gang Wei, Kairong Cui, Arito Yamane, Wolfgang
-Resch, Ruoning Wang, et al. 2012. {``C-{M}yc Is a Universal Amplifier of
-Expressed Genes in Lymphocytes and Embryonic Stem Cells.''} \emph{Cell}
-151 (1): 68--79. \url{https://doi.org/10.1016/j.cell.2012.08.033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nixon2023scale}{}}%
-Nixon, Michelle Pistner, Jeffrey Letourneau, Lawrence A. David, Nicole
-A. Lazar, Sayan Mukherjee, and Justin D. Silverman. 2023. {``Scale
-Reliant Inference.''} \url{https://arxiv.org/abs/2201.03616}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Open-Science-Collaboration:2015aa}{}}%
-Open Science Collaboration. 2015. {``Estimating the Reproducibility of
-Psychological Science.''} \emph{Science} 349 (6251): aac4716.
-\url{https://doi.org/10.1126/science.aac4716}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Props:2017aa}{}}%
-Props, Ruben, Frederiek-Maarten Kerckhof, Peter Rubbens, Jo De Vrieze,
-Emma Hernandez Sanabria, Willem Waegeman, Pieter Monsieurs, Frederik
-Hammes, and Nico Boon. 2017. {``Absolute Quantification of Microbial
-Taxon Abundances.''} \emph{ISME J} 11 (2): 584--87.
-\url{https://doi.org/10.1038/ismej.2016.117}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Quinn:2018aa}{}}%
-Quinn, Thomas P, Tamsyn M Crowley, and Mark F Richardson. 2018.
-{``Benchmarking Differential Expression Analysis Tools for RNA-Seq:
-Normalization-Based Vs. Log-Ratio Transformation-Based Methods.''}
-\emph{BMC Bioinformatics} 19 (1): 274.
-\url{https://doi.org/10.1186/s12859-018-2261-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ravel:2010}{}}%
-Ravel, Jacques, Pawel Gajer, Zaid Abdo, G Maria Schneider, Sara S K
-Koenig, Stacey L McCulle, Shara Karlebach, et al. 2011. {``Vaginal
-Microbiome of Reproductive-Age Women.''} \emph{Proc Natl Acad Sci U S
-A}, no. 108: 4680--87. \url{https://doi.org/doi/10.1073/pnas.100611107}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Robinson:2010}{}}%
-Robinson, Mark D, Davis J McCarthy, and Gordon K Smyth. 2010. {``edgeR:
-A Bioconductor Package for Differential Expression Analysis of Digital
-Gene Expression Data.''} \emph{Bioinformatics} 26 (1): 139--40.
-\url{https://doi.org/10.1093/bioinformatics/btp616}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Robinson:2010a}{}}%
-Robinson, Mark D, and Alicia Oshlack. 2010. {``A Scaling Normalization
-Method for Differential Expression Analysis of {RNA-seq} Data.''}
-\emph{Genome Biol} 11 (3): R25.1--9.
-\url{https://doi.org/10.1186/gb-2010-11-3-r25}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Rodgers:2021aa}{}}%
-Rodgers, Peter, and Andy Collings. 2021. {``What Have We Learned?''}
-\emph{Elife} 10 (December). \url{https://doi.org/10.7554/eLife.75830}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Schurch:2016aa}{}}%
-Schurch, Nicholas J, Pietá Schofield, Marek Gierliński, Christian Cole,
-Alexander Sherstnev, Vijender Singh, Nicola Wrobel, et al. 2016. {``How
-Many Biological Replicates Are Needed in an RNA-Seq Experiment and Which
-Differential Expression Tool Should You Use?''} \emph{RNA} 22 (6):
-839--51. \url{https://doi.org/10.1261/rna.053959.115}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SEQCux2fMAQC-III-Consortium:2014aa}{}}%
-SEQC/MAQC-III Consortium. 2014. {``A Comprehensive Assessment of RNA-Seq
-Accuracy, Reproducibility and Information Content by the Sequencing
-Quality Control Consortium.''} \emph{Nat Biotechnol} 32 (9): 903--14.
-\url{https://doi.org/10.1038/nbt.2957}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Soneson:2013}{}}%
-Soneson, Charlotte, and Mauro Delorenzi. 2013. {``A Comparison of
-Methods for Differential Expression Analysis of {RNA-seq} Data.''}
-\emph{BMC Bioinformatics} 14: 91.
-\url{https://doi.org/10.1186/1471-2105-14-91}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Taniguchi:2010aa}{}}%
-Taniguchi, Yuichi, Paul J Choi, Gene-Wei Li, Huiyi Chen, Mohan Babu,
-Jeremy Hearn, Andrew Emili, and X Sunney Xie. 2010. {``Quantifying e.
-Coli Proteome and Transcriptome with Single-Molecule Sensitivity in
-Single Cells.''} \emph{Science} 329 (5991): 533--38.
-\url{https://doi.org/10.1126/science.1188308}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Thellin:1999aa}{}}%
-Thellin, O, W Zorzi, B Lakaye, B De Borman, B Coumans, G Hennen, T
-Grisar, A Igout, and E Heinen. 1999.
-{``\href{https://www.ncbi.nlm.nih.gov/pubmed/10617337}{Housekeeping
-Genes as Internal Standards: Use and Limits}.''} \emph{J Biotechnol} 75
-(2-3): 291--95.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Thorsen:2016aa}{}}%
-Thorsen, Jonathan, Asker Brejnrod, Martin Mortensen, Morten A Rasmussen,
-Jakob Stokholm, Waleed Abu Al-Soud, Søren Sørensen, Hans Bisgaard, and
-Johannes Waage. 2016. {``Large-Scale Benchmarking Reveals False
-Discoveries and Count Transformation Sensitivity in 16{S} r{RNA} Gene
-Amplicon Data Analysis Methods Used in Microbiome Studies.''}
-\emph{Microbiome} 4 (1): 62.
-\url{https://doi.org/10.1186/s40168-016-0208-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Vandeputte:2017aa}{}}%
-Vandeputte, Doris, Gunter Kathagen, Kevin D'hoe, Sara Vieira-Silva,
-Mireia Valles-Colomer, João Sabino, Jun Wang, et al. 2017.
-{``Quantitative Microbiome Profiling Links Gut Community Variation to
-Microbial Load.''} \emph{Nature} 551 (7681): 507--11.
-\url{https://doi.org/10.1038/nature24460}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wagner:tpm}{}}%
-Wagner, Günter P, Koryu Kin, and Vincent J Lynch. 2012. {``Measurement
-of mRNA Abundance Using RNA-Seq Data: RPKM Measure Is Inconsistent Among
-Samples.''} \emph{Theory Biosci} 131 (4): 281--85.
-\url{https://doi.org/10.1007/s12064-012-0162-3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Weiss:2017aa}{}}%
-Weiss, Sophie, Zhenjiang Zech Xu, Shyamal Peddada, Amnon Amir, Kyle
-Bittinger, Antonio Gonzalez, Catherine Lozupone, et al. 2017.
-{``Normalization and Microbial Differential Abundance Strategies Depend
-Upon Data Characteristics.''} \emph{Microbiome} 5 (1): 27.
-\url{https://doi.org/10.1186/s40168-017-0237-y}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Wu2021}{}}%
-Wu, Jia R., Jean M. Macklaim, Briana L. Genge, and Gregory B. Gloor.
-2021. {``Finding the Centre: Compositional Asymmetry in High-Throughput
-Sequencing Datasets.''} In \emph{Advances in Compositional Data
-Analysis: Festschrift in Honour of Vera Pawlowsky-Glahn}, edited by
-Peter Filzmoser, Karel Hron, Josep Antoni Martìn-Fernàndez, and Javier
-Palarea-Albaladejo, 329--46. Cham: Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-71175-7_17}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Yoshikawa:2011aa}{}}%
-Yoshikawa, Katsunori, Tadamasa Tanaka, Yoshihiro Ida, Chikara Furusawa,
-Takashi Hirasawa, and Hiroshi Shimizu. 2011. {``Comprehensive Phenotypic
-Analysis of Single-Gene Deletion and Overexpression Strains of
-Saccharomyces Cerevisiae.''} \emph{Yeast} 28 (5): 349--61.
-\url{https://doi.org/10.1002/yea.1843}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Zozaya:2010}{}}%
-Zozaya-Hinchliffe, Marcela, Rebecca Lillis, David H Martin, and Michael
-J Ferris. 2010. {``Quantitative PCR Assessments of Bacterial Species in
-Women with and Without Bacterial Vaginosis.''} \emph{J Clin Microbiol}
-48 (5): 1812--19. \url{https://doi.org/10.1128/JCM.00851-09}.
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{1}
+\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
+1. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
+Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
+Darzi,Y., \emph{et al.} (2017)
+\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
+profiling links gut community variation to microbial load}.
+\emph{Nature}, \textbf{551}, 507--511.
+
+\bibitem[\citeproctext]{ref-Props:2017aa}
+2. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
+Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
+\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
+microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.
+
+\bibitem[\citeproctext]{ref-Thorsen:2016aa}
+3. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
+Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
+\href{https://doi.org/10.1186/s40168-016-0208-8}{Large-scale
+benchmarking reveals false discoveries and count transformation
+sensitivity in 16{S} r{RNA} gene amplicon data analysis methods used in
+microbiome studies}. \emph{Microbiome}, \textbf{4}, 62.
+
+\bibitem[\citeproctext]{ref-Quinn:2018aa}
+4. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
+\href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
+differential expression analysis tools for RNA-seq: Normalization-based
+vs. Log-ratio transformation-based methods}. \emph{BMC Bioinformatics},
+\textbf{19}, 274.
+
+\bibitem[\citeproctext]{ref-Zhang:2009aa}
+5. Zhang,S. and Cao,J. (2009)
+\href{https://doi.org/10.1186/1471-2105-10-402}{A close examination of
+double filtering with fold change and t test in microarray analysis}.
+\emph{BMC Bioinformatics}, \textbf{10}, 402.
+
+\bibitem[\citeproctext]{ref-Ebrahimpoor:2021aa}
+6. Ebrahimpoor,M. and Goeman,J.J. (2021)
+\href{https://doi.org/10.1093/bib/bbab053}{Inflated false discovery rate
+due to volcano plots: Problem and solutions}. \emph{Brief Bioinform},
+\textbf{22}.
+
+\bibitem[\citeproctext]{ref-Wu2021}
+7. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
+\href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
+Compositional asymmetry in high-throughput sequencing datasets}. In
+Filzmoser,P., Hron,K., Martìn-Fernàndez,J.A., Palarea-Albaladejo,J.
+(eds), \emph{Advances in compositional data analysis: Festschrift in
+honour of vera pawlowsky-glahn}. Springer International Publishing,
+Cham, pp. 329--346.
+
+\bibitem[\citeproctext]{ref-nixon2023scale}
+8. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
+Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
+reliant inference}.
+
+\bibitem[\citeproctext]{ref-Nixon2024B}
+9. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
+normalization: Incorporating scale uncertainty in microbiome and gene
+expression analysis. \emph{bioRxiv},
+\href{https://doi.org/10.1101/2024.04.01.587602}{10.1101/2024.04.01.587602}.
+
+\bibitem[\citeproctext]{ref-lovell:2011}
+10. Lovell,D., Müller,W., Taylor,J., Zwart,A. and Helliwell,C. (2011)
+Proportions, percentages, ppm: Do the molecular biosciences treat
+compositional data right? In Pawlowsky-Glahn,V., Buccianti,A. (eds),
+\emph{Compositional Data Analysis: Theory and Applications}. John Wiley;
+Sons New York, NY, London, pp. 193--207.
+
+\bibitem[\citeproctext]{ref-Quinn:2019aa}
+11. Quinn,T.P., Erb,I., Gloor,G., Notredame,C., Richardson,M.F. and
+Crowley,T.M. (2019) \href{https://doi.org/10.1093/gigascience/giz107}{A
+field guide for the compositional analysis of any-omics data}.
+\emph{Gigascience}, \textbf{8}.
+
+\bibitem[\citeproctext]{ref-Lovell:2015}
+12. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
+Bähler,J. (2015)
+\href{https://doi.org/10.1371/journal.pcbi.1004075}{Proportionality: A
+valid alternative to correlation for relative data}. \emph{PLoS Comput
+Biol}, \textbf{11}, e1004075.
+
+\bibitem[\citeproctext]{ref-Nie:2012aa}
+13. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
+Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
+\href{https://doi.org/10.1016/j.cell.2012.08.033}{C-{M}yc is a universal
+amplifier of expressed genes in lymphocytes and embryonic stem cells}.
+\emph{Cell}, \textbf{151}, 68--79.
+
+\bibitem[\citeproctext]{ref-Loven:2012aa}
+14. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
+Burge,C.B., Levens,D.L., Lee,T.I. and Young,R.A. (2012)
+\href{https://doi.org/10.1016/j.cell.2012.10.012}{Revisiting global gene
+expression analysis}. \emph{Cell}, \textbf{151}, 476--82.
+
+\bibitem[\citeproctext]{ref-Scott:2010}
+15. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
+\href{https://doi.org/10.1126/science.1192588}{Interdependence of cell
+growth and gene expression: Origins and consequences}. \emph{Science},
+\textbf{330}, 1099--102.
+
+\bibitem[\citeproctext]{ref-Yoshikawa:2011aa}
+16. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
+Shimizu,H. (2011) \href{https://doi.org/10.1002/yea.1843}{Comprehensive
+phenotypic analysis of single-gene deletion and overexpression strains
+of saccharomyces cerevisiae}. \emph{Yeast}, \textbf{28}, 349--61.
+
+\bibitem[\citeproctext]{ref-Lin:2018aa}
+17. Lin,J. and Amir,A. (2018)
+\href{https://doi.org/10.1038/s41467-018-06714-z}{Homeostasis of protein
+and mRNA concentrations in growing cells}. \emph{Nat Commun},
+\textbf{9}, 4496.
+
+\bibitem[\citeproctext]{ref-Zozaya:2010}
+18. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
+\href{https://doi.org/10.1128/JCM.00851-09}{Quantitative PCR assessments
+of bacterial species in women with and without bacterial vaginosis}.
+\emph{J Clin Microbiol}, \textbf{48}, 1812--9.
+
+\bibitem[\citeproctext]{ref-Ravel:2010}
+19. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
+McCulle,S.L., Karlebach,S., Gorle,R., Russell,J., Tacket,C.O., \emph{et
+al.} (2011) Vaginal microbiome of reproductive-age women. \emph{Proc
+Natl Acad Sci U S A},
+\href{https://doi.org/doi/10.1073/pnas.100611107}{doi/10.1073/pnas.100611107}.
+
+\bibitem[\citeproctext]{ref-Hummelen:2010}
+20. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
+Changalucha,J., Gloor,G.B. and Reid,G. (2010)
+\href{https://doi.org/10.1371/journal.pone.0012078}{Deep sequencing of
+the vaginal microbiota of women with {HIV}}. \emph{PLoS One},
+\textbf{5}, e12078.
+
+\bibitem[\citeproctext]{ref-Bullard:2010}
+21. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
+\href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
+methods for normalization and differential expression in m{RNA-seq}
+experiments}. \emph{BMC Bioinformatics}, \textbf{11}, 94.
+
+\bibitem[\citeproctext]{ref-Dillies:2013}
+22. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
+Jeanmougin,M., Servant,N., Keime,C., Marot,G., Castel,D., Estelle,J.,
+\emph{et al.} (2013) \href{https://doi.org/10.1093/bib/bbs046}{A
+comprehensive evaluation of normalization methods for {Illumina}
+high-throughput {RNA} sequencing data analysis}. \emph{Brief Bioinform},
+\textbf{14}, 671--83.
+
+\bibitem[\citeproctext]{ref-Weiss:2017aa}
+23. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
+Lozupone,C., Zaneveld,J.R., Vázquez-Baeza,Y., Birmingham,A., \emph{et
+al.} (2017)
+\href{https://doi.org/10.1186/s40168-017-0237-y}{Normalization and
+microbial differential abundance strategies depend upon data
+characteristics}. \emph{Microbiome}, \textbf{5}, 27.
+
+\bibitem[\citeproctext]{ref-Nearing:2022aa}
+24. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
+Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
+\emph{et al.} (2022)
+\href{https://doi.org/10.1038/s41467-022-28034-z}{Microbiome
+differential abundance methods produce different results across 38
+datasets}. \emph{Nat Commun}, \textbf{13}, 342.
+
+\bibitem[\citeproctext]{ref-Hughes:2005tu}
+25. Hughes,J.B. and Hellmann,J.J. (2005)
+\href{https://doi.org/10.1016/S0076-6879(05)97017-1}{The application of
+rarefaction techniques to molecular inventories of microbial diversity}.
+\emph{Methods Enzymol}, \textbf{397}, 292--308.
+
+\bibitem[\citeproctext]{ref-Mortazavi:2008}
+26. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
+(2008) \href{https://doi.org/10.1038/nmeth.1226}{Mapping and quantifying
+mammalian transcriptomes by {RNA-seq}}. \emph{Nat Methods}, \textbf{5},
+621--8.
+
+\bibitem[\citeproctext]{ref-wagner:tpm}
+27. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
+\href{https://doi.org/10.1007/s12064-012-0162-3}{Measurement of mRNA
+abundance using RNA-seq data: RPKM measure is inconsistent among
+samples}. \emph{Theory Biosci}, \textbf{131}, 281--5.
+
+\bibitem[\citeproctext]{ref-Robinson:2010a}
+28. Robinson,M.D. and Oshlack,A. (2010)
+\href{https://doi.org/10.1186/gb-2010-11-3-r25}{A scaling normalization
+method for differential expression analysis of {RNA-seq} data}.
+\emph{Genome Biol}, \textbf{11}, R25.1--R25.9.
+
+\bibitem[\citeproctext]{ref-aitchison1982}
+29. Aitchison,J. (1982) The statistical analysis of compositional data.
+\emph{Journal of the Royal Statistical Society: Series B
+(Methodological)}, \textbf{44}, 139--160.
+
+\bibitem[\citeproctext]{ref-Anders:2010}
+30. Anders,S. and Huber,W. (2010)
+\href{https://doi.org/10.1186/gb-2010-11-10-r106}{Differential
+expression analysis for sequence count data}. \emph{Genome Biol},
+\textbf{11}, R106.
+
+\bibitem[\citeproctext]{ref-fernandes:2013}
+31. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
+(2013) \href{https://doi.org/10.1371/journal.pone.0067019}{ANOVA-like
+differential expression (ALDEx) analysis for mixed population RNA-seq}.
+\emph{PLoS One}, \textbf{8}, e67019.
+
+\bibitem[\citeproctext]{ref-Yerke:2024aa}
+32. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
+\href{https://doi.org/10.1186/s40168-023-01747-z}{Proportion-based
+normalizations outperform compositional data transformations in machine
+learning applications}. \emph{Microbiome}, \textbf{12}, 45.
+
+\bibitem[\citeproctext]{ref-fernandes:2014}
+33. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
+Edgell,D.R. and Gloor,G.B. (2014)
+\href{https://doi.org/10.1186/2049-2618-2-15}{Unifying the analysis of
+high-throughput sequencing datasets: Characterizing {RNA}-seq, 16{S}
+r{RNA} gene sequencing and selective growth experiments by compositional
+data analysis}. \emph{Microbiome}, \textbf{2}, 15.1--15.13.
+
+\bibitem[\citeproctext]{ref-hawinkel2017}
+34. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
+\href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
+Microbiome differential abundance methods do not control the false
+discovery rate}. \emph{BRIEFINGS IN BIOINFORMATICS}.
+
+\bibitem[\citeproctext]{ref-Li:2022aa}
+35. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+\href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
+positives by popular differential expression methods when analyzing
+human population samples}. \emph{Genome Biol}, \textbf{23}, 79.
+
+\bibitem[\citeproctext]{ref-Cui:2003aa}
+36. Cui,X. and Churchill,G.A. (2003)
+\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
+for differential expression in cDNA microarray experiments}.
+\emph{Genome Biol}, \textbf{4}, 210.1--210.10.
+
+\bibitem[\citeproctext]{ref-Schurch:2016aa}
+37. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
+Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
+al.} (2016) \href{https://doi.org/10.1261/rna.053959.115}{How many
+biological replicates are needed in an RNA-seq experiment and which
+differential expression tool should you use?} \emph{RNA}, \textbf{22},
+839--51.
+
+\bibitem[\citeproctext]{ref-Gierlinski:2015aa}
+38. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
+Singh,V., Wrobel,N., Gharbi,K., Simpson,G., Owen-Hughes,T., \emph{et
+al.} (2015)
+\href{https://doi.org/10.1093/bioinformatics/btv425}{Statistical models
+for RNA-seq data derived from a two-condition 48-replicate experiment}.
+\emph{Bioinformatics}, \textbf{31}, 3625--3630.
+
+\bibitem[\citeproctext]{ref-Halsey:2015aa}
+39. Halsey,L.G., Curran-Everett,D., Vowler,S.L. and Drummond,G.B. (2015)
+\href{https://doi.org/10.1038/nmeth.3288}{The fickle p value generates
+irreproducible results}. \emph{Nat Methods}, \textbf{12}, 179--85.
+
+\bibitem[\citeproctext]{ref-benjamini:1995}
+40. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
+rate: A practical and powerful approach to multiple testing.
+\emph{Journal of the Royal Statistical Society. Series B
+(Methodological)}, \textbf{57}, 289--300.
+
+\bibitem[\citeproctext]{ref-Soneson:2013}
+41. Soneson,C. and Delorenzi,M. (2013)
+\href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
+for differential expression analysis of {RNA-seq} data}. \emph{BMC
+Bioinformatics}, \textbf{14}, 91.
+
+\bibitem[\citeproctext]{ref-maza2013}
+42. Maza,E., Frasse,P., Senin,P., Bouzayen,M. and Zouine,M. (2013)
+\href{https://doi.org/10.4161/cib.25849}{Comparison of normalization
+methods for differential gene expression analysis in RNA-seq
+experiments: A matter of relative size of studied transcriptomes}.
+\emph{Commun Integr Biol}, \textbf{6}, e25849.
+
+\bibitem[\citeproctext]{ref-gloor:effect}
+43. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
+\href{https://doi.org/10.1080/10618600.2015.1131161}{Displaying
+variation in large datasets: Plotting a visual summary of effect sizes}.
+\emph{Journal of Computational and Graphical Statistics}, \textbf{25},
+971--979.
+
+\bibitem[\citeproctext]{ref-Rocha:2020aa}
+44. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
+(2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
+expression analysis in bacteria by RT-qPCR}. \emph{Methods Mol Biol},
+\textbf{2065}, 119--137.
+
+\bibitem[\citeproctext]{ref-Taniguchi:2010aa}
+45. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
+Emili,A. and Xie,X.S. (2010)
+\href{https://doi.org/10.1126/science.1188308}{Quantifying e. Coli
+proteome and transcriptome with single-molecule sensitivity in single
+cells}. \emph{Science}, \textbf{329}, 533--8.
+
+\bibitem[\citeproctext]{ref-dosSantos:2024}
+46. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
+Gloor,G.B. (2024) Vaginal metatranscriptome meta-analysis reveals
+functional BV subgroups and novel colonisation strategies.
+\emph{bioRxiv},
+\href{https://doi.org/10.1101/2024.04.24.590967}{10.1101/2024.04.24.590967}.
+
+\bibitem[\citeproctext]{ref-macklaim:2013}
+47. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
+and Gloor,G.B. (2013)
+\href{https://doi.org/10.1186/2049-2618-1-12}{Comparative meta-{RNA}-seq
+of the vaginal microbiota and differential expression by lactobacillus
+iners in health and dysbiosis}. \emph{Microbiome}, \textbf{1}, 12.
+
+\bibitem[\citeproctext]{ref-Denge00262-18}
+48. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
+Wagner-Döbler,I. (2018)
+\href{https://doi.org/10.1128/mSphereDirect.00262-18}{Metatranscriptome
+analysis of the vaginal microbiota reveals potential mechanisms for
+protection against metronidazole in bacterial vaginosis}.
+\emph{mSphere}, \textbf{3}.
+
+\bibitem[\citeproctext]{ref-Fettweis:2019aa}
+49. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
+Parikh,H.I., Huang,B., Arodz,T.J., Edupuganti,L., Glascock,A.L.,
+\emph{et al.} (2019)
+\href{https://doi.org/10.1038/s41591-019-0450-2}{The vaginal microbiome
+and preterm birth}. \emph{Nat Med}, \textbf{25}, 1012--1021.
+
+\bibitem[\citeproctext]{ref-Okuda:2008}
+50. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
+Goto,S. and Kanehisa,M. (2008)
+\href{https://doi.org/10.1093/nar/gkn282}{KEGG atlas mapping for global
+analysis of metabolic pathways}. \emph{Nucleic Acids Res}, \textbf{36},
+W423--6.
+
+\bibitem[\citeproctext]{ref-yeast-absolute}
+51. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
+Bähler,J. (2012)
+\href{https://doi.org/10.1016/j.cell.2012.09.019}{Quantitative analysis
+of fission yeast transcriptomes and proteomes in proliferating and
+quiescent cells}. \emph{Cell}, \textbf{151}, 671--83.
+
+\bibitem[\citeproctext]{ref-McMurdie:2014a}
+52. McMurdie,P.J. and Holmes,S. (2014)
+\href{https://doi.org/10.1371/journal.pcbi.1003531}{Waste not, want not:
+Why rarefying microbiome data is inadmissible}. \emph{PLoS Comput Biol},
+\textbf{10}, e1003531.
+
+\bibitem[\citeproctext]{ref-Brooks:2024aa}
+53. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
+\href{https://doi.org/10.1038/s41576-023-00679-6}{Challenges and best
+practices in omics benchmarking}. \emph{Nat Rev Genet}, \textbf{25},
+326--339.
+
+\bibitem[\citeproctext]{ref-gustafson2015bayesian}
+54. Gustafson,P. (2015) Bayesian inference for partially identified
+models: Exploring the limits of limited data CRC Press.
+
+\bibitem[\citeproctext]{ref-Thellin:1999aa}
+55. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
+Hennen,G., Grisar,T., Igout,A. and Heinen,E. (1999)
+\href{https://www.ncbi.nlm.nih.gov/pubmed/10617337}{Housekeeping genes
+as internal standards: Use and limits}. \emph{J Biotechnol},
+\textbf{75}, 291--5.
+
+\bibitem[\citeproctext]{ref-SEQCux2fMAQC-III-Consortium:2014aa}
+56. SEQC/MAQC-III Consortium (2014)
+\href{https://doi.org/10.1038/nbt.2957}{A comprehensive assessment of
+RNA-seq accuracy, reproducibility and information content by the
+sequencing quality control consortium}. \emph{Nat Biotechnol},
+\textbf{32}, 903--14.
+
+\bibitem[\citeproctext]{ref-Song:2023aa}
+57. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
+Klatt,N.R., Hensley-McBain,T. and Wu,M.C. (2023)
+\href{https://doi.org/10.1186/s12859-023-05147-w}{Accommodating multiple
+potential normalizations in microbiome associations studies}. \emph{BMC
+Bioinformatics}, \textbf{24}, 22.
 

### 2024-08-13 code/fix_authors.sh

@@ -18,7 +18,6 @@ rm go3.aux
 rm go3.out
 rm go3.toc
 rm go3.log
-rm go3.tex
 rm go3-author.aux
 rm go3-author.log
 rm go3-author.blg

### 2024-08-22 go3-author.tex
with Justin's suggestions
Send back to Justin for comment

@@
 \hypersetup{
-  pdftitle={Normalizations are not what you think; Explicit Scale Simulation in ALDEx2},
+  pdftitle={Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2},
   hidelinks,
@@
 
-\title{Normalizations are not what you think; Explicit Scale Simulation
-in ALDEx2}
+\title{Explicit Scale Simulation for analysis of RNA-sequencing with
+ALDEx2}
 \author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4}$}
@@
 in sequencing depth is driven by technical factors, and not by variation
-in the scale (e.g., total size, microbial load, or total gene
+in the scale (e.g., total size, microbial load, or total mRNA
 expression) of the underlying biological systems. Typically a
@@
 sensitive to scale; e.g., differential abundance and differential
-expression analyses. Recently we showed that all normalizations make
+expression analyses. We recently showed that all normalizations make
 implicit assumptions about the unmeasured system scale and that errors
-in these assumptions can lead to dramatic increases in false positive
-and false negative rates. Here we describe updates to the ALDEx2 R
-package that mitigate these problems by directly modeling uncertainty in
-the unmeasured system scale through the use of a \textit{scale model}.
-Scale models generalize the idea of normalizations and can be thought of
-as explicitly modeling the error in normalization by examining a
-distribution over all possible normalizations. Beyond enhancing the
-robustness of HTS analyses, the use of scale models within ALDEx2
-enhances the transparency and reproducibility of analyses by making
-implicit normalizing assumptions an explicit part of the model building
-process.
+in these assumptions can dramatically increase false positive and false
+negative rates. We demonstrated that these errors can be mitigated by
+accounting for uncertainty about scale using a \emph{scale model}, which
+we integrated into the ALDEx2 R package. This article provides new
+insights into those methods, focusing on the application to
+transcriptomic analysis. Here we provide transcriptomic case studies
+demonstrating how scale models, rather than traditional normalizations,
+can reduce false positive and false negative rates in practice while
+enhancing the transparency and reproducibility of analyses. We show that
+these scale models replace the need for dual cutoff approaches often
+used to address the disconnect between practical and statistical
+significance. We demonstrate the utility of that scale models built
+based on known housekeeping genes in complex metatranscriptomic
+datasets. Thus this work provides example and practical guidance on how
+to incorporate scale into transcriptomic analysis.
 \end{abstract}
@@
 \section{Introduction}\label{introduction}
+
+\doublespacing
 
@@
 (merging) multiple libraries together, and then applying a sample of the
-multiplexed library to the flow cell. Each of these steps is a
+multiplexed library to a flow cell. Each of these steps is a
 compositional sampling step as only a fixed-size subsample of nucleic
@@
 (e.g., size, microbial load, or total gene expression) of the measured
-biological system is degraded or lost. Increasingly, researchers are
-turning to modified experimental protocols (e.g., DNA spike-ins or cell
-counting) in an attempt to recover the lost biological variation in
-scale; see for example (1, 2). However, these protocols often do not
-recover meaningful biological variation but instead recover scale
-variation at the intermediate step in the sample preparation protocol
-where the DNA spike-in was added. In short, the disconnect between
-sample-to-sample variation in sequencing depth and biological variation
-in scale remains an outstanding challenge.
+biological system is degraded or lost. In the end, the information
+contained in the data relates only to relative abundances and has an
+arbitrary scale imposed by the sequencing process(1--3). Increasingly,
+researchers are turning to modified experimental protocols in an attempt
+to recover the lost biological variation in scale; see for example (4,
+5). However, these protocols only recover scale variation at the
+intermediate step in the sample preparation protocol where the
+intervention was made. In short, the disconnect between sample-to-sample
+variation in sequencing depth and biological variation in scale remains
+an open challenge.
 
@@
 traced, in whole or in part, to misspecification of scale. The first
-issue is poor control of the false discovery rate (FDR)
-{[}(3);(4);Nearing:\href{mailto:2022aa;@Li}{\nolinkurl{2022aa;@Li}}:2022aa{]},
-exhibited as dataset-dependent FDR control. The FDR problem is connected
-to the double-filtering approach that is often used, but which is known
-not to have appropriate FDR control (5, 6). The second issue is poor
-performance when analyzing asymmetric data; data that arises because of
-a bias in the direction of change. This type of data frequently arises
-in in-vitro selection experiments (SELEX), transcriptome analysis, and
-microbiome analysis (7). The third issue is that these problems become
-more pronounced as more samples are collected; that is, more information
-results in a worsening of the accuracy of the analysis (9). The final
-problem is that these data are relative data, i.e.~compositional (10,
-11), and substantial effort has been put into removing this constraint.
-
-The first three problems were recently shown by Nixon and Silverman (8)
-to be a result of a mismatch between the underlying size or scale of the
+issue is poor control of the false discovery rate (FDR) (6--9),
+exhibited as dataset-dependent FDR control and by the disconnect between
+statistical and biological significance(10). In current practice, these
+issues are addressed by a dual-filtering method, whereby both a low
+p-value (or equivalently a low q-value following FDR correction (11))
+and a large difference between groups is used to identify interesting
+transcripts or genes for follow-up analysis(10). This double-filtering
+approach is graphically exemplified by the volcano plot (12), but is
+known to not appropriately control the FDR\\
+(13, 14). The second issue is poor performance when analyzing data where
+the mean change between groups is non-zero. Such asymmetric data can
+arise when a gene set is expressed in one group but not the other, or
+when one group contains different gene content from the other. This type
+of data frequently arises in in-vitro selection experiments (SELEX),
+transcriptome analysis, and microbiome analysis (15). The third issue is
+that these problems become more pronounced as more samples are
+collected; that is, more information results in a worsening of the
+accuracy of the analysis (16).
+
+The three problems were recently shown by Nixon and Silverman (3) to be
+a result of a mismatch between the underlying size or scale of the
 system and the assumptions of the normalizations used for the analysis
 of HTS. Biological variation in scale often represents an important
-unmeasured confounder in HTS analyses (12). For example, cells
+unmeasured confounder in HTS analyses (17). For example, cells
 transformed by the cMyc oncogene have about 3 times the amount of mRNA
-and about twice the rRNA content than non-transformed cells (13), and
-this dramatically skews transcriptome analysis (14). In addition,
+and about twice the rRNA content than non-transformed cells (18), and
+this dramatically skews transcriptome analysis (19). In addition,
 wild-type and mutant strains of cell lines, yeast or bacteria have
@@
 which affect our ability to identify truly differentially abundant genes
-(15--17). As another example, the total bacterial load of the vaginal
+(20--22). As another example, the total bacterial load of the vaginal
 microbiome differs by 1-2 orders of magnitude in absolute abundance
-between the healthy and bacterial vaginosis states (18), and the
-composition between these states is dramatically different (19, 20).
+between the healthy and bacterial vaginosis states (23), and the
+composition between these states is dramatically different (24, 25).
 Thus, a full description of any of these systems includes both relative
@@
 access only the compositional information yet make implicit assumptions
-about the scale.
-
-Recently, Nixon et al. (8) showed that the challenge of non-biological
+about the scale (16).
+
+Recently, Nixon et al. (3) showed that the challenge of non-biological
 variation in sequencing depth be viewed as a problem of
 partially-identified models. They showed that \emph{all} normalizations
-make some assumption about scale. These implicit assumptions are often
-difficult to interpret, and different normalizations provide different
-outputs when applied to the same dataset (3, 21--24). Intuitively,
-normalizations in widespread use assume that either all samples have the
-same scale, e.g.~proportions, rarefaction (25), RPKM (26, 27), etc; or
-that a subset of features in one sample can be chosen as a reference to
-which the others are scaled e.g.~the TMM (28), or LVHA (7) or the
-additive log-ratio (29); or that different sub-parts of each sample
-maintain a constant scale across samples e.g.~the RLE (30); or that the
-geometric mean of the parts is appropriate e.g the CLR and its
-derivatives.
-
-The original ALDEx2 (31) model unwittingly made a strict assumption
-about scale through the CLR normalization (8). While this assumption
-could be useful in many cases it could never be exactly true, and others
-have shown that it is not always the best option (32). Nixon et al. (8)
-showed that better scale assumptions resulted in more reproducible data
-analysis including better control of both false positive and false
-negative results. In essence ALDEx2 has been modified to explicitly
-model the scale over a range of reasonable normalization parameters.
-Here, we briefly introduce these modifications and then show how scale
-uncertainty can greatly improve modeling in transcriptome and
-meta-transcriptome datasets to provide more robust and reproducible
-results.
+make some assumption about scale but these implicit assumptions are
+often inappropriate and difficult to interpret. Because of this
+different normalizations provide different outputs when applied to the
+same dataset (6, 8, 26--28). Intuitively, normalizations in widespread
+use assume that either all samples have the same scale,
+e.g.~proportions, rarefaction (29), RPKM (30, 31), etc; or that a subset
+of features in one sample can be chosen as a reference to which the
+others are scaled e.g.~the TMM (32), or LVHA (15) or the additive
+log-ratio (33); or that different sub-parts of each sample maintain a
+constant scale across samples e.g.~the RLE (34); or that the geometric
+mean of the parts is appropriate e.g the CLR (35) and its derivatives.
+
+The original naive ALDEx2 (36) model unwittingly made a strict
+assumption about scale through the CLR normalization (3). While this
+assumption could be useful in many cases it could never be exactly true,
+and others have shown that it is not always the best option (37). Nixon
+et al. (3) showed that better scale assumptions resulted in more
+reproducible data analysis including better control of both false
+positive and false negative results. We recently modified ALDEx2 to
+explicitly model the scale over a range of reasonable normalization
+parameters, and showed significant improvements in performance in
+microbiome and in-vitro selection experiments (16). Here, we briefly
+introduce review modifications and then show how scale uncertainty can
+greatly improve modeling in transcriptome and meta-transcriptome
+datasets to provide more robust and reproducible results.
 
@@
 
-To be concrete, we let \(\mathbf{Y}\) denote the \emph{measured}
-\(D \times N\) matrix of sequence counts with elements
+Formal and expanded descriptions of the concepts that follow are given
+in (3, 16). To be concrete, we let \(\mathbf{Y}\) denote the
+\emph{measured} \(D \times N\) matrix of sequence counts with elements
 \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
-mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
-Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount of
-class \(d\) in the biological system from which sample \(n\) was
+mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
+\(n\). Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount
+of class \(d\) in the biological system from which sample \(n\) was
 obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
@@
 element-wise combination of scale and composition:
-\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
-
-All the normalizations in current use are data transformations that can
-be stated as ratios of the form
-\(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
-denominator is determined by some function of the observation. We use
-the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization
-is attempting to provide an estimate of the true value. The technical
+\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\),
+or as the logarithm
+\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).
+
+Many of the normalizations in widespread use in tools such as DESeq2
+(38), edgeR (32), metagenomeSeq (39) ALDEx2 (40) can be stated as ratios
+of the form
+\(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where
+the denominator is determined by some function of the observation. We
+use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate
+that the output is an estimate of the true value. The technical
 variation in sequencing depth
@@
 
-The ALDEx2 R package (31, 33) is a general purpose toolbox for Bayesian
-modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
-form as a tool for estimating the magnitude and statistical significance
-of Log-Fold-Changes (LFC) (e.g., differential abundance or differential
-expression analysis), but note that it can be used to fit more complex
-linear models. At a high-level, ALDEx2 involves three steps: 1) a
-Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\)
-given observations \(\mathbf{Y}\); 2) a normalization is used to
-estimate \(\mathbf{\hat{W}}\) given the estimate
-\(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates
-are used to estimate LFC as part of differential abundance (or
-expression) analyses. For more details on ALDEx2 see (8, 31).
-
-By default, ALDEx2 uses the CLR normalization which can be written as
-\(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\)
-where \(G_{n}\) denotes the geometric mean of the vector
-\((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR
-normalization equates to an implicit assumption that
-\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)(8) and even slight errors in
-this assumption introduces bias into the LFC estimate that is not
-accounted for when estimating uncertainty (e.g., confidence intervals or
-credible sets). In fact, Nixon et al. (8) showed that the only way in
-which the ALDEx2 model, or any normalization-based model, could ever be
-calibrated (e.g., control Type-I Error rates) was if this assumption was
-exactly true. In support of this assertion it is known that false
-discovery rates vary widely by analysis, dataset, and normalization
-method (34, 35). To fix this, Nixon et al. (8) showed that models should
-incorporate potential error in the assumptions implied by
-normalizations.
-
-Nixon et al.~(2023) generalized the concept of normalizations by
-introducing the concept of a \textit{scale model} to account for
-potential error. Scale models can be incorporated into ALDEx2, turning
-the ALDEx2 model into a specialized type of statistical model which they
-called a \textit{Scale Simulation Random   Variable} (SSRV). They did
-this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). Since
-the CLR normalization makes the assumption
-\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be
-generalized by considering probability models for the scale
-\(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
-example, the following scale model generalizes the CLR:
-
-\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
-
-where \(\gamma\) is a tunable parameter drawn from a Gaussian
-distribution that controls the degree of uncertainty in the CLR
-assumption and \(x_{n}\) denotes a binary condition indicator (e.g.,
-\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). We have made
-those modifications a permanent fixture of ALDEx2 which now represents
-the first software package designed for SSRV-based inference.
+The ALDEx2 R package (36, 40) is a general purpose toolbox to model the
+uncertainty of HTS data and to use that model to estimate LFC
+significance. At a high-level, ALDEx2 has three connected components to
+estimate the uncertainty inherent in HTS datasets. First, the tool
+accounts for the uncertainty of the sequencing counts using Dirichlet
+multinomial sampling to build a probabilistic model of the data; i.e.,
+\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
+Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
+data (36). However, this step was modified recently to account for scale
+uncertainty and misspecification (16) via a scale model, explained with
+more details in (3, 16) and summarized in the next paragraph. Finally, a
+standard null-hypothesis test and a non-parametric estimate of mean
+standardized difference are used to report on the finite sample
+variation. These sources of uncertainty and variation are combined via
+reporting the expected values from a Monte-Carlo simulation framework.
+For simplicity, we use the term `difference' to refer to the absolute
+difference between groups, and `dispersion' to refer to the
+within-condition difference or pooled variance as defined in (36). These
+are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
+16, 36, 40).
+
+Scale models can be incorporated into ALDEx2, turning the ALDEx2 model
+into a specialized type of statistical model which Nixon et al. (3) term
+a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon
+et al. (3) generalized the concept of normalizations by introducing the
+concept of a \textit{scale model} to account for potential error in the
+centred log-ratio normalization step. They did this by including a model
+for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR method used by ALDEx2
+makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where
+\(G_n\) is the geometric mean of sample n, which while being a random
+variable, is essentially constant across each Monte-Carlo replicate.
+With this modificaiton, ALDEx2 can be generalized by considering
+probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that
+have mean \(1/G_{n}\). For example, the following scale model
+generalizes the CLR:
+
+\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2})\].
+
+This formulation is quite flexible (3, 16). In the simple or `default'
+configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
+from a log-Normal distribution that adds scale uncertainty. The default
+approach controls only the degree of uncertainty in the CLR assumption
+for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes
+case and \(x_{n}=0\) denotes control). In the advanced or `informed'
+configuration, \(\mu\) takes different values for each group, and this
+controls the location of the LFC assumption. An example of both the
+default and informed approaches is given for a microbiome dataset in
+(16), and here we show that these approaches also work well in
+transcriptome and metatranscriptome datasets. These modifications are
+instantiated in ALDEx2 which is the first software package designed for
+SSRV-based inference.
 
@@
 
-It is standard practice in many fields of HTS, but especially
-transcriptomics, to use the dual cutoff approach graphically exemplified
-by volcano plots (36, 37) because not all statistically significant
-differences are biologically relevant. Paraphrasing this we can say that
-in in some types of datasets more samples leads to a majority of
-features being statistically significant. A detailed reasoning for this
-using modelled data is explained in two reports by Nixon et al. (8, 9),
-which shows that scaled models completely address this analytic issue.
-
-We use the data of Gierliński et al. (38) who conducted a highly
-replicated yeast transcriptome experiment comparing a wild-type strain
-with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to
-argue that a dual cutoff approach is appropriate to limit the number of
-significant parts and for purposes of reproducibility (37). However, in
-this benchmarking study, the number of significantly different
-transcripts varied between 65\% to \textgreater80\% of all transcripts
-depending on the tool; in essence the desired behavior was that almost
-all transcripts were significantly different. The guidance on
-reproducibility runs counter to standard statistical practice where
-power is intrinsically linked to sample size and very large sample sizes
-are indeed desired (39). Furthermore, the dual-cutoff approach is known
-not to provide appropriate FDR control (5, 6). Schurch et al. (37)
-further suggested that different tools might be better in some
-conditions or datasets than others because each tool has different
-intrinsic statistical power and Type 1 and Type 2 errors. Through the
-lens of scale uncertainty, the behaviour of the tools in this study show
-unacknowledged bias; false confidence in the precision of the estimate
-as sample size increases because o mis-specification in the scale of the
-data. This certainty is driven by the assumptions of the tool not the
-actual experiment being investigated (9).
-
-Using either DESeq2 or ALDEx2, a majority of transcripts are
-statistically significantly different between groups with a
-Benjamini-Hochberg (40) false discovery rate (FDR) of 0.05; i.e.~4636
-(79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
-large numbers of significant transcripts seems biologically unrealistic
-and furthermore breaks the necessary assumption made by the
-normalization methods that the majority of the features must be
-invariant. That 118 transcripts are identified by ALDEx2 and not DESeq2,
-while DESeq2 identifies 582 transcripts that ALDEx2 does not, suggests
-that the choice of normalization plays a role in which results are
-returned as significant and that some if not the majority are driven by
-technical differences in the analysis (22, 23, 37, 41, 42).
+Gierliński et al. (41) generated a highly replicated yeast transcriptome
+dataset to compare gene expression between a wild-type strain and a snf2
+gene knockout, \(\Delta\)snf2. This dataset was used to test several
+RNA-seq tools for their power to detect the set of differentially
+abundant transcripts identified in the full dataset when the data was
+subset (10). In this study each tool had its own `gold standard' set of
+transcripts with different tools identifying between between 65\% to
+\textgreater80\% of all transcripts as being significantly different.
+Since the majority of transcripts were significantly different, the
+authors suggested that it was more appropriate to apply a dual cutoff
+composed of both a Benjamini-Hochberg (42) corrected p-value (q-value)
+plus a difference cutoff to limit the number of identified transcripts
+to a much smaller fraction of the total. Nixon et al, (16) showed that
+adding even a small amount of scale uncertainty with ALDEx2 dramatically
+reduced the number of significant transcripts identified, removing the
+need for the dual cutoff approach in this dataset and others. Below we
+include an intuitive explanation of how incorporating scale uncertainty
+achieved this outcome using a setting of \(\gamma=0.5\), which is in
+line with the recommendations of (16) and that gives results comparable
+to those proposed in (10).
+
+We start with the assumption that not all statistically significant
+differences are biologically relevant (43), and that such a large number
+of significantly different transcripts breaks the necessary assumption
+for DA/DE expression that most parts be invariant (27). As noted,
+transcriptomics commonly uses a dual cutoff approach that is graphically
+exemplified by volcano plots (10, 12). Using either DESeq2 or ALDEx2, a
+majority of transcripts are statistically significantly different
+between groups with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%,
+DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. These values are
+in line with those observed by (10). Such large numbers of statistically
+significant transcripts seems biologically unrealistic. That 118
+transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
+identifies 582 transcripts that ALDEx2 does not, suggests that the
+choice of normalization plays a role in which results are returned as
+significant and that some, if not the majority, are driven by technical
+differences in the analysis (9, 27).
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/plot1-1.pdf}}
-\caption{Effect and volcano plots for unscaled and scaled transcriptome
-analysis. ALDEx2 was used to conduct a differential abundance (DA)
+\caption{Volcano and effect plots for unscaled and scaled transcriptome
+analysis. ALDEx2 was used to conduct a differential expression (DE)
 analysis on the yeast transcriptome dataset. The results were plotted to
 show the relationship between difference and dispersion using effect
-plots ) or difference and the Benjamini-Hochberg corrected p-values
-(volcano plot). Panels A,C are for the unscaled analysis, and Panels B,D
-are for the scaled analysis. Each point represents the values for one
+plots or difference and the q-values using volcano plots. Panels A,C are
+for the naive analyses, and Panels B,D are for the default analyses that
+include scale uncertainty. Each point represents the values for one
 transcript, with the color indicating if that transcript was significant
-in the scaled analysis and unscaled analysis (red) or in the unscaled
-analysis only (orange). Points in grey are not statistically
-signficantly different with any analysis. The horizontal dashed lines
-represent a log2(difference) of \(\pm 1.4\).}
+in the both analyses (red) or in the naive analysis only (orange).
+Points in grey are not statistically signficantly different under any
+condition. The horizontal dashed lines represent a \(\log_2\)difference
+of \(\pm 1.4\).}
 \end{figure}
 
-The effect plots (43) in Figure 1A (ALDEx2) and Supplementary Figure 1
-(DESeq2) shows that the majority of significant transcripts (red,
-orange) have negligible differences between groups and very low
-dispersion and we suggest that this is driven by the experimental design
-(37). Scale uncertainty can be incorporated using the \texttt{gamma}
-parameter that controls the amount uncertainty added to the CLR mean
-assumption when we call either \texttt{aldex()}, or
-\texttt{aldex.clr()}. Figure 1B shows that setting \(\gamma=0.5\)
-results in far fewer transcripts being significant (205) and we observe
-that the minimum dispersion increases from 0.12 (unscaled) to 0.67
-(scaled). The Volcano plots in Figure 1 C and D show a similar story.
-Here we can see that adding scale increases the minimum FDR value and
-increases the concordance between the FDR value and the difference
-between groups (compare panels C and D).
+The Volcano plots in Figure 1 A and B show that adding scale uncertainty
+increases the minimum q-value and increases the concordance between the
+q-value and the difference between groups (compare panels A and B). The
+effect plots (44) in Figure 1C shows that the majority of significant
+transcripts (red, orange) have negligible differences between groups and
+very low dispersion. We suggest that this low dispersion is driven by
+the experimental design which is actually a technical wet lab
+replication rather than a true biological replication design (41). Scale
+uncertainty can be incorporated using the \texttt{gamma} parameter that
+controls the amount of noise added to the CLR mean assumption when we
+call either \texttt{aldex()}, or \texttt{aldex.clr()}. Figure 1 B,D
+shows that setting \(\gamma=0.5\) results in 205 which is far fewer
+significant transcripts than in the naive analysis and we observe that
+the minimum dispersion increases from 0.12 (\(\gamma = 0\) ) to 0.67
+(\(\gamma=0.5\)).
 
@@
 shows the difference \(\Delta\), the red line is the population of
-significant transcripts (*) with no scale, the orange line is the
+significant transcripts (*) with \(\gamma = 0\), the orange line is the
 population of significant transcripts with a difference threshold (T) of
@@
 
-As shown by the effect plot in Figure 1 the root cause of the many
-statistically significant positive transcripts is the very large number
-of transcripts with negligible variance. This phenomenon is not unique
-to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon
-occurs in DESeq2 (and presumably other methods although the relevant
-parameters are not exposed). This issue is not unique to this dataset
-and it is common practice to use a dual-cutoff by choosing transcripts
-based on a thresholds for both corrected p-values and fold-changes (37)
-(here set at \(\pm 2^{1.4}\) for the latter), although considerable
-variation in cutoff values is observed. These limits are shown by the
-dashed grey lines. Here, applying a dual-cutoff using a heuristic of at
-least a \(2^{1.4}\) fold change reduces the number of significant
-outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with
-that observed with that found by ALDEx2 with \(\gamma = 0.5\) which
-identifies 205. Indeed, Supplementary Figure 2, shows that even adding a
-very small amount of scale \(\gamma = 0.1\) reduces the number of
-significant transcripts by more than half and the
-\texttt{aldex.scaleSim()} function can be used to identify those
-transcripts that are significant only because of an absence of scale.
-These results begs the question: why bother with significance tests at
-all if all transcripts with \(> 1.4\)-fold expression change are
-statistically significant?
-
-The effect on dispersion with increasing amounts of scale are shown in
-Figure 2A, where we can see that the dispersion increases as scale is
-added. Note that the dispersion in the unscaled data in Figure 2A
-reaches a minimum near the mid-point of the distribution, and also does
-so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
-This shows more clearly that dispersion of many transcripts is almost
-negligible in the absence of scale and makes the counter-intuitive
-suggestion that the variance in expression of the majority of genes with
-moderate expression is more predictable than highly-expressed genes or
-of housekeeping genes (44). This is at odds with the known biology of
-cells where single cell counting of highly-expressed transcripts shows
-that they have little intrinsic variation (45).
-
-Adding scale by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases
-the minimum dispersion as shown in Figure 2A by the red and blue data
-points, and by the colored lines of fit through the centre of mass of
-the data. Less obvious is that the scaled dispersion estimates are
-rotated. Figure 2B shows a plot of the difference between the
-\(\gamma= 0\) and \(\gamma= 1\) data to show this more clearly and here
-we can see that the scale is preferentially increasing the dispersion of
-the mid-expressed transcripts that formerly had negligible dispersion;
-examine the grey line of best fit (overlaid by the red line) for the
-trend. Panel B also shows the trend of the expression-dispersion
+It is common practice to use a dual-cutoff by choosing transcripts based
+on a thresholds for both q-values and fold-changes (10, 12). Note that
+there is considerable variation in recommended cutoff values(10). Here,
+applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold
+change reduces the number of significant outputs to 193 for DESeq2 and
+to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line
+with the recommendations of (10). the fold change limits are shown by
+the dashed grey lines in Figure 1. The \(2^{1.4}\) fold change cutoff
+identifies a similar number of transcripts as does ALDEx2 using
+\(\gamma = 0.5\) which identifies 205. Supplementary Figure 2 shows how
+to use the \texttt{aldex.senAnalysis()} function to identify those
+transcripts that are very sensitive to scale uncertainty. In this
+supplementary figure we see that even adding a very small amount of
+scale \(\gamma = 0.1\) reduces the number of significant transcripts by
+more than half. This allows us to those low-dispertion transcripts that
+were significant only because of an absence of scale uncertainty. In
+practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
+is realistic for most experimental designs (16).
+
+The effect on dispersion with increasing amounts of scale uncertainty
+are shown in Figure 2A, where we can see that the dispersion increases
+as uncertainty is added. Note that the dispersion in the unscaled
+analysis in Figure 2A reaches a minimum near the mid-point of the
+distribution, and also does so when the analysis is conducted with
+DESeq2 (Supplementary Figure 3). This shows more clearly that dispersion
+of many transcripts is almost negligible in the absence of scale
+uncertainty. This plot makes the counter-intuitive suggestion that the
+variance in expression of the majority of genes with moderate expression
+is more predictable than highly-expressed genes or of housekeeping genes
+(45). This is at odds with the known biology of cells where single cell
+counting of highly-expressed transcripts shows that they have little
+intrinsic variation (20, 46).
+
+Adding scale uncertainty by setting \(\gamma=0.5\) , or
+\(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 2A
+by the red and blue data points, and by the colored lines of fit through
+the centre of mass of the data. Less obvious is that the additional
+dispersion is not applied equally to all points. Figure 2B shows a plot
+of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data and
+here we can see that scale uncertainty is preferentially increasing the
+dispersion of the mid-expressed transcripts that formerly had negligible
+dispersion; examine the grey line of best fit (overlaid by the red line)
+for the trend. Panel B also shows the trend of the expression-dispersion
 relationship for transcripts that are classed as statistically
-significant. The red line shows the trendline with no added scale, and
-this trendline exactly overlays with the grey trendline of the bulk of
-transcripts. The orange trendline indicates those transcripts that are
-both statistically significant and that have a thresholded expression
-level of \(\pm 1.4\), and the dark blue and cyan lines show the
-statistically significant trendline for \(\gamma=0.5\) , or \(1.0\).
-Note that this has the effect of changing the distribution of parts
-identified as significant and that the substantially fewer significantly
-different genes are in the very high abundance but low dispersion
-category.
-
-\subsection{Housekeeping genes can be used to guide scale model
-choices.}\label{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}
-
-Dos Santos et al. (46) used a vaginal metatranscriptome dataset to
+significant. The red line shows the trendline with no added scale
+uncertainty, and this trendline exactly overlays with the grey trendline
+for the bulk of transcripts. The orange trendline indicates those
+transcripts that are both statistically significant and that have a
+thresholded expression level of \(\pm 1.4\), and the dark blue and cyan
+lines show the statistically significant trendline for \(\gamma=0.5\) ,
+or \(1.0\).
+
+Thus, taking Figures 1 and 2 together adding scale uncertainty has the
+desirable effect of changing the distribution of transcripts identified
+as significantly different between groups. Those parts that were
+statistically significantly different \emph{only because of low
+dispersion} are preferentially excluded from statistical significance
+while those parts that were significantly different because of a high
+difference between groups remain.
+
+\subsection{Housekeeping genes and functions to guide scale model
+choices.}\label{housekeeping-genes-and-functions-to-guide-scale-model-choices.}
+
+Dos Santos et al. (47) used a vaginal metatranscriptome dataset to
 compare the gene expression in bacteria collected from healthy (H) and
@@
 relative abundance of species between groups and the gene expression
-level within a species is different (47). Additionally, prior research
+level within a species is different (48). Additionally, prior research
 suggests that the total number of bacteria is about 10 times more in the
-BV than in the H condition (18). Thus, this is an extremely challenging
-environment in which to determine differential abundance as there are
-both compositional and scale changes between conditions. The usual
-method to analyze vaginal metratranscriptome data is on a taxon-by-taxon
-basis (47--49) because the scale confounding can be ignored. Attempts at
-system-wide analysis show that many housekeeping functions are returned
-as differentially abundant between groups; a result likely due to a
-disconnect between the scale assumptions of the normalization used (7).
-
-In this example, we show how to specify a user-defined, or
-\emph{informed}, scale model explicitly that can account for some of
-these modeling difficulties. An informed scale model can control for
-both the mean difference of scale between groups (e.g., directly
-incorporate information on the differences in total number of bacteria
-between the BV and H conditions) as well as the uncertainty assumed in
+BV than in the H condition (23). Thus, these are extremely challenging
+datasets in which to determine differential abundance as there are both
+compositional and scale changes between conditions. The usual method to
+analyze vaginal metratranscriptome data is to do so on an
+organism-by-organism basis (48--50) because the scale confounding of the
+environment is less pronounced. One attempt at system-wide analysis
+returned several housekeeping functions as differentially expressed
+between groups (49); a result likely due to a disconnect between the
+assumptions of the normalization used and the actual scale of the
+environment (15).
+
+In this example, we show how to specify and interpret an informed scale
+model that can explicitly account for some of these modeling
+difficulties (16) even in a difficult dataset. An informed scale model
+can control for both the mean difference of scale between groups (e.g.,
+directly incorporate information on the differences in total number of
+bacteria between the BV and H conditions) as well as the uncertainty of
 that difference. To specify a user-defined scale model, we can pass a
-matrix of scale values instead of a single estimate of gamma to
-\texttt{aldex.clr()}. This matrix should have the same number of rows as
-the of Monte-Carlo Dirichlet samples, and the same number of columns as
-the number of samples. While this matrix can be computed from scratch by
-the analyst, there is an \texttt{aldex.makeScaleModel()} function that
-can be used to simplify this step in most cases. This encodes the scale
-model as \(\Lambda \sim N(log2 \mu_n, \gamma^{2})\), where \(\mu_n\)
-represents the actual scale value for each sample. This can be a
-measured value (cell count, nucleic acid input, etc), or an imputed
-value. Nixon et al. (9) showed that only the ratio between the scale
-values for each sample was important; see Supplementary Figures 4 and 5
-for a demonstration of this point.
+matrix of scale values instead of an estimate of just the scale
+uncertainty to \texttt{aldex.clr()}. This matrix should have the same
+number of rows as the of Monte-Carlo Dirichlet samples, and the same
+number of columns as the number of samples. While this matrix can be
+computed from scratch by the analyst, there is an
+\texttt{aldex.makeScaleModel()} function that can be used to simplify
+this step in most cases. This encodes the scale model as
+\(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents
+the scale value for each sample or group and gamma is the uncertainty as
+before. The scale estimate can be a measured value (cell count, nucleic
+acid input, etc) or an estimate. Nixon et al. (16) showed that only the
+ratio of the means are important when providing values for \(\mu_n\);
+i.e., the ratio between the \(\log_2 \mu_i\) and \(\log_2 \mu_j\)
+values. See the supplement to Nixon et al. (16) for more information.
 
 Figure 3A shows an effect plot of the data where reads are grouped by
-function, corresponding to grouping homologous sequences regardless of
-the organism of origin. Each point represents one of 3728 KEGG functions
-(50). There are many more functions represented in the BV group (bottom)
-than in the healthy group (top). This is because the
-\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
-reduced genome content relative to the anaerobic organisms that dominate
-in BV, because there is a greater diversity of organisms in BV than in H
-samples, and because the BV condition has at least an order of magnitude
-more bacteria than does the H condition.
-
-There are 101 functions with low dispersion that appear to be share by
+homologous sequence regardless of the organism of origin. Each point
+represents one of 3728 KEGG functions (51). There are many more
+functions represented in the BV group (bottom) than in the healthy group
+(top). This is because the \textit{Lactobacilli} that dominate a healthy
+vaginal microbiome have reduced genome content relative to the anaerobic
+organisms that dominate in BV, because there is a greater diversity of
+organisms in BV than in H samples, and because the BV condition has at
+least an order of magnitude more bacteria than does the H condition.
+
+There are 101 functions with low dispersion that appear to be shared by
 both groups (boxed area in Figure 3A, and colored in cyan). Inspection
@@
 chaperones, etc (Supplementary file housekeeping.txt). The transcripts
-of many of these are commonly used as invariant reference sequences (44)
-and so would not be expected to contribute to differences in ecosystem
-behaviour and should be centred on 0 difference. These should not be
-scored as among the most differentially abundant. The major group of
-these housekeeping functions is located off the line of no difference
-(being approximately located at +1.5). While changes in the abundance of
-housekeeping functions is a useful proxy for relative abundance of
-species in the environment, they tell us nothing about the functional
-capacity of the two groups as these are common to every organism. Of
-more interest is determining the functions that are different between
-groups because these are unique or over-expressed in one group relative
-to the other.
+of many of these are commonly used as invariant reference sequences in
+wet lab experiments (45) and so are not be expected to contribute to
+differences in ecosystem behaviour. The average location of these should
+be centred on 0 difference to represent an internal reference set.
+However, without an informed scale model, the mean of these housekeeping
+functions is approximately located at +2.3.
 
@@
 have negative values. Highlighed in the box are KOs that are almost
-exlusively housekeeping functions and are colored cyan. These
+exlusively housekeeping functions; these and are colored cyan. These
 housekeeping functions should be located on the midline of no
@@
 which increase the minimum dispersion as before. Panel C shows the same
-data scaled with \(\gamma = 0.5\) and a 0.15 fold difference in
+data scaled with \(\gamma = 0.5\) and a 0.14 fold difference in
 dispersion applied to the BV samples relative to the H samples. In these
-plots statistically significant (FDR \textless{} 0.01) functions in the
-informed model are in red, false positive functions are in blue,
+plots statistically significant (q-value \textless{} 0.01) functions in
+the informed model are in red, false positive functions are in blue,
 non-significant functions in black and false negative functions are in
@@
 Applying the default scale model of \(\gamma=0.5\) increases the
-dispersion as expected but does little to move the large number of
-housekeeping functions toward the midline of no difference. This is as
-expected; the mean of the default scale model is based on the CLR
-normalization so no shift in location would be expected over the
-original ALDEx2 model. Nevertheless, about 30\% of the housekeeping
-functions are no longer statistically significantly different. Note that
-this change is simple to conduct, has no additional computational
-complexity and requires only a slight modification from the analyst.
-
-Up to this point, scale uncertainty has been applied as an extension of
-the CLR normalization via the default scale model, but a user-defined
-scale adjustment can be applied to each condition, or even each sample
-independently through a custom scale matrix. Our starting point for this
-is to identify the naive scale estimate from the data, and this can be
-done by calculating the mean scale value for each group. The scale
-estimate for samples can be accessed in the \texttt{@scaleSamps} slot
-from the \texttt{aldex.clr} output. In this datset the scale estimate
-for the healthy group is 17.41 and for the BV group is 14.59 for a
-difference of 2.82. This is interpreted as the scale of the H group of
-samples being 7.06 greater than the BV group and this precise but
-incorrect estimate places the location of the housekeeping genes off the
-midline of no difference.
+dispersion slightly but does not move the housekeeping functions toward
+the midline. This is as expected; the mean of the default scale model is
+based on the CLR normalization so no shift in location would be expected
+over the original ALDEx2 model. Nevertheless, about 30\% of the
+housekeeping functions are no longer statistically significantly
+different. Note that this change is simple to conduct, has no additional
+computational complexity and requires only a slight modification for the
+analyst.
 
 We desire a scale model that approximately centres the housekeeping
-functions. We anticipate that housekeeping functions should be nearly
-invariant and so an appropriate scale in this dataset is likely closer
-to 0 than the naive estimate. While a user-defined scale model can be
-quite flexible with relative scales that are distinct for each group (or
-even each sample) along with their uncertainties, here we focus on using
-the \texttt{aldex.makeScaleMatrix()} function. This function uses a
-logNormal distribution to build a scale matrix given a user-specified
-mean difference between groups and uncertainty level. Applying a
-per-group relative differential scale of 0.15 moves the housekeeping
-functions to the midline of no difference (Figure 3C), and applying a
-gamma of 0.5 provides the same dispersion as in panel B). Note that now
-a significant number of functions are differentially up in BV that were
-formerly classed as not different without the full scale model (orange),
-or when only a default scale was applied. Inspection of the functions
-shows that these are largely missing from the \emph{Lactobacillus}
-species and so should actually be captured as differentially abundant in
-the BV group. Thus, applying a differential scale allows us to
-distinguish between both false positives (housekeeping functions in
-cyan, and others in blue) and false negatives (orange functions) even in
-a very difficult to analyze dataset. The remarkable improvements in
-biological interpretation afforded by this full scale model, and the
-transferrability of it between sample cohorts of the same condition is
-outlined elsewhere (46). We suggest that the default scale model is
-sufficient when the data are approximately centred. However, an informed
-model is more appropriate with datasets are not well centred or when the
-investigator has prior information about the underlying biology.
+functions. Our starting point for this is to identify the naive scale
+estimate from the data, and this can be done by calculating the mean
+scale value for each group with a negligible scale value; i.e.,
+\(\gamma=1e-3\). The scale estimate for samples can be accessed in the
+\texttt{@scaleSamps} slot from the \texttt{aldex.clr} output. In this
+datset the naive scale estimate for the healthy group is 17.41 and for
+the BV group is 14.59 for a difference of 2.82. This is interpreted as
+the scale of the H group of samples being 7.06 greater than the BV
+group. This precise but incorrect estimate places the location of the
+housekeeping genes off the midline of no difference.
+
+We anticipate that housekeeping functions should be nearly invariant and
+so an appropriate scale in this dataset is likely closer to 0 than the
+naive estimate. One way to choose an appropriate value for \(\mu_n\) is
+to us the \texttt{aldex.clr} function on only the presumed invariant
+functions setting \(\gamma > 0\), and then accessing the
+\texttt{@scaleSamps} slot as before. Doing so suggests that the
+difference in scale should be about 14\%. A second approach would be to
+identify the functions used as the denominator with the
+\texttt{denom="lvha"} option (15) for the \texttt{aldex.clr} function,
+and then to use these values as before. This approach suggests a 5\%
+difference in scale, and is potentially less subject to user
+interpretation.
+
+For the purposes of this example, if we assume a 14\% difference in
+scale, we can set \(\mu_i = 1\) and \(\mu_j = 1.14\) using the
+\texttt{makeScaleMatrix} function. This function uses a logNormal
+distribution to build a scale matrix given a user-specified mean
+difference between groups and uncertainty level. Applying a per-group
+relative differential scale of 0.14 moves the housekeeping functions to
+the midline of no difference (Figure 3C, 14\% mean difference = -0.24,
+5\% difference = -0.34), and applying a gamma of 0.5 provides the same
+dispersion as in panel B of Figure 3. Note that now a significant number
+of functions are differentially up in BV that were formerly classed as
+not different without the full scale model (orange), or when only a
+default scale was applied. Inspection of the functions shows that these
+are largely missing from the \emph{Lactobacillus} species and so should
+actually be captured as differentially abundant in the BV group. The
+supplement shows that the using either the 5\% or the 14\% scale
+difference give imperceptibly different results suggesting that an
+informed scale model does not have to precisely estimate the scale
+difference to be useful. Nixon et al, (16) found that multiple
+reasonable estimates for the \(\mu_n\) part of the informed scale model
+were similarly useful in microbiome data.
+
+Thus, applying an informed scale allows us to distinguish between both
+false positives (housekeeping functions in cyan, and others in blue) and
+false negatives (orange functions) even in a very difficult to analyze
+dataset. The remarkable improvements in biological interpretation
+afforded by an informed scale model, and the transferrability of it
+between sample cohorts of the same condition is outlined elsewhere (47).
+We suggest that the default scale model is sufficient when the data are
+approximately centred. However, an informed model is more appropriate
+with datasets are not well centred or when the investigator has prior
+information about the underlying biology.
 
@@
 
-\doublespacing
-\singlespacing
-
-Biological systems are both predictably variable and stochastic (45) and
-current measurement methods that rely on high throughput sequencing fail
-to capture all of that variation, particularly variation due to scale
-(8, 9). In the absence of external information (such as spike-in probes
-(14), cell counting (1), FISH (51) etc) sequencing depth normalisation
-methods cannot recapture the scale information (14), and can only
-normalize for the technical variation due to sequencing depth.
-
-Many groups have conducted benchmarking studies on different tools and
-normalizations used for the analysis of datasets such as
-transcriptomes(4, 21, 35, 37, 41) and microbiomes (3, 23, 24, 32, 34,
-52). Generically, it is observed that the actual agreement between
-analysis methods can be modest, and no single method outperforms all
-others in every dataset or type of experiment. That different tools
-appear to work more reliably in different datasets from different
-sources can be explained as different normalizations being a better fit
-to the scale of a particular dataset by chance.
+Biological systems are both predictably variable and stochastic (46) and
+systems biology experiments show that there are transcripts with
+approximately constant concentrations in the cell and those with large
+variability under different growth conditions (20). Current measurement
+methods that rely on high throughput sequencing fail to capture all of
+the variation, particularly variation due to scale (3, 16). In the
+absence of external information (4, 19, 52) sequencing depth
+normalisation methods cannot recapture the scale information (19), and
+can only normalize for the technical variation due to sequencing depth.
+Here we demonstrated that even approximate estimates of scale and the
+uncertainty of measuring it can aid in the interpretation of
+RNA-sequencing experiments.
+
+Nixon et al. (3) introduced the idea of explicitly modelling the scale
+of a HTS dataset, and showed how to incorporate these models in the
+analysis of microbiome and other datasets (16). They demonstrated that
+many tools commonly used to analyze HTS datasets had substantial Type 1
+and Type 2 error rates, in line with recent findings by others (9). A
+version of ALDEx2 with the ability to include scale uncertainty was
+shown to be able to correct for the high Type 1 error rate for that
+tool, albeit with some loss of sensitivity. Finally, they showed that
+incorporating an informed scale model incorporating both location and
+scale uncertainty estimates could both control for Type 1 and Type 2
+error rates (16).
+
+Building and using a scale model, thus has substantial benefits relative
+to the dual cutoff approach that is advocated for many gene expression
+experiments (10, 12). In particular, the dual cutoff approach has long
+been known to not control for Type 1 errors (13, 14), and the frequent
+lack of concordance between tools when benchmarked on transcriptomes(7,
+9, 10, 26, 53) and microbiomes (6, 8, 28, 37, 54, 55) suggests poor
+control of Type 2 errors as well (7, 9). Thus, incorporating a scale
+model during the analysis of HTS data promises the best of both worlds.
+A default scale model can control for Type 1 errors with minimal prior
+knowledge of the environment and this can be done with essentially no
+additional computational overhead. Furthermore, this work and previous
+(16) show that even minimal information about the underlying environment
+can be used to build a relatively robust scale model that controls for
+both Type 1 and 2 error rates.
 
 In the analysis of HTS data it is often observed that larger datasets
-converge on the majority of parts being significantly different (8, 35,
-37), and that different analytic approaches result in different parts
-being chosen as significantly different (4, 24, 37, 41). Li et al. (35)
-conducted a permutation-based benchmarking study and found that popular
-tools performed worse than simple Wilcoxon rank-sum tests in controlling
-the FDR when sample sizes became large. They suggested that the presence
-of outliers were one of the factors driving this observation. Brooks et
-al. (53) suggested that inappropriate choice of benchmarking methods are
-also a major contributing factor. From our perspective, the disagreement
-between tools can be explained by the partially identifiable nature of
-the data; that is, the observed data are consistent with multiple ways
-of estimating the parameters (8). Partially identifiable data exhibit
-multiple pathologies, chief among them being that that different
-analytic approaches can produce different parameter estimates and that
-more data produces worse estimates because the additional data increases
-the precision of a flawed estimate (54).
-
-Scale simulation is now build into ALDEx2 to address the issue of
-partially identifiability, and through this lens the two main root
-causes to common HTS data pathologies can be proposed. The first
-contributing factor is the observed very low dispersion estimate for
-many features that is a by-product of experimental design, sequencing
-and normalization. In the Schurch et al. (37) dataset, the data were
-from single colonies derived from a single culture. Thus, it is more
-accurate to describe the 96 samples as wet-lab technical replicates
-rather than independent samples. However, this type of replication
+converge on the majority of parts being significantly different (3, 9,
+10). Li et al. (9) conducted a permutation-based benchmarking study and
+found that widely used tools performed worse than simple Wilcoxon
+rank-sum tests in controlling the FDR when sample sizes became large.
+They suggested that the presence of outliers were one of the factors
+driving this observation. Brooks et al. (56) suggested that
+inappropriate choice of benchmarking methods are also a major
+contributing factor and that objective standards of truth are important.
+From the perspective of our work the disagreement between tools can be
+explained by the observation that different analytic approaches produce
+different parameter estimates. Thus, more data produces worse estimates
+because the additional data simply increases the precision of a flawed
+estimate (3, 57).
+
+Scale simulation is now build into ALDEx2 (16) and this work suggest
+that there are two main root causes to common HTS data pathologies. The
+first contributing factor is the observed very low dispersion estimate
+for many features that is a by-product of experimental design,
+sequencing and normalization. In the Schurch et al. (10) dataset, the
+data were from single colonies derived from a single culture. Thus, it
+is more accurate to describe the 96 samples as wet-lab technical
+replicates rather than independent samples. This type of replication
 approach is standard in the molecular literature, and would be expected
-to result in the very low dispersion that is observed.
-
-In the yeast transcriptome dataset, applying the default scale model
-with \(\gamma=0.5\) a large number of transcripts with near 0 dispersion
-have had their dispersion increased (Figure 1D), and this results in
-modest number of transcripts, 205, being called significantly different
-as shown in the volcano plot in Figure 1D (red points). In addition,
-there was now a strong concordance between the difference and p-values
-(Figure 1D), this is not surprising. In hindsight, what is not obvious
-is why the unscaled volcano plot shows such poor correspondence. We
-suggest that this can be explained by random fluctuations in the many
-variance estimates with very low values, and this is supported by the
-plot shown in Figure 2B. Furthermore, overplotting the significant
-transcripts identified after adding scale uncertainty on the un-scaled
-analysis shows that adding scale uncertainty removes the need for the
-dual cutoff. Indeed, adding scale uncertainty reduces the significant
-transcripts to a subset of those identified with the dual cutoff that
-have the largest effect size. Thus, incorporating scale uncertainty
-through the default scale model allows us to determine which variables
-are likely to be significant due to sequencing and normalization, and
-which are significantly different even with scale uncertainty included.
-
-While the actual scale of the underlying environment is inaccessible
-post-sequencing, we can study the sensitivity of the results to the
-choice of normalization and thus scale. Varying amounts of scale
-uncertainty is incorporated using the \texttt{gamma} parameter that
-controls the amount uncertainty added to the CLR mean assumption when we
-call either \texttt{aldex()}, or \texttt{aldex.clr()}. The ALDEx2
-package contains a sensitivity analysis function,
-\texttt{aldex.senAnalysis()}, that can be used to explore the effect of
-different amounts of scale uncertainty, and an example is shown in
-Supplementary Figure 2 for the yeast transcriptome dataset. Here it is
-clear that even tiny amounts of scale preclude the majority of
-transcripts from being considered statistically significant. In
-practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
-is realistic for most experimental designs, but further work is needed.
+to result in the very low dispersion that is observed. Applying the
+default scale model with \(\gamma=0.5\) a large number of transcripts
+have their dispersion increased (Figure 1D), with the effect being
+largest for those with the lowest initial dispersion (Figure 2). Adding
+scale uncertainty results in modest number of transcripts, 205, being
+called significantly different as shown in the volcano plot in Figure 1B
+(red points). In addition, there is now a strong concordance between the
+difference and p-values. In hindsight, what is not obvious is why the
+unscaled volcano plot shows such poor correspondence. We suggest that
+this can be explained by random fluctuations in the many variance
+estimates with very low values, and this is supported by the plots shown
+in Figure 2.
 
 The second contributing factor is unacknowledged asymmetry in many
-datasets. In the case of asymmetry, the use of a user-specified scale
-model can be very useful for otherwise difficult-to-analyze datasets
-such as meta-transcriptomes and in-vitro selection datasets where the
-majority of features can change. We showed one such example for the
-metatranscriptome dataset in Figure 3. Here the dataset was highly
-asymmetrical, and the TMM and RLE normalizations could not fully move
-all the housekeeping genes to the midline of no difference or the tools
-exhibited other pathologies (Supplemental Figure 6). Incorporating
-differential scale on a per-group basis moves the mass of the
-housekeeping functions towards the midline of no difference and so
-affects both Type I and Type II error rates. It is also of note that in
-the case of true biological replicates (different individuals) that
+datasets (15); i.e., different gene content or a directional change in
+the majority of features. In the case of asymmetry, the use of a
+user-specified scale model can be very useful for otherwise
+difficult-to-analyze datasets such as meta-transcriptomes and in-vitro
+selection datasets where the majority of features can change. We showed
+one such example for the metatranscriptome dataset in Figure 3. Here the
+dataset was highly asymmetrical. Incorporating differential scale on a
+per-group basis moves the mass of the housekeeping functions towards the
+midline of no difference and so affects both Type I and Type II error
+rates. We showed two ways of estimating the scale difference between
+groups and found that any reasonable estimate is an improvement over the
+default scale model. This is in line with the observations by Nixon et
+al (16) in a 16S rRNA gene sequencing dataset. It is also of note that
+in the case of true biological replicates (different individuals) that
 adding a modest amount of scale \(\gamma=0.5\) had little effect on the
-concordance between the difference between groups and the p-values as
-shown in Supplementary Figure 7. Thus, in this dataset the scale
-mis-specification was affecting mainly the location of the difference
-between groups.
-
-In the metatranscriptome analysis, transcripts that were previously not
-classed as differentially abundant were now called as significantly
-different, and the housekeeping transcripts move from being
-significantly different to not being identified as such. Note the
-assumption that housekeeping genes should not generally be included in
-differential abundance analysis is implicit in the dual
-p-value:fold-change cutoff approach in widespread use. While we
-acknowledge that some prior information on which housekeeping
-transcripts should not be classed as differentially abundant is needed,
-we suggest that this information is widely available and is already used
-when performing the gold-standard quantitative PCR test of differential
-abundance (55, 56). Thus, the use of this prior knowledge is not unique
-to our approach.
+the difference between groups and on the dispersion. Thus, in this
+dataset the scale mis-specification was affecting mainly the location of
+the difference between groups. While we acknowledge that some prior
+information on which housekeeping transcripts should not be classed as
+differentially abundant is needed, we suggest that this information is
+widely available and is already used when performing the gold-standard
+quantitative PCR test of differential abundance (58, 59).
 
@@
 uncertainty essentially tests the model over a range of normalizations
-(8) and so can replace the consensus approach that has been proposed by
-some groups (24, 57) with no additional computational overhead. Thus, an
+(3) and so can replace the consensus approach that has been proposed by
+some groups (8, 60) with no additional computational overhead. Thus, an
 advantage of incorporating scale is that analyses can be made much more
 robust such that actual or potential differences in scale can be tested
-and accounted for explicitly. This occurs because rather than using an
-implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the
-CLR normalization), scale models can make this an explicit part of the
-model building process. While it is beyond the scope of the present
-article, we note that there are many ways of building scale models that
-enhance the interpretability of the parameters and assumptions and a
-detailed description of these points is describe elsewhere (8). In this
-work, we simply use the parameter \(\gamma\) in the default scale model
-as a term that represents uncertainty in the CLR assumption and note
-that larger values of \(\gamma\) correspond to more uncertainty in this
-assumption. In the metatranscriptome example and elsewhere (9), we also
-show that the ratio of scales between conditions can be used to build a
-full scale model when the conditions have dramatically different
-underlying scales.
-
-The ALDEx2 R package is readily amenable for incorporating scale
-uncertainty. Originally, this tool used the only the Dirichlet
-distribution to sample compositional uncertainty, but scale uncertainty
-can be added through the use of a scale model with no additional
-computational complexity. By default, ALDEx2 samples scale from a
-logNormal distribution inspired by the CLR normalization. However, scale
-uncertainty can be sampled from any distribution depending on prior
-knowledge or preference, using the option to introduce a full informed
-scale model that encapsulates both uncertainty and asymmetry in
-underlying scale.
-
-In summary, we supply a toolkit that makes incorporating scale simple to
-incorporate for any type of HTS dataset. While the underlying scale of
-the system is generally inaccessible, the effect of scale on the
-analysis outcomes can be modelled and can help explain some of the
-underlying biology, and known issues with the analysis of HTS data.
-Adding scale information to the analysis allows for more robust
-inference because the features that are sensitive to scale can be
+and accounted for explicitly. While it is beyond the scope of the
+present article, we note that there are many ways of building scale
+models that enhance the interpretability of the parameters and
+assumptions and a detailed description of these points is describe
+elsewhere (3).
+
+In summary, we supply a toolkit that makes incorporating scale
+uncertainty and location information simple to incorporate for
+transcriptomes or indeed any type of HTS dataset. While the underlying
+scale of the system is generally inaccessible, the effect of scale on
+the analysis outcomes can be modelled and can help explain some of the
+underlying biology, and help to expose known issues with the analysis of
+HTS data. Adding scale information to the analysis allows for more
+robust inference because the features that are sensitive to scale can be
 identified and their impact on conclusions weighted accordingly.
-Additionally, the use of user-defined scale models permits difficult to
+Additionally, the use of informed scale models permits difficult to
 analyze datasets to be examined in a robust and principled manner even
 when the majority of features are asymmetrically distributed or
-expressed (or both) in the groups (46). Thus, reporting scale
+expressed (or both) in the groups (47). Thus, using and reporting scale
 uncertainty should become a standard practice in the analysis of HTS
-datasets as a way to identify which features are most robust to
-differences in the underlying system.
+datasets.
+
+\singlespacing
 
@@
 \begin{CSLReferences}{1}{1}
+\bibitem[\citeproctext]{ref-lovell:2011}
+1. Lovell,D., Müller,W., Taylor,J., Zwart,A. and Helliwell,C. (2011)
+Proportions, percentages, ppm: Do the molecular biosciences treat
+compositional data right? In Pawlowsky-Glahn,V., Buccianti,A. (eds),
+\emph{Compositional Data Analysis: Theory and Applications}. John Wiley;
+Sons New York, NY, London, pp. 193--207.
+
+\bibitem[\citeproctext]{ref-Quinn:2019aa}
+2. Quinn,T.P., Erb,I., Gloor,G., Notredame,C., Richardson,M.F. and
+Crowley,T.M. (2019) \href{https://doi.org/10.1093/gigascience/giz107}{A
+field guide for the compositional analysis of any-omics data}.
+\emph{Gigascience}, \textbf{8}.
+
+\bibitem[\citeproctext]{ref-nixon2023scale}
+3. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
+Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
+reliant inference}. \emph{https://arxiv.org/abs/2201.03616}.
+
 \bibitem[\citeproctext]{ref-Vandeputte:2017aa}
-1. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
+4. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
 Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
@@
 \bibitem[\citeproctext]{ref-Props:2017aa}
-2. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
+5. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
 Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
@@
 \bibitem[\citeproctext]{ref-Thorsen:2016aa}
-3. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
+6. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
 Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
@@
 \bibitem[\citeproctext]{ref-Quinn:2018aa}
-4. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
+7. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
 \href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
@@
 
+\bibitem[\citeproctext]{ref-Nearing:2022aa}
+8. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
+Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
+\emph{et al.} (2022)
+\href{https://doi.org/10.1038/s41467-022-28034-z}{Microbiome
+differential abundance methods produce different results across 38
+datasets}. \emph{Nat Commun}, \textbf{13}, 342.
+
+\bibitem[\citeproctext]{ref-Li:2022aa}
+9. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+\href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
+positives by popular differential expression methods when analyzing
+human population samples}. \emph{Genome Biol}, \textbf{23}, 79.
+
+\bibitem[\citeproctext]{ref-Schurch:2016aa}
+10. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
+Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
+al.} (2016) \href{https://doi.org/10.1261/rna.053959.115}{How many
+biological replicates are needed in an RNA-seq experiment and which
+differential expression tool should you use?} \emph{RNA}, \textbf{22},
+839--51.
+
+\bibitem[\citeproctext]{ref-storey2003positive}
+11. Storey,J.D. (2003) The positive false discovery rate: A bayesian
+interpretation and the q-value. \emph{The annals of statistics},
+\textbf{31}, 2013--2035.
+
+\bibitem[\citeproctext]{ref-Cui:2003aa}
+12. Cui,X. and Churchill,G.A. (2003)
+\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
+for differential expression in cDNA microarray experiments}.
+\emph{Genome Biol}, \textbf{4}, 210.1--210.10.
+
 \bibitem[\citeproctext]{ref-Zhang:2009aa}
-5. Zhang,S. and Cao,J. (2009)
+13. Zhang,S. and Cao,J. (2009)
 \href{https://doi.org/10.1186/1471-2105-10-402}{A close examination of
@@
 \bibitem[\citeproctext]{ref-Ebrahimpoor:2021aa}
-6. Ebrahimpoor,M. and Goeman,J.J. (2021)
+14. Ebrahimpoor,M. and Goeman,J.J. (2021)
 \href{https://doi.org/10.1093/bib/bbab053}{Inflated false discovery rate
@@
 \bibitem[\citeproctext]{ref-Wu2021}
-7. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
+15. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
 \href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
@@
 
-\bibitem[\citeproctext]{ref-nixon2023scale}
-8. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
-Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
-reliant inference}.
-
 \bibitem[\citeproctext]{ref-Nixon2024B}
-9. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
+16. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
 normalization: Incorporating scale uncertainty in microbiome and gene
@@
 
-\bibitem[\citeproctext]{ref-lovell:2011}
-10. Lovell,D., Müller,W., Taylor,J., Zwart,A. and Helliwell,C. (2011)
-Proportions, percentages, ppm: Do the molecular biosciences treat
-compositional data right? In Pawlowsky-Glahn,V., Buccianti,A. (eds),
-\emph{Compositional Data Analysis: Theory and Applications}. John Wiley;
-Sons New York, NY, London, pp. 193--207.
-
-\bibitem[\citeproctext]{ref-Quinn:2019aa}
-11. Quinn,T.P., Erb,I., Gloor,G., Notredame,C., Richardson,M.F. and
-Crowley,T.M. (2019) \href{https://doi.org/10.1093/gigascience/giz107}{A
-field guide for the compositional analysis of any-omics data}.
-\emph{Gigascience}, \textbf{8}.
-
 \bibitem[\citeproctext]{ref-Lovell:2015}
-12. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
+17. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
 Bähler,J. (2015)
@@
 \bibitem[\citeproctext]{ref-Nie:2012aa}
-13. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
+18. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
 Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
@@
 \bibitem[\citeproctext]{ref-Loven:2012aa}
-14. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
+19. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
 Burge,C.B., Levens,D.L., Lee,T.I. and Young,R.A. (2012)
@@
 \bibitem[\citeproctext]{ref-Scott:2010}
-15. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
+20. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
 \href{https://doi.org/10.1126/science.1192588}{Interdependence of cell
@@
 \bibitem[\citeproctext]{ref-Yoshikawa:2011aa}
-16. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
+21. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
 Shimizu,H. (2011) \href{https://doi.org/10.1002/yea.1843}{Comprehensive
@@
 \bibitem[\citeproctext]{ref-Lin:2018aa}
-17. Lin,J. and Amir,A. (2018)
+22. Lin,J. and Amir,A. (2018)
 \href{https://doi.org/10.1038/s41467-018-06714-z}{Homeostasis of protein
@@
 \bibitem[\citeproctext]{ref-Zozaya:2010}
-18. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
+23. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
 \href{https://doi.org/10.1128/JCM.00851-09}{Quantitative PCR assessments
@@
 \bibitem[\citeproctext]{ref-Ravel:2010}
-19. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
+24. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
 McCulle,S.L., Karlebach,S., Gorle,R., Russell,J., Tacket,C.O., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hummelen:2010}
-20. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
+25. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
 Changalucha,J., Gloor,G.B. and Reid,G. (2010)
@@
 \bibitem[\citeproctext]{ref-Bullard:2010}
-21. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
+26. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
 \href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
@@
 \bibitem[\citeproctext]{ref-Dillies:2013}
-22. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
+27. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
 Jeanmougin,M., Servant,N., Keime,C., Marot,G., Castel,D., Estelle,J.,
@@
 \bibitem[\citeproctext]{ref-Weiss:2017aa}
-23. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
+28. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
 Lozupone,C., Zaneveld,J.R., Vázquez-Baeza,Y., Birmingham,A., \emph{et
@@
 
-\bibitem[\citeproctext]{ref-Nearing:2022aa}
-24. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
-Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
-\emph{et al.} (2022)
-\href{https://doi.org/10.1038/s41467-022-28034-z}{Microbiome
-differential abundance methods produce different results across 38
-datasets}. \emph{Nat Commun}, \textbf{13}, 342.
-
 \bibitem[\citeproctext]{ref-Hughes:2005tu}
-25. Hughes,J.B. and Hellmann,J.J. (2005)
+29. Hughes,J.B. and Hellmann,J.J. (2005)
 \href{https://doi.org/10.1016/S0076-6879(05)97017-1}{The application of
@@
 \bibitem[\citeproctext]{ref-Mortazavi:2008}
-26. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
+30. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
 (2008) \href{https://doi.org/10.1038/nmeth.1226}{Mapping and quantifying
@@
 \bibitem[\citeproctext]{ref-wagner:tpm}
-27. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
+31. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
 \href{https://doi.org/10.1007/s12064-012-0162-3}{Measurement of mRNA
@@
 \bibitem[\citeproctext]{ref-Robinson:2010a}
-28. Robinson,M.D. and Oshlack,A. (2010)
+32. Robinson,M.D. and Oshlack,A. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-3-r25}{A scaling normalization
@@
 \bibitem[\citeproctext]{ref-aitchison1982}
-29. Aitchison,J. (1982) The statistical analysis of compositional data.
+33. Aitchison,J. (1982) The statistical analysis of compositional data.
 \emph{Journal of the Royal Statistical Society: Series B
@@
 \bibitem[\citeproctext]{ref-Anders:2010}
-30. Anders,S. and Huber,W. (2010)
+34. Anders,S. and Huber,W. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-10-r106}{Differential
@@
 
+\bibitem[\citeproctext]{ref-Aitchison:1986}
+35. Aitchison,J. (1986) The statistical analysis of compositional data
+Chapman \& Hall, London, England.
+
 \bibitem[\citeproctext]{ref-fernandes:2013}
-31. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
+36. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
 (2013) \href{https://doi.org/10.1371/journal.pone.0067019}{ANOVA-like
@@
 \bibitem[\citeproctext]{ref-Yerke:2024aa}
-32. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
+37. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
 \href{https://doi.org/10.1186/s40168-023-01747-z}{Proportion-based
@@
 
+\bibitem[\citeproctext]{ref-Love:2014aa}
+38. Love,M.I., Huber,W. and Anders,S. (2014)
+\href{https://doi.org/10.1186/s13059-014-0550-8}{Moderated estimation of
+fold change and dispersion for RNA-seq data with DESeq2}. \emph{Genome
+Biol}, \textbf{15}, 550.1--550.21.
+
+\bibitem[\citeproctext]{ref-Paulson:2013aa}
+39. Paulson,J.N., Stine,O.C., Bravo,H.C. and Pop,M. (2013)
+\href{https://doi.org/10.1038/nmeth.2658}{Differential abundance
+analysis for microbial marker-gene surveys}. \emph{Nat Methods},
+\textbf{10}, 1200--2.
+
 \bibitem[\citeproctext]{ref-fernandes:2014}
-33. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
+40. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
 Edgell,D.R. and Gloor,G.B. (2014)
@@
 
-\bibitem[\citeproctext]{ref-hawinkel2017}
-34. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
-\href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
-Microbiome differential abundance methods do not control the false
-discovery rate}. \emph{BRIEFINGS IN BIOINFORMATICS}.
-
-\bibitem[\citeproctext]{ref-Li:2022aa}
-35. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
-\href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
-positives by popular differential expression methods when analyzing
-human population samples}. \emph{Genome Biol}, \textbf{23}, 79.
-
-\bibitem[\citeproctext]{ref-Cui:2003aa}
-36. Cui,X. and Churchill,G.A. (2003)
-\href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
-for differential expression in cDNA microarray experiments}.
-\emph{Genome Biol}, \textbf{4}, 210.1--210.10.
-
-\bibitem[\citeproctext]{ref-Schurch:2016aa}
-37. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
-Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
-al.} (2016) \href{https://doi.org/10.1261/rna.053959.115}{How many
-biological replicates are needed in an RNA-seq experiment and which
-differential expression tool should you use?} \emph{RNA}, \textbf{22},
-839--51.
-
 \bibitem[\citeproctext]{ref-Gierlinski:2015aa}
-38. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
+41. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G., Owen-Hughes,T., \emph{et
@@
 
-\bibitem[\citeproctext]{ref-Halsey:2015aa}
-39. Halsey,L.G., Curran-Everett,D., Vowler,S.L. and Drummond,G.B. (2015)
-\href{https://doi.org/10.1038/nmeth.3288}{The fickle p value generates
-irreproducible results}. \emph{Nat Methods}, \textbf{12}, 179--85.
-
 \bibitem[\citeproctext]{ref-benjamini:1995}
-40. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
+42. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
 rate: A practical and powerful approach to multiple testing.
@@
 
-\bibitem[\citeproctext]{ref-Soneson:2013}
-41. Soneson,C. and Delorenzi,M. (2013)
-\href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
-for differential expression analysis of {RNA-seq} data}. \emph{BMC
-Bioinformatics}, \textbf{14}, 91.
-
-\bibitem[\citeproctext]{ref-maza2013}
-42. Maza,E., Frasse,P., Senin,P., Bouzayen,M. and Zouine,M. (2013)
-\href{https://doi.org/10.4161/cib.25849}{Comparison of normalization
-methods for differential gene expression analysis in RNA-seq
-experiments: A matter of relative size of studied transcriptomes}.
-\emph{Commun Integr Biol}, \textbf{6}, e25849.
+\bibitem[\citeproctext]{ref-efron2008FDR}
+43. Efron,B. (2008) Microarrays, empirical bayes and the two-groups
+model. \emph{Statist. Sci.}, \textbf{23}, 1--22.
 
 \bibitem[\citeproctext]{ref-gloor:effect}
-43. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
+44. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
 \href{https://doi.org/10.1080/10618600.2015.1131161}{Displaying
@@
 \bibitem[\citeproctext]{ref-Rocha:2020aa}
-44. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
+45. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
 (2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
@@
 \bibitem[\citeproctext]{ref-Taniguchi:2010aa}
-45. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
+46. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
 Emili,A. and Xie,X.S. (2010)
@@
 \bibitem[\citeproctext]{ref-dosSantos:2024}
-46. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
+47. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
 Gloor,G.B. (2024) Vaginal metatranscriptome meta-analysis reveals
@@
 \bibitem[\citeproctext]{ref-macklaim:2013}
-47. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
+48. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
 and Gloor,G.B. (2013)
@@
 \bibitem[\citeproctext]{ref-Denge00262-18}
-48. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
+49. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
 Wagner-Döbler,I. (2018)
@@
 \bibitem[\citeproctext]{ref-Fettweis:2019aa}
-49. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
+50. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
 Parikh,H.I., Huang,B., Arodz,T.J., Edupuganti,L., Glascock,A.L.,
@@
 \bibitem[\citeproctext]{ref-Okuda:2008}
-50. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
+51. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
 Goto,S. and Kanehisa,M. (2008)
@@
 \bibitem[\citeproctext]{ref-yeast-absolute}
-51. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
+52. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
 Bähler,J. (2012)
@@
 
+\bibitem[\citeproctext]{ref-Soneson:2013}
+53. Soneson,C. and Delorenzi,M. (2013)
+\href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
+for differential expression analysis of {RNA-seq} data}. \emph{BMC
+Bioinformatics}, \textbf{14}, 91.
+
 \bibitem[\citeproctext]{ref-McMurdie:2014a}
-52. McMurdie,P.J. and Holmes,S. (2014)
+54. McMurdie,P.J. and Holmes,S. (2014)
 \href{https://doi.org/10.1371/journal.pcbi.1003531}{Waste not, want not:
@@
 
+\bibitem[\citeproctext]{ref-hawinkel2017}
+55. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
+\href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
+Microbiome differential abundance methods do not control the false
+discovery rate}. \emph{BRIEFINGS IN BIOINFORMATICS}.
+
 \bibitem[\citeproctext]{ref-Brooks:2024aa}
-53. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
+56. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
 \href{https://doi.org/10.1038/s41576-023-00679-6}{Challenges and best
@@
 \bibitem[\citeproctext]{ref-gustafson2015bayesian}
-54. Gustafson,P. (2015) Bayesian inference for partially identified
+57. Gustafson,P. (2015) Bayesian inference for partially identified
 models: Exploring the limits of limited data CRC Press.
@@
 \bibitem[\citeproctext]{ref-Thellin:1999aa}
-55. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
+58. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
 Hennen,G., Grisar,T., Igout,A. and Heinen,E. (1999)
@@
 \bibitem[\citeproctext]{ref-SEQCux2fMAQC-III-Consortium:2014aa}
-56. SEQC/MAQC-III Consortium (2014)
+59. SEQC/MAQC-III Consortium (2014)
 \href{https://doi.org/10.1038/nbt.2957}{A comprehensive assessment of
@@
 \bibitem[\citeproctext]{ref-Song:2023aa}
-57. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
+60. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
 Klatt,N.R., Hensley-McBain,T. and Wu,M.C. (2023)

### 2024-08-22 go3.Rmd
with Justin's suggestions
Send back to Justin for comment

@@
 ---
-title: "Normalizations are not what you think; Explicit Scale Simulation in ALDEx2
-"
+title: "Explicit Scale Simulation for analysis  of RNA-sequencing  with ALDEx2"
 shorttitle: "Scale ALDEx2"
@@
 package: ALDEx2
-abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total gene expression) of the underlying biological systems.
+abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total mRNA expression) of the underlying biological systems.
 Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses.
-Recently we showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can lead to dramatic increases in false positive and false negative rates.
-Here we describe updates to the ALDEx2 R package that mitigate these problems by directly modeling uncertainty in the unmeasured system scale through the use of a \textit{scale model}.
-Scale models generalize the idea of normalizations and can be thought of as explicitly modeling the error in normalization by examining a distribution over all possible  normalizations.
-Beyond enhancing the robustness of HTS analyses,  the use of scale models within ALDEx2 enhances the transparency and  reproducibility of analyses by making implicit normalizing assumptions an explicit part of the model building process.
+We recently showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can dramatically increase false positive and false negative rates.
+We demonstrated that these errors can be mitigated by accounting for uncertainty about scale using a *scale model*, which we integrated into the ALDEx2 R package.
+This article provides new insights into those methods, focusing on the application to transcriptomic analysis.
+Here we provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses.
+We show that these scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance.
+We demonstrate the utility of that scale models   built based on known housekeeping genes in complex metatranscriptomic datasets.
+Thus this work provides example and practical guidance on how to incorporate scale into transcriptomic analysis.
 
@@
 # Introduction
+
+\doublespacing
 
@@
 (merging) multiple libraries together, and then applying a sample of the
-multiplexed library to the flow cell. Each of these steps is a 
+multiplexed library to a flow cell. Each of these steps is a 
 compositional sampling step as only a fixed-size subsample of nucleic acid is 
-carried over to subsequent steps. Thus, with
+carried over to subsequent steps. 
+Thus, with
 each sampling step the connection between the size of the sampled DNA
@@
 expression) of the measured biological system is degraded or lost.
+In the end, the information contained in the data relates only to relative abundances and  has an arbitrary scale imposed by the sequencing process[@lovell:2011;@Quinn:2019aa;@nixon2023scale].
 Increasingly, researchers are turning to modified experimental protocols
-(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
+ in an attempt to recover the lost
 biological variation in scale; see for example  [@Vandeputte:2017aa;@Props:2017aa]. 
-However, these protocols often do not recover
-meaningful biological variation but instead recover scale variation at
-the intermediate step in the sample preparation protocol where the DNA spike-in was added. 
+However, these protocols only recover scale variation at
+the intermediate step in the sample preparation protocol where the intervention was made. 
 In short, the disconnect
 between sample-to-sample variation in sequencing depth and biological
-variation in scale remains an outstanding challenge.
+variation in scale remains an open challenge.
 
@@
 traced, in whole or in part, to  misspecification of scale. The first issue 
-is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;Nearing:2022aa;@Li:2022aa], 
+is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Li:2022aa], 
   exhibited as dataset-dependent 
-FDR control.  The FDR problem is connected to the double-filtering approach
-that is often used, but which is known not to have appropriate FDR control 
+FDR control and by the disconnect between statistical and biological significance[@Schurch:2016aa].
+In current practice,  these issues are addressed  by a dual-filtering method, whereby both a low p-value (or equivalently a low q-value following FDR correction [@storey2003positive]) and  a large difference between groups is used to identify interesting transcripts or genes for follow-up analysis[@Schurch:2016aa].
+This double-filtering approach is graphically exemplified by the volcano plot [@Cui:2003aa], but
+ is known to not   appropriately control the  FDR  
 [@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
-analyzing asymmetric data; data that arises because of a bias in the direction of change.
+analyzing data where the mean change between groups is non-zero.
+Such asymmetric data can arise when a gene set is expressed in one group but not the other, or when one group contains different gene content from the other.
 This type of data frequently arises in in-vitro selection experiments (SELEX), 
-transcriptome analysis, and microbiome analysis [@Wu2021]. The third issue is that these
+transcriptome analysis, and microbiome analysis [@Wu2021]. 
+The third issue is that these
 problems become more pronounced as more samples are collected; that is,
-more information results in a worsening of the accuracy of the analysis [@nixon2023scale,@Nixon2024B]. 
-The final problem is that these data are relative data, i.e. compositional [@lovell:2011;@Quinn:2019aa], 
-and substantial effort has been put into removing this constraint.
+more information results in a worsening of the accuracy of the analysis [@Li:2022aa;@nixon2023scale,@Nixon2024B].
 
-The first three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.
+The  three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.
 Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Lovell:2015].
@@
 Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).
-Current methods access only the compositional information yet make implicit assumptions about the scale.
+Current methods access only the compositional information yet make implicit assumptions about the scale [@nixon2023scale,@Nixon2024B].
 
@@
 [-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models.
-They showed that \emph{all} normalizations make some assumption about  scale.
-These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
+They showed that \emph{all} normalizations make some assumption about  scale but these implicit assumptions are often inappropriate and difficult to interpret.
+Because of this different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
 Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g.
@@
 the TMM [@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g.
-the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR  and its derivatives.
+the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@Aitchison:1986] and its derivatives.
 
-The original ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale].
+The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale].
 While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa].
 Nixon et al.
-[-@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results.
-In essence ALDEx2 has been modified to explicitly model the scale over a range of reasonable normalization parameters.
-Here, we briefly introduce these modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.
+[@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results.
+We recently modified ALDEx2  to explicitly model the scale over a range of reasonable normalization parameters, and showed significant improvements in performance in microbiome and in-vitro selection experiments [@Nixon2024B].
+Here, we briefly introduce review modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.
 
@@
 
-To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
+Formal and expanded descriptions of the concepts that follow are given in [@nixon2023scale;@Nixon2024B].
+To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).
 Likewise, we can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained.
@@
 That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
-Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
+Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).
 
-All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation.
-We use the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization is attempting to provide an estimate of the true value.
+Many of  the normalizations in widespread use in tools such as DESeq2 [@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2 [@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation.
+We use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate that the output is an estimate of the true value.
 The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al.
@@
 
-The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox for Bayesian modeling of HTS data.
-For brevity, we discuss ALDEx2 in its simplest form as a tool for estimating the magnitude and statistical significance of Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models.
-At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{\hat{W}}\) given the estimate \(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses.
-For more details on ALDEx2 see [@fernandes:2013;@nixon2023scale].
+The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox to  model the uncertainty of HTS data and to use that model to estimate LFC significance.
+At a high-level, ALDEx2  has three connected components to estimate the uncertainty inherent in HTS datasets.
+First, the tool accounts for the uncertainty of the sequencing counts using Dirichlet multinomial sampling to build a probabilistic model of the data; i.e., \(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
+Secondly, ALDEx2 uses the centred log-ratio transformation to scale the data [@fernandes:2013].
+However, this step was modified recently to account for scale uncertainty and misspecification [@nixon2023scale,@Nixon2024B] via a scale model, explained  with more details in [@nixon2023scale;@Nixon2024B] and summarized in the next paragraph.
+Finally, a standard null-hypothesis test and a non-parametric estimate of mean standardized difference are used to report on the finite sample variation.
+These sources of uncertainty and variation are combined via reporting the expected values from a Monte-Carlo simulation framework.
+For simplicity, we use the term 'difference' to refer to the absolute difference between groups, and 'dispersion' to refer to the  within-condition difference or pooled variance as defined in [@fernandes:2013].
+These are calculated on a \(\log_2\) scale.
+For more details on ALDEx2 see [@fernandes:2013;@fernandes:2014;@nixon2023scale;@Nixon2024B].
 
-By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\) where \(G_{n}\) denotes the geometric mean of the vector \((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\).
-The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale;@Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets).
-In fact, Nixon et al.
-[-@nixon2023scale] showed that the only way in which the ALDEx2 model, or any normalization-based model, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.
-In support of this assertion it is known that false discovery rates vary widely by analysis, dataset, and normalization method [@hawinkel2017;@Li:2022aa,Nixon2024B].
-To fix this, Nixon et al.
-[-@nixon2023scale] showed that models should incorporate potential error in the assumptions implied by normalizations.
-
-Nixon et al.
-(2023) generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error.
-Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV).
+Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which Nixon et al.
+[@nixon2023scale] term  a \textit{Scale Simulation Random   Variable} (SSRV).
+To do this, Nixon et al.
+[@nixon2023scale] generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error in the centred log-ratio normalization step.
 They did this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\).
-Since the CLR normalization makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\).
+The CLR method used by ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the geometric mean of sample n, which while being a random variable, is essentially constant across each Monte-Carlo replicate.
+With this modificaiton, ALDEx2 can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\).
 For example, the following scale model generalizes the CLR:
 
-\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]
+\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2})\].
 
-where \(\gamma\) is a tunable parameter drawn from a Gaussian distribution that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control).
-We have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.
+This formulation is quite flexible [@nixon2023scale;@Nixon2024B].
+In the simple or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter drawn from a log-Normal distribution that adds scale uncertainty.
+The default approach controls only the degree of uncertainty in the CLR assumption for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control).
+In the advanced or 'informed' configuration, \(\mu\) takes different values for each group, and this controls the location of the LFC assumption.
+An example of both the default and informed approaches is given for a microbiome dataset in @Nixon2024B, and here we show that these approaches also work well in transcriptome and metatranscriptome datasets.
+These modifications are  instantiated in ALDEx2 which is  the first software package designed for SSRV-based inference.
 
@@
 
-It is standard practice in many fields of HTS, but especially transcriptomics, to use the dual cutoff approach graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa] because not all statistically significant differences are biologically relevant.
-Paraphrasing this we can say that in in some types of datasets more samples leads to a majority of features being statistically significant.
-A detailed reasoning for this using modelled data is explained in two reports by Nixon et al.
-[-@nixon2023scale;-@Nixon2024B], which shows that scaled models completely address this analytic issue.
-
-We use the data of Gierliński et al.
-[-@Gierlinski:2015aa] who conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf2 gene knockout, \(\Delta\)snf2.
-This dataset has been used to argue that a dual cutoff approach is appropriate to limit the number of significant parts and for purposes of reproducibility [@Schurch:2016aa].
-However, in this benchmarking study,  the number of significantly different transcripts varied between 65% to >80% of all transcripts depending on the tool; in essence the desired behavior was that almost all transcripts were significantly different.
-The guidance on reproducibility runs counter to standard statistical practice where power is intrinsically linked to sample size and very large sample sizes are indeed desired [@Halsey:2015aa].
-Furthermore, the dual-cutoff approach is  known not to provide appropriate FDR control [@Zhang:2009aa; @Ebrahimpoor:2021aa].
-Schurch et al.
--@Schurch:2016aa further suggested that different tools might be better in some conditions or datasets than others because each tool has different intrinsic statistical power and Type 1 and Type 2 errors.
-Through the lens of scale uncertainty, the behaviour of the tools in this study show unacknowledged bias; false confidence in the precision of the estimate as sample size increases because o mis-specification  in the scale of the data.
-This certainty is driven by the assumptions of the tool not the actual experiment being investigated [@Nixon2024B].
+Gierliński et al.
+[@Gierlinski:2015aa]  generated a  highly replicated yeast transcriptome dataset to compare gene expression between a wild-type strain and a snf2 gene knockout, \(\Delta\)snf2.
+This dataset was  used to test several RNA-seq tools for their power  to detect the set of differentially abundant transcripts identified in the full dataset when the data was subset [@Schurch:2016aa].
+In this  study each tool had its own 'gold standard' set of transcripts with different tools identifying between between 65% to >80% of all transcripts as being significantly different.
+Since the majority of transcripts were significantly different, the authors suggested that it was more appropriate to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995] corrected p-value (q-value) plus a difference cutoff  to limit the number of identified transcripts to a much smaller fraction of the total.
+Nixon et al, @Nixon2024B showed that adding even a small amount of  scale uncertainty with ALDEx2  dramatically reduced the number of significant transcripts identified, removing the need for the dual cutoff approach in this dataset and others.
+Below we include an intuitive explanation of how incorporating scale uncertainty achieved this outcome using a setting of \(\gamma=0.5\), which is in line with the recommendations of [@Nixon2024B] and that gives results comparable to those proposed in [@Schurch:2016aa].
 
@@
 
-Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.05; i.e.
+We start with the assumption that not all statistically significant differences are biologically relevant [@efron2008FDR], and that such a large number of significantly different transcripts breaks the  necessary assumption for DA/DE expression that most parts be invariant [@Dillies:2013].
+As noted,  transcriptomics commonly uses a dual cutoff approach that is graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa].
+Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a  q-value cutoff of \( \le 0.05\); i.e.
 `r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or `r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2) of the `r nrow(yst.all)` transcripts.
-Such  large numbers of significant transcripts seems biologically unrealistic  and furthermore  breaks the necessary assumption made by the normalization methods that the majority of the features must be invariant.
-That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some if not the majority are driven by technical differences in the analysis [@Soneson:2013;@maza2013;@Dillies:2013;@Weiss:2017aa; @Schurch:2016aa].
+These values are in line with those observed by [@Schurch:2016aa].
+Such  large numbers of statistically significant transcripts seems biologically unrealistic.
+That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some, if not the majority, are driven by technical differences in the analysis [@Dillies:2013;@Li:2022aa].
 
-```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis.
-ALDEx2 was used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset.
-The results were plotted to show the relationship between difference and dispersion using effect plots ) or difference and the Benjamini-Hochberg corrected p-values (volcano plot).
-Panels A,C are for the unscaled analysis, and Panels B,D are for the scaled analysis.
-Each point represents the values for one transcript, with the color indicating if that transcript was significant in the scaled analysis and unscaled analysis (red) or in the unscaled analysis only (orange).
-Points in grey are not statistically signficantly different with any analysis.
-The horizontal dashed lines represent a log2(difference) of \\(\\pm 1.4\\)."}
+```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis.
+ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset.
+The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots.
+Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty.
+Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange).
+Points in grey are not statistically signficantly different under any condition.
+The horizontal dashed lines represent a \\(\\log_2\\)difference of \\(\\pm 1.4\\)."}
 
 par(mfrow=c(2,2))
+# volcano
+#plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
+#  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
+#title('D: DESeq2 volcano', adj=0, line= 0.8)
+#points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
+#  pch=19, cex=0.5)
+#points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
+#  pch=19, cex=0.5)
+#  abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+
+l10q <- expression(-log[10](q))
+l2d <- expression(log[2]~ Difference)
+
+plot(yst.all$diff.btw, -1*log10(yst.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
+   xlab=l2d, ylab=l10q)
+title(expression(paste("A: Volcano: ", gamma, " = 0")), adj=0, line= 0.8)
+points(yst.all$diff.btw[sig.ald], -1*log10(yst.all$we.eBH[sig.ald]	+1e-70), 
+  col=rgb(1,.66,0,.3), cex=0.5, pch=19)
+points(yst.all$diff.btw[sig.s.ald], -1*log10(yst.all$we.eBH[sig.s.ald]	+1e-70), 
+  col=rgb(1,0,0,1), cex=0.5, pch=19)
+abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+
+plot(yst.s.all$diff.btw, -1*log10(yst.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
+   xlab=l2d, ylab=l10q)
+title(expression(paste("B: Volcano: ", gamma, " = 0.5")), adj=0, line= 0.8)
+points(yst.s.all$diff.btw[sig.s.ald], -1*log10(yst.s.all$we.eBH[sig.s.ald]+1e-15),
+  col=rgb(1,0,0,1), cex=0.5, pch=19)
+	abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
@@
 plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), 
-   xlab='Dispersion', ylab='log2 Difference')
-title(expression(paste("A: Effect: ", gamma, " = 0")), adj=0, line= 0.8)
+   xlab='Dispersion', ylab=l2d)
+title(expression(paste("C: Effect: ", gamma, " = 0")), adj=0, line= 0.8)
 points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald], 
@@
 plot(yst.s.all$diff.win, yst.s.all$diff.btw, col=rgb(0,0,0,0.1),
-   xlab='Dispersion', ylab='log2 Difference', xlim=c(0.1,5))
-title(expression(paste("B: Effect: ", gamma, " = 0.5")), adj=0, line= 0.8)
+   xlab='Dispersion', ylab=l2d, xlim=c(0.1,5))
+title(expression(paste("D: Effect: ", gamma, " = 0.5")), adj=0, line= 0.8)
 points(yst.s.all$diff.win[sig.s.ald], yst.s.all$diff.btw[sig.s.ald], 
@@
 
-# volcano
-#plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
-#  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
-#title('D: DESeq2 volcano', adj=0, line= 0.8)
-#points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
-#  pch=19, cex=0.5)
-#points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
-#  pch=19, cex=0.5)
-#  abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
-
-plot(yst.all$diff.btw, -1*log10(yst.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
-   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
-title(expression(paste("C: Volcano: ", gamma, " = 0")), adj=0, line= 0.8)
-points(yst.all$diff.btw[sig.ald], -1*log10(yst.all$we.eBH[sig.ald]	+1e-70), 
-  col=rgb(1,.66,0,.3), cex=0.5, pch=19)
-points(yst.all$diff.btw[sig.s.ald], -1*log10(yst.all$we.eBH[sig.s.ald]	+1e-70), 
-  col=rgb(1,0,0,1), cex=0.5, pch=19)
-abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
-
-plot(yst.s.all$diff.btw, -1*log10(yst.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
-   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
-title(expression(paste("D: Volcano: ", gamma, " = 0.5")), adj=0, line= 0.8)
-points(yst.s.all$diff.btw[sig.s.ald], -1*log10(yst.s.all$we.eBH[sig.s.ald]+1e-15),
-  col=rgb(1,0,0,1), cex=0.5, pch=19)
-	abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
-
 ```
 
-The effect plots  [@gloor:effect] in Figure 1A (ALDEx2) and Supplementary Figure 1 (DESeq2) shows that the majority of significant transcripts (red, orange) have negligible differences between groups and  very low dispersion and we suggest that this is driven by the experimental design [@Schurch:2016aa].
-Scale uncertainty can be incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
-Figure 1B shows that  setting  \( \gamma=0.5 \) results in far fewer transcripts being significant (`r length(sig.s.ald)`) and we observe that  the minimum dispersion increases from  `r round(min(yst.all$diff.win), 2)` (unscaled) to  `r round(min(yst.s.all$diff.win), 2)` (scaled).
-The Volcano plots in Figure 1 C and D show a similar story.
-Here we can see that adding scale increases the minimum FDR value and increases the concordance between the FDR value and the difference between groups (compare panels C and D).
+The Volcano plots in Figure 1 A and B show  that adding scale uncertainty increases the minimum q-value and increases the concordance between the q-value and the difference between groups (compare panels A and B).
+The effect plots  [@gloor:effect] in Figure 1C  shows that the majority of significant transcripts (red, orange) have negligible differences between groups and  very low dispersion.
+We suggest that this low dispersion is driven by the experimental design which is actually a technical wet lab replication  rather than a true biological replication design [@Gierlinski:2015aa].
+Scale uncertainty can be incorporated using the \texttt{gamma} parameter that controls the amount of noise added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
+Figure 1 B,D shows that  setting  \( \gamma=0.5 \) results in `r length(sig.s.ald)` which is far fewer significant transcripts than in the naive analysis  and we observe that  the minimum dispersion increases from  `r round(min(yst.all$diff.win), 2)` (\(\gamma = 0\) ) to  `r round(min(yst.s.all$diff.win), 2)` (\(\gamma=0.5\)).
 
@@
 The colored lines indicate the lowess line of fit through the centre of mass of the plot for the various  populations of points.
-The grey line is the total population and shows the difference \\(\\Delta\\), the red line is the population of significant transcripts (\\*) with no scale, the orange line is the population of significant transcripts with a difference threshold (T) of about \\(\\pm 2^{1.4}\\)-fold change, the blue line is the population of significant transcripts with \\(\\gamma = 0.5\\), and the cyan line is the significant population with \\(\\gamma = 1\\).
+The grey line is the total population and shows the difference \\(\\Delta\\), the red line is the population of significant transcripts (\\*) with \\(\\gamma = 0\\), the orange line is the population of significant transcripts with a difference threshold (T) of about \\(\\pm 2^{1.4}\\)-fold change, the blue line is the population of significant transcripts with \\(\\gamma = 0.5\\), and the cyan line is the significant population with \\(\\gamma = 1\\).
 \\(\\Delta\\): Difference, *: significant, T: thresholded.
@@
 plot(yst.all$rab.all, yst.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
-  xlab='rAbundance', ylab='dispersion')
+  xlab='rAbundance', ylab='Dispersion')
 lines(lowess(no.scale, f=0.1), col='grey', lwd=3)
@@
 # plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3), ylim=c(0,4.5), 
-#   xlab='g=0 dispersion', ylab='g=0.5 | g=1 dispersion')
+#   xlab='g=0 Dispersion', ylab='g=0.5 | g=1 Dispersion')
 # points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
@@
 
-ylab1 <- expression(paste("dispersion: ", gamma, "1 - ", gamma, "0"))
+ylab1 <- expression(paste("Dispersion: ", gamma, "1 - ", gamma, "0"))
 
@@
 
-As shown by the effect plot in Figure 1 the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance.
-This phenomenon is not unique to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon occurs in DESeq2 (and presumably other methods although the relevant parameters are not exposed).
-This issue is not unique to this dataset and it is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both corrected p-values and fold-changes  [@Schurch:2016aa] (here set at \(\pm 2^{1.4}\) for the latter), although considerable variation in cutoff values is observed.
-These limits are shown by the dashed grey lines.
-Here, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with that observed with that found by ALDEx2 with \( \gamma = 0.5 \) which identifies `r length(sig.s.ald)`.
-Indeed, Supplementary Figure 2, shows that even adding a very small amount of scale \( \gamma = 0.1 \) reduces the number of significant transcripts by more than half and the \texttt{aldex.scaleSim()} function can be used to identify those transcripts that are significant only because of an absence of scale.
-These results begs the question: why bother with significance tests at all if all transcripts with \( > 1.4\)-fold expression change are statistically significant?
+It is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both q-values and fold-changes  [@Cui:2003aa;@Schurch:2016aa].
+Note that there is considerable variation in recommended cutoff values[@Schurch:2016aa].
+Here, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
+This cutoff was chosen for convenience and is in-line with the recommendations of [@Schurch:2016aa].
+the fold change limits are shown by the dashed grey lines in Figure 1.
+The  \(2^{1.4}\) fold change cutoff identifies a similar number of transcripts as does  ALDEx2 using \( \gamma = 0.5 \) which identifies `r length(sig.s.ald)`.
+Supplementary Figure 2 shows how to use the \texttt{aldex.senAnalysis()} function to identify those transcripts that are very sensitive to scale uncertainty.
+In this supplementary figure we see that even adding a very small amount of scale \( \gamma = 0.1 \) reduces the number of significant transcripts by more than half.
+This allows us to those low-dispertion transcripts that were  significant only because of an absence of scale uncertainty.
+In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs [@Nixon2024B].
 
-The effect on dispersion with increasing  amounts of scale are shown in Figure 2A, where we can see that the  dispersion increases as scale is added.
-Note that the dispersion in the unscaled data in Figure 2A reaches a minimum near the mid-point of the distribution, and also does so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
-This shows more clearly that dispersion of many transcripts is  almost negligible in the absence of scale and makes the counter-intuitive suggestion that the variance in expression of the majority of genes with moderate expression is more predictable  than highly-expressed genes or of housekeeping genes [@Rocha:2020aa].
-This is at odds with the known biology of cells where single cell counting of highly-expressed transcripts shows that they have little intrinsic variation [@Taniguchi:2010aa].
+The effect on dispersion with increasing  amounts of scale uncertainty are shown in Figure 2A, where we can see that the  dispersion increases as uncertainty is added.
+Note that the dispersion in the unscaled analysis in Figure 2A reaches a minimum near the mid-point of the distribution, and also does so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
+This shows more clearly that dispersion of many transcripts is  almost negligible in the absence of scale uncertainty.
+This plot makes the counter-intuitive suggestion that the variance in expression of the majority of genes with moderate expression is more predictable  than highly-expressed genes or of housekeeping genes [@Rocha:2020aa].
+This is at odds with the known biology of cells where single cell counting of highly-expressed transcripts shows that they have little intrinsic variation [@Taniguchi:2010aa;@Scott:2010].
 
-Adding scale by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 2A by the red and blue data points, and by the colored lines of fit through the centre of mass of the data.
-Less obvious is that the scaled dispersion estimates are rotated.
-Figure 2B shows a plot of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data to show this more clearly and here we can see that the scale is preferentially increasing the dispersion of the mid-expressed transcripts that formerly had negligible dispersion; examine the grey line of best fit (overlaid by the red line) for the trend.
+Adding scale uncertainty by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 2A by the red and blue data points, and by the colored lines of fit through the centre of mass of the data.
+Less obvious is that the additional dispersion is  not applied equally to all points.
+Figure 2B shows a plot of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data  and here we can see that  scale uncertainty is preferentially increasing the dispersion of the mid-expressed transcripts that formerly had negligible dispersion; examine the grey line of best fit (overlaid by the red line) for the trend.
 Panel B also shows the trend of the expression-dispersion relationship for transcripts that are classed as statistically significant.
-The red line shows the trendline with no added scale, and this trendline exactly overlays with the grey trendline of the bulk of transcripts.
+The red line shows the trendline with no added scale uncertainty, and this trendline exactly overlays with the grey trendline for the bulk of transcripts.
 The orange trendline indicates those transcripts that are both statistically significant and that have a thresholded expression level of \( \pm 1.4\), and the dark blue and cyan lines show the statistically significant trendline for \(\gamma=0.5\) , or \(1.0\).
-Note that this has the effect of changing the distribution of parts identified as significant and that the substantially fewer significantly different genes are  in the very high abundance but low dispersion category.
+
+Thus, taking Figures 1 and 2 together adding scale uncertainty has the desirable effect of changing the distribution of transcripts identified as significantly different between groups.
+Those parts that were statistically significantly different *only because of low dispersion* are preferentially excluded from statistical significance while those parts that were significantly different because of a high difference between groups remain.
 
@@
 
-## Housekeeping genes can be used to guide scale model choices.
+## Housekeeping genes and functions to guide scale model choices.
 
 Dos Santos et al.
--@dosSantos:2024  used a vaginal metatranscriptome dataset  to compare the gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
+@dosSantos:2024  used a vaginal metatranscriptome dataset  to compare the gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
 In this environment, both the relative abundance of species between groups and the gene expression level within a species is different [@macklaim:2013].
 Additionally, prior research suggests that the total number of bacteria is about 10 times more in the BV than in the H condition [@Zozaya:2010].
-Thus, this is an extremely challenging environment in which to determine differential abundance as there are both compositional and scale changes between conditions.
-The usual method to analyze vaginal metratranscriptome data is on a taxon-by-taxon basis [@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding can be ignored.
-Attempts at system-wide analysis show that many housekeeping functions are returned as differentially abundant between groups; a result likely due to a disconnect between the scale assumptions of the normalization used [@Wu2021].
+Thus, these are extremely challenging datasets in which to determine differential abundance as there are both compositional and scale changes between conditions.
+The usual method to analyze vaginal metratranscriptome data is to do so on an organism-by-organism basis [@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding of the environment is less pronounced.
+One attempt at system-wide analysis returned several housekeeping functions as differentially expressed between groups [@Denge00262-18]; a result likely due to a disconnect between the  assumptions of the normalization used and the actual scale of the environment [@Wu2021].
 
-In this example, we show how to specify a user-defined, or \emph{informed}, scale model explicitly that can account for some of these modeling difficulties.
-An informed scale model can control for both the mean difference of scale between groups (e.g., directly incorporate information on the differences in total number of bacteria between the BV and H conditions) as well as the uncertainty assumed in that difference.
-To specify a user-defined scale model, we can pass a matrix of scale values instead of a single estimate of gamma to \texttt{aldex.clr()}.
+In this example, we show how to specify and interpret an informed scale model  that can explicitly account for some of these modeling difficulties [@Nixon2024B] even in a difficult dataset.
+An informed scale model can control for both the mean difference of scale between groups (e.g., directly incorporate information on the differences in total number of bacteria between the BV and H conditions) as well as the uncertainty of that difference.
+To specify a user-defined scale model, we can pass a matrix of scale values instead of an estimate of just the scale uncertainty  to \texttt{aldex.clr()}.
 This matrix should have the same number of rows as the of Monte-Carlo Dirichlet samples, and the same number of columns as the number of samples.
@@
 This encodes the scale model as
- \(\Lambda \sim N(log2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents the actual scale value for each sample.
-This can be a measured value (cell count, nucleic acid input, etc), or an imputed value.
+ \(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents the scale value for each sample or group and gamma is the uncertainty as before.
+The scale estimate can be a measured value (cell count, nucleic acid input, etc) or an estimate.
 Nixon et al.
--@Nixon2024B showed that only the ratio between the scale values for each sample was important; see Supplementary Figures 4 and 5 for a demonstration of this point.
+-@Nixon2024B showed that only the ratio of the means are important when providing values for \( \mu_n\); i.e., the ratio between the \( \log_2 \mu_i\) and \(\log_2 \mu_j\) values.
+See the supplement to  Nixon et al.
+-@Nixon2024B for more information.
 
-Figure 3A shows an effect plot of the data where reads are grouped by function, corresponding  to grouping homologous sequences regardless of the organism of origin.
+Figure 3A shows an effect plot of the data where reads are grouped by  homologous sequence regardless of the organism of origin.
 Each point represents one of  3728 KEGG  functions [@Okuda:2008].
@@
 
-There are 101  functions with low dispersion that appear to be share by both groups (boxed area in Figure 3A, and colored in cyan).
+There are 101  functions with low dispersion that appear to be shared by both groups (boxed area in Figure 3A, and colored in cyan).
 Inspection shows that these   largely correspond to core metabolic functions such as transcription, translation, ribosomal proteins, glycolysis, replication, chaperones, etc (Supplementary file housekeeping.txt).
-The transcripts of many of these are commonly used as invariant reference sequences [@Rocha:2020aa] and so would not be expected	 to contribute to differences in ecosystem behaviour and should be centred on 0 difference.
-These should not be scored as among the most differentially abundant.
-The major group of these housekeeping functions is located off the line of no difference (being approximately located at +1.5).
-While changes in the abundance of housekeeping functions is a useful proxy for relative abundance of species in the environment, they tell us nothing about the functional capacity of the two groups as these are common to every organism.
-Of more interest is determining the functions that are different between groups because these are unique or over-expressed in one group relative to the other.
+The transcripts of many of these are commonly used as invariant reference sequences in wet lab experiments [@Rocha:2020aa] and so are not be expected	 to contribute to differences in ecosystem behaviour.
+The average location of these should be centred on 0 difference to represent an internal reference set.
+However, without an informed scale model, the mean  of these housekeeping functions is  approximately located at +2.3.
 
 ```{r ribo, echo=F}
-# list of ribosomal assocaited functions in the datset
+# list of ribosomal associated functions in the datset
 # TIL about complete.cases()
@@
 Panel A shows an effect plot for the default analysis where the functions that are elevated in  the  healthy individuals have positive values and functions that are elevated in BV have negative values.
-Highlighed in the box are KOs that are almost exlusively housekeeping functions and are  colored cyan.
+Highlighed in the box are KOs that are almost exlusively housekeeping functions; these and are  colored cyan.
 These housekeeping functions should be located on the midline of no difference.
 Panel B shows the same data scaled with \\(\\gamma = 0.5\\), which increase the minimum dispersion as before.
-Panel C shows the same data scaled with \\(\\gamma = 0.5\\) and a 0.15 fold difference in dispersion applied to the BV samples relative to the H samples.
-In these plots statistically significant (FDR < 0.01) functions in the informed model are  in red, false positive functions are in blue,  non-significant functions in black and false negative functions are in orange.
+Panel C shows the same data scaled with \\(\\gamma = 0.5\\) and a 0.14 fold difference in dispersion applied to the BV samples relative to the H samples.
+In these plots statistically significant (q-value < 0.01) functions in the informed model are  in red, false positive functions are in blue,  non-significant functions in black and false negative functions are in orange.
 " }
@@
 load('analysis/xt.m.Rda')
+load('analysis/xt.lv.Rda')
 
@@
 par(mfrow=c(1,3))
-aldex.plot(xt.all, cutoff.pval=0.01)
+aldex.plot(xt.all, cutoff.pval=0.01, xlim=c(0.3,9), xlab="Dispersion", ylab=l2d)
 points(xt.all[fp,'diff.win'], xt.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
@@
 
-aldex.plot(xg.all, xlim=c(0.3,9), cutoff.pval=0.01)
+aldex.plot(xg.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion", ylab=l2d)
 points(xg.all[fp,'diff.win'], xg.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
@@
 
-aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01)
+aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion", ylab=l2d)
 points(xt.m.all[fp,'diff.win'], xt.m.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
@@
 
-Applying the default scale model of \(\gamma=0.5\) increases the dispersion as expected but does little to move the large number of housekeeping functions toward the midline of no difference.
+Applying the default scale model of \(\gamma=0.5\) increases the dispersion slightly but does not move the   housekeeping functions toward the midline.
 This is as expected; the mean of the default scale model is based on the CLR normalization so no shift in location would be expected over the original ALDEx2 model.
 Nevertheless, about 30\% of the housekeeping functions are no longer statistically significantly different.
-Note that this change is simple to conduct, has no additional computational complexity and requires only a slight modification from the analyst.
-
-Up to this point, scale uncertainty has been applied as an extension of the CLR normalization via the default scale model, but a user-defined scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix.
-Our starting point for this is to identify the naive scale estimate from the data, and this can be done by calculating the mean scale value for each group.
-The scale estimate for samples can be accessed in the \texttt{@scaleSamps} slot from the \texttt{aldex.clr} output.
-In this datset the scale estimate for the healthy group is `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])),2)` and for the BV group is  `r round(mean(rowMeans(xg@scaleSamps[c(9:36),])),2)` for a difference of `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),])),2)`.
-This is interpreted as the scale of the H group of samples being `r round(2^(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),]))),2)` greater than the BV group and this precise but incorrect estimate places the location of the housekeeping genes off the midline of no difference.
+Note that this change is simple to conduct, has no additional computational complexity and requires only a slight modification for the analyst.
 
 We desire a scale model that approximately centres the housekeeping functions.
+Our starting point for this is to identify the naive scale estimate from the data, and this can be done by calculating the mean scale value for each group with a negligible scale value; i.e., \(\gamma=1e-3\).
+The scale estimate for samples can be accessed in the \texttt{@scaleSamps} slot from the \texttt{aldex.clr} output.
+In this datset the naive scale estimate for the healthy group is `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])),2)` and for the BV group is  `r round(mean(rowMeans(xg@scaleSamps[c(9:36),])),2)` for a difference of `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),])),2)`.
+This is interpreted as the scale of the H group of samples being `r round(2^(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),]))),2)` greater than the BV group.
+This precise but incorrect estimate places the location of the housekeeping genes off the midline of no difference.
+
 We anticipate that housekeeping functions should be nearly invariant and so an appropriate scale in this dataset is likely closer to 0 than the naive estimate.
-While a user-defined scale model can be quite flexible with relative scales that are distinct for each group (or even each sample) along with their uncertainties, here we focus on using the  \texttt{aldex.makeScaleMatrix()} function.
+One way to choose an appropriate value for \(\mu_n\) is to us the \texttt{aldex.clr} function on only the presumed invariant functions setting \(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot as before.
+Doing so suggests that the difference in scale should be about 14%.
+A second approach would be to identify the functions used as the denominator with the \texttt{denom="lvha"} option [@Wu2021] for the \texttt{aldex.clr} function, and then to use these values as before.
+This approach suggests a 5% difference in scale, and is potentially less subject to user interpretation.
+
+For the purposes of this example, if we assume a 14% difference in scale, we can set \(\mu_i = 1\) and \(\mu_j = 1.14\) using the \texttt{makeScaleMatrix} function.
 This function uses a logNormal distribution to build a scale matrix given a user-specified mean difference between groups and uncertainty level.
-Applying a per-group relative differential scale of 0.15 moves the housekeeping functions to the midline of no difference (Figure 3C), and applying a gamma of 0.5 provides the same dispersion as in panel B).
+Applying a per-group relative differential scale of 0.14 moves the housekeeping functions to the midline of no difference (Figure 3C, 14% mean difference = `r round(mean(xt.m.all[hk.off,"diff.btw"]),2)`, 5% difference = `r round(mean(xt.lv.all[hk.off,"diff.btw"]),2)`), and applying a gamma of 0.5 provides the same dispersion as in panel B of Figure 3.
 Note that now a significant number of functions are differentially up in BV that were formerly classed as not different without the full scale model (orange), or when only a default scale was applied.
 Inspection of the functions shows that these are largely missing from the \emph{Lactobacillus} species and so should actually be captured as differentially abundant in the BV group.
-Thus, applying a differential scale allows us to distinguish between both false positives (housekeeping functions in cyan, and others in blue) and false negatives (orange functions) even in a very difficult to analyze dataset.
-The remarkable improvements in biological interpretation afforded by this full scale model, and the transferrability of it between sample cohorts of the same condition is outlined elsewhere [@dosSantos:2024].
+The supplement shows that the using either the 5% or the 14%  scale difference  give  imperceptibly different results suggesting that an informed scale model does not have to precisely estimate the scale difference to be useful.
+Nixon et al, @Nixon2024B also found that multiple reasonable estimates for the \(\mu_n\) part of the informed scale model were similarly useful in microbiome data.
+
+ Thus, applying an informed scale allows us to distinguish between both false positives (housekeeping functions in cyan, and others in blue) and false negatives (orange functions) even in a very difficult to analyze dataset.
+The remarkable improvements in biological interpretation afforded by an informed scale model, and the transferrability of it between sample cohorts of the same condition is outlined elsewhere [@dosSantos:2024].
 We suggest that the default scale model  is sufficient when the data are approximately  centred.
@@
 # Discussion
-\doublespacing
-\singlespacing
-Biological systems are both predictably variable and stochastic [@Taniguchi:2010aa] and current measurement methods that rely on high throughput sequencing fail to capture all of that variation, particularly variation due to scale [@nixon2023scale;@Nixon2024B].
-In the absence of external information (such as spike-in probes [@Loven:2012aa], cell counting [@Vandeputte:2017aa], FISH [@yeast-absolute] etc) sequencing depth normalisation methods cannot recapture the scale information [@Loven:2012aa], and can only normalize for the technical variation due to sequencing depth.
 
-Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa].
-Generically, it is observed that the actual agreement between analysis methods can be modest, and  no single method outperforms all others in every dataset or type of experiment.
-That different tools appear to work more reliably in different datasets from different sources can be explained as different  normalizations being a better fit to the scale of a particular dataset by chance.
+Biological systems are both predictably variable and stochastic [@Taniguchi:2010aa] and systems biology experiments show that there are transcripts with approximately constant concentrations in the cell and those with large variability under different growth conditions [@Scott:2010].
+Current measurement methods that rely on high throughput sequencing fail to capture all of the variation, particularly variation due to scale [@nixon2023scale;@Nixon2024B].
+In the absence of external information [@Loven:2012aa;@Vandeputte:2017aa;@yeast-absolute] sequencing depth normalisation methods cannot recapture the scale information [@Loven:2012aa;@Lovell:2015], and can only normalize for the technical variation due to sequencing depth.
+Here we demonstrated that even approximate estimates of scale and the uncertainty of measuring it can aid in the interpretation of RNA-sequencing experiments.
 
-In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@Li:2022aa;@nixon2023scale], and that different analytic approaches result in different parts being chosen as significantly different [@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Nearing:2022aa].
+Nixon et al.
+[@nixon2023scale] introduced the idea of explicitly modelling the scale of a HTS dataset, and showed how to incorporate these models in the analysis of microbiome and other datasets [@Nixon2024B].
+They  demonstrated that many tools commonly used to analyze HTS datasets had substantial Type 1 and Type 2 error rates, in line with recent findings by others [@Quinn:2018aa;@Li:2022aa].
+A version of ALDEx2 with the ability to include scale uncertainty was shown to be able to correct for the high Type 1 error rate for that tool, albeit with some loss of sensitivity.
+Finally, they showed that incorporating an informed scale model incorporating both location and scale uncertainty estimates could both control for Type 1 and Type 2 error rates [@Nixon2024B].
+
+Building and using a scale model, thus has substantial benefits relative to the dual cutoff approach that is advocated for many gene expression experiments [@Cui:2003aa;@Schurch:2016aa].
+In particular, the dual cutoff approach has long been known to not control for Type 1 errors [@Zhang:2009aa;@Ebrahimpoor:2021aa], and the frequent lack of concordance between tools when benchmarked on transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa] suggests poor control of Type 2 errors as well [@Quinn:2018aa;@Li:2022aa].
+Thus, incorporating a scale model during the analysis of HTS data promises the best of both worlds.
+A default scale model can control for Type 1 errors with minimal prior knowledge of the environment and this can be done with essentially no additional computational overhead.
+Furthermore, this work and previous [@Nixon2024B] show that even minimal information about the underlying environment can be used to build a relatively robust informed scale model that controls for both Type 1 and 2 error rates.
+
+In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@Li:2022aa;@nixon2023scale].
 Li et al.
--@Li:2022aa conducted a permutation-based benchmarking study and found that popular tools performed worse than simple Wilcoxon rank-sum tests in controlling the FDR when sample sizes became large.
+[@Li:2022aa] conducted a permutation-based benchmarking study and found that widely used tools performed worse than simple Wilcoxon rank-sum tests in controlling the FDR when sample sizes became large.
 They suggested that the presence of outliers were one of the factors driving this observation.
 Brooks et al.
--@Brooks:2024aa suggested that inappropriate choice of benchmarking methods are also a major contributing factor.
-From our perspective, the disagreement between tools can be explained by the partially identifiable nature of the data; that is,  the  observed data are consistent with multiple ways of estimating the parameters [@nixon2023scale].
-Partially identifiable data exhibit multiple pathologies, chief among them being that that different analytic approaches can produce  different parameter estimates and that more data produces worse estimates because the additional data increases the precision of a flawed estimate [@gustafson2015bayesian].
+[@Brooks:2024aa] suggested that inappropriate choice of benchmarking methods are also a major contributing factor and that objective standards of truth are important.
+From the perspective of our work the disagreement between tools can be explained by the observation that different analytic approaches  produce  different parameter estimates for location or scale.
+Thus, more data produces worse estimates because the additional data simply increases the precision of a flawed estimate [@gustafson2015bayesian;@nixon2023scale].
 
-Scale simulation  is now build into ALDEx2 to address the issue of partially identifiability, and through this lens the two main root causes to common HTS data pathologies can be proposed.
-The first contributing factor is the observed very low dispersion estimate for many features that is a by-product of experimental design, sequencing and normalization.
+Scale simulation  is now build into ALDEx2 [@Nixon2024B] and this work suggest that  there are two main root causes to common HTS data pathologies.
+The first contributing factor is the observed very low dispersion estimate for many features that is a by-product of some experimental designs and of normalization.
 In the Schurch et al.
@@
 Thus, it is more accurate to describe the 96 samples as wet-lab technical replicates rather than independent samples.
-However, this type of replication approach is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.
+This type of replication approach is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.
+Applying the default scale model with \( \gamma=0.5 \) a large number of transcripts have  their dispersion increased (Figure 1D), with the effect being largest for those with the lowest initial dispersion (Figure 2).
+Adding scale uncertainty results in modest number of transcripts, `r length(sig.s.ald)`, being called significantly different as shown in the volcano plot in Figure 1B (red points).
+In addition, there is now a strong concordance between the difference and q-values.
+In hindsight, it is not obvious  why the unscaled volcano plot shows such poor correspondence.
+We suggest that this is explained by random fluctuations of the many very low variance estimates and this is supported by the  plots shown in Figure 2.
 
-In the yeast transcriptome dataset, applying the default scale model with \( \gamma=0.5 \) a large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 1D), and this results in modest number of transcripts, `r length(sig.s.ald)`, being called significantly different as shown in the volcano plot in Figure 1D (red points).
-In addition, there was now a strong concordance between the difference and p-values (Figure 1D), this is not surprising.
-In hindsight, what is not obvious is why the unscaled volcano plot shows such poor correspondence.
-We suggest that this can be explained by random fluctuations in the many variance estimates with  very low values, and this is supported by the  plot shown in Figure 2B.
-Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.
-Indeed, adding scale uncertainty reduces the significant transcripts to a subset of those identified with the dual cutoff that have the largest effect size.
-Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.
-
-While the actual scale of the underlying environment is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale.
-Varying amounts of scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
-The ALDEx2 package  contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty, and an example is shown in  Supplementary Figure 2 for the yeast transcriptome dataset.
-Here it is clear that even tiny amounts of scale preclude the majority of transcripts from being considered statistically significant.
-In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs, but further work is needed.
-
-The second contributing factor is unacknowledged asymmetry in many datasets.
+The second contributing factor is unacknowledged asymmetry in many datasets [@Wu2021]; i.e., different gene content or a directional change in the majority of features.
 In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult-to-analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
-We showed one such example for the metatranscriptome dataset in Figure 3.
-Here the dataset was highly asymmetrical, and the TMM and RLE normalizations could not fully move all the housekeeping genes to the midline of no difference or the tools exhibited other pathologies (Supplemental Figure 6).
+We showed one such example for the metatranscriptome dataset in Figure 3. Here the dataset was highly asymmetrical.
 Incorporating differential scale on a per-group basis moves the mass of the housekeeping functions towards the midline of no difference and so affects both Type I and Type II error rates.
-It is also of note that in the case of true biological replicates (different individuals) that adding a modest amount of scale \(\gamma=0.5\) had little effect on the concordance between the difference between groups and the p-values as shown in Supplementary Figure 7.
+We showed two ways of estimating the  scale difference between groups and found that any reasonable estimate is an improvement over the naive approach and also over the default scale model.
+This is in line with the observations by Nixon et al [@Nixon2024B] in a 16S rRNA gene sequencing dataset.
+It is also of note that in the case of true biological replicates (different individuals) that adding a modest amount of scale \(\gamma=0.5\) had little effect on the the difference between groups and on the dispersion.
 Thus, in this dataset the scale mis-specification was affecting mainly the location of the difference between groups.
-
-In the metatranscriptome analysis, transcripts that were previously not classed as differentially abundant were now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such.
-Note  the assumption that housekeeping genes should not generally be included in differential abundance analysis is implicit in the dual p-value:fold-change cutoff approach in widespread use.
 While we acknowledge that some prior information on which housekeeping transcripts should not be classed as differentially abundant is needed, we suggest that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
-Thus, the use of this prior knowledge is not unique to our approach.
 
@@
 Thus, an advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
-This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models can make this an explicit part of the model building process.
 While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions and a detailed description of these points is describe elsewhere [@nixon2023scale].
-In this work, we simply use the parameter \(\gamma\) in the default scale model as a term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption.
-In the metatranscriptome example and elsewhere [@Nixon2024B], we also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.
 
-The ALDEx2 R package is readily amenable for incorporating scale uncertainty.
-Originally, this tool used the only the Dirichlet distribution to sample compositional uncertainty, but scale uncertainty can be added through the use of a scale model with no additional computational complexity.
-By default, ALDEx2 samples scale from a logNormal distribution inspired by the CLR normalization. 
-However, scale uncertainty can be sampled from any distribution depending on prior knowledge or preference, using the option to introduce a full informed scale model that  encapsulates both uncertainty and asymmetry in underlying scale.
+In summary, we supply a toolkit that makes incorporating scale uncertainty and location information simple to incorporate for transcriptomes or indeed any type of HTS dataset.
+While the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and can help explain some of the underlying biology, and help to expose known issues with the analysis of HTS data.
+Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on conclusions weighted accordingly.
+Additionally, the use of informed  scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups [@dosSantos:2024].
+Thus, using and reporting scale uncertainty should become a standard practice in the analysis of HTS datasets.
 
-In summary, we supply a toolkit that makes incorporating scale simple to incorporate for any type of HTS dataset.
-While the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and can help explain some of the underlying biology, and known issues with the analysis of HTS data.
-Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on conclusions weighted accordingly.
-Additionally, the use of user-defined  scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups [@dosSantos:2024].
-Thus, reporting scale uncertainty should become a standard practice in the analysis of HTS datasets as a way to identify which features are most robust to differences in the underlying system.
+\singlespacing
 

### 2024-08-22 go3.tex
with Justin's suggestions
Send back to Justin for comment

@@
 
-\title{Normalizations are not what you think; Explicit Scale Simulation
-in ALDEx2}
+\title{Explicit Scale Simulation for analysis of RNA-sequencing with
+ALDEx2}
 \author{true \and true \and true}

### 2024-08-22 code/meta-aldex.R

@@ -7,10 +7,26 @@ load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
 # make a vector of conditions
 conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8)) 
 
-xt <- aldex.clr(ko.both, conds.K0)
+# practically is identical to 0, and fills scale info slot
+xt <- aldex.clr(ko.both, conds.K0, gamma=1e-3)
 xt.e <- aldex.effect(xt, include.sample.summary=T)
 xt.t <- aldex.ttest(xt)
 xt.all <- cbind(xt.e, xt.t)
+
+##### find scale estimate for hk functions
+# commented out since just need values
+# hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3
+# xhk <- aldex.clr(ko.both[hk.off,], conds.K0, gamma=1e-3)
+# round(2^(mean(xhk@scaleSamps[c(1:8,37:44),]) -  mean(xhk@scaleSamps[c(9:36),])),2) # 1.14
+# 
+# x.lvha <- aldex.clr(ko.both, conds.K0, denom="lvha")
+# denom <- x.lvha@denom
+# 
+# xd <- aldex.clr(ko.both[denom,], conds.K0, gamma=1e-3)
+# round(2^(mean(xd@scaleSamps[c(1:8,37:44),]) -  mean(xd@scaleSamps[c(9:36),])),2) # 1.05
+
+#####
+
 
 save(xt.all, file='analysis/xt.Rda')
 
@@ -20,21 +36,31 @@ xg.e <- aldex.effect(xg)
 xg.t <- aldex.ttest(xg)
 xg.all <- cbind(xg.e, xg.t)
 
+# can get the mean scale values
+#  mean( xt@scaleSamps[c(1:8,37:44),]) #H
+#  mean( xt@scaleSamps[c(9:36),]) #BV
 for(i in 1:length(xt@dirichletData)){xt@dirichletData[[1]] <- NULL}
 save(xt, file='analysis/xt.clr.Rda')
 for(i in 1:length(xg@dirichletData)){xg@dirichletData[[1]] <- NULL}
 save(xg, file='analysis/xg.clr.Rda')
 
 save(xg.all, file='analysis/xg.Rda')
-mu <- c(1,1.15) # 15% difference in 
[…]

### 2025-03-11 go3.Rmd
submitted version Dec 6

@@
   - name: Michelle Pistner Nixon
-    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
+    affiliation: Department of Population Health Sciences, Geisinger, Danville, PA 
   - name: Justin Silverman
-    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
+    affiliation: 
+      College of Information Sciences and Technology and Department of Medicine, Pennsylvania   State University
+      Department of Medicine, Pennsylvania State University
+      Institute for Computational and Data Science, Pennsylvania State University
+      Department of Statistics, Pennsylvania State University
+cite-method: citeproc
 citeproc: TRUE
-cite-method: citeproc
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
 csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
-pdf-engine: latex
 output:
-  pdf_document 
+    pdf_document:
+      keep_tex: true
 header-includes:
@@
 package: ALDEx2
-abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total mRNA expression) of the underlying biological systems.
-Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses.
+
+abstract: 
+  In high-throughput sequencing (HTS) studies,  sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system.
+Typically a statistical normalization removes unwanted technical variation in the data or the parameters of the model to enable differential abundance analyses.
 We recently showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can dramatically increase false positive and false negative rates.
-We demonstrated that these errors can be mitigated by accounting for uncertainty about scale using a *scale model*, which we integrated into the ALDEx2 R package.
-This article provides new insights into those methods, focusing on the application to transcriptomic analysis.
-Here we provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses.
-We show that these scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance.
-We demonstrate the utility of that scale models   built based on known housekeeping genes in complex metatranscriptomic datasets.
-Thus this work provides example and practical guidance on how to incorporate scale into transcriptomic analysis.
+We demonstrated that these errors can be mitigated by accounting for uncertainty using a *scale model*, which we integrated into the ALDEx2 R package.
+This article provides new insights focusing on the application to transcriptomic analysis.
+We provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses.
+These scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance.
+We demonstrate the utility of scale models built based on known housekeeping genes in complex metatranscriptomic datasets.
+Thus this work provides guidance on how to incorporate scale into transcriptomic data sets.
 
@@
 \doublespacing 
+\singlespacing
 
-High-throughput sequencing (HTS) is a ubiquitous tool used to explore
-many biological phenomenon such as gene expression (single-cell
-sequencing, RNA-sequencing, meta-transcriptomics), microbial community
-composition (16S rRNA gene sequencing, shotgun metagenomics) and
-differential enzyme activity (selex, CRISPR killing). HTS proceeds by
-taking a sample from the environment, making a library, multiplexing
-(merging) multiple libraries together, and then applying a sample of the
-multiplexed library to a flow cell. Each of these steps is a 
-compositional sampling step as only a fixed-size subsample of nucleic acid is 
-carried over to subsequent steps. 
-Thus, with
-each sampling step the connection between the size of the sampled DNA
-pool and the scale (e.g., size, microbial load, or total gene
-expression) of the measured biological system is degraded or lost.
-In the end, the information contained in the data relates only to relative abundances and  has an arbitrary scale imposed by the sequencing process[@lovell:2011;@Quinn:2019aa;@nixon2023scale].
-Increasingly, researchers are turning to modified experimental protocols
- in an attempt to recover the lost
-biological variation in scale; see for example  [@Vandeputte:2017aa;@Props:2017aa]. 
-However, these protocols only recover scale variation at
-the intermediate step in the sample preparation protocol where the intervention was made. 
-In short, the disconnect
-between sample-to-sample variation in sequencing depth and biological
-variation in scale remains an open challenge.
+High-throughput sequencing (HTS) is a ubiquitous
+tool used to explore many biological phenomenon such as gene expression
+(single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial
+community composition (16S rRNA gene sequencing, shotgun metagenomics) and
+differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a
+sample from the environment, making a library, multiplexing (merging) multiple
+libraries together, and then applying a sample of the multiplexed library to a
+flow cell. Each of these steps is a compositional sampling step as only a
+fixed-size subsample of nucleic acid is carried over to subsequent steps. Thus,
+with each sampling step the connection between the actual size of the sampled
+DNA pool and the scale (e.g., size, microbial load, or total gene expression) of
+the measured biological system is degraded or lost. In the end, the information
+contained in the data relates only to relative abundances and  has an arbitrary
+scale imposed by the sequencing
+process [@lovell:2011;@Quinn:2019aa;@nixon2024scale]. Researchers
+can use modified experimental protocols [@Jiang:2011aa;@Loven:2012aa;@Vandeputte:2017aa;@Props:2017aa] or machine learning methods [@Nishijima:2024aa] to uncover the
+biological variation in scale. However,  wet-lab protocols only provide information on
+the size of the data downstream of the step in the sample preparation protocol where the
+intervention was made and  introduce an additional source of variation that must be accounted for [@Loven:2012aa].
+The computational methods developed for microbiome analysis have low correlation with the actual scale but are useful [@Nishijima:2024aa].
+In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an open
+challenge.
 
 The analysis of  HTS data suffers from several known problems that can be
-traced, in whole or in part, to  misspecification of scale. The first issue 
-is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Li:2022aa], 
-  exhibited as dataset-dependent 
-FDR control and by the disconnect between statistical and biological significance[@Schurch:2016aa].
-In current practice,  these issues are addressed  by a dual-filtering method, whereby both a low p-value (or equivalently a low q-value following FDR correction [@storey2003positive]) and  a large difference between groups is used to identify interesting transcripts or genes for follow-up analysis[@Schurch:2016aa].
-This double-filtering approach is graphically exemplified by the volcano plot [@Cui:2003aa], but
- is known to not   appropriately control the  FDR  
-[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
-analyzing data where the mean change between groups is non-zero.
-Such asymmetric data can arise when a gene set is expressed in one group but not the other, or when one group contains different gene content from the other.
-This type of data frequently arises in in-vitro selection experiments (SELEX), 
-transcriptome analysis, and microbiome analysis [@Wu2021]. 
-The third issue is that these
-problems become more pronounced as more samples are collected; that is,
-more information results in a worsening of the accuracy of the analysis [@Li:2022aa;@nixon2023scale,@Nixon2024B].
+traced, in whole or in part, to  misspecification of scale. The first issue is
+poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Ge:2021aa;@Li:2022aa], exhibited
+as dataset-dependent FDR control and by the disconnect between statistical and
+biological significance [@Schurch:2016aa]. In current practice,  these issues are
+addressed  by a dual-filtering method, whereby both a low p-value (or
+equivalently a low q-value following FDR correction [@storey2003positive]) and a
+large difference between groups is used to identify interesting transcripts or
+genes for follow-up analysis [@Cui:2003aa;@Schurch:2016aa]. This double-filtering approach is
+graphically exemplified by the volcano plot [@Cui:2003aa], but is known to not
+appropriately control the  FDR [@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second
+issue is poor performance when analyzing data where the mean change between
+groups is non-zero [@nixon2024scale]. Such asymmetric data can arise when a gene set is expressed
+in one group but not the other, or when one group contains different gene
+content from the other. This type of data frequently arises in in-vitro
+selection experiments (SELEX), transcriptome analysis, and microbiome analysis
+[@Wu2021]. The third issue is that the actual scale of the environment is often
+a major confounding variable during analysis [@nixon2024scale;@Nishijima:2024aa]. The final issue is that these problems become more pronounced as more
+samples are collected; that is, more information results in a worsening of the
+accuracy of the analysis [@Li:2022aa;@nixon2024scale;@Nixon2024B].
 
-The  three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.
-Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Lovell:2015].
-For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa].
-In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa].
-As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010].
-Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).
-Current methods access only the compositional information yet make implicit assumptions about the scale [@nixon2023scale,@Nixon2024B].
+The  four problems were recently shown by Nixon et al. [@nixon2024scale]
+to be a result of a mismatch between the underlying size or scale of the system
+and the assumptions of the normalizations used for the analysis of HTS.
+Biological variation in scale often represents an important unmeasured
+confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the
+cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA
+content than  non-transformed cells [@Nie:2012aa], and this dramatically skews
+transcriptome analysis unless spike-in approaches are used [@Loven:2012aa]. In addition, wild-type and mutant
+strains of cell lines, yeast or bacteria  have different growth rates and RNA
+contents under different conditions, which  affect our ability to identify truly
+differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As
+another example, the total bacterial load of the vaginal microbiome differs by
+1-2 orders of magnitude in absolute abundance between the healthy and bacterial
+vaginosis states [@Zozaya:2010], and the composition between these states is
+dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of
+any of these systems includes both relative change (composition) and absolute
+abundance (scale).  Current methods access only the compositional information
+yet make implicit assumptions about the scale [@nixon2024scale,@Nixon2024B].
 
-Recently, Nixon et al.
-[-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models.
-They showed that \emph{all} normalizations make some assumption about  scale but these implicit assumptions are often inappropriate and difficult to interpret.
-Because of this different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
-Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g.
-proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g.
-the TMM [@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g.
-the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@Aitchison:1986] and its derivatives.
+Recently, Nixon et al. [-@nixon2024scale] showed that the challenge of
+non-biological variation in sequencing depth be viewed as a problem of
+partially-identified models. They showed that \emph{all} normalizations make
+some assumption about  scale but these implicit assumptions are often
+inappropriate and difficult to interpret. This causes different normalizations
+to provide different outputs when applied to the same dataset
+[@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
+Intuitively, normalizations in widespread use  assume that either all samples
+have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM
+[@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample
+can be chosen as a reference to which the others are scaled e.g. the TMM
+[@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982];
+or that different sub-parts of each sample maintain a constant scale across
+samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is
+appropriate e.g the CLR [@Aitchison:1986] and its derivatives.
 
-The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale].
-While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa].
-Nixon et al.
-[@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results.
-We recently modified ALDEx2  to explicitly model the scale over a range of reasonable normalization parameters, and showed significant improvements in performance in microbiome and in-vitro selection experiments [@Nixon2024B].
-Here, we briefly introduce review modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.
+The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict
+assumption about scale through the CLR normalization [@nixon2024scale].
+This assumption was often close enough to the true value to be useful, but was not always the a good estimate and could be outperformed by other normalizations [@Yerke:2024aa].
+Nixon et al. [@nixon2024scale] showed that better scale assumptions resulted in
+more reproducible data analysis including better control of both false positive
+and false negative results. We recently modified ALDEx2  to explicitly model the
+scale over a range of reasonable normalization parameters, and showed
+significant improvements in performance in microbiome and in-vitro selection
+experiments [@Nixon2024B]. Here, we briefly  review these modifications and show how scale uncertainty can greatly improve modeling in transcriptome
+and meta-transcriptome datasets to provide more robust and reproducible results.
 
@@
 
-Formal and expanded descriptions of the concepts that follow are given in [@nixon2023scale;@Nixon2024B].
-To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).
-Likewise, we can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained.
-We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions).
-That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
-Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).
+Formal and expanded descriptions of the concepts that follow are given in
+[@nixon2024scale;@Nixon2024B].   To be  concrete, we let \(\mathbf{Y}\) denote
+the \emph{measured} \(D \times N\) matrix of sequence counts with elements
+\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to
+feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).  Likewise, we
+can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the
+biological system from which sample \(n\) was obtained. We can think of
+\(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\)
+(e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e.,
+proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
+\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
+\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
+\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note
+that with these definitions \(\mathbf{W}\) can be written as the element-wise
+combination of scale and composition:
+\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the
+logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
+\mathbf{W}^{\perp}_{n}\).
 
-Many of  the normalizations in widespread use in tools such as DESeq2 [@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2 [@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation.
-We use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate that the output is an estimate of the true value.
-The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al.
-2011).
+Many of  the normalizations in widespread use in tools such as DESeq2
+[@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2
+[@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}
+\approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by
+some function of the observation. We use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\
+\)) notation to indicate that the output is an estimate of the true value.  The
+technical variation in sequencing depth
+(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data
+\(\mathbf{Y}\) provides us with information about the system composition
+\(\mathbf{W}^{\parallel}\) but little to no information in the system scale
+\(\mathbf{W}^{\perp}\) (Lovell et al. 2011).
 
@@
 
-The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox to  model the uncertainty of HTS data and to use that model to estimate LFC significance.
-At a high-level, ALDEx2  has three connected components to estimate the uncertainty inherent in HTS datasets.
-First, the tool accounts for the uncertainty of the sequencing counts using Dirichlet multinomial sampling to build a probabilistic model of the data; i.e., \(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
-Secondly, ALDEx2 uses the centred log-ratio transformation to scale the data [@fernandes:2013].
-However, this step was modified recently to account for scale uncertainty and misspecification [@nixon2023scale,@Nixon2024B] via a scale model, explained  with more details in [@nixon2023scale;@Nixon2024B] and summarized in the next paragraph.
-Finally, a standard null-hypothesis test and a non-parametric estimate of mean standardized difference are used to report on the finite sample variation.
-These sources of uncertainty and variation are combined via reporting the expected values from a Monte-Carlo simulation framework.
-For simplicity, we use the term 'difference' to refer to the absolute difference between groups, and 'dispersion' to refer to the  within-condition difference or pooled variance as defined in [@fernandes:2013].
-These are calculated on a \(\log_2\) scale.
-For more details on ALDEx2 see [@fernandes:2013;@fernandes:2014;@nixon2023scale;@Nixon2024B].
+The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose
+toolbox to  model the uncertainty of HTS data and to use that model to estimate
+the underlying LFC (log-fold change) significance.  At a high-level, ALDEx2  has
+three connected components to estimate the uncertainty inherent in HTS datasets.
+First, the tool accounts for the uncertainty of the sequencing counts using
+Dirichlet multinomial sampling to build a probabilistic model of the data; i.e.,
+\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).  Secondly,
+ALDEx2 uses the centred log-ratio transformation to scale the data
+[@fernandes:2013]. However, this step was modified recently to account for scale
+uncertainty and misspecification [@nixon2024scale,@Nixon2024B] via a scale
+model, explained  with more details in [@nixon2024scale;@Nixon2024B] and
+summarized in the next paragraph. Finally, a standard null-hypothesis test and a
+non-parametric estimate of mean standardized difference are used to report on
+the finite sample variation. These sources of uncertainty and variation are
+combined via reporting the expected values from a Monte-Carlo simulation
+framework. For simplicity, we use the term 'difference' to refer to the absolute
+difference between groups, and 'dispersion' to refer to the  within-condition
+difference or pooled variance as defined in [@fernandes:2013]. These are
+calculated on a \(\log_2\) scale. For more details on ALDEx2 see
+[@fernandes:2013;@fernandes:2014;@nixon2024scale;@Nixon2024B].
 
-Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which Nixon et al.
-[@nixon2023scale] term  a \textit{Scale Simulation Random   Variable} (SSRV).
-To do this, Nixon et al.
-[@nixon2023scale] generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error in the centred log-ratio normalization step.
-They did this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\).
-The CLR method used by ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the geometric mean of sample n, which while being a random variable, is essentially constant across each Monte-Carlo replicate.
-With this modificaiton, ALDEx2 can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\).
-For example, the following scale model generalizes the CLR:
+Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a
+specialized type of statistical model which Nixon et al. [@nixon2024scale] term
+a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon et al.
+[@nixon2024scale] generalized the concept of normalizations by introducing  the
+concept of a \textit{scale model} to account for potential error in the centred
+log-ratio normalization step.  They did this by including a model for
+\(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by ALDEx2 makes the
+assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the
+geometric mean of sample n, which while being a random variable, is essentially
+constant across each Monte-Carlo replicate, but that differs between samples.
+With this modification, ALDEx2 can be generalized by considering probability
+models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean
+\(1/G_{n}\). For example, the following scale model generalizes the CLR:
 
-\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2})\].
+\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda
+\sim N(\mu, \gamma^{2})\].
 
-This formulation is quite flexible [@nixon2023scale;@Nixon2024B].
-In the simple or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter drawn from a log-Normal distribution that adds scale uncertainty.
-The default approach controls only the degree of uncertainty in the CLR assumption for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control).
-In the advanced or 'informed' configuration, \(\mu\) takes different values for each group, and this controls the location of the LFC assumption.
-An example of both the default and informed approaches is given for a microbiome dataset in @Nixon2024B, and here we show that these approaches also work well in transcriptome and metatranscriptome datasets.
-These modifications are  instantiated in ALDEx2 which is  the first software package designed for SSRV-based inference.
+This formulation is quite flexible [@nixon2024scale;@Nixon2024B]. In the simple
+or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter
+drawn from a log-Normal distribution[@nixon2024scale]. Adding scale uncertainty with the \(\gamma\) paramenter  controls only the degree of uncertainty of the CLR assumption for the
+\(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and
+\(x_{n}=0\) denotes control). In the advanced or 'informed' configuration,
+\(\mu\) takes different values for each group and  controls the location of
+the LFC assumption; combining \(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the location and the scale. An example of both the default and informed approaches is
+given for a microbiome dataset in [@Nixon2024B] showing increased sensitivity and specificity. Here we show that these
+approaches also work well in transcriptome and metatranscriptome datasets. These
+modifications are  instantiated in ALDEx2 which is  the first software package
+designed for SSRV-based inference.
 
@@
 
-Gierliński et al.
-[@Gierlinski:2015aa]  generated a  highly replicated yeast transcriptome dataset to compare gene expression between a wild-type strain and a snf2 gene knockout, \(\Delta\)snf2.
-This dataset was  used to test several RNA-seq tools for their power  to detect the set of differentially abundant transcripts identified in the full dataset when the data was subset [@Schurch:2016aa].
-In this  study each tool had its own 'gold standard' set of transcripts with different tools identifying between between 65% to >80% of all transcripts as being significantly different.
-Since the majority of transcripts were significantly different, the authors suggested that it was more appropriate to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995] corrected p-value (q-value) plus a difference cutoff  to limit the number of identified transcripts to a much smaller fraction of the total.
-Nixon et al, @Nixon2024B showed that adding even a small amount of  scale uncertainty with ALDEx2  dramatically reduced the number of significant transcripts identified, removing the need for the dual cutoff approach in this dataset and others.
-Below we include an intuitive explanation of how incorporating scale uncertainty achieved this outcome using a setting of \(\gamma=0.5\), which is in line with the recommendations of [@Nixon2024B] and that gives results comparable to those proposed in [@Schurch:2016aa].
+Gierliński et al. [@Gierlinski:2015aa]  generated a  highly replicated yeast
+transcriptome dataset to compare gene expression between a wild-type strain and
+a snf2 gene knockout, \(\Delta\)snf2. This dataset was  used to test several
+RNA-seq tools for their power  to detect the set of differentially abundant
+transcripts identified in the full dataset when the data was subset
+[@Schurch:2016aa]. In this  study each tool had its own 'gold standard' set of
+transcripts with different tools identifying between between 65% to >80% of all
+transcripts as being significantly different. Since the majority of transcripts
+were significantly different, the authors suggested that it was more appropriate
+to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995]
+corrected p-value (q-value) plus a difference cutoff  to limit the number of
+identified transcripts to a much smaller fraction of the total.  Nixon et al,
+@Nixon2024B showed that adding even a small amount of  scale uncertainty with
+ALDEx2  dramatically reduced the number of significant transcripts identified,
+removing the need for the dual cutoff approach in this dataset and others. Below
+we include an intuitive explanation of why and how incorporating scale uncertainty
+achieved this outcome using a setting of \(\gamma=0.5\). The approach is in line with
+the recommendations of [@Nixon2024B] and  gives results comparable to those
+proposed in [@Schurch:2016aa].
 
@@
 sig.all <- yst.all$we.eBH < 0.05 
-sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
+sig.t <- yst.all$we.eBH < 0.05 &
+abs(yst.all$diff.btw) > 1.4 
 sig.s <- yst.s.all$we.eBH < 0.05 
@@
 
-We start with the assumption that not all statistically significant differences are biologically relevant [@efron2008FDR], and that such a large number of significantly different transcripts breaks the  necessary assumption for DA/DE expression that most parts be invariant [@Dillies:2013].
-As noted,  transcriptomics commonly uses a dual cutoff approach that is graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa].
-Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a  q-value cutoff of \( \le 0.05\); i.e.
-`r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or `r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2) of the `r nrow(yst.all)` transcripts.
-These values are in line with those observed by [@Schurch:2016aa].
-Such  large numbers of statistically significant transcripts seems biologically unrealistic.
-That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some, if not the majority, are driven by technical differences in the analysis [@Dillies:2013;@Li:2022aa].
+We start with the assumption that not all statistically significant differences
+are biologically relevant [@efron2008FDR], and that such a large number of
+significantly different transcripts breaks the  necessary assumption for DA/DE
+expression that most parts be invariant [@Dillies:2013]. As noted,
+transcriptomics commonly uses a dual cutoff approach that is graphically
+exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa]. Using either DESeq2
+or ALDEx2, a majority of transcripts are statistically significantly different
+between groups  with a  q-value cutoff of \( \le 0.05\); i.e. `r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or
+`r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2)
+of the `r nrow(yst.all)` transcripts. These values are in line with those
+observed by [@Schurch:2016aa]. Such  large numbers of statistically significant
+transcripts seems biologically unrealistic. That `r length(setdiff(sig.ald,
+sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
+identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does
+not, suggests that the choice of normalization plays a role in which results are
+returned as significant and that some, if not the majority, are driven by
+technical differences in the analysis [@Dillies:2013;@Li:2022aa].
 
-```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis.
+```{r plot1, echo=FALSE, warning=FALSE, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis.
 ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset.
@@
 par(mfrow=c(2,2))
-# volcano
-#plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
+# volcano plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300),
 # col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
-#title('D: DESeq2 volcano', adj=0, line= 0.8)
-#points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
-#  pch=19, cex=0.5)
-#points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
-#  pch=19, cex=0.5)
-#  abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+# title('D: DESeq2 volcano', adj=0, line= 0.8)
+# points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des
+# ] + 1e-300), col=rgb(1,.66,0,0.3), pch=19, cex=0.5)
+# points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s
+# .ald] + 1e-300), col=rgb(1,0,0,1), pch=19, cex=0.5) abline(v=c(-1.4,1.4),
+# lty=2,lwd=2, col='grey')
 
@@
 
-plot(yst.all$diff.btw, -1*log10(yst.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
-   xlab=l2d, ylab=l10q)
+plot(yst.all$diff.btw, -1*log10(yst.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), xlab=l2d, ylab=l10q) 
 title(expression(paste("A: Volcano: ", gamma, " = 0")), adj=0, line= 0.8) 
-points(yst.all$diff.btw[sig.ald], -1*log10(yst.all$we.eBH[sig.ald]	+1e-70), 
-  col=rgb(1,.66,0,.3), cex=0.5, pch=19)
+points(yst.all$diff.btw[sig.ald],
+-1*log10(yst.all$we.eBH[sig.ald]  +1e-70), col=rgb(1,.66,0,.3), cex=0.5, pch=19)
 points(yst.all$diff.btw[sig.s.ald], -1*log10(yst.all$we.eBH[sig.s.ald]  +1e-70),
@@
 title(expression(paste("B: Volcano: ", gamma, " = 0.5")), adj=0, line= 0.8) 
-points(yst.s.all$diff.btw[sig.s.ald], -1*log10(yst.s.all$we.eBH[sig.s.ald]+1e-15),
-  col=rgb(1,0,0,1), cex=0.5, pch=19)
+points(yst.s.all$diff.btw[sig.s.ald], -1*log10(yst.s.all$we.eBH[sig.s.ald]+1e-15), col=rgb(1,0,0,1), cex=0.5, pch=19)
 abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
-#  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
+#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange,
+#xlim=c(0,5), col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
 #title('A: DESeq2 effect', adj=0, line= 0.8)
 #points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
-#  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
-#  pch=19, cex=0.5)
+#res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), pch=19, cex=0.5)
 #points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
-#  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
-#  pch=19, cex=0.5)
+#res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), pch=19, cex=0.5)
 #abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), 
-   xlab='Dispersion', ylab=l2d)
-title(expression(paste("C: Effect: ", gamma, " = 0")), adj=0, line= 0.8)
+plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), xlab='Dispersion',
+ylab=l2d) 
+title(expression(paste("C: Effect: ", gamma, " = 0")), adj=0, line=
+0.8) 
 points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald],
 col=rgb(1,.66,0,0.3), cex=0.5, pch=19) 
-	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+abline(h=c(-1.4,1.4), lty=2,lwd=2,
+col='grey') 
 points(yst.all$diff.win[sig.s.ald], yst.all$diff.btw[sig.s.ald],
 col=rgb(1,0,0,0.6), cex=0.5, pch=19) 
-	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+abline(h=c(-1.4,1.4), lty=2,lwd=2,
+col='grey')
 
@@
 xlab='Dispersion', ylab=l2d, xlim=c(0.1,5)) 
-title(expression(paste("D: Effect: ", gamma, " = 0.5")), adj=0, line= 0.8)
-points(yst.s.all$diff.win[sig.s.ald], yst.s.all$diff.btw[sig.s.ald], 
-  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
+title(expression(paste("D: Effect:
+", gamma, " = 0.5")), adj=0, line= 0.8) 
+points(yst.s.all$diff.win[sig.s.ald],
+yst.s.all$diff.btw[sig.s.ald], col=rgb(1,0,0,0.5), cex=0.5, pch=19)
 abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
@@
 
-The Volcano plots in Figure 1 A and B show  that adding scale uncertainty increases the minimum q-value and increases the concordance between the q-value and the difference between groups (compare panels A and B).
-The effect plots  [@gloor:effect] in Figure 1C  shows that the majority of significant transcripts (red, orange) have negligible differences between groups and  very low dispersion.
-We suggest that this low dispersion is driven by the experimental design which is actually a technical wet lab replication  rather than a true biological replication design [@Gierlinski:2015aa].
-Scale uncertainty can be incorporated using the \texttt{gamma} parameter that controls the amount of noise added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
-Figure 1 B,D shows that  setting  \( \gamma=0.5 \) results in `r length(sig.s.ald)` which is far fewer significant transcripts than in the naive analysis  and we observe that  the minimum dispersion increases from  `r round(min(yst.all$diff.win), 2)` (\(\gamma = 0\) ) to  `r round(min(yst.s.all$diff.win), 2)` (\(\gamma=0.5\)).
+The Volcano plots in Figure 1 A and B show  that adding scale uncertainty
+increases the minimum q-value and increases the concordance between the q-value
+and the difference between groups (compare panels A and B). The effect plots
+[@gloor:effect] in Figure 1C  shows that the majority of significant transcripts
+(red, orange) have negligible differences between groups and  very low
+dispersion. We suggest that this low dispersion is driven by the experimental
+design which is actually a technical wet lab replication  rather than a true
+biological replication design [@Gierlinski:2015aa]. Scale uncertainty can be
+incorporated using the \texttt{gamma} parameter that controls the amount of
+noise added to the CLR mean assumption when we call either \texttt{aldex()}, or
+\texttt{aldex.clr()}. Figure 1 B,D shows that  setting  \( \gamma=0.5 \) results
+in `r length(sig.s.ald)` which is far fewer significant transcripts than in the
+naive analysis  and we observe that  the minimum dispersion increases from  `r round(min(yst.all$diff.win), 2)` (\(\gamma = 0\) ) to  `r round(min(yst.s.all$diff.win), 2)` (\(\gamma=0.5\)).
 
@@
 xlab='rAbundance', ylab='Dispersion') 
-lines(lowess(no.scale, f=0.1), col='grey', lwd=3)
-points(yst.s.all$rab.all, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+lines(lowess(no.scale, f=0.1), col='grey',
+lwd=3) 
+points(yst.s.all$rab.all, yst.s.all$diff.win, pch=19, cex=0.5,
+col=rgb(1,0,0,0.3)) 
 lines(lowess(half.scale, f=0.1), col='pink', lwd=3)
-points(yst.1.all$rab.all, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3))
+points(yst.1.all$rab.all, yst.1.all$diff.win, pch=19, cex=0.5,
+col=rgb(0,0,1,0.3)) 
 lines(lowess(full.scale, f=0.1), col='royalblue1', lwd=3)
@@
 abline(h=mn.mid.s, lty=2, col='red') 
-abline(h=mn.mid.1, lty=2, col='blue')
+abline(h=mn.mid.1,
+lty=2, col='blue') 
 title(main='A', line=-1.2, adj=0.9)
@@
 g1 <- expression(paste(gamma, " = 1")) 
-legend(1,4, legend=c(g0, g5, g1), 
-  pch=19, col=c("black", "red", "blue"))
+legend(1,4, legend=c(g0, g5, g1), pch=19, col=c("black", "red", "blue"))
 
-# plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3), ylim=c(0,4.5), 
-#   xlab='g=0 Dispersion', ylab='g=0.5 | g=1 Dispersion')
-# points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
-# abline(0,1, lty=2)
-# title(main='B', line=-1.2, adj=0.5)
-# legend(0,4, legend=c( "0.5", "1"), 
-#   pch=19, col=c("red", "blue"))
+# plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5,
+# col=rgb(0,0,1,0.3), ylim=c(0,4.5), xlab='g=0 Dispersion', ylab='g=0.5 | g=1
+# Dispersion') points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5,
+# col=rgb(1,0,0,0.3)) abline(0,1, lty=2) title(main='B', line=-1.2, adj=0.5)
+# legend(0,4, legend=c( "0.5", "1"), pch=19, col=c("red", "blue"))
 
-diff.df <- data.frame(yst.all$rab.all, 
-  yst.1.all$diff.win- yst.all$diff.win)
-sig.df <- data.frame(yst.all$rab.all[sig.all],
-   yst.1.all$diff.win[sig.all]- yst.all$diff.win[sig.all])
+diff.df <- data.frame(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win)
+sig.df <- data.frame(yst.all$rab.all[sig.all], yst.1.all$diff.win[sig.all]-
+yst.all$diff.win[sig.all]) 
 sig.t.df <- data.frame(yst.all$rab.al[sig.t],
 yst.1.all$diff.win[sig.t] - yst.all$diff.win[sig.t]) 
-sig.s.df <- data.frame(yst.all$rab.all[sig.s], 
-  yst.1.all$diff.win[sig.s] - yst.all$diff.win[sig.s])
+sig.s.df <-
+data.frame(yst.all$rab.all[sig.s], yst.1.all$diff.win[sig.s] -
+yst.all$diff.win[sig.s]) 
 sig.1.df <- data.frame(yst.all$rab.all[sig.1],
@@
 ylab=ylab1, pch=19, cex=0.5, xlim=c(-11,10), ylim=c(.1,1.5)) 
-title(main='B', line=-1.2, adj=0.9)
+title(main='B',
+line=-1.2, adj=0.9) 
 lines(lowess(diff.df, f=0.1), col='grey60', lwd=3)
 lines(lowess(sig.df, f=0.1), col='red', lwd=3) 
-lines(lowess(sig.t.df, f=0.5), col='orange', lwd=3)
+lines(lowess(sig.t.df, f=0.5),
+col='orange', lwd=3) 
 lines(lowess(sig.s.df, f=0.5), col='royalblue1', lwd=3)
@@
 
-legend(-12,1.6, legend=c( d0, dstar, dstarT,dg5, dg1), 
-  pch="-", col=c("grey", "red", "orange", "royalblue1", "cyan"))
+legend(-12,1.6, legend=c( d0, dstar, dstarT,dg5, dg1), pch="-", col=c("grey",
+"red", "orange", "royalblue1", "cyan"))
 
@@
 
-It is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both q-values and fold-changes  [@Cui:2003aa;@Schurch:2016aa].
-Note that there is considerable variation in recommended cutoff values[@Schurch:2016aa].
-Here, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
-This cutoff was chosen for convenience and is in-line with the recommendations of [@Schurch:2016aa].
-the fold change limits are shown by the dashed grey lines in Figure 1.
-The  \(2^{1.4}\) fold change cutoff identifies a similar number of transcripts as does  ALDEx2 using \( \gamma = 0.5 \) which identifies `r length(sig.s.ald)`.
-Supplementary Figure 2 shows how to use the \texttt{aldex.senAnalysis()} function to identify those transcripts that are very sensitive to scale uncertainty.
-In this supplementary figure we see that even adding a very small amount of scale \( \gamma = 0.1 \) reduces the number of significant transcripts by more than half.
-This allows us to those low-dispertion transcripts that were  significant only because of an absence of scale uncertainty.
-In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs [@Nixon2024B].
+It is common practice to use a dual-cutoff by choosing transcripts based on a
+thresholds for both q-values and fold-changes  [@Cui:2003aa;@Schurch:2016aa].
+Note that there is considerable variation in recommended cutoff
+values[@Schurch:2016aa]. Here, applying a dual-cutoff using a heuristic of at
+least a \(2^{1.4}\) fold change  reduces the number of significant outputs to
+193 for DESeq2 and to 186 for ALDEx2. This cutoff was chosen for convenience and
+is in-line with the recommendations of [@Schurch:2016aa] with the fold change limits
+ shown by the dashed grey lines in Figure 1. The  \(2^{1.4}\) fold change
+cutoff identifies a similar number of transcripts as does  ALDEx2 using \(
+\gamma = 0.5 \) which identifies `r length(sig.s.ald)`.  Supplementary Figures 1
+and 2 shows how to use the \texttt{aldex.senAnalysis()} function to identify
+those transcripts that are very sensitive to scale uncertainty. In this
+supplementary figure we see that even adding a very small amount of scale \(
+\gamma = 0.1 \) reduces the number of significant transcripts by more than half.
+This allows us to ignore those low-dispersion transcripts that were  significant only
+because of an absence of scale uncertainty. In practice, we suggest that a
+\texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental
+designs [@Nixon2024B].
 
-The effect on dispersion with increasing  amounts of scale uncertainty are shown in Figure 2A, where we can see that the  dispersion increases as uncertainty is added.
-Note that the dispersion in the unscaled analysis in Figure 2A reaches a minimum near the mid-point of the distribution, and also does so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
-This shows more clearly that dispersion of many transcripts is  almost negligible in the absence of scale uncertainty.
-This plot makes the counter-intuitive suggestion that the variance in expression of the majority of genes with moderate expression is more predictable  than highly-expressed genes or of housekeeping genes [@Rocha:2020aa].
-This is at odds with the known biology of cells where single cell counting of highly-expressed transcripts shows that they have little intrinsic variation [@Taniguchi:2010aa;@Scott:2010].
+The effect on dispersion with increasing  amounts of scale uncertainty are shown
+in Figure 2A, where we can see that the  dispersion increases as uncertainty is
+added. Note that the dispersion in the unscaled analysis in Figure 2A reaches a
+minimum near the mid-point of the distribution, and also does so when the
+analysis is conducted with DESeq2 (Supplementary Figure 3). This shows more
+clearly that dispersion of many transcripts is  almost negligible in the absence
+of scale uncertainty. This plot makes the counter-intuitive suggestion that the
+variance in expression of the majority of genes with moderate expression is more
+predictable  than highly-expressed genes or of housekeeping genes
+[@Rocha:2020aa].  This is at odds with the known biology of cells where single
+cell counting of highly-expressed transcripts shows that they have little
+intrinsic variation [@Taniguchi:2010aa;@Scott:2010].
 
-Adding scale uncertainty by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 2A by the red and blue data points, and by the colored lines of fit through the centre of mass of the data.
-Less obvious is that the additional dispersion is  not applied equally to all points.
-Figure 2B shows a plot of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data  and here we can see that  scale uncertainty is preferentially increasing the dispersion of the mid-expressed transcripts that formerly had negligible dispersion; examine the grey line of best fit (overlaid by the red line) for the trend.
-Panel B also shows the trend of the expression-dispersion relationship for transcripts that are classed as statistically significant.
-The red line shows the trendline with no added scale uncertainty, and this trendline exactly overlays with the grey trendline for the bulk of transcripts.
-The orange trendline indicates those transcripts that are both statistically significant and that have a thresholded expression level of \( \pm 1.4\), and the dark blue and cyan lines show the statistically significant trendline for \(\gamma=0.5\) , or \(1.0\).
+Adding scale uncertainty by setting \(\gamma=0.5\) , or \(\gamma = 1.0\),
+increases the minimum dispersion as shown in Figure 2A by the red and blue data
+points, and by the colored lines of fit through the centre of mass of the data.
+Less obvious is that the additional dispersion is  not applied equally to all
+points. Figure 2B shows a plot of the difference between the \(\gamma= 0\) and
+\(\gamma= 1\) data  and here we can see that  scale uncertainty is
+preferentially increasing the dispersion of the mid-expressed transcripts that
+formerly had negligible dispersion; examine the grey line of best fit (overlaid
+by the red line) for the trend. Panel B also shows the trend of the
+expression-dispersion relationship for transcripts that are classed as
+statistically significant. The red line shows the trendline with no added scale
+uncertainty, and this trendline exactly overlays with the grey trendline for the
+bulk of transcripts. The orange trendline indicates those transcripts that are
+both statistically significant and that have a thresholded expression level of
+\( \pm 1.4\), and the dark blue and cyan lines show the statistically
+significant trendline for \(\gamma=0.5\) , or \(1.0\).
 
-Thus, taking Figures 1 and 2 together adding scale uncertainty has the desirable effect of changing the distribution of transcripts identified as significantly different between groups.
-Those parts that were statistically significantly different *only because of low dispersion* are preferentially excluded from statistical significance while those parts that were significantly different because of a high difference between groups remain.
+Thus, taking Figures 1 and 2 together adding scale uncertainty has the desirable
+effect of changing the distribution of transcripts identified as significantly
+different between groups. Those parts that were statistically significantly
+different *only because of low dispersion* are preferentially excluded from
+statistical significance while those parts that were significantly different
+because of a high difference between groups remain.
 
-<!-- 
-Interestingly, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions; yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
-The CLR normalization assumes that the scale of the wild-type is exactly 98.4\% of the snf1 strain, and so is \( \theta = 0.016\).
+<!-- Interestingly, the yeast transcriptome experiment has an unacknowledged
+difference in scale between conditions; yeast deficient for snf1 are smaller,
+grow more slowly and are sensitive to a variety of common agents that cause cell
+stress [@Yoshikawa:2011aa]. The CLR normalization assumes that the scale of the
+wild-type is exactly 98.4\% of the snf1 strain, and so is \( \theta = 0.016\).
 Given the per-generation time difference of about 15\%, this seems implausible.
-Setting \(\gamma=0.5\) can be interpreted as acknowledging uncertainty in the difference in the underlaying scale between groups.
-Nixon et al.
--@Nixon2024B showed that the 95\% bounds on the scale uncertainty for a given \(\gamma\) value, say 0.5, can be calculated as \(2^{-(2 \gamma + \theta)}, 2^{(2 \gamma + \theta)}\), or \(2^{-(2*0.5 + 0.016)}, 2^{(2*0.5 + 0.016)} = 0.494,2.022\).
-Thus, by adding scale uncertainty, we are acknowledging that the data are uncertain and that any results are robust to that uncertainty, within the bounds of our model.
--->
+Setting \(\gamma=0.5\) can be interpreted as acknowledging uncertainty in the
+difference in the underlaying scale between groups. Nixon et al. -@Nixon2024B
+showed that the 95\% bounds on the scale uncertainty for a given \(\gamma\)
+value, say 0.5, can be calculated as \(2^{-(2 \gamma + \theta)}, 2^{(2 \gamma +
+\theta)}\), or \(2^{-(2*0.5 + 0.016)}, 2^{(2*0.5 + 0.016)} = 0.494,2.022\).
+Thus, by adding scale uncertainty, we are acknowledging that the data are
+uncertain and that any results are robust to that uncertainty, within the bounds
+of our model. -->
 
@@
 
-Dos Santos et al.
-@dosSantos:2024  used a vaginal metatranscriptome dataset  to compare the gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
-In this environment, both the relative abundance of species between groups and the gene expression level within a species is different [@macklaim:2013].
-Additionally, prior research suggests that the total number of bacteria is about 10 times more in the BV than in the H condition [@Zozaya:2010].
-Thus, these are extremely challenging datasets in which to determine differential abundance as there are both compositional and scale changes between conditions.
-The usual method to analyze vaginal metratranscriptome data is to do so on an organism-by-organism basis [@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding of the environment is less pronounced.
-One attempt at system-wide analysis returned several housekeeping functions as differentially expressed between groups [@Denge00262-18]; a result likely due to a disconnect between the  assumptions of the normalization used and the actual scale of the environment [@Wu2021].
+Dos Santos et al. [@dosSantos:2024]  used a vaginal metatranscriptome dataset  to
+compare the gene expression in bacteria collected from healthy (H) and bacterial
+vaginosis (BV) affected women. In this environment, both the relative abundance
+of species between groups and the gene expression level within a species is
+different [@macklaim:2013]. Additionally, prior research suggests that the total
+number of bacteria is about 10 times more in the BV than in the H condition
+[@Zozaya:2010]. Thus, these are extremely challenging datasets in which to
+determine differential abundance as there are both compositional and scale
+changes between conditions. The usual method to analyze vaginal
+metratranscriptome data is to do so on an organism-by-organism basis
+[@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding
+of the environment is less pronounced. One attempt at system-wide analysis
+returned several housekeeping functions as differentially expressed between
+groups [@Denge00262-18]; a result likely due to a disconnect between the
+assumptions of the normalization used and the actual scale of the environment
+[@Wu2021].
 
-In this example, we show how to specify and interpret an informed scale model  that can explicitly account for some of these modeling difficulties [@Nixon2024B] even in a difficult dataset.
-An informed scale model can control for both the mean difference of scale between groups (e.g., directly incorporate information on the differences in total number of bacteria between the BV and H conditions) as well as the uncertainty of that difference.
-To specify a user-defined scale model, we can pass a matrix of scale values instead of an estimate of just the scale uncertainty  to \texttt{aldex.clr()}.
-This matrix should have the same number of rows as the of Monte-Carlo Dirichlet samples, and the same number of columns as the number of samples.
-While this matrix can be computed from scratch by the analyst, there is an \texttt{aldex.makeScaleModel()} function that can be used to simplify this step in most cases.
-This encodes the scale model as
- \(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents the scale value for each sample or group and gamma is the uncertainty as before.
-The scale estimate can be a measured value (cell count, nucleic acid input, etc) or an estimate.
-Nixon et al.
--@Nixon2024B showed that only the ratio of the means are important when providing values for \( \mu_n\); i.e., the ratio between the \( \log_2 \mu_i\) and \(\log_2 \mu_j\) values.
-See the supplement to  Nixon et al.
--@Nixon2024B for more information.
+In this example, we show how to specify and interpret an informed scale model
+that can explicitly account for some of these modeling difficulties
+[@Nixon2024B] even in a difficult dataset. An informed scale model can control
+for both the mean difference of scale between groups (e.g., directly incorporate
+information on the differences in total number of bacteria between the BV and H
+conditions) as well as the uncertainty of that difference. To specify a
+user-defined scale model, we can pass a matrix of scale values instead of an
+estimate of just the scale uncertainty  to \texttt{aldex.clr()}. This matrix
+should have the same number of rows as the of Monte-Carlo Dirichlet samples, and
+the same number of columns as the number of samples. While this matrix can be
+computed from scratch by the analyst, there is an
+\texttt{aldex.makeScaleModel()} function that can be used to simplify this step
+in most cases. This encodes the scale model as \(\Lambda \sim N(log_2 \mu_n,
+\gamma^{2})\), where \(\mu_n\) represents the scale value for each sample or
+group and gamma is the uncertainty as before. The scale estimate can be a
+measured value (cell count, nucleic acid input, etc) or an estimate.  Nixon et
+al. [@nixon2024scale;@Nixon2024B] showed that only the ratio of the means are important when
+providing values for \( \mu_n\); i.e., the ratio between the \( \log_2 \mu_i\)
+and \(\log_2 \mu_j\) values. See the supplement to  Nixon et al. [@Nixon2024B]
+for more information.
 
-Figure 3A shows an effect plot of the data where reads are grouped by  homologous sequence regardless of the organism of origin.
-Each point represents one of  3728 KEGG  functions [@Okuda:2008].
-There are many more functions represented in the BV group (bottom) than in the healthy group (top).
-This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because	there is a greater diversity of organisms in BV than in H samples, and because the BV condition has at least an order of magnitude more bacteria than does the H condition.
+Figure 3A shows an effect plot of the data where reads are grouped by homologous
+function regardless of the organism of origin. Each point represents one of 
+3728 KEGG  functions [@Okuda:2008]. There are many more functions represented in
+the BV group (bottom) than in the healthy group (top). This is because the
+\textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced
+genome content relative to the anaerobic organisms that dominate in BV, because
+there is a greater diversity of organisms in BV than in H samples, and because
+the BV condition has about  an order of magnitude more bacteria than does the
+H condition.
 
-There are 101  functions with low dispersion that appear to be shared by both groups (boxed area in Figure 3A, and colored in cyan).
-Inspection shows that these   largely correspond to core metabolic functions such as transcription, translation, ribosomal proteins, glycolysis, replication, chaperones, etc (Supplementary file housekeeping.txt).
-The transcripts of many of these are commonly used as invariant reference sequences in wet lab experiments [@Rocha:2020aa] and so are not be expected	 to contribute to differences in ecosystem behaviour.
-The average location of these should be centred on 0 difference to represent an internal reference set.
-However, without an informed scale model, the mean  of these housekeeping functions is  approximately located at +2.3.
+The naive scale model appears to be reflecting the bacterial load as observed by calculating the mean scale value for each group.
+Using a negligible scale value; i.e., \(\gamma=1e-3\) exposes the naive scale estimate for
+samples  in the \texttt{@scaleSamps} slot from the
+\texttt{aldex.clr} output. the naive scale estimate for the
+`r load('analysis/xg.clr.Rda')`
+healthy group is `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])),2)` and
+for the BV group is  `r round(mean(rowMeans(xg@scaleSamps[c(9:36),])),2)` for a
+difference of `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),])),2)`. This is interpreted as the scale of
+the H group of samples being 
+`r round(2^(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),]))),2)` 
+greater than the BV group. This
+precise but incorrect estimate places the location of the housekeeping genes off
+the midline of no difference.
+
+Applying the default scale model of \(\gamma=0.5\) increases the dispersion
+slightly but does not move the   housekeeping functions toward the midline. This
+is as expected; the mean of the default scale model is based on the CLR
+normalization so no shift in location would be expected over the original ALDEx2
+model. Nevertheless, about 30\% of the housekeeping functions are no longer
+statistically significantly different. Note that this change is simple to
+conduct, has no additional computational complexity and requires only a slight
+modification for the analyst.
 
 ```{r ribo, echo=F}
-# list of ribosomal associated functions in the datset
-# TIL about complete.cases()
-ribo.v <- c("K02863","K02864","K02867","K02871","K02874","K02876","K02878","K02879","K02881","K02884","K02886","K02887","K02888","K02890","K02892","K02895","K02897","K02899","K02902","K02904","K02906","K02907","K02909","K02911","K02913","K02914","K02916","K02926","K02931","K02933","K02935","K02939","K02945","K02946","K02948","K02950","K02952","K02954","K02956","K02959","K02961","K02963","K02965","K02967","K02968","K02970","K02982","K02986","K02988","K02990","K02992","K02994","K02996")
+# list of ribosomal associated functions in the datset TIL about
+# complete.cases()
+ribo.v <-c("K02863","K02864","K02867","K02871","K02874","K02876","K02878","K02879","
+K02881","K02884","K02886","K02887","K02888","K02890","K02892","K02895","K02897",
+"K02899","K02902","K02904","K02906","K02907","K02909","K02911","K02913","K02914"
+,"K02916","K02926","K02931","K02933","K02935","K02939","K02945","K02946","K02948
+","K02950","K02952","K02954","K02956","K02959","K02961","K02963","K02965","
+K02967","K02968","K02970","K02982","K02986","K02988","K02990","K02992","K02994",
+"K02996") 
 ```
+
+There are 101  functions with low dispersion that appear to be shared by both
+groups (boxed area in Figure 3A, and colored in cyan). Inspection shows that
+these   largely correspond to core metabolic functions such as transcription,
+translation, ribosomal functions, glycolysis, replication, chaperones, etc
+(Supplementary file housekeeping.txt).  The transcripts of many of these are
+commonly used as invariant reference sequences in wet lab experiments
+[@Rocha:2020aa] and so are not be expected   to contribute to differences in
+ecosystem behaviour. The average location of these should be centred on 0
+difference to represent an internal reference set.  However, without an informed
+scale model, the mean  of these housekeeping functions is  approximately located
+at +2.3.
 
@@
 
-# all code for ALDEx2 and DESeq is being moved to the code/directory
-# all variables recreated using the makefile
-# from code/meta_aldex.R
-load('analysis/xg.clr.Rda')
+# all code for ALDEx2 and DESeq is being moved to the code/directory all
+# variables recreated using the makefile from code/meta_aldex.R
+
 load('analysis/xt.Rda') 
@@
 load('analysis/xt.lv.Rda')
+load('analysis/xg.clr.Rda')
 
-hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3]
-hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3
+hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 &
+xt.all$diff.btw < 3] 
+hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 &
+xt.all$diff.btw < 3
 
 fn <- rownames(xt.m.all)[xt.m.all$we.eBH < 0.01 & xt.m.all$diff.btw < 1 & xt.all$we.eBH > 0.01] 
-fp <- rownames(xt.all)[xt.all$we.eBH < 0.01 & xt.m.all$we.eBH > 0.01 & !xt.m.all$we.eBH < 0.01]
+fp <- rownames(xt.all)[xt.all$we.eBH < 0.01 &
+xt.m.all$we.eBH > 0.01 & !xt.m.all$we.eBH < 0.01]
 
@@
 par(mfrow=c(1,3)) 
-aldex.plot(xt.all, cutoff.pval=0.01, xlim=c(0.3,9), xlab="Dispersion", ylab=l2d)
-points(xt.all[fp,'diff.win'], xt.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
-points(xt.all$diff.win[hk.off], xt.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
-points(xt.all[fn,'diff.win'], xt.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19)
+aldex.plot(xt.all, cutoff.pval=0.01, xlim=c(0.3,9),xlab="Dispersion", ylab=l2d) 
+points(xt.all[fp,'diff.win'],xt.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
+
+points(xt.all$diff.win[hk.off], xt.all$diff.btw[hk.off], pch=19, cex=0.4,
+col=rgb(0,1,1,0.5)) 
+points(xt.all[fn,'diff.win'], xt.all[fn,'diff.btw'],
+col='orange', cex=0.4, pch=19) 
 rect(0.5,1,2.5,3, col=rgb(0,0,0,0.1), lty=2, lwd=3) 
 title('A: ALDEx2 unscaled', adj=0, line= 0.8) 
-legend(0,12, legend=c("HK", "TP", "FP", "FN"), pch=19, cex=0.5, col=c("cyan", "red", "orange", "blue"))
+legend(0,12, legend=c("HK",
+"TP", "FP", "FN"), pch=19, cex=0.5, col=c("cyan", "red", "orange", "blue"))
 
 aldex.plot(xg.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion", ylab=l2d)
-points(xg.all[fp,'diff.win'], xg.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
-points(xg.all$diff.win[hk.off], xg.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
-points(xg.all[fn,'diff.win'], xg.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19)
-title('B: ALDEx2 gamma', adj=0, line= 0.8)
+points(xg.all[fp,'diff.win'], xg.all[fp,'diff.btw'], col='blue', cex=0.4,
+pch=19) 
+points(xg.all$diff.win[hk.off], xg.all$diff.btw[hk.off], pch=19,
+cex=0.4, col=rgb(0,1,1,0.5)) 
+points(xg.all[fn,'diff.win'],
+xg.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19) 
+title('B: ALDEx2 gamma',
+adj=0, line= 0.8)
 
-aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion", ylab=l2d)
-points(xt.m.all[fp,'diff.win'], xt.m.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19)
-points(xt.m.all$diff.win[hk.off], xt.m.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5))
-points(xt.m.all[fn,'diff.win'], xt.m.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19)
-title('C: ALDEx2 both', adj=0, line= 0.8)
+aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion",
+ylab=l2d) 
+points(xt.m.all[fp,'diff.win'], xt.m.all[fp,'diff.btw'], col='blue',
+cex=0.4, pch=19) 
+points(xt.m.all$diff.win[hk.off], xt.m.all$diff.btw[hk.off],
+pch=19, cex=0.4, col=rgb(0,1,1,0.5)) 
+points(xt.m.all[fn,'diff.win'],
+xt.m.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19) 
+title('C: ALDEx2 both',
+adj=0, line= 0.8)
 
@@
 
-Applying the default scale model of \(\gamma=0.5\) increases the dispersion slightly but does not move the   housekeeping functions toward the midline.
-This is as expected; the mean of the default scale model is based on the CLR normalization so no shift in location would be expected over the original ALDEx2 model.
-Nevertheless, about 30\% of the housekeeping functions are no longer statistically significantly different.
-Note that this change is simple to conduct, has no additional computational complexity and requires only a slight modification for the analyst.
+We desire a scale model that approximately centres the housekeeping functions, because we expect housekeeping functions tp be nearly invariant; thus an
+appropriate scale in this dataset for functional analysis is likely closer to 0 than the naive estimate.
+One way to choose an appropriate value for \(\mu_n\) is to use the
+\texttt{aldex.clr} function on only the presumed invariant functions setting
+\(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot as before.
+Doing so suggests that the difference in scale should be about 14%.   A second
+approach would be to identify the functions used as the denominator with the
+\texttt{denom="lvha"} option [@Wu2021] for the \texttt{aldex.clr} function, and
+then to use these values as before. This approach suggests a 5% difference in
+scale, and is potentially less subject to user interpretation.
 
-We desire a scale model that approximately centres the housekeeping functions.
-Our starting point for this is to identify the naive scale estimate from the data, and this can be done by calculating the mean scale value for each group with a negligible scale value; i.e., \(\gamma=1e-3\).
-The scale estimate for samples can be accessed in the \texttt{@scaleSamps} slot from the \texttt{aldex.clr} output.
-In this datset the naive scale estimate for the healthy group is `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])),2)` and for the BV group is  `r round(mean(rowMeans(xg@scaleSamps[c(9:36),])),2)` for a difference of `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),])),2)`.
-This is interpreted as the scale of the H group of samples being `r round(2^(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),]))),2)` greater than the BV group.
-This precise but incorrect estimate places the location of the housekeeping genes off the midline of no difference.
+`r load('analysis/xt.m.Rda')`
 
-We anticipate that housekeeping functions should be nearly invariant and so an appropriate scale in this dataset is likely closer to 0 than the naive estimate.
-One way to choose an appropriate value for \(\mu_n\) is to us the \texttt{aldex.clr} function on only the presumed invariant functions setting \(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot as before.
-Doing so suggests that the difference in scale should be about 14%.
-A second approach would be to identify the functions used as the denominator with the \texttt{denom="lvha"} option [@Wu2021] for the \texttt{aldex.clr} function, and then to use these values as before.
-This approach suggests a 5% difference in scale, and is potentially less subject to user interpretation.
+For the purposes of this example, if we assume a 14% difference in scale, we can
+set \(\mu_i = 1\) and \(\mu_j = 1.14\) using the \texttt{makeScaleMatrix}
+function. This function uses a logNormal distribution to build a scale matrix
+given a user-specified mean difference between groups and uncertainty level.
+Applying a per-group relative differential scale of 0.14 moves the housekeeping
+functions close to the midline of no difference (Figure 3C, assuming 14% mean difference = 
+`r round(mean(xt.m.all[hk.off,"diff.btw"]),2)`, assuming a 5% mean difference = 
+`r round(mean(xt.lv.all[hk.off,"diff.btw"]),2)`), and applying a gamma of 0.5
+provides the same dispersion as in panel B of Figure 3. Note that now a
+significant number of functions are differentially up in BV that were formerly
+classed as not different without the full scale model (orange), or when only a
+default scale was applied. Inspection of the functions shows that these are
+largely missing from the \emph{Lactobacillus} species and so should actually be
+captured as differentially abundant in the BV group. Supplementary Figure 4
+shows that the using either the 5% or the 14%  scale difference  give
+imperceptibly different results suggesting that an informed scale model does not
+have to precisely estimate the scale difference to be useful. Nixon et al,
+@Nixon2024B also found that multiple reasonable estimates for the \(\mu_n\) part
+of the informed scale model were similarly useful in microbiome data.
 
-For the purposes of this example, if we assume a 14% difference in scale, we can set \(\mu_i = 1\) and \(\mu_j = 1.14\) using the \texttt{makeScaleMatrix} function.
-This function uses a logNormal distribution to build a scale matrix given a user-specified mean difference between groups and uncertainty level.
-Applying a per-group relative differential scale of 0.14 moves the housekeeping functions to the midline of no difference (Figure 3C, 14% mean difference = `r round(mean(xt.m.all[hk.off,"diff.btw"]),2)`, 5% difference = `r round(mean(xt.lv.all[hk.off,"diff.btw"]),2)`), and applying a gamma of 0.5 provides the same dispersion as in panel B of Figure 3.
-Note that now a significant number of functions are differentially up in BV that were formerly classed as not different without the full scale model (orange), or when only a default scale was applied.
-Inspection of the functions shows that these are largely missing from the \emph{Lactobacillus} species and so should actually be captured as differentially abundant in the BV group.
-The supplement shows that the using either the 5% or the 14%  scale difference  give  imperceptibly different results suggesting that an informed scale model does not have to precisely estimate the scale difference to be useful.
-Nixon et al, @Nixon2024B also found that multiple reasonable estimates for the \(\mu_n\) part of the informed scale model were similarly useful in microbiome data.
-
- Thus, applying an informed scale allows us to distinguish between both false positives (housekeeping functions in cyan, and others in blue) and false negatives (orange functions) even in a very difficult to analyze dataset.
-The remarkable improvements in biological interpretation afforded by an informed scale model, and the transferrability of it between sample cohorts of the same condition is outlined elsewhere [@dosSantos:2024].
-We suggest that the default scale model  is sufficient when the data are approximately  centred.
-However, an informed model is more appropriate with datasets are not well centred or when the investigator has prior information about the underlying biology.
+Thus, applying an informed scale allows us to distinguish between both false
+positives (housekeeping functions in cyan, and others in blue) and false
+negatives (orange functions) even in a very difficult to analyze dataset. The
+remarkable improvements in biological interpretation afforded by an informed
+scale model, and the transferrability of it between sample cohorts of the same
+condition is outlined elsewhere [@dosSantos:2024]. We suggest that the default
+scale model  is sufficient when the data are approximately  centred. However,
+an informed model is more appropriate with datasets are not well centred or
+when the investigator has prior information about the underlying biology.
 
@@
 
-Biological systems are both predictably variable and stochastic [@Taniguchi:2010aa] and systems biology experiments show that there are transcripts with approximately constant concentrations in the cell and those with large variability under different growth conditions [@Scott:2010].
-Current measurement methods that rely on high throughput sequencing fail to capture all of the variation, particularly variation due to scale [@nixon2023scale;@Nixon2024B].
-In the absence of external information [@Loven:2012aa;@Vandeputte:2017aa;@yeast-absolute] sequencing depth normalisation methods cannot recapture the scale information [@Loven:2012aa;@Lovell:2015], and can only normalize for the technical variation due to sequencing depth.
-Here we demonstrated that even approximate estimates of scale and the uncertainty of measuring it can aid in the interpretation of RNA-sequencing experiments.
+Biological systems are both predictably variable and stochastic
+[@Taniguchi:2010aa] and systems biology experiments show that there are
+transcripts with approximately constant concentrations in the cell and those
+with large variability under different growth conditions [@Scott:2010].  Current
+measurement methods that rely on high throughput sequencing fail to capture all
+of the variation, particularly variation due to scale
+[@nixon2024scale;@Nixon2024B].  In the absence of external information
+[@Loven:2012aa;@Vandeputte:2017aa;@yeast-absolute] sequencing depth
+normalisation methods cannot recapture the scale information
+[@Loven:2012aa;@Lovell:2015], and can only normalize for the technical variation
+due to sequencing depth. Here we demonstrated that even approximate estimates of
+the true system scale and the uncertainty of measuring it can aid in the interpretation of
+RNA-sequencing experiments.
 
-Nixon et al.
-[@nixon2023scale] introduced the idea of explicitly modelling the scale of a HTS dataset, and showed how to incorporate these models in the analysis of microbiome and other datasets [@Nixon2024B].
-They  demonstrated that many tools commonly used to analyze HTS datasets had substantial Type 1 and Type 2 error rates, in line with recent findings by others [@Quinn:2018aa;@Li:2022aa].
-A version of ALDEx2 with the ability to include scale uncertainty was shown to be able to correct for the high Type 1 error rate for that tool, albeit with some loss of sensitivity.
-Finally, they showed that incorporating an informed scale model incorporating both location and scale uncertainty estimates could both control for Type 1 and Type 2 error rates [@Nixon2024B].
+Nixon et al. [@nixon2024scale] introduced the idea of explicitly modeling the
+scale of a HTS dataset, and showed how to incorporate these models in the
+analysis of microbiome and other datasets [@Nixon2024B]. They  demonstrated that
+many tools commonly used to analyze HTS datasets had substantial Type 1 and Type
+2 error rates, in line with recent findings by others
+[@Quinn:2018aa;@Ge:2021aa;@Li:2022aa]. A version of ALDEx2 with the ability to
+include scale uncertainty was shown to be able to correct for the high Type 1
+error rate for that tool, albeit with some loss of sensitivity. Finally, they
+showed that incorporating an informed scale model incorporating both location
+and scale uncertainty estimates could both control for Type 1 and Type 2 error
+rates [@Nixon2024B].
 
-Building and using a scale model, thus has substantial benefits relative to the dual cutoff approach that is advocated for many gene expression experiments [@Cui:2003aa;@Schurch:2016aa].
-In particular, the dual cutoff approach has long been known to not control for Type 1 errors [@Zhang:2009aa;@Ebrahimpoor:2021aa], and the frequent lack of concordance between tools when benchmarked on transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa] suggests poor control of Type 2 errors as well [@Quinn:2018aa;@Li:2022aa].
-Thus, incorporating a scale model during the analysis of HTS data promises the best of both worlds.
-A default scale model can control for Type 1 errors with minimal prior knowledge of the environment and this can be done with essentially no additional computational overhead.
-Furthermore, this work and previous [@Nixon2024B] show that even minimal information about the underlying environment can be used to build a relatively robust informed scale model that controls for both Type 1 and 2 error rates.
+Building and using a scale model thus has substantial benefits relative to the
+dual cutoff approach that is advocated for many gene expression experiments
+[@Cui:2003aa;@Schurch:2016aa]. In particular, the dual cutoff approach has long
+been known to not control for Type 1 errors [@Zhang:2009aa;@Ebrahimpoor:2021aa],
+and the frequent lack of concordance between tools when benchmarked on
+transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Ge:2021aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa;
+@Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa] suggests poor
+control of Type 2 errors as well [@Quinn:2018aa;@Li:2022aa]. Thus, incorporating
+a scale model during the analysis of HTS data promises the best of both worlds.
+A default scale model can control for Type 1 errors with minimal prior knowledge
+of the environment and this can be done with essentially no additional
+computational overhead. Furthermore, this work and previous [@Nixon2024B] show
+that even minimal information about the underlying environment can be used to
+build a relatively robust informed scale model that controls for both Type 1 and
+2 error rates.
 
-In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@Li:2022aa;@nixon2023scale].
-Li et al.
-[@Li:2022aa] conducted a permutation-based benchmarking study and found that widely used tools performed worse than simple Wilcoxon rank-sum tests in controlling the FDR when sample sizes became large.
-They suggested that the presence of outliers were one of the factors driving this observation.
-Brooks et al.
-[@Brooks:2024aa] suggested that inappropriate choice of benchmarking methods are also a major contributing factor and that objective standards of truth are important.
-From the perspective of our work the disagreement between tools can be explained by the observation that different analytic approaches  produce  different parameter estimates for location or scale.
-Thus, more data produces worse estimates because the additional data simply increases the precision of a flawed estimate [@gustafson2015bayesian;@nixon2023scale].
+In the analysis of HTS data it is often observed that larger datasets converge
+on the majority of parts being significantly different
+[@Schurch:2016aa;@Li:2022aa;@nixon2024scale]. Li et al. [@Li:2022aa] conducted a
+permutation-based benchmarking study and found that widely used tools performed
+worse than simple Wilcoxon rank-sum tests in controlling the FDR when sample
+sizes became large. They suggested that the presence of outliers were one of the
+factors driving this observation. Brooks et al. [@Brooks:2024aa] suggested that
+inappropriate choice of benchmarking methods are also a major contributing
+factor and that objective standards of truth are important. From the perspective
+of our work the disagreement between tools can be explained by the observation
+that different analytic approaches  produce  different parameter estimates for
+location or scale or for both. Thus, more data produces worse estimates because the
+additional data simply increases the precision of a flawed estimate
+[@gustafson2015bayesian;@nixon2024scale].
 
-Scale simulation  is now build into ALDEx2 [@Nixon2024B] and this work suggest that  there are two main root causes to common HTS data pathologies.
-The first contributing factor is the observed very low dispersion estimate for many features that is a by-product of some experimental designs and of normalization.
-In the Schurch et al.
-@Schurch:2016aa dataset, the data were  from single colonies derived from a single culture.
-Thus, it is more accurate to describe the 96 samples as wet-lab technical replicates rather than independent samples.
-This type of replication approach is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.
-Applying the default scale model with \( \gamma=0.5 \) a large number of transcripts have  their dispersion increased (Figure 1D), with the effect being largest for those with the lowest initial dispersion (Figure 2).
-Adding scale uncertainty results in modest number of transcripts, `r length(sig.s.ald)`, being called significantly different as shown in the volcano plot in Figure 1B (red points).
-In addition, there is now a strong concordance between the difference and q-values.
-In hindsight, it is not obvious  why the unscaled volcano plot shows such poor correspondence.
-We suggest that this is explained by random fluctuations of the many very low variance estimates and this is supported by the  plots shown in Figure 2.
+Scale simulation  is now built into ALDEx2 [@Nixon2024B] and here we suggest
+that  there are two main root causes to common HTS data pathologies. The first
+contributing factor is the observed very low dispersion estimate for many
+features that is a by-product of some experimental designs and of normalization.
+In the Schurch et al. [@Schurch:2016aa dataset], the data were  from single
+colonies derived from a single culture. Thus, it is more accurate to describe
+the 96 samples as wet-lab technical replicates rather than independent samples.
+This type of replication approach is standard in the molecular literature, and
+would be expected to result in the very low dispersion that is observed.
+Applying the default scale model with \( \gamma=0.5 \) a large number of
+transcripts have  their dispersion increased (Figure 1D), with the effect being
+largest for those with the lowest initial dispersion (Figure 2). Adding scale
+uncertainty results in modest number of transcripts, `r length(sig.s.ald)`,
+being called significantly different as shown in the volcano plot in Figure 1B
+(red points). In addition, there is now a strong concordance between the
+difference and q-values. In hindsight, it is not obvious  why the unscaled
+volcano plot shows such poor correspondence. We suggest that this is explained
+by random fluctuations of the many very low variance estimates and this is
+supported by the  plots shown in Figure 2.
 
-The second contributing factor is unacknowledged asymmetry in many datasets [@Wu2021]; i.e., different gene content or a directional change in the majority of features.
-In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult-to-analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change.
-We showed one such example for the metatranscriptome dataset in Figure 3. Here the dataset was highly asymmetrical.
-Incorporating differential scale on a per-group basis moves the mass of the housekeeping functions towards the midline of no difference and so affects both Type I and Type II error rates.
-We showed two ways of estimating the  scale difference between groups and found that any reasonable estimate is an improvement over the naive approach and also over the default scale model.
-This is in line with the observations by Nixon et al [@Nixon2024B] in a 16S rRNA gene sequencing dataset.
-It is also of note that in the case of true biological replicates (different individuals) that adding a modest amount of scale \(\gamma=0.5\) had little effect on the the difference between groups and on the dispersion.
-Thus, in this dataset the scale mis-specification was affecting mainly the location of the difference between groups.
-While we acknowledge that some prior information on which housekeeping transcripts should not be classed as differentially abundant is needed, we suggest that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
+The second contributing factor is unacknowledged asymmetry in many datasets
+[@Wu2021]; i.e., different gene content or a directional change in the majority
+of features. In the case of asymmetry, the use of a user-specified scale model
+can be very useful for otherwise difficult-to-analyze datasets such as
+meta-transcriptomes and in-vitro selection datasets where the majority of
+features can change. We showed one such example for the metatranscriptome
+dataset in Figure 3. Here the dataset was highly asymmetrical. Incorporating
+differential scale on a per-group basis moves the mass of the housekeeping
+functions towards the midline of no difference and so affects both Type I and
+Type II error rates. We showed two ways of estimating the  scale difference
+between groups and found that any reasonable estimate is an improvement over the
+naive approach and also over the default scale model. This is in line with the
+observations by Nixon et al [@Nixon2024B] in a 16S rRNA gene sequencing dataset.
+It is also of note that in the case of true biological replicates (different
+individuals) that adding a modest amount of scale \(\gamma=0.5\) had little
+effect on the the difference between groups and on the dispersion. Thus, in this
+dataset the scale mis-specification was affecting mainly the location of the
+difference between groups.  While we acknowledge that some prior information on
+which housekeeping transcripts should not be classed as differentially abundant
+is needed, we suggest that this information is widely available and is already
+used when performing the gold-standard quantitative PCR test of differential
+abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
 
-Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses.
-The addition of scale uncertainty essentially tests the model over a range of normalizations [@nixon2023scale] and so can replace the consensus approach that has been proposed by some groups [@Nearing:2022aa;@Song:2023aa] with no additional computational overhead.
-Thus, an advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
-While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions and a detailed description of these points is describe elsewhere [@nixon2023scale].
+Beyond concerns of fidelity and rigor, scale models also enhance the
+reproducibility and transparency of HTS analyses. The addition of scale
+uncertainty essentially tests the model over a range of normalizations
+[@nixon2024scale] and so can replace the consensus approach that has been
+proposed by some groups [@Nearing:2022aa;@Song:2023aa] with no additional
+computational overhead. Thus, an advantage of incorporating scale is that
+analyses can be made much more robust such that actual or potential differences
+in scale can be tested and accounted for explicitly. While it is beyond the
+scope of the present article, we note that there are many ways of building scale
+models that enhance the interpretability of the parameters and assumptions and a
+detailed description of these points is describe elsewhere [@nixon2024scale,].
 
-In summary, we supply a toolkit that makes incorporating scale uncertainty and location information simple to incorporate for transcriptomes or indeed any type of HTS dataset.
-While the underlying scale of the system is generally inaccessible, the effect of scale on the analysis outcomes can be modelled and can help explain some of the underlying biology, and help to expose known issues with the analysis of HTS data.
-Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on conclusions weighted accordingly.
-Additionally, the use of informed  scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups [@dosSantos:2024].
-Thus, using and reporting scale uncertainty should become a standard practice in the analysis of HTS datasets.
+In summary, we supply a toolkit that makes incorporating scale uncertainty and
+location information simple to incorporate for transcriptomes or indeed any type
+of HTS dataset. While the underlying scale of the system is generally
+inaccessible, the effect of scale on the analysis outcomes can be modelled and
+can help explain some of the underlying biology, and help to expose known issues
+with the analysis of HTS data. Adding scale information to the analysis allows
+for more robust inference because the features that are sensitive to scale can
+be identified and their impact on conclusions weighted accordingly.
+Additionally, the use of informed  scale models permits difficult to analyze
+datasets to be examined in a robust and principled manner even when the majority
+of features are asymmetrically distributed or expressed (or both) in the groups
+[@dosSantos:2024]. Thus, using and reporting scale uncertainty should become a
+standard practice in the analysis of HTS datasets.
 

### 2025-03-11 go3.tex
submitted version Dec 6

@@
 in sequencing depth is driven by technical factors, and not by variation
-in the scale (e.g., total size, microbial load, or total mRNA
-expression) of the underlying biological systems. Typically a
-statistical normalization is used to remove unwanted technical variation
-in the data or the parameters of the model to enable analyses that are
-sensitive to scale; e.g., differential abundance and differential
-expression analyses. We recently showed that all normalizations make
-implicit assumptions about the unmeasured system scale and that errors
-in these assumptions can dramatically increase false positive and false
-negative rates. We demonstrated that these errors can be mitigated by
-accounting for uncertainty about scale using a \emph{scale model}, which
-we integrated into the ALDEx2 R package. This article provides new
-insights into those methods, focusing on the application to
-transcriptomic analysis. Here we provide transcriptomic case studies
-demonstrating how scale models, rather than traditional normalizations,
-can reduce false positive and false negative rates in practice while
-enhancing the transparency and reproducibility of analyses. We show that
-these scale models replace the need for dual cutoff approaches often
-used to address the disconnect between practical and statistical
-significance. We demonstrate the utility of that scale models built
-based on known housekeeping genes in complex metatranscriptomic
-datasets. Thus this work provides example and practical guidance on how
-to incorporate scale into transcriptomic analysis.
+in the scale (size) of the biological system. Typically a statistical
+normalization removes unwanted technical variation in the data or the
+parameters of the model to enable differential abundance analyses. We
+recently showed that all normalizations make implicit assumptions about
+the unmeasured system scale and that errors in these assumptions can
+dramatically increase false positive and false negative rates. We
+demonstrated that these errors can be mitigated by accounting for
+uncertainty using a \emph{scale model}, which we integrated into the
+ALDEx2 R package. This article provides new insights focusing on the
+application to transcriptomic analysis. We provide transcriptomic case
+studies demonstrating how scale models, rather than traditional
+normalizations, can reduce false positive and false negative rates in
+practice while enhancing the transparency and reproducibility of
+analyses. These scale models replace the need for dual cutoff approaches
+often used to address the disconnect between practical and statistical
+significance. We demonstrate the utility of scale models built based on
+known housekeeping genes in complex metatranscriptomic datasets. Thus
+this work provides guidance on how to incorporate scale into
+transcriptomic data sets.
 \end{abstract}
@@
 \doublespacing 
+\singlespacing
 
@@
 acid is carried over to subsequent steps. Thus, with each sampling step
-the connection between the size of the sampled DNA pool and the scale
-(e.g., size, microbial load, or total gene expression) of the measured
-biological system is degraded or lost. In the end, the information
-contained in the data relates only to relative abundances and has an
-arbitrary scale imposed by the sequencing process(1--3). Increasingly,
-researchers are turning to modified experimental protocols in an attempt
-to recover the lost biological variation in scale; see for example (4,
-5). However, these protocols only recover scale variation at the
-intermediate step in the sample preparation protocol where the
-intervention was made. In short, the disconnect between sample-to-sample
-variation in sequencing depth and biological variation in scale remains
-an open challenge.
+the connection between the actual size of the sampled DNA pool and the
+scale (e.g., size, microbial load, or total gene expression) of the
+measured biological system is degraded or lost. In the end, the
+information contained in the data relates only to relative abundances
+and has an arbitrary scale imposed by the sequencing process (1--3).
+Researchers can use modified experimental protocols (4--7) or machine
+learning methods (8) to uncover the biological variation in scale.
+However, wet-lab protocols only provide information on the size of the
+data downstream of the step in the sample preparation protocol where the
+intervention was made and introduce an additional source of variation
+that must be accounted for (5). The computational methods developed for
+microbiome analysis have low correlation with the actual scale but are
+useful (8). In short, the disconnect between sample-to-sample variation
+in sequencing depth and biological variation in scale remains an open
+challenge.
 
@@
 traced, in whole or in part, to misspecification of scale. The first
-issue is poor control of the false discovery rate (FDR) (6--9),
+issue is poor control of the false discovery rate (FDR) (9--13),
 exhibited as dataset-dependent FDR control and by the disconnect between
-statistical and biological significance(10). In current practice, these
+statistical and biological significance (14). In current practice, these
 issues are addressed by a dual-filtering method, whereby both a low
-p-value (or equivalently a low q-value following FDR correction (11))
+p-value (or equivalently a low q-value following FDR correction (15))
 and a large difference between groups is used to identify interesting
-transcripts or genes for follow-up analysis(10). This double-filtering
-approach is graphically exemplified by the volcano plot (12), but is
-known to not appropriately control the FDR\\
-(13, 14). The second issue is poor performance when analyzing data where
-the mean change between groups is non-zero. Such asymmetric data can
-arise when a gene set is expressed in one group but not the other, or
-when one group contains different gene content from the other. This type
-of data frequently arises in in-vitro selection experiments (SELEX),
-transcriptome analysis, and microbiome analysis (15). The third issue is
-that these problems become more pronounced as more samples are
-collected; that is, more information results in a worsening of the
-accuracy of the analysis (16).
-
-The three problems were recently shown by Nixon and Silverman (3) to be
-a result of a mismatch between the underlying size or scale of the
-system and the assumptions of the normalizations used for the analysis
-of HTS. Biological variation in scale often represents an important
-unmeasured confounder in HTS analyses (17). For example, cells
-transformed by the cMyc oncogene have about 3 times the amount of mRNA
-and about twice the rRNA content than non-transformed cells (18), and
-this dramatically skews transcriptome analysis (19). In addition,
-wild-type and mutant strains of cell lines, yeast or bacteria have
-different growth rates and RNA contents under different conditions,
+transcripts or genes for follow-up analysis (14, 16). This
+double-filtering approach is graphically exemplified by the volcano plot
+(16), but is known to not appropriately control the FDR (17, 18). The
+second issue is poor performance when analyzing data where the mean
+change between groups is non-zero (3). Such asymmetric data can arise
+when a gene set is expressed in one group but not the other, or when one
+group contains different gene content from the other. This type of data
+frequently arises in in-vitro selection experiments (SELEX),
+transcriptome analysis, and microbiome analysis (19). The third issue is
+that the actual scale of the environment is often a major confounding
+variable during analysis (3, 8). The final issue is that these problems
+become more pronounced as more samples are collected; that is, more
+information results in a worsening of the accuracy of the analysis (3,
+13, 20).
+
+The four problems were recently shown by Nixon et al. (3) to be a result
+of a mismatch between the underlying size or scale of the system and the
+assumptions of the normalizations used for the analysis of HTS.
+Biological variation in scale often represents an important unmeasured
+confounder in HTS analyses (21). For example, cells transformed by the
+cMyc oncogene have about 3 times the amount of mRNA and about twice the
+rRNA content than non-transformed cells (22), and this dramatically
+skews transcriptome analysis unless spike-in approaches are used (5). In
+addition, wild-type and mutant strains of cell lines, yeast or bacteria
+have different growth rates and RNA contents under different conditions,
 which affect our ability to identify truly differentially abundant genes
-(20--22). As another example, the total bacterial load of the vaginal
+(23--25). As another example, the total bacterial load of the vaginal
 microbiome differs by 1-2 orders of magnitude in absolute abundance
-between the healthy and bacterial vaginosis states (23), and the
-composition between these states is dramatically different (24, 25).
+between the healthy and bacterial vaginosis states (26), and the
+composition between these states is dramatically different (27, 28).
 Thus, a full description of any of these systems includes both relative
@@
 access only the compositional information yet make implicit assumptions
-about the scale (16).
+about the scale (20).
 
@@
 make some assumption about scale but these implicit assumptions are
-often inappropriate and difficult to interpret. Because of this
-different normalizations provide different outputs when applied to the
-same dataset (6, 8, 26--28). Intuitively, normalizations in widespread
-use assume that either all samples have the same scale,
-e.g.~proportions, rarefaction (29), RPKM (30, 31), etc; or that a subset
-of features in one sample can be chosen as a reference to which the
-others are scaled e.g.~the TMM (32), or LVHA (15) or the additive
-log-ratio (33); or that different sub-parts of each sample maintain a
-constant scale across samples e.g.~the RLE (34); or that the geometric
-mean of the parts is appropriate e.g the CLR (35) and its derivatives.
-
-The original naive ALDEx2 (36) model unwittingly made a strict
-assumption about scale through the CLR normalization (3). While this
-assumption could be useful in many cases it could never be exactly true,
-and others have shown that it is not always the best option (37). Nixon
-et al. (3) showed that better scale assumptions resulted in more
-reproducible data analysis including better control of both false
-positive and false negative results. We recently modified ALDEx2 to
-explicitly model the scale over a range of reasonable normalization
-parameters, and showed significant improvements in performance in
-microbiome and in-vitro selection experiments (16). Here, we briefly
-introduce review modifications and then show how scale uncertainty can
+often inappropriate and difficult to interpret. This causes different
+normalizations to provide different outputs when applied to the same
+dataset (9, 11, 29--31). Intuitively, normalizations in widespread use
+assume that either all samples have the same scale, e.g.~proportions,
+rarefaction (32), RPKM (33, 34), etc; or that a subset of features in
+one sample can be chosen as a reference to which the others are scaled
+e.g.~the TMM (35), or LVHA (19) or the additive log-ratio (36); or that
+different sub-parts of each sample maintain a constant scale across
+samples e.g.~the RLE (37); or that the geometric mean of the parts is
+appropriate e.g the CLR (38) and its derivatives.
+
+The original naive ALDEx2 (39) model unwittingly made a strict
+assumption about scale through the CLR normalization (3). This
+assumption was often close enough to the true value to be useful, but
+was not always the a good estimate and could be outperformed by other
+normalizations (40). Nixon et al. (3) showed that better scale
+assumptions resulted in more reproducible data analysis including better
+control of both false positive and false negative results. We recently
+modified ALDEx2 to explicitly model the scale over a range of reasonable
+normalization parameters, and showed significant improvements in
+performance in microbiome and in-vitro selection experiments (20). Here,
+we briefly review these modifications and show how scale uncertainty can
 greatly improve modeling in transcriptome and meta-transcriptome
@@
 Formal and expanded descriptions of the concepts that follow are given
-in (3, 16). To be concrete, we let \(\mathbf{Y}\) denote the
+in (3, 20). To be concrete, we let \(\mathbf{Y}\) denote the
 \emph{measured} \(D \times N\) matrix of sequence counts with elements
@@
 or as the logarithm
-\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).
+\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
+\mathbf{W}^{\perp}_{n}\).
 
 Many of the normalizations in widespread use in tools such as DESeq2
-(38), edgeR (32), metagenomeSeq (39) ALDEx2 (40) can be stated as ratios
-of the form
-\(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where
-the denominator is determined by some function of the observation. We
-use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate
-that the output is an estimate of the true value. The technical
-variation in sequencing depth
+(41), edgeR (35), metagenomeSeq (42) ALDEx2 (43) can be stated as ratios
+of the form \(\hat{{\mathbf{W}}}_{dn}
+\approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is
+determined by some function of the observation. We use the
+\(\hat{{\mathbf{W}}}\) (\(\ \hat{}\
+\)) notation to indicate that the output is an estimate of the true
+value. The technical variation in sequencing depth
 (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
@@
 
-The ALDEx2 R package (36, 40) is a general purpose toolbox to model the
-uncertainty of HTS data and to use that model to estimate LFC
-significance. At a high-level, ALDEx2 has three connected components to
-estimate the uncertainty inherent in HTS datasets. First, the tool
-accounts for the uncertainty of the sequencing counts using Dirichlet
-multinomial sampling to build a probabilistic model of the data; i.e.,
+The ALDEx2 R package (39, 43) is a general purpose toolbox to model the
+uncertainty of HTS data and to use that model to estimate the underlying
+LFC (log-fold change) significance. At a high-level, ALDEx2 has three
+connected components to estimate the uncertainty inherent in HTS
+datasets. First, the tool accounts for the uncertainty of the sequencing
+counts using Dirichlet multinomial sampling to build a probabilistic
+model of the data; i.e.,
 \(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
 Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
-data (36). However, this step was modified recently to account for scale
-uncertainty and misspecification (16) via a scale model, explained with
-more details in (3, 16) and summarized in the next paragraph. Finally, a
+data (39). However, this step was modified recently to account for scale
+uncertainty and misspecification (20) via a scale model, explained with
+more details in (3, 20) and summarized in the next paragraph. Finally, a
 standard null-hypothesis test and a non-parametric estimate of mean
@@
 difference between groups, and `dispersion' to refer to the
-within-condition difference or pooled variance as defined in (36). These
+within-condition difference or pooled variance as defined in (39). These
 are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
-16, 36, 40).
+20, 39, 43).
 
@@
 centred log-ratio normalization step. They did this by including a model
-for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR method used by ALDEx2
-makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where
-\(G_n\) is the geometric mean of sample n, which while being a random
-variable, is essentially constant across each Monte-Carlo replicate.
-With this modificaiton, ALDEx2 can be generalized by considering
-probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that
-have mean \(1/G_{n}\). For example, the following scale model
-generalizes the CLR:
-
-\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2})\].
-
-This formulation is quite flexible (3, 16). In the simple or `default'
+for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by
+ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\),
+where \(G_n\) is the geometric mean of sample n, which while being a
+random variable, is essentially constant across each Monte-Carlo
+replicate, but that differs between samples. With this modification,
+ALDEx2 can be generalized by considering probability models for the
+scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
+example, the following scale model generalizes the CLR:
+
+\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda
+\sim N(\mu, \gamma^{2})\].
+
+This formulation is quite flexible (3, 20). In the simple or `default'
 configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
-from a log-Normal distribution that adds scale uncertainty. The default
-approach controls only the degree of uncertainty in the CLR assumption
-for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes
-case and \(x_{n}=0\) denotes control). In the advanced or `informed'
-configuration, \(\mu\) takes different values for each group, and this
-controls the location of the LFC assumption. An example of both the
-default and informed approaches is given for a microbiome dataset in
-(16), and here we show that these approaches also work well in
-transcriptome and metatranscriptome datasets. These modifications are
-instantiated in ALDEx2 which is the first software package designed for
-SSRV-based inference.
+from a log-Normal distribution(3). Adding scale uncertainty with the
+\(\gamma\) paramenter controls only the degree of uncertainty of the CLR
+assumption for the \(x_{n}\) binary condition indicator (e.g.,
+\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). In the
+advanced or `informed' configuration, \(\mu\) takes different values for
+each group and controls the location of the LFC assumption; combining
+\(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the
+location and the scale. An example of both the default and informed
+approaches is given for a microbiome dataset in (20) showing increased
+sensitivity and specificity. Here we show that these approaches also
+work well in transcriptome and metatranscriptome datasets. These
+modifications are instantiated in ALDEx2 which is the first software
+package designed for SSRV-based inference.
 
@@
 
-Gierliński et al. (41) generated a highly replicated yeast transcriptome
+Gierliński et al. (44) generated a highly replicated yeast transcriptome
 dataset to compare gene expression between a wild-type strain and a snf2
@@
 abundant transcripts identified in the full dataset when the data was
-subset (10). In this study each tool had its own `gold standard' set of
+subset (14). In this study each tool had its own `gold standard' set of
 transcripts with different tools identifying between between 65\% to
@@
 authors suggested that it was more appropriate to apply a dual cutoff
-composed of both a Benjamini-Hochberg (42) corrected p-value (q-value)
+composed of both a Benjamini-Hochberg (45) corrected p-value (q-value)
 plus a difference cutoff to limit the number of identified transcripts
-to a much smaller fraction of the total. Nixon et al, (16) showed that
+to a much smaller fraction of the total. Nixon et al, (20) showed that
 adding even a small amount of scale uncertainty with ALDEx2 dramatically
@@
 need for the dual cutoff approach in this dataset and others. Below we
-include an intuitive explanation of how incorporating scale uncertainty
-achieved this outcome using a setting of \(\gamma=0.5\), which is in
-line with the recommendations of (16) and that gives results comparable
-to those proposed in (10).
+include an intuitive explanation of why and how incorporating scale
+uncertainty achieved this outcome using a setting of \(\gamma=0.5\). The
+approach is in line with the recommendations of (20) and gives results
+comparable to those proposed in (14).
 
 We start with the assumption that not all statistically significant
-differences are biologically relevant (43), and that such a large number
+differences are biologically relevant (46), and that such a large number
 of significantly different transcripts breaks the necessary assumption
-for DA/DE expression that most parts be invariant (27). As noted,
+for DA/DE expression that most parts be invariant (30). As noted,
 transcriptomics commonly uses a dual cutoff approach that is graphically
-exemplified by volcano plots (10, 12). Using either DESeq2 or ALDEx2, a
+exemplified by volcano plots (14, 16). Using either DESeq2 or ALDEx2, a
 majority of transcripts are statistically significantly different
@@
 DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. These values are
-in line with those observed by (10). Such large numbers of statistically
+in line with those observed by (14). Such large numbers of statistically
 significant transcripts seems biologically unrealistic. That 118
@@
 significant and that some, if not the majority, are driven by technical
-differences in the analysis (9, 27).
+differences in the analysis (13, 30).
 
@@
 q-value and the difference between groups (compare panels A and B). The
-effect plots (44) in Figure 1C shows that the majority of significant
+effect plots (47) in Figure 1C shows that the majority of significant
 transcripts (red, orange) have negligible differences between groups and
@@
 the experimental design which is actually a technical wet lab
-replication rather than a true biological replication design (41). Scale
+replication rather than a true biological replication design (44). Scale
 uncertainty can be incorporated using the \texttt{gamma} parameter that
@@
 It is common practice to use a dual-cutoff by choosing transcripts based
-on a thresholds for both q-values and fold-changes (10, 12). Note that
-there is considerable variation in recommended cutoff values(10). Here,
+on a thresholds for both q-values and fold-changes (14, 16). Note that
+there is considerable variation in recommended cutoff values(14). Here,
 applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold
@@
 to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line
-with the recommendations of (10). the fold change limits are shown by
+with the recommendations of (14) with the fold change limits shown by
 the dashed grey lines in Figure 1. The \(2^{1.4}\) fold change cutoff
 identifies a similar number of transcripts as does ALDEx2 using
-\(\gamma = 0.5\) which identifies 205. Supplementary Figure 2 shows how
-to use the \texttt{aldex.senAnalysis()} function to identify those
-transcripts that are very sensitive to scale uncertainty. In this
+\(\gamma = 0.5\) which identifies 205. Supplementary Figures 1 and 2
+shows how to use the \texttt{aldex.senAnalysis()} function to identify
+those transcripts that are very sensitive to scale uncertainty. In this
 supplementary figure we see that even adding a very small amount of
 scale \(\gamma = 0.1\) reduces the number of significant transcripts by
-more than half. This allows us to those low-dispertion transcripts that
-were significant only because of an absence of scale uncertainty. In
-practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
-is realistic for most experimental designs (16).
+more than half. This allows us to ignore those low-dispersion
+transcripts that were significant only because of an absence of scale
+uncertainty. In practice, we suggest that a \texttt{gamma} parameter
+between 0.5 and 1 is realistic for most experimental designs (20).
 
@@
 is more predictable than highly-expressed genes or of housekeeping genes
-(45). This is at odds with the known biology of cells where single cell
+(48). This is at odds with the known biology of cells where single cell
 counting of highly-expressed transcripts shows that they have little
-intrinsic variation (20, 46).
+intrinsic variation (23, 49).
 
@@
 
-Dos Santos et al. (47) used a vaginal metatranscriptome dataset to
+Dos Santos et al. (50) used a vaginal metatranscriptome dataset to
 compare the gene expression in bacteria collected from healthy (H) and
@@
 relative abundance of species between groups and the gene expression
-level within a species is different (48). Additionally, prior research
+level within a species is different (51). Additionally, prior research
 suggests that the total number of bacteria is about 10 times more in the
-BV than in the H condition (23). Thus, these are extremely challenging
+BV than in the H condition (26). Thus, these are extremely challenging
 datasets in which to determine differential abundance as there are both
@@
 analyze vaginal metratranscriptome data is to do so on an
-organism-by-organism basis (48--50) because the scale confounding of the
+organism-by-organism basis (51--53) because the scale confounding of the
 environment is less pronounced. One attempt at system-wide analysis
 returned several housekeeping functions as differentially expressed
-between groups (49); a result likely due to a disconnect between the
+between groups (52); a result likely due to a disconnect between the
 assumptions of the normalization used and the actual scale of the
-environment (15).
+environment (19).
 
@@
 model that can explicitly account for some of these modeling
-difficulties (16) even in a difficult dataset. An informed scale model
+difficulties (20) even in a difficult dataset. An informed scale model
 can control for both the mean difference of scale between groups (e.g.,
@@
 this step in most cases. This encodes the scale model as
-\(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents
-the scale value for each sample or group and gamma is the uncertainty as
-before. The scale estimate can be a measured value (cell count, nucleic
-acid input, etc) or an estimate. Nixon et al. (16) showed that only the
-ratio of the means are important when providing values for \(\mu_n\);
-i.e., the ratio between the \(\log_2 \mu_i\) and \(\log_2 \mu_j\)
-values. See the supplement to Nixon et al. (16) for more information.
+\(\Lambda \sim N(log_2 \mu_n,
+\gamma^{2})\), where \(\mu_n\) represents the scale value for each
+sample or group and gamma is the uncertainty as before. The scale
+estimate can be a measured value (cell count, nucleic acid input, etc)
+or an estimate. Nixon et al. (3, 20) showed that only the ratio of the
+means are important when providing values for \(\mu_n\); i.e., the ratio
+between the \(\log_2 \mu_i\) and \(\log_2 \mu_j\) values. See the
+supplement to Nixon et al. (20) for more information.
 
 Figure 3A shows an effect plot of the data where reads are grouped by
-homologous sequence regardless of the organism of origin. Each point
-represents one of 3728 KEGG functions (51). There are many more
+homologous function regardless of the organism of origin. Each point
+represents one of 3728 KEGG functions (54). There are many more
 functions represented in the BV group (bottom) than in the healthy group
@@
 organisms that dominate in BV, because there is a greater diversity of
-organisms in BV than in H samples, and because the BV condition has at
-least an order of magnitude more bacteria than does the H condition.
+organisms in BV than in H samples, and because the BV condition has
+about an order of magnitude more bacteria than does the H condition.
+
+The naive scale model appears to be reflecting the bacterial load as
+observed by calculating the mean scale value for each group. Using a
+negligible scale value; i.e., \(\gamma=1e-3\) exposes the naive scale
+estimate for samples in the \texttt{@scaleSamps} slot from the
+\texttt{aldex.clr} output. the naive scale estimate for the
+
+healthy group is 17.41 and for the BV group is 14.59 for a difference of
+2.82. This is interpreted as the scale of the H group of samples being
+7.06 greater than the BV group. This precise but incorrect estimate
+places the location of the housekeeping genes off the midline of no
+difference.
+
+Applying the default scale model of \(\gamma=0.5\) increases the
+dispersion slightly but does not move the housekeeping functions toward
+the midline. This is as expected; the mean of the default scale model is
+based on the CLR normalization so no shift in location would be expected
+over the original ALDEx2 model. Nevertheless, about 30\% of the
+housekeeping functions are no longer statistically significantly
+different. Note that this change is simple to conduct, has no additional
+computational complexity and requires only a slight modification for the
+analyst.
 
@@
 shows that these largely correspond to core metabolic functions such as
-transcription, translation, ribosomal proteins, glycolysis, replication,
-chaperones, etc (Supplementary file housekeeping.txt). The transcripts
-of many of these are commonly used as invariant reference sequences in
-wet lab experiments (45) and so are not be expected to contribute to
-differences in ecosystem behaviour. The average location of these should
-be centred on 0 difference to represent an internal reference set.
-However, without an informed scale model, the mean of these housekeeping
-functions is approximately located at +2.3.
+transcription, translation, ribosomal functions, glycolysis,
+replication, chaperones, etc (Supplementary file housekeeping.txt). The
+transcripts of many of these are commonly used as invariant reference
+sequences in wet lab experiments (48) and so are not be expected to
+contribute to differences in ecosystem behaviour. The average location
+of these should be centred on 0 difference to represent an internal
+reference set. However, without an informed scale model, the mean of
+these housekeeping functions is approximately located at +2.3.
 
@@
 
-Applying the default scale model of \(\gamma=0.5\) increases the
-dispersion slightly but does not move the housekeeping functions toward
-the midline. This is as expected; the mean of the default scale model is
-based on the CLR normalization so no shift in location would be expected
-over the original ALDEx2 model. Nevertheless, about 30\% of the
-housekeeping functions are no longer statistically significantly
-different. Note that this change is simple to conduct, has no additional
-computational complexity and requires only a slight modification for the
-analyst.
-
 We desire a scale model that approximately centres the housekeeping
-functions. Our starting point for this is to identify the naive scale
-estimate from the data, and this can be done by calculating the mean
-scale value for each group with a negligible scale value; i.e.,
-\(\gamma=1e-3\). The scale estimate for samples can be accessed in the
-\texttt{@scaleSamps} slot from the \texttt{aldex.clr} output. In this
-datset the naive scale estimate for the healthy group is 17.41 and for
-the BV group is 14.59 for a difference of 2.82. This is interpreted as
-the scale of the H group of samples being 7.06 greater than the BV
-group. This precise but incorrect estimate places the location of the
-housekeeping genes off the midline of no difference.
-
-We anticipate that housekeeping functions should be nearly invariant and
-so an appropriate scale in this dataset is likely closer to 0 than the
-naive estimate. One way to choose an appropriate value for \(\mu_n\) is
-to us the \texttt{aldex.clr} function on only the presumed invariant
-functions setting \(\gamma > 0\), and then accessing the
-\texttt{@scaleSamps} slot as before. Doing so suggests that the
-difference in scale should be about 14\%. A second approach would be to
-identify the functions used as the denominator with the
-\texttt{denom="lvha"} option (15) for the \texttt{aldex.clr} function,
-and then to use these values as before. This approach suggests a 5\%
-difference in scale, and is potentially less subject to user
-interpretation.
+functions, because we expect housekeeping functions tp be nearly
+invariant; thus an appropriate scale in this dataset for functional
+analysis is likely closer to 0 than the naive estimate. One way to
+choose an appropriate value for \(\mu_n\) is to use the
+\texttt{aldex.clr} function on only the presumed invariant functions
+setting \(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot
+as before. Doing so suggests that the difference in scale should be
+about 14\%. A second approach would be to identify the functions used as
+the denominator with the \texttt{denom="lvha"} option (19) for the
+\texttt{aldex.clr} function, and then to use these values as before.
+This approach suggests a 5\% difference in scale, and is potentially
+less subject to user interpretation.
 
@@
 difference between groups and uncertainty level. Applying a per-group
-relative differential scale of 0.14 moves the housekeeping functions to
-the midline of no difference (Figure 3C, 14\% mean difference = -0.24,
-5\% difference = -0.34), and applying a gamma of 0.5 provides the same
-dispersion as in panel B of Figure 3. Note that now a significant number
-of functions are differentially up in BV that were formerly classed as
-not different without the full scale model (orange), or when only a
-default scale was applied. Inspection of the functions shows that these
-are largely missing from the \emph{Lactobacillus} species and so should
-actually be captured as differentially abundant in the BV group. The
-supplement shows that the using either the 5\% or the 14\% scale
+relative differential scale of 0.14 moves the housekeeping functions
+close to the midline of no difference (Figure 3C, assuming 14\% mean
+difference = -0.24, assuming a 5\% mean difference = -0.34), and
+applying a gamma of 0.5 provides the same dispersion as in panel B of
+Figure 3. Note that now a significant number of functions are
+differentially up in BV that were formerly classed as not different
+without the full scale model (orange), or when only a default scale was
+applied. Inspection of the functions shows that these are largely
+missing from the \emph{Lactobacillus} species and so should actually be
+captured as differentially abundant in the BV group. Supplementary
+Figure 4 shows that the using either the 5\% or the 14\% scale
 difference give imperceptibly different results suggesting that an
 informed scale model does not have to precisely estimate the scale
-difference to be useful. Nixon et al, (16) found that multiple
+difference to be useful. Nixon et al, (20) also found that multiple
 reasonable estimates for the \(\mu_n\) part of the informed scale model
@@
 afforded by an informed scale model, and the transferrability of it
-between sample cohorts of the same condition is outlined elsewhere (47).
+between sample cohorts of the same condition is outlined elsewhere (50).
 We suggest that the default scale model is sufficient when the data are
@@
 
-Biological systems are both predictably variable and stochastic (46) and
+Biological systems are both predictably variable and stochastic (49) and
 systems biology experiments show that there are transcripts with
 approximately constant concentrations in the cell and those with large
-variability under different growth conditions (20). Current measurement
+variability under different growth conditions (23). Current measurement
 methods that rely on high throughput sequencing fail to capture all of
-the variation, particularly variation due to scale (3, 16). In the
-absence of external information (4, 19, 52) sequencing depth
-normalisation methods cannot recapture the scale information (19), and
-can only normalize for the technical variation due to sequencing depth.
-Here we demonstrated that even approximate estimates of scale and the
-uncertainty of measuring it can aid in the interpretation of
-RNA-sequencing experiments.
-
-Nixon et al. (3) introduced the idea of explicitly modelling the scale
-of a HTS dataset, and showed how to incorporate these models in the
-analysis of microbiome and other datasets (16). They demonstrated that
+the variation, particularly variation due to scale (3, 20). In the
+absence of external information (5, 6, 55) sequencing depth
+normalisation methods cannot recapture the scale information (5, 21),
+and can only normalize for the technical variation due to sequencing
+depth. Here we demonstrated that even approximate estimates of the true
+system scale and the uncertainty of measuring it can aid in the
+interpretation of RNA-sequencing experiments.
+
+Nixon et al. (3) introduced the idea of explicitly modeling the scale of
+a HTS dataset, and showed how to incorporate these models in the
+analysis of microbiome and other datasets (20). They demonstrated that
 many tools commonly used to analyze HTS datasets had substantial Type 1
-and Type 2 error rates, in line with recent findings by others (9). A
-version of ALDEx2 with the ability to include scale uncertainty was
-shown to be able to correct for the high Type 1 error rate for that
+and Type 2 error rates, in line with recent findings by others (10, 12,
+13). A version of ALDEx2 with the ability to include scale uncertainty
+was shown to be able to correct for the high Type 1 error rate for that
 tool, albeit with some loss of sensitivity. Finally, they showed that
@@
 scale uncertainty estimates could both control for Type 1 and Type 2
-error rates (16).
-
-Building and using a scale model, thus has substantial benefits relative
+error rates (20).
+
+Building and using a scale model thus has substantial benefits relative
 to the dual cutoff approach that is advocated for many gene expression
-experiments (10, 12). In particular, the dual cutoff approach has long
-been known to not control for Type 1 errors (13, 14), and the frequent
-lack of concordance between tools when benchmarked on transcriptomes(7,
-9, 10, 26, 53) and microbiomes (6, 8, 28, 37, 54, 55) suggests poor
-control of Type 2 errors as well (7, 9). Thus, incorporating a scale
+experiments (14, 16). In particular, the dual cutoff approach has long
+been known to not control for Type 1 errors (17, 18), and the frequent
+lack of concordance between tools when benchmarked on transcriptomes(10,
+12--14, 29, 56) and microbiomes (9, 11, 31, 40, 57, 58) suggests poor
+control of Type 2 errors as well (10, 13). Thus, incorporating a scale
 model during the analysis of HTS data promises the best of both worlds.
@@
 additional computational overhead. Furthermore, this work and previous
-(16) show that even minimal information about the underlying environment
-can be used to build a relatively robust scale model that controls for
-both Type 1 and 2 error rates.
+(20) show that even minimal information about the underlying environment
+can be used to build a relatively robust informed scale model that
+controls for both Type 1 and 2 error rates.
 
 In the analysis of HTS data it is often observed that larger datasets
-converge on the majority of parts being significantly different (3, 9,
-10). Li et al. (9) conducted a permutation-based benchmarking study and
+converge on the majority of parts being significantly different (3, 13,
+14). Li et al. (13) conducted a permutation-based benchmarking study and
 found that widely used tools performed worse than simple Wilcoxon
@@
 They suggested that the presence of outliers were one of the factors
-driving this observation. Brooks et al. (56) suggested that
+driving this observation. Brooks et al. (59) suggested that
 inappropriate choice of benchmarking methods are also a major
@@
 explained by the observation that different analytic approaches produce
-different parameter estimates. Thus, more data produces worse estimates
-because the additional data simply increases the precision of a flawed
-estimate (3, 57).
-
-Scale simulation is now build into ALDEx2 (16) and this work suggest
-that there are two main root causes to common HTS data pathologies. The
-first contributing factor is the observed very low dispersion estimate
-for many features that is a by-product of experimental design,
-sequencing and normalization. In the Schurch et al. (10) dataset, the
-data were from single colonies derived from a single culture. Thus, it
-is more accurate to describe the 96 samples as wet-lab technical
-replicates rather than independent samples. This type of replication
-approach is standard in the molecular literature, and would be expected
-to result in the very low dispersion that is observed. Applying the
-default scale model with \(\gamma=0.5\) a large number of transcripts
-have their dispersion increased (Figure 1D), with the effect being
-largest for those with the lowest initial dispersion (Figure 2). Adding
-scale uncertainty results in modest number of transcripts, 205, being
-called significantly different as shown in the volcano plot in Figure 1B
-(red points). In addition, there is now a strong concordance between the
-difference and p-values. In hindsight, what is not obvious is why the
-unscaled volcano plot shows such poor correspondence. We suggest that
-this can be explained by random fluctuations in the many variance
-estimates with very low values, and this is supported by the plots shown
-in Figure 2.
+different parameter estimates for location or scale or for both. Thus,
+more data produces worse estimates because the additional data simply
+increases the precision of a flawed estimate (3, 60).
+
+Scale simulation is now built into ALDEx2 (20) and here we suggest that
+there are two main root causes to common HTS data pathologies. The first
+contributing factor is the observed very low dispersion estimate for
+many features that is a by-product of some experimental designs and of
+normalization. In the Schurch et al. (14 dataset), the data were from
+single colonies derived from a single culture. Thus, it is more accurate
+to describe the 96 samples as wet-lab technical replicates rather than
+independent samples. This type of replication approach is standard in
+the molecular literature, and would be expected to result in the very
+low dispersion that is observed. Applying the default scale model with
+\(\gamma=0.5\) a large number of transcripts have their dispersion
+increased (Figure 1D), with the effect being largest for those with the
+lowest initial dispersion (Figure 2). Adding scale uncertainty results
+in modest number of transcripts, 205, being called significantly
+different as shown in the volcano plot in Figure 1B (red points). In
+addition, there is now a strong concordance between the difference and
+q-values. In hindsight, it is not obvious why the unscaled volcano plot
+shows such poor correspondence. We suggest that this is explained by
+random fluctuations of the many very low variance estimates and this is
+supported by the plots shown in Figure 2.
 
 The second contributing factor is unacknowledged asymmetry in many
-datasets (15); i.e., different gene content or a directional change in
+datasets (19); i.e., different gene content or a directional change in
 the majority of features. In the case of asymmetry, the use of a
@@
 groups and found that any reasonable estimate is an improvement over the
-default scale model. This is in line with the observations by Nixon et
-al (16) in a 16S rRNA gene sequencing dataset. It is also of note that
-in the case of true biological replicates (different individuals) that
-adding a modest amount of scale \(\gamma=0.5\) had little effect on the
-the difference between groups and on the dispersion. Thus, in this
-dataset the scale mis-specification was affecting mainly the location of
-the difference between groups. While we acknowledge that some prior
-information on which housekeeping transcripts should not be classed as
-differentially abundant is needed, we suggest that this information is
-widely available and is already used when performing the gold-standard
-quantitative PCR test of differential abundance (58, 59).
+naive approach and also over the default scale model. This is in line
+with the observations by Nixon et al (20) in a 16S rRNA gene sequencing
+dataset. It is also of note that in the case of true biological
+replicates (different individuals) that adding a modest amount of scale
+\(\gamma=0.5\) had little effect on the the difference between groups
+and on the dispersion. Thus, in this dataset the scale mis-specification
+was affecting mainly the location of the difference between groups.
+While we acknowledge that some prior information on which housekeeping
+transcripts should not be classed as differentially abundant is needed,
+we suggest that this information is widely available and is already used
+when performing the gold-standard quantitative PCR test of differential
+abundance (61, 62).
 
@@
 (3) and so can replace the consensus approach that has been proposed by
-some groups (8, 60) with no additional computational overhead. Thus, an
+some groups (11, 63) with no additional computational overhead. Thus, an
 advantage of incorporating scale is that analyses can be made much more
@@
 assumptions and a detailed description of these points is describe
-elsewhere (3).
+elsewhere (3,).
 
@@
 when the majority of features are asymmetrically distributed or
-expressed (or both) in the groups (47). Thus, using and reporting scale
+expressed (or both) in the groups (50). Thus, using and reporting scale
 uncertainty should become a standard practice in the analysis of HTS
@@
 
-\bibitem[\citeproctext]{ref-nixon2023scale}
-3. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
-Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
-reliant inference}. \emph{https://arxiv.org/abs/2201.03616}.
+\bibitem[\citeproctext]{ref-nixon2024scale}
+3. Nixon,M.P., McGovern,K.C., Letourneau,J., David,L.A., Lazar,N.A.,
+Mukherjee,S. and Silverman,J.D. (2024)
+\href{https://arxiv.org/abs/2201.03616}{Scale reliant inference}.
+
+\bibitem[\citeproctext]{ref-Jiang:2011aa}
+4. Jiang,L., Schlesinger,F., Davis,C.A., Zhang,Y., Li,R., Salit,M.,
+Gingeras,T.R. and Oliver,B. (2011)
+\href{https://doi.org/10.1101/gr.121095.111}{Synthetic spike-in
+standards for RNA-seq experiments}. \emph{Genome Res}, \textbf{21},
+1543--51.
+
+\bibitem[\citeproctext]{ref-Loven:2012aa}
+5. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B., Burge,C.B.,
+Levens,D.L., Lee,T.I. and Young,R.A. (2012)
+\href{https://doi.org/10.1016/j.cell.2012.10.012}{Revisiting global gene
+expression analysis}. \emph{Cell}, \textbf{151}, 476--82.
 
 \bibitem[\citeproctext]{ref-Vandeputte:2017aa}
-4. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
+6. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
 Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
@@
 \bibitem[\citeproctext]{ref-Props:2017aa}
-5. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
+7. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
 Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
@@
 
+\bibitem[\citeproctext]{ref-Nishijima:2024aa}
+8. Nishijima,S., Stankevic,E., Aasmets,O., Schmidt,T.S.B., Nagata,N.,
+Keller,M.I., Ferretti,P., Juel,H.B., Fullam,A., Robbani,S.M., \emph{et
+al.} (2024) Fecal microbial load is a major determinant of gut
+microbiome variation and a confounder for disease associations.
+\emph{Cell},
+\href{https://doi.org/10.1016/j.cell.2024.10.022}{10.1016/j.cell.2024.10.022}.
+
 \bibitem[\citeproctext]{ref-Thorsen:2016aa}
-6. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
+9. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
 Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
@@
 \bibitem[\citeproctext]{ref-Quinn:2018aa}
-7. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
+10. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
 \href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
@@
 \bibitem[\citeproctext]{ref-Nearing:2022aa}
-8. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
+11. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
 Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
@@
 
+\bibitem[\citeproctext]{ref-Ge:2021aa}
+12. Ge,X., Chen,Y.E., Song,D., McDermott,M., Woyshner,K.,
+Manousopoulou,A., Wang,N., Li,W., Wang,L.D. and Li,J.J. (2021)
+\href{https://doi.org/10.1186/s13059-021-02506-9}{Clipper: P-value-free
+FDR control on high-throughput data from two conditions}. \emph{Genome
+Biol}, \textbf{22}, 288.
+
 \bibitem[\citeproctext]{ref-Li:2022aa}
-9. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+13. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
 \href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
@@
 \bibitem[\citeproctext]{ref-Schurch:2016aa}
-10. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
+14. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-storey2003positive}
-11. Storey,J.D. (2003) The positive false discovery rate: A bayesian
+15. Storey,J.D. (2003) The positive false discovery rate: A bayesian
 interpretation and the q-value. \emph{The annals of statistics},
@@
 \bibitem[\citeproctext]{ref-Cui:2003aa}
-12. Cui,X. and Churchill,G.A. (2003)
+16. Cui,X. and Churchill,G.A. (2003)
 \href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
@@
 \bibitem[\citeproctext]{ref-Zhang:2009aa}
-13. Zhang,S. and Cao,J. (2009)
+17. Zhang,S. and Cao,J. (2009)
 \href{https://doi.org/10.1186/1471-2105-10-402}{A close examination of
@@
 \bibitem[\citeproctext]{ref-Ebrahimpoor:2021aa}
-14. Ebrahimpoor,M. and Goeman,J.J. (2021)
+18. Ebrahimpoor,M. and Goeman,J.J. (2021)
 \href{https://doi.org/10.1093/bib/bbab053}{Inflated false discovery rate
@@
 \bibitem[\citeproctext]{ref-Wu2021}
-15. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
+19. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
 \href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
@@
 \bibitem[\citeproctext]{ref-Nixon2024B}
-16. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
+20. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
 normalization: Incorporating scale uncertainty in microbiome and gene
@@
 \bibitem[\citeproctext]{ref-Lovell:2015}
-17. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
+21. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
 Bähler,J. (2015)
@@
 \bibitem[\citeproctext]{ref-Nie:2012aa}
-18. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
+22. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
 Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
@@
 
-\bibitem[\citeproctext]{ref-Loven:2012aa}
-19. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
-Burge,C.B., Levens,D.L., Lee,T.I. and Young,R.A. (2012)
-\href{https://doi.org/10.1016/j.cell.2012.10.012}{Revisiting global gene
-expression analysis}. \emph{Cell}, \textbf{151}, 476--82.
-
 \bibitem[\citeproctext]{ref-Scott:2010}
-20. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
+23. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
 \href{https://doi.org/10.1126/science.1192588}{Interdependence of cell
@@
 \bibitem[\citeproctext]{ref-Yoshikawa:2011aa}
-21. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
+24. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
 Shimizu,H. (2011) \href{https://doi.org/10.1002/yea.1843}{Comprehensive
@@
 \bibitem[\citeproctext]{ref-Lin:2018aa}
-22. Lin,J. and Amir,A. (2018)
+25. Lin,J. and Amir,A. (2018)
 \href{https://doi.org/10.1038/s41467-018-06714-z}{Homeostasis of protein
@@
 \bibitem[\citeproctext]{ref-Zozaya:2010}
-23. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
+26. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
 \href{https://doi.org/10.1128/JCM.00851-09}{Quantitative PCR assessments
@@
 \bibitem[\citeproctext]{ref-Ravel:2010}
-24. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
+27. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
 McCulle,S.L., Karlebach,S., Gorle,R., Russell,J., Tacket,C.O., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hummelen:2010}
-25. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
+28. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
 Changalucha,J., Gloor,G.B. and Reid,G. (2010)
@@
 \bibitem[\citeproctext]{ref-Bullard:2010}
-26. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
+29. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
 \href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
@@
 \bibitem[\citeproctext]{ref-Dillies:2013}
-27. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
+30. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
 Jeanmougin,M., Servant,N., Keime,C., Marot,G., Castel,D., Estelle,J.,
@@
 \bibitem[\citeproctext]{ref-Weiss:2017aa}
-28. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
+31. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
 Lozupone,C., Zaneveld,J.R., Vázquez-Baeza,Y., Birmingham,A., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hughes:2005tu}
-29. Hughes,J.B. and Hellmann,J.J. (2005)
+32. Hughes,J.B. and Hellmann,J.J. (2005)
 \href{https://doi.org/10.1016/S0076-6879(05)97017-1}{The application of
@@
 \bibitem[\citeproctext]{ref-Mortazavi:2008}
-30. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
+33. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
 (2008) \href{https://doi.org/10.1038/nmeth.1226}{Mapping and quantifying
@@
 \bibitem[\citeproctext]{ref-wagner:tpm}
-31. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
+34. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
 \href{https://doi.org/10.1007/s12064-012-0162-3}{Measurement of mRNA
@@
 \bibitem[\citeproctext]{ref-Robinson:2010a}
-32. Robinson,M.D. and Oshlack,A. (2010)
+35. Robinson,M.D. and Oshlack,A. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-3-r25}{A scaling normalization
@@
 \bibitem[\citeproctext]{ref-aitchison1982}
-33. Aitchison,J. (1982) The statistical analysis of compositional data.
+36. Aitchison,J. (1982) The statistical analysis of compositional data.
 \emph{Journal of the Royal Statistical Society: Series B
@@
 \bibitem[\citeproctext]{ref-Anders:2010}
-34. Anders,S. and Huber,W. (2010)
+37. Anders,S. and Huber,W. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-10-r106}{Differential
@@
 \bibitem[\citeproctext]{ref-Aitchison:1986}
-35. Aitchison,J. (1986) The statistical analysis of compositional data
+38. Aitchison,J. (1986) The statistical analysis of compositional data
 Chapman \& Hall, London, England.
@@
 \bibitem[\citeproctext]{ref-fernandes:2013}
-36. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
+39. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
 (2013) \href{https://doi.org/10.1371/journal.pone.0067019}{ANOVA-like
@@
 \bibitem[\citeproctext]{ref-Yerke:2024aa}
-37. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
+40. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
 \href{https://doi.org/10.1186/s40168-023-01747-z}{Proportion-based
@@
 \bibitem[\citeproctext]{ref-Love:2014aa}
-38. Love,M.I., Huber,W. and Anders,S. (2014)
+41. Love,M.I., Huber,W. and Anders,S. (2014)
 \href{https://doi.org/10.1186/s13059-014-0550-8}{Moderated estimation of
@@
 \bibitem[\citeproctext]{ref-Paulson:2013aa}
-39. Paulson,J.N., Stine,O.C., Bravo,H.C. and Pop,M. (2013)
+42. Paulson,J.N., Stine,O.C., Bravo,H.C. and Pop,M. (2013)
 \href{https://doi.org/10.1038/nmeth.2658}{Differential abundance
@@
 \bibitem[\citeproctext]{ref-fernandes:2014}
-40. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
+43. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
 Edgell,D.R. and Gloor,G.B. (2014)
@@
 \bibitem[\citeproctext]{ref-Gierlinski:2015aa}
-41. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
+44. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-benjamini:1995}
-42. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
+45. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
 rate: A practical and powerful approach to multiple testing.
@@
 \bibitem[\citeproctext]{ref-efron2008FDR}
-43. Efron,B. (2008) Microarrays, empirical bayes and the two-groups
+46. Efron,B. (2008) Microarrays, empirical bayes and the two-groups
 model. \emph{Statist. Sci.}, \textbf{23}, 1--22.
@@
 \bibitem[\citeproctext]{ref-gloor:effect}
-44. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
+47. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
 \href{https://doi.org/10.1080/10618600.2015.1131161}{Displaying
@@
 \bibitem[\citeproctext]{ref-Rocha:2020aa}
-45. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
+48. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
 (2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
@@
 \bibitem[\citeproctext]{ref-Taniguchi:2010aa}
-46. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
+49. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
 Emili,A. and Xie,X.S. (2010)
@@
 \bibitem[\citeproctext]{ref-dosSantos:2024}
-47. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
+50. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
 Gloor,G.B. (2024) Vaginal metatranscriptome meta-analysis reveals
@@
 \bibitem[\citeproctext]{ref-macklaim:2013}
-48. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
+51. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
 and Gloor,G.B. (2013)
@@
 \bibitem[\citeproctext]{ref-Denge00262-18}
-49. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
+52. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
 Wagner-Döbler,I. (2018)
@@
 \bibitem[\citeproctext]{ref-Fettweis:2019aa}
-50. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
+53. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
 Parikh,H.I., Huang,B., Arodz,T.J., Edupuganti,L., Glascock,A.L.,
@@
 \bibitem[\citeproctext]{ref-Okuda:2008}
-51. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
+54. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
 Goto,S. and Kanehisa,M. (2008)
@@
 \bibitem[\citeproctext]{ref-yeast-absolute}
-52. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
+55. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
 Bähler,J. (2012)
@@
 \bibitem[\citeproctext]{ref-Soneson:2013}
-53. Soneson,C. and Delorenzi,M. (2013)
+56. Soneson,C. and Delorenzi,M. (2013)
 \href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
@@
 \bibitem[\citeproctext]{ref-McMurdie:2014a}
-54. McMurdie,P.J. and Holmes,S. (2014)
+57. McMurdie,P.J. and Holmes,S. (2014)
 \href{https://doi.org/10.1371/journal.pcbi.1003531}{Waste not, want not:
@@
 \bibitem[\citeproctext]{ref-hawinkel2017}
-55. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
+58. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
 \href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
 Microbiome differential abundance methods do not control the false
-discovery rate}. \emph{BRIEFINGS IN BIOINFORMATICS}.
+discovery rate}. \emph{Briefings in Bioinformatics}.
 
 \bibitem[\citeproctext]{ref-Brooks:2024aa}
-56. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
+59. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
 \href{https://doi.org/10.1038/s41576-023-00679-6}{Challenges and best
@@
 \bibitem[\citeproctext]{ref-gustafson2015bayesian}
-57. Gustafson,P. (2015) Bayesian inference for partially identified
+60. Gustafson,P. (2015) Bayesian inference for partially identified
 models: Exploring the limits of limited data CRC Press.
@@
 \bibitem[\citeproctext]{ref-Thellin:1999aa}
-58. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
+61. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
 Hennen,G., Grisar,T., Igout,A. and Heinen,E. (1999)
@@
 \bibitem[\citeproctext]{ref-SEQCux2fMAQC-III-Consortium:2014aa}
-59. SEQC/MAQC-III Consortium (2014)
+62. SEQC/MAQC-III Consortium (2014)
 \href{https://doi.org/10.1038/nbt.2957}{A comprehensive assessment of
@@
 \bibitem[\citeproctext]{ref-Song:2023aa}
-60. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
+63. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
 Klatt,N.R., Hensley-McBain,T. and Wu,M.C. (2023)

### 2025-03-11 information-supplement.Rmd
submitted version Dec 6

--- 2024-05-16 supplement_figs.Rmd
@@
 ---
-title: "Supplement: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
-"
-shorttitle: "Scale ALDEx2 Supp"
+title: "In high throughput sequencing all normalizations are wrong but some are often useful"
+shorttitle: "Scale explains normalization"
 author:
-- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
+- name: Greg Gloor, Justin Silverman
   affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
   email: ggloor@uwo.ca
+citeproc: TRUE
+cite-method: citeproc
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
+csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
+pdf-engine: latex
+  \usepackage{amsmath}
 output:
-  BiocStyle::pdf_document: default
-package: ALDEx2 1.33.1
-abstract: |
-    Introduction to scale simulation and FDR correction with ALDEx2.
-date: 21 August 2023
-link-citations: true
+  pdf_document
 ---
+
+```{r echo=F, warning=F, message=F,comment=F}
+# DESeq2
+load('analysis/res.Rda')
+library(zCompositions)
+#ALDEx2
+library(ALDEx2)
+data(selex)
+load(file="analysis/yst.all.Rda")
+load(file="analysis/yst.s.all.Rda")
+load(file="analysis/yst.1.all.Rda")
+# with gamma = 0.5
+load('analysis/x.s.all.Rda')
+
+sig.des <- which(res@listData$padj < 0.05)
+sig.ald <- which(yst.all$we.eBH < 0.05)
+sig.s.ald <- which(yst.s.all$we.eBH < 0.05)
+
+sig.all <- yst.all$we.eBH < 0.05
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
+sig.s <- yst.s.all$we.eBH < 0.05
+sig.1 <- yst.1.all$we.eBH < 0.05
+
+tsc <- read.table("~/Documents/0_git/datasets/transcriptome.tsv", header=T, row.names=1, sep="\t")
+```
 
@@
   x.p <- x.vec/sum(x.vec)
+  x.p <- x.p[! x.p %in% 0]
   return(-1 * sum(x.p * log2(x.p)))
@@
 # note mean is only used when 
-HvG <- function(n.random, n.parts, test='U', mean=30){
+HvG <- function(n.random, n.parts, test='U', mean=30, sd=1){
 
@@
     for(j in 1:n.random){	 
-     if(test=='N'){ d <- rnorm(val[i], mean=mean) }
+     if(test=='N'){ d <- rnorm(val[i], mean=mean, sd=sd) }
      else if (test == 'U'){ d <- runif(val[i]) * mean }
@@
 
-HvG.plot <- function(HvG.out){
+HvG.plot <- function(HvG.out, points=FALSE){
 
@@
    xlab='G', ylab='H')
-
+if(points==TRUE){
+  points(HvG.out$mat.log, HvG.out$mat.h, pch=19, cex=0.5, col=rgb(1,0,0,0.2))
+}
 #abline(lm((HvG.out$mat.log[,2]-HvG.out$mat.h[,2]) ~ val), col='red')
@@
 
+```{r echo=F, eval=F}
+prior <- 0:10/10
+matG <- matrix(data=NA, ncol=14, nrow=11)
+matH <- matrix(data=NA, ncol=14, nrow=11)
+for(i in 1:11){
+  matG[i,] <- apply(selex+prior[i], 2, G)
+  matH[i,] <- apply(selex+prior[i], 2, H)
+}
+
+plot(prior, matH[,1]-matH[,8], ylim=c(6,8), pch=19, col=rgb(0,0,1,0.3))
+points(prior, matG[,1]-matG[,8], pch=19, col=rgb(1,0,1,0.3))
+abline(lm(matG[2:11,1]-matG[2:11,8] ~ prior[2:11]), col='red')
+abline(lm(matH[2:11,1]-matH[2:11,8] ~ prior[2:11]), col='blue')
+for(i in 2:7){ 
+  points(prior, matH[,i]-matH[,i+7], pch=19, col=rgb(0,0,1,0.3)) 
+  abline(lm(matH[2:11,i]-matH[2:11,i+7] ~ prior[2:11]), col='blue', lty=i)
+}
+for(i in 2:7){ 
+  points(prior, matG[,i]-matG[,i+7], pch=19, col=rgb(1,0,0,0.3)) 
+  abline(lm(matG[2:11,i]-matG[2:11,i+7] ~ prior[2:11]), col='red', lty=i)
+}
+
+```
+# Scale in high throughput sequencing
+
+High throughput sequencing generates  data where the total number of reads is not informative about the environment.
+These data are compositional, that is, the data is constrained to a fixed upper limit and can be represented by their corresponding proportions (or probabilities) without losing any information.
+These data must be normalized and many of the normalizations in widespread use can be represented as ratios, further the normalized data are almost universally log-transformed prior to analysis.
+Thus, all of the aforementioned normalizations are  actually log-ratios of some sort.
+There has been much debate in the literature, and much misunderstanding by the user base as to which normalization is the best because until now the actual assumptions of those normalizations could not be understood inside a common framework.
+
+Recently, Silverman and colleagues proved that log-ratios could be understood as  implicit assumptions about the actual scale (size) of the environment.
+Informally, they recognized that an underlying environment \(W\) containing \(D\) parts in \(N\) samples  could be represented by a count matrix decomposed into proportional (p) and total (t) sub-matrices for each sample \(n\)
+
+\begin{equation}
+ W_{dn} = W_{dn}^p \cdot  W_n^t
+ \label{envt}
+\end{equation}
+
+or the logarithmic form
+
+\begin{equation}
+ \log W_{dn} = \log W_{dn}^p  + \log W_n^t,
+ \label{logenvt}
+\end{equation}
+
+where \begin{equation} W_n^t=\sum{W_{1 \ldots D, n}} \label{sumenvt}\end{equation}  and \begin{equation}W_{dn}^p = \frac{W_{dn}}{ W_n^t} . \label{propenvt}\end{equation}
+
+The logarithmic form in equation 2 is strikingly similar to the logarithmic formula for the centred log-ratio (CLR) transformation used in compositional data analysis (CoDa).
+Here the observed data matrix  \(Y\) can be transformed by the CLR
+\footnote{The formula for the CLR gives the same answer if the matrix  \(Y\) is the raw count values or if its values have been converted to proportions, or indeed by any other linear transformation.
+This is the scale-invariance property of CoDa in action.}
+  by a taking the log-ratio of each part with the sample geometric mean which we will denote as \(h\):
+
+\begin{equation}
+ CLR(Y_{dn}) = \log Y_{dn}  + h_n,
+ \label{clr}
+\end{equation}
+
+where  \begin{equation} h_n = - \frac{1}{D} \sum log Y_{1 \dots D, n}. \label{geometric}\end{equation}
+
+In other words, if \(Y\) is a sample of \(W\), then  \(h_n\) is an implicit estimate of the scale  \( \log W_n^t\).
+Of course, just because the equations are similar does not guarantee that the values estimated are correct.
+They went on to show that in the rare instance when the geometric mean was a good estimate of the scale then analyses were trustworthy, but more commonly that the geometric mean was a biased scale estimator of of the data.
+Nixon et al made two further important  conclusions: first,  that any estimate that used the logarithm of a ratio approach was also estimating the scale; and second, that the only the relative scale was important.
+We will discuss the implications of these conclusions in a later section.
+
+All CoDa based approaches are constrained to real vectors  greater than 0
+
+\begin{equation}
+X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x > 0, \sum_X = 1.
+\end{equation}
+
+Despite this limitation some advantages of the CoDa framework are that it has true distance metrics and it has a solid geometric interpretation.
+The usual ways of dealing with 0 values in CoDa approaches are to either impute 0 values, to add a prior to all values, or to restrict the analysis to the parts that include only non-0 values.
+All these methods restrict the generality of CoDa methods and are rightly pointed out as being weaknesses of the approaches.
+
+Information theory is another framework in which proportional vectors can be examined.
+Information theory  developed as a way to examine the amount of information needed to encode, transmit and decode information.
+Thus the concept of size was build in from the beginning and this has been one of the most useful advances in all of computation and statistics.
+The domain of information theory is subtly different from CoDa in that 0 values are permitted:
+
+\begin{equation}
+X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x \ge 0, \sum_X = 1.
+\end{equation}
+
+ In information theory, the elemental amount of information or surprisal for \(x_i\) is the inverse of the logarithm of the elemental probability [@rezaInfoIntro].
+This measure is often called self-information or the surprisal and we will denote this as \(I\):
+
+\begin{equation}
+I(x_i) = \log_2(\frac{1}{x_i}) = -\log_2{x_i}.
+\end{equation}
+
+Logarithms of probabilities are additive so that for multiple independent probabilities the entropy \(H\) which is the expected information content of the random variable \(x\) is
+
+\begin{equation}
+H= \sum_{X} x_i I(x_i) = -\sum_{X} x_i\log_2(x_i).
+\end{equation}
+
+In this formula Shannon defined  \(x_i\log_2(x_i) = 0, x_i=0\) and this effectively removed the 0 constraint when using information theory.
+Less well known is that a sample average entropy was also defined
+
+\begin{equation}
+h = - \frac{1}{D}\sum log_2(x_i),
+\end{equation}
+
+and this is simply the geometric mean.
+Usefully, \(h\) measure converges on \(H\) as the number of measurements from a "typical subset" of the data approaches infinity; this is the Asymptotic Equipartition Property of information as described in [@mackay2003information;@wilde_2017].
+As shown in Chapter 4 of Mackay [@mackay2003information] the typical subset is approximately Gaussian and so if both \(h\) and \(H\) are drawn from this distribution they should be identical in the limit.
+Another way to think about this is that the two measures converge as the distribution of \(p(x_i\) approaches the uniform \(\frac{1}{N}\).
+Thus the relationship between \(H\) and \(h\) demonstrates independently that the log of the geometric mean can be interpreted as a size of a probability vector.
+
+One the challenges of the scale-based approach has been to identify appropriate base scale values for informed scale models.
+It has been noted before that \(h\) can be a good or a poor initial estimate.
+Recall that the actual scale value is not important, but that the ratio between scale values in each group is important (or the difference in log(scale) values).
+Figure 1 shows that one potential confounder is that the value of \(h\) is very sensitive to small changes in initial assumptions when the data are sparse.
+For example, one method of treating 0 values is to add a small prior to each value in the matrix.
+Figure 1 shows how \(H\) and \(h\) vary for values of this prior in the range of 0.01 to 1.
+Note that entropy is essentially not affected by these prior values, but that  \(h\) can produce a very different initial estimate of the scale with different priors.
+Another method of treating 0s is to replace the 0 values with a pseudocount, this approach  produces a result that is similar to adding a relatively large prior in sparse data and similar to adding a small prior in non-sparse data.
+A final method is to impute 0 values using a CoDa-based approach.
+This method
+
+As shown in Figure 1, one reason for this is that the geometric mean cannot be calculated when one or more of the entries in the probability vector is 0.
+Thus, Furthermore, the tight relationship between can be exploited
+
+```{r zeroh, echo=F, fig.cap="Plots of H and h with different values added to 0 count parts.
+The non-sparse plot shows that H and h are linearly related and have a very narrow range over when samples contain no 0 values.
+The sparse plot shows that H still varies in a small range (although it is about 10-fold larger) but that h now varies has through an approximate 8-fold range when 0 replacement values between 0.1 and 1.
+The outlier point is a 0 replacement value of 0.01.
+The difference plot is the mean difference in either h or H between the non-sparse and sparse samples in the dataset.
+"}
+
+### inpiration can pass function outputs (even nested) as parameters
+# ald.ys.h <- aldex(tsc, conditions=y.co, gamma=aldex.makeScaleMatrix(0.5, mu=apply(tsc, 2, function(x)-H(x)), conditions=y.co))
+# x.ys.h <- aldex.clr(tsc, conds=y.co, gamma=aldex.makeScaleMatrix(0.5, mu=apply(tsc, 2, function(x)-H(x)), conditions=y.co))
+
+## SELEX
+prior <- c(0:10) / 10
+p.out <- matrix(data=NA, nrow=14, ncol=6)
+prior[1] <- 0.001
+for(i in 1:length(prior)){
+p.out[i,1] <- -G(selex[,1]+prior[i])
+p.out[i,2] <- -G(selex[,8]+prior[i])
+p.out[i,3] <- H(selex[,1]+prior[i])
+p.out[i,4] <- H(selex[,8]+prior[i])
+Gx <- -(apply(selex+prior[i], 2, G))
+p.out[i,5] <- mean(Gx[1:7]-Gx[8:14])
+Hx <- -(apply(selex+prior[i], 2, H))
+p.out[i,6] <- mean(Hx[1:7]-Hx[8:14])
+}
+selrp1 <- selex 
+selrp1[selrp1==0] <- 1
+p.out[12,1] <- -G(selrp1[,1])
+p.out[12,2] <- -G(selrp1[,8])
+p.out[12,3] <- H(selrp1[,1])
+p.out[12,4] <- H(selrp1[,8])
+Gx <- -(apply(selrp1, 2, G))
+p.out[12,5] <- mean(Gx[1:7]-Gx[8:14])
+Hx <- -(apply(selrp1, 2, H))
+p.out[12,6] <- mean(Hx[1:7]-Hx[8:14])
+
+sel.GBM <- t(cmultRepl(t(selex), label=0, method="GBM"))
+p.out[13,1] <- -G(sel.GBM[,1])
+p.out[13,2] <- -G(sel.GBM[,8])
+p.out[13,3] <- H(sel.GBM[,1])
+p.out[13,4] <- H(sel.GBM[,8])
+Gx <- -(apply(sel.GBM, 2, G))
+p.out[13,5] <- mean(Gx[1:6]-Gx[8:13])
+Hx <- -(apply(sel.GBM, 2, H))
+p.out[13,6] <- mean(Hx[1:6]-Hx[8:13])
+
+sel.CZM <- t(cmultRepl(t(selex), label=0, method="CZM"))
+p.out[14,1] <- -G(sel.CZM[,1])
+p.out[14,2] <- -G(sel.CZM[,8])
+p.out[14,3] <- H(sel.CZM[,1])
+p.out[14,4] <- H(sel.CZM[,8])
+Gx <- -(apply(sel.CZM, 2, G))
+p.out[14,5] <- mean(Gx[1:6]-Gx[8:13])
+Hx <- -(apply(sel.CZM, 2, H))
+p.out[14,6] <- mean(Hx[1:6]-Hx[8:13])
+
+cex=c(rep(0.6,11), 1.5,1.5,1.5)
+pch=c(rep(0,12), 1,2)
+par(mfrow=c(1,3))
+plot(p.out[,1], p.out[,3], xlab="h", ylab="H", main="nonsparse", cex=cex, pch=pch)
+plot(p.out[,2], p.out[,4], xlab="h", ylab="H", main="sparse", cex=cex, pch=pch)
+stripchart(list(p.out[,5], p.out[,6]), method="jitter", vertical=TRUE, main="nonsparse - sparse", group.names=c('h diff', 'H diff') ,cex=cex, pch=pch)
+
+#TSCOME
+prior <- c(0:10) / 10
+t.out <- matrix(data=NA, nrow=12, ncol=7)
+colnames(t.out) <- c("h1", "h2", "H1",H2","mh", 'mH','prior')
+prior[1] <- 0.001
+for(i in 1:length(prior)){
+t.out[i,1] <- -G(tsc[,1]+prior[i])
+t.out[i,2] <- -G(tsc[,49]+prior[i])
+t.out[i,3] <- H(tsc[,1]+prior[i])
+t.out[i,4] <- H(tsc[,49]+prior[i])
+Gx <- -(apply(tsc+prior[i], 2, G))
+t.out[i,5] <- mean(Gx[1:48]-Gx[49:96])
+Hx <- -(apply(tsc+prior[i], 2, H))
+t.out[i,6] <- mean(Hx[1:48]-Hx[49:96])
+t.out[i,7] <- prior[i]
+}
+selrp1 <- tsc 
+selrp1[selrp1==0] <- 1
+t.out[12,1] <- -G(selrp1[,1])
+t.out[12,2] <- -G(selrp1[,49])
+t.out[12,3] <- H(selrp1[,1])
+t.out[12,4] <- H(selrp1[,49])
+Gx <- -(apply(selrp1, 2, G))
+t.out[12,5] <- mean(Gx[1:48]-Gx[49:96])
+Hx <- -(apply(selrp1, 2, H))
+t.out[12,6] <- mean(Hx[1:48]-Hx[49:96])
+t.out[12,7] <- 1
+
+par(mfrow=c(1,3))
+plot(t.out[,1], t.out[,3], xlab="h", ylab="H", main="nonsparse", cex=c(rep(0.6,11), 1.5))
+plot(t.out[,2], t.out[,4], xlab="h", ylab="H", main="sparse", cex=c(rep(0.6,11), 1.5))
+stripchart(list(t.out[,5], t.out[,6]), method="jitter", vertical=TRUE, main="nonsparse - sparse", group.names=c('h diff', 'H diff') ,cex=c(rep(0.6,11), 1.5))
+
+### SS
+
+url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
+ss <- read.table(url, header=T, row.names=1, sep='\t')
+ss <- ss[,c(1:100,1502:1601)]
+
+prior <- c(0:10) / 10
+ss.out <- matrix(data=NA, nrow=12, ncol=6)
+prior[1] <- 0.001
+for(i in 1:length(prior)){
+ss.out[i,1] <- -G(ss[,1]+prior[i])
+ss.out[i,2] <- -G(ss[,101]+prior[i])
+ss.out[i,3] <- H(ss[,1]+prior[i])
+ss.out[i,4] <- H(ss[,101]+prior[i])
+Gx <- -(apply(ss+prior[i], 2, G))
+ss.out[i,5] <- mean(Gx[1:100]-Gx[101:200])
+Hx <- -(apply(ss+prior[i], 2, H))
+ss.out[i,6] <- mean(Hx[1:100]-Hx[101:200])
+}
+ssrp1 <- ss 
+ssrp1[ssrp1==0] <- 1
+ss.out[12,1] <- -G(ssrp1[,1])
+ss.out[12,2] <- -G(ssrp1[,101])
+ss.out[12,3] <- H(ssrp1[,1])
+ss.out[12,4] <- H(ssrp1[,101])
+Gx <- -(apply(ssrp1, 2, G))
+ss.out[12,5] <- mean(Gx[1:100]-Gx[101:200])
+Hx <- -(apply(ssrp1, 2, H))
+ss.out[12,6] <- mean(Hx[1:100]-Hx[101:200])
+
+par(mfrow=c(1,3))
+plot(ss.out[,1], ss.out[,3], xlab="h", ylab="H", main="nonsparse", cex=c(rep(0.6,11), 1.5))
+plot(ss.out[,2], ss.out[,4], xlab="h", ylab="H", main="sparse", cex=c(rep(0.6,11), 1.5))
+stripchart(list(ss.out[,5], ss.out[,6]), method="jitter", vertical=TRUE, main="nonsparse - sparse", group.names=c('h diff', 'H diff') ,cex=c(rep(0.6,11), 1.5))
+
+```
+
+Many such normalizations have been developed and are in widespread use.
+These include simple proportions, relative log expression (RLE), trimmed mean of M values (TMM), cumulative sum scaling (CSS), LVHA, IQLR and others.
+In the case of proportions the assumption being made is that the total sum of \(Y_n\) is the same (or at least directly proportional to) the total sum of \(W_n\).
+All the other normalizations make the assumption that a subset of the parts in \(Y_n\) are the same (or at least directly proportional to) the same subset of parts in \(W_n\).
+Thus, while the actual details of how the normalization is conducted vary, the end result is always the same: analyses are conducted with a log of a ratio that can be expressed as a function of the observed data
+
+\begin{equation}
+N(Y_{dn}) = \log Y_{dn}  + f(\cdot n),
+\label{function} 
+\end{equation}
+
+where \(\cdot n\) represents the whole or a subset of the parts in sample \(n\).
+
+Returning to equation \eqref{function} we are now in a position to think about  how the concept of scale allows us to understand the assumptions made by different normalizations and to understand how scale normalization supersedes these individual normalizations, including the CLR.
+
+The simplest normalization is the proportion calculated as in equation \eqref{propenvt}.
+For proportions the values in the output all sum to 1, and so it  is easy to see that the assumption being made is that the scale of all the \(n\) samples is the same.
+This holds if we replace 0 values, add priors or remove 0 count parts.
+In all cases the total remains 1 and the assumption holds.
+In many cases the assumption that the scale of all samples is nearly constant is likely good enough.
+For example, it likely holds approximately in data taken from closely controlled experiments where only a small perturbation was made.
+
+However, early on it was noted that simple proportions introduced bias into the analyses and so alternative normalizations were developed.
+These normalizations were explicitly developed to attempt to scale the output data so that it more closely reflected the environment.
+They did this by making assumptions that appropriate reference parts, or subsets of parts, could be chosen that would allow a good estimate of \(W_d^n\) from \(Y_d^n\).
+Multiple such normalizations exist but they differ only in how the subset is chosen.
+For example the RLE as
+
+There are several ways of determining the appropriate denominator for a composition.
+The most general is to determine a matrix of all possible pairwise log-ratios; this moves the data onto a real space that is completely interpretable and symmetric [@Aitchison:1986].
+However, for large datasets such as those generated by high-throughput sequencing or mass spectroscopy of biological samples where there are thousands of parts, the number of possible ratios becomes computationally infeasible.
+Thus, a number of other approaches have been proposed that involve different ways of choosing appropriate denominators for the log-ratio.
+The main problem now being that the choice of denominator can affect the interpretation of the result.
+
+The earliest method used by Aitchison was the ALR (additive log ratio), where a presumed invariant part was chosen as the reference for the denominator.
+The log-ratio was calculated between each part and the reference resulting in a dataset with 1 less entry than the original.
+The ALR has the advantage of simplicity and ease of interpretation [@aitchison1982;@Greenacre:2021aa].
+The primary disadvantage of the ALR is that it is not isometric, that is, the ALR does not by default  recapitulate the geometry of the all-vs-all log-ratio.
+Greenacre [@Greenacre:2021aa] showed that it was possible to identify a reference for most datasets that was isometric for all practical purposes.
+It is noteworthy that the ALR has long been used, without naming it such, in fields such as molecular biology and mass spectrometry where it is common to use a single reference as a standard.
+
+Another approach proposed was the CLR (centred log-ratio), where it is presumed that the geometric mean of the parts in each sample is invariant.
+The CLR is isometric, and captures all of the ratio information of the all-vs-all approach and has the further advantage of being computationally tractable.
+
+Intermediate denominators have also been proposed. For example, the LVHA (low variance, high abundance) log-ratio
+
+choices of reference give slightly different H ( CLR, ALR, hybrid
+  - choice of denominator affects interpretation
+  - Greenacre
+
+Silverman and colleagues introduced concept of scale which still uses a log-ratio.
+The actual method of choosing the reference is nearly immaterial, instead it is the relationship between the denominators for each group that determines if the log-ratio is useful.
+The key finding was that the denominator chosen was a proxy for the underlying size or scale of the environment.
+The original work by Silverman and colleagues proved this relationship, here we provide an intuitive demonstration based on information theory.
+
+Information theory and CoDa are two ways to deal with compositional data, but information theory is more general. CoDa approaches are applicable to any real vector of numbers
+
+\begin{equation}
+X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x > 0, \sum_X = 1.
+\end{equation}
+
+Information Theory approaches are subtly different in that here 0 values are permitted:
+
+\begin{equation}
+X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x \ge 0, \sum_X = 1.
+\end{equation}
+
+ This s in the l to
+
+Scale is related to entropy
+  - Shannon's entropy measures size of the data
+  - define entropy
+  - define size or scale
+  - H is insensitive to 0
+  - G \(\sim\) H
+    - can use diff H as base
+
 # GM is related to Information and Shannon's entropy in HTS datasets
@@
 
-Information  is a fundamental property of all measured systems.
-For discrete probability vectors, Shannon defined their information properties and launched the field of information theory for communications.
-Famously, for any probability vector, Shannon set the total entropy as the inverse of the sum of the probability weighted logarithm of the probabilities.
-Less well known is that an unweighted average entropy was also defined, and that this turns out to be the inverse of the logarithm of the geometric mean of the probability vector.
-Thus, information theory and compositional data analysis intersect through the geometric mean of a probability vector.
-Here I show that the information theoretic interpretation can help us understand the scale of a system as defined by Nixon and Silverman, and that this interpretation can help in the interpretation of highly asymmetric datasets.
-I will show the Asymptotic Equipartition Property of information can be defined as a volume for discrete distributions, and how that volume relates to the scale of a system.
-Finally, I will show how Shannon's entropy can substitute for the geometric mean in common CoDa operations and how that alters the interpretation of the results.
+Information content is a fundamental property of all measured systems.
+Shannon defined the information properties of discrete probability vectors in the field of communications and launched the field of information theory.
+Information theory concerns itself with  how to encode a dataset optimally to minimize its size when transmitted.
+
+The following example is adapted from Chapter 2 of [@wilde_2017]. If we have information encoded in bits--i.e., either a "0" or "1"--we can imagine a uniform encoding scheme for four symbols
+
+\begin{equation}
+x  = [A,C,G,T]
+\end{equation}
+
+where the underlying probability of observing each symbol is
+
+\begin{equation}
+Pr_x=[\frac{1}{2},\frac{1}{4}, \frac{1}{8}, \frac{1}{8}]. 
+\end{equation}
+
+Now we can define two encodings; a uniform encoding
+
+\begin{equation}
+U_x=[00,01, 10, 11] 
+\end{equation}
+
+and a non-uniform encoding
+
+\begin{equation}
+N_x=[0,10, 110, 111]. 
+\end{equation}
+
+In the uniform encoding the number of bits per symbol is always 2 and in the non-uniform encoding it varies from one to three, with the most common symbol encoded by one bit and the rarest symbols encoded by three bits.
+Lets now examine the amount of information needed to encode a string of characters that has a similar frequency to the true background.
+The 11 character string is:
+
+\begin{equation}
+aacbdabbaab.
+\end{equation}
+
+In the uniform encoding this needs 22 bits because each symbol requires two bits:
+
+\begin{equation}
+\sum{U_x} = 5a + 4b + c + d = 10 + 8 + 2 + 2 = 22.
+\end{equation}
+
+In the non-uniform encoding this needs 19 bits because the number of bits for each symbol is encoded by a variable number of bits:
+
+\begin{equation}
+\sum{N_x}  = 5a + 4b + c + d = 5(1) + 4(2) + 1(3) + 1(3) = 19.
+\end{equation}
+
+Going back to our known background frequencies we can calculate the expected number of bits for a given set of symbols and frequencies. For the uniform encoding we get 
+\begin{equation}
+\frac{1}{2}(2), \frac{1}{4}(2), \frac{1}{8}(2), \frac{1}{8}(2) = \frac{16}{8} = \frac{8}{4} = 2
+\end{equation}
+
+which is 2 bits per symbol. While for the variable length encoding we get
+
+\begin{equation}
+p_i = \frac{1}{2}(1), \frac{1}{4}(2), \frac{1}{8}(3), \frac{1}{8}(3) = \frac{14}{8} = \frac{7}{4}
+\end{equation}
+
+which is smaller.
+
+We are now able to understand the formula for entropy.
+For notational simplicity assume we have a single discrete random variable to represent a probability distribution with d elements; i.e.
+\( X =  \mathbf{p}_{i=(1 \dots d )} \).
+In information theory, the elemental amount of information or surprisal for \(p_i\) is the inverse of the logarithm of the elemental probability [@rezaInfoIntro].
+This measure is often called self-information or the surprisal and I will denote this as \(I\):
+
+\begin{equation}
+I(X_i) = \log_2(\frac{1}{X_i}) = -\log_2{X_i}.
+\end{equation}
+
+Logarithms of probabilities are additive so that for multiple independent probabilities the expected information content of the random variable \(x\) is
+
+\begin{equation}
+\sum_{X} p(X_i) I(X_i) = -\sum_{X}  p(X_i)\log_2(X_i).
+\end{equation}
+
+And this is the formula for entropy \( H \)!
+For any encoding of information in a probability vector Shannon showed that \(H\) is the expected information content of that probability vector.
+We can demonstrate this property  if we remember that the base 2 logarithms of the fractions from equation 2  are -1, -2, -3, -3.
+The expected entropy to base 2 is thus:
+
+\begin{equation}
+H(x) = -\Bigl(\frac{1}{2}(-1) + \frac{1}{4}(-2) + \frac{1}{8}(-3) + \frac{1}{8}(-3) \Bigr) = \frac{7}{4}
+\end{equation}
+
+which is precisely the same result as in equation 9 where we  calculated the average bit length of the infinite length message.
+Thus, entropy  measures the average information encoded by each symbol in a message.
+One conclusion of the above discussion is that information has a size and that we can measure that size as the average number of bits needed to encode the information.
+
+Less well known is that a sample average entropy was also defined
+
+\begin{equation}
+h = - \frac{1}{d}\sum log_2(p),
+\end{equation}
+
+and that this measure converges on \(H\) as the number of measurements from a "typical subset" of the data approaches infinity; this is the Asymptotic Equipartition Property of information as described in [@mackay2003information;@wilde_2017].
+As shown in Chapter 4 of Mackay [@mackay2003information] the typical subset is approximately Gaussian and so if both \(h\) and \(H\) are drawn from this distribution they should be identical in the limit.
+Another way to think about this is that the two measures converge as the distribution of \(p_i\) approaches \(\frac{1}{n}\).
+
+```{r asymptote, echo=F, fig.cap="Plot showing that \\(H\\) and \\(h\\) converge as the size of the dataset increases.  "}
+set.seed(13)
+coeff <- vector()
+s.coeff <- vector()
+
+mn <- c(5,10,30,60,100,200,400,800,1000,2000)
+std <- c(10,7,5,2,1,.8,.6,.5,.3,.2)
+for(i in 1:10){
+  X <- HvG(50, 500, test='N', mean=mn[i], sd=1)
+  jnk <- lm((X$mat.h[,2]~X$mat.log[,2]))
+  coeff[i] <- as.numeric(jnk$coefficients[1])
+  X <- HvG(50, 500, test='N', mean=1000, sd=std[i])
+  jnk <- lm((X$mat.h[,2]~X$mat.log[,2]))
+  s.coeff[i] <- as.numeric(jnk$coefficients[1])
+}
+par(mfrow=c(1,1))
+plot(mn, -coeff, log='y', xlab="vector mean", ylab="H vs. G intercept")
+```
+
+Compositional data analysis is concerned with the analysis of compositions; datasets that are strictly positive and that have an arbitrary upper limit.
+These can always be reduced to a proportion (or probability) without loss of information.
+Compositional approaches have become popular for the analysis of high throughput sequencing data, and other high dimensional data generated by 'omics platforms.
+In  compositional data analysis (CoDa)  one popular data transformation is the centred log-ratio (CLR) transform, which is:
+
+\begin{equation}
+clr_x = log_2(x_i) - mean(log_2(x) =  log_2(x_i) + h_x,
+\end{equation}
+
+and we can see immediately that the second term is equivalent to \(h(x)\). Thus, information theory and compositional data analysis intersect through the geometric mean of a probability vector.
+
+This intersection of information theory and CoDa  can help us understand the scale of a system as defined by Nixon and Silverman which can help in the interpretation of many high throughput sequencing datasets.
+Since \(H\) can be defined as a volume for discrete distributions it is reasonable to treat \(h\)  treated similarly.
+
+Thus, the use of the geometric mean as a measure of size is acceptable both from the information theoretic and from a CoDa perspective.
+
+Biological context of H - average amount of information in each entry. CLR is diff between sample avg entropy and each i(x)
 
 We can think about  scale  from an information theoretic point of view as a measure of how much information, or total uncertainty, is encoded in a particular sample [@Shannon.48;@Jaynes:2003].
-In the geometric interpretation of information theory used in quantum information theory, formally described as  the  Asymptotic Equipartition Property of information [@Cover:1991;@wilde_2017], entropy can be interpreted as the volume occupied by a probability distribution.
-See chapters 4 of the PhD thesis of Lecamwasam [-@ruvi.blog] for a nice explanation of this.
+In the geometric interpretation of information theory used in quantum information theory of information [@Cover:1991;@wilde_2017], entropy can be interpreted as the volume occupied by a probability distribution relative to the maximum total entropy.
+See chapters 4 of the PhD thesis of Lecamwasam [-@ruvi.blog] for a more complete explanation of this.
+Furthermore, the  Asymptotic Equipartition Property says that the expected value of h(X) for a large number of random variables approaches H(X) as the number of variables approaches infinity [@Cover:1991;@wilde_2017].
 
-For notational simplicity assume we have a single discrete random variable to represent a probability distribution with d elements; i.e.
-\( X =  \mathbf{p}_{i=(1 \dots d )} \).
-In information theory, the elemental amount of information or surprisal for \(p_i\) is the inverse of the logarithm of the elemental probability, \( -log_2(p_i)\)[@rezaInfoIntro].
-This measure is often called self-information.
+The weighted average entropy of the system  \( H(X) \) is the weighted sum of the elemental information;
 
-The total entropy of the system  \( H(X) \) is the weighted sum of the elemental information;
+\begin{equation}
+H(X) = - \sum_{i=1}^{d} p_i \log_2 p_i
+\end{equation}
 
-\[
-H(X) = - \sum_{i=1}^{d} p_i \log_2 p_i
-\]
-
-\( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \).
-We can also calculate the expected amount of information for each observation in the random variable, and this is the mean of the elemental probability.
+\( H(X) \) corresponds to the mean amount of information in each entry that is needed  to reproduce  \( X \).
+We can also calculate the unweighted expected amount of information for each observation in the random variable, and this is the mean of the elemental probability.
 This measure is also called the sample average of the information [@wilde_2017];
 
-\[
+\begin{equation}
  h(X) = - \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
-\]
+\end{equation}
 
@@
 
-\[
+\begin{equation}
  l2G(X) = \log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
-\];
+\end{equation};
 
@@
 par(mfrow=c(1,3))
-U <- HvG(20, 500, test='U', mean=1)
+U <- HvG(50, 500, test='U', mean=100)
 U.plot <- HvG.plot(U)
@@
 title(main='random uniform')
-B <- HvG(20, 500, test='B', mean=1)
+abline(0,-1, lty=2)
+B <- HvG(50, 500, test='B', mean=100)
 B.plot <- HvG.plot(B)
 title(main='Extreme skew')
-N <- HvG(20, 500, test='N', mean=30)
+abline(0,-1, lty=2)
+N <- HvG(50, 500, test='N', mean=100, sd=1)
 N.plot <- HvG.plot(N)
-title(main='N, mean=300')
+title(main='N, mean=100')
+abline(0,-1, lty=2)
+
 # intercept
@@
 abline(lm(N$mat.h[15,]~N$mat.log[15,]), lty=2, col='grey', lwd=2)
+abline(0,1, col='red', lty=3)
 plot(U$mat.log[15,], U$mat.h[15,], main="Uniform", xlab="G", ylab="H")
@@
 
+While not necessary in all datasets (e.g., the transcriptome example discussed in the previous section and see the Supplementary material), it can greatly improve modeling in certain cases, especially when the scale is highly asymmetric between conditions.
+In order to specify a scale model from scratch, we need to revisit the concept that all normalizations in widespread use are actually ratios with the denominator implied by the normalization.
+Therefore, we can easily deviate from a certain normalization (e.g., the geometric mean assumption implied by the CLR) by specifying a total model based on the mean difference between conditions.
+While knowing the mean difference between conditions may seem cumbersome in practice, it is the \emph{relationship} between the group scale values that is important, not their raw values (Nixon et al.
+2023).
+This can be illustrated quite simply by starting with the mean ratio in \(G\) between groups for the yeast transcriptome dataset which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for WT; their ratio being 1.17, or a 0.17-fold difference.
+Using this information, we can recapitulate the differential abundance analysis in Figure 2B and 2C exactly by using setting the mean denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental Figure 2.
+This ratio can be adjusted to alter the mean assumption placed on the group scale values.
+
 We can see that for most datasets the difference between the \(\overline{G}\) in each group is relatively small.
@@
 
-conds <- c(rep('NS',7), rep('S',7))
-
-x.all <- aldex(selex, conditions=conds, CI=T, gamma=0.5)
-
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
-x.1.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio as 1:50 explicitly
-#sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
-scale <- c(rep(1,7), rep(50,7)) 
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=FALSE)
-x.5.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio in log space
-# 2^-5.64 : 2^0  = 0.02 : 1
-# this also works, but reversed because of factor()  
-# sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(0,-5.64), conds, log=TRUE)
-scale <- c(rep(-5.6,7), rep(0,7))
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=TRUE)
-
-x.2.all <- aldex(selex,conditions=conds, CI=T, gamma=sel.mu)
+# from selex.aldex.mu.R
+load(file="analysis/sel.2.all.Rda")
+load(file="analysis/sel.5.all.Rda")
+load(file="analysis/sel.1.all.Rda")
+load(file="analysis/sel.all.Rda")
 
 par(mfrow=c(1,4))
-aldex.plot(x.all, main=('gamma=0.5\nmu=G'))
-aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
-aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,50'))
-aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.02, 1'))
+aldex.plot(sel.all, main=('gamma=0.5\nmu=G'))
+aldex.plot(sel.1.all, main=('gamma=0.5\nmu=1,1'))
+aldex.plot(sel.5.all, main=('gamma=0.5\nmu=1,50'))
+aldex.plot(sel.2.all, main=('gamma=0.5\nmu=0.02, 1'))
 ```
@@
 
-Thus far we have observed that adding scale uncertainty increases the minimum dispersion value with little effect on the difference between values as in panel A.
-Panel B shows the change in dispersion relative to the rAbundance of the features.
-Here we can see that the minimum dispersion is increased as gamma increases.
-The horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5.
-Panel C shows that this increase is non-linear, being more pronounced among those features that had minimal dispersion when gamma=0.
-Panel D shows that increasing gamma actually shrinks the dispersion estimates towards the mean dispersion, while increasing the total dispersion as shown in panel C.
-
-Thus, adding scale uncertainty has two effects both on dispersion.
-The first is to increase the dispersion estimate for each part and this has its primary effect to reduce the  p-values and standardized effect sizes returned by ALDEx2.
-The second is to reduce the range of dispersions such that the parts with the lowest dispersions have the greatest increase.
-This increase in dispersion comes about because the underlying distributions are widened because of the additional uncertainty added by the inclusion of scale uncertainty.
-
-```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="jnk"}
-yst.conds <- c(rep('S',44),rep('W',42))
-x <- aldex.clr(yst, yst.conds, gamma=NULL)
-xs <- aldex.clr(yst, yst.conds, gamma=0.5)
-x1 <- aldex.clr(yst, yst.conds, gamma=1)
-
-x.e <- aldex.effect(x)
-xs.e <- aldex.effect(xs)
-x1.e <- aldex.effect(x1)
-
-par(mfrow=c(2,2))
-# we imperceptibly change difference
-plot(x.e$diff.btw, x1.e$diff.btw, xlab='diff g=0', ylab='diff g=1')
-abline(0,1, col='grey', lwd=2, lty=2)
-title(main='A', line=-1.2, adj=0.5)
-#=-1
-cuts <- x.e$rab.all > -0.5 & x.e$rab.all < 0.5
-# the residual of the 
-mn.mid <- median(x.e$diff.win[cuts])
-mn.mid.s <- median(xs.e$diff.win[cuts])
-mn.mid.1 <- median(x1.e$diff.win[cuts])
-
-# we change dispersion a lot for the bulk of the features
-plot(x.e$rab.all, x.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
-  xlab='rAbundance', ylab='dispersion')
-points(xs.e$rab.all, xs.e$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
-points(x1.e$rab.all, x1.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3))
-abline(h=mn.mid, lty=2)
-abline(h=mn.mid.s, lty=2, col='red')
-abline(h=mn.mid.1, lty=2, col='blue')
-title(main='B', line=-1.2, adj=0.5)
-
-plot(x.e$diff.win, x1.e$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3), ylim=c(0,4.5), 
-  xlab='g=0 dispersion', ylab='g=0.5 | g=1 dispersion')
-points(x.e$diff.win, xs.e$diff.win, pch=19, cex=0.5, col=rgb(1,.4,.1,0.3))
-abline(0,1, lty=2)
-title(main='C', line=-1.2, adj=0.5)
-
-strip.data <- list(x.e$diff.win[cuts] - mn.mid, xs.e$diff.win[cuts] - mn.mid.s, x1.e$diff.win[cuts] - mn.mid.1)
-
-stripchart( strip.data, col=c(rgb(0,0,0,0.3), rgb(1,0,0,0.3), rgb(0,0,1,0.3)), vertical=T, method='jitter', pch=19, group.names=c('g=0', 'g=0.5', 'g=1'), ylab='residual dispersion')
-title(main='D', line=-1.2, adj=0.5)
-
+```{r, echo=F}
+load(file="analysis/yst.all.Rda")
+load(file="analysis/yst.s.all.Rda")
+load(file="analysis/yst.1.all.Rda")
 ```
@@
 load('analysis/xt.Rda')
-hk.off <- xt.all$diff.win < 5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
+load('analysis/xg.Rda')
+hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3]
+hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3
 
@@
 
+```{r eval=T, echo=F, fig.cap="Shown here are two volcano plots that plot the mean log2 fold change plotted vs the log of the FDR for the yeast transcriptome dataset and for the metatranscriptome dataset.
+Data were generated in ALDEx2 with \\(\\gamma = [0, 0.5] \\) in black or in red."}
+
+par(mfrow=c(1,2))
+plot(yst.all$diff.btw, -log10(yst.all$we.eBH +1e-70), pch=19, cex=0.5, col=rgb(0,0,0,0.3), 
+  ylab='-log10(FDR)', xlab="log2(diff)", main='yeast')
+points(yst.s.all$diff.btw, -log10(yst.s.all$we.eBH + 1e-20), pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+legend(4,60, legend=c("gamma=0", "gamma=0.5"), pch=19, cex=0.5, col=c("black", "red"))
+
+plot(xt.all$diff.btw, -log10(xt.all$we.eBH), pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  ylab='-log10(FDR)', xlab="log2(diff)", main='meta')
+points(xg.all$diff.btw, -log10(xg.all$we.eBH), pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+
+```
+
+Adding a small amount of scale, \(\gamma=0.5\), has a major effect on the volcano plot for the yeast transcriptome dataset, but mimimal effect on the vaginal metatranscriptome dataset.
+In both datasets the FDR values are raised, but no effect is observed on the difference between values.
+However, the concordance between the FDR values and the difference between become very tight for the yeast transcriptome dataset, but the concordance is not visibly unchanged for the other dataset.
+This is likely because both the difference between and the dispersion were both very small in the former and substantially larger in the latter.
+
 # Checking the scale assumptions of the RLE and TMM normalizations
@@
 
-# References
+Thus far we have observed that adding scale uncertainty has a negligible effect on either the differences between groups or the relative abundance measure with a correlation of `r round(cor(yst.all$diff.btw, yst.1.all$diff.btw), 3)` and `r round(cor(yst.all$rab.all, yst.1.all$rab.all), 3)`.
+The effect is entirely restricted to the dispersion estimats as shown below.
+Panel A shows the change in dispersion relative to the rAbundance of the features.
+Here we can see that the minimum dispersion is increased as gamma increases.
+The horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5.
+Panel B shows that this increase is non-linear, being more pronounced among those features that had minimal dispersion when gamma=0.
+Panel C shows that increasing gamma rotates the dispersion estimates at high relative abundances, such that housekeeping genes no longer have larger mean dispersions (rAB > 5) than do regulated genes (rAB >0, < 5).
+Overplotted are the lowess lines of fit through the densities of transcripts that are judged as significantly different using either scaled or unscaled data, and with or without thresholding.
+Statistical significance was determined with a FDR < 0.05.
+
+Thus, adding scale uncertainty has two effects both on dispersion.
+The first is to increase the dispersion estimate for each part and this has its primary effect to reduce the  p-values and standardized effect sizes returned by ALDEx2.
+The second is to reduce the range of dispersions, and rotate them such that the parts with the lowest dispersions have the greatest increase.
+This increase in dispersion comes about because the underlying distributions are widened with the inclusion of scale uncertainty.
+
+```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="Adding scale uncertainty changes the dispersion distribution.
+Panel A shows a plot of the expected value for relative abundance vs the expected value for the pooled dispersion as output by aldex.effect.
+The dashed  horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5.
+Panel B plots the unscaled vs scaled dispersion, note the non-linear relationship.
+Panel C plots relative abundance vs the  difference between dispersion with gamma set to 0 and to 1 to highlight the rotation that is evident in Panel A.
+The colored lines indicate the lowess line of fit through the centre of mass of the plot.
+Line colors represent different populations of points.
+The grey line is the total population, the red line is the population of significant transcripts with no scale, the orange line is the population of significant transcripts with a difference threshold of about 2.5-fold change, the blue line is the population of significant transcripts with gamma =0.5, and the cyan line is the significant population with gamma = 1.
+"}
+
+cuts <- yst.all$rab.all > -0.5 & yst.all$rab.all < 5
+par(mfrow=c(1,3))
+mn.mid <- median(yst.all$diff.win[cuts])
+mn.mid.s <- median(yst.s.all$diff.win[cuts])
+mn.mid.1 <- median(yst.1.all$diff.win[cuts])
+
+# we change dispersion a lot for the bulk of the features
+plot(yst.all$rab.all, yst.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  xlab='rAbundance', ylab='dispersion')
+points(yst.s.all$rab.all, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+points(yst.1.all$rab.all, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3))
+abline(h=mn.mid, lty=2)
+abline(h=mn.mid.s, lty=2, col='red')
+abline(h=mn.mid.1, lty=2, col='blue')
+title(main='A', line=-1.2, adj=0.5)
+legend(0,4, legend=c("0", "0.5", "1"), 
+  pch=19, col=c("black", "red", "blue"))
+
+plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3), ylim=c(0,4.5), 
+  xlab='g=0 dispersion', ylab='g=0.5 | g=1 dispersion')
+points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+abline(0,1, lty=2)
+title(main='B', line=-1.2, adj=0.5)
+legend(0,4, legend=c( "0.5", "1"), 
+  pch=19, col=c("red", "blue"))
+
+sig.all <- yst.all$we.eBH < 0.05
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
+sig.s <- yst.s.all$we.eBH < 0.05
+sig.1 <- yst.1.all$we.eBH < 0.05
+
+diff.df <- data.frame(yst.all$rab.all, 
+  yst.1.all$diff.win- yst.all$diff.win)
+sig.df <- data.frame(yst.all$rab.all[sig.all],
+   yst.1.all$diff.win[sig.all]- yst.all$diff.win[sig.all])
+sig.t.df <- data.frame(yst.all$rab.al[sig.t], 
+  yst.1.all$diff.win[sig.t] - yst.all$diff.win[sig.t])
+sig.s.df <- data.frame(yst.all$rab.all[sig.s], 
+  yst.1.all$diff.win[sig.s] - yst.all$diff.win[sig.s])
+sig.1.df <- data.frame(yst.all$rab.all[sig.1], 
+  yst.1.all$diff.win[sig.1]- yst.all$diff.win[sig.1])
+
+plot(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win, xlab='rAbundance', ylab="disp:1 - disp:0", pch=19, cex=0.5)
+title(main='C', line=-1.2, adj=0.5)
+lines(lowess(diff.df, f=0.1), col='grey60', lwd=3)
+lines(lowess(sig.df, f=0.1), col='red', lwd=3)
+lines(lowess(sig.t.df, f=0.5), col='orange', lwd=3)
+lines(lowess(sig.s.df, f=0.5), col='royalblue', lwd=3)
+lines(lowess(sig.1.df, f=0.7), col='cyan', lwd=3)
+
+legend(-10,1.3, legend=c( "diff", "sig diff", "t + sig","0.5 sig", "1 sig"), 
+  pch="-", col=c("grey", "red", "orange", "royalblue", "cyan"))
+
+```

### 2025-03-11 submitted/go3-author-sub.tex
submitted version Dec 6

--- 2025-03-11 go3.tex
@@
 ALDEx2}
-\author{true \and true \and true}
-\date{}
+\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4,5}$}
+\date{\today}
 
@@
 \maketitle
+
+$^1$Department of Biochemistry, University of Western Ontario
+$^2$Department of Population Health Sciences, Geisinger, Danville, PA
+$^3$Department of Medicine, Pennsylvania State University
+$^4$Institute for Computational and Data Science, Pennsylvania State University
+$^5$Department of Statistics, Pennsylvania State University
+
 \begin{abstract}

### 2025-03-11 supplement_figs.Rmd
submitted version Dec 6

--- 2023-10-17 supplement_figs.Rmd
@@
 ---
-title: "Supplement: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
+title: "Supplement: Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2
 "
 shorttitle: "Scale ALDEx2 Supp"
-author:
-- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
-  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
-  email: ggloor@uwo.ca
+citeproc: TRUE
+cite-method: citeproc
 bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
+csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
+pdf-engine: latex
 output:
-  BiocStyle::pdf_document: default
-package: ALDEx2 1.33.1
-abstract: |
-    Introduction to scale simulation and FDR correction with ALDEx2.
-date: 21 August 2023
-link-citations: true
+  pdf_document
 ---
 
-# GM is related to Information and Shannon's entropy in HTS datasets
+```{r echo=F, warning=F, message=F}
+library(ALDEx2)
+data(selex)
+```
+```{r echo=FALSE, include=TRUE, out.width="0.8\\linewidth",  fig.align="center", fig.cap=c("The aldex.senAnalysis() function was used to generate a dataset with \\(\\gamma\\) values of 1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7 and 1.
+The plotGamma() function was used to plot the result.
+Transcripts that are statistically significant are shown in black, and if not significant are in grey.
+The \\(\\gamma\\) values, standardized effect size and number of significant transcripts at each value are given on the axes.
+")}
+knitr::include_graphics("analysis/scaleSim-yeast.pdf")
+## plotGamma needs to be cleaned up for production
+## so this figure was done by hand with the example
+## snippet below - ugly sorry
+#plotGamma(x.sens, blackWhite=T)
+#text(6.2,345, label=length(which(x.sens[[8]]$we.eBH < 0.05)))
+#text(-2,345, label=length(which(x.sens[[7]]$we.eBH < 0.05)))
+#text(-6.9,345, label=length(which(x.sens[[6]]$we.eBH < 0.05)))
+#text(-9.6,345, label=length(which(x.sens[[5]]$we.eBH < 0.05)))
+#text(-12.2,345, label=length(which(x.sens[[4]]$we.eBH < 0.05)))
+#text(-14.8,345, label=length(which(x.sens[[3]]$we.eBH < 0.05)))
+#text(-17.4,345, label=length(which(x.sens[[2]]$we.eBH < 0.05)))
+#text(-20,345, label=length(which(x.sens[[1]]$we.eBH < 0.05)))
+#text(-6.9,355, label="N significant")
+```
 
-Information content is a fundamental property of all systems.
-We can think about  scale  from an information theoretic point of view as a measure of how much information is encoded in a particular sample.
-Intuitively,   systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).
-Below, we show that the scale of a system as defined by Nixon et al.
-[-@nixon2023scale] is inversely related to the information content and entropy of the data.
+ One root cause of the large number of significant parts is the very low dispersion of transcripts.
+Figure 1 shows a graphical output from the texttt{aldex.scaleSim()} function with the yeast transcriptome dataset described in the text [@Gierlinski:2015aa;@Schurch:2016aa].
+This allows us to examine which transcripts are sensitive to even minimal amounts of scale uncertainty using \(\gamma = (1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7, 1) \).
+Here it is obvious that even a negligible amount of scale uncertainty removes over half of the transcripts that were formerly significantly different, and all of these had very small effect sizes.
 
-Recall the underlying system is described by a \(D \times N\) matrix of counts \(\mathbf{W}\) decomposed into the proportions for the \(n^{th}\) sample \(\mathbf{W}^{\parallel}_n\) (or the equivalent probability distribution \( \mathbf{p}(w_n) \) ), and its scale \(\mathbf{W}^{\perp}_n\), such that \(\mathbf{W}=\mathbf{W}^{\parallel}\mathbf{W}^{\perp} \).
-Sequencing  returns counts which are related to the underlying proportion; i.e., \(\mathbf{Y}^{\parallel}_n \sim \mathbf{W}^{\parallel}_n\)
+We recommend a minimum scale uncertainty of 0.5, and suggest that the texttt{aldex.senAnalysis()} function be run on all analyses.
+The individual \texttt{aldex()} outputs can be accessed as sequential entries in the list output, or the analysis as a whole can be plotted with the \texttt{plotGamma} function.
 
-For notational simplicity assume we will use a single discrete random variable to represent the probability distribution; i.e.
-\( X =  \mathbf{p}(w_n) \)  over \(1 \dots d \) elements.
-The geometric mean can be calculated as:
+```{r, echo=F, warning=F, message=F,comment=F, result=F} 
+library(DESeq2)
+library(ALDEx2)
+#ALDEx2
+load(file="analysis/yst.all.Rda")
+load(file="analysis/yst.s.all.Rda")
+load(file="analysis/yst.1.all.Rda")
+# with gamma = 0.5
+load('analysis/x.s.all.Rda')
+load('analysis/yst.sens.Rda')
+nsig <- x.sens[[2]]$we.eBH > 0.05 & x.sens[[1]]$we.eBH < 0.05
 
-\[
-\log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
-\];
+# data to compare different scale means
+load(file='analysis/xt.lv.Rda')
+load(file='analysis/xt.m.Rda')
+load(file='analysis/xg.Rda')
+```
 
-here every element in \(log_2 G(X)\) is the logarithm of the elemental probability \( log_2(p_i)\).
+Figure 2 compares the significant transcripts using effect and volcano plots for the first two \(\gamma\) parameters.
+We can see that the `r length(nsig)` transcripts are excluded by even the smallest setting of \(\gamma=0.1\).
+All of these have very low difference between and include the majority of the transcripts with very low dispersion.
 
-In information theory, the elemental amount of information or uncertainty for \(p_i\) is the inverse \(I(X)_i = -log_2(p_i)\)[@rezaInfoIntro].
+```{r echo=FALSE, include=TRUE, out.width="0.8\\linewidth",  fig.align="center", fig.cap=c('Effect and volcano plots showing the significant transcripts for the first two \\(\\gamma\\) values of 1e-3 and 0.1.
+Transcripts that are significantly different with a q-value <= 0.5 with both \\(\\gamma\\) values are in red, those significant with \\(\\gamma = 0.1 \\) are in orange.
+Those that are not significant are in dark grey.')}
 
-The total entropy of the system  \( H(x) \) is the weighted sum of the elemental information
+par(mfrow=c(1,2))
+aldex.plot(x.sens[[1]])
+points(x.sens[[1]]$diff.win[nsig], x.sens[[1]]$diff.btw[nsig], pch=19, cex=0.5, col='orange')
 
-\[
-H(x) = -\sum_{i=1}^{d} p_i \log_2 p_i
-\]
+aldex.plot(x.sens[[1]], type='volcano')
+points(x.sens[[1]]$diff.btw[nsig], -log10(x.sens[[1]]$we.eBH[nsig]), , pch=19, cex=0.5, col='orange')
+```
+\newpage
 
-In this formulation, \( \log_2(G(X)) \) and \( H(X) \) can be understood  to be an unweighted and a weighted measure  of 'surprisal' in the dataset.
-The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.
-\( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \).
-Conversely, \( G \) is an unweighted mean of the same distribution and  \( G \) is the denominator to calculate the centred log-ratio normalization (CLR).
-Note that when calculating the CLR we are taking a ratio, and similarly for calculating entropy, but that the scale formula takes the product.
-Thus, the scale of a system is inversely related to the other two measures.
+Figure 3 shows how dispersion behaves in linear and log space.
+The variance or dispersion always increases with increasing raw (or normalized) read count but decreases when measured on the log-ratio transformed data [@fernandes:2013;@Love:2014aa], reaching a minimum at some mid-point of the distribution.
+This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes.
+This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa].
+Furthermore, the dispersion is exceeding small being, for many transcripts, almost negligible.
+To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead to many false positives.
+Indeed, benchmarking and comparison studies  repeatedly show that the choice of normalization plays a role in which results are returned as significant [@maza2013;@Dillies:2013;@Weiss:2017aa].
+From the perspective of Nixon et al.
+[-@nixon2023scale], it is reasonable to conclude that some of these results may be due to the choice of normalization.
 
-Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(\mathbf{Y}^{\parallel}_{n})\) (Supplemental Figure 1 and Table 1).
-These two measures are expected to have different behaviours with different distributions of \( p_i \).
-In the case of a uniform distribution both \( H(\mathbf{Y}^{\parallel}_{n}) \) and \( G(\mathbf{Y}^{\parallel}_{n}) \) are maximal since \( p(x) \) is equally and identically distributed.
-Thus, we expect that they are positively correlated here.
-In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values.
-In very sparse  datasets, the two measures could become uncoupled because \( H(\mathbf{Y}^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(\mathbf{Y}^{\parallel}_{n}) \) would tend to be very small.
-Here these two measures could be either uncorrelated or exhibit negative correlation.
-We can see this distributional behaviour in different datasets.
+```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T,fig.dim=c(7,3), fig.cap="Plot of abundance v dispersion for the yeast  transcriptome dataset as counts,	 as logarithms of counts, and as CLR values.
+Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale.
+Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first.
+Panel C shows the equivalent values calculated by ALDEx2	 in which the expected CLR value for each transcipt are plotted vs.
+the expected dispersion.
+Panel D shows the output for ALDEx2 with \\( \\gamma=0.5 \\).
+The red line in each panel shows the LOESS line of fit to the mid-point of the distributions.
+In panels B and C the amount of dispersion reaches a minimum at moderate	 values.
+The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value.
+The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }
 
-```{r info, echo=F,warning=F, message=F,comment=F, fig.cap="Plot of Shannon's entropy (H) vs geometric mean (G) for each sample in different datasets.
-The groups that each sample belong to are highlighted as filled or open circles.
-Each group in each dataset has different entropy with the groups in the selex and metatranscriptome datasets being highly distinct."}
-# Gm ~ Im
-# plotting geometric mean vs Shannon's entropy
-# shows differential information by group
-devtools::load_all('~/Documents/0_git/ALDEx_bioc')
-data(selex)
-
-load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
 url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
 yst <- read.table(url, header=T, row.names=1)
-url <- "https://raw.githubusercontent.com/ggloor/datasets/main/meta16S.tsv"
-rRNA <- read.table(url, header=T, row.names=1, sep='\t')
-url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
-ss <- read.table(url, header=T, row.names=1, sep='\t')
-ss <- ss[,c(1:100,1502:1601)]
 # remove the one gene with 0 reads
+
 yst <- yst[rownames(yst) != "YOR072W-B",]
+
 # Gierlinski:2015aa
@@
 
-HS <- apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GS <- apply(selex+0.5, 2, function(x) mean(log2(x/sum(x))))
+conds <- c(rep('S', 44), rep('W', 42))
+coldata <- data.frame(conds)
 
-HY <- apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GY <- apply(yst+0.5, 2, function(x) mean(log2(x/sum(x))))
+# DESeq2
+load('analysis/res.Rda')
 
-HM <- apply(ko.both+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-GM <- apply(ko.both+0.5, 2, function(x) mean(log2(x/sum(x))))
+#ALDEx2
+load('analysis/x.all.Rda')
+load('analysis/x.s.all.Rda')
 
-H16 <- apply(rRNA+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-G16 <- apply(rRNA+0.5, 2, function(x) mean(log2(x/sum(x))))
+# with gamma = 1
+load('analysis/x.s.all.Rda')
+sig.des <- which(res@listData$padj < 0.05)
+sig.ald <- which(x.all$we.eBH < 0.05)
+sig.s.ald <- which(x.s.all$we.eBH < 0.05)
 
-Hss <- apply(ss+0.1, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
-Gss <- apply(ss+0.1, 2, function(x) mean(log2(x/sum(x))))
+# get and plot mean/var or mean/SE for counts
+# and log counts
+yst.mn <- apply(yst, 1, mean)
+yst.mn.log <- apply(yst, 1, function(x) mean(log2(x)))
+yst.mn.log[is.infinite(yst.mn.log)] <- 0
 
-par(mfrow=c(2,3))
-plot(HS, GS, ylab="Geometric mean", xlab="Shannon's entropy",
-  pch=c(rep(19,7),rep(1,7)), col=rgb(0,0,0,0.5))
-title(main='selex')
+yst.v <- apply(yst, 1, var)
 
-plot(HY,GY, 
-  ylab="Geometric mean", xlab="Shannon's entropy", pch=c(rep(19,44),rep(1,42)),
-  col=rgb(1,0.6,0,0.5))
-title(main='transcriptome')
+par(mfrow=c(1,4))
+plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='Mean count', ylab='Variance')
+#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
+def <- data.frame(log2(yst.mn),log2(yst.v))
+lines(lowess(def, f=0.1), col=2, lwd=2)
+title('A', adj=0, line= 0.8)
+abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 
-plot(HM, GM, col=rgb(1,0,0,0.5), pch=c(rep(1,8), rep(19,28), rep(1,8)), 
-  ylab="Geometric mean", xlab="Shannon's entropy")
-title(main='meta-transcriptome')
+l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
+plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
+lines(lowess(l.df, f=0.25), col=2, lwd=2)
+abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.0275, col='grey', lty=2)
+title('B', adj=0, line= 0.8)
 
-plot(H16,G16, col=rgb(0,1,1,0.5),
-  pch=c(rep(1,198), rep(19,161)), 
-  ylab="Geometric mean", xlab="Shannon's entropy")
-title(main='16S rRNA')
+l.df <- data.frame(x.all$rab.all, x.all$diff.win)
+plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='E(clr)', ylab='E(disp)')
+#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
+lines(lowess(l.df, f=.1), col=2, lwd=2)
+abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.208, col='grey', lty=2)
+title('C', adj=0, line= 0.8)
 
-plot(Hss,Gss, col=rgb(0.5,1,0,0.5),
- pch=c(rep(1,100), rep(19,100)), 
-  ylab="Geometric mean", xlab="Shannon's entropy")
-title(main="single cell")
+l.s.df <- data.frame(x.s.all$rab.all, x.s.all$diff.win)
+plot(x.s.all$rab.all, x.s.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
+  xlab='E(clr)', ylab='E(disp)', ylim=c(0,4.5))
+#points(x.s.all$rab.all[sig.ald], x.s.all$diff.win[sig.ald], col='orange', cex=0.4)
+lines(lowess(l.s.df, f=.1), col=2, lwd=2)
+abline(h=min(lowess(l.s.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
+ abline(h=0.208, col='grey', lty=2)
+title('D', adj=0, line= 0.8)
+```
 
-# empty plot for legend
-plot(selex[,1], selex[,2], xlab=NA, ylab=NA, axes=F,xlim=c(-10,-1), ylim=c(-10,-1))
-legend(-10,-1, legend=c('selex', 'transcriptome','metatranscriptome', '16S', 'SS', 'low entropy group', 'high entropy group'),
-  col=c('darkgrey', 'orange', 'red', 'cyan', 'greenyellow','grey','grey'), pch=c(19,19,19,19,19,1,19))
+\newpage
+
+Figure 4 shows that the informed models with 5% or 14% difference in location between groups and \(\gamma=0.5\) provide nearly the same output for q-values, effect sizes and difference between groups (black).
+The line of identity is given as the grey diagonal.
+However, using the default CLR values to specify location are very different.
+In the q-value and effect plots, there is multiple populations of points that indicate the Type 1 and 2 errors that occur when the location is not specified properly.
+In the difference between groups plot, we see only a shift in the location that corresponds to moving the mass of the data points down by about 2.2 units (See Figure 3 in the main text)
+
+```{r echo=FALSE, include=TRUE, fig.dim=c(7,3),  fig.align="center", fig.cap=c("Plots showing the similarity of outputs with different scale parameters.
+The black points show that using either a an informed model with 5% or 14% difference in location has a minimal effect on either the q-values, the effect size or difference between groups.
+In red, the same values  are plotted with the default model that uses a naive estimate of the location derived from the CLR.")}
+
+par(mfrow=c(1,3))
+plot(xt.m.all$we.eBH,xt.lv.all$we.eBH, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  main="q-value", log='xy', ylab="5% diff", xlab="14% diff")
+points(xt.m.all$we.eBH, xg.all$we.eBH, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+abline(0,1, lty=2, lwd=2, col='grey')
+plot(xt.m.all$effect,xt.lv.all$effect, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  main="effect", ylab="5% diff", xlab="14% diff")
+points(xt.m.all$effect, xg.all$effect, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+abline(0,1, lty=2, lwd=2, col='grey')
+
+plot(xt.m.all$diff.btw,xt.lv.all$diff.btw, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
+  main="diff.btw", ylab="5% diff", xlab="14% diff")
+points(xt.m.all$diff.btw, xg.all$diff.btw, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
+
+abline(0,1, lty=2, lwd=2, col='grey')
 
 ```
-
-The table below summarizes the mean values for, and the correlation between, \( G \) and \( H \)  (cor)  and the sparsity defined as the proportion of features with less than 1 count per sample (spar) for each association in each group of samples:
-
-| Dataset | group | \(\overline{G}\) | \( \overline{H} \) | cor | spar |
-|---------|-------|---|---|-----|-------|
-| Selex   | control | -11.2 | 10.2 | 0.99 | 0 |
-| $~~~$"  | selected | -17.8 | 2.8 | -0.88 | 0.802 |
-| yeast   | snf2 ko  | -14.0 | 10.7 | 0.99 | 0.004 |
-| $~~~$"  | WT    | -14.2 | 10.4 | 0.99 | 0.007 |
-| Meta    | H     | -18.8 | 8.6 | 0.78 | 0.451 |
-| $~~~$"  | BV   | -18.2 | 8.9 | 0.79 | 0.238 |
-| 16S     | Pup   | -14.7 | 5.4 | 0.68 | 0.079 |
-| $~~~$"  | Cent | -15.2 | 5.4 | 0.53 | 0.251 |
-| SS      | A     | -13.0 | 8.2 | 0.83 | 0.978 |
-| $~~~$"  | B    | -12.9 | 8.3 | 0.80 | 0.977 |
-
-We can see that for most datasets the difference between the \(\overline{G}\) in each group is relatively small.
-Most significantly, the selex dataset has a very large difference of about 100-fold, and both the 16S and the metatranscriptome dataset have about a 1.5 fold difference.
-These three datasets are candidates for a full scale model correction.
-
-```{r full.scale, echo=F, fig.cap="Plot of the offset of the mean differnece between groups as a function of scale ratio.
-For this, the default scale of 1:1 was altered in increments of 0.1 keeping the gamma parameter (dispersion) at 0.5.
-The filled circle shows the outcome when the calculation is done using the geometric mean and the same gamma parameter."}
-
-load('analysis/data.out.Rda' )
-
-plot(data.out[,6],data.out[,1], pch=c(rep(1,19),19), ylim=c(-11, 4),
-  xlab='full scale offset', ylab='mean difference between groups')
-points(data.out[,6],data.out[,2], pch=c(rep(1,19),19), col='red')
-points(data.out[,6],data.out[,3], pch=c(rep(1,19),19), col='orange')
-points(data.out[,6],data.out[,4], pch=c(rep(1,19),19), col='cyan')
-points(data.out[,6],data.out[,5], pch=c(rep(1,19),19), col='blue')
-abline(h=0, col='grey', lty=2)
-abline(v=0, col='grey', lty=2)
-legend(0.7, 0, legend=c('16S','selex','tscome','metatscome','ss'),
-  pch=19, col=c('black','red','orange','cyan','blue'))
-
-```
-
-Nixon et al.
-[-@nixon2023scale] showed that many operations on HTS datasets relied on both the proportion and scale components of the data.
-Moreover, all normalizations impose a scale model on the data, but the appropriateness of these models has never been explicitly acknowledged or tested.
-Thus, the full scale model option allows the investigator to set both the offset between the geometric means of the features and their dispersion and observe how this affects the analysis outcome.
-
-In the offset plot above we can see the cause and the effect of the full model with a fixed gamma of 0.5 and a base scale of 1 in each group.
-We see that the mean location of well centred datasets (yeast transcriptome, single-cell transcriptome) are close to 0, but could be centred better with small changes in scale ranging from 0 (single cell) to 1:1.1 for the yeast transcriptome dataset.
-In contrast, centring the 16S dataset requires about a 1:1.3 fold change.
-The metatranscriptome dataset would require about a 0.5:1 change in relative scale between groups, but as shown in the main text, centring the housekeeping genes is more apt.
-
-The in vitro selection dataset is clearly an outlier in both the difference between the average group geometric mean, and in the offset plot.
-However, this dataset can  be used to illustrate the power of the full scale model and the relationship between \( G_n \) and scale.
-In Figure 3 we can see that the default output of ALDEx2 has a centered output.
-This occurs largely by chance, as the high sparsity of the selected (S) group is balanced almost exactly by the arbitrarily chosen sequencing depth so the non-selected group (NS) appears to have a similar location as the S group.
-The difference in entropy between the two groups and the differences in geometric mean are very large, with the difference in \(\log_2{\overline{G}(S)}\) and \(log_2{\overline{G}(NS)}\) being about \(2^{6.6}\).
-Setting the scale of both the S and NS groups to 1 we find that the difference in location is approximately \(2^{7.7}\) in close agreement with the difference the geometric means.
-For this dataset to be centered we need to have a scale ratio \( \approx \) 1:50 or more.
-Note that the scale ratio is inverse to the ratio of geometric means as described above.
-In fact, in this dataset the relative abundances of the majority of features are nearly invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014], thus changing the scale of the data.
-Neither DESeq nor edgeR are able to provide a reasonable analysis of this dataset because the normalizations used assume equivalent scales [@gloorAJS:2016].
-
-Figure 3 shows an effect plot of various scale models with this dataset.
-The full scale model, where the strong assumption that the mean \(G\) is assumed to be 1:1 between the two groups, dramatically skews the output and the large number of relatively invariant features are now identified as significantly different.
-While not wrong as long as the assumption that the scales are identical is stated, this is not a useful analysis outcome.
-Modifying the mean scale difference between the NS:S groups to be \( \approx \) 1:50 different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained using \(G_n\) as the scale estimate where the ratio is \( \sim 100:1 \).
-Note that we get exactly the same answer (within random sampling error) with a scale of .02 for group NS and a scale of 1 for group S, or using a scale of 1 for group NS and a scale of 50 for group S.
-This shows that it is the relative difference between scales that is important in this dataset, not the absolute values.
-From this result we can conclude that, on average, the difference in underlying scale in the system is about \(\approx 50\)-fold, and this is congruent with the circa 100-fold difference in  \(\overline{G}\) between groups; the discrepancy being explained because the default scale model is applied uniformly to all samples, whereas the different values of the within-group geometric means ranging over a \{ > 2.5 \) fold range.
-Thus, an advantage of a full scale model is that we have gained both information and understanding about the drivers of asymmetry underlying system.
-
-```{r selex, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters.
-All scales are calculated with a logNormal distribution to ensure symmetry for the user.
-"}
-
-conds <- c(rep('NS',7), rep('S',7))
-
-x.all <- aldex(selex, conditions=conds, CI=T, gamma=0.5)
-
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
-x.1.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio as 1:50 explicitly
-#sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
-scale <- c(rep(1,7), rep(50,7)) 
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=FALSE)
-x.5.all <- aldex(selex, conditions=conds, CI=T, gamma=sel.mu)
-
-# setting ratio in log space
-# 2^-5.64 : 2^0  = 0.02 : 1
-# this also works, but reversed because of factor()  
-# sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(0,-5.64), conds, log=TRUE)
-scale <- c(rep(-5.6,7), rep(0,7))
-sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=scale, conds, log=TRUE)
-
-x.2.all <- aldex(selex,conditions=conds, CI=T, gamma=sel.mu)
-
-par(mfrow=c(1,4))
-aldex.plot(x.all, main=('gamma=0.5\nmu=G'))
-aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
-aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,50'))
-aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.02, 1'))
-```
-
-# Issues with DESeq2 and edgeR
-
-DESeq2 and edgeR are two of the most commonly used tools for differential abundance analysis of bule RNA sequencing datasets.
-They both operate by finding a scaling factor that makes all the samples commesurate.
-DESeq2 does this by finding a midpoint feature that can be used as a reference in each sample; this can be different for different samples.
-The edgeR reference finds the midpoint of the 'typical' sample instead.
-In both cases the data are then scaled by dividing by a small factor that makes the read counts commesurate.
-Differential abundance analysis is then performed on the scaled values after taking their logarithm to base 2.
-In some ways this is similar to the log-ratio approach used by ALDEx2, but is more prone to dataset and sample effects than is the log-ratio method [@GloorAJS2023]
-
-```{r DESedg, echo=F, fig.cap="Shown here are the mean log2 fold change as a density plot, and a Volcano plot showing the location and adjusted p-value for each feature in the metatranscriptomic dataset.
-The DESeq2 approach does a good job of centring this data, while edgeR is less suitable.
-The volcano plots show dramatically different outcomes.
-The DESeq2 algorithm assigns very large fold changes to features that have only moderate change, and further identifies a very large proportion of features as significantly different.
-In contrast, edgeR exhibits a much smaller number of differentially abundant features.
-In both volcano plots, the housekeeping genes in the main Figure 3 are shown in orange.
-We can see that these are asymmetrically distributed in both plots.
-Additionally the location of the features taht DESeq2 identified as having a very large difference are shown in the edgeR volcano plot as blue circles."}
-
-# find housekeeping from ALDEx2
-
-load('analysis/xt.m.Rda')
-load('analysis/xt.Rda')
-hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3
-
-load('analysis/meta.DES.res.Rda')
-load('analysis/meta.edge.qlf.Rda')
-
-# these are all very rare K0s
-zero.var <-  which(meta.DES.res@listData$lfcSE*sqrt(44) == 0)
-
-weird.des <- which(meta.DES.res@listData$log2FoldChange < -20 )
-
-# housekeeping genes centred with DESeq2 - nice
-
-sig.des <- which(meta.DES.res@listData$padj < 0.01)
-sig.edge <- which(p.adjust(meta.edg.qlf$PValue) < 0.01)
-par(mfrow=c(2,2))
-
-plot(density(meta.DES.res@listData$log2FoldChange[hk.off]), main='DESeq2 FC')
-abline(v=0, col='red', lty=2)
-
-plot(meta.DES.res@listData$log2FoldChange, -1*log10(meta.DES.res@listData$padj), 
-  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)', main='DESeq2 volcano')
-points(meta.DES.res@listData$log2FoldChange[sig.des],-1*log10(meta.DES.res@listData$padj[sig.des] + 1e-70), col='red')
-points(meta.DES.res@listData$log2FoldChange[hk.off],-1*log10(meta.DES.res@listData$padj[hk.off] + 1e-70), col='orange', cex=0.6)
-
-plot(density(meta.edg.qlf$logFC[hk.off]), main='edgeR FC')
-abline(v=0, col='red', lty=2)
-
-plot(meta.edg.qlf$logFC, -1*log10(p.adjust(meta.edg.qlf$PValue)), 
-  col=rgb(0,0,0,0.1), xlab='LFC', ylab='-1log10(p)', main='edgeR volcano')
-points(meta.edg.qlf$logFC[sig.edge], -1*log10(p.adjust(meta.edg.qlf$PValue)[sig.edge]), 
-  col='red')
-points(meta.edg.qlf$logFC[hk.off], -1*log10(p.adjust(meta.edg.qlf$PValue)[hk.off]), 
-  col='orange', cex=0.6)
-  points(meta.edg.qlf$logFC[weird.des], -1*log10(p.adjust(meta.edg.qlf$PValue)[weird.des]), 
-  col='blue', cex=0.6)
-
-```
-
-Examining the plots, edge R clearly not centred with the median housekeeping functions offset by `r median(meta.edg.qlf$logFC[hk.off==T])` and minimum FDR value  is `r min(p.adjust(meta.edg.qlf$PValue)[hk.off==T])`.
-Additionally, as shown in Figure 4there is little range in the p-values relative to those seen for DESeq2, and the values are more in line with the range seen with ALDEx2.
-
-DESeq2 is better centred with median housekeeping functions offset by `r median(meta.DES.res@listData$log2FoldChange[hk.off==T])` but the minimum FDR value  is `r min(meta.DES.res@listData$padj[hk.off==T])`.
-In addition there are a large number of functions with 0 variance, these are very low count functions with very high sparsity in the dataset.
-These are not differential in the DESeq2 analysis.
-However, there are a number of functions in the DESeq2 analysis that are very differentially abundant, with a log2 fold change of < -20.
-Examination of the raw counts shows that these are uniformly 0 or 1 in the H dataset, and present at high counts in some, but not all of the BV dataset.
-That these stand out is odd, since inspection of the count table shows that there are many other functions that are missing from the H dataset, and have higher and more uniform counts in the BV dataset.
-Thus, the importance of the outlier functions seen in the DESeq2 volcano plot should be viewed with suspicion and may be an artefact of the normalization used.
-When overplotted on the edgeR volcano plot (blue), it is clear that these functions have non-signficant p-values.
-
-# Checking the scale assumptions of the RLE and TMM normalizations
-
-We can show that the normalizations built into the edgeR Bioconductor package (RLE, TMM, TMMwsp, upperquantile) are scale assumptions by using the normalization factor as an input to `aldex.makeScaleMatrix()` and then measure the mean location of the data as above.
-The ideal behavior of a normalization is that the mean location of the data should be close to 0 or unchanged.
-This behavior will ensure that Type 1 and Type 2 errors due to scale assumptions are minimized.
-
-The results, shown in the tables below compare these normalizations to the no scale assumption (iso) and the geometric mean normalization (GM) assuming that the normalizations are either on a log scale or a linear scale.
-That these affect scale can be observed by their effect on the mean location of the data.
-For the most part assuming no scale differences between groups centres the date less well than does the geometric mean.
-In most cases the other normalizations perform poorer than assuming no scale at all for the rRNA dataset and about as well as the no scale assumption in the metatranscriptome datasets and the single-cell transcriptome dataset.
-All normalizations perform poorly in the selex dataset.
-Overall, these results may not be surprising because the TMM and RLE (and related normalizations) were developed specifically to scale and normalize transcriptome datasets [@Robinson:2010a;@Anders:2010], but have become widely used in the analysis of other data modalities; e.g.
-[@McMurdie:2014a].
-
-```{r getNorm, echo=F}
-# output from code/scale_norms.R
-
-load('analysis/normalizations.Rda')
-knitr::kable(norm.data.out, 'simple', caption='Log scale models')
-```
-
-```{r getNormNL, echo=F}
-# output from code/scale_norms.R
-
-load('analysis/normalizations.nl.Rda')
-knitr::kable(norm.nl.data.out, 'simple', caption='Linear scale models')
-```
+\newpage
 

### 2025-03-11 code/fix_authors.sh

@@ -1,12 +1,10 @@
 #!/bin/zsh
 
-Rscript code/markdown.R
-
-awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin D. Silverman$^{2,3,4}$")};1' go3.tex > jnk.tex
+awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin D. Silverman$^{2,3,4,5}$")};1' go3.tex > jnk.tex
 
 awk '{sub(/\\date{}/,"\\date{\\today}")};1' jnk.tex > jnk2.tex
 
-awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$ Department of Biochemistry, University of Western Ontario\n\$^2\$ College of Information Sciences and Technology, Pennsylvania  State University\n\$^3\$ Department of Medicine, Pennsylvania  State University\n\$^4\$ Institute for Computational and Data Science, Pennsylvania State University\n")};1' jnk2.tex > go3-author.tex
+awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$Department of Biochemistry, University of Western Ontario\n\$^2\$Department of Population Health Sciences, Geisinger, Danville, PA\n\$^3\$Department of Medicine, Pennsylvania State University\n\$^4\$Institute for Computational and Data Science, Pennsylvania State University\n\$^5\$Department of Statistics, Pennsylvania State University\n")};1' jnk2.tex > go3-author.tex
 
 pdflatex go3-author.tex -o go3-author.pdf
 bibtex go3-author
@@ -14,6 +12,7 @@ pdflatex go3-author.tex -o go3-author.pdf
 
 rm jnk.tex
 rm jnk2.tex
+rm go3.pdf
 rm go3.aux
 rm go3.out
 rm go3.toc

### 2025-03-11 code/meta.DESeq.R

@@ -4,7 +4,7 @@ load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.Rda'))
 
 conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8)) 
 
-# DESeq2
+# DESeq2 
 library(DESeq2)
 
 # make a vector of conditions

### 2025-03-11 code/selex.aldex.mu.R

@@ -1,6 +1,13 @@
 library(ALDEx2)
 data(selex)
 
+# this is pretty counter-intuitive but here we go
+# we start by getting our prior values from the point clr
+# we need to do this to log(e)  because the rlnorm function
+# uses this base.  
+# tldr: the makeScaleMatrix function is messed up and needs fixing
+
+
 conds <- c(rep('NS',7), rep('S',7))
 
 sel.all <- aldex(selex, conditions=conds, CI=T, gamma=0.5)

### 2025-03-11 code/yst-DESeq.R

@@ -18,4 +18,3 @@ dds <- DESeqDataSetFromMatrix(countData = yst,
 dds <- DESeq(dds)
 res <- results(dds, name="conds_W_vs_S")
 save(res, file='analysis/res.Rda')
-

### 2025-03-11 code/yst-FP.R

@@ -1,6 +1,6 @@
 #devtools::load_all('~/Documents/0_git/ALDEx_bioc')
 library(ALDEx2)
-
+### permutations below se li Li:2022aa
 url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
 yst <- read.table(url, header=T, row.names=1)
 # remove the one gene with 0 reads
@@ -108,4 +108,34 @@ save(x.sub, file="analysis/x.sub.Rda")
 save(x.ref, file="analysis/x.ref.Rda")
 save(xs.sub, file="analysis/xs.sub.Rda")
 save(xs.ref, file="analysis/xs.ref.Rda")
+
+
+#### Try some permutations
+#### neither ALDEx2 nor DESeq2 returned FP on permutation
+nperms <- 25
+tally.BH <- vector(mode='numeric', length=nperms)
+tally.p <- vector(mode='numeric', length=nperms)
+tally.DES <- vector(mode='numeric', length=nperms)
+
+for(i in 1:nperms){
+yst.perm <- yst[,sample(colnames(yst))]
+x.p <- aldex.clr(yst.perm, conds, gamma=1e-3)
+x.pt <- aldex.ttest(x.p)
+aldex.plot(x.pall)
+tally.BH[i] <- sum(x.pt$we.eBH < 0.05)
+tally.p[i] <- sum(x.pt$we.ep < 0.05)
+
+dds <- DESeqDataSetFromMatrix(countData = yst.perm,
+          colData = coldata, design= ~ conds)
+dds <- DESeq(dds)
+meta.DES.res <- results(dds, name="conds_W_vs_S")
+tally.DES[i] <- sum(meta.DES.res@listData$p.adj < 0.05)
+}
+
+
+dds <- DESeqDataSetFromMatrix(countData = sample(yst),
+          colData = coldata, design= ~ conds)
+dds <- DESeq(dds)
+meta.DES.res <- results(dds, name="conds.W_vs_S")
+sum(meta.DES.res@listData$p.adj < 0.05)
 

### 2025-04-14 response.Rmd
Penultimate before cleaning

---
title: "Response to Reviewers"
author:
  - name: Greg Gloor 
    email: ggloor@uwo.ca 
cite-method: citeproc 
citeproc: TRUE 
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib 
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
output:
    pdf_document:
      keep_tex: true
header-includes:
  - \usepackage{setspace}
---

Reviewer: 1

Comments to the Author

Gloor, Nixon, and Silverman provide a clear and concisely written manuscript that describes a significant improvement of the ALDEx2 package to incorporate scale uncertainty in analyses of differential expression and abundance. These scale models relieve the reliance on commonly used but difficult to rationalize cutoffs of q-values and log-fold change thresholds, and facilitate adoption through the ALDEx2 R package. This work represents a valuable addition to an already impactful and versatile tool for the community. I have several minor comments that I am confident can be easily addressed by the authors.

1. The authors focus here on the application of scale-informed ALDEx2 to transcriptomic analysis. Can the authors further discuss whether applications to other types of measurements is equally appropriate? If so, there may be an opportunity to generalize the title to use “counts” instead of “RNA-sequencing”, as that broadens applicability to other data types (e.g. proteomics, single cell epigenome data, etc).

    - We have changed the title to reflect that it is count data, and have included information on other data types in the introduction. However, the focus of this report is intended to be squarely on transcriptome and metatranscriptome data in an attempt to more tightly target that analytic audience. Companion articles by by McGovern et al. [@McGovern:2023] discussing the use in gene set enrichment and by  Nixon et al. [@Nixon2024B] targeting the microbiome community are published in PLoS Comp. Bio. and under second revision at Genome Biol. respectively.

2. Relevant to statements like this: “The original naive ALDEx2 (39) model unwittingly made a strict assumption about scale through the CLR normalization (3).” - Can “scale-naïve” be used instead to more precisely indicate the distinction between the previous and current defaults in ALDEx2?

    - We thank the reviewer for this suggestion of more precise and descriptive terminology.

3. Why this threshold in Figure 1? “The horizontal dashed lines represent a log2 difference of ±1.4”
The text goes on to say the following: (line 41-44, pg 6): “Note that there is considerable variation in recommended cutoff values(14). Here, applying a dual-cutoff using a heuristic of at least a 2^1.4 fold change reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line with the recommendations of (14) with the fold change limits shown by the dashed grey lines in Figure 1.” - Please add a rationale for selection of this cutoff, and briefly state the guideline used here, especially given the variation in recommendations in ref 14.

    - We agree that the explanation for this cutoff was not well explained, and now include a section in the results explaining that all fold-change cutoffs are arbitrarily chosen and this one was chosen to be comparable in number of transcripts identified to our gamma parameter. There is now an example in the text (Figure 3) and in the supplement (Sup. Figure 2) showing that adding scale uncertainty is superior in controlling FDR when compared to fold-change cutoffs.(lines 19-20, 160-169)

4. Supplementary Fig 1 is an excellent demonstration of how scale uncertainty affects the number of DE genes – the recommendation in that section that this is be used a diagnostic plot should be emphasized in the main text as well, possibly on line 50-51, or in the discussion.

    - Thank you we have enlarged the discussion of this and included a second example using a biologically replicated dataset with modeled TP information.

5. This dataset (ref 44) included many technical replicates.  How do these conclusions port to datasets that have biological replicates instead?

    - we have included two worked examples that use biological datasets from human-derived studies along with modeled TP data. (lines 177-237)

6. Figure 2B, and pg 9 lines 11-14: “Figure 2B shows a plot of the difference between the \(\gamma = 0\) and \(\gamma = 1\) data and here we can see that scale uncertainty is preferentially increasing the dispersion of the mid-expressed transcripts that formerly had negligible dispersion; examine the grey line of best fit (overlaid by the red line) for the trend.” - Can the 5 lines be separately shown on 5 panels, rather than overlaid? They are difficult to visually distinguish as presented.

    - We realize that this figure was confusing in concept, and have moved it to the supplement as Supp. Figure 5. This figure is now only discussed in support of figures that we believe better make the point that there is not an across-the-board change to dispersion and hence p-values. This is now better described by Figure 2 panels C and D, and Supp. Figure 3 panels C and D. This information is now only briefly discussed in lines 243-245, and is no longer central to our argument.

7. Is the conclusion from the first part of the results that by adding scale uncertainty (\(\gamma =1\)), one can do without setting a change threshold (T=1.4 in this case).  This should be emphasized in the discussion.

    - Yes, and the inclusion of the modeled data now makes this point explicitly. The discussion has been simplified to more clearly highlight this point. We are explicit in lines 246-251, 367-374 and 381-388

8. The examples outlined in the results would benefit from a more explicit description of key experimental features – e.g. the number of technical replicates in the ref 44 dataset (mentioned in discussion, but would be useful in results); same for the ref 50 data, so the reader can easily contextualize the results, and draw parallels to their datasets of interest to which ALDEx2 may be applicable.

    - We have included more description of the datasets and made clear what is technical and what is biological replication especially at lines 128-134, 177-185, 253-259

Some typos:

Line 25, pg 10, typo: "functions tp be nearly invariant"

Grammatical error: This assumption was often close enough to the true value to be useful, but was not always the a good estimate and could be outperformed by other normalizations (40).

  - Thank you, we have tried to catch all typos and grammatical errors.

Reviewer: 2

Comments to the Author
Summary:
In this manuscript Gloor et al. highlight the application of ALDEx2 and the recent addition of scale models to help improve differential expression analysis in RNA seq experiments. This manuscript builds upon previous work that introduced scale models to ALDEx2 and highlights that differential analysis in high throughput sequencing often suffers from scale assumptions introduced during commonly applied normalization techniques. In this manuscript they used both a transcriptomic and metatranscriptomic experiment to highlight the improvements that can be achieved by using the scale models incorporated in ALDEx2. The manuscript is well written and highlights an important problem within the field, however, there are some major comments that should be addressed by the authors.

Major comments:

1. Throughout the manuscript a clear definition of what the “truth” is for each experimental set up is needed (i.e., is it transcripts relative to species abundance or bulk absolute transcripts or something else?). This is critically important in the metatranscriptomics section where high levels of differing biomass could result in higher levels of absolute transcripts across the board but not higher levels of transcripts relative to the underlying species abundance. The attempt to center the data on housekeeping genes suggests the latter is the objective but this is not clear throughout the manuscript. In addition to this if the authors are interested in identifying the transcript levels relative to species abundance, they may be interested in comparing their tests against a method that normalizes transcript levels against underlying DNA abundances(https://pubmed.ncbi.nlm.nih.gov/34465175/).

    - We have added two semi-synthetic datasets to show how the FDR is controlled much better by scale uncertainty than by fold-change cutoffs. Ultimately, the choice of what to use as the baseline for normalization of scale between samples is a choice that must be made by the analyst. In some situations the biological question may be best addressed by normalizing transcript levels to species abundance (although to be clear, in the method suggested both the RNA and species levels are relative abundances). However, such a normalization is discussed with reference to species-level metatranscriptomics rather than systems-level metatranscriptomics as shown in our report and in [@Dos-Santos:2024aa]. We would need to redo all of the analysis in the cited paper in order to make this comparison. Indeed, the scale approach advocated in this report can use any measure of absolute abundance as an anchor and we have been explicit throughout about this. (lines 177-237, 377-380)

2. The claims in Figure 1 should be further validated through simulated data and or a dataset where the true underlying number of differentially expressed transcripts is known. While the reduction in transcripts that the scale model achieves is likely inline with the biology it is unclear if the model is being overconservative.

    - We thank the reviewer for this suggestion while the original paper by Nixon et al. [@nixon2024scale] included extensive simulation and theoretic justification we recognize that this is not yet published and a clear demonstration of the utility in transcriptome datasets would be helpful. We have now included two simulated datasets from a recent benchmarking paper [2Li:2022aa] showing that the scale-reliant approach is substantially better in controlling the FDR and, at least in one example, has minimal effect on power (Fig 2, Supp Fig 3). This has resulted in a major re-write of the results section and has made a substantial improvement for how we understand scale uncertainty and its effect on data analys. We again thank the reviewer for pushing us to do this analysis. (lines 177-237)

3. In the metatranscriptomics section it is claimed that biomass differs by 20 fold between the groups that are being compared, however, the scaled model only includes a difference of 14%. Please address this discrepancy. Furthermore, based on Figure 3B, the gamma model by itself was ineffective in this setting, so the manuscript should clarify when only the gamma parameter versus the gamma and mu parameters must be informed.

    -  We have tried to include more insights in the results (Supp Figure 4) and the discussion around when to use an informed vs a simple scale model. In essence, the centre of mass of the difference between groups should be close to 0 (no difference) to avoid FP and FN [@Wu2021]. (lines 358-365, 398-410)

4. Due to the limited number of datasets shown in this manuscript the robustness of the model for various biological and technical conditions is difficult to assess.

  - We have added two additional datasets showing generality. (see above)

5. Data availability: Scripts used to run the analysis (beyond just the model itself) should be provided in a well-documented GitHub (or equivalent).

  - Thank you for pointing out this oversight. The entire analysis is on github and I forgot to flip this to public access.

Minor comments:
1. Without external information, the Lambda distribution appears entirely determined by the prior. It is unclear how robust the choice of Lambda is across differing HTS experiments. Fig. 3 seems to provide a case in which lambda alone is not enough.

  - We have added more guidance and examples on how to choose appropriate parameters as outlined in the response to reviewer 1. (lines 398-410)

2. “The computational methods developed for microbiome analysis have low correlation with the actual scale but are useful (8).”
Well this may be true it has been shown that in general coefficients from the relative scale correlate strongly with those from the absolute scale. Especially in cases where biomass is not significantly changing between groups. This should be addressed.

  - Unfortunately, this statement is only true in the context of the machine learning model in microbiome analyses. This statement is intended to point out that the inclusion of differences in the location in scale between groups *coupled with uncertainty in scale* are robust. This is also the intent of the Supplementary Figure 7 showing that 14% and 5% scale offset are essentially equivalent as long a some uncertainty is accepted.

3. The authors claim that “the dispersion in the unscaled analysis in Figure 2A reaches a minimum near the mid-point of the distribution” which “makes the counterintuitive suggestion that the variance in expression of the majority of genes with moderate expression is more predictable than highly-expressed genes or of housekeeping genes.” However, the dispersion seems essentially constant beyond the midpoint, with potentially a small difference in the tail. Whether or not that is biologically important and to what degree that impacts traditional differential expression analysis is unclear and not presented in a convincing manner. I wouldn’t say this small change in the tail is inconsistent with the notation that  sufficiently expressed genes all have equally predictable expression.

  - we agree that this figure was not optimal for the points we were trying to make. This figure has been moved to the supplement and de-emphasized. As indicated in the response to reviewer 1, the point that the addition of scale uncertainty is affecting transcripts based on dispersion and not on difference between is now made much better by Figure 2 and Supplementary Figure 3. (lines 19-20, 160-169)

4. In the third paragraph of the discussion the authors claim that their method can control type 1 and 2 errors robustly. However, the authors never directly present any evidence that the “hits” their tools are finding are better at controlling these errors than others when a known ground truth is present (they only show that they have reduced numbers of hits and that they may be more reasonable with the underlying biology). The authors may want to benchmark their work on previous simulated data from ref. 18 if they want to make this claim.

    - We have now included two modelled datasets to bolster this claim with biological data and modeled true positive (TP) transcripts.

5. The clause “wet-lab protocols only provide information on the size of the data downstream of the step in the sample preparation protocol where the intervention was made” should be reviewed as previous works have shown that these methods can give relatively accurate log fold changes under real experimental conditions.

    - We agree that spike-ins can be useful, and indeed the use of full scale models is a general tool to do so. However, to give a trivial example, spiking in at the time the RNA library is made will not control for the number of cells in the environment.

6. The authors should make it clear that their works are referring to the absolute and not compositional scale as some researchers may be interested in the composition of transcripts within their sample rather than attempting to infer absolute changes from that composition (which is what is highlighted within this works).

    - We have included additional information around what scale is and what we are trying to measure. Figure 1 is now included to make this clearer. (line 48 now states this explicitly and we give examples throughout regarding what absolute abundance information can be included)

7. In Fig. 2, what is rAbundance referring to? Is it log2(percent abundance)?

    - this is now supplementary Figure 5 and the rAbundance term has been clarified

8. The Fig. 3 caption is inconsistent with the legend in the top of 3A. The text caption seems to be correct, but this should be fixed.

    - We have added information to the legend in now Figure 4 that the first panel has 0 scale uncertainty.

8. I’m not entirely convinced that low dispersion is to blame for issues presenting in HTS data. The example given was of technical replicates which will naturally have lower dispersion rates than true biological replicates. I would suggest that the authors include some evidence that low dispersion is present across biological replicates if they want to make the claim that it is one of the largest sources of HTS data pathologies.

    - We have now included two biological datasets and modeling data

9. In the metatranscriptomic’s example it is not entirely clear why housekeeping genes would have a non-zero log fold change between the groups in the first place. Well the authors suggest that misidentification of these genes is due to inappropriate scale assumptions, it would help to identify what assumptions made by the normalization result in this potential pathology to begin with.

    - The reviewer is correct in that HK genes may not have non-0 LFC, but that the assumption that is almost always made is that these are appropriate references. Note that the analysis in the metatranscriptome is not at the gene level, but at the aggregated functional level. We have tried to clarify that the difference in occurrence is a major driver of this asymmetry.  (lines 358-365)

10. Could the authors comment on whether they believe their scale model implementation will work well under other normalization conditions other than CLR.

    - Nixon et al. show that in theory, any interpretable log-ratio will work with this schema and other papers are in preparation showing this. For the purposes of this manuscript the CLR is the normalization used by ALDEx2

Associate Editor: Erb, Ionas
Comments to the Author:
Dear authors,

Please focus your revision on the following aspects: State what the true scale of the data should describe, i.e., number of transcripts relative to species abundance or similar. Does this change between data sets?  Include a comparison with other approaches, e.g., comparing with an approach that uses DNA abundance as described in the paper in Annual Reviews (see reviewer 2).

  - we have added a paragraph in the discussion about this. Fundamentally, while the scale approach is generalizable and could be applied to the types of data given in the review paper (the actual method is in [@Zhang:2021ab]), but that approach is designed to ask a different type of question than is addressed by scale. (Lines 374-380 and above)

A comparison with the use of a moderated statistic (see, e.g.,  the classical Smyth 2004 paper) which also addresses the low dispersion problem in data with low sample size would add considerable value, even if only as a discussion.

   - we have added some discussion of the moderated approach in the discussion. However, fundamentally this approach is trying to solve a problem of missing information in the data. The use of scale uncertainty directly addresses this issue because it adds in uncertainty around the missing information rather than try to model it from the data. This is the fundamental problem of partially specified models that we bring into the introduction, when the information is not available, one is better to acknowledge this rather than to try and pull it out of data that is missing it. (lines 51-53)

Additionally, the inclusion of a simulated or otherwise known ground truth (as pointed out by reviewer 2) would validate the claims in Figure 1.

  - This has been done with two different datasets and the result has significantly strengthend the paper and our interpretations

Discuss effect of biological replicates vs. technical replicates (see both reviewers).

  - This has been added in multiple places and again the introduction of the simulated dataset has helped with this point.

Please indicate if the preprints referenced in (3) and (20) are submitted / currently under review.

  - The companion articles by by McGovern et al. [@McGovern:2023] discussing the use in gene set enrichment and by  Nixon et al. [@Nixon2024B] targeting the microbiome community are published in PLoS Comp. Bio. and under second revision at Genome Biol. respectively. The companion article introducing the theory and the initial modeling by Nixon et al. [@nixon2024scale] is still under review at Annals of applied Statistics which is a frustration for the authors as much as the reviewers

### 2025-04-14 response.tex
Penultimate before cleaning

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Response to Reviewers},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Response to Reviewers}
\author{true}
\date{}

\begin{document}
\maketitle

Reviewer: 1

Comments to the Author

Gloor, Nixon, and Silverman provide a clear and concisely written
manuscript that describes a significant improvement of the ALDEx2
package to incorporate scale uncertainty in analyses of differential
expression and abundance. These scale models relieve the reliance on
commonly used but difficult to rationalize cutoffs of q-values and
log-fold change thresholds, and facilitate adoption through the ALDEx2 R
package. This work represents a valuable addition to an already
impactful and versatile tool for the community. I have several minor
comments that I am confident can be easily addressed by the authors.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The authors focus here on the application of scale-informed ALDEx2 to
  transcriptomic analysis. Can the authors further discuss whether
  applications to other types of measurements is equally appropriate? If
  so, there may be an opportunity to generalize the title to use
  ``counts'' instead of ``RNA-sequencing'', as that broadens
  applicability to other data types (e.g.~proteomics, single cell
  epigenome data, etc).

  \begin{itemize}
  \tightlist
  \item
    We have changed the title to reflect that it is count data, and have
    included information on other data types in the introduction.
    However, the focus of this report is intended to be squarely on
    transcriptome and metatranscriptome data in an attempt to more
    tightly target that analytic audience. Companion articles by by
    McGovern et al. (1) discussing the use in gene set enrichment and by
    Nixon et al. (2) targeting the microbiome community are published in
    PLoS Comp. Bio. and under second revision at Genome Biol.
    respectively.
  \end{itemize}
\item
  Relevant to statements like this: ``The original naive ALDEx2 (39)
  model unwittingly made a strict assumption about scale through the CLR
  normalization (3).'' - Can ``scale-naïve'' be used instead to more
  precisely indicate the distinction between the previous and current
  defaults in ALDEx2?

  \begin{itemize}
  \tightlist
  \item
    We thank the reviewer for this suggestion of more precise and
    descriptive terminology.
  \end{itemize}
\item
  Why this threshold in Figure 1? ``The horizontal dashed lines
  represent a log2 difference of ±1.4'' The text goes on to say the
  following: (line 41-44, pg 6): ``Note that there is considerable
  variation in recommended cutoff values(14). Here, applying a
  dual-cutoff using a heuristic of at least a 2\^{}1.4 fold change
  reduces the number of significant outputs to 193 for DESeq2 and to 186
  for ALDEx2. This cutoff was chosen for convenience and is in-line with
  the recommendations of (14) with the fold change limits shown by the
  dashed grey lines in Figure 1.'' - Please add a rationale for
  selection of this cutoff, and briefly state the guideline used here,
  especially given the variation in recommendations in ref 14.

  \begin{itemize}
  \tightlist
  \item
    We agree that the explanation for this cutoff was not well
    explained, and now include a section in the results explaining that
    all fold-change cutoffs are arbitrarily chosen and this one was
    chosen to be comparable in number of transcripts identified to our
    gamma parameter. There is now an example in the text (Figure 3) and
    in the supplement (Sup. Figure 2) showing that adding scale
    uncertainty is superior in controlling FDR when compared to
    fold-change cutoffs.(lines 19-20, 160-169)
  \end{itemize}
\item
  Supplementary Fig 1 is an excellent demonstration of how scale
  uncertainty affects the number of DE genes -- the recommendation in
  that section that this is be used a diagnostic plot should be
  emphasized in the main text as well, possibly on line 50-51, or in the
  discussion.

  \begin{itemize}
  \tightlist
  \item
    Thank you we have enlarged the discussion of this and included a
    second example using a biologically replicated dataset with modeled
    TP information.
  \end{itemize}
\item
  This dataset (ref 44) included many technical replicates. ~How do
  these conclusions port to datasets that have biological replicates
  instead?

  \begin{itemize}
  \tightlist
  \item
    we have included two worked examples that use biological datasets
    from human-derived studies along with modeled TP data. (lines
    177-237)
  \end{itemize}
\item
  Figure 2B, and pg 9 lines 11-14: ``Figure 2B shows a plot of the
  difference between the \(\gamma = 0\) and \(\gamma = 1\) data and here
  we can see that scale uncertainty is preferentially increasing the
  dispersion of the mid-expressed transcripts that formerly had
  negligible dispersion; examine the grey line of best fit (overlaid by
  the red line) for the trend.'' - Can the 5 lines be separately shown
  on 5 panels, rather than overlaid? They are difficult to visually
  distinguish as presented.

  \begin{itemize}
  \tightlist
  \item
    We realize that this figure was confusing in concept, and have moved
    it to the supplement as Supp. Figure 5. This figure is now only
    discussed in support of figures that we believe better make the
    point that there is not an across-the-board change to dispersion and
    hence p-values. This is now better described by Figure 2 panels C
    and D, and Supp. Figure 3 panels C and D. This information is now
    only briefly discussed in lines 243-245, and is no longer central to
    our argument.
  \end{itemize}
\item
  Is the conclusion from the first part of the results that by adding
  scale uncertainty (\(\gamma =1\)), one can do without setting a change
  threshold (T=1.4 in this case). ~This should be emphasized in the
  discussion.

  \begin{itemize}
  \tightlist
  \item
    Yes, and the inclusion of the modeled data now makes this point
    explicitly. The discussion has been simplified to more clearly
    highlight this point. We are explicit in lines 246-251, 367-374 and
    381-388
  \end{itemize}
\item
  The examples outlined in the results would benefit from a more
  explicit description of key experimental features -- e.g.~the number
  of technical replicates in the ref 44 dataset (mentioned in
  discussion, but would be useful in results); same for the ref 50 data,
  so the reader can easily contextualize the results, and draw parallels
  to their datasets of interest to which ALDEx2 may be applicable.

  \begin{itemize}
  \tightlist
  \item
    We have included more description of the datasets and made clear
    what is technical and what is biological replication especially at
    lines 128-134, 177-185, 253-259
  \end{itemize}
\end{enumerate}

Some typos:

Line 25, pg 10, typo: ``functions tp be nearly invariant''

Grammatical error: This assumption was often close enough to the true
value to be useful, but was not always the a good estimate and could be
outperformed by other normalizations (40).

\begin{itemize}
\tightlist
\item
  Thank you, we have tried to catch all typos and grammatical errors.
\end{itemize}

Reviewer: 2

Comments to the Author Summary: In this manuscript Gloor et
al.~highlight the application of ALDEx2 and the recent addition of scale
models to help improve differential expression analysis in RNA seq
experiments. This manuscript builds upon previous work that introduced
scale models to ALDEx2 and highlights that differential analysis in high
throughput sequencing often suffers from scale assumptions introduced
during commonly applied normalization techniques. In this manuscript
they used both a transcriptomic and metatranscriptomic experiment to
highlight the improvements that can be achieved by using the scale
models incorporated in ALDEx2. The manuscript is well written and
highlights an important problem within the field, however, there are
some major comments that should be addressed by the authors.

Major comments:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Throughout the manuscript a clear definition of what the ``truth'' is
  for each experimental set up is needed (i.e., is it transcripts
  relative to species abundance or bulk absolute transcripts or
  something else?). This is critically important in the
  metatranscriptomics section where high levels of differing biomass
  could result in higher levels of absolute transcripts across the board
  but not higher levels of transcripts relative to the underlying
  species abundance. The attempt to center the data on housekeeping
  genes suggests the latter is the objective but this is not clear
  throughout the manuscript. In addition to this if the authors are
  interested in identifying the transcript levels relative to species
  abundance, they may be interested in comparing their tests against a
  method that normalizes transcript levels against underlying DNA
  abundances(\url{https://pubmed.ncbi.nlm.nih.gov/34465175/}).

  \begin{itemize}
  \tightlist
  \item
    We have added two semi-synthetic datasets to show how the FDR is
    controlled much better by scale uncertainty than by fold-change
    cutoffs. Ultimately, the choice of what to use as the baseline for
    normalization of scale between samples is a choice that must be made
    by the analyst. In some situations the biological question may be
    best addressed by normalizing transcript levels to species abundance
    (although to be clear, in the method suggested both the RNA and
    species levels are relative abundances). However, such a
    normalization is discussed with reference to species-level
    metatranscriptomics rather than systems-level metatranscriptomics as
    shown in our report and in (3). We would need to redo all of the
    analysis in the cited paper in order to make this comparison.
    Indeed, the scale approach advocated in this report can use any
    measure of absolute abundance as an anchor and we have been explicit
    throughout about this. (lines 177-237, 377-380)
  \end{itemize}
\item
  The claims in Figure 1 should be further validated through simulated
  data and or a dataset where the true underlying number of
  differentially expressed transcripts is known. While the reduction in
  transcripts that the scale model achieves is likely inline with the
  biology it is unclear if the model is being overconservative.

  \begin{itemize}
  \tightlist
  \item
    We thank the reviewer for this suggestion while the original paper
    by Nixon et al. (4) included extensive simulation and theoretic
    justification we recognize that this is not yet published and a
    clear demonstration of the utility in transcriptome datasets would
    be helpful. We have now included two simulated datasets from a
    recent benchmarking paper {[}2Li:2022aa{]} showing that the
    scale-reliant approach is substantially better in controlling the
    FDR and, at least in one example, has minimal effect on power (Fig
    2, Supp Fig 3). This has resulted in a major re-write of the results
    section and has made a substantial improvement for how we understand
    scale uncertainty and its effect on data analys. We again thank the
    reviewer for pushing us to do this analysis. (lines 177-237)
  \end{itemize}
\item
  In the metatranscriptomics section it is claimed that biomass differs
  by 20 fold between the groups that are being compared, however, the
  scaled model only includes a difference of 14\%. Please address this
  discrepancy. Furthermore, based on Figure 3B, the gamma model by
  itself was ineffective in this setting, so the manuscript should
  clarify when only the gamma parameter versus the gamma and mu
  parameters must be informed.

  \begin{itemize}
  \tightlist
  \item
    We have tried to include more insights in the results (Supp Figure
    4) and the discussion around when to use an informed vs a simple
    scale model. In essence, the centre of mass of the difference
    between groups should be close to 0 (no difference) to avoid FP and
    FN (5). (lines 358-365, 398-410)
  \end{itemize}
\item
  Due to the limited number of datasets shown in this manuscript the
  robustness of the model for various biological and technical
  conditions is difficult to assess.
\end{enumerate}

\begin{itemize}
\tightlist
\item
  We have added two additional datasets showing generality. (see above)
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Data availability: Scripts used to run the analysis (beyond just the
  model itself) should be provided in a well-documented GitHub (or
  equivalent).
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Thank you for pointing out this oversight. The entire analysis is on
  github and I forgot to flip this to public access.
\end{itemize}

Minor comments: 1. Without external information, the Lambda distribution
appears entirely determined by the prior. It is unclear how robust the
choice of Lambda is across differing HTS experiments. Fig. 3 seems to
provide a case in which lambda alone is not enough.

\begin{itemize}
\tightlist
\item
  We have added more guidance and examples on how to choose appropriate
  parameters as outlined in the response to reviewer 1. (lines 398-410)
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  ``The computational methods developed for microbiome analysis have low
  correlation with the actual scale but are useful (8).'' Well this may
  be true it has been shown that in general coefficients from the
  relative scale correlate strongly with those from the absolute scale.
  Especially in cases where biomass is not significantly changing
  between groups. This should be addressed.
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Unfortunately, this statement is only true in the context of the
  machine learning model in microbiome analyses. This statement is
  intended to point out that the inclusion of differences in the
  location in scale between groups \emph{coupled with uncertainty in
  scale} are robust. This is also the intent of the Supplementary Figure
  7 showing that 14\% and 5\% scale offset are essentially equivalent as
  long a some uncertainty is accepted.
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  The authors claim that ``the dispersion in the unscaled analysis in
  Figure 2A reaches a minimum near the mid-point of the distribution''
  which ``makes the counterintuitive suggestion that the variance in
  expression of the majority of genes with moderate expression is more
  predictable than highly-expressed genes or of housekeeping genes.''
  However, the dispersion seems essentially constant beyond the
  midpoint, with potentially a small difference in the tail. Whether or
  not that is biologically important and to what degree that impacts
  traditional differential expression analysis is unclear and not
  presented in a convincing manner. I wouldn't say this small change in
  the tail is inconsistent with the notation that ~sufficiently
  expressed genes all have equally predictable expression.
\end{enumerate}

\begin{itemize}
\tightlist
\item
  we agree that this figure was not optimal for the points we were
  trying to make. This figure has been moved to the supplement and
  de-emphasized. As indicated in the response to reviewer 1, the point
  that the addition of scale uncertainty is affecting transcripts based
  on dispersion and not on difference between is now made much better by
  Figure 2 and Supplementary Figure 3. (lines 19-20, 160-169)
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\item
  In the third paragraph of the discussion the authors claim that their
  method can control type 1 and 2 errors robustly. However, the authors
  never directly present any evidence that the ``hits'' their tools are
  finding are better at controlling these errors than others when a
  known ground truth is present (they only show that they have reduced
  numbers of hits and that they may be more reasonable with the
  underlying biology). The authors may want to benchmark their work on
  previous simulated data from ref. 18 if they want to make this claim.

  \begin{itemize}
  \tightlist
  \item
    We have now included two modelled datasets to bolster this claim
    with biological data and modeled true positive (TP) transcripts.
  \end{itemize}
\item
  The clause ``wet-lab protocols only provide information on the size of
  the data downstream of the step in the sample preparation protocol
  where the intervention was made'' should be reviewed as previous works
  have shown that these methods can give relatively accurate log fold
  changes under real experimental conditions.

  \begin{itemize}
  \tightlist
  \item
    We agree that spike-ins can be useful, and indeed the use of full
    scale models is a general tool to do so. However, to give a trivial
    example, spiking in at the time the RNA library is made will not
    control for the number of cells in the environment.
  \end{itemize}
\item
  The authors should make it clear that their works are referring to the
  absolute and not compositional scale as some researchers may be
  interested in the composition of transcripts within their sample
  rather than attempting to infer absolute changes from that composition
  (which is what is highlighted within this works).

  \begin{itemize}
  \tightlist
  \item
    We have included additional information around what scale is and
    what we are trying to measure. Figure 1 is now included to make this
    clearer. (line 48 now states this explicitly and we give examples
    throughout regarding what absolute abundance information can be
    included)
  \end{itemize}
\item
  In Fig. 2, what is rAbundance referring to? Is it log2(percent
  abundance)?

  \begin{itemize}
  \tightlist
  \item
    this is now supplementary Figure 5 and the rAbundance term has been
    clarified
  \end{itemize}
\item
  The Fig. 3 caption is inconsistent with the legend in the top of 3A.
  The text caption seems to be correct, but this should be fixed.

  \begin{itemize}
  \tightlist
  \item
    We have added information to the legend in now Figure 4 that the
    first panel has 0 scale uncertainty.
  \end{itemize}
\item
  I'm not entirely convinced that low dispersion is to blame for issues
  presenting in HTS data. The example given was of technical replicates
  which will naturally have lower dispersion rates than true biological
  replicates. I would suggest that the authors include some evidence
  that low dispersion is present across biological replicates if they
  want to make the claim that it is one of the largest sources of HTS
  data pathologies.

  \begin{itemize}
  \tightlist
  \item
    We have now included two biological datasets and modeling data
  \end{itemize}
\item
  In the metatranscriptomic's example it is not entirely clear why
  housekeeping genes would have a non-zero log fold change between the
  groups in the first place. Well the authors suggest that
  misidentification of these genes is due to inappropriate scale
  assumptions, it would help to identify what assumptions made by the
  normalization result in this potential pathology to begin with.

  \begin{itemize}
  \tightlist
  \item
    The reviewer is correct in that HK genes may not have non-0 LFC, but
    that the assumption that is almost always made is that these are
    appropriate references. Note that the analysis in the
    metatranscriptome is not at the gene level, but at the aggregated
    functional level. We have tried to clarify that the difference in
    occurrence is a major driver of this asymmetry. (lines 358-365)
  \end{itemize}
\item
  Could the authors comment on whether they believe their scale model
  implementation will work well under other normalization conditions
  other than CLR.

  \begin{itemize}
  \tightlist
  \item
    Nixon et al.~show that in theory, any interpretable log-ratio will
    work with this schema and other papers are in preparation showing
    this. For the purposes of this manuscript the CLR is the
    normalization used by ALDEx2
  \end{itemize}
\end{enumerate}

Associate Editor: Erb, Ionas Comments to the Author: Dear authors,

Please focus your revision on the following aspects: State what the true
scale of the data should describe, i.e., number of transcripts relative
to species abundance or similar. Does this change between data sets?
Include a comparison with other approaches, e.g., comparing with an
approach that uses DNA abundance as described in the paper in Annual
Reviews (see reviewer 2).

\begin{itemize}
\tightlist
\item
  we have added a paragraph in the discussion about this. Fundamentally,
  while the scale approach is generalizable and could be applied to the
  types of data given in the review paper (the actual method is in (6)),
  but that approach is designed to ask a different type of question than
  is addressed by scale. (Lines 374-380 and above)
\end{itemize}

A comparison with the use of a moderated statistic (see, e.g., ~the
classical Smyth 2004 paper) which also addresses the low dispersion
problem in data with low sample size would add considerable value, even
if only as a discussion.

\begin{itemize}
\tightlist
\item
  we have added some discussion of the moderated approach in the
  discussion. However, fundamentally this approach is trying to solve a
  problem of missing information in the data. The use of scale
  uncertainty directly addresses this issue because it adds in
  uncertainty around the missing information rather than try to model it
  from the data. This is the fundamental problem of partially specified
  models that we bring into the introduction, when the information is
  not available, one is better to acknowledge this rather than to try
  and pull it out of data that is missing it. (lines 51-53)
\end{itemize}

Additionally, the inclusion of a simulated or otherwise known ground
truth (as pointed out by reviewer 2) would validate the claims in Figure
1.

\begin{itemize}
\tightlist
\item
  This has been done with two different datasets and the result has
  significantly strengthend the paper and our interpretations
\end{itemize}

Discuss effect of biological replicates vs.~technical replicates (see
both reviewers).

\begin{itemize}
\tightlist
\item
  This has been added in multiple places and again the introduction of
  the simulated dataset has helped with this point.
\end{itemize}

Please indicate if the preprints referenced in (3) and (20) are
submitted / currently under review.

\begin{itemize}
\tightlist
\item
  The companion articles by by McGovern et al. (1) discussing the use in
  gene set enrichment and by Nixon et al. (2) targeting the microbiome
  community are published in PLoS Comp. Bio. and under second revision
  at Genome Biol. respectively. The companion article introducing the
  theory and the initial modeling by Nixon et al. (4) is still under
  review at Annals of applied Statistics which is a frustration for the
  authors as much as the reviewers
\end{itemize}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-McGovern:2023}
1. McGovern,K.C., Nixon,M.P. and Silverman,J.D. (2023)
\href{https://doi.org/10.1371/journal.pcbi.1011659}{Addressing erroneous
scale assumptions in microbe and gene set enrichment analysis}.
\emph{PLoS Comput Biol}, \textbf{19}, e1011659.

\bibitem[\citeproctext]{ref-Nixon2024B}
2. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
normalization: Incorporating scale uncertainty in microbiome and gene
expression analysis. \emph{bioRxiv},
\href{https://doi.org/10.1101/2024.04.01.587602}{10.1101/2024.04.01.587602}.

\bibitem[\citeproctext]{ref-Dos-Santos:2024aa}
3. Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and Gloor,G.B.
(2024) \href{https://doi.org/10.1186/s40168-024-01992-w}{Vaginal
metatranscriptome meta-analysis reveals functional {BV} subgroups and
novel colonisation strategies}. \emph{Microbiome}, \textbf{12}, 271.

\bibitem[\citeproctext]{ref-nixon2024scale}
4. Nixon,M.P., McGovern,K.C., Letourneau,J., David,L.A., Lazar,N.A.,
Mukherjee,S. and Silverman,J.D. (2024)
\href{https://arxiv.org/abs/2201.03616}{Scale reliant inference}.

\bibitem[\citeproctext]{ref-Wu2021}
5. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
\href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
Compositional asymmetry in high-throughput sequencing datasets}. In
Filzmoser,P., Hron,K., Martìn-Fernàndez,J.A., Palarea-Albaladejo,J.
(eds), \emph{Advances in compositional data analysis: Festschrift in
honour of vera pawlowsky-glahn}. Springer International Publishing,
Cham, pp. 329--346.

\bibitem[\citeproctext]{ref-Zhang:2021ab}
6. Zhang,Y., Thompson,K.N., Huttenhower,C. and Franzosa,E.A. (2021)
\href{https://doi.org/10.1093/bioinformatics/btab327}{Statistical
approaches for differential expression analysis in metatranscriptomics}.
\emph{Bioinformatics}, \textbf{37}, i34--i41.

\end{CSLReferences}

\end{document}

### 2025-04-14 go3-author.tex
Penultimate before cleaning

@@
 \usepackage{setspace}
+\usepackage[left]{lineno}
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2},
+  pdftitle={Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2},
   hidelinks,
@@
 
-\title{Explicit Scale Simulation for analysis of RNA-sequencing with
-ALDEx2}
-\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4}$}
+\title{Explicit Scale Simulation for analysis of RNA-sequencing count
+data with ALDEx2}
+\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{3,4,5}$}
 \date{\today}
@@
 
-$^1$ Department of Biochemistry, University of Western Ontario
-$^2$ College of Information Sciences and Technology, Pennsylvania  State University
-$^3$ Department of Medicine, Pennsylvania  State University
-$^4$ Institute for Computational and Data Science, Pennsylvania State University
+$^1$Department of Biochemistry, University of Western Ontario
+$^2$Department of Population Health Sciences, Geisinger, Danville, PA
+$^3$Department of Medicine, Pennsylvania State University
+$^4$Institute for Computational and Data Science, Pennsylvania State University
+$^5$Department of Statistics, Pennsylvania State University
 
@@
 
-\doublespacing
+\doublespacing \singlespacing 
+\linenumbers
 
@@
 acid is carried over to subsequent steps. Thus, with each sampling step
-the connection between the size of the sampled DNA pool and the scale
-(e.g., size, microbial load, or total gene expression) of the measured
-biological system is degraded or lost. In the end, the information
-contained in the data relates only to relative abundances and has an
-arbitrary scale imposed by the sequencing process(1--3). Increasingly,
-researchers are turning to modified experimental protocols in an attempt
-to recover the lost biological variation in scale; see for example (4,
-5). However, these protocols only recover scale variation at the
-intermediate step in the sample preparation protocol where the
-intervention was made. In short, the disconnect between sample-to-sample
-variation in sequencing depth and biological variation in scale remains
-an open challenge.
+the connection between the actual number of molecules in the sampled DNA
+pool and the environmental scale (e.g., total number of molecules,
+microbial load, or total gene expression) of the measured biological
+system is degraded or lost. In the end, the information contained in the
+data relates only to relative abundances and has an arbitrary scale
+imposed by the sequencing process (1--3).
 
 The analysis of HTS data suffers from several known problems that can be
-traced, in whole or in part, to misspecification of scale. The first
-issue is poor control of the false discovery rate (FDR) (6--9),
-exhibited as dataset-dependent FDR control and by the disconnect between
-statistical and biological significance(10). In current practice, these
-issues are addressed by a dual-filtering method, whereby both a low
-p-value (or equivalently a low q-value following FDR correction (11))
-and a large difference between groups is used to identify interesting
-transcripts or genes for follow-up analysis(10). This double-filtering
-approach is graphically exemplified by the volcano plot (12), but is
-known to not appropriately control the FDR\\
-(13, 14). The second issue is poor performance when analyzing data where
-the mean change between groups is non-zero. Such asymmetric data can
+traced, in whole or in part, to misspecification of scale in the output
+data. The first issue is poor control of the false discovery rate (FDR)
+(4--8), exhibited as dataset-dependent FDR control which is observed as
+a disconnect between statistical and biological significance (8, 9). In
+current practice, this issue is addressed by a dual-filtering method,
+whereby both a low p-value (or equivalently a low q-value following FDR
+correction (10)) and a large difference between groups is used to
+identify transcripts or genes of interest for follow-up analysis (9,
+11). This double-filtering approach is graphically exemplified by the
+volcano plot (11), but is known to not appropriately control the FDR
+(12, 13). Since there is no standard way of determining what fold-change
+cutoff should be used researchers have unlimited degrees of freedom
+which is known to lead to unreliable inference (14). In particular Li et
+al. (8) used patient or clinical-derived transcriptome datasets and
+found that many methods suffer from an extremely high false positive
+rate. The second issue is poor performance when analyzing data where the
+mean change between groups is non-zero (3). Such asymmetric data can
 arise when a gene set is expressed in one group but not the other, or
@@
 transcriptome analysis, and microbiome analysis (15). The third issue is
-that these problems become more pronounced as more samples are
-collected; that is, more information results in a worsening of the
-accuracy of the analysis (16).
-
-The three problems were recently shown by Nixon and Silverman (3) to be
-a result of a mismatch between the underlying size or scale of the
-system and the assumptions of the normalizations used for the analysis
-of HTS. Biological variation in scale often represents an important
-unmeasured confounder in HTS analyses (17). For example, cells
+that the actual scale of the environment is often a major confounding
+variable during analysis (3, 16). This has long been known in
+transcriptome analysis and was a major driver for the development of
+normalization factors (17) and the use of molecular spike-ins to provide
+a reference set of known counts (18--21) to estimate scale. While often
+useful, spike-in methods only provide information downstream of the step
+in the sample preparation protocol where the intervention was made and
+introduce an additional source of variation that must be accounted for
+(19). In the microbiome field, a recent landmark paper showed that
+biological scale was a major unacknowledged confounder in many human
+analyses (16). These authors built a machine learning model to uncover
+the biological variation in scale and including this information was
+useful despite exhibiting only a modest correlation (estimated to be
+0.6) with the scale of data external to the training set (16). The final
+issue is that all the above problems become more pronounced as more
+samples are collected; that is, more information results in optimizing
+for a precise but inaccurate analysis (3, 8, 22).
+
+The four problems were recently shown by Nixon et al. (3) to be a result
+of a mismatch between the underlying size or scale of the system and the
+assumptions of the normalizations used for the analysis of HTS.
+Biological variation in scale often represents an important unmeasured
+confounder in HTS analyses (15, 16, 19, 23). For example, cells
 transformed by the cMyc oncogene have about 3 times the amount of mRNA
-and about twice the rRNA content than non-transformed cells (18), and
-this dramatically skews transcriptome analysis (19). In addition,
-wild-type and mutant strains of cell lines, yeast or bacteria have
-different growth rates and RNA contents under different conditions,
-which affect our ability to identify truly differentially abundant genes
-(20--22). As another example, the total bacterial load of the vaginal
-microbiome differs by 1-2 orders of magnitude in absolute abundance
-between the healthy and bacterial vaginosis states (23), and the
-composition between these states is dramatically different (24, 25).
-Thus, a full description of any of these systems includes both relative
-change (composition) and absolute abundance (scale). Current methods
-access only the compositional information yet make implicit assumptions
-about the scale (16).
-
-Recently, Nixon et al. (3) showed that the challenge of non-biological
-variation in sequencing depth be viewed as a problem of
+and about twice the rRNA content than non-transformed cells (24), and
+this dramatically skews transcriptome analysis unless spike-in
+approaches are used (19). In addition, wild-type and mutant strains of
+cell lines, yeast or bacteria have different growth rates and RNA
+contents under different conditions, which affect our ability to
+identify truly differentially abundant genes (25--27). As another
+example, the total bacterial load of the vaginal microbiome differs by
+1-2 orders of magnitude in absolute abundance between the healthy and
+bacterial vaginosis states (28), and the cell and RNA composition
+between these states is dramatically different (29, 30). Thus, a full
+description of any of these systems includes both relative change
+(composition) and absolute abundance (scale). Current methods access
+only the compositional information yet make implicit assumptions about
+the scale (22).
+
+Nixon et al. (3) showed that the challenge of non-biological variation
+in sequencing depth could be explained as a problem of
 partially-identified models. They showed that \emph{all} normalizations
 make some assumption about scale but these implicit assumptions are
-often inappropriate and difficult to interpret. Because of this
-different normalizations provide different outputs when applied to the
-same dataset (6, 8, 26--28). Intuitively, normalizations in widespread
+often inappropriate and difficult to interpret. As a result, different
+normalizations provide different outputs when applied to the same
+dataset (4, 6, 17, 31, 32). Intuitively, normalizations in widespread
 use assume that either all samples have the same scale,
-e.g.~proportions, rarefaction (29), RPKM (30, 31), etc; or that a subset
+e.g.~proportions, rarefaction (33), RPKM (34, 35), etc; or that a subset
 of features in one sample can be chosen as a reference to which the
-others are scaled e.g.~the TMM (32), or LVHA (15) or the additive
-log-ratio (33); or that different sub-parts of each sample maintain a
-constant scale across samples e.g.~the RLE (34); or that the geometric
-mean of the parts is appropriate e.g the CLR (35) and its derivatives.
-
-The original naive ALDEx2 (36) model unwittingly made a strict
-assumption about scale through the CLR normalization (3). While this
-assumption could be useful in many cases it could never be exactly true,
-and others have shown that it is not always the best option (37). Nixon
-et al. (3) showed that better scale assumptions resulted in more
-reproducible data analysis including better control of both false
-positive and false negative results. We recently modified ALDEx2 to
-explicitly model the scale over a range of reasonable normalization
-parameters, and showed significant improvements in performance in
-microbiome and in-vitro selection experiments (16). Here, we briefly
-introduce review modifications and then show how scale uncertainty can
-greatly improve modeling in transcriptome and meta-transcriptome
-datasets to provide more robust and reproducible results.
+others are scaled e.g.~the TMM (36), the LVHA (15) qPCR (37), the
+additive log-ratio (38); or that different sub-parts of each sample
+maintain a constant scale across samples e.g.~the RLE (39); or that the
+geometric mean of the parts is appropriate e.g the CLR (40) and its
+derivatives. Notably when the assumption of similar scale across samples
+is not violated, or violated only weakly, any method will provide a
+reasonable analysis. The problem arises when this assumption is
+violated, in which case all tools fail without warning.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.6\linewidth]{./go3_files/figure-latex/scale-explained}
+
+}
+
+\caption{Mismatches between estimated scale and true scale lead to poor estimation of high throughput sequencing data.
+All normalizations used for differential abundance analysis make some strict assumption about the scale of the environment as shown in line 1.
+In this example all normalizations produce a biased estimate of the environmental scale, but estimate C is the closest to the truth.
+Adding uncertainty to normalization C as represented by the distribution in the line 2 leads to less bias and now includes the actual environmental scale in the assumption.
+As shown in line 3 in some cases it may be useful to adjust the centre of the scale uncertainty estimate if the initial normalizations give very poor estimates of the underlying environment.}\label{fig:unnamed-chunk-1}
+\end{figure}
+
+The original scale-naive ALDEx2 (41) model unwittingly made a strict
+assumption about scale through the CLR normalization and we found that
+the CLR was very sensitive to violations of the assumption of scale
+identity between groups (15). Moreover, when the assumption of identity
+was not true the CLR used by ALDEx2 could be outperformed by other
+normalizations in simulation studies (42). As illustrated graphically in
+Figure 1, Nixon et al. (3) showed through simulation that introducing
+uncertainty in the scale assumptions, and in extreme cases altering the
+location of the scale assumption, resulted in more reproducible data
+analysis including better control of both false positive and false
+negative results. We modified ALDEx2 to explicitly model uncertainty in
+scale over a range of reasonable normalization parameters, and showed
+significant improvements in performance in microbiome and in-vitro
+selection experiments (22) and in a vaginal metatranscriptome analysis
+(43). Here, we briefly review these modifications and show how adding
+scale uncertainty can greatly improve modeling in transcriptome and
+meta-transcriptome datasets to provide substantially more robust and
+reproducible results.
 
@@
 Formal and expanded descriptions of the concepts that follow are given
-in (3, 16). To be concrete, we let \(\mathbf{Y}\) denote the
+in (3, 22). To be concrete, we let \(\mathbf{Y}\) denote the
 \emph{measured} \(D \times N\) matrix of sequence counts with elements
@@
 mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
-\(n\). Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount
-of class \(d\) in the biological system from which sample \(n\) was
-obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
-scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
-\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
-\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
-\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
-\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
-\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
-Note that with these definitions \(\mathbf{W}\) can be written as the
+\(n\). Likewise, we can denote \(\mathbf{W}_{dn}\) as the \emph{true}
+amount of class \(d\) in the biological system from which sample \(n\)
+was obtained. We can think of \(\mathbf{W}\) as consisting of two parts,
+the scale \(\mathbf{W}^{Tot}\) (e.g., totals) and the composition
+\(\mathbf{W}^{Comp}\) (i.e., proportions). That is, \(\mathbf{W}^{Tot}\)
+is a \(N\)-vector with elements
+\(\mathbf{W}^{Tot}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
+\(\mathbf{W}^{Comp}\) is a \(D \times N\) matrix with elements
+\(\mathbf{W}^{Comp}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{Tot}_{n}\). Note
+that with these definitions \(\mathbf{W}\) can be written as the
 element-wise combination of scale and composition:
-\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\),
-or as the logarithm
-\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).
-
-Many of the normalizations in widespread use in tools such as DESeq2
-(38), edgeR (32), metagenomeSeq (39) ALDEx2 (40) can be stated as ratios
-of the form
+\(\mathbf{W}_{dn}=\mathbf{W}^{Comp}_{dn}\mathbf{W}^{Tot}_{n}\), or as
+the logarithm
+\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{Comp}_{dn} + \log \mathbf{W}^{Tot}_{n}\).
+
+Many of the normalizations used in tools such as DESeq2 (44), edgeR
+(36), metagenomeSeq (45) ALDEx2 (46) can be stated as ratios of the form
 \(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where
 the denominator is determined by some function of the observation. We
-use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate
-that the output is an estimate of the true value. The technical
-variation in sequencing depth
-(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
-observed data \(\mathbf{Y}\) provides us with information about the
-system composition \(\mathbf{W}^{\parallel}\) but little to no
-information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
-al.~2011).
+use the hat notation (\(\ \hat{}\ \)) to indicate that the output is an
+estimate of the true value. The technical variation in sequencing depth,
+which is often called ``library size'' has no relationship with the
+actual number of molecules in the sampled environment (1). In other
+words (\(\mathbf{Y}^{Tot}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) the observed
+data \(\mathbf{Y}\) provides us with information about the system
+composition \(\mathbf{W}^{Comp}\) but little to no information in the
+system scale \(\mathbf{W}^{Tot}\) (Lovell et al.~2011).
 
@@
 
-The ALDEx2 R package (36, 40) is a general purpose toolbox to model the
-uncertainty of HTS data and to use that model to estimate LFC
-significance. At a high-level, ALDEx2 has three connected components to
-estimate the uncertainty inherent in HTS datasets. First, the tool
-accounts for the uncertainty of the sequencing counts using Dirichlet
-multinomial sampling to build a probabilistic model of the data; i.e.,
-\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
-Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
-data (36). However, this step was modified recently to account for scale
-uncertainty and misspecification (16) via a scale model, explained with
-more details in (3, 16) and summarized in the next paragraph. Finally, a
-standard null-hypothesis test and a non-parametric estimate of mean
-standardized difference are used to report on the finite sample
-variation. These sources of uncertainty and variation are combined via
-reporting the expected values from a Monte-Carlo simulation framework.
-For simplicity, we use the term `difference' to refer to the absolute
-difference between groups, and `dispersion' to refer to the
-within-condition difference or pooled variance as defined in (36). These
-are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
-16, 36, 40).
-
-Scale models can be incorporated into ALDEx2, turning the ALDEx2 model
+The ALDEx2 R package (41, 46) is a general purpose toolbox to model the
+uncertainty of HTS data and to use that model to estimate the
+significance of the underlying LFC (log-fold change). At a high-level,
+ALDEx2 has three connected components to estimate the uncertainty
+inherent in HTS datasets. First, the tool accounts for the uncertainty
+of the sequencing counts using Dirichlet multinomial sampling to build a
+probabilistic model of the data; i.e.,
+\(\mathbf{\hat{W}}^{Comp} \approx \mathrm{Dir}(\mathbf{Y})\). Secondly,
+ALDEx2 uses the centred log-ratio transformation to scale the data (41).
+It was this step that was modified to account for scale uncertainty and
+misspecification (22) explained with more details in (3, 22) and
+summarized in the next paragraph. Finally, a standard null-hypothesis
+test and a non-parametric estimate of mean standardized difference are
+used to report on the finite sample variation. These sources of
+uncertainty and variation are combined via reporting the expected values
+from a Monte-Carlo simulation framework. For simplicity, we use the term
+`difference' to refer to the absolute difference between groups, and
+`dispersion' to refer to the within-condition difference or pooled
+variance as defined in (41). These are calculated on a \(\log_2\) scale.
+For more details on ALDEx2 see (3, 22, 41, 46).
+
+Scale models were incorporated into ALDEx2, turning the ALDEx2 model
 into a specialized type of statistical model which Nixon et al. (3) term
@@
 centred log-ratio normalization step. They did this by including a model
-for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR method used by ALDEx2
-makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where
-\(G_n\) is the geometric mean of sample n, which while being a random
-variable, is essentially constant across each Monte-Carlo replicate.
-With this modificaiton, ALDEx2 can be generalized by considering
-probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that
-have mean \(1/G_{n}\). For example, the following scale model
-generalizes the CLR:
-
-\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2})\].
-
-This formulation is quite flexible (3, 16). In the simple or `default'
+for \(\mathbf{\hat{W}}^{Tot}_{n}\). The CLR normalization used by ALDEx2
+makes the assumption \(\mathbf{\hat{W}}^{Tot}_{n}=1/G_{n}\), where
+\(G_n\) is the geometric mean of the counts (or the corresponding
+proportions)of each part in sample n, which while being a random
+variable, is essentially constant across each Monte-Carlo replicate, but
+that differs between samples. With this modification, ALDEx2 can be
+generalized by considering probability models for the scale
+\(\mathbf{\hat{W}}^{Tot}_{n}\) that have mean \(1/G_{n}\). For example,
+the following scale model generalizes the CLR:
+
+\[\log \mathbf{\hat{W}}^{Tot}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2}).\]
+
+This formulation is quite flexible (3, 22). In the simple or `default'
 configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
-from a log-Normal distribution that adds scale uncertainty. The default
-approach controls only the degree of uncertainty in the CLR assumption
-for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes
-case and \(x_{n}=0\) denotes control). In the advanced or `informed'
-configuration, \(\mu\) takes different values for each group, and this
-controls the location of the LFC assumption. An example of both the
-default and informed approaches is given for a microbiome dataset in
-(16), and here we show that these approaches also work well in
-transcriptome and metatranscriptome datasets. These modifications are
-instantiated in ALDEx2 which is the first software package designed for
-SSRV-based inference.
+from a log-Normal distribution (3). Adding scale uncertainty with the
+\(\gamma\) paramenter (as shown in Figure 1:1) controls only the degree
+of uncertainty of the CLR assumption for the \(x_{n}\) binary condition
+indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes
+control). In the advanced or `informed' configuration, \(\mu\) takes
+different values for each group and controls the location of the LFC
+assumption (as shown in Figure 1:2); combining \(\mu\) with a \(\gamma\)
+estimate allows for uncertainty in both the location and the scale. In
+the manuscript where the idea of scale was original derived, Nixon et
+al. (3) conducted extensive simulations showing that both the default
+and informed approaches exhibit increased sensitivity and specificity
+(22). In this report we show that these approaches also work well to
+control the FDR in transcriptome and metatranscriptome datasets. We also
+provide some additional insights into how this is achieved. These
+modifications are instantiated in ALDEx2 which is the first software
+package designed for SSRV-based inference.
 
@@
 
-Gierliński et al. (41) generated a highly replicated yeast transcriptome
+Gierliński et al. (47) generated a highly replicated yeast transcriptome
 dataset to compare gene expression between a wild-type strain and a snf2
-gene knockout, \(\Delta\)snf2. This dataset was used to test several
-RNA-seq tools for their power to detect the set of differentially
-abundant transcripts identified in the full dataset when the data was
-subset (10). In this study each tool had its own `gold standard' set of
-transcripts with different tools identifying between between 65\% to
+gene knockout, \(\Delta\)snf2. This dataset of 86 samples (44 and 42 per
+group) is an example of technical growth replicate experiments common in
+the literature. The dataset was used to test several RNA-seq tools for
+their power to detect the set of differentially abundant transcripts
+identified in the full dataset when the data was subset (9). In this
+original study each tool had its own `gold standard' set of transcripts
+with different tools identifying between between 65\% to
 \textgreater80\% of all transcripts as being significantly different.
@@
 authors suggested that it was more appropriate to apply a dual cutoff
-composed of both a Benjamini-Hochberg (42) corrected p-value (q-value)
+composed of both a Benjamini-Hochberg (48) corrected p-value (q-value)
 plus a difference cutoff to limit the number of identified transcripts
-to a much smaller fraction of the total. Nixon et al, (16) showed that
-adding even a small amount of scale uncertainty with ALDEx2 dramatically
-reduced the number of significant transcripts identified, removing the
-need for the dual cutoff approach in this dataset and others. Below we
-include an intuitive explanation of how incorporating scale uncertainty
-achieved this outcome using a setting of \(\gamma=0.5\), which is in
-line with the recommendations of (16) and that gives results comparable
-to those proposed in (10).
+to a much smaller fraction of the total. In other datasets, Nixon et al,
+(22) showed that adding even a small amount of scale uncertainty with
+ALDEx2 dramatically reduced the number of significant transcripts
+identified, removing the need for the dual cutoff approach in this
+dataset and others.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{go3_files/figure-latex/yst-plot-1}
+
+}
+
+\caption{Volcano and effect plots for unscaled and scaled transcriptome analysis.
+ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset.
+The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots.
+Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty.
+Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange).
+Points in grey are not statistically signficantly different under any condition.
+The horizontal dashed lines represent a \(\log_2\)difference of \(\pm 1.4\).}\label{fig:yst-plot}
+\end{figure}
 
 We start with the assumption that not all statistically significant
-differences are biologically relevant (43), and that such a large number
-of significantly different transcripts breaks the necessary assumption
-for DA/DE expression that most parts be invariant (27). As noted,
-transcriptomics commonly uses a dual cutoff approach that is graphically
-exemplified by volcano plots (10, 12). Using either DESeq2 or ALDEx2, a
-majority of transcripts are statistically significantly different
-between groups with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%,
-DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. These values are
-in line with those observed by (10). Such large numbers of statistically
-significant transcripts seems biologically unrealistic. That 118
-transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
-identifies 582 transcripts that ALDEx2 does not, suggests that the
-choice of normalization plays a role in which results are returned as
-significant and that some, if not the majority, are driven by technical
-differences in the analysis (9, 27).
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/plot1-1.pdf}}
-\caption{Volcano and effect plots for unscaled and scaled transcriptome
-analysis. ALDEx2 was used to conduct a differential expression (DE)
-analysis on the yeast transcriptome dataset. The results were plotted to
-show the relationship between difference and dispersion using effect
-plots or difference and the q-values using volcano plots. Panels A,C are
-for the naive analyses, and Panels B,D are for the default analyses that
-include scale uncertainty. Each point represents the values for one
-transcript, with the color indicating if that transcript was significant
-in the both analyses (red) or in the naive analysis only (orange).
-Points in grey are not statistically signficantly different under any
-condition. The horizontal dashed lines represent a \(\log_2\)difference
-of \(\pm 1.4\).}
-\end{figure}
-
-The Volcano plots in Figure 1 A and B show that adding scale uncertainty
+differences are biologically relevant (49), and that a result where the
+majority of transcripts are significant breaks the necessary assumption
+for DA/DE expression that most parts be invariant (31). Transcriptomic
+analysis commonly uses a dual cutoff approach graphically exemplified by
+volcano plots (9, 11). Using either DESeq2 or ALDEx2, a majority of
+transcripts are statistically significantly different between groups
+with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%, DESeq2) or 4172
+(71\%, ALDEx2) of the 5891 transcripts. These values are in line with
+those observed by (9). Such large numbers of statistically significant
+transcripts seems biologically unrealistic. That 118 transcripts are
+identified by ALDEx2 and not DESeq2, while DESeq2 identifies 582
+transcripts that ALDEx2 does not, suggests that the choice of
+normalization plays a role in which results are returned as significant
+and that some, if not the majority, are driven by technical differences
+in the analysis (8, 31) or are false positives.
+
+The Volcano plots in Figure 2 A and B show that adding scale uncertainty
 increases the minimum q-value and increases the concordance between the
 q-value and the difference between groups (compare panels A and B). The
-effect plots (44) in Figure 1C shows that the majority of significant
+effect plots (50) in Figure 2C shows that the majority of significant
 transcripts (red, orange) have negligible differences between groups and
@@
 the experimental design which is actually a technical wet lab
-replication rather than a true biological replication design (41). Scale
+replication rather than a true biological replication design (47). Scale
 uncertainty can be incorporated using the \texttt{gamma} parameter that
 controls the amount of noise added to the CLR mean assumption when we
-call either \texttt{aldex()}, or \texttt{aldex.clr()}. Figure 1 B,D
-shows that setting \(\gamma=0.5\) results in 205 which is far fewer
-significant transcripts than in the naive analysis and we observe that
-the minimum dispersion increases from 0.12 (\(\gamma = 0\) ) to 0.67
-(\(\gamma=0.5\)).
+call either \texttt{aldex()}, or \texttt{aldex.clr()}. Figure 2 B,D
+shows that setting \(\gamma=0.5\) now results in 205 which is far fewer
+statistically significant transcripts than in the naive analysis and we
+observe that the minimum dispersion increases from 0.12 (\(\gamma = 0\)
+) to 0.67 (\(\gamma=0.5\)).
+
+It is common practice to use a dual-cutoff by choosing transcripts based
+on a thresholds for both q-values and fold-changes (9), and these were
+first proposed for microarray experiments through volcano plots (11).
+Note that there is considerable variation in recommended cutoff values
+(9), and that this controversy has persisted ever since fold-change was
+suggested (51). Unfortunately, universal cutoff fold-change values
+cannot be identified in part because different tools have intrinsically
+different variance in their log\_2-fold change ranges (52). This has led
+to the widespread practice of applying a post-hoc fold-change cutoff to
+reduce the number of positive identifications to a manageable proportion
+of the whole dataset. Here, we applied the dual-cutoff method using a
+fold-change of at least a \(2^{1.4}\) fold change that reduces the
+number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
+This cutoff was chosen for convenience and is mid-way between the high
+and low fold-change recommendations of (9). These limits are shown by
+the dashed grey lines in Figure 2 and we can see that a \(2^{1.4}\) fold
+change \((\sim 2.6\) fold) cutoff identifies a similar number of
+transcripts as does ALDEx2 using \(\gamma = 0.5\) which identifies 205
+transcripts.
+
+Supplementary Figures 1 shows an example of the
+\texttt{aldex.senAnalysis()} function to identify those transcripts that
+are very sensitive to scale uncertainty in this dataset. Here we see
+that adding a very small amount of scale \(\gamma = 0.1\) reduces the
+number of significant transcripts by more than half in the yeast
+dataset. This allows the analyst to ignore those low-dispersion
+transcripts that were significant only because of an absence of scale
+uncertainty.
+
+We next examined how adding scale would alter the analysis in a real
+dataset to which synthetically generated true positive counts had been
+added. We show results from the anti-PD-1 therapy RNA-seq dataset (53)
+which examined changes in gene expression when cells were exposed or not
+to a cell-cycle checkpoint inhibitor. This dataset was used by Li et al.
+(8) as an example of the dangers of relying on tools with high false
+positive error rates when analyzing clinical or clinically-related
+transcriptome samples. Indeed, in the benchmarking analysis done by this
+group, they found that parameter based methods such as DESeq2 and edgeR
+that on reported p-value and fold-change cutoffs often led to
+conclusions in the original dataset that were indistinguishable from
+permutations of the dataset. In other words, that the analysis of
+transcriptome datasets from patient-derived samples often exhibited many
+false-positive identifications.
 
 \begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/disp-1.pdf}}
-\caption{Adding scale uncertainty changes the dispersion distribution.
-Panel A shows a plot of the expected value for relative abundance vs the
-expected value for the pooled dispersion as output by
-\texttt{aldex.effect}. The dashed horizontal lines show the median value
-for the features with a rAbundance between -0.5 and 0.5, and the light
-colored lines are lowess lines of fit through the center of mass of the
-data. Panel B plots the dispersion difference between \(\gamma = 1\) and
-\(\gamma = 0\); note the non-linear relationship that highlights the
-rotation that is evident in Panel A. The colored lines indicate the
-lowess line of fit through the centre of mass of the plot for the
-various populations of points. The grey line is the total population and
-shows the difference \(\Delta\), the red line is the population of
-significant transcripts (*) with \(\gamma = 0\), the orange line is the
-population of significant transcripts with a difference threshold (T) of
-about \(\pm 2^{1.4}\)-fold change, the blue line is the population of
-significant transcripts with \(\gamma = 0.5\), and the cyan line is the
-significant population with \(\gamma = 1\). \(\Delta\): Difference, *:
-significant, T: thresholded.}
+
+{\centering \includegraphics[width=0.8\linewidth]{go3_files/figure-latex/immuno-1}
+
+}
+
+\caption{Results of modeling true positive (TP) differences in the PD-1 dataset.
+For this the data were shuffled and  5\% of the transcripts were modeled to have TP differences between groups where the differences were drawn from a Normal distribution with a mean difference of 0 and a sd of 2.
+Panel A shows the mean FDR of 10 instances for scale-naive ALDEx2 (A 0), and ALDEx2 with \(\gamma=0.2\) (A 0.2) or \(\gamma=0.2\) (A 0.5), and where significant features were identified by  DESeq2 without (D 0), or with (D 0.5) a 0.5-fold change difference.
+The x-axis shows how the FDR changes for each tool as a function of the estimated modeled difference between groups.
+Panel B shows the mean sensitivity (TP found /all TP).
+Panels C and D show volcano and effect plots for the ALDEx2 output with the transcripts identified as significant at each scale setting as colored points, and non-significant transcripts in black.}\label{fig:immuno}
 \end{figure}
 
-It is common practice to use a dual-cutoff by choosing transcripts based
-on a thresholds for both q-values and fold-changes (10, 12). Note that
-there is considerable variation in recommended cutoff values(10). Here,
-applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold
-change reduces the number of significant outputs to 193 for DESeq2 and
-to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line
-with the recommendations of (10). the fold change limits are shown by
-the dashed grey lines in Figure 1. The \(2^{1.4}\) fold change cutoff
-identifies a similar number of transcripts as does ALDEx2 using
-\(\gamma = 0.5\) which identifies 205. Supplementary Figure 2 shows how
-to use the \texttt{aldex.senAnalysis()} function to identify those
-transcripts that are very sensitive to scale uncertainty. In this
-supplementary figure we see that even adding a very small amount of
-scale \(\gamma = 0.1\) reduces the number of significant transcripts by
-more than half. This allows us to those low-dispertion transcripts that
-were significant only because of an absence of scale uncertainty. In
-practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
-is realistic for most experimental designs (16).
-
-The effect on dispersion with increasing amounts of scale uncertainty
-are shown in Figure 2A, where we can see that the dispersion increases
-as uncertainty is added. Note that the dispersion in the unscaled
-analysis in Figure 2A reaches a minimum near the mid-point of the
-distribution, and also does so when the analysis is conducted with
-DESeq2 (Supplementary Figure 3). This shows more clearly that dispersion
-of many transcripts is almost negligible in the absence of scale
-uncertainty. This plot makes the counter-intuitive suggestion that the
-variance in expression of the majority of genes with moderate expression
-is more predictable than highly-expressed genes or of housekeeping genes
-(45). This is at odds with the known biology of cells where single cell
-counting of highly-expressed transcripts shows that they have little
-intrinsic variation (20, 46).
-
-Adding scale uncertainty by setting \(\gamma=0.5\) , or
-\(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 2A
-by the red and blue data points, and by the colored lines of fit through
-the centre of mass of the data. Less obvious is that the additional
-dispersion is not applied equally to all points. Figure 2B shows a plot
-of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data and
-here we can see that scale uncertainty is preferentially increasing the
-dispersion of the mid-expressed transcripts that formerly had negligible
-dispersion; examine the grey line of best fit (overlaid by the red line)
-for the trend. Panel B also shows the trend of the expression-dispersion
-relationship for transcripts that are classed as statistically
-significant. The red line shows the trendline with no added scale
-uncertainty, and this trendline exactly overlays with the grey trendline
-for the bulk of transcripts. The orange trendline indicates those
-transcripts that are both statistically significant and that have a
-thresholded expression level of \(\pm 1.4\), and the dark blue and cyan
-lines show the statistically significant trendline for \(\gamma=0.5\) ,
-or \(1.0\).
-
-Thus, taking Figures 1 and 2 together adding scale uncertainty has the
-desirable effect of changing the distribution of transcripts identified
-as significantly different between groups. Those parts that were
-statistically significantly different \emph{only because of low
-dispersion} are preferentially excluded from statistical significance
-while those parts that were significantly different because of a high
-difference between groups remain.
+Figure 3 shows the results of ten permutations of this dataset while
+simulating 5\% of the transcripts to be true positives where the
+difference from no change was derived from a Normal distribution (54).
+This simulation was conducted with the \texttt{seqgendiff} R package
+(54) to both permute the dataset and to add true positive features. We
+did ten permutations and kept track of the number of true and false
+positive identifications. Figure 3A shows the actual false discovery
+rate for ALDEx2 with and without the addition of scale uncertainty and
+for DESeq2 with or without a fold-change cutoff of \(2^{0.5}\). When the
+modeled difference was greater than 0, ALDEx2 exhibited a FDR of
+\(\gamma = 0\): 0.002, \(\gamma = 0.2\): 0, and \(\gamma = 0.5\): 0,
+while DESeq2 with no fold-change cutoff had an FDR of 0.32 and with a
+0.5 fold-change cutoff an FDR of 0.31. Figure 3 also plots these results
+as a function of the minimum modelled fold change of the true positive
+transcripts. First, we can see that this analysis recapitulates the
+observations of Li et al (8) in that DESeq2 has very poor false positive
+control at a nominal FDR of 0.05. The FDR is not controlled any better
+when a fold-change cutoff is applied, and this agrees with previous work
+showing that fold-change cutoffs do not materially improve FDR control
+in high throughput datasets (12, 13). Figure 3B shows that DESeq2 has
+higher sensitivity than does ALDEx2, and not surprisingly this
+sensitivity increases as the difference between groups increases. For
+the case where the mean difference is 0 or greater, ALDEx2 exhibited a
+sensitivity of \(\gamma = 0\): 0.66, \(\gamma = 0.2\): 0.54, and
+\(\gamma = 0.5\): 0.35, while DESeq2 with no fold-change cutoff had a
+sensitivity of 0.76 and with a 0.5 fold-change cutoff a sensitivity of
+0.73. In this example, scale-naive ALDEx2 has near perfect FDR control
+and reasonable sensitivity at low modeled difference between groups.
+However, when the modeled difference between groups becomes large, then
+the scale-naive version of ALDEx2 begins to exhibit unacceptable rates
+of false positives and the false positive rate for DESeq2 also
+increases. Adding in even small amounts of scale uncertainty drops the
+true FDR rate to 0, but at the expense of sensitivity. The major
+contributor to the increase in FDR with larger modeled differences is
+that the tools are identifying as positives those transcripts that are
+modeled to have differences just below the threshold. We need to
+recognize that there is no such thing as a statistical free lunch; the
+analyst can have high sensitivity but low confidence that any individual
+transcript is truly different, or have lower sensitivity but have very
+high confidence that the difference is real. In other words, the
+sensitivity of a method is directly tied to how much error the
+investigator is willing to tolerate.
+
+Examination of the volcano plot (11) and effect plot (50) in Figure 3C
+and D provides some insight into why adding scale uncertainty provides
+better FDR control than does the approach of using a p-value and a
+fold-change cutoff. In the volcano plot, adding scale uncertainty
+differentially excludes transcripts with a combination of marginal
+p-values and low difference between groups, and that this becomes more
+pronounced with a larger scale value. This effect is also seen in Figure
+2 but is more nuanced. In contrast, the fold-change cutoff does not
+incorporate the magnitude of the p-value and so transcripts with large
+differences, but marginal p-values are retained. The effect plot in
+Figure 3D shows that the transcripts with marginal p values and large
+differences that are excluded when scale uncertainty is added are those
+that have a large dispersion.
+
+As a concrete example consider the point that is circled in panels C and
+D with a marginal p-value, with a difference between of nearly 4 and a
+dispersion of greater than 6. This transcript is no longer significant
+when \(\gamma = 0.2\), but would require a very large fold-change cutoff
+to be excluded by the standard approach. In addition, transcripts with
+very small dispersion and very small differences are also excluded when
+scale uncertainty is added. Thus, the addition of scale uncertainty
+achieves the desired outcome of lowering the FDR for those transcripts
+that are either marginally differentially abundant, or where the
+underlying dispersion--and hence the uncertainty in measurement--is very
+high, or for transcripts that fit both criteria.
+
+Supplementary Figure 2 shows a second permuted dataset with the addition
+of modeled differences between groups. Here we used another real dataset
+with over 200 biological replicates of BRCA1 tumor and control tissue
+samples from Li et al. (8). This Supplementary Figure shows that the FDR
+control of DESeq2 is somewhat better than in the PD-1 dataset, although
+still much greater than the the anticipated 5\%. Further, a fold-change
+cutoff reduces the FDR of DESeq2 from about 30\% to just over 20\% with
+a power of over 80\%. However, we can see that scale-naive ALDEx2
+performs substantially better with a negligible FDR and comparable
+power. Adding scale uncertainty again improves FDR even for those
+transcripts modeled to have larger differences and the power is
+substantially better than in the PD-1 dataset, reaching the same power
+as DESeq2 or scale-naive ALDEx2 when the modeled difference is large. As
+before, this is driven by removing from consideration those transcripts
+with either a small difference between or a marginal p-value, or both.
+
+Supplementary Figures 3 shows that effect of applying \(\gamma = 0.5\)
+to this dataset results in reducing the number of positive transcripts
+from being \(\sim 70\)\% of the whole dataset to less than 10\% of the
+dataset and that this is largely because of a reduction in significance
+of those transcripts with low dispersion. Supplementary Figures 4 shows
+a sensitivity analysis of the BRCA1 dataset showing that different scale
+uncertainty amounts alter the number of significant transcripts in a
+biological replicate experiment similarly to a technical replicate
+experiment. Supplementary Figures 5 and 6 delve into how adding scale
+uncertainty affects the variace-abundance relationship in subtle ways,
+and may help readers to understand the observations seen in Figure 3 and
+supplementary Figure 2.
+
+Together the results in this section show that adding scale uncertainty
+has the desirable effect of altering the transcripts identified as
+significantly different between groups in a way that exhibits better
+control of FDR albeit with a corresponding reduction in sensitivity.
+Those parts that were statistically significantly different \emph{only
+because of low dispersion} or that \emph{had marginal p-values} or both,
+are now preferentially excluded from statistical significance. In
+practice, we suggest that a \texttt{gamma} parameter of between 0.2 and
+0.5 is realistic for most experimental designs (22) regardless if the
+replication is technical or biological.
 
@@
 
-Dos Santos et al. (47) used a vaginal metatranscriptome dataset to
+Dos Santos et al. (55) used a vaginal metatranscriptome dataset to
 compare the gene expression in bacteria collected from healthy (H) and
-bacterial vaginosis (BV) affected women. In this environment, both the
-relative abundance of species between groups and the gene expression
-level within a species is different (48). Additionally, prior research
-suggests that the total number of bacteria is about 10 times more in the
-BV than in the H condition (23). Thus, these are extremely challenging
-datasets in which to determine differential abundance as there are both
-compositional and scale changes between conditions. The usual method to
-analyze vaginal metratranscriptome data is to do so on an
-organism-by-organism basis (48--50) because the scale confounding of the
-environment is less pronounced. One attempt at system-wide analysis
-returned several housekeeping functions as differentially expressed
-between groups (49); a result likely due to a disconnect between the
-assumptions of the normalization used and the actual scale of the
-environment (15).
-
-In this example, we show how to specify and interpret an informed scale
-model that can explicitly account for some of these modeling
-difficulties (16) even in a difficult dataset. An informed scale model
-can control for both the mean difference of scale between groups (e.g.,
-directly incorporate information on the differences in total number of
-bacteria between the BV and H conditions) as well as the uncertainty of
-that difference. To specify a user-defined scale model, we can pass a
-matrix of scale values instead of an estimate of just the scale
-uncertainty to \texttt{aldex.clr()}. This matrix should have the same
-number of rows as the of Monte-Carlo Dirichlet samples, and the same
-number of columns as the number of samples. While this matrix can be
-computed from scratch by the analyst, there is an
-\texttt{aldex.makeScaleModel()} function that can be used to simplify
-this step in most cases. This encodes the scale model as
-\(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents
-the scale value for each sample or group and gamma is the uncertainty as
-before. The scale estimate can be a measured value (cell count, nucleic
-acid input, etc) or an estimate. Nixon et al. (16) showed that only the
-ratio of the means are important when providing values for \(\mu_n\);
-i.e., the ratio between the \(\log_2 \mu_i\) and \(\log_2 \mu_j\)
-values. See the supplement to Nixon et al. (16) for more information.
-
-Figure 3A shows an effect plot of the data where reads are grouped by
-homologous sequence regardless of the organism of origin. Each point
-represents one of 3728 KEGG functions (51). There are many more
-functions represented in the BV group (bottom) than in the healthy group
-(top). This is because the \textit{Lactobacilli} that dominate a healthy
-vaginal microbiome have reduced genome content relative to the anaerobic
-organisms that dominate in BV, because there is a greater diversity of
-organisms in BV than in H samples, and because the BV condition has at
-least an order of magnitude more bacteria than does the H condition.
-
-There are 101 functions with low dispersion that appear to be shared by
-both groups (boxed area in Figure 3A, and colored in cyan). Inspection
-shows that these largely correspond to core metabolic functions such as
-transcription, translation, ribosomal proteins, glycolysis, replication,
-chaperones, etc (Supplementary file housekeeping.txt). The transcripts
-of many of these are commonly used as invariant reference sequences in
-wet lab experiments (45) and so are not be expected to contribute to
-differences in ecosystem behaviour. The average location of these should
-be centred on 0 difference to represent an internal reference set.
-However, without an informed scale model, the mean of these housekeeping
-functions is approximately located at +2.3.
+bacterial vaginosis (BV) affected women. This dataset is derived from
+two publicly available datasets composed of a set of 20 non-pregnant
+women from London, Ontario Canada (56) and a subset of 22 non-pregnant
+women collected from German women who underwent metronidazole treatment
+for BV (57). Batch effects for these two groups were removed with
+ComBat-seq (58) and the two datasets were merged into one, giving a
+total of 16 H and 26 BV samples. In the Dos Santos paper, all results
+from this initial analysis were replicated in a much larger dataset
+derived from the MOMS-PI study (59)
+
+In this vaginal environment, both the relative abundance of species
+between groups and the gene expression level within a species is
+different (60). Additionally, prior research suggests that the total
+number of bacteria is about 10 times more in the BV than in the H
+condition (28). Thus, these are extremely challenging datasets in which
+to determine differential abundance as there are both compositional and
+scale changes between conditions. The usual method to analyze vaginal
+metratranscriptome data is to do so on an organism-by-organism basis
+(57, 59, 60) because the scale confounding of the environment is less
+pronounced. One attempt at system-wide analysis returned several
+housekeeping functions as differentially expressed between groups (57);
+a result likely due to a disconnect between the assumptions of the
+normalization used and the actual scale of the environment (15).
 
@@
 
+In this example, we show how to specify and interpret a user defined or
+\emph{informed} scale model that can explicitly account for some of
+these modeling difficulties (22) even in a difficult to analyze dataset.
+An informed scale model can control for both the mean difference of
+scale between groups (e.g., directly incorporate information on the
+differences in total number of bacteria between the BV and H conditions)
+as well as the uncertainty of that difference as illustrated in Figure
+1:3. To specify a user-defined scale model, we can pass a matrix of
+scale values instead of an estimate of just the scale uncertainty to
+\texttt{aldex.clr()}. This matrix should have the same number of rows as
+the of Monte-Carlo Dirichlet samples, and the same number of columns as
+the number of samples. While this matrix can be computed from scratch by
+the analyst, there is an \texttt{aldex.makeScaleModel()} function that
+can be used to simplify this step in most cases. This encodes the scale
+model as \(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\)
+represents the scale value for each sample or group and \(gamma\) is the
+uncertainty as before. The scale estimate can be a measured value (cell
+count, nucleic acid input, spike-ins, etc) or an estimate. Nixon et al.
+(3, 22) showed that only the ratio of the means are important when
+providing values for \(\mu_n\); i.e., the ratio between the
+\(\log_2 \mu_i\) and \(\log_2 \mu_j\) values. See the supplement to
+Nixon et al. (22) for more information.
+
+Figure 4A shows an effect plot of the data where reads are grouped by
+homologous function regardless of the organism of origin. Each point
+represents one of 3728 KEGG functions (61). There are many more
+functions represented in the BV group (bottom) than in the healthy group
+(top). This is because the \textit{Lactobacilli} that dominate a healthy
+vaginal microbiome have reduced genome content relative to the anaerobic
+organisms that dominate in BV, because there is a greater diversity of
+organisms in BV than in H samples, and because the BV condition has
+about an order of magnitude more bacteria than does the H condition.
+
+The naive scale model appears to be reflecting the bacterial load as
+observed by calculating the mean scale value for each group. Using a
+negligible scale value; i.e., \(\gamma=1e-3\) exposes the naive scale
+estimate for samples in the \texttt{@scaleSamps} slot from the
+\texttt{aldex.clr} output. The naive scale estimate for the healthy
+group is 17.41 and for the BV group is 14.59 for a difference of 2.82.
+This is interpreted as the scale of the H group of samples being 7.06
+greater than the BV group.
+
 Applying the default scale model of \(\gamma=0.5\) increases the
@@
 based on the CLR normalization so no shift in location would be expected
-over the original ALDEx2 model. Nevertheless, about 30\% of the
+over the scale-naive ALDEx2 model. Nevertheless, about 30\% of the
 housekeeping functions are no longer statistically significantly
@@
 
-We desire a scale model that approximately centres the housekeeping
-functions. Our starting point for this is to identify the naive scale
-estimate from the data, and this can be done by calculating the mean
-scale value for each group with a negligible scale value; i.e.,
-\(\gamma=1e-3\). The scale estimate for samples can be accessed in the
-\texttt{@scaleSamps} slot from the \texttt{aldex.clr} output. In this
-datset the naive scale estimate for the healthy group is 17.41 and for
-the BV group is 14.59 for a difference of 2.82. This is interpreted as
-the scale of the H group of samples being 7.06 greater than the BV
-group. This precise but incorrect estimate places the location of the
-housekeeping genes off the midline of no difference.
-
-We anticipate that housekeeping functions should be nearly invariant and
-so an appropriate scale in this dataset is likely closer to 0 than the
-naive estimate. One way to choose an appropriate value for \(\mu_n\) is
-to us the \texttt{aldex.clr} function on only the presumed invariant
-functions setting \(\gamma > 0\), and then accessing the
-\texttt{@scaleSamps} slot as before. Doing so suggests that the
-difference in scale should be about 14\%. A second approach would be to
-identify the functions used as the denominator with the
-\texttt{denom="lvha"} option (15) for the \texttt{aldex.clr} function,
-and then to use these values as before. This approach suggests a 5\%
-difference in scale, and is potentially less subject to user
-interpretation.
+There are 101 functions with low dispersion that appear to be shared by
+both groups (boxed area in Figure 3A, and colored in cyan). Inspection
+shows that these largely correspond to core metabolic functions such as
+transcription, translation, ribosomal functions, glycolysis,
+replication, chaperones, etc (Supplementary file housekeeping.txt). The
+transcripts of many of these are commonly used as invariant reference
+sequences in wet lab experiments (62) and so are not be expected to
+contribute to differences in ecosystem behaviour. Because we expect
+housekeeping functions to be nearly invariant in their expression and to
+occur in all organisms, the average location of these should be centred
+on 0 difference to represent an internal reference set. However, with
+the naive scale model, the mean of these housekeeping functions is
+approximately located at 2.3. Thus, we desire a scale model that
+approximately centres the housekeeping functions and an appropriate
+scale in this dataset for functional analysis will place these functions
+closer to 0 than does the naive estimate. One way to choose an
+appropriate value for \(\mu_n\) is to use the \texttt{aldex.clr}
+function on only the presumed invariant functions setting
+\(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot as
+before. Doing so suggests that the difference in scale should be about
+14\%. A second approach would be to identify the functions used as the
+denominator with the \texttt{denom="lvha"} option (15) for the
+\texttt{aldex.clr} function, and then to use these values as before.
+This approach suggests a 5\% difference in scale, and is potentially
+less subject to user interpretation.
 
@@
 difference between groups and uncertainty level. Applying a per-group
-relative differential scale of 0.14 moves the housekeeping functions to
-the midline of no difference (Figure 3C, 14\% mean difference = -0.24,
-5\% difference = -0.34), and applying a gamma of 0.5 provides the same
-dispersion as in panel B of Figure 3. Note that now a significant number
-of functions are differentially up in BV that were formerly classed as
-not different without the full scale model (orange), or when only a
-default scale was applied. Inspection of the functions shows that these
-are largely missing from the \emph{Lactobacillus} species and so should
-actually be captured as differentially abundant in the BV group. The
-supplement shows that the using either the 5\% or the 14\% scale
+relative differential scale of 0.14 moves the housekeeping functions
+close to the midline of no difference (Figure 3C, assuming 14\% mean
+difference = -0.24, assuming a 5\% mean difference = -0.34), and
+applying a gamma of 0.5 provides the same dispersion as in panel B of
+Figure 3. Note that now a significant number of functions are
+differentially up in BV that were formerly classed as not different
+without the full scale model (orange), or when only a default scale was
+applied. Inspection of the functions shows that these are largely
+missing from the \emph{Lactobacillus} species and so should actually be
+captured as differentially abundant in the BV group. Supplementary
+Figure 7 shows that the using either the 5\% or the 14\% scale
 difference give imperceptibly different results suggesting that an
 informed scale model does not have to precisely estimate the scale
-difference to be useful. Nixon et al, (16) found that multiple
+difference to be useful. Nixon et al, (22) also found that multiple
 reasonable estimates for the \(\mu_n\) part of the informed scale model
@@
 false negatives (orange functions) even in a very difficult to analyze
-dataset. The remarkable improvements in biological interpretation
-afforded by an informed scale model, and the transferrability of it
-between sample cohorts of the same condition is outlined elsewhere (47).
-We suggest that the default scale model is sufficient when the data are
-approximately centred. However, an informed model is more appropriate
-with datasets are not well centred or when the investigator has prior
-information about the underlying biology.
+dataset. We used this scale model to uncover hiter-to-now unknown
+differences in microbiome functional activity between the Healthy and BV
+cohorts that were missed in previos analyses and that explain important
+clinical differences between them (55). The remarkable improvements in
+biological interpretation afforded by an informed scale model, and the
+transferrability of it between sample cohorts of the same condition is
+outlined in dos Santos et al. (55). We suggest that the default scale
+model is sufficient when the data are approximately centred but that an
+informed model is more appropriate with datasets are not well centred or
+when the investigator has prior information about the underlying
+biology.
 
@@
 
-Biological systems are both predictably variable and stochastic (46) and
+Scale estimates affect two parts of the analysis. Modeling uncertainty
+in scale prevents false certainty in the precision of estimation and
+controls false positive identification. Modeling between-group scale
+relaxes the assumption of identity between the sizes of the environments
+and allows better control of false negative identification. The scale
+estimates can be derived from the total number of molecules in the
+environment or from other estimates of input size (cell counts, initial
+concentrations, spike-ins, growth rates, etc).
+
+Biological systems are both predictably variable and stochastic (63) and
 systems biology experiments show that there are transcripts with
 approximately constant concentrations in the cell and those with large
-variability under different growth conditions (20). Current measurement
+variability under different growth conditions (25). Current measurement
 methods that rely on high throughput sequencing fail to capture all of
-the variation, particularly variation due to scale (3, 16). In the
-absence of external information (4, 19, 52) sequencing depth
-normalisation methods cannot recapture the scale information (19), and
-can only normalize for the technical variation due to sequencing depth.
-Here we demonstrated that even approximate estimates of scale and the
-uncertainty of measuring it can aid in the interpretation of
-RNA-sequencing experiments.
-
-Nixon et al. (3) introduced the idea of explicitly modelling the scale
-of a HTS dataset, and showed how to incorporate these models in the
-analysis of microbiome and other datasets (16). They demonstrated that
+the variation, particularly variation due to scale (3, 22). In the
+absence of external information (19, 20, 64) sequencing depth
+normalisation methods cannot recapture the scale information (19, 23),
+and can only normalize for the technical variation due to sequencing
+depth. Here we demonstrated that even approximate estimates of the true
+system scale and the uncertainty of measuring it can aid in the
+interpretation of RNA-sequencing experiments.
+
+Nixon et al. (3) introduced the idea of explicitly modeling the scale of
+a HTS dataset, and showed how to incorporate these models in the
+analysis of microbiome and other datasets (22). They demonstrated that
 many tools commonly used to analyze HTS datasets had substantial Type 1
-and Type 2 error rates, in line with recent findings by others (9). A
-version of ALDEx2 with the ability to include scale uncertainty was
-shown to be able to correct for the high Type 1 error rate for that
-tool, albeit with some loss of sensitivity. Finally, they showed that
+and Type 2 error rates in line with recent findings by others (5, 7, 8).
+A version of ALDEx2 with the ability to include scale uncertainty was
+shown to be able to correct for high Type 1 error rate for that tool,
+albeit with some loss of sensitivity. Finally, they showed that
 incorporating an informed scale model incorporating both location and
 scale uncertainty estimates could both control for Type 1 and Type 2
-error rates (16).
-
-Building and using a scale model, thus has substantial benefits relative
+error rates (22, 43).
+
+The process of choosing the parameters are experiment-specific and can
+be anchored in known information such as cell counts, spike-ins,
+information from the literature or similar (3, 22, 65). In the
+metatranscriptome example used in this report {[}(43);{]} the choice of
+parameter was driven by the assumption inherent in the biology that core
+housekeeping functions would serve as an appropriate standard (37). The
+choice of parameters must be guided by the experimental question and
+other approaches in the literature which suggest normalizing the
+transcript levels to the bacterial metagenomic levels (66) could be used
+to set the scale parameters, but in this case are more granular at the
+individual organism level rather than at the systemic functional level.
+
+Building and using a scale model thus has substantial benefits relative
 to the dual cutoff approach that is advocated for many gene expression
-experiments (10, 12). In particular, the dual cutoff approach has long
-been known to not control for Type 1 errors (13, 14), and the frequent
-lack of concordance between tools when benchmarked on transcriptomes(7,
-9, 10, 26, 53) and microbiomes (6, 8, 28, 37, 54, 55) suggests poor
-control of Type 2 errors as well (7, 9). Thus, incorporating a scale
+experiments (9, 11). In particular, the dual cutoff approach has long
+been known to not control for Type 1 errors (12, 13), and the frequent
+lack of concordance between tools when benchmarked on transcriptomes (5,
+7--9, 17, 67) and microbiomes (4, 6, 32, 42, 68, 69) suggests poor
+control of Type 2 errors as well (5, 8). Thus, incorporating a scale
 model during the analysis of HTS data promises the best of both worlds.
@@
 additional computational overhead. Furthermore, this work and previous
-(16) show that even minimal information about the underlying environment
-can be used to build a relatively robust scale model that controls for
-both Type 1 and 2 error rates.
+(22, 43) show that even minimal information about the underlying
+environment can be used to build a relatively robust informed scale
+model that controls for both Type 1 and 2 error rates. It is important
+to note that the approach advocated here is distinct from that suggested
+by Zhang et al. (66, 70) where the DNA amount for a gene is a covariate
+in the model for transcriptomic differential abundance. In our analysis
+we grouped all the transcript information to functional level regardless
+of organism, instead of modeling per-organism gene abundances. In the
+future we anticipate being able to build more complex models similar to
+those used by Zhang et al.~with the additional information of
+uncertainty in the underlying gene count.
 
 In the analysis of HTS data it is often observed that larger datasets
-converge on the majority of parts being significantly different (3, 9,
-10). Li et al. (9) conducted a permutation-based benchmarking study and
+converge on the majority of parts being significantly different (3, 8,
+9). Li et al. (8) conducted a permutation-based benchmarking study and
 found that widely used tools performed worse than simple Wilcoxon
-rank-sum tests in controlling the FDR when sample sizes became large.
-They suggested that the presence of outliers were one of the factors
-driving this observation. Brooks et al. (56) suggested that
-inappropriate choice of benchmarking methods are also a major
-contributing factor and that objective standards of truth are important.
+rank-sum tests coupled with the TPM normalization in controlling the FDR
+when sample sizes became large. Li et al.~suggested that the presence of
+outliers were one of the factors driving the extreme FDR in some tests.
+We found that when the Wilcoxon test was used within the ALDEx2
+framework that it had essentially the same outputs as did the t-test.
+For example, in the PD-1 dataset where \(\gamma = 0\) the ALDEx2 t-test
+exhibited a mean FDR of 0.2\% and mean sensitivity of 65.9\% while the
+corresponding values from the ALDEx2 Wilcoxon test were 0.3\% and
+68.9\%. This result again supports that the assumptions of the
+normalizations are as important or more important than the statistical
+test. Brooks et al. (71) suggested that inappropriate choice of
+benchmarking methods are also a major contributing factor and that
+better objective standards of truth are needed. In this report we
+generated semi-synthetic test data used binomial thinning which produces
+data that more closely mimic the properties of real high throughput
+sequencing data, and so can more rigorously test different tools (54).
 From the perspective of our work the disagreement between tools can be
 explained by the observation that different analytic approaches produce
-different parameter estimates. Thus, more data produces worse estimates
+different parameter estimates for either location or scale, or for both,
+as suggested in Figure 1. Thus, more data produces worse estimates
 because the additional data simply increases the precision of a flawed
-estimate (3, 57).
-
-Scale simulation is now build into ALDEx2 (16) and this work suggest
-that there are two main root causes to common HTS data pathologies. The
-first contributing factor is the observed very low dispersion estimate
-for many features that is a by-product of experimental design,
-sequencing and normalization. In the Schurch et al. (10) dataset, the
-data were from single colonies derived from a single culture. Thus, it
-is more accurate to describe the 96 samples as wet-lab technical
-replicates rather than independent samples. This type of replication
-approach is standard in the molecular literature, and would be expected
-to result in the very low dispersion that is observed. Applying the
-default scale model with \(\gamma=0.5\) a large number of transcripts
-have their dispersion increased (Figure 1D), with the effect being
-largest for those with the lowest initial dispersion (Figure 2). Adding
-scale uncertainty results in modest number of transcripts, 205, being
-called significantly different as shown in the volcano plot in Figure 1B
-(red points). In addition, there is now a strong concordance between the
-difference and p-values. In hindsight, what is not obvious is why the
-unscaled volcano plot shows such poor correspondence. We suggest that
-this can be explained by random fluctuations in the many variance
-estimates with very low values, and this is supported by the plots shown
-in Figure 2.
-
-The second contributing factor is unacknowledged asymmetry in many
-datasets (15); i.e., different gene content or a directional change in
-the majority of features. In the case of asymmetry, the use of a
-user-specified scale model can be very useful for otherwise
-difficult-to-analyze datasets such as meta-transcriptomes and in-vitro
-selection datasets where the majority of features can change. We showed
-one such example for the metatranscriptome dataset in Figure 3. Here the
-dataset was highly asymmetrical. Incorporating differential scale on a
-per-group basis moves the mass of the housekeeping functions towards the
-midline of no difference and so affects both Type I and Type II error
-rates. We showed two ways of estimating the scale difference between
-groups and found that any reasonable estimate is an improvement over the
-default scale model. This is in line with the observations by Nixon et
-al (16) in a 16S rRNA gene sequencing dataset. It is also of note that
-in the case of true biological replicates (different individuals) that
-adding a modest amount of scale \(\gamma=0.5\) had little effect on the
-the difference between groups and on the dispersion. Thus, in this
-dataset the scale mis-specification was affecting mainly the location of
-the difference between groups. While we acknowledge that some prior
-information on which housekeeping transcripts should not be classed as
-differentially abundant is needed, we suggest that this information is
-widely available and is already used when performing the gold-standard
-quantitative PCR test of differential abundance (58, 59).
+estimate (3, 72).
+
+Scale simulation is now built into ALDEx2 (22) and in this report we
+suggest that there are two main root causes to common HTS data
+pathologies. The first contributing factor is the observed very low
+dispersion estimate for many features that is a by-product of some
+experimental designs and normalizations (Figure 2). Supplying additional
+uncertainty alleviates many FP but in a way that more appropriately
+controls the FDR as shown in Figure 3. The second contributing factor is
+unacknowledged asymmetry in many datasets (15); i.e., different gene
+content or a directional change in the majority of features. In the case
+of asymmetry, the use of a user-specified scale model can be very useful
+for otherwise difficult-to-analyze datasets such as meta-transcriptomes
+and in-vitro selection datasets where the majority of features can
+change as shown in Figure 4. We showed two ways of estimating the scale
+difference between groups and found that any reasonable estimate is an
+improvement over the naive approach and also over the default scale
+model. This is in line with the observations by Nixon et al (22) in a
+16S rRNA gene sequencing dataset. While we acknowledge that some prior
+information is needed that this information is widely available and is
+already used when performing the gold-standard quantitative PCR test of
+differential abundance (73, 74).
 
 Beyond concerns of fidelity and rigor, scale models also enhance the
-reproducibility and transparency of HTS analyses. The addition of scale
-uncertainty essentially tests the model over a range of normalizations
-(3) and so can replace the consensus approach that has been proposed by
-some groups (8, 60) with no additional computational overhead. Thus, an
-advantage of incorporating scale is that analyses can be made much more
-robust such that actual or potential differences in scale can be tested
-and accounted for explicitly. While it is beyond the scope of the
-present article, we note that there are many ways of building scale
-models that enhance the interpretability of the parameters and
-assumptions and a detailed description of these points is describe
-elsewhere (3).
+reproducibility and transparency of HTS analyses. The development of HTS
+and the associated problems of very high dimensional data that was not
+always statistically well-behaved led to many different proposed
+solutions including multiple normalizations and moderated test
+statistics. That these perform poorly in real data is shown by the
+simulation data in this report and elsewhere where both DESeq2 (which we
+used) and edgeR (which uses moderated statistical tests) performed
+poorly in controlling the FDR (8). While these approaches often work in
+many datasets they fail to address the underlying problems of
+information that is missing in the data which is what is supplied by
+adding uncertainty around the information we have about that data. The
+addition of scale uncertainty directly addresses the missing information
+by testing the model over a range of normalizations (3). In doing so,
+the scale-based approach removes the need for moderated statistics and
+can replace the consensus approach that has been proposed by some groups
+(6, 75) with no additional computational overhead. Thus, an advantage of
+incorporating scale is that analyses can be made much more robust such
+that actual or potential differences in scale can be tested and
+accounted for explicitly. While it is beyond the scope of the present
+article, we note that there are many ways of building scale models that
+enhance the interpretability of the parameters and assumptions and a
+detailed description of these points is describe elsewhere (3).
 
@@
 transcriptomes or indeed any type of HTS dataset. While the underlying
-scale of the system is generally inaccessible, the effect of scale on
-the analysis outcomes can be modelled and can help explain some of the
-underlying biology, and help to expose known issues with the analysis of
-HTS data. Adding scale information to the analysis allows for more
-robust inference because the features that are sensitive to scale can be
-identified and their impact on conclusions weighted accordingly.
-Additionally, the use of informed scale models permits difficult to
-analyze datasets to be examined in a robust and principled manner even
-when the majority of features are asymmetrically distributed or
-expressed (or both) in the groups (47). Thus, using and reporting scale
+scale of the system is generally inaccessible, the effect of scale
+uncertainty on the analysis outcomes can be modelled and can help
+explain some of the underlying biology. Adding scale information to the
+analysis allows for more robust inference because the features that are
+sensitive to scale can be identified and their impact on conclusions
+weighted accordingly. The use of informed scale models permits difficult
+to analyze datasets to be examined in a robust and principled manner
+even when the majority of features are asymmetrically distributed or
+expressed (or both) in the groups (55). Thus, using and reporting scale
 uncertainty should become a standard practice in the analysis of HTS
@@
 
-\bibitem[\citeproctext]{ref-nixon2023scale}
-3. Nixon,M.P., Letourneau,J., David,L.A., Lazar,N.A., Mukherjee,S. and
-Silverman,J.D. (2023) \href{https://arxiv.org/abs/2201.03616}{Scale
-reliant inference}. \emph{https://arxiv.org/abs/2201.03616}.
-
-\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
-4. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
-Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
-Darzi,Y., \emph{et al.} (2017)
-\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
-profiling links gut community variation to microbial load}.
-\emph{Nature}, \textbf{551}, 507--511.
-
-\bibitem[\citeproctext]{ref-Props:2017aa}
-5. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
-Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
-\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
-microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.
+\bibitem[\citeproctext]{ref-nixon2024scale}
+3. Nixon,M.P., McGovern,K.C., Letourneau,J., David,L.A., Lazar,N.A.,
+Mukherjee,S. and Silverman,J.D. (2024)
+\href{https://arxiv.org/abs/2201.03616}{Scale reliant inference}.
 
 \bibitem[\citeproctext]{ref-Thorsen:2016aa}
-6. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
+4. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
 Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
@@
 \bibitem[\citeproctext]{ref-Quinn:2018aa}
-7. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
+5. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
 \href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
@@
 \bibitem[\citeproctext]{ref-Nearing:2022aa}
-8. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
+6. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
 Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
@@
 
+\bibitem[\citeproctext]{ref-Ge:2021aa}
+7. Ge,X., Chen,Y.E., Song,D., McDermott,M., Woyshner,K.,
+Manousopoulou,A., Wang,N., Li,W., Wang,L.D. and Li,J.J. (2021)
+\href{https://doi.org/10.1186/s13059-021-02506-9}{Clipper: P-value-free
+FDR control on high-throughput data from two conditions}. \emph{Genome
+Biol}, \textbf{22}, 288.
+
 \bibitem[\citeproctext]{ref-Li:2022aa}
-9. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+8. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
 \href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
@@
 \bibitem[\citeproctext]{ref-Schurch:2016aa}
-10. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
+9. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-storey2003positive}
-11. Storey,J.D. (2003) The positive false discovery rate: A bayesian
+10. Storey,J.D. (2003) The positive false discovery rate: A bayesian
 interpretation and the q-value. \emph{The annals of statistics},
@@
 \bibitem[\citeproctext]{ref-Cui:2003aa}
-12. Cui,X. and Churchill,G.A. (2003)
+11. Cui,X. and Churchill,G.A. (2003)
 \href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
@@
 \bibitem[\citeproctext]{ref-Zhang:2009aa}
-13. Zhang,S. and Cao,J. (2009)
+12. Zhang,S. and Cao,J. (2009)
 \href{https://doi.org/10.1186/1471-2105-10-402}{A close examination of
@@
 \bibitem[\citeproctext]{ref-Ebrahimpoor:2021aa}
-14. Ebrahimpoor,M. and Goeman,J.J. (2021)
+13. Ebrahimpoor,M. and Goeman,J.J. (2021)
 \href{https://doi.org/10.1093/bib/bbab053}{Inflated false discovery rate
@@
 \textbf{22}.
+
+\bibitem[\citeproctext]{ref-simmons2011false}
+14. Simmons,J.P., Nelson,L.D. and Simonsohn,U. (2011) False-positive
+psychology: Undisclosed flexibility in data collection and analysis
+allows presenting anything as significant. \emph{Psychological science},
+\textbf{22}, 1359--1366.
 
@@
 
-\bibitem[\citeproctext]{ref-Nixon2024B}
-16. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
-normalization: Incorporating scale uncertainty in microbiome and gene
-expression analysis. \emph{bioRxiv},
-\href{https://doi.org/10.1101/2024.04.01.587602}{10.1101/2024.04.01.587602}.
-
-\bibitem[\citeproctext]{ref-Lovell:2015}
-17. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
-Bähler,J. (2015)
-\href{https://doi.org/10.1371/journal.pcbi.1004075}{Proportionality: A
-valid alternative to correlation for relative data}. \emph{PLoS Comput
-Biol}, \textbf{11}, e1004075.
-
-\bibitem[\citeproctext]{ref-Nie:2012aa}
-18. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
-Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
-\href{https://doi.org/10.1016/j.cell.2012.08.033}{C-{M}yc is a universal
-amplifier of expressed genes in lymphocytes and embryonic stem cells}.
-\emph{Cell}, \textbf{151}, 68--79.
+\bibitem[\citeproctext]{ref-Nishijima:2024aa}
+16. Nishijima,S., Stankevic,E., Aasmets,O., Schmidt,T.S.B., Nagata,N.,
+Keller,M.I., Ferretti,P., Juel,H.B., Fullam,A., Robbani,S.M., \emph{et
+al.} (2024) Fecal microbial load is a major determinant of gut
+microbiome variation and a confounder for disease associations.
+\emph{Cell},
+\href{https://doi.org/10.1016/j.cell.2024.10.022}{10.1016/j.cell.2024.10.022}.
+
+\bibitem[\citeproctext]{ref-Bullard:2010}
+17. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
+\href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
+methods for normalization and differential expression in m{RNA-seq}
+experiments}. \emph{BMC Bioinformatics}, \textbf{11}, 94.
+
+\bibitem[\citeproctext]{ref-Jiang:2011aa}
+18. Jiang,L., Schlesinger,F., Davis,C.A., Zhang,Y., Li,R., Salit,M.,
+Gingeras,T.R. and Oliver,B. (2011)
+\href{https://doi.org/10.1101/gr.121095.111}{Synthetic spike-in
+standards for RNA-seq experiments}. \emph{Genome Res}, \textbf{21},
+1543--51.
 
@@
 
+\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
+20. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
+Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
+Darzi,Y., \emph{et al.} (2017)
+\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
+profiling links gut community variation to microbial load}.
+\emph{Nature}, \textbf{551}, 507--511.
+
+\bibitem[\citeproctext]{ref-Props:2017aa}
+21. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
+Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
+\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
+microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.
+
+\bibitem[\citeproctext]{ref-Nixon2024B}
+22. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
+normalization: Incorporating scale uncertainty in microbiome and gene
+expression analysis. \emph{bioRxiv},
+\href{https://doi.org/10.1101/2024.04.01.587602}{10.1101/2024.04.01.587602}.
+
+\bibitem[\citeproctext]{ref-Lovell:2015}
+23. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
+Bähler,J. (2015)
+\href{https://doi.org/10.1371/journal.pcbi.1004075}{Proportionality: A
+valid alternative to correlation for relative data}. \emph{PLoS Comput
+Biol}, \textbf{11}, e1004075.
+
+\bibitem[\citeproctext]{ref-Nie:2012aa}
+24. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
+Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
+\href{https://doi.org/10.1016/j.cell.2012.08.033}{C-{M}yc is a universal
+amplifier of expressed genes in lymphocytes and embryonic stem cells}.
+\emph{Cell}, \textbf{151}, 68--79.
+
 \bibitem[\citeproctext]{ref-Scott:2010}
-20. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
+25. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
 \href{https://doi.org/10.1126/science.1192588}{Interdependence of cell
@@
 \bibitem[\citeproctext]{ref-Yoshikawa:2011aa}
-21. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
+26. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
 Shimizu,H. (2011) \href{https://doi.org/10.1002/yea.1843}{Comprehensive
@@
 \bibitem[\citeproctext]{ref-Lin:2018aa}
-22. Lin,J. and Amir,A. (2018)
+27. Lin,J. and Amir,A. (2018)
 \href{https://doi.org/10.1038/s41467-018-06714-z}{Homeostasis of protein
@@
 \bibitem[\citeproctext]{ref-Zozaya:2010}
-23. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
+28. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
 \href{https://doi.org/10.1128/JCM.00851-09}{Quantitative PCR assessments
@@
 \bibitem[\citeproctext]{ref-Ravel:2010}
-24. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
+29. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
 McCulle,S.L., Karlebach,S., Gorle,R., Russell,J., Tacket,C.O., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hummelen:2010}
-25. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
+30. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
 Changalucha,J., Gloor,G.B. and Reid,G. (2010)
@@
 
-\bibitem[\citeproctext]{ref-Bullard:2010}
-26. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
-\href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
-methods for normalization and differential expression in m{RNA-seq}
-experiments}. \emph{BMC Bioinformatics}, \textbf{11}, 94.
-
 \bibitem[\citeproctext]{ref-Dillies:2013}
-27. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
+31. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
 Jeanmougin,M., Servant,N., Keime,C., Marot,G., Castel,D., Estelle,J.,
@@
 \bibitem[\citeproctext]{ref-Weiss:2017aa}
-28. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
+32. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
 Lozupone,C., Zaneveld,J.R., Vázquez-Baeza,Y., Birmingham,A., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hughes:2005tu}
-29. Hughes,J.B. and Hellmann,J.J. (2005)
+33. Hughes,J.B. and Hellmann,J.J. (2005)
 \href{https://doi.org/10.1016/S0076-6879(05)97017-1}{The application of
@@
 \bibitem[\citeproctext]{ref-Mortazavi:2008}
-30. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
+34. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
 (2008) \href{https://doi.org/10.1038/nmeth.1226}{Mapping and quantifying
@@
 \bibitem[\citeproctext]{ref-wagner:tpm}
-31. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
+35. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
 \href{https://doi.org/10.1007/s12064-012-0162-3}{Measurement of mRNA
@@
 \bibitem[\citeproctext]{ref-Robinson:2010a}
-32. Robinson,M.D. and Oshlack,A. (2010)
+36. Robinson,M.D. and Oshlack,A. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-3-r25}{A scaling normalization
@@
 
+\bibitem[\citeproctext]{ref-Vandesompele:2002aa}
+37. Vandesompele,J., De Preter,K., Pattyn,F., Poppe,B., Van Roy,N., De
+Paepe,A. and Speleman,F. (2002)
+\href{https://www.ncbi.nlm.nih.gov/pubmed/12184808}{Accurate
+normalization of real-time quantitative RT-PCR data by geometric
+averaging of multiple internal control genes}. \emph{Genome Biol},
+\textbf{3}, RESEARCH0034.
+
 \bibitem[\citeproctext]{ref-aitchison1982}
-33. Aitchison,J. (1982) The statistical analysis of compositional data.
+38. Aitchison,J. (1982) The statistical analysis of compositional data.
 \emph{Journal of the Royal Statistical Society: Series B
@@
 \bibitem[\citeproctext]{ref-Anders:2010}
-34. Anders,S. and Huber,W. (2010)
+39. Anders,S. and Huber,W. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-10-r106}{Differential
@@
 \bibitem[\citeproctext]{ref-Aitchison:1986}
-35. Aitchison,J. (1986) The statistical analysis of compositional data
+40. Aitchison,J. (1986) The statistical analysis of compositional data
 Chapman \& Hall, London, England.
@@
 \bibitem[\citeproctext]{ref-fernandes:2013}
-36. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
-(2013) \href{https://doi.org/10.1371/journal.pone.0067019}{ANOVA-like
-differential expression (ALDEx) analysis for mixed population RNA-seq}.
-\emph{PLoS One}, \textbf{8}, e67019.
+41. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
+(2013) \href{https://doi.org/10.1371/journal.pone.0067019}{{ANOVA}-like
+differential expression (ALDEx) analysis for mixed population
+{RNA}-seq}. \emph{PLoS One}, \textbf{8}, e67019.
 
 \bibitem[\citeproctext]{ref-Yerke:2024aa}
-37. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
+42. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
 \href{https://doi.org/10.1186/s40168-023-01747-z}{Proportion-based
@@
 
+\bibitem[\citeproctext]{ref-Dos-Santos:2024aa}
+43. Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and Gloor,G.B.
+(2024) \href{https://doi.org/10.1186/s40168-024-01992-w}{Vaginal
+metatranscriptome meta-analysis reveals functional {BV} subgroups and
+novel colonisation strategies}. \emph{Microbiome}, \textbf{12}, 271.
+
 \bibitem[\citeproctext]{ref-Love:2014aa}
-38. Love,M.I., Huber,W. and Anders,S. (2014)
+44. Love,M.I., Huber,W. and Anders,S. (2014)
 \href{https://doi.org/10.1186/s13059-014-0550-8}{Moderated estimation of
@@
 \bibitem[\citeproctext]{ref-Paulson:2013aa}
-39. Paulson,J.N., Stine,O.C., Bravo,H.C. and Pop,M. (2013)
+45. Paulson,J.N., Stine,O.C., Bravo,H.C. and Pop,M. (2013)
 \href{https://doi.org/10.1038/nmeth.2658}{Differential abundance
@@
 \bibitem[\citeproctext]{ref-fernandes:2014}
-40. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
+46. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
 Edgell,D.R. and Gloor,G.B. (2014)
@@
 \bibitem[\citeproctext]{ref-Gierlinski:2015aa}
-41. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
+47. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-benjamini:1995}
-42. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
+48. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
 rate: A practical and powerful approach to multiple testing.
@@
 \bibitem[\citeproctext]{ref-efron2008FDR}
-43. Efron,B. (2008) Microarrays, empirical bayes and the two-groups
+49. Efron,B. (2008) Microarrays, empirical bayes and the two-groups
 model. \emph{Statist. Sci.}, \textbf{23}, 1--22.
@@
 \bibitem[\citeproctext]{ref-gloor:effect}
-44. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
+50. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
 \href{https://doi.org/10.1080/10618600.2015.1131161}{Displaying
@@
 
-\bibitem[\citeproctext]{ref-Rocha:2020aa}
-45. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
-(2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
-expression analysis in bacteria by RT-qPCR}. \emph{Methods Mol Biol},
-\textbf{2065}, 119--137.
-
-\bibitem[\citeproctext]{ref-Taniguchi:2010aa}
-46. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
-Emili,A. and Xie,X.S. (2010)
-\href{https://doi.org/10.1126/science.1188308}{Quantifying e. Coli
-proteome and transcriptome with single-molecule sensitivity in single
-cells}. \emph{Science}, \textbf{329}, 533--8.
+\bibitem[\citeproctext]{ref-witten2007comparison}
+51. Witten,D. and Tibshirani,R. (2007) A comparison of fold-change and
+the t-statistic for microarray data analysis. \emph{Analysis},
+\textbf{1776}, 58--85.
+
+\bibitem[\citeproctext]{ref-Liu:2022aa}
+52. Liu,X., Zhao,J., Xue,L., Zhao,T., Ding,W., Han,Y. and Ye,H. (2022)
+\href{https://doi.org/10.1186/s12864-022-08465-0}{A comparison of
+transcriptome analysis methods with reference genome}. \emph{BMC
+Genomics}, \textbf{23}, 232.
+
+\bibitem[\citeproctext]{ref-Riaz:2017aa}
+53. Riaz,N., Havel,J.J., Makarov,V., Desrichard,A., Urba,W.J.,
+Sims,J.S., Hodi,F.S., Martín-Algarra,S., Mandal,R., Sharfman,W.H.,
+\emph{et al.} (2017)
+\href{https://doi.org/10.1016/j.cell.2017.09.028}{Tumor and
+microenvironment evolution during immunotherapy with nivolumab}.
+\emph{Cell}, \textbf{171}, 934--949.e16.
+
+\bibitem[\citeproctext]{ref-Gerard:2020aa}
+54. Gerard,D. (2020)
+\href{https://doi.org/10.1186/s12859-020-3450-9}{Data-based RNA-seq
+simulations by binomial thinning}. \emph{BMC Bioinformatics},
+\textbf{21}, 206.
 
 \bibitem[\citeproctext]{ref-dosSantos:2024}
-47. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
+55. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
 Gloor,G.B. (2024) Vaginal metatranscriptome meta-analysis reveals
@@
 
-\bibitem[\citeproctext]{ref-macklaim:2013}
-48. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
-and Gloor,G.B. (2013)
-\href{https://doi.org/10.1186/2049-2618-1-12}{Comparative meta-{RNA}-seq
-of the vaginal microbiota and differential expression by lactobacillus
-iners in health and dysbiosis}. \emph{Microbiome}, \textbf{1}, 12.
+\bibitem[\citeproctext]{ref-Macklaim:2018aa}
+56. Macklaim,J.M. and Gloor,G.B. (2018)
+\href{https://doi.org/10.1007/978-1-4939-8728-3/_13}{From {RNA}-seq to
+biological inference: Using compositional data analysis in
+meta-transcriptomics}. \emph{Methods Mol Biol}, \textbf{1849}, 193--213.
 
 \bibitem[\citeproctext]{ref-Denge00262-18}
-49. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
+57. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
 Wagner-Döbler,I. (2018)
@@
 
+\bibitem[\citeproctext]{ref-Zhang:2020ab}
+58. Zhang,Y., Parmigiani,G. and Johnson,W.E. (2020)
+\href{https://doi.org/10.1093/nargab/lqaa078}{ComBat-seq: Batch effect
+adjustment for RNA-seq count data}. \emph{NAR Genom Bioinform},
+\textbf{2}, lqaa078.
+
 \bibitem[\citeproctext]{ref-Fettweis:2019aa}
-50. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
+59. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
 Parikh,H.I., Huang,B., Arodz,T.J., Edupuganti,L., Glascock,A.L.,
@@
 
+\bibitem[\citeproctext]{ref-macklaim:2013}
+60. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
+and Gloor,G.B. (2013)
+\href{https://doi.org/10.1186/2049-2618-1-12}{Comparative meta-{RNA}-seq
+of the vaginal microbiota and differential expression by
+\emph{{L}actobacillus iners} in health and dysbiosis}.
+\emph{Microbiome}, \textbf{1}, 12.
+
 \bibitem[\citeproctext]{ref-Okuda:2008}
-51. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
+61. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
 Goto,S. and Kanehisa,M. (2008)
@@
 
+\bibitem[\citeproctext]{ref-Rocha:2020aa}
+62. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
+(2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
+expression analysis in bacteria by RT-qPCR}. \emph{Methods Mol Biol},
+\textbf{2065}, 119--137.
+
+\bibitem[\citeproctext]{ref-Taniguchi:2010aa}
+63. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
+Emili,A. and Xie,X.S. (2010)
+\href{https://doi.org/10.1126/science.1188308}{Quantifying e. Coli
+proteome and transcriptome with single-molecule sensitivity in single
+cells}. \emph{Science}, \textbf{329}, 533--8.
+
 \bibitem[\citeproctext]{ref-yeast-absolute}
-52. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
+64. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
 Bähler,J. (2012)
@@
 
+\bibitem[\citeproctext]{ref-McGovern:2023}
+65. McGovern,K.C., Nixon,M.P. and Silverman,J.D. (2023)
+\href{https://doi.org/10.1371/journal.pcbi.1011659}{Addressing erroneous
+scale assumptions in microbe and gene set enrichment analysis}.
+\emph{PLoS Comput Biol}, \textbf{19}, e1011659.
+
+\bibitem[\citeproctext]{ref-Zhang:2021ab}
+66. Zhang,Y., Thompson,K.N., Huttenhower,C. and Franzosa,E.A. (2021)
+\href{https://doi.org/10.1093/bioinformatics/btab327}{Statistical
+approaches for differential expression analysis in metatranscriptomics}.
+\emph{Bioinformatics}, \textbf{37}, i34--i41.
+
 \bibitem[\citeproctext]{ref-Soneson:2013}
-53. Soneson,C. and Delorenzi,M. (2013)
+67. Soneson,C. and Delorenzi,M. (2013)
 \href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
@@
 \bibitem[\citeproctext]{ref-McMurdie:2014a}
-54. McMurdie,P.J. and Holmes,S. (2014)
+68. McMurdie,P.J. and Holmes,S. (2014)
 \href{https://doi.org/10.1371/journal.pcbi.1003531}{Waste not, want not:
@@
 \bibitem[\citeproctext]{ref-hawinkel2017}
-55. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
+69. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
 \href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
 Microbiome differential abundance methods do not control the false
-discovery rate}. \emph{BRIEFINGS IN BIOINFORMATICS}.
+discovery rate}. \emph{Briefings in Bioinformatics}.
+
+\bibitem[\citeproctext]{ref-Zhang:2021aa}
+70. Zhang,Y., Thompson,K.N., Branck,T., Yan,Y., Nguyen,L.H.,
+Franzosa,E.A. and Huttenhower,C. (2021)
+\href{https://doi.org/10.1146/annurev-biodatasci-031121-103035}{Metatranscriptomics
+for the human microbiome and microbial community functional profiling}.
+\emph{Annu Rev Biomed Data Sci}, \textbf{4}, 279--311.
 
 \bibitem[\citeproctext]{ref-Brooks:2024aa}
-56. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
+71. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
 \href{https://doi.org/10.1038/s41576-023-00679-6}{Challenges and best
@@
 \bibitem[\citeproctext]{ref-gustafson2015bayesian}
-57. Gustafson,P. (2015) Bayesian inference for partially identified
+72. Gustafson,P. (2015) Bayesian inference for partially identified
 models: Exploring the limits of limited data CRC Press.
@@
 \bibitem[\citeproctext]{ref-Thellin:1999aa}
-58. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
+73. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
 Hennen,G., Grisar,T., Igout,A. and Heinen,E. (1999)
@@
 \bibitem[\citeproctext]{ref-SEQCux2fMAQC-III-Consortium:2014aa}
-59. SEQC/MAQC-III Consortium (2014)
+74. SEQC/MAQC-III Consortium (2014)
 \href{https://doi.org/10.1038/nbt.2957}{A comprehensive assessment of
@@
 \bibitem[\citeproctext]{ref-Song:2023aa}
-60. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
+75. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
 Klatt,N.R., Hensley-McBain,T. and Wu,M.C. (2023)

### 2025-04-14 go3.Rmd
Penultimate before cleaning

@@
 ---
-title: "Explicit Scale Simulation for analysis of RNA-sequencing  with ALDEx2"
+title: "Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2"
 shorttitle: "Scale ALDEx2"
@@
   - name: Justin Silverman 
-    affiliation: 
-      College of Information Sciences and Technology and Department of Medicine, Pennsylvania   State University
-      Department of Medicine, Pennsylvania State University
-      Institute for Computational and Data Science, Pennsylvania State University
-      Department of Statistics, Pennsylvania State University
+    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania State University, Department of Medicine, Pennsylvania State University, Institute for Computational and Data Science, Pennsylvania State University, Department of Statistics, Pennsylvania State University
 cite-method: citeproc 
@@
   - \usepackage{setspace}
+  - \usepackage[left]{lineno}
+
 package: ALDEx2
 
-abstract: 
-  In high-throughput sequencing (HTS) studies,  sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system.
+abstract: In high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system.
 Typically a statistical normalization removes unwanted technical variation in the data or the parameters of the model to enable differential abundance analyses.
@@
 
-\doublespacing 
-\singlespacing
+\doublespacing \singlespacing 
+\linenumbers
 
-High-throughput sequencing (HTS) is a ubiquitous
-tool used to explore many biological phenomenon such as gene expression
-(single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial
-community composition (16S rRNA gene sequencing, shotgun metagenomics) and
-differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a
-sample from the environment, making a library, multiplexing (merging) multiple
-libraries together, and then applying a sample of the multiplexed library to a
-flow cell. Each of these steps is a compositional sampling step as only a
-fixed-size subsample of nucleic acid is carried over to subsequent steps. Thus,
-with each sampling step the connection between the actual size of the sampled
-DNA pool and the scale (e.g., size, microbial load, or total gene expression) of
-the measured biological system is degraded or lost. In the end, the information
-contained in the data relates only to relative abundances and  has an arbitrary
-scale imposed by the sequencing
-process [@lovell:2011;@Quinn:2019aa;@nixon2024scale]. Researchers
-can use modified experimental protocols [@Jiang:2011aa;@Loven:2012aa;@Vandeputte:2017aa;@Props:2017aa] or machine learning methods [@Nishijima:2024aa] to uncover the
-biological variation in scale. However,  wet-lab protocols only provide information on
-the size of the data downstream of the step in the sample preparation protocol where the
-intervention was made and  introduce an additional source of variation that must be accounted for [@Loven:2012aa].
-The computational methods developed for microbiome analysis have low correlation with the actual scale but are useful [@Nishijima:2024aa].
-In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an open
-challenge.
+High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing).
+HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to a flow cell.
+Each of these steps is a compositional sampling step as only a fixed-size subsample of nucleic acid is carried over to subsequent steps.
+Thus, with each sampling step the connection between the actual number of molecules in the sampled DNA pool and the environmental scale (e.g., total number of molecules, microbial load, or total gene expression) of the measured biological system is degraded or lost.
+In the end, the information contained in the data relates only to relative abundances and  has an arbitrary scale imposed by the sequencing process [@lovell:2011;@Quinn:2019aa;@nixon2024scale].
 
-The analysis of  HTS data suffers from several known problems that can be
-traced, in whole or in part, to  misspecification of scale. The first issue is
-poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Ge:2021aa;@Li:2022aa], exhibited
-as dataset-dependent FDR control and by the disconnect between statistical and
-biological significance [@Schurch:2016aa]. In current practice,  these issues are
-addressed  by a dual-filtering method, whereby both a low p-value (or
-equivalently a low q-value following FDR correction [@storey2003positive]) and a
-large difference between groups is used to identify interesting transcripts or
-genes for follow-up analysis [@Cui:2003aa;@Schurch:2016aa]. This double-filtering approach is
-graphically exemplified by the volcano plot [@Cui:2003aa], but is known to not
-appropriately control the  FDR [@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second
-issue is poor performance when analyzing data where the mean change between
-groups is non-zero [@nixon2024scale]. Such asymmetric data can arise when a gene set is expressed
-in one group but not the other, or when one group contains different gene
-content from the other. This type of data frequently arises in in-vitro
-selection experiments (SELEX), transcriptome analysis, and microbiome analysis
-[@Wu2021]. The third issue is that the actual scale of the environment is often
-a major confounding variable during analysis [@nixon2024scale;@Nishijima:2024aa]. The final issue is that these problems become more pronounced as more
-samples are collected; that is, more information results in a worsening of the
-accuracy of the analysis [@Li:2022aa;@nixon2024scale;@Nixon2024B].
+The analysis of  HTS data suffers from several known problems that can be traced, in whole or in part, to misspecification of scale in the output data.
+The first issue is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Ge:2021aa;@Li:2022aa], exhibited as dataset-dependent FDR control which is observed as a disconnect between statistical and biological significance [@Schurch:2016aa;@Li:2022aa].
+In current practice, this issue is addressed  by a dual-filtering method, whereby both a low p-value (or equivalently a low q-value following FDR correction [@storey2003positive]) and a large difference between groups is used to identify  transcripts or genes of interest for follow-up analysis [@Cui:2003aa;@Schurch:2016aa].
+This double-filtering approach is graphically exemplified by the volcano plot [@Cui:2003aa], but is known to not appropriately control the  FDR [@Zhang:2009aa;@Ebrahimpoor:2021aa].
+Since there is no standard way of determining what fold-change cutoff should be used researchers have unlimited degrees of  freedom which is known to lead to unreliable inference [@simmons2011false].
+In particular Li et al.
+[@Li:2022aa] used patient or clinical-derived transcriptome datasets and found that many methods suffer from an extremely high false positive rate.
+The second issue is poor performance when analyzing data where the mean change between groups is non-zero [@nixon2024scale].
+Such asymmetric data can arise when a gene set is expressed in one group but not the other, or when one group contains different gene content from the other.
+This type of data frequently arises in in-vitro selection experiments (SELEX), transcriptome analysis, and microbiome analysis [@Wu2021].
+The third issue is that the actual scale of the environment is often a major confounding variable during analysis [@nixon2024scale;@Nishijima:2024aa].
+This has long been known in transcriptome analysis and was a major driver for the development of normalization factors [@Bullard:2010] and the use of molecular spike-ins  to provide a reference set of known counts [@Jiang:2011aa;@Loven:2012aa;@Vandeputte:2017aa;@Props:2017aa] to estimate scale.
+While often useful, spike-in methods only provide information downstream of the step in the sample preparation protocol where the intervention was made and  introduce an additional source of variation that must be accounted for [@Loven:2012aa].
+In the microbiome field, a recent landmark paper showed that biological scale was a major unacknowledged confounder in many human  analyses [@Nishijima:2024aa].
+These authors built a machine learning model   to uncover the biological variation in scale and including this information  was useful despite exhibiting only a  modest correlation (estimated to be 0.6)  with the  scale of data external to the training set [@Nishijima:2024aa].
+The final issue is that all the above problems become more pronounced as more samples are collected; that is, more information results in optimizing for a precise but inaccurate analysis [@Li:2022aa;@nixon2024scale;@Nixon2024B].
 
-The  four problems were recently shown by Nixon et al. [@nixon2024scale]
-to be a result of a mismatch between the underlying size or scale of the system
-and the assumptions of the normalizations used for the analysis of HTS.
-Biological variation in scale often represents an important unmeasured
-confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the
-cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA
-content than  non-transformed cells [@Nie:2012aa], and this dramatically skews
-transcriptome analysis unless spike-in approaches are used [@Loven:2012aa]. In addition, wild-type and mutant
-strains of cell lines, yeast or bacteria  have different growth rates and RNA
-contents under different conditions, which  affect our ability to identify truly
-differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As
-another example, the total bacterial load of the vaginal microbiome differs by
-1-2 orders of magnitude in absolute abundance between the healthy and bacterial
-vaginosis states [@Zozaya:2010], and the composition between these states is
-dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of
-any of these systems includes both relative change (composition) and absolute
-abundance (scale).  Current methods access only the compositional information
-yet make implicit assumptions about the scale [@nixon2024scale,@Nixon2024B].
+The  four problems were recently shown by Nixon et al.
+[@nixon2024scale]  to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.
+Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Loven:2012aa;@Lovell:2015;@Wu2021;@Nishijima:2024aa].
+For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis unless spike-in approaches are used [@Loven:2012aa].
+In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa].
+As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the cell and RNA composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010].
+Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).
+Current methods access only the compositional information yet make implicit assumptions about the scale [@nixon2024scale,@Nixon2024B].
 
-Recently, Nixon et al. [-@nixon2024scale] showed that the challenge of
-non-biological variation in sequencing depth be viewed as a problem of
-partially-identified models. They showed that \emph{all} normalizations make
-some assumption about  scale but these implicit assumptions are often
-inappropriate and difficult to interpret. This causes different normalizations
-to provide different outputs when applied to the same dataset
-[@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
-Intuitively, normalizations in widespread use  assume that either all samples
-have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM
-[@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample
-can be chosen as a reference to which the others are scaled e.g. the TMM
-[@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982];
-or that different sub-parts of each sample maintain a constant scale across
-samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is
-appropriate e.g the CLR [@Aitchison:1986] and its derivatives.
+Nixon et al.
+[-@nixon2024scale] showed that the challenge of non-biological variation in sequencing depth could be explained as a problem of partially-identified models.
+They showed that \emph{all} normalizations make some assumption about  scale but these implicit assumptions are often inappropriate and difficult to interpret.
+As a result, different normalizations  provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
+Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g.
+proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g.
+the TMM [@Robinson:2010a], the LVHA [@Wu2021]  qPCR [@Vandesompele:2002aa],  the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g.
+the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@Aitchison:1986] and its derivatives.
+Notably when the assumption of similar scale across samples is not violated, or violated only weakly,  any method will provide a reasonable analysis.
+The problem arises  when this assumption is violated, in which case all tools fail without warning.
 
-The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict
-assumption about scale through the CLR normalization [@nixon2024scale].
-This assumption was often close enough to the true value to be useful, but was not always the a good estimate and could be outperformed by other normalizations [@Yerke:2024aa].
-Nixon et al. [@nixon2024scale] showed that better scale assumptions resulted in
-more reproducible data analysis including better control of both false positive
-and false negative results. We recently modified ALDEx2  to explicitly model the
-scale over a range of reasonable normalization parameters, and showed
-significant improvements in performance in microbiome and in-vitro selection
-experiments [@Nixon2024B]. Here, we briefly  review these modifications and show how scale uncertainty can greatly improve modeling in transcriptome
-and meta-transcriptome datasets to provide more robust and reproducible results.
+```{r, out.width="0.6\\linewidth", , echo=FALSE, include=TRUE, fig.align="center", fig.cap=c("Mismatches between estimated scale and true scale lead to poor estimation of high throughput sequencing data.
+All normalizations used for differential abundance analysis make some strict assumption about the scale of the environment as shown in line 1.
+In this example all normalizations produce a biased estimate of the environmental scale, but estimate C is the closest to the truth.
+Adding uncertainty to normalization C as represented by the distribution in the line 2 leads to less bias and now includes the actual environmental scale in the assumption.
+As shown in line 3 in some cases it may be useful to adjust the centre of the scale uncertainty estimate if the initial normalizations give very poor estimates of the underlying environment.")}
+knitr::include_graphics("./go3_files/figure-latex/scale-explained.pdf")
+```
+
+The original scale-naive ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization and we found that the CLR was very sensitive to violations of the assumption of scale identity between groups [@Wu2021].
+Moreover, when the assumption of identity was not true the CLR used by ALDEx2 could be outperformed by other normalizations in simulation studies [@Yerke:2024aa].
+As illustrated graphically in Figure 1, Nixon et al.
+[@nixon2024scale] showed through simulation that introducing uncertainty in the scale assumptions, and in extreme cases altering the location of the scale assumption, resulted in more reproducible data analysis including better control of both false positive and false negative results.
+We  modified ALDEx2  to explicitly model uncertainty in scale over a range of reasonable normalization parameters, and showed significant improvements in performance in microbiome and in-vitro selection experiments [@Nixon2024B] and in a vaginal metatranscriptome analysis [@Dos-Santos:2024aa].
+Here, we briefly  review these modifications and show how adding scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide substantially more robust and reproducible results.
 
@@
 
-Formal and expanded descriptions of the concepts that follow are given in
-[@nixon2024scale;@Nixon2024B].   To be  concrete, we let \(\mathbf{Y}\) denote
-the \emph{measured} \(D \times N\) matrix of sequence counts with elements
-\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to
-feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).  Likewise, we
-can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the
-biological system from which sample \(n\) was obtained. We can think of
-\(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\)
-(e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e.,
-proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
-\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
-\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
-\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note
-that with these definitions \(\mathbf{W}\) can be written as the element-wise
-combination of scale and composition:
-\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the
-logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
-\mathbf{W}^{\perp}_{n}\).
+Formal and expanded descriptions of the concepts that follow are given in [@nixon2024scale;@Nixon2024B].
+To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).
+Likewise, we can denote \(\mathbf{W}_{dn}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained.
+We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{Tot}\) (e.g., totals) and the composition \(\mathbf{W}^{Comp}\) (i.e., proportions).
+That is, \(\mathbf{W}^{Tot}\) is a \(N\)-vector with elements \(\mathbf{W}^{Tot}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{Comp}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{Comp}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{Tot}_{n}\).
+Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{Comp}_{dn}\mathbf{W}^{Tot}_{n}\), or as the logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{Comp}_{dn} + \log \mathbf{W}^{Tot}_{n}\).
 
-Many of  the normalizations in widespread use in tools such as DESeq2
-[@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2
-[@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}
-\approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by
-some function of the observation. We use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\
-\)) notation to indicate that the output is an estimate of the true value.  The
-technical variation in sequencing depth
-(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data
-\(\mathbf{Y}\) provides us with information about the system composition
-\(\mathbf{W}^{\parallel}\) but little to no information in the system scale
-\(\mathbf{W}^{\perp}\) (Lovell et al. 2011).
+Many of  the normalizations used in tools such as DESeq2 [@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2 [@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation.
+We use the hat  notation (\(\ \hat{}\ \)) to indicate that the output is an estimate of the true value.
+The technical variation in sequencing depth, which is often called "library size" has no relationship with the actual number of molecules in the sampled environment [@lovell:2011].
+In other words (\(\mathbf{Y}^{Tot}_{n}=\sum_{d}\mathbf{Y}_{dn}\))  the observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{Comp}\) but little to no information in the system scale \(\mathbf{W}^{Tot}\) (Lovell et al.
+2011).
 
@@
 
-The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose
-toolbox to  model the uncertainty of HTS data and to use that model to estimate
-the underlying LFC (log-fold change) significance.  At a high-level, ALDEx2  has
-three connected components to estimate the uncertainty inherent in HTS datasets.
-First, the tool accounts for the uncertainty of the sequencing counts using
-Dirichlet multinomial sampling to build a probabilistic model of the data; i.e.,
-\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).  Secondly,
-ALDEx2 uses the centred log-ratio transformation to scale the data
-[@fernandes:2013]. However, this step was modified recently to account for scale
-uncertainty and misspecification [@nixon2024scale,@Nixon2024B] via a scale
-model, explained  with more details in [@nixon2024scale;@Nixon2024B] and
-summarized in the next paragraph. Finally, a standard null-hypothesis test and a
-non-parametric estimate of mean standardized difference are used to report on
-the finite sample variation. These sources of uncertainty and variation are
-combined via reporting the expected values from a Monte-Carlo simulation
-framework. For simplicity, we use the term 'difference' to refer to the absolute
-difference between groups, and 'dispersion' to refer to the  within-condition
-difference or pooled variance as defined in [@fernandes:2013]. These are
-calculated on a \(\log_2\) scale. For more details on ALDEx2 see
-[@fernandes:2013;@fernandes:2014;@nixon2024scale;@Nixon2024B].
+The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox to  model the uncertainty of HTS data and to use that model to estimate the significance of the underlying LFC (log-fold change).
+At a high-level, ALDEx2  has three connected components to estimate the uncertainty inherent in HTS datasets.
+First, the tool accounts for the uncertainty of the sequencing counts using Dirichlet multinomial sampling to build a probabilistic model of the data; i.e., \(\mathbf{\hat{W}}^{Comp} \approx \mathrm{Dir}(\mathbf{Y})\).
+Secondly, ALDEx2 uses the centred log-ratio transformation to scale the data [@fernandes:2013].
+It was this step that was modified  to account for scale uncertainty and misspecification [@nixon2024scale,@Nixon2024B] explained  with more details in [@nixon2024scale;@Nixon2024B] and summarized in the next paragraph.
+Finally, a standard null-hypothesis test and a non-parametric estimate of mean standardized difference are used to report on the finite sample variation.
+These sources of uncertainty and variation are combined via reporting the expected values from a Monte-Carlo simulation framework.
+For simplicity, we use the term 'difference' to refer to the absolute difference between groups, and 'dispersion' to refer to the  within-condition difference or pooled variance as defined in [@fernandes:2013].
+These are calculated on a \(\log_2\) scale.
+For more details on ALDEx2 see [@fernandes:2013;@fernandes:2014;@nixon2024scale;@Nixon2024B].
 
-Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a
-specialized type of statistical model which Nixon et al. [@nixon2024scale] term
-a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon et al.
-[@nixon2024scale] generalized the concept of normalizations by introducing  the
-concept of a \textit{scale model} to account for potential error in the centred
-log-ratio normalization step.  They did this by including a model for
-\(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by ALDEx2 makes the
-assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the
-geometric mean of sample n, which while being a random variable, is essentially
-constant across each Monte-Carlo replicate, but that differs between samples.
-With this modification, ALDEx2 can be generalized by considering probability
-models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean
-\(1/G_{n}\). For example, the following scale model generalizes the CLR:
+Scale models were incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which Nixon et al.
+[@nixon2024scale] term a \textit{Scale Simulation Random   Variable} (SSRV).
+To do this, Nixon et al.
+[@nixon2024scale] generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error in the centred log-ratio normalization step.
+They did this by including a model for \(\mathbf{\hat{W}}^{Tot}_{n}\).
+The CLR normalization used by ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{Tot}_{n}=1/G_{n}\), where \(G_n\) is the geometric mean of the counts (or the corresponding proportions)of each part in sample n, which while being a random variable, is essentially constant across each Monte-Carlo replicate, but that differs between samples.
+With this modification, ALDEx2 can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{Tot}_{n}\) that have mean \(1/G_{n}\).
+For example, the following scale model generalizes the CLR:
 
-\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda
-\sim N(\mu, \gamma^{2})\].
+\[\log \mathbf{\hat{W}}^{Tot}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2}).\]
 
-This formulation is quite flexible [@nixon2024scale;@Nixon2024B]. In the simple
-or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter
-drawn from a log-Normal distribution[@nixon2024scale]. Adding scale uncertainty with the \(\gamma\) paramenter  controls only the degree of uncertainty of the CLR assumption for the
-\(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and
-\(x_{n}=0\) denotes control). In the advanced or 'informed' configuration,
-\(\mu\) takes different values for each group and  controls the location of
-the LFC assumption; combining \(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the location and the scale. An example of both the default and informed approaches is
-given for a microbiome dataset in [@Nixon2024B] showing increased sensitivity and specificity. Here we show that these
-approaches also work well in transcriptome and metatranscriptome datasets. These
-modifications are  instantiated in ALDEx2 which is  the first software package
-designed for SSRV-based inference.
+This formulation is quite flexible [@nixon2024scale;@Nixon2024B].
+In the simple or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter drawn from a log-Normal distribution [@nixon2024scale].
+Adding scale uncertainty with the \(\gamma\) paramenter (as shown in Figure 1:1) controls only the degree of uncertainty of the CLR assumption for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control).
+In the advanced or 'informed' configuration, \(\mu\) takes different values for each group and  controls the location of the LFC assumption (as shown in Figure 1:2); combining \(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the location and the scale.
+In the manuscript where the idea of scale was original derived, Nixon et al.
+[@nixon2024scale] conducted extensive simulations showing that both the default and informed approaches exhibit   increased sensitivity and specificity [@Nixon2024B].
+In this report we show that these approaches also work well to control the FDR in transcriptome and metatranscriptome datasets.
+We also provide some additional insights into how this is achieved.
+These modifications are  instantiated in ALDEx2 which is  the first software package designed for SSRV-based inference.
 
@@
 
-Gierliński et al. [@Gierlinski:2015aa]  generated a  highly replicated yeast
-transcriptome dataset to compare gene expression between a wild-type strain and
-a snf2 gene knockout, \(\Delta\)snf2. This dataset was  used to test several
-RNA-seq tools for their power  to detect the set of differentially abundant
-transcripts identified in the full dataset when the data was subset
-[@Schurch:2016aa]. In this  study each tool had its own 'gold standard' set of
-transcripts with different tools identifying between between 65% to >80% of all
-transcripts as being significantly different. Since the majority of transcripts
-were significantly different, the authors suggested that it was more appropriate
-to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995]
-corrected p-value (q-value) plus a difference cutoff  to limit the number of
-identified transcripts to a much smaller fraction of the total.  Nixon et al,
-@Nixon2024B showed that adding even a small amount of  scale uncertainty with
-ALDEx2  dramatically reduced the number of significant transcripts identified,
-removing the need for the dual cutoff approach in this dataset and others. Below
-we include an intuitive explanation of why and how incorporating scale uncertainty
-achieved this outcome using a setting of \(\gamma=0.5\). The approach is in line with
-the recommendations of [@Nixon2024B] and  gives results comparable to those
-proposed in [@Schurch:2016aa].
+Gierliński et al.
+[@Gierlinski:2015aa]  generated a  highly replicated yeast transcriptome dataset to compare gene expression between a wild-type strain and a snf2 gene knockout, \(\Delta\)snf2.
+This dataset of 86 samples (44 and 42 per group) is an example of technical growth replicate experiments common in the literature.
+The dataset was  used to test several RNA-seq tools for their power  to detect the set of differentially abundant transcripts identified in the full dataset when the data was subset [@Schurch:2016aa].
+In this original study each tool had its own 'gold standard' set of transcripts with different tools identifying between between 65% to >80% of all transcripts as being significantly different.
+Since the majority of transcripts were significantly different, the authors suggested that it was more appropriate to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995] corrected p-value (q-value) plus a difference cutoff  to limit the number of identified transcripts to a much smaller fraction of the total.
+In other datasets, Nixon et al, @Nixon2024B showed that adding even a small amount of  scale uncertainty with ALDEx2  dramatically reduced the number of significant transcripts identified, removing the need for the dual cutoff approach in this dataset and others.
 
@@
 #ALDEx2 
-library(ALDEx2) 
+library(ALDEx2, warn.conflicts=F) 
 load(file="analysis/yst.all.Rda") 
 load(file="analysis/yst.s.all.Rda") 
-load(file="analysis/yst.1.all.Rda")
-# DESeq2
+load(file="analysis/yst.1.all.Rda") # DESeq2 
 load('analysis/res.Rda')
@@
 sig.all <- yst.all$we.eBH < 0.05 
-sig.t <- yst.all$we.eBH < 0.05 &
-abs(yst.all$diff.btw) > 1.4 
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4 
 sig.s <- yst.s.all$we.eBH < 0.05 
@@
 
-We start with the assumption that not all statistically significant differences
-are biologically relevant [@efron2008FDR], and that such a large number of
-significantly different transcripts breaks the  necessary assumption for DA/DE
-expression that most parts be invariant [@Dillies:2013]. As noted,
-transcriptomics commonly uses a dual cutoff approach that is graphically
-exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa]. Using either DESeq2
-or ALDEx2, a majority of transcripts are statistically significantly different
-between groups  with a  q-value cutoff of \( \le 0.05\); i.e. `r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or
-`r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2)
-of the `r nrow(yst.all)` transcripts. These values are in line with those
-observed by [@Schurch:2016aa]. Such  large numbers of statistically significant
-transcripts seems biologically unrealistic. That `r length(setdiff(sig.ald,
-sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
-identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does
-not, suggests that the choice of normalization plays a role in which results are
-returned as significant and that some, if not the majority, are driven by
-technical differences in the analysis [@Dillies:2013;@Li:2022aa].
-
-```{r plot1, echo=FALSE, warning=FALSE, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis.
+```{r yst-plot, echo=FALSE, warning=FALSE, fig.align="center", out.width="0.8\\linewidth", fig.dim=c(6,7), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis.
 ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset.
@@
 
+load(file="analysis/yst.all.Rda") 
+load(file="analysis/yst.s.all.Rda") 
+load(file="analysis/yst.1.all.Rda") # DESeq2 
+load('analysis/res.Rda')
+
+# with gamma = 0.5 
+load('analysis/x.s.all.Rda') 
+sig.des <- which(res@listData$padj < 0.05) 
+sig.ald <- which(yst.all$we.eBH < 0.05) 
+sig.s.ald <- which(yst.s.all$we.eBH < 0.05)
+
+sig.all <- yst.all$we.eBH < 0.05 
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4 
+sig.s <- yst.s.all$we.eBH < 0.05 
+sig.1 <- yst.1.all$we.eBH < 0.05
+
 par(mfrow=c(2,2)) 
-# volcano plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300),
+# volcano 
+# plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
 #  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
+#   
 # title('D: DESeq2 volcano', adj=0, line= 0.8) 
-# points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des
-# ] + 1e-300), col=rgb(1,.66,0,0.3), pch=19, cex=0.5)
-# points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s
-# .ald] + 1e-300), col=rgb(1,0,0,1), pch=19, cex=0.5) abline(v=c(-1.4,1.4),
-# lty=2,lwd=2, col='grey')
+# 
+# points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), 
+#   col=rgb(1,.66,0,0.3), pch=19, cex=0.5) 
+# points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300),     
+#   col=rgb(1,0,0,1), pch=19, cex=0.5) 
+#  
+# abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
@@
 title(expression(paste("A: Volcano: ", gamma, " = 0")), adj=0, line= 0.8) 
-points(yst.all$diff.btw[sig.ald],
--1*log10(yst.all$we.eBH[sig.ald]  +1e-70), col=rgb(1,.66,0,.3), cex=0.5, pch=19)
-points(yst.all$diff.btw[sig.s.ald], -1*log10(yst.all$we.eBH[sig.s.ald]  +1e-70),
-col=rgb(1,0,0,1), cex=0.5, pch=19) 
+points(yst.all$diff.btw[sig.ald], -1*log10(yst.all$we.eBH[sig.ald]  +1e-70), col=rgb(1,.66,0,.3), cex=0.5, pch=19) 
+points(yst.all$diff.btw[sig.s.ald], -1*log10(yst.all$we.eBH[sig.s.ald]  +1e-70), col=rgb(1,0,0,1), cex=0.5, pch=19) 
 abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-plot(yst.s.all$diff.btw, -1*log10(yst.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
-xlab=l2d, ylab=l10q) 
+plot(yst.s.all$diff.btw, -1*log10(yst.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1), xlab=l2d, ylab=l10q) 
 title(expression(paste("B: Volcano: ", gamma, " = 0.5")), adj=0, line= 0.8) 
@@
 
-#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange,
-#xlim=c(0,5), col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
+#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, #xlim=c(0,5), col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference') 
 #title('A: DESeq2 effect', adj=0, line= 0.8) 
 #points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)), 
-#res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), pch=19, cex=0.5)
-#points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
-#res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), pch=19, cex=0.5)
-#abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
+#res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), pch=19, cex=0.5) #points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)), 
+#res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), pch=19, cex=0.5) #abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), xlab='Dispersion',
-ylab=l2d) 
-title(expression(paste("C: Effect: ", gamma, " = 0")), adj=0, line=
-0.8) 
-points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald],
-col=rgb(1,.66,0,0.3), cex=0.5, pch=19) 
-abline(h=c(-1.4,1.4), lty=2,lwd=2,
-col='grey') 
-points(yst.all$diff.win[sig.s.ald], yst.all$diff.btw[sig.s.ald],
-col=rgb(1,0,0,0.6), cex=0.5, pch=19) 
-abline(h=c(-1.4,1.4), lty=2,lwd=2,
-col='grey')
+plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), xlab='Dispersion', ylab=l2d) 
+title(expression(paste("C: Effect: ", gamma, " = 0")), adj=0, line= 0.8) 
+points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald], col=rgb(1,.66,0,0.3), cex=0.5, pch=19) 
+abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey') 
+points(yst.all$diff.win[sig.s.ald], yst.all$diff.btw[sig.s.ald], col=rgb(1,0,0,0.6), cex=0.5, pch=19) 
+abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
 
-plot(yst.s.all$diff.win, yst.s.all$diff.btw, col=rgb(0,0,0,0.1),
-xlab='Dispersion', ylab=l2d, xlim=c(0.1,5)) 
-title(expression(paste("D: Effect:
-", gamma, " = 0.5")), adj=0, line= 0.8) 
-points(yst.s.all$diff.win[sig.s.ald],
-yst.s.all$diff.btw[sig.s.ald], col=rgb(1,0,0,0.5), cex=0.5, pch=19)
+plot(yst.s.all$diff.win, yst.s.all$diff.btw, col=rgb(0,0,0,0.1), xlab='Dispersion', ylab=l2d, xlim=c(0.1,5)) 
+title(expression(paste("D: Effect: ", gamma, " = 0.5")), adj=0, line= 0.8) 
+points(yst.s.all$diff.win[sig.s.ald], yst.s.all$diff.btw[sig.s.ald], col=rgb(1,0,0,0.5), cex=0.5, pch=19) 
 abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
@@
 
-The Volcano plots in Figure 1 A and B show  that adding scale uncertainty
-increases the minimum q-value and increases the concordance between the q-value
-and the difference between groups (compare panels A and B). The effect plots
-[@gloor:effect] in Figure 1C  shows that the majority of significant transcripts
-(red, orange) have negligible differences between groups and  very low
-dispersion. We suggest that this low dispersion is driven by the experimental
-design which is actually a technical wet lab replication  rather than a true
-biological replication design [@Gierlinski:2015aa]. Scale uncertainty can be
-incorporated using the \texttt{gamma} parameter that controls the amount of
-noise added to the CLR mean assumption when we call either \texttt{aldex()}, or
-\texttt{aldex.clr()}. Figure 1 B,D shows that  setting  \( \gamma=0.5 \) results
-in `r length(sig.s.ald)` which is far fewer significant transcripts than in the
-naive analysis  and we observe that  the minimum dispersion increases from  `r round(min(yst.all$diff.win), 2)` (\(\gamma = 0\) ) to  `r round(min(yst.s.all$diff.win), 2)` (\(\gamma=0.5\)).
+We start with the assumption that not all statistically significant differences are biologically relevant [@efron2008FDR], and that a result where the majority of transcripts are significant breaks the  necessary  assumption for DA/DE expression that most parts be invariant [@Dillies:2013].
+Transcriptomic analysis commonly uses a dual cutoff approach graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa].
+Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a  q-value cutoff of \( \le 0.05\); i.e.
+`r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or `r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2) of the `r nrow(yst.all)` transcripts.
+These values are in line with those observed by [@Schurch:2016aa].
+Such  large numbers of statistically significant transcripts seems biologically unrealistic.
+That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some, if not the majority, are driven by technical differences in the analysis [@Dillies:2013;@Li:2022aa] or are false positives.
 
-```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="Adding scale uncertainty changes the dispersion distribution.
-Panel A shows a plot of the expected value for relative abundance vs the expected value for the pooled dispersion as output by \\texttt{aldex.effect}.
-The dashed  horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5, and the light colored lines are lowess lines of fit through the center of mass of the data.
-Panel B plots the dispersion difference between  \\(\\gamma = 1\\) and \\(\\gamma = 0\\); note the non-linear relationship that highlights  the rotation that is evident in Panel A.
-The colored lines indicate the lowess line of fit through the centre of mass of the plot for the various  populations of points.
-The grey line is the total population and shows the difference \\(\\Delta\\), the red line is the population of significant transcripts (\\*) with \\(\\gamma = 0\\), the orange line is the population of significant transcripts with a difference threshold (T) of about \\(\\pm 2^{1.4}\\)-fold change, the blue line is the population of significant transcripts with \\(\\gamma = 0.5\\), and the cyan line is the significant population with \\(\\gamma = 1\\).
-\\(\\Delta\\): Difference, *: significant, T: thresholded.
-"}
+The Volcano plots in Figure  2 A and B show  that adding scale uncertainty increases the minimum q-value and increases the concordance between the q-value and the difference between groups (compare panels A and B).
+The effect plots [@gloor:effect] in Figure 2C  shows that the majority of significant transcripts (red, orange) have negligible differences between groups and  very low dispersion.
+We suggest that this low dispersion is driven by the experimental design which is actually a technical wet lab replication  rather than a true biological replication design [@Gierlinski:2015aa].
+Scale uncertainty can be incorporated using the \texttt{gamma} parameter that controls the amount of noise added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
+Figure 2 B,D shows that  setting  \( \gamma=0.5 \) now results in `r length(sig.s.ald)` which is far fewer statistically significant transcripts than in the naive analysis  and we observe that  the minimum dispersion increases from  `r round(min(yst.all$diff.win), 2)` (\(\gamma = 0\) ) to  `r round(min(yst.s.all$diff.win), 2)` (\(\gamma=0.5\)).
 
-cuts <- yst.all$rab.all > -0.5 & yst.all$rab.all < 5 
-par(mfrow=c(1,2)) 
-mn.mid <- median(yst.all$diff.win[cuts]) 
-mn.mid.s <- median(yst.s.all$diff.win[cuts])
-mn.mid.1 <- median(yst.1.all$diff.win[cuts])
+It is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both q-values and fold-changes [@Schurch:2016aa], and these were first proposed for microarray experiments through volcano plots [@Cui:2003aa].
+Note that there is considerable variation in recommended cutoff values [@Schurch:2016aa], and that this controversy has persisted ever since fold-change was suggested [@witten2007comparison].
+Unfortunately, universal cutoff fold-change values cannot be identified in part because different tools have intrinsically different variance in their log_2-fold change ranges [@Liu:2022aa].
+This has led to the widespread practice of applying a post-hoc fold-change cutoff to reduce the number of positive identifications to a manageable proportion of the whole dataset.
+Here, we applied the dual-cutoff method using a fold-change of at least a \(2^{1.4}\) fold change that reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
+This cutoff was chosen for convenience and is mid-way between the high and low fold-change recommendations of [@Schurch:2016aa].
+<!-- there is no clear guidance provided about the appropriate cutoffs in the literature.
+--> These limits are shown by the dashed grey lines in Figure 2 and we can see that a \(2^{1.4}\) fold change \((\sim 2.6\) fold) cutoff identifies a similar number of transcripts as does  ALDEx2 using \( \gamma = 0.5 \) which identifies `r length(sig.s.ald)` transcripts.
 
-no.scale <- data.frame(yst.all$rab.all, yst.all$diff.win) 
-half.scale <- data.frame(yst.s.all$rab.all, yst.s.all$diff.win) 
-full.scale <- data.frame(yst.1.all$rab.all, yst.1.all$diff.win)
+Supplementary Figures 1 shows an example of the \texttt{aldex.senAnalysis()} function to identify those transcripts that are very sensitive to scale uncertainty in this dataset.
+Here we see that adding a very small amount of scale \(\gamma = 0.1 \) reduces the number of significant transcripts by more than half in the yeast dataset.
+This allows the analyst to ignore those low-dispersion transcripts that were  significant only because of an absence of scale uncertainty.
 
-# we change dispersion a lot for the bulk of the features
-plot(yst.all$rab.all, yst.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3),
-xlab='rAbundance', ylab='Dispersion') 
-lines(lowess(no.scale, f=0.1), col='grey',
-lwd=3) 
-points(yst.s.all$rab.all, yst.s.all$diff.win, pch=19, cex=0.5,
-col=rgb(1,0,0,0.3)) 
-lines(lowess(half.scale, f=0.1), col='pink', lwd=3)
-points(yst.1.all$rab.all, yst.1.all$diff.win, pch=19, cex=0.5,
-col=rgb(0,0,1,0.3)) 
-lines(lowess(full.scale, f=0.1), col='royalblue1', lwd=3)
-abline(h=mn.mid, lty=2) 
-abline(h=mn.mid.s, lty=2, col='red') 
-abline(h=mn.mid.1,
-lty=2, col='blue') 
-title(main='A', line=-1.2, adj=0.9)
+We next examined how adding scale would alter the analysis in a real dataset to which synthetically generated true positive counts had been added.
+We show results from the  anti-PD-1 therapy RNA-seq dataset [@Riaz:2017aa] which examined  changes in gene expression when cells were exposed or not to a cell-cycle checkpoint inhibitor.
+This dataset was used by Li et al.
+[@Li:2022aa] as an example of the dangers of relying on tools with high false positive error rates when analyzing clinical or clinically-related transcriptome samples.
+Indeed, in the benchmarking analysis done by this group, they found that  parameter based methods such as DESeq2 and edgeR that on reported p-value and fold-change cutoffs often led to conclusions in the original dataset that were indistinguishable from permutations of the dataset.
+In other words, that the analysis of transcriptome datasets from patient-derived samples often exhibited many false-positive identifications.
 
-g0 <- expression(paste(gamma, " = 0")) 
-g5 <- expression(paste(gamma, " = 0.5"))
-g1 <- expression(paste(gamma, " = 1")) 
-legend(1,4, legend=c(g0, g5, g1), pch=19, col=c("black", "red", "blue"))
+```{r immuno, echo=F, warning=F, message=F, include=TRUE, fig.dim=c(8,8),  fig.align="center", out.width="0.8\\linewidth",  fig.cap=c('Results of modeling true positive (TP) differences in the PD-1 dataset.
+For this the data were shuffled and  5% of the transcripts were modeled to have TP differences between groups where the differences were drawn from a Normal distribution with a mean difference of 0 and a sd of 2.
+Panel A shows the mean FDR of 10 instances for scale-naive ALDEx2 (A 0), and ALDEx2 with \\(\\gamma=0.2\\) (A 0.2) or \\(\\gamma=0.2\\) (A 0.5), and where significant features were identified by  DESeq2 without (D 0), or with (D 0.5) a 0.5-fold change difference.
+The x-axis shows how the FDR changes for each tool as a function of the estimated modeled difference between groups.
+Panel B shows the mean sensitivity (TP found /all TP).
+Panels C and D show volcano and effect plots for the ALDEx2 output with the transcripts identified as significant at each scale setting as colored points, and non-significant transcripts in black.')}
 
-# plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5,
-# col=rgb(0,0,1,0.3), ylim=c(0,4.5), xlab='g=0 Dispersion', ylab='g=0.5 | g=1
-# Dispersion') points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5,
-# col=rgb(1,0,0,0.3)) abline(0,1, lty=2) title(main='B', line=-1.2, adj=0.5)
-# legend(0,4, legend=c( "0.5", "1"), pch=19, col=c("red", "blue"))
+# this contains the code to run the simulation
+# and to process the output of the simulation
+# the simulation is only run is the output does not
+# exist in the analysis folder
+# look for thin_sim_data.out.Rda
 
-diff.df <- data.frame(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win)
-sig.df <- data.frame(yst.all$rab.all[sig.all], yst.1.all$diff.win[sig.all]-
-yst.all$diff.win[sig.all]) 
-sig.t.df <- data.frame(yst.all$rab.al[sig.t],
-yst.1.all$diff.win[sig.t] - yst.all$diff.win[sig.t]) 
-sig.s.df <-
-data.frame(yst.all$rab.all[sig.s], yst.1.all$diff.win[sig.s] -
-yst.all$diff.win[sig.s]) 
-sig.1.df <- data.frame(yst.all$rab.all[sig.1],
-yst.1.all$diff.win[sig.1]- yst.all$diff.win[sig.1])
+source("code/immuno.R")
 
-ylab1 <- expression(paste("Dispersion: ", gamma, "1 - ", gamma, "0"))
+# this code block is for plotting the data only
 
-plot(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win, xlab='rAbundance',
-ylab=ylab1, pch=19, cex=0.5, xlim=c(-11,10), ylim=c(.1,1.5)) 
-title(main='B',
-line=-1.2, adj=0.9) 
-lines(lowess(diff.df, f=0.1), col='grey60', lwd=3)
-lines(lowess(sig.df, f=0.1), col='red', lwd=3) 
-lines(lowess(sig.t.df, f=0.5),
-col='orange', lwd=3) 
-lines(lowess(sig.s.df, f=0.5), col='royalblue1', lwd=3)
-lines(lowess(sig.1.df, f=0.7), col='cyan', lwd=3)
+par(mfrow=c(2,2))
+plot(means$coef[ald.row], means$FDR[ald.row], ylim=c(0,0.6), type='b',lwd=3,
+  xlab="modeled minimum difference", ylab="mean FDR", col='blue')
+title("A: FDR", adj=0, line= 0.8) 
+#points(means.wi$coef[ald.row], means.wi$FDR[ald.row],  type='l', lwd=3,  col='cyan', lty=3)
+points(means$coef[ald2.row], means$FDR[ald2.row],  type='b', lwd=3, 
+  col='grey')
+#points(means.wi$coef[ald2.row], means.wi$FDR[ald2.row],  type='l', lwd=3, col='cyan', lty=3)
+points(means$coef[ald5.row], means$FDR[ald5.row],  type='b', lwd=3, 
+  col='grey50')
+#points(means.wi$coef[ald5.row], means.wi$FDR[ald5.row],  type='l', lwd=3, col='cyan', lty=3)
+points(means$coef[des.row], means$FDR[des.row],  type='b', lwd=3, col='red')
+points(means$coef[des5.row], means$FDR[des5.row],  type='b', lwd=3,
+  col='darkred')
+legend(0,0.3, legend=c('A 0', 'A 0.2', 'A 0.5'), pch=19,
+  col=c('blue', 'grey', 'grey50'), bty="n", cex=1.2) 
+legend(0.25,0.3, legend=c('D 0', 'D 0.5'), pch=19,
+  col=c('red', 'darkred'), bty="n", cex=1.2) 
+rect(xleft=0, xright=0.55,  ybottom=0.1, ytop=0.3)
 
-d0 <- expression(paste(Delta)) 
-dstar <- expression(paste(Delta,"*")) 
-dstarT <- expression(paste(Delta,"*T")) 
-dg5 <- expression(paste(Delta,"*", gamma, "0.5"))
-dg1 <- expression(paste(Delta,"*", gamma, "1"))
+plot(means$coef[ald.row], means$SEN[ald.row], ylim=c(0,1), lwd=3, type='b',
+  xlab="modeled minimum difference", ylab="mean Sensitivity", col='blue')
+title("B: Sensitivity", adj=0, line= 0.8) 
+#points(means.wi$coef[ald.row], means.wi$SEN[ald.row],  lwd=3, type='l', col='cyan', lty=3)
+points(means$coef[ald2.row], means$SEN[ald2.row],  lwd=3, type='b', col='grey')
+#points(means.wi$coef[ald2.row], means.wi$SEN[ald2.row],  lwd=3, type='l', lty=3, col='cyan')
+#points(means.wi$coef[ald5.row], means.wi$SEN[ald5.row],  lwd=3, type='l', col='cyan', lty=3)
+points(means$coef[ald5.row], means$SEN[ald5.row],  lwd=3, type='b', col='grey50')
+points(means$coef[des.row], means$SEN[des.row],  lwd=3, type='b', col='red')
+points(means$coef[des5.row], means$SEN[des5.row], lwd=3,  type='b', col='darkred')
 
-legend(-12,1.6, legend=c( d0, dstar, dstarT,dg5, dg1), pch="-", col=c("grey",
-"red", "orange", "royalblue1", "cyan"))
+addmin <- min(data.out[[10]]$ald0$we.eBH[data.out[[10]]$ald0$we.eBH > 0])/10
+# row 2677
+aldex.plot(data.out[[10]]$ald0, type='volcano', main="")
+title("C: Volcano", adj=0, line= 0.8) 
+points(data.out[[10]]$ald0$diff.btw[which(data.out[[10]]$ald2$we.eBH < 0.05)], 
+  -log10(data.out[[10]]$ald0$we.eBH[which(data.out[[10]]$ald2$we.eBH < 0.05)]+ addmin), cex=0.6, pch=19, 
+  col='orange')
+points(data.out[[10]]$ald0$diff.btw[which(data.out[[10]]$ald5$we.eBH < 0.05)], 
+  -log10(data.out[[10]]$ald0$we.eBH[which(data.out[[10]]$ald5$we.eBH < 0.05)]+ addmin), col='blue', 
+  cex=0.6, pch=19)
+points(data.out[[10]]$ald0[2677,"diff.btw"], 
+  -log10(data.out[[10]]$ald0[2677,"we.eBH"]+ addmin), col='orange', cex=1.5)
+
+abline(v=c(-0.5,0.5), lty=2, col='darkgrey')
+abline(v=c(-1.5,1.5), lty=2, col='darkgrey')
+
+gam0 <- expression(paste(gamma,"=0"))
+gam2 <- expression(paste(gamma,"=0.2"))
+gam5 <- expression(paste(gamma,"=0.5"))
+legend(2, 60, legend=c("NS", gam0, gam2, gam5), pch=19, col=c("black", "red", "orange", "blue"), cex=1.2)
+
+aldex.plot(data.out[[10]]$ald0, type='MW')
+title("D: Effect", adj=0, line= 0.8) 
+points(data.out[[10]]$ald0$diff.win[which(data.out[[10]]$ald2$we.eBH < 0.05)], 
+  (data.out[[10]]$ald0$diff.btw[which(data.out[[10]]$ald2$we.eBH < 0.05)]), cex=0.6, pch=19, 
+  col='orange')
+points(data.out[[10]]$ald0$diff.win[which(data.out[[10]]$ald5$we.eBH < 0.05)], 
+  (data.out[[10]]$ald0$diff.btw[which(data.out[[10]]$ald5$we.eBH < 0.05)]), col='blue', 
+  cex=0.6, pch=19)
+points(data.out[[10]]$ald0[2677,"diff.win"], 
+  (data.out[[10]]$ald0[2677,"diff.btw"]), col='orange', cex=1.5)
+
+abline(v=c(-0.5,0.5), lty=2, col='darkgrey')
+abline(v=c(-1.5,1.5), lty=2, col='darkgrey')
 
@@
 
-It is common practice to use a dual-cutoff by choosing transcripts based on a
-thresholds for both q-values and fold-changes  [@Cui:2003aa;@Schurch:2016aa].
-Note that there is considerable variation in recommended cutoff
-values[@Schurch:2016aa]. Here, applying a dual-cutoff using a heuristic of at
-least a \(2^{1.4}\) fold change  reduces the number of significant outputs to
-193 for DESeq2 and to 186 for ALDEx2. This cutoff was chosen for convenience and
-is in-line with the recommendations of [@Schurch:2016aa] with the fold change limits
- shown by the dashed grey lines in Figure 1. The  \(2^{1.4}\) fold change
-cutoff identifies a similar number of transcripts as does  ALDEx2 using \(
-\gamma = 0.5 \) which identifies `r length(sig.s.ald)`.  Supplementary Figures 1
-and 2 shows how to use the \texttt{aldex.senAnalysis()} function to identify
-those transcripts that are very sensitive to scale uncertainty. In this
-supplementary figure we see that even adding a very small amount of scale \(
-\gamma = 0.1 \) reduces the number of significant transcripts by more than half.
-This allows us to ignore those low-dispersion transcripts that were  significant only
-because of an absence of scale uncertainty. In practice, we suggest that a
-\texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental
-designs [@Nixon2024B].
+Figure 3 shows the results of ten permutations of this dataset while simulating 5% of the transcripts to be true positives where the difference from no change was derived from a Normal distribution [@Gerard:2020aa].
+This simulation was conducted with the `seqgendiff` R package [@Gerard:2020aa] to both permute the dataset and to add true positive features.
+We did ten permutations and kept track of the number of true and false positive identifications.
+Figure 3A shows the actual false discovery rate for ALDEx2 with and without the addition of scale uncertainty and for DESeq2 with or without a fold-change cutoff of \(2^{0.5}\).
+When the modeled difference was greater than 0, ALDEx2 exhibited a FDR of \(\gamma = 0\): `r round(means$FDR[ald.row[1]],3)`,  \(\gamma = 0.2\): `r round(means$FDR[ald2.row[1]],3)`, and  \(\gamma = 0.5\): `r round(means$FDR[ald5.row[1]],3)`, while DESeq2 with no fold-change cutoff had an FDR of `r round(means$FDR[des.row[1]],2)` and with a 0.5 fold-change cutoff an FDR of `r round(means$FDR[des5.row[1]],2)`.
+Figure 3 also plots these results as a function of the minimum modelled fold change of the true positive transcripts.
+First, we can see that this analysis recapitulates the observations of Li et al [@Li:2022aa] in that DESeq2 has very poor false positive control at a nominal FDR of 0.05.
+The FDR is not controlled any better when a fold-change cutoff is applied, and this agrees with previous work showing that fold-change cutoffs do not materially improve FDR control in high throughput datasets [@Zhang:2009aa;@Ebrahimpoor:2021aa].
+Figure 3B shows that DESeq2 has higher sensitivity than does ALDEx2, and not surprisingly this sensitivity increases as the difference between groups increases.
+For the case where the mean difference is 0 or greater, ALDEx2 exhibited a sensitivity of \(\gamma = 0\): `r round(means$SEN[ald.row[1]],2)`,  \(\gamma = 0.2\): `r round(means$SEN[ald2.row[1]],2)`, and  \(\gamma = 0.5\): `r round(means$SEN[ald5.row[1]],2)`, while DESeq2 with no fold-change cutoff had a sensitivity of `r round(means$SEN[des.row[1]],2)` and with a 0.5 fold-change cutoff a sensitivity of `r round(means$SEN[des5.row[1]],2)`.
+In this example, scale-naive ALDEx2 has near perfect FDR control and reasonable sensitivity at low modeled difference between groups.
+However, when the modeled difference between groups becomes large, then the scale-naive version of ALDEx2 begins to exhibit unacceptable rates of false positives and the false positive rate for DESeq2 also increases.
+Adding in even small amounts of scale uncertainty drops the true FDR rate to 0, but at the expense of sensitivity.
+The major contributor to the increase in FDR with larger modeled differences is that the tools are identifying as positives those transcripts that are modeled to have differences just below the threshold.
+We need to recognize that there is no such thing as a statistical free lunch; the analyst can have high sensitivity but low confidence that any individual transcript is truly different, or have lower sensitivity but have very high confidence that the difference is real.
+In other words, the sensitivity of a method is directly tied to how much error the investigator is willing to tolerate.
 
-The effect on dispersion with increasing  amounts of scale uncertainty are shown
-in Figure 2A, where we can see that the  dispersion increases as uncertainty is
-added. Note that the dispersion in the unscaled analysis in Figure 2A reaches a
-minimum near the mid-point of the distribution, and also does so when the
-analysis is conducted with DESeq2 (Supplementary Figure 3). This shows more
-clearly that dispersion of many transcripts is  almost negligible in the absence
-of scale uncertainty. This plot makes the counter-intuitive suggestion that the
-variance in expression of the majority of genes with moderate expression is more
-predictable  than highly-expressed genes or of housekeeping genes
-[@Rocha:2020aa].  This is at odds with the known biology of cells where single
-cell counting of highly-expressed transcripts shows that they have little
-intrinsic variation [@Taniguchi:2010aa;@Scott:2010].
+Examination of the volcano plot [@Cui:2003aa] and effect plot [@gloor:effect] in Figure 3C and D provides some insight into why adding scale uncertainty provides better FDR control than does the approach of using a p-value and a fold-change cutoff.
+In the volcano plot, adding scale uncertainty differentially excludes transcripts with a combination of marginal p-values and low difference between groups, and that this becomes more pronounced with a larger scale value.
+This effect is also seen in Figure 2 but is more nuanced.
+In contrast, the fold-change cutoff does not incorporate the magnitude of the p-value and so transcripts with large differences, but marginal p-values are retained.
+The effect plot in Figure 3D shows that the transcripts with marginal p values and large differences that are excluded when scale uncertainty is added are those that have  a large dispersion.
 
-Adding scale uncertainty by setting \(\gamma=0.5\) , or \(\gamma = 1.0\),
-increases the minimum dispersion as shown in Figure 2A by the red and blue data
-points, and by the colored lines of fit through the centre of mass of the data.
-Less obvious is that the additional dispersion is  not applied equally to all
-points. Figure 2B shows a plot of the difference between the \(\gamma= 0\) and
-\(\gamma= 1\) data  and here we can see that  scale uncertainty is
-preferentially increasing the dispersion of the mid-expressed transcripts that
-formerly had negligible dispersion; examine the grey line of best fit (overlaid
-by the red line) for the trend. Panel B also shows the trend of the
-expression-dispersion relationship for transcripts that are classed as
-statistically significant. The red line shows the trendline with no added scale
-uncertainty, and this trendline exactly overlays with the grey trendline for the
-bulk of transcripts. The orange trendline indicates those transcripts that are
-both statistically significant and that have a thresholded expression level of
-\( \pm 1.4\), and the dark blue and cyan lines show the statistically
-significant trendline for \(\gamma=0.5\) , or \(1.0\).
+As a concrete example consider the point that is circled in panels C and D with a marginal p-value, with a difference between of nearly 4 and a dispersion of greater than 6.
+This transcript is no longer significant when \(\gamma = 0.2\), but would require a very large fold-change cutoff to be excluded by the standard approach.
+In addition,  transcripts with very small dispersion and very small differences are also excluded when scale uncertainty is added.
+Thus, the addition of scale uncertainty achieves the desired outcome of lowering the FDR for those transcripts that are either marginally differentially abundant, or where the underlying dispersion--and hence the uncertainty in measurement--is very high, or for transcripts that fit both criteria.
 
-Thus, taking Figures 1 and 2 together adding scale uncertainty has the desirable
-effect of changing the distribution of transcripts identified as significantly
-different between groups. Those parts that were statistically significantly
-different *only because of low dispersion* are preferentially excluded from
-statistical significance while those parts that were significantly different
-because of a high difference between groups remain.
+Supplementary Figure 2 shows a second permuted  dataset with the addition of modeled differences between groups.
+Here we used another real dataset with over 200 biological replicates of BRCA1 tumor and control tissue samples from Li et al.
+[@Li:2022aa].
+This Supplementary Figure shows that the FDR control of DESeq2 is somewhat better than in the PD-1 dataset, although still much greater than the the anticipated 5%.
+Further, a fold-change cutoff reduces the FDR of DESeq2 from about 30% to just over 20% with a power of  over 80%.
+However, we can see that scale-naive ALDEx2 performs substantially better with a negligible FDR and comparable power.
+Adding scale uncertainty again improves FDR even for those transcripts modeled to have larger differences and the power is substantially better than in the PD-1 dataset, reaching the same power as DESeq2 or scale-naive ALDEx2 when the modeled difference is large.
+As before, this is driven by removing from consideration those transcripts with either a small difference between or a marginal p-value, or both.
 
-<!-- Interestingly, the yeast transcriptome experiment has an unacknowledged
-difference in scale between conditions; yeast deficient for snf1 are smaller,
-grow more slowly and are sensitive to a variety of common agents that cause cell
-stress [@Yoshikawa:2011aa]. The CLR normalization assumes that the scale of the
-wild-type is exactly 98.4\% of the snf1 strain, and so is \( \theta = 0.016\).
+ Supplementary Figures 3  shows that effect of applying \(\gamma = 0.5 \) to this dataset results in reducing the number of positive transcripts from being \(\sim 70\)\% of the whole dataset to less than 10\% of the dataset and that this is largely because of a reduction in significance of those transcripts with low dispersion.
+Supplementary Figures 4 shows a sensitivity analysis of the BRCA1 dataset showing that different scale uncertainty amounts alter the number of significant transcripts in a biological replicate experiment similarly to a technical replicate experiment.
+Supplementary Figures 5 and 6 delve into how adding scale uncertainty affects the variace-abundance relationship in subtle ways, and may help readers to understand the observations seen in Figure 3 and supplementary Figure 2.
+
+Together the results in this section show that adding scale uncertainty has the desirable effect of altering the transcripts identified as significantly different between groups in a way that exhibits better control of FDR albeit with a corresponding reduction in sensitivity.
+Those parts that were statistically significantly different *only because of low dispersion* or that *had marginal p-values* or both, are now preferentially excluded from statistical significance.
+In practice, we suggest that a \texttt{gamma} parameter of between 0.2 and 0.5 is realistic for most experimental designs [@Nixon2024B] regardless if the replication is technical or biological.
+
+<!-- Interestingly, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions; yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].
+The CLR normalization assumes that the scale of the wild-type is exactly 98.4\% of the snf1 strain, and so is \( \theta = 0.016\).
 Given the per-generation time difference of about 15\%, this seems implausible.
-Setting \(\gamma=0.5\) can be interpreted as acknowledging uncertainty in the
-difference in the underlaying scale between groups. Nixon et al. -@Nixon2024B
-showed that the 95\% bounds on the scale uncertainty for a given \(\gamma\)
-value, say 0.5, can be calculated as \(2^{-(2 \gamma + \theta)}, 2^{(2 \gamma +
-\theta)}\), or \(2^{-(2*0.5 + 0.016)}, 2^{(2*0.5 + 0.016)} = 0.494,2.022\).
-Thus, by adding scale uncertainty, we are acknowledging that the data are
-uncertain and that any results are robust to that uncertainty, within the bounds
-of our model. -->
+Setting \(\gamma=0.5\) can be interpreted as acknowledging uncertainty in the difference in the underlaying scale between groups.
+Nixon et al.
+-@Nixon2024B showed that the 95\% bounds on the scale uncertainty for a given \(\gamma\) value, say 0.5, can be calculated as \(2^{-(2 \gamma + \theta)}, 2^{(2 \gamma + \theta)}\), or \(2^{-(2*0.5 + 0.016)}, 2^{(2*0.5 + 0.016)} = 0.494,2.022\).
+Thus, by adding scale uncertainty, we are acknowledging that the data are uncertain and that any results are robust to that uncertainty, within the bounds of our model.
+-->
 
@@
 
-Dos Santos et al. [@dosSantos:2024]  used a vaginal metatranscriptome dataset  to
-compare the gene expression in bacteria collected from healthy (H) and bacterial
-vaginosis (BV) affected women. In this environment, both the relative abundance
-of species between groups and the gene expression level within a species is
-different [@macklaim:2013]. Additionally, prior research suggests that the total
-number of bacteria is about 10 times more in the BV than in the H condition
-[@Zozaya:2010]. Thus, these are extremely challenging datasets in which to
-determine differential abundance as there are both compositional and scale
-changes between conditions. The usual method to analyze vaginal
-metratranscriptome data is to do so on an organism-by-organism basis
-[@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding
-of the environment is less pronounced. One attempt at system-wide analysis
-returned several housekeeping functions as differentially expressed between
-groups [@Denge00262-18]; a result likely due to a disconnect between the
-assumptions of the normalization used and the actual scale of the environment
-[@Wu2021].
+Dos Santos et al.
+[@dosSantos:2024]  used a vaginal metatranscriptome dataset  to compare the gene expression in bacteria collected from healthy (H) and bacterial vaginosis (BV) affected women.
+This dataset is derived from two publicly available datasets composed of a set of 20 non-pregnant women from London, Ontario Canada [@Macklaim:2018aa] and a subset of 22 non-pregnant women collected from German women who underwent metronidazole treatment for BV [@Denge00262-18].
+Batch effects for these two groups were removed with ComBat-seq [@Zhang:2020ab] and the two datasets were merged into one, giving a total of 16 H and 26 BV samples.
+In the Dos Santos paper, all results from this initial analysis were replicated in a much larger dataset derived from the MOMS-PI study [@Fettweis:2019aa]
 
-In this example, we show how to specify and interpret an informed scale model
-that can explicitly account for some of these modeling difficulties
-[@Nixon2024B] even in a difficult dataset. An informed scale model can control
-for both the mean difference of scale between groups (e.g., directly incorporate
-information on the differences in total number of bacteria between the BV and H
-conditions) as well as the uncertainty of that difference. To specify a
-user-defined scale model, we can pass a matrix of scale values instead of an
-estimate of just the scale uncertainty  to \texttt{aldex.clr()}. This matrix
-should have the same number of rows as the of Monte-Carlo Dirichlet samples, and
-the same number of columns as the number of samples. While this matrix can be
-computed from scratch by the analyst, there is an
-\texttt{aldex.makeScaleModel()} function that can be used to simplify this step
-in most cases. This encodes the scale model as \(\Lambda \sim N(log_2 \mu_n,
-\gamma^{2})\), where \(\mu_n\) represents the scale value for each sample or
-group and gamma is the uncertainty as before. The scale estimate can be a
-measured value (cell count, nucleic acid input, etc) or an estimate.  Nixon et
-al. [@nixon2024scale;@Nixon2024B] showed that only the ratio of the means are important when
-providing values for \( \mu_n\); i.e., the ratio between the \( \log_2 \mu_i\)
-and \(\log_2 \mu_j\) values. See the supplement to  Nixon et al. [@Nixon2024B]
-for more information.
-
-Figure 3A shows an effect plot of the data where reads are grouped by homologous
-function regardless of the organism of origin. Each point represents one of 
-3728 KEGG  functions [@Okuda:2008]. There are many more functions represented in
-the BV group (bottom) than in the healthy group (top). This is because the
-\textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced
-genome content relative to the anaerobic organisms that dominate in BV, because
-there is a greater diversity of organisms in BV than in H samples, and because
-the BV condition has about  an order of magnitude more bacteria than does the
-H condition.
-
-The naive scale model appears to be reflecting the bacterial load as observed by calculating the mean scale value for each group.
-Using a negligible scale value; i.e., \(\gamma=1e-3\) exposes the naive scale estimate for
-samples  in the \texttt{@scaleSamps} slot from the
-\texttt{aldex.clr} output. the naive scale estimate for the
-`r load('analysis/xg.clr.Rda')`
-healthy group is `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])),2)` and
-for the BV group is  `r round(mean(rowMeans(xg@scaleSamps[c(9:36),])),2)` for a
-difference of `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),])),2)`. This is interpreted as the scale of
-the H group of samples being 
-`r round(2^(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),]))),2)` 
-greater than the BV group. This
-precise but incorrect estimate places the location of the housekeeping genes off
-the midline of no difference.
-
-Applying the default scale model of \(\gamma=0.5\) increases the dispersion
-slightly but does not move the   housekeeping functions toward the midline. This
-is as expected; the mean of the default scale model is based on the CLR
-normalization so no shift in location would be expected over the original ALDEx2
-model. Nevertheless, about 30\% of the housekeeping functions are no longer
-statistically significantly different. Note that this change is simple to
-conduct, has no additional computational complexity and requires only a slight
-modification for the analyst.
+In this vaginal environment, both the relative abundance of species between groups and the gene expression level within a species is different [@macklaim:2013].
+Additionally, prior research suggests that the total number of bacteria is about 10 times more in the BV than in the H condition [@Zozaya:2010].
+Thus, these are extremely challenging datasets in which to determine differential abundance as there are both compositional and scale changes between conditions.
+The usual method to analyze vaginal metratranscriptome data is to do so on an organism-by-organism basis [@macklaim:2013; @Denge00262-18; @Fettweis:2019aa] because the scale confounding of the environment is less pronounced.
+One attempt at system-wide analysis returned several housekeeping functions as differentially expressed between groups [@Denge00262-18]; a result likely due to a disconnect between the assumptions of the normalization used and the actual scale of the environment [@Wu2021].
 
@@
 # list of ribosomal associated functions in the datset TIL about 
-# complete.cases()
-ribo.v <-c("K02863","K02864","K02867","K02871","K02874","K02876","K02878","K02879","
-K02881","K02884","K02886","K02887","K02888","K02890","K02892","K02895","K02897",
-"K02899","K02902","K02904","K02906","K02907","K02909","K02911","K02913","K02914"
-,"K02916","K02926","K02931","K02933","K02935","K02939","K02945","K02946","K02948
-","K02950","K02952","K02954","K02956","K02959","K02961","K02963","K02965","
-K02967","K02968","K02970","K02982","K02986","K02988","K02990","K02992","K02994",
-"K02996") 
+# complete.cases() ribo.v <-c("K02863","K02864","K02867","K02871","K02874","K02876","K02878","K02879"," K02881","K02884","K02886","K02887","K02888","K02890","K02892","K02895","K02897", "K02899","K02902","K02904","K02906","K02907","K02909","K02911","K02913","K02914" ,"K02916","K02926","K02931","K02933","K02935","K02939","K02945","K02946","K02948 ","K02950","K02952","K02954","K02956","K02959","K02961","K02963","K02965"," K02967","K02968","K02970","K02982","K02986","K02988","K02990","K02992","K02994", "K02996")
+
 ```
-
-There are 101  functions with low dispersion that appear to be shared by both
-groups (boxed area in Figure 3A, and colored in cyan). Inspection shows that
-these   largely correspond to core metabolic functions such as transcription,
-translation, ribosomal functions, glycolysis, replication, chaperones, etc
-(Supplementary file housekeeping.txt).  The transcripts of many of these are
-commonly used as invariant reference sequences in wet lab experiments
-[@Rocha:2020aa] and so are not be expected   to contribute to differences in
-ecosystem behaviour. The average location of these should be centred on 0
-difference to represent an internal reference set.  However, without an informed
-scale model, the mean  of these housekeeping functions is  approximately located
-at +2.3.
 
@@
 
+l10q <- expression(-log[10](q)) 
+l2d <- expression(log[2]~ Difference)
+
 load('analysis/xt.Rda') 
@@
 
-hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 &
-xt.all$diff.btw < 3] 
-hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 &
-xt.all$diff.btw < 3
+hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3] 
+hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 1 & xt.all$diff.btw < 3
 
 fn <- rownames(xt.m.all)[xt.m.all$we.eBH < 0.01 & xt.m.all$diff.btw < 1 & xt.all$we.eBH > 0.01] 
-fp <- rownames(xt.all)[xt.all$we.eBH < 0.01 &
-xt.m.all$we.eBH > 0.01 & !xt.m.all$we.eBH < 0.01]
+fp <- rownames(xt.all)[xt.all$we.eBH < 0.01 & xt.m.all$we.eBH > 0.01 & !xt.m.all$we.eBH < 0.01]
 
@@
 
-points(xt.all$diff.win[hk.off], xt.all$diff.btw[hk.off], pch=19, cex=0.4,
-col=rgb(0,1,1,0.5)) 
-points(xt.all[fn,'diff.win'], xt.all[fn,'diff.btw'],
-col='orange', cex=0.4, pch=19) 
+points(xt.all$diff.win[hk.off], xt.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5)) 
+points(xt.all[fn,'diff.win'], xt.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19) 
 rect(0.5,1,2.5,3, col=rgb(0,0,0,0.1), lty=2, lwd=3) 
 title('A: ALDEx2 unscaled', adj=0, line= 0.8) 
-legend(0,12, legend=c("HK",
-"TP", "FP", "FN"), pch=19, cex=0.5, col=c("cyan", "red", "orange", "blue"))
+legend(0,12, legend=c("HK", "TP", "FP", "FN"), pch=19, cex=0.5, col=c("cyan", "red", "orange", "blue"))
 
 aldex.plot(xg.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion", ylab=l2d) 
-points(xg.all[fp,'diff.win'], xg.all[fp,'diff.btw'], col='blue', cex=0.4,
-pch=19) 
-points(xg.all$diff.win[hk.off], xg.all$diff.btw[hk.off], pch=19,
-cex=0.4, col=rgb(0,1,1,0.5)) 
-points(xg.all[fn,'diff.win'],
-xg.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19) 
-title('B: ALDEx2 gamma',
-adj=0, line= 0.8)
+points(xg.all[fp,'diff.win'], xg.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19) 
+points(xg.all$diff.win[hk.off], xg.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5)) 
+points(xg.all[fn,'diff.win'], xg.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19) 
+title('B: ALDEx2 gamma', adj=0, line= 0.8)
 
-aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion",
-ylab=l2d) 
-points(xt.m.all[fp,'diff.win'], xt.m.all[fp,'diff.btw'], col='blue',
-cex=0.4, pch=19) 
-points(xt.m.all$diff.win[hk.off], xt.m.all$diff.btw[hk.off],
-pch=19, cex=0.4, col=rgb(0,1,1,0.5)) 
-points(xt.m.all[fn,'diff.win'],
-xt.m.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19) 
-title('C: ALDEx2 both',
-adj=0, line= 0.8)
+aldex.plot(xt.m.all, xlim=c(0.3,9), cutoff.pval=0.01, xlab="Dispersion", ylab=l2d) 
+points(xt.m.all[fp,'diff.win'], xt.m.all[fp,'diff.btw'], col='blue', cex=0.4, pch=19) 
+points(xt.m.all$diff.win[hk.off], xt.m.all$diff.btw[hk.off], pch=19, cex=0.4, col=rgb(0,1,1,0.5)) 
+points(xt.m.all[fn,'diff.win'], xt.m.all[fn,'diff.btw'], col='orange', cex=0.4, pch=19) 
+title('C: ALDEx2 both', adj=0, line= 0.8)
 
@@
 
-We desire a scale model that approximately centres the housekeeping functions, because we expect housekeeping functions tp be nearly invariant; thus an
-appropriate scale in this dataset for functional analysis is likely closer to 0 than the naive estimate.
-One way to choose an appropriate value for \(\mu_n\) is to use the
-\texttt{aldex.clr} function on only the presumed invariant functions setting
-\(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot as before.
-Doing so suggests that the difference in scale should be about 14%.   A second
-approach would be to identify the functions used as the denominator with the
-\texttt{denom="lvha"} option [@Wu2021] for the \texttt{aldex.clr} function, and
-then to use these values as before. This approach suggests a 5% difference in
-scale, and is potentially less subject to user interpretation.
+In this example, we show how to specify and interpret a user defined or *informed* scale model that can explicitly account for some of these modeling difficulties [@Nixon2024B] even in a difficult to analyze dataset.
+An informed scale model can control for both the mean difference of scale between groups (e.g., directly incorporate information on the differences in total number of bacteria between the BV and H conditions) as well as the uncertainty of that difference as illustrated in Figure 1:3.
+To specify a user-defined scale model, we can pass a matrix of scale values instead of an estimate of just the scale uncertainty  to \texttt{aldex.clr()}.
+This matrix should have the same number of rows as the of Monte-Carlo Dirichlet samples, and the same number of columns as the number of samples.
+While this matrix can be computed from scratch by the analyst, there is an \texttt{aldex.makeScaleModel()} function that can be used to simplify this step in most cases.
+This encodes the scale model as \(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents the scale value for each sample or group and \(gamma\) is the uncertainty as before.
+The scale estimate can be a measured value (cell count, nucleic acid input, spike-ins, etc) or an estimate.
+Nixon et al.
+[@nixon2024scale;@Nixon2024B] showed that only the ratio of the means are important when providing values for \( \mu_n\); i.e., the ratio between the \( \log_2 \mu_i\) and \(\log_2 \mu_j\) values.
+See the supplement to  Nixon et al.
+[@Nixon2024B] for more information.
+
+Figure 4A shows an effect plot of the data where reads are grouped by homologous function regardless of the organism of origin.
+Each point represents one of 3728 KEGG  functions [@Okuda:2008].
+There are many more functions represented in the BV group (bottom) than in the healthy group (top).
+This is because the \textit{Lactobacilli} that dominate a healthy vaginal microbiome have reduced genome content relative to the anaerobic organisms that dominate in BV, because there is a greater diversity of organisms in BV than in H samples, and because the BV condition has about  an order of magnitude more bacteria than does the H condition.
+
+The naive scale model appears to be reflecting the bacterial load as observed by calculating the mean scale value for each group.
+Using a negligible scale value; i.e., \(\gamma=1e-3\) exposes the naive scale estimate for samples  in the \texttt{@scaleSamps} slot from the \texttt{aldex.clr} output.
+The naive scale estimate for the `r load('analysis/xg.clr.Rda')` healthy group is `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])),2)` and for the BV group is  `r round(mean(rowMeans(xg@scaleSamps[c(9:36),])),2)` for a difference of `r round(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),])),2)`.
+This is interpreted as the scale of the H group of samples being `r round(2^(mean(rowMeans(xg@scaleSamps[c(1:8,37:44),])) - mean(rowMeans(xg@scaleSamps[c(9:36),]))),2)` greater than the BV group.
+
+Applying the default scale model of \(\gamma=0.5\) increases the dispersion slightly but does not move the   housekeeping functions toward the midline.
+This is as expected; the mean of the default scale model is based on the CLR normalization so no shift in location would be expected over the scale-naive ALDEx2 model.
+Nevertheless, about 30\% of the housekeeping functions are no longer statistically significantly different.
+Note that this change is simple to conduct, has no additional computational complexity and requires only a slight modification for the analyst.
+
+There are 101  functions with low dispersion that appear to be shared by both groups (boxed area in Figure 3A, and colored in cyan).
+Inspection shows that these   largely correspond to core metabolic functions such as transcription, translation, ribosomal functions, glycolysis, replication, chaperones, etc (Supplementary file housekeeping.txt).
+The transcripts of many of these are commonly used as invariant reference sequences in wet lab experiments [@Rocha:2020aa] and so are not be expected   to contribute to differences in ecosystem behaviour.
+Because we expect housekeeping functions to be nearly invariant in their expression and to occur in all organisms, the average location of these should be centred on 0 difference to represent an internal reference set.
+However, with the naive scale model, the mean  of these housekeeping functions is  approximately located at 2.3.
+Thus, we  desire a scale model that approximately centres the housekeeping functions and an appropriate scale in this dataset for functional analysis  will place these functions closer to 0 than does the naive estimate.
+One way to choose an appropriate value for \(\mu_n\) is to use the \texttt{aldex.clr} function on only the presumed invariant functions setting \(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot as before.
+Doing so suggests that the difference in scale should be about 14%.
+A second approach would be to identify the functions used as the denominator with the \texttt{denom="lvha"} option [@Wu2021] for the \texttt{aldex.clr} function, and then to use these values as before.
+This approach suggests a 5% difference in scale, and is potentially less subject to user interpretation.
 
@@
 
-For the purposes of this example, if we assume a 14% difference in scale, we can
-set \(\mu_i = 1\) and \(\mu_j = 1.14\) using the \texttt{makeScaleMatrix}
-function. This function uses a logNormal distribution to build a scale matrix
-given a user-specified mean difference between groups and uncertainty level.
-Applying a per-group relative differential scale of 0.14 moves the housekeeping
-functions close to the midline of no difference (Figure 3C, assuming 14% mean difference = 
-`r round(mean(xt.m.all[hk.off,"diff.btw"]),2)`, assuming a 5% mean difference = 
-`r round(mean(xt.lv.all[hk.off,"diff.btw"]),2)`), and applying a gamma of 0.5
-provides the same dispersion as in panel B of Figure 3. Note that now a
-significant number of functions are differentially up in BV that were formerly
-classed as not different without the full scale model (orange), or when only a
-default scale was applied. Inspection of the functions shows that these are
-largely missing from the \emph{Lactobacillus} species and so should actually be
-captured as differentially abundant in the BV group. Supplementary Figure 4
-shows that the using either the 5% or the 14%  scale difference  give
-imperceptibly different results suggesting that an informed scale model does not
-have to precisely estimate the scale difference to be useful. Nixon et al,
-@Nixon2024B also found that multiple reasonable estimates for the \(\mu_n\) part
-of the informed scale model were similarly useful in microbiome data.
+For the purposes of this example, if we assume a 14% difference in scale, we can set \(\mu_i = 1\) and \(\mu_j = 1.14\) using the \texttt{makeScaleMatrix} function.
+This function uses a logNormal distribution to build a scale matrix given a user-specified mean difference between groups and uncertainty level.
+Applying a per-group relative differential scale of 0.14 moves the housekeeping functions close to the midline of no difference (Figure 3C, assuming 14% mean difference = `r round(mean(xt.m.all[hk.off,"diff.btw"]),2)`, assuming a 5% mean difference = `r round(mean(xt.lv.all[hk.off,"diff.btw"]),2)`), and applying a gamma of 0.5 provides the same dispersion as in panel B of Figure 3.
+Note that now a significant number of functions are differentially up in BV that were formerly classed as not different without the full scale model (orange), or when only a default scale was applied.
+Inspection of the functions shows that these are largely missing from the \emph{Lactobacillus} species and so should actually be captured as differentially abundant in the BV group.
+Supplementary Figure 7 shows that the using either the 5% or the 14%  scale difference  give imperceptibly different results suggesting that an informed scale model does not have to precisely estimate the scale difference to be useful.
+Nixon et al, @Nixon2024B also found that multiple reasonable estimates for the \(\mu_n\) part of the informed scale model were similarly useful in microbiome data.
 
-Thus, applying an informed scale allows us to distinguish between both false
-positives (housekeeping functions in cyan, and others in blue) and false
-negatives (orange functions) even in a very difficult to analyze dataset. The
-remarkable improvements in biological interpretation afforded by an informed
-scale model, and the transferrability of it between sample cohorts of the same
-condition is outlined elsewhere [@dosSantos:2024]. We suggest that the default
-scale model  is sufficient when the data are approximately  centred. However,
-an informed model is more appropriate with datasets are not well centred or
-when the investigator has prior information about the underlying biology.
+Thus, applying an informed scale allows us to distinguish between both false positives (housekeeping functions in cyan, and others in blue) and false negatives (orange functions) even in a very difficult to analyze dataset.
+We used this scale model to uncover hiter-to-now unknown differences in microbiome functional activity between the Healthy and BV cohorts that were missed in previos analyses and that explain important clinical differences between them [@dosSantos:2024].
+The remarkable improvements in biological interpretation afforded by an informed scale model, and the transferrability of it between sample cohorts of the same condition is outlined in dos Santos et al.
+[@dosSantos:2024].
+We suggest that the default scale model  is sufficient when the data are approximately  centred but that an informed model is more appropriate with datasets are not well centred or when the investigator has prior information about the underlying biology.
 
@@
 
-Biological systems are both predictably variable and stochastic
-[@Taniguchi:2010aa] and systems biology experiments show that there are
-transcripts with approximately constant concentrations in the cell and those
-with large variability under different growth conditions [@Scott:2010].  Current
-measurement methods that rely on high throughput sequencing fail to capture all
-of the variation, particularly variation due to scale
-[@nixon2024scale;@Nixon2024B].  In the absence of external information
-[@Loven:2012aa;@Vandeputte:2017aa;@yeast-absolute] sequencing depth
-normalisation methods cannot recapture the scale information
-[@Loven:2012aa;@Lovell:2015], and can only normalize for the technical variation
-due to sequencing depth. Here we demonstrated that even approximate estimates of
-the true system scale and the uncertainty of measuring it can aid in the interpretation of
-RNA-sequencing experiments.
+Scale estimates affect two parts of the analysis.
+Modeling uncertainty in scale prevents false certainty in the precision of estimation and controls false positive identification.
+Modeling between-group scale  relaxes the assumption of identity between the sizes of the environments and allows better control of false  negative identification.
+The scale estimates can be derived from the total number of molecules in the environment or from other estimates  of input size (cell counts, initial concentrations, spike-ins, growth rates, etc).
 
-Nixon et al. [@nixon2024scale] introduced the idea of explicitly modeling the
-scale of a HTS dataset, and showed how to incorporate these models in the
-analysis of microbiome and other datasets [@Nixon2024B]. They  demonstrated that
-many tools commonly used to analyze HTS datasets had substantial Type 1 and Type
-2 error rates, in line with recent findings by others
-[@Quinn:2018aa;@Ge:2021aa;@Li:2022aa]. A version of ALDEx2 with the ability to
-include scale uncertainty was shown to be able to correct for the high Type 1
-error rate for that tool, albeit with some loss of sensitivity. Finally, they
-showed that incorporating an informed scale model incorporating both location
-and scale uncertainty estimates could both control for Type 1 and Type 2 error
-rates [@Nixon2024B].
+Biological systems are both predictably variable and stochastic [@Taniguchi:2010aa] and systems biology experiments show that there are transcripts with approximately constant concentrations in the cell and those with large variability under different growth conditions [@Scott:2010].
+Current measurement methods that rely on high throughput sequencing fail to capture all of the variation, particularly variation due to scale [@nixon2024scale;@Nixon2024B].
+In the absence of external information [@Loven:2012aa;@Vandeputte:2017aa;@yeast-absolute] sequencing depth normalisation methods cannot recapture the scale information [@Loven:2012aa;@Lovell:2015], and can only normalize for the technical variation due to sequencing depth.
+Here we demonstrated that even approximate estimates of the true system scale and the uncertainty of measuring it can aid in the interpretation of RNA-sequencing experiments.
 
-Building and using a scale model thus has substantial benefits relative to the
-dual cutoff approach that is advocated for many gene expression experiments
-[@Cui:2003aa;@Schurch:2016aa]. In particular, the dual cutoff approach has long
-been known to not control for Type 1 errors [@Zhang:2009aa;@Ebrahimpoor:2021aa],
-and the frequent lack of concordance between tools when benchmarked on
-transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Ge:2021aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa;
-@Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa] suggests poor
-control of Type 2 errors as well [@Quinn:2018aa;@Li:2022aa]. Thus, incorporating
-a scale model during the analysis of HTS data promises the best of both worlds.
-A default scale model can control for Type 1 errors with minimal prior knowledge
-of the environment and this can be done with essentially no additional
-computational overhead. Furthermore, this work and previous [@Nixon2024B] show
-that even minimal information about the underlying environment can be used to
-build a relatively robust informed scale model that controls for both Type 1 and
-2 error rates.
+Nixon et al.
+[@nixon2024scale] introduced the idea of explicitly modeling the scale of a HTS dataset, and showed how to incorporate these models in the analysis of microbiome and other datasets [@Nixon2024B].
+They  demonstrated that many tools commonly used to analyze HTS datasets had substantial Type 1 and Type 2 error rates in line with recent findings by others [@Quinn:2018aa;@Ge:2021aa;@Li:2022aa].
+A version of ALDEx2 with the ability to include scale uncertainty was shown to be able to correct for  high Type 1 error rate for that tool, albeit with some loss of sensitivity.
+Finally, they showed that incorporating an informed scale model incorporating both location and scale uncertainty estimates could both control for Type 1 and Type 2 error rates [@Nixon2024B;@Dos-Santos:2024aa].
 
-In the analysis of HTS data it is often observed that larger datasets converge
-on the majority of parts being significantly different
-[@Schurch:2016aa;@Li:2022aa;@nixon2024scale]. Li et al. [@Li:2022aa] conducted a
-permutation-based benchmarking study and found that widely used tools performed
-worse than simple Wilcoxon rank-sum tests in controlling the FDR when sample
-sizes became large. They suggested that the presence of outliers were one of the
-factors driving this observation. Brooks et al. [@Brooks:2024aa] suggested that
-inappropriate choice of benchmarking methods are also a major contributing
-factor and that objective standards of truth are important. From the perspective
-of our work the disagreement between tools can be explained by the observation
-that different analytic approaches  produce  different parameter estimates for
-location or scale or for both. Thus, more data produces worse estimates because the
-additional data simply increases the precision of a flawed estimate
-[@gustafson2015bayesian;@nixon2024scale].
+The process of choosing the parameters are experiment-specific and can be anchored in known information such as cell counts, spike-ins, information from the literature  or similar [@McGovern:2023;@Nixon2024B;@nixon2024scale].
+In the metatranscriptome example used in this report [@Dos-Santos:2024aa;] the choice of parameter was driven by the assumption inherent in the biology that core housekeeping functions would serve as an appropriate standard [@Vandesompele:2002aa].
+The choice of parameters must be guided by the experimental question  and other approaches in the literature which suggest normalizing the transcript levels to the bacterial metagenomic levels [@Zhang:2021ab] could be used to set the scale parameters, but in this case are  more granular at the individual organism level rather than at the systemic functional level.
 
-Scale simulation  is now built into ALDEx2 [@Nixon2024B] and here we suggest
-that  there are two main root causes to common HTS data pathologies. The first
-contributing factor is the observed very low dispersion estimate for many
-features that is a by-product of some experimental designs and of normalization.
-In the Schurch et al. [@Schurch:2016aa dataset], the data were  from single
-colonies derived from a single culture. Thus, it is more accurate to describe
-the 96 samples as wet-lab technical replicates rather than independent samples.
-This type of replication approach is standard in the molecular literature, and
-would be expected to result in the very low dispersion that is observed.
-Applying the default scale model with \( \gamma=0.5 \) a large number of
-transcripts have  their dispersion increased (Figure 1D), with the effect being
-largest for those with the lowest initial dispersion (Figure 2). Adding scale
-uncertainty results in modest number of transcripts, `r length(sig.s.ald)`,
-being called significantly different as shown in the volcano plot in Figure 1B
-(red points). In addition, there is now a strong concordance between the
-difference and q-values. In hindsight, it is not obvious  why the unscaled
-volcano plot shows such poor correspondence. We suggest that this is explained
-by random fluctuations of the many very low variance estimates and this is
-supported by the  plots shown in Figure 2.
+Building and using a scale model thus has substantial benefits relative to the dual cutoff approach that is advocated for many gene expression experiments [@Cui:2003aa;@Schurch:2016aa].
+In particular, the dual cutoff approach has long been known to not control for Type 1 errors [@Zhang:2009aa;@Ebrahimpoor:2021aa], and the frequent lack of concordance between tools when benchmarked on transcriptomes [@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Ge:2021aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa] suggests poor control of Type 2 errors as well [@Quinn:2018aa;@Li:2022aa].
+Thus, incorporating a scale model during the analysis of HTS data promises the best of both worlds.
+A default scale model can control for Type 1 errors with minimal prior knowledge of the environment and this can be done with essentially no additional computational overhead.
+Furthermore, this work and previous [@Nixon2024B;@Dos-Santos:2024aa] show that even minimal information about the underlying environment can be used to build a relatively robust informed scale model that controls for both Type 1 and 2 error rates.
+It is important to note that the approach advocated here is distinct from that suggested by Zhang et al.
+[@Zhang:2021aa;@Zhang:2021ab] where the DNA amount for a gene is a covariate in the model for transcriptomic differential abundance.
+In our analysis we grouped all the transcript information to functional level regardless of organism, instead of modeling per-organism gene abundances.
+In the future we anticipate being able to build more complex models similar to those used by Zhang et al.
+with the additional information of uncertainty in the underlying gene count.
 
-The second contributing factor is unacknowledged asymmetry in many datasets
-[@Wu2021]; i.e., different gene content or a directional change in the majority
-of features. In the case of asymmetry, the use of a user-specified scale model
-can be very useful for otherwise difficult-to-analyze datasets such as
-meta-transcriptomes and in-vitro selection datasets where the majority of
-features can change. We showed one such example for the metatranscriptome
-dataset in Figure 3. Here the dataset was highly asymmetrical. Incorporating
-differential scale on a per-group basis moves the mass of the housekeeping
-functions towards the midline of no difference and so affects both Type I and
-Type II error rates. We showed two ways of estimating the  scale difference
-between groups and found that any reasonable estimate is an improvement over the
-naive approach and also over the default scale model. This is in line with the
-observations by Nixon et al [@Nixon2024B] in a 16S rRNA gene sequencing dataset.
-It is also of note that in the case of true biological replicates (different
-individuals) that adding a modest amount of scale \(\gamma=0.5\) had little
-effect on the the difference between groups and on the dispersion. Thus, in this
-dataset the scale mis-specification was affecting mainly the location of the
-difference between groups.  While we acknowledge that some prior information on
-which housekeeping transcripts should not be classed as differentially abundant
-is needed, we suggest that this information is widely available and is already
-used when performing the gold-standard quantitative PCR test of differential
-abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
+In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@Li:2022aa;@nixon2024scale].
+Li et al.
+[@Li:2022aa] conducted a permutation-based benchmarking study and found that widely used tools performed worse than simple Wilcoxon rank-sum tests coupled with the TPM normalization in controlling the FDR when sample sizes became large.
+Li et al.
+suggested that the presence of outliers were one of the factors driving the extreme FDR in some tests.
+We found that when the Wilcoxon test was used within the ALDEx2 framework that it had essentially the same outputs as did the t-test.
+For example, in the PD-1 dataset where \(\gamma = 0\)  the ALDEx2 t-test exhibited a mean FDR of `r round(means$FDR[ald.row[1]],3)*100`% and mean sensitivity of  `r round(means$SEN[ald.row[1]],3)*100`% while the corresponding values from the  ALDEx2 Wilcoxon test  were `r round(means.wi$FDR[ald.row[1]],3)*100`% and `r round(means.wi$SEN[ald.row[1]],3)*100`%.
+This result again supports that the assumptions of the normalizations are as important or more important than the statistical test.
+Brooks et al.
+[@Brooks:2024aa] suggested that inappropriate choice of benchmarking methods are also a major contributing factor and that better objective standards of truth are needed.
+In this report we generated semi-synthetic test data used binomial thinning which produces data that more closely mimic the properties of real high throughput sequencing data, and so can more rigorously test different tools [@Gerard:2020aa].
+From the perspective of our work the disagreement between tools can be explained by the observation that different analytic approaches  produce  different parameter estimates for either location or scale, or for both, as suggested in Figure 1.
+Thus, more data produces worse estimates because the additional data simply increases the precision of a flawed estimate [@gustafson2015bayesian;@nixon2024scale].
 
-Beyond concerns of fidelity and rigor, scale models also enhance the
-reproducibility and transparency of HTS analyses. The addition of scale
-uncertainty essentially tests the model over a range of normalizations
-[@nixon2024scale] and so can replace the consensus approach that has been
-proposed by some groups [@Nearing:2022aa;@Song:2023aa] with no additional
-computational overhead. Thus, an advantage of incorporating scale is that
-analyses can be made much more robust such that actual or potential differences
-in scale can be tested and accounted for explicitly. While it is beyond the
-scope of the present article, we note that there are many ways of building scale
-models that enhance the interpretability of the parameters and assumptions and a
-detailed description of these points is describe elsewhere [@nixon2024scale,].
+Scale simulation  is now built into ALDEx2 [@Nixon2024B] and in this report we suggest that  there are two main root causes to common HTS data pathologies.
+The first contributing factor is the observed very low dispersion estimate for many features that is a by-product of some experimental designs and  normalizations (Figure 2).
+Supplying additional uncertainty alleviates many FP but in a way that more appropriately controls the FDR as shown in Figure 3.
+The second contributing factor is unacknowledged asymmetry in many datasets [@Wu2021]; i.e., different gene content or a directional change in the majority of features.
+In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult-to-analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change as shown in Figure 4.
+We showed two ways of estimating the  scale difference between groups and found that any reasonable estimate is an improvement over the naive approach and also over the default scale model.
+This is in line with the observations by Nixon et al [@Nixon2024B] in a 16S rRNA gene sequencing dataset.
+While we acknowledge that some prior information is needed that this information is widely available and is already used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa;@SEQC/MAQC-III-Consortium:2014aa].
 
-In summary, we supply a toolkit that makes incorporating scale uncertainty and
-location information simple to incorporate for transcriptomes or indeed any type
-of HTS dataset. While the underlying scale of the system is generally
-inaccessible, the effect of scale on the analysis outcomes can be modelled and
-can help explain some of the underlying biology, and help to expose known issues
-with the analysis of HTS data. Adding scale information to the analysis allows
-for more robust inference because the features that are sensitive to scale can
-be identified and their impact on conclusions weighted accordingly.
-Additionally, the use of informed  scale models permits difficult to analyze
-datasets to be examined in a robust and principled manner even when the majority
-of features are asymmetrically distributed or expressed (or both) in the groups
-[@dosSantos:2024]. Thus, using and reporting scale uncertainty should become a
-standard practice in the analysis of HTS datasets.
+Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses.
+The development of HTS and the associated problems of very high dimensional data that was not always statistically well-behaved led to many different proposed solutions including multiple normalizations and moderated test statistics.
+That these perform poorly in real data is shown by the simulation data in this report and elsewhere  where both DESeq2 (which we used) and edgeR (which uses moderated statistical tests) performed poorly in controlling the FDR [@Li:2022aa].
+While these approaches often work in many datasets  they fail to address the underlying problems of information that is missing in the data which is what is supplied by adding uncertainty around the information we have about that data.
+The addition of scale uncertainty directly addresses the missing information by  testing the model over a range of normalizations [@nixon2024scale].
+In doing so, the scale-based approach removes the need for moderated statistics and can replace the consensus approach that has been proposed by some groups [@Nearing:2022aa;@Song:2023aa] with no additional computational overhead.
+Thus, an advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
+While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions and a detailed description of these points is describe elsewhere [@nixon2024scale].
+
+In summary, we supply a toolkit that makes incorporating scale uncertainty and location information simple to incorporate for transcriptomes or indeed any type of HTS dataset.
+While the underlying scale of the system is generally inaccessible, the effect of scale uncertainty on the analysis outcomes can be modelled and can help explain some of the underlying biology.
+Adding scale information to the analysis allows for more robust inference because the features that are sensitive to scale can be identified and their impact on conclusions weighted accordingly.
+The use of informed  scale models permits difficult to analyze datasets to be examined in a robust and principled manner even when the majority of features are asymmetrically distributed or expressed (or both) in the groups [@dosSantos:2024].
+Thus, using and reporting scale uncertainty should become a standard practice in the analysis of HTS datasets.
 

### 2025-04-14 go3.tex
Penultimate before cleaning

@@
 
-\title{Explicit Scale Simulation for analysis of RNA-sequencing with
-ALDEx2}
+\title{Explicit Scale Simulation for analysis of RNA-sequencing count
+data with ALDEx2}
 \author{true \and true \and true}
@@
 
-\doublespacing 
-\singlespacing
+\doublespacing \singlespacing 
+\linenumbers
 
@@
 acid is carried over to subsequent steps. Thus, with each sampling step
-the connection between the actual size of the sampled DNA pool and the
-scale (e.g., size, microbial load, or total gene expression) of the
-measured biological system is degraded or lost. In the end, the
-information contained in the data relates only to relative abundances
-and has an arbitrary scale imposed by the sequencing process (1--3).
-Researchers can use modified experimental protocols (4--7) or machine
-learning methods (8) to uncover the biological variation in scale.
-However, wet-lab protocols only provide information on the size of the
-data downstream of the step in the sample preparation protocol where the
-intervention was made and introduce an additional source of variation
-that must be accounted for (5). The computational methods developed for
-microbiome analysis have low correlation with the actual scale but are
-useful (8). In short, the disconnect between sample-to-sample variation
-in sequencing depth and biological variation in scale remains an open
-challenge.
+the connection between the actual number of molecules in the sampled DNA
+pool and the environmental scale (e.g., total number of molecules,
+microbial load, or total gene expression) of the measured biological
+system is degraded or lost. In the end, the information contained in the
+data relates only to relative abundances and has an arbitrary scale
+imposed by the sequencing process (1--3).
 
 The analysis of HTS data suffers from several known problems that can be
-traced, in whole or in part, to misspecification of scale. The first
-issue is poor control of the false discovery rate (FDR) (9--13),
-exhibited as dataset-dependent FDR control and by the disconnect between
-statistical and biological significance (14). In current practice, these
-issues are addressed by a dual-filtering method, whereby both a low
-p-value (or equivalently a low q-value following FDR correction (15))
-and a large difference between groups is used to identify interesting
-transcripts or genes for follow-up analysis (14, 16). This
-double-filtering approach is graphically exemplified by the volcano plot
-(16), but is known to not appropriately control the FDR (17, 18). The
-second issue is poor performance when analyzing data where the mean
-change between groups is non-zero (3). Such asymmetric data can arise
-when a gene set is expressed in one group but not the other, or when one
-group contains different gene content from the other. This type of data
-frequently arises in in-vitro selection experiments (SELEX),
-transcriptome analysis, and microbiome analysis (19). The third issue is
+traced, in whole or in part, to misspecification of scale in the output
+data. The first issue is poor control of the false discovery rate (FDR)
+(4--8), exhibited as dataset-dependent FDR control which is observed as
+a disconnect between statistical and biological significance (8, 9). In
+current practice, this issue is addressed by a dual-filtering method,
+whereby both a low p-value (or equivalently a low q-value following FDR
+correction (10)) and a large difference between groups is used to
+identify transcripts or genes of interest for follow-up analysis (9,
+11). This double-filtering approach is graphically exemplified by the
+volcano plot (11), but is known to not appropriately control the FDR
+(12, 13). Since there is no standard way of determining what fold-change
+cutoff should be used researchers have unlimited degrees of freedom
+which is known to lead to unreliable inference (14). In particular Li et
+al. (8) used patient or clinical-derived transcriptome datasets and
+found that many methods suffer from an extremely high false positive
+rate. The second issue is poor performance when analyzing data where the
+mean change between groups is non-zero (3). Such asymmetric data can
+arise when a gene set is expressed in one group but not the other, or
+when one group contains different gene content from the other. This type
+of data frequently arises in in-vitro selection experiments (SELEX),
+transcriptome analysis, and microbiome analysis (15). The third issue is
 that the actual scale of the environment is often a major confounding
-variable during analysis (3, 8). The final issue is that these problems
-become more pronounced as more samples are collected; that is, more
-information results in a worsening of the accuracy of the analysis (3,
-13, 20).
+variable during analysis (3, 16). This has long been known in
+transcriptome analysis and was a major driver for the development of
+normalization factors (17) and the use of molecular spike-ins to provide
+a reference set of known counts (18--21) to estimate scale. While often
+useful, spike-in methods only provide information downstream of the step
+in the sample preparation protocol where the intervention was made and
+introduce an additional source of variation that must be accounted for
+(19). In the microbiome field, a recent landmark paper showed that
+biological scale was a major unacknowledged confounder in many human
+analyses (16). These authors built a machine learning model to uncover
+the biological variation in scale and including this information was
+useful despite exhibiting only a modest correlation (estimated to be
+0.6) with the scale of data external to the training set (16). The final
+issue is that all the above problems become more pronounced as more
+samples are collected; that is, more information results in optimizing
+for a precise but inaccurate analysis (3, 8, 22).
 
@@
 Biological variation in scale often represents an important unmeasured
-confounder in HTS analyses (21). For example, cells transformed by the
-cMyc oncogene have about 3 times the amount of mRNA and about twice the
-rRNA content than non-transformed cells (22), and this dramatically
-skews transcriptome analysis unless spike-in approaches are used (5). In
-addition, wild-type and mutant strains of cell lines, yeast or bacteria
-have different growth rates and RNA contents under different conditions,
-which affect our ability to identify truly differentially abundant genes
-(23--25). As another example, the total bacterial load of the vaginal
-microbiome differs by 1-2 orders of magnitude in absolute abundance
-between the healthy and bacterial vaginosis states (26), and the
-composition between these states is dramatically different (27, 28).
-Thus, a full description of any of these systems includes both relative
-change (composition) and absolute abundance (scale). Current methods
-access only the compositional information yet make implicit assumptions
-about the scale (20).
-
-Recently, Nixon et al. (3) showed that the challenge of non-biological
-variation in sequencing depth be viewed as a problem of
+confounder in HTS analyses (15, 16, 19, 23). For example, cells
+transformed by the cMyc oncogene have about 3 times the amount of mRNA
+and about twice the rRNA content than non-transformed cells (24), and
+this dramatically skews transcriptome analysis unless spike-in
+approaches are used (19). In addition, wild-type and mutant strains of
+cell lines, yeast or bacteria have different growth rates and RNA
+contents under different conditions, which affect our ability to
+identify truly differentially abundant genes (25--27). As another
+example, the total bacterial load of the vaginal microbiome differs by
+1-2 orders of magnitude in absolute abundance between the healthy and
+bacterial vaginosis states (28), and the cell and RNA composition
+between these states is dramatically different (29, 30). Thus, a full
+description of any of these systems includes both relative change
+(composition) and absolute abundance (scale). Current methods access
+only the compositional information yet make implicit assumptions about
+the scale (22).
+
+Nixon et al. (3) showed that the challenge of non-biological variation
+in sequencing depth could be explained as a problem of
 partially-identified models. They showed that \emph{all} normalizations
 make some assumption about scale but these implicit assumptions are
-often inappropriate and difficult to interpret. This causes different
-normalizations to provide different outputs when applied to the same
-dataset (9, 11, 29--31). Intuitively, normalizations in widespread use
-assume that either all samples have the same scale, e.g.~proportions,
-rarefaction (32), RPKM (33, 34), etc; or that a subset of features in
-one sample can be chosen as a reference to which the others are scaled
-e.g.~the TMM (35), or LVHA (19) or the additive log-ratio (36); or that
-different sub-parts of each sample maintain a constant scale across
-samples e.g.~the RLE (37); or that the geometric mean of the parts is
-appropriate e.g the CLR (38) and its derivatives.
-
-The original naive ALDEx2 (39) model unwittingly made a strict
-assumption about scale through the CLR normalization (3). This
-assumption was often close enough to the true value to be useful, but
-was not always the a good estimate and could be outperformed by other
-normalizations (40). Nixon et al. (3) showed that better scale
-assumptions resulted in more reproducible data analysis including better
-control of both false positive and false negative results. We recently
-modified ALDEx2 to explicitly model the scale over a range of reasonable
-normalization parameters, and showed significant improvements in
-performance in microbiome and in-vitro selection experiments (20). Here,
-we briefly review these modifications and show how scale uncertainty can
-greatly improve modeling in transcriptome and meta-transcriptome
-datasets to provide more robust and reproducible results.
+often inappropriate and difficult to interpret. As a result, different
+normalizations provide different outputs when applied to the same
+dataset (4, 6, 17, 31, 32). Intuitively, normalizations in widespread
+use assume that either all samples have the same scale,
+e.g.~proportions, rarefaction (33), RPKM (34, 35), etc; or that a subset
+of features in one sample can be chosen as a reference to which the
+others are scaled e.g.~the TMM (36), the LVHA (15) qPCR (37), the
+additive log-ratio (38); or that different sub-parts of each sample
+maintain a constant scale across samples e.g.~the RLE (39); or that the
+geometric mean of the parts is appropriate e.g the CLR (40) and its
+derivatives. Notably when the assumption of similar scale across samples
+is not violated, or violated only weakly, any method will provide a
+reasonable analysis. The problem arises when this assumption is
+violated, in which case all tools fail without warning.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.6\linewidth]{./go3_files/figure-latex/scale-explained}
+
+}
+
+\caption{Mismatches between estimated scale and true scale lead to poor estimation of high throughput sequencing data.
+All normalizations used for differential abundance analysis make some strict assumption about the scale of the environment as shown in line 1.
+In this example all normalizations produce a biased estimate of the environmental scale, but estimate C is the closest to the truth.
+Adding uncertainty to normalization C as represented by the distribution in the line 2 leads to less bias and now includes the actual environmental scale in the assumption.
+As shown in line 3 in some cases it may be useful to adjust the centre of the scale uncertainty estimate if the initial normalizations give very poor estimates of the underlying environment.}\label{fig:unnamed-chunk-1}
+\end{figure}
+
+The original scale-naive ALDEx2 (41) model unwittingly made a strict
+assumption about scale through the CLR normalization and we found that
+the CLR was very sensitive to violations of the assumption of scale
+identity between groups (15). Moreover, when the assumption of identity
+was not true the CLR used by ALDEx2 could be outperformed by other
+normalizations in simulation studies (42). As illustrated graphically in
+Figure 1, Nixon et al. (3) showed through simulation that introducing
+uncertainty in the scale assumptions, and in extreme cases altering the
+location of the scale assumption, resulted in more reproducible data
+analysis including better control of both false positive and false
+negative results. We modified ALDEx2 to explicitly model uncertainty in
+scale over a range of reasonable normalization parameters, and showed
+significant improvements in performance in microbiome and in-vitro
+selection experiments (22) and in a vaginal metatranscriptome analysis
+(43). Here, we briefly review these modifications and show how adding
+scale uncertainty can greatly improve modeling in transcriptome and
+meta-transcriptome datasets to provide substantially more robust and
+reproducible results.
 
@@
 Formal and expanded descriptions of the concepts that follow are given
-in (3, 20). To be concrete, we let \(\mathbf{Y}\) denote the
+in (3, 22). To be concrete, we let \(\mathbf{Y}\) denote the
 \emph{measured} \(D \times N\) matrix of sequence counts with elements
@@
 mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
-\(n\). Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount
-of class \(d\) in the biological system from which sample \(n\) was
-obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
-scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
-\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
-\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
-\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
-\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
-\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
-Note that with these definitions \(\mathbf{W}\) can be written as the
+\(n\). Likewise, we can denote \(\mathbf{W}_{dn}\) as the \emph{true}
+amount of class \(d\) in the biological system from which sample \(n\)
+was obtained. We can think of \(\mathbf{W}\) as consisting of two parts,
+the scale \(\mathbf{W}^{Tot}\) (e.g., totals) and the composition
+\(\mathbf{W}^{Comp}\) (i.e., proportions). That is, \(\mathbf{W}^{Tot}\)
+is a \(N\)-vector with elements
+\(\mathbf{W}^{Tot}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
+\(\mathbf{W}^{Comp}\) is a \(D \times N\) matrix with elements
+\(\mathbf{W}^{Comp}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{Tot}_{n}\). Note
+that with these definitions \(\mathbf{W}\) can be written as the
 element-wise combination of scale and composition:
-\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\),
-or as the logarithm
-\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
-\mathbf{W}^{\perp}_{n}\).
-
-Many of the normalizations in widespread use in tools such as DESeq2
-(41), edgeR (35), metagenomeSeq (42) ALDEx2 (43) can be stated as ratios
-of the form \(\hat{{\mathbf{W}}}_{dn}
-\approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is
-determined by some function of the observation. We use the
-\(\hat{{\mathbf{W}}}\) (\(\ \hat{}\
-\)) notation to indicate that the output is an estimate of the true
-value. The technical variation in sequencing depth
-(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
-observed data \(\mathbf{Y}\) provides us with information about the
-system composition \(\mathbf{W}^{\parallel}\) but little to no
-information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
-al.~2011).
+\(\mathbf{W}_{dn}=\mathbf{W}^{Comp}_{dn}\mathbf{W}^{Tot}_{n}\), or as
+the logarithm
+\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{Comp}_{dn} + \log \mathbf{W}^{Tot}_{n}\).
+
+Many of the normalizations used in tools such as DESeq2 (44), edgeR
+(36), metagenomeSeq (45) ALDEx2 (46) can be stated as ratios of the form
+\(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where
+the denominator is determined by some function of the observation. We
+use the hat notation (\(\ \hat{}\ \)) to indicate that the output is an
+estimate of the true value. The technical variation in sequencing depth,
+which is often called ``library size'' has no relationship with the
+actual number of molecules in the sampled environment (1). In other
+words (\(\mathbf{Y}^{Tot}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) the observed
+data \(\mathbf{Y}\) provides us with information about the system
+composition \(\mathbf{W}^{Comp}\) but little to no information in the
+system scale \(\mathbf{W}^{Tot}\) (Lovell et al.~2011).
 
@@
 
-The ALDEx2 R package (39, 43) is a general purpose toolbox to model the
-uncertainty of HTS data and to use that model to estimate the underlying
-LFC (log-fold change) significance. At a high-level, ALDEx2 has three
-connected components to estimate the uncertainty inherent in HTS
-datasets. First, the tool accounts for the uncertainty of the sequencing
-counts using Dirichlet multinomial sampling to build a probabilistic
-model of the data; i.e.,
-\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
-Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
-data (39). However, this step was modified recently to account for scale
-uncertainty and misspecification (20) via a scale model, explained with
-more details in (3, 20) and summarized in the next paragraph. Finally, a
-standard null-hypothesis test and a non-parametric estimate of mean
-standardized difference are used to report on the finite sample
-variation. These sources of uncertainty and variation are combined via
-reporting the expected values from a Monte-Carlo simulation framework.
-For simplicity, we use the term `difference' to refer to the absolute
-difference between groups, and `dispersion' to refer to the
-within-condition difference or pooled variance as defined in (39). These
-are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
-20, 39, 43).
-
-Scale models can be incorporated into ALDEx2, turning the ALDEx2 model
+The ALDEx2 R package (41, 46) is a general purpose toolbox to model the
+uncertainty of HTS data and to use that model to estimate the
+significance of the underlying LFC (log-fold change). At a high-level,
+ALDEx2 has three connected components to estimate the uncertainty
+inherent in HTS datasets. First, the tool accounts for the uncertainty
+of the sequencing counts using Dirichlet multinomial sampling to build a
+probabilistic model of the data; i.e.,
+\(\mathbf{\hat{W}}^{Comp} \approx \mathrm{Dir}(\mathbf{Y})\). Secondly,
+ALDEx2 uses the centred log-ratio transformation to scale the data (41).
+It was this step that was modified to account for scale uncertainty and
+misspecification (22) explained with more details in (3, 22) and
+summarized in the next paragraph. Finally, a standard null-hypothesis
+test and a non-parametric estimate of mean standardized difference are
+used to report on the finite sample variation. These sources of
+uncertainty and variation are combined via reporting the expected values
+from a Monte-Carlo simulation framework. For simplicity, we use the term
+`difference' to refer to the absolute difference between groups, and
+`dispersion' to refer to the within-condition difference or pooled
+variance as defined in (41). These are calculated on a \(\log_2\) scale.
+For more details on ALDEx2 see (3, 22, 41, 46).
+
+Scale models were incorporated into ALDEx2, turning the ALDEx2 model
 into a specialized type of statistical model which Nixon et al. (3) term
@@
 centred log-ratio normalization step. They did this by including a model
-for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by
-ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\),
-where \(G_n\) is the geometric mean of sample n, which while being a
-random variable, is essentially constant across each Monte-Carlo
-replicate, but that differs between samples. With this modification,
-ALDEx2 can be generalized by considering probability models for the
-scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
-example, the following scale model generalizes the CLR:
-
-\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda
-\sim N(\mu, \gamma^{2})\].
-
-This formulation is quite flexible (3, 20). In the simple or `default'
+for \(\mathbf{\hat{W}}^{Tot}_{n}\). The CLR normalization used by ALDEx2
+makes the assumption \(\mathbf{\hat{W}}^{Tot}_{n}=1/G_{n}\), where
+\(G_n\) is the geometric mean of the counts (or the corresponding
+proportions)of each part in sample n, which while being a random
+variable, is essentially constant across each Monte-Carlo replicate, but
+that differs between samples. With this modification, ALDEx2 can be
+generalized by considering probability models for the scale
+\(\mathbf{\hat{W}}^{Tot}_{n}\) that have mean \(1/G_{n}\). For example,
+the following scale model generalizes the CLR:
+
+\[\log \mathbf{\hat{W}}^{Tot}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2}).\]
+
+This formulation is quite flexible (3, 22). In the simple or `default'
 configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
-from a log-Normal distribution(3). Adding scale uncertainty with the
-\(\gamma\) paramenter controls only the degree of uncertainty of the CLR
-assumption for the \(x_{n}\) binary condition indicator (e.g.,
-\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). In the
-advanced or `informed' configuration, \(\mu\) takes different values for
-each group and controls the location of the LFC assumption; combining
-\(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the
-location and the scale. An example of both the default and informed
-approaches is given for a microbiome dataset in (20) showing increased
-sensitivity and specificity. Here we show that these approaches also
-work well in transcriptome and metatranscriptome datasets. These
+from a log-Normal distribution (3). Adding scale uncertainty with the
+\(\gamma\) paramenter (as shown in Figure 1:1) controls only the degree
+of uncertainty of the CLR assumption for the \(x_{n}\) binary condition
+indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes
+control). In the advanced or `informed' configuration, \(\mu\) takes
+different values for each group and controls the location of the LFC
+assumption (as shown in Figure 1:2); combining \(\mu\) with a \(\gamma\)
+estimate allows for uncertainty in both the location and the scale. In
+the manuscript where the idea of scale was original derived, Nixon et
+al. (3) conducted extensive simulations showing that both the default
+and informed approaches exhibit increased sensitivity and specificity
+(22). In this report we show that these approaches also work well to
+control the FDR in transcriptome and metatranscriptome datasets. We also
+provide some additional insights into how this is achieved. These
 modifications are instantiated in ALDEx2 which is the first software
@@
 
-Gierliński et al. (44) generated a highly replicated yeast transcriptome
+Gierliński et al. (47) generated a highly replicated yeast transcriptome
 dataset to compare gene expression between a wild-type strain and a snf2
-gene knockout, \(\Delta\)snf2. This dataset was used to test several
-RNA-seq tools for their power to detect the set of differentially
-abundant transcripts identified in the full dataset when the data was
-subset (14). In this study each tool had its own `gold standard' set of
-transcripts with different tools identifying between between 65\% to
+gene knockout, \(\Delta\)snf2. This dataset of 86 samples (44 and 42 per
+group) is an example of technical growth replicate experiments common in
+the literature. The dataset was used to test several RNA-seq tools for
+their power to detect the set of differentially abundant transcripts
+identified in the full dataset when the data was subset (9). In this
+original study each tool had its own `gold standard' set of transcripts
+with different tools identifying between between 65\% to
 \textgreater80\% of all transcripts as being significantly different.
@@
 authors suggested that it was more appropriate to apply a dual cutoff
-composed of both a Benjamini-Hochberg (45) corrected p-value (q-value)
+composed of both a Benjamini-Hochberg (48) corrected p-value (q-value)
 plus a difference cutoff to limit the number of identified transcripts
-to a much smaller fraction of the total. Nixon et al, (20) showed that
-adding even a small amount of scale uncertainty with ALDEx2 dramatically
-reduced the number of significant transcripts identified, removing the
-need for the dual cutoff approach in this dataset and others. Below we
-include an intuitive explanation of why and how incorporating scale
-uncertainty achieved this outcome using a setting of \(\gamma=0.5\). The
-approach is in line with the recommendations of (20) and gives results
-comparable to those proposed in (14).
+to a much smaller fraction of the total. In other datasets, Nixon et al,
+(22) showed that adding even a small amount of scale uncertainty with
+ALDEx2 dramatically reduced the number of significant transcripts
+identified, removing the need for the dual cutoff approach in this
+dataset and others.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{go3_files/figure-latex/yst-plot-1}
+
+}
+
+\caption{Volcano and effect plots for unscaled and scaled transcriptome analysis.
+ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset.
+The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots.
+Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty.
+Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange).
+Points in grey are not statistically signficantly different under any condition.
+The horizontal dashed lines represent a \(\log_2\)difference of \(\pm 1.4\).}\label{fig:yst-plot}
+\end{figure}
 
 We start with the assumption that not all statistically significant
-differences are biologically relevant (46), and that such a large number
-of significantly different transcripts breaks the necessary assumption
-for DA/DE expression that most parts be invariant (30). As noted,
-transcriptomics commonly uses a dual cutoff approach that is graphically
-exemplified by volcano plots (14, 16). Using either DESeq2 or ALDEx2, a
-majority of transcripts are statistically significantly different
-between groups with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%,
-DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. These values are
-in line with those observed by (14). Such large numbers of statistically
-significant transcripts seems biologically unrealistic. That 118
-transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
-identifies 582 transcripts that ALDEx2 does not, suggests that the
-choice of normalization plays a role in which results are returned as
-significant and that some, if not the majority, are driven by technical
-differences in the analysis (13, 30).
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/plot1-1.pdf}}
-\caption{Volcano and effect plots for unscaled and scaled transcriptome
-analysis. ALDEx2 was used to conduct a differential expression (DE)
-analysis on the yeast transcriptome dataset. The results were plotted to
-show the relationship between difference and dispersion using effect
-plots or difference and the q-values using volcano plots. Panels A,C are
-for the naive analyses, and Panels B,D are for the default analyses that
-include scale uncertainty. Each point represents the values for one
-transcript, with the color indicating if that transcript was significant
-in the both analyses (red) or in the naive analysis only (orange).
-Points in grey are not statistically signficantly different under any
-condition. The horizontal dashed lines represent a \(\log_2\)difference
-of \(\pm 1.4\).}
-\end{figure}
-
-The Volcano plots in Figure 1 A and B show that adding scale uncertainty
+differences are biologically relevant (49), and that a result where the
+majority of transcripts are significant breaks the necessary assumption
+for DA/DE expression that most parts be invariant (31). Transcriptomic
+analysis commonly uses a dual cutoff approach graphically exemplified by
+volcano plots (9, 11). Using either DESeq2 or ALDEx2, a majority of
+transcripts are statistically significantly different between groups
+with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%, DESeq2) or 4172
+(71\%, ALDEx2) of the 5891 transcripts. These values are in line with
+those observed by (9). Such large numbers of statistically significant
+transcripts seems biologically unrealistic. That 118 transcripts are
+identified by ALDEx2 and not DESeq2, while DESeq2 identifies 582
+transcripts that ALDEx2 does not, suggests that the choice of
+normalization plays a role in which results are returned as significant
+and that some, if not the majority, are driven by technical differences
+in the analysis (8, 31) or are false positives.
+
+The Volcano plots in Figure 2 A and B show that adding scale uncertainty
 increases the minimum q-value and increases the concordance between the
 q-value and the difference between groups (compare panels A and B). The
-effect plots (47) in Figure 1C shows that the majority of significant
+effect plots (50) in Figure 2C shows that the majority of significant
 transcripts (red, orange) have negligible differences between groups and
@@
 the experimental design which is actually a technical wet lab
-replication rather than a true biological replication design (44). Scale
+replication rather than a true biological replication design (47). Scale
 uncertainty can be incorporated using the \texttt{gamma} parameter that
 controls the amount of noise added to the CLR mean assumption when we
-call either \texttt{aldex()}, or \texttt{aldex.clr()}. Figure 1 B,D
-shows that setting \(\gamma=0.5\) results in 205 which is far fewer
-significant transcripts than in the naive analysis and we observe that
-the minimum dispersion increases from 0.12 (\(\gamma = 0\) ) to 0.67
-(\(\gamma=0.5\)).
+call either \texttt{aldex()}, or \texttt{aldex.clr()}. Figure 2 B,D
+shows that setting \(\gamma=0.5\) now results in 205 which is far fewer
+statistically significant transcripts than in the naive analysis and we
+observe that the minimum dispersion increases from 0.12 (\(\gamma = 0\)
+) to 0.67 (\(\gamma=0.5\)).
+
+It is common practice to use a dual-cutoff by choosing transcripts based
+on a thresholds for both q-values and fold-changes (9), and these were
+first proposed for microarray experiments through volcano plots (11).
+Note that there is considerable variation in recommended cutoff values
+(9), and that this controversy has persisted ever since fold-change was
+suggested (51). Unfortunately, universal cutoff fold-change values
+cannot be identified in part because different tools have intrinsically
+different variance in their log\_2-fold change ranges (52). This has led
+to the widespread practice of applying a post-hoc fold-change cutoff to
+reduce the number of positive identifications to a manageable proportion
+of the whole dataset. Here, we applied the dual-cutoff method using a
+fold-change of at least a \(2^{1.4}\) fold change that reduces the
+number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
+This cutoff was chosen for convenience and is mid-way between the high
+and low fold-change recommendations of (9). These limits are shown by
+the dashed grey lines in Figure 2 and we can see that a \(2^{1.4}\) fold
+change \((\sim 2.6\) fold) cutoff identifies a similar number of
+transcripts as does ALDEx2 using \(\gamma = 0.5\) which identifies 205
+transcripts.
+
+Supplementary Figures 1 shows an example of the
+\texttt{aldex.senAnalysis()} function to identify those transcripts that
+are very sensitive to scale uncertainty in this dataset. Here we see
+that adding a very small amount of scale \(\gamma = 0.1\) reduces the
+number of significant transcripts by more than half in the yeast
+dataset. This allows the analyst to ignore those low-dispersion
+transcripts that were significant only because of an absence of scale
+uncertainty.
+
+We next examined how adding scale would alter the analysis in a real
+dataset to which synthetically generated true positive counts had been
+added. We show results from the anti-PD-1 therapy RNA-seq dataset (53)
+which examined changes in gene expression when cells were exposed or not
+to a cell-cycle checkpoint inhibitor. This dataset was used by Li et al.
+(8) as an example of the dangers of relying on tools with high false
+positive error rates when analyzing clinical or clinically-related
+transcriptome samples. Indeed, in the benchmarking analysis done by this
+group, they found that parameter based methods such as DESeq2 and edgeR
+that on reported p-value and fold-change cutoffs often led to
+conclusions in the original dataset that were indistinguishable from
+permutations of the dataset. In other words, that the analysis of
+transcriptome datasets from patient-derived samples often exhibited many
+false-positive identifications.
 
 \begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/disp-1.pdf}}
-\caption{Adding scale uncertainty changes the dispersion distribution.
-Panel A shows a plot of the expected value for relative abundance vs the
-expected value for the pooled dispersion as output by
-\texttt{aldex.effect}. The dashed horizontal lines show the median value
-for the features with a rAbundance between -0.5 and 0.5, and the light
-colored lines are lowess lines of fit through the center of mass of the
-data. Panel B plots the dispersion difference between \(\gamma = 1\) and
-\(\gamma = 0\); note the non-linear relationship that highlights the
-rotation that is evident in Panel A. The colored lines indicate the
-lowess line of fit through the centre of mass of the plot for the
-various populations of points. The grey line is the total population and
-shows the difference \(\Delta\), the red line is the population of
-significant transcripts (*) with \(\gamma = 0\), the orange line is the
-population of significant transcripts with a difference threshold (T) of
-about \(\pm 2^{1.4}\)-fold change, the blue line is the population of
-significant transcripts with \(\gamma = 0.5\), and the cyan line is the
-significant population with \(\gamma = 1\). \(\Delta\): Difference, *:
-significant, T: thresholded.}
+
+{\centering \includegraphics[width=0.8\linewidth]{go3_files/figure-latex/immuno-1}
+
+}
+
+\caption{Results of modeling true positive (TP) differences in the PD-1 dataset.
+For this the data were shuffled and  5\% of the transcripts were modeled to have TP differences between groups where the differences were drawn from a Normal distribution with a mean difference of 0 and a sd of 2.
+Panel A shows the mean FDR of 10 instances for scale-naive ALDEx2 (A 0), and ALDEx2 with \(\gamma=0.2\) (A 0.2) or \(\gamma=0.2\) (A 0.5), and where significant features were identified by  DESeq2 without (D 0), or with (D 0.5) a 0.5-fold change difference.
+The x-axis shows how the FDR changes for each tool as a function of the estimated modeled difference between groups.
+Panel B shows the mean sensitivity (TP found /all TP).
+Panels C and D show volcano and effect plots for the ALDEx2 output with the transcripts identified as significant at each scale setting as colored points, and non-significant transcripts in black.}\label{fig:immuno}
 \end{figure}
 
-It is common practice to use a dual-cutoff by choosing transcripts based
-on a thresholds for both q-values and fold-changes (14, 16). Note that
-there is considerable variation in recommended cutoff values(14). Here,
-applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold
-change reduces the number of significant outputs to 193 for DESeq2 and
-to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line
-with the recommendations of (14) with the fold change limits shown by
-the dashed grey lines in Figure 1. The \(2^{1.4}\) fold change cutoff
-identifies a similar number of transcripts as does ALDEx2 using
-\(\gamma = 0.5\) which identifies 205. Supplementary Figures 1 and 2
-shows how to use the \texttt{aldex.senAnalysis()} function to identify
-those transcripts that are very sensitive to scale uncertainty. In this
-supplementary figure we see that even adding a very small amount of
-scale \(\gamma = 0.1\) reduces the number of significant transcripts by
-more than half. This allows us to ignore those low-dispersion
-transcripts that were significant only because of an absence of scale
-uncertainty. In practice, we suggest that a \texttt{gamma} parameter
-between 0.5 and 1 is realistic for most experimental designs (20).
-
-The effect on dispersion with increasing amounts of scale uncertainty
-are shown in Figure 2A, where we can see that the dispersion increases
-as uncertainty is added. Note that the dispersion in the unscaled
-analysis in Figure 2A reaches a minimum near the mid-point of the
-distribution, and also does so when the analysis is conducted with
-DESeq2 (Supplementary Figure 3). This shows more clearly that dispersion
-of many transcripts is almost negligible in the absence of scale
-uncertainty. This plot makes the counter-intuitive suggestion that the
-variance in expression of the majority of genes with moderate expression
-is more predictable than highly-expressed genes or of housekeeping genes
-(48). This is at odds with the known biology of cells where single cell
-counting of highly-expressed transcripts shows that they have little
-intrinsic variation (23, 49).
-
-Adding scale uncertainty by setting \(\gamma=0.5\) , or
-\(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 2A
-by the red and blue data points, and by the colored lines of fit through
-the centre of mass of the data. Less obvious is that the additional
-dispersion is not applied equally to all points. Figure 2B shows a plot
-of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data and
-here we can see that scale uncertainty is preferentially increasing the
-dispersion of the mid-expressed transcripts that formerly had negligible
-dispersion; examine the grey line of best fit (overlaid by the red line)
-for the trend. Panel B also shows the trend of the expression-dispersion
-relationship for transcripts that are classed as statistically
-significant. The red line shows the trendline with no added scale
-uncertainty, and this trendline exactly overlays with the grey trendline
-for the bulk of transcripts. The orange trendline indicates those
-transcripts that are both statistically significant and that have a
-thresholded expression level of \(\pm 1.4\), and the dark blue and cyan
-lines show the statistically significant trendline for \(\gamma=0.5\) ,
-or \(1.0\).
-
-Thus, taking Figures 1 and 2 together adding scale uncertainty has the
-desirable effect of changing the distribution of transcripts identified
-as significantly different between groups. Those parts that were
-statistically significantly different \emph{only because of low
-dispersion} are preferentially excluded from statistical significance
-while those parts that were significantly different because of a high
-difference between groups remain.
+Figure 3 shows the results of ten permutations of this dataset while
+simulating 5\% of the transcripts to be true positives where the
+difference from no change was derived from a Normal distribution (54).
+This simulation was conducted with the \texttt{seqgendiff} R package
+(54) to both permute the dataset and to add true positive features. We
+did ten permutations and kept track of the number of true and false
+positive identifications. Figure 3A shows the actual false discovery
+rate for ALDEx2 with and without the addition of scale uncertainty and
+for DESeq2 with or without a fold-change cutoff of \(2^{0.5}\). When the
+modeled difference was greater than 0, ALDEx2 exhibited a FDR of
+\(\gamma = 0\): 0.002, \(\gamma = 0.2\): 0, and \(\gamma = 0.5\): 0,
+while DESeq2 with no fold-change cutoff had an FDR of 0.32 and with a
+0.5 fold-change cutoff an FDR of 0.31. Figure 3 also plots these results
+as a function of the minimum modelled fold change of the true positive
+transcripts. First, we can see that this analysis recapitulates the
+observations of Li et al (8) in that DESeq2 has very poor false positive
+control at a nominal FDR of 0.05. The FDR is not controlled any better
+when a fold-change cutoff is applied, and this agrees with previous work
+showing that fold-change cutoffs do not materially improve FDR control
+in high throughput datasets (12, 13). Figure 3B shows that DESeq2 has
+higher sensitivity than does ALDEx2, and not surprisingly this
+sensitivity increases as the difference between groups increases. For
+the case where the mean difference is 0 or greater, ALDEx2 exhibited a
+sensitivity of \(\gamma = 0\): 0.66, \(\gamma = 0.2\): 0.54, and
+\(\gamma = 0.5\): 0.35, while DESeq2 with no fold-change cutoff had a
+sensitivity of 0.76 and with a 0.5 fold-change cutoff a sensitivity of
+0.73. In this example, scale-naive ALDEx2 has near perfect FDR control
+and reasonable sensitivity at low modeled difference between groups.
+However, when the modeled difference between groups becomes large, then
+the scale-naive version of ALDEx2 begins to exhibit unacceptable rates
+of false positives and the false positive rate for DESeq2 also
+increases. Adding in even small amounts of scale uncertainty drops the
+true FDR rate to 0, but at the expense of sensitivity. The major
+contributor to the increase in FDR with larger modeled differences is
+that the tools are identifying as positives those transcripts that are
+modeled to have differences just below the threshold. We need to
+recognize that there is no such thing as a statistical free lunch; the
+analyst can have high sensitivity but low confidence that any individual
+transcript is truly different, or have lower sensitivity but have very
+high confidence that the difference is real. In other words, the
+sensitivity of a method is directly tied to how much error the
+investigator is willing to tolerate.
+
+Examination of the volcano plot (11) and effect plot (50) in Figure 3C
+and D provides some insight into why adding scale uncertainty provides
+better FDR control than does the approach of using a p-value and a
+fold-change cutoff. In the volcano plot, adding scale uncertainty
+differentially excludes transcripts with a combination of marginal
+p-values and low difference between groups, and that this becomes more
+pronounced with a larger scale value. This effect is also seen in Figure
+2 but is more nuanced. In contrast, the fold-change cutoff does not
+incorporate the magnitude of the p-value and so transcripts with large
+differences, but marginal p-values are retained. The effect plot in
+Figure 3D shows that the transcripts with marginal p values and large
+differences that are excluded when scale uncertainty is added are those
+that have a large dispersion.
+
+As a concrete example consider the point that is circled in panels C and
+D with a marginal p-value, with a difference between of nearly 4 and a
+dispersion of greater than 6. This transcript is no longer significant
+when \(\gamma = 0.2\), but would require a very large fold-change cutoff
+to be excluded by the standard approach. In addition, transcripts with
+very small dispersion and very small differences are also excluded when
+scale uncertainty is added. Thus, the addition of scale uncertainty
+achieves the desired outcome of lowering the FDR for those transcripts
+that are either marginally differentially abundant, or where the
+underlying dispersion--and hence the uncertainty in measurement--is very
+high, or for transcripts that fit both criteria.
+
+Supplementary Figure 2 shows a second permuted dataset with the addition
+of modeled differences between groups. Here we used another real dataset
+with over 200 biological replicates of BRCA1 tumor and control tissue
+samples from Li et al. (8). This Supplementary Figure shows that the FDR
+control of DESeq2 is somewhat better than in the PD-1 dataset, although
+still much greater than the the anticipated 5\%. Further, a fold-change
+cutoff reduces the FDR of DESeq2 from about 30\% to just over 20\% with
+a power of over 80\%. However, we can see that scale-naive ALDEx2
+performs substantially better with a negligible FDR and comparable
+power. Adding scale uncertainty again improves FDR even for those
+transcripts modeled to have larger differences and the power is
+substantially better than in the PD-1 dataset, reaching the same power
+as DESeq2 or scale-naive ALDEx2 when the modeled difference is large. As
+before, this is driven by removing from consideration those transcripts
+with either a small difference between or a marginal p-value, or both.
+
+Supplementary Figures 3 shows that effect of applying \(\gamma = 0.5\)
+to this dataset results in reducing the number of positive transcripts
+from being \(\sim 70\)\% of the whole dataset to less than 10\% of the
+dataset and that this is largely because of a reduction in significance
+of those transcripts with low dispersion. Supplementary Figures 4 shows
+a sensitivity analysis of the BRCA1 dataset showing that different scale
+uncertainty amounts alter the number of significant transcripts in a
+biological replicate experiment similarly to a technical replicate
+experiment. Supplementary Figures 5 and 6 delve into how adding scale
+uncertainty affects the variace-abundance relationship in subtle ways,
+and may help readers to understand the observations seen in Figure 3 and
+supplementary Figure 2.
+
+Together the results in this section show that adding scale uncertainty
+has the desirable effect of altering the transcripts identified as
+significantly different between groups in a way that exhibits better
+control of FDR albeit with a corresponding reduction in sensitivity.
+Those parts that were statistically significantly different \emph{only
+because of low dispersion} or that \emph{had marginal p-values} or both,
+are now preferentially excluded from statistical significance. In
+practice, we suggest that a \texttt{gamma} parameter of between 0.2 and
+0.5 is realistic for most experimental designs (22) regardless if the
+replication is technical or biological.
 
@@
 
-Dos Santos et al. (50) used a vaginal metatranscriptome dataset to
+Dos Santos et al. (55) used a vaginal metatranscriptome dataset to
 compare the gene expression in bacteria collected from healthy (H) and
-bacterial vaginosis (BV) affected women. In this environment, both the
-relative abundance of species between groups and the gene expression
-level within a species is different (51). Additionally, prior research
-suggests that the total number of bacteria is about 10 times more in the
-BV than in the H condition (26). Thus, these are extremely challenging
-datasets in which to determine differential abundance as there are both
-compositional and scale changes between conditions. The usual method to
-analyze vaginal metratranscriptome data is to do so on an
-organism-by-organism basis (51--53) because the scale confounding of the
-environment is less pronounced. One attempt at system-wide analysis
-returned several housekeeping functions as differentially expressed
-between groups (52); a result likely due to a disconnect between the
-assumptions of the normalization used and the actual scale of the
-environment (19).
-
-In this example, we show how to specify and interpret an informed scale
-model that can explicitly account for some of these modeling
-difficulties (20) even in a difficult dataset. An informed scale model
-can control for both the mean difference of scale between groups (e.g.,
-directly incorporate information on the differences in total number of
-bacteria between the BV and H conditions) as well as the uncertainty of
-that difference. To specify a user-defined scale model, we can pass a
-matrix of scale values instead of an estimate of just the scale
-uncertainty to \texttt{aldex.clr()}. This matrix should have the same
-number of rows as the of Monte-Carlo Dirichlet samples, and the same
-number of columns as the number of samples. While this matrix can be
-computed from scratch by the analyst, there is an
-\texttt{aldex.makeScaleModel()} function that can be used to simplify
-this step in most cases. This encodes the scale model as
-\(\Lambda \sim N(log_2 \mu_n,
-\gamma^{2})\), where \(\mu_n\) represents the scale value for each
-sample or group and gamma is the uncertainty as before. The scale
-estimate can be a measured value (cell count, nucleic acid input, etc)
-or an estimate. Nixon et al. (3, 20) showed that only the ratio of the
-means are important when providing values for \(\mu_n\); i.e., the ratio
-between the \(\log_2 \mu_i\) and \(\log_2 \mu_j\) values. See the
-supplement to Nixon et al. (20) for more information.
-
-Figure 3A shows an effect plot of the data where reads are grouped by
-homologous function regardless of the organism of origin. Each point
-represents one of 3728 KEGG functions (54). There are many more
-functions represented in the BV group (bottom) than in the healthy group
-(top). This is because the \textit{Lactobacilli} that dominate a healthy
-vaginal microbiome have reduced genome content relative to the anaerobic
-organisms that dominate in BV, because there is a greater diversity of
-organisms in BV than in H samples, and because the BV condition has
-about an order of magnitude more bacteria than does the H condition.
-
-The naive scale model appears to be reflecting the bacterial load as
-observed by calculating the mean scale value for each group. Using a
-negligible scale value; i.e., \(\gamma=1e-3\) exposes the naive scale
-estimate for samples in the \texttt{@scaleSamps} slot from the
-\texttt{aldex.clr} output. the naive scale estimate for the
-
-healthy group is 17.41 and for the BV group is 14.59 for a difference of
-2.82. This is interpreted as the scale of the H group of samples being
-7.06 greater than the BV group. This precise but incorrect estimate
-places the location of the housekeeping genes off the midline of no
-difference.
-
-Applying the default scale model of \(\gamma=0.5\) increases the
-dispersion slightly but does not move the housekeeping functions toward
-the midline. This is as expected; the mean of the default scale model is
-based on the CLR normalization so no shift in location would be expected
-over the original ALDEx2 model. Nevertheless, about 30\% of the
-housekeeping functions are no longer statistically significantly
-different. Note that this change is simple to conduct, has no additional
-computational complexity and requires only a slight modification for the
-analyst.
-
-There are 101 functions with low dispersion that appear to be shared by
-both groups (boxed area in Figure 3A, and colored in cyan). Inspection
-shows that these largely correspond to core metabolic functions such as
-transcription, translation, ribosomal functions, glycolysis,
-replication, chaperones, etc (Supplementary file housekeeping.txt). The
-transcripts of many of these are commonly used as invariant reference
-sequences in wet lab experiments (48) and so are not be expected to
-contribute to differences in ecosystem behaviour. The average location
-of these should be centred on 0 difference to represent an internal
-reference set. However, without an informed scale model, the mean of
-these housekeeping functions is approximately located at +2.3.
+bacterial vaginosis (BV) affected women. This dataset is derived from
+two publicly available datasets composed of a set of 20 non-pregnant
+women from London, Ontario Canada (56) and a subset of 22 non-pregnant
+women collected from German women who underwent metronidazole treatment
+for BV (57). Batch effects for these two groups were removed with
+ComBat-seq (58) and the two datasets were merged into one, giving a
+total of 16 H and 26 BV samples. In the Dos Santos paper, all results
+from this initial analysis were replicated in a much larger dataset
+derived from the MOMS-PI study (59)
+
+In this vaginal environment, both the relative abundance of species
+between groups and the gene expression level within a species is
+different (60). Additionally, prior research suggests that the total
+number of bacteria is about 10 times more in the BV than in the H
+condition (28). Thus, these are extremely challenging datasets in which
+to determine differential abundance as there are both compositional and
+scale changes between conditions. The usual method to analyze vaginal
+metratranscriptome data is to do so on an organism-by-organism basis
+(57, 59, 60) because the scale confounding of the environment is less
+pronounced. One attempt at system-wide analysis returned several
+housekeeping functions as differentially expressed between groups (57);
+a result likely due to a disconnect between the assumptions of the
+normalization used and the actual scale of the environment (15).
 
@@
 
-We desire a scale model that approximately centres the housekeeping
-functions, because we expect housekeeping functions tp be nearly
-invariant; thus an appropriate scale in this dataset for functional
-analysis is likely closer to 0 than the naive estimate. One way to
-choose an appropriate value for \(\mu_n\) is to use the
-\texttt{aldex.clr} function on only the presumed invariant functions
-setting \(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot
-as before. Doing so suggests that the difference in scale should be
-about 14\%. A second approach would be to identify the functions used as
-the denominator with the \texttt{denom="lvha"} option (19) for the
+In this example, we show how to specify and interpret a user defined or
+\emph{informed} scale model that can explicitly account for some of
+these modeling difficulties (22) even in a difficult to analyze dataset.
+An informed scale model can control for both the mean difference of
+scale between groups (e.g., directly incorporate information on the
+differences in total number of bacteria between the BV and H conditions)
+as well as the uncertainty of that difference as illustrated in Figure
+1:3. To specify a user-defined scale model, we can pass a matrix of
+scale values instead of an estimate of just the scale uncertainty to
+\texttt{aldex.clr()}. This matrix should have the same number of rows as
+the of Monte-Carlo Dirichlet samples, and the same number of columns as
+the number of samples. While this matrix can be computed from scratch by
+the analyst, there is an \texttt{aldex.makeScaleModel()} function that
+can be used to simplify this step in most cases. This encodes the scale
+model as \(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\)
+represents the scale value for each sample or group and \(gamma\) is the
+uncertainty as before. The scale estimate can be a measured value (cell
+count, nucleic acid input, spike-ins, etc) or an estimate. Nixon et al.
+(3, 22) showed that only the ratio of the means are important when
+providing values for \(\mu_n\); i.e., the ratio between the
+\(\log_2 \mu_i\) and \(\log_2 \mu_j\) values. See the supplement to
+Nixon et al. (22) for more information.
+
+Figure 4A shows an effect plot of the data where reads are grouped by
+homologous function regardless of the organism of origin. Each point
+represents one of 3728 KEGG functions (61). There are many more
+functions represented in the BV group (bottom) than in the healthy group
+(top). This is because the \textit{Lactobacilli} that dominate a healthy
+vaginal microbiome have reduced genome content relative to the anaerobic
+organisms that dominate in BV, because there is a greater diversity of
+organisms in BV than in H samples, and because the BV condition has
+about an order of magnitude more bacteria than does the H condition.
+
+The naive scale model appears to be reflecting the bacterial load as
+observed by calculating the mean scale value for each group. Using a
+negligible scale value; i.e., \(\gamma=1e-3\) exposes the naive scale
+estimate for samples in the \texttt{@scaleSamps} slot from the
+\texttt{aldex.clr} output. The naive scale estimate for the healthy
+group is 17.41 and for the BV group is 14.59 for a difference of 2.82.
+This is interpreted as the scale of the H group of samples being 7.06
+greater than the BV group.
+
+Applying the default scale model of \(\gamma=0.5\) increases the
+dispersion slightly but does not move the housekeeping functions toward
+the midline. This is as expected; the mean of the default scale model is
+based on the CLR normalization so no shift in location would be expected
+over the scale-naive ALDEx2 model. Nevertheless, about 30\% of the
+housekeeping functions are no longer statistically significantly
+different. Note that this change is simple to conduct, has no additional
+computational complexity and requires only a slight modification for the
+analyst.
+
+There are 101 functions with low dispersion that appear to be shared by
+both groups (boxed area in Figure 3A, and colored in cyan). Inspection
+shows that these largely correspond to core metabolic functions such as
+transcription, translation, ribosomal functions, glycolysis,
+replication, chaperones, etc (Supplementary file housekeeping.txt). The
+transcripts of many of these are commonly used as invariant reference
+sequences in wet lab experiments (62) and so are not be expected to
+contribute to differences in ecosystem behaviour. Because we expect
+housekeeping functions to be nearly invariant in their expression and to
+occur in all organisms, the average location of these should be centred
+on 0 difference to represent an internal reference set. However, with
+the naive scale model, the mean of these housekeeping functions is
+approximately located at 2.3. Thus, we desire a scale model that
+approximately centres the housekeeping functions and an appropriate
+scale in this dataset for functional analysis will place these functions
+closer to 0 than does the naive estimate. One way to choose an
+appropriate value for \(\mu_n\) is to use the \texttt{aldex.clr}
+function on only the presumed invariant functions setting
+\(\gamma > 0\), and then accessing the \texttt{@scaleSamps} slot as
+before. Doing so suggests that the difference in scale should be about
+14\%. A second approach would be to identify the functions used as the
+denominator with the \texttt{denom="lvha"} option (15) for the
 \texttt{aldex.clr} function, and then to use these values as before.
@@
 captured as differentially abundant in the BV group. Supplementary
-Figure 4 shows that the using either the 5\% or the 14\% scale
+Figure 7 shows that the using either the 5\% or the 14\% scale
 difference give imperceptibly different results suggesting that an
 informed scale model does not have to precisely estimate the scale
-difference to be useful. Nixon et al, (20) also found that multiple
+difference to be useful. Nixon et al, (22) also found that multiple
 reasonable estimates for the \(\mu_n\) part of the informed scale model
@@
 false negatives (orange functions) even in a very difficult to analyze
-dataset. The remarkable improvements in biological interpretation
-afforded by an informed scale model, and the transferrability of it
-between sample cohorts of the same condition is outlined elsewhere (50).
-We suggest that the default scale model is sufficient when the data are
-approximately centred. However, an informed model is more appropriate
-with datasets are not well centred or when the investigator has prior
-information about the underlying biology.
+dataset. We used this scale model to uncover hiter-to-now unknown
+differences in microbiome functional activity between the Healthy and BV
+cohorts that were missed in previos analyses and that explain important
+clinical differences between them (55). The remarkable improvements in
+biological interpretation afforded by an informed scale model, and the
+transferrability of it between sample cohorts of the same condition is
+outlined in dos Santos et al. (55). We suggest that the default scale
+model is sufficient when the data are approximately centred but that an
+informed model is more appropriate with datasets are not well centred or
+when the investigator has prior information about the underlying
+biology.
 
@@
 
-Biological systems are both predictably variable and stochastic (49) and
+Scale estimates affect two parts of the analysis. Modeling uncertainty
+in scale prevents false certainty in the precision of estimation and
+controls false positive identification. Modeling between-group scale
+relaxes the assumption of identity between the sizes of the environments
+and allows better control of false negative identification. The scale
+estimates can be derived from the total number of molecules in the
+environment or from other estimates of input size (cell counts, initial
+concentrations, spike-ins, growth rates, etc).
+
+Biological systems are both predictably variable and stochastic (63) and
 systems biology experiments show that there are transcripts with
 approximately constant concentrations in the cell and those with large
-variability under different growth conditions (23). Current measurement
+variability under different growth conditions (25). Current measurement
 methods that rely on high throughput sequencing fail to capture all of
-the variation, particularly variation due to scale (3, 20). In the
-absence of external information (5, 6, 55) sequencing depth
-normalisation methods cannot recapture the scale information (5, 21),
+the variation, particularly variation due to scale (3, 22). In the
+absence of external information (19, 20, 64) sequencing depth
+normalisation methods cannot recapture the scale information (19, 23),
 and can only normalize for the technical variation due to sequencing
@@
 a HTS dataset, and showed how to incorporate these models in the
-analysis of microbiome and other datasets (20). They demonstrated that
+analysis of microbiome and other datasets (22). They demonstrated that
 many tools commonly used to analyze HTS datasets had substantial Type 1
-and Type 2 error rates, in line with recent findings by others (10, 12,
-13). A version of ALDEx2 with the ability to include scale uncertainty
-was shown to be able to correct for the high Type 1 error rate for that
-tool, albeit with some loss of sensitivity. Finally, they showed that
+and Type 2 error rates in line with recent findings by others (5, 7, 8).
+A version of ALDEx2 with the ability to include scale uncertainty was
+shown to be able to correct for high Type 1 error rate for that tool,
+albeit with some loss of sensitivity. Finally, they showed that
 incorporating an informed scale model incorporating both location and
 scale uncertainty estimates could both control for Type 1 and Type 2
-error rates (20).
+error rates (22, 43).
+
+The process of choosing the parameters are experiment-specific and can
+be anchored in known information such as cell counts, spike-ins,
+information from the literature or similar (3, 22, 65). In the
+metatranscriptome example used in this report {[}(43);{]} the choice of
+parameter was driven by the assumption inherent in the biology that core
+housekeeping functions would serve as an appropriate standard (37). The
+choice of parameters must be guided by the experimental question and
+other approaches in the literature which suggest normalizing the
+transcript levels to the bacterial metagenomic levels (66) could be used
+to set the scale parameters, but in this case are more granular at the
+individual organism level rather than at the systemic functional level.
 
@@
 to the dual cutoff approach that is advocated for many gene expression
-experiments (14, 16). In particular, the dual cutoff approach has long
-been known to not control for Type 1 errors (17, 18), and the frequent
-lack of concordance between tools when benchmarked on transcriptomes(10,
-12--14, 29, 56) and microbiomes (9, 11, 31, 40, 57, 58) suggests poor
-control of Type 2 errors as well (10, 13). Thus, incorporating a scale
+experiments (9, 11). In particular, the dual cutoff approach has long
+been known to not control for Type 1 errors (12, 13), and the frequent
+lack of concordance between tools when benchmarked on transcriptomes (5,
+7--9, 17, 67) and microbiomes (4, 6, 32, 42, 68, 69) suggests poor
+control of Type 2 errors as well (5, 8). Thus, incorporating a scale
 model during the analysis of HTS data promises the best of both worlds.
@@
 additional computational overhead. Furthermore, this work and previous
-(20) show that even minimal information about the underlying environment
-can be used to build a relatively robust informed scale model that
-controls for both Type 1 and 2 error rates.
+(22, 43) show that even minimal information about the underlying
+environment can be used to build a relatively robust informed scale
+model that controls for both Type 1 and 2 error rates. It is important
+to note that the approach advocated here is distinct from that suggested
+by Zhang et al. (66, 70) where the DNA amount for a gene is a covariate
+in the model for transcriptomic differential abundance. In our analysis
+we grouped all the transcript information to functional level regardless
+of organism, instead of modeling per-organism gene abundances. In the
+future we anticipate being able to build more complex models similar to
+those used by Zhang et al.~with the additional information of
+uncertainty in the underlying gene count.
 
 In the analysis of HTS data it is often observed that larger datasets
-converge on the majority of parts being significantly different (3, 13,
-14). Li et al. (13) conducted a permutation-based benchmarking study and
+converge on the majority of parts being significantly different (3, 8,
+9). Li et al. (8) conducted a permutation-based benchmarking study and
 found that widely used tools performed worse than simple Wilcoxon
-rank-sum tests in controlling the FDR when sample sizes became large.
-They suggested that the presence of outliers were one of the factors
-driving this observation. Brooks et al. (59) suggested that
-inappropriate choice of benchmarking methods are also a major
-contributing factor and that objective standards of truth are important.
+rank-sum tests coupled with the TPM normalization in controlling the FDR
+when sample sizes became large. Li et al.~suggested that the presence of
+outliers were one of the factors driving the extreme FDR in some tests.
+We found that when the Wilcoxon test was used within the ALDEx2
+framework that it had essentially the same outputs as did the t-test.
+For example, in the PD-1 dataset where \(\gamma = 0\) the ALDEx2 t-test
+exhibited a mean FDR of 0.2\% and mean sensitivity of 65.9\% while the
+corresponding values from the ALDEx2 Wilcoxon test were 0.3\% and
+68.9\%. This result again supports that the assumptions of the
+normalizations are as important or more important than the statistical
+test. Brooks et al. (71) suggested that inappropriate choice of
+benchmarking methods are also a major contributing factor and that
+better objective standards of truth are needed. In this report we
+generated semi-synthetic test data used binomial thinning which produces
+data that more closely mimic the properties of real high throughput
+sequencing data, and so can more rigorously test different tools (54).
 From the perspective of our work the disagreement between tools can be
 explained by the observation that different analytic approaches produce
-different parameter estimates for location or scale or for both. Thus,
-more data produces worse estimates because the additional data simply
-increases the precision of a flawed estimate (3, 60).
-
-Scale simulation is now built into ALDEx2 (20) and here we suggest that
-there are two main root causes to common HTS data pathologies. The first
-contributing factor is the observed very low dispersion estimate for
-many features that is a by-product of some experimental designs and of
-normalization. In the Schurch et al. (14 dataset), the data were from
-single colonies derived from a single culture. Thus, it is more accurate
-to describe the 96 samples as wet-lab technical replicates rather than
-independent samples. This type of replication approach is standard in
-the molecular literature, and would be expected to result in the very
-low dispersion that is observed. Applying the default scale model with
-\(\gamma=0.5\) a large number of transcripts have their dispersion
-increased (Figure 1D), with the effect being largest for those with the
-lowest initial dispersion (Figure 2). Adding scale uncertainty results
-in modest number of transcripts, 205, being called significantly
-different as shown in the volcano plot in Figure 1B (red points). In
-addition, there is now a strong concordance between the difference and
-q-values. In hindsight, it is not obvious why the unscaled volcano plot
-shows such poor correspondence. We suggest that this is explained by
-random fluctuations of the many very low variance estimates and this is
-supported by the plots shown in Figure 2.
-
-The second contributing factor is unacknowledged asymmetry in many
-datasets (19); i.e., different gene content or a directional change in
-the majority of features. In the case of asymmetry, the use of a
-user-specified scale model can be very useful for otherwise
-difficult-to-analyze datasets such as meta-transcriptomes and in-vitro
-selection datasets where the majority of features can change. We showed
-one such example for the metatranscriptome dataset in Figure 3. Here the
-dataset was highly asymmetrical. Incorporating differential scale on a
-per-group basis moves the mass of the housekeeping functions towards the
-midline of no difference and so affects both Type I and Type II error
-rates. We showed two ways of estimating the scale difference between
-groups and found that any reasonable estimate is an improvement over the
-naive approach and also over the default scale model. This is in line
-with the observations by Nixon et al (20) in a 16S rRNA gene sequencing
-dataset. It is also of note that in the case of true biological
-replicates (different individuals) that adding a modest amount of scale
-\(\gamma=0.5\) had little effect on the the difference between groups
-and on the dispersion. Thus, in this dataset the scale mis-specification
-was affecting mainly the location of the difference between groups.
-While we acknowledge that some prior information on which housekeeping
-transcripts should not be classed as differentially abundant is needed,
-we suggest that this information is widely available and is already used
-when performing the gold-standard quantitative PCR test of differential
-abundance (61, 62).
+different parameter estimates for either location or scale, or for both,
+as suggested in Figure 1. Thus, more data produces worse estimates
+because the additional data simply increases the precision of a flawed
+estimate (3, 72).
+
+Scale simulation is now built into ALDEx2 (22) and in this report we
+suggest that there are two main root causes to common HTS data
+pathologies. The first contributing factor is the observed very low
+dispersion estimate for many features that is a by-product of some
+experimental designs and normalizations (Figure 2). Supplying additional
+uncertainty alleviates many FP but in a way that more appropriately
+controls the FDR as shown in Figure 3. The second contributing factor is
+unacknowledged asymmetry in many datasets (15); i.e., different gene
+content or a directional change in the majority of features. In the case
+of asymmetry, the use of a user-specified scale model can be very useful
+for otherwise difficult-to-analyze datasets such as meta-transcriptomes
+and in-vitro selection datasets where the majority of features can
+change as shown in Figure 4. We showed two ways of estimating the scale
+difference between groups and found that any reasonable estimate is an
+improvement over the naive approach and also over the default scale
+model. This is in line with the observations by Nixon et al (22) in a
+16S rRNA gene sequencing dataset. While we acknowledge that some prior
+information is needed that this information is widely available and is
+already used when performing the gold-standard quantitative PCR test of
+differential abundance (73, 74).
 
 Beyond concerns of fidelity and rigor, scale models also enhance the
-reproducibility and transparency of HTS analyses. The addition of scale
-uncertainty essentially tests the model over a range of normalizations
-(3) and so can replace the consensus approach that has been proposed by
-some groups (11, 63) with no additional computational overhead. Thus, an
-advantage of incorporating scale is that analyses can be made much more
-robust such that actual or potential differences in scale can be tested
-and accounted for explicitly. While it is beyond the scope of the
-present article, we note that there are many ways of building scale
-models that enhance the interpretability of the parameters and
-assumptions and a detailed description of these points is describe
-elsewhere (3,).
+reproducibility and transparency of HTS analyses. The development of HTS
+and the associated problems of very high dimensional data that was not
+always statistically well-behaved led to many different proposed
+solutions including multiple normalizations and moderated test
+statistics. That these perform poorly in real data is shown by the
+simulation data in this report and elsewhere where both DESeq2 (which we
+used) and edgeR (which uses moderated statistical tests) performed
+poorly in controlling the FDR (8). While these approaches often work in
+many datasets they fail to address the underlying problems of
+information that is missing in the data which is what is supplied by
+adding uncertainty around the information we have about that data. The
+addition of scale uncertainty directly addresses the missing information
+by testing the model over a range of normalizations (3). In doing so,
+the scale-based approach removes the need for moderated statistics and
+can replace the consensus approach that has been proposed by some groups
+(6, 75) with no additional computational overhead. Thus, an advantage of
+incorporating scale is that analyses can be made much more robust such
+that actual or potential differences in scale can be tested and
+accounted for explicitly. While it is beyond the scope of the present
+article, we note that there are many ways of building scale models that
+enhance the interpretability of the parameters and assumptions and a
+detailed description of these points is describe elsewhere (3).
 
@@
 transcriptomes or indeed any type of HTS dataset. While the underlying
-scale of the system is generally inaccessible, the effect of scale on
-the analysis outcomes can be modelled and can help explain some of the
-underlying biology, and help to expose known issues with the analysis of
-HTS data. Adding scale information to the analysis allows for more
-robust inference because the features that are sensitive to scale can be
-identified and their impact on conclusions weighted accordingly.
-Additionally, the use of informed scale models permits difficult to
-analyze datasets to be examined in a robust and principled manner even
-when the majority of features are asymmetrically distributed or
-expressed (or both) in the groups (50). Thus, using and reporting scale
+scale of the system is generally inaccessible, the effect of scale
+uncertainty on the analysis outcomes can be modelled and can help
+explain some of the underlying biology. Adding scale information to the
+analysis allows for more robust inference because the features that are
+sensitive to scale can be identified and their impact on conclusions
+weighted accordingly. The use of informed scale models permits difficult
+to analyze datasets to be examined in a robust and principled manner
+even when the majority of features are asymmetrically distributed or
+expressed (or both) in the groups (55). Thus, using and reporting scale
 uncertainty should become a standard practice in the analysis of HTS
@@
 
-\bibitem[\citeproctext]{ref-Jiang:2011aa}
-4. Jiang,L., Schlesinger,F., Davis,C.A., Zhang,Y., Li,R., Salit,M.,
-Gingeras,T.R. and Oliver,B. (2011)
-\href{https://doi.org/10.1101/gr.121095.111}{Synthetic spike-in
-standards for RNA-seq experiments}. \emph{Genome Res}, \textbf{21},
-1543--51.
-
-\bibitem[\citeproctext]{ref-Loven:2012aa}
-5. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B., Burge,C.B.,
-Levens,D.L., Lee,T.I. and Young,R.A. (2012)
-\href{https://doi.org/10.1016/j.cell.2012.10.012}{Revisiting global gene
-expression analysis}. \emph{Cell}, \textbf{151}, 476--82.
-
-\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
-6. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
-Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
-Darzi,Y., \emph{et al.} (2017)
-\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
-profiling links gut community variation to microbial load}.
-\emph{Nature}, \textbf{551}, 507--511.
-
-\bibitem[\citeproctext]{ref-Props:2017aa}
-7. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
-Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
-\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
-microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.
-
-\bibitem[\citeproctext]{ref-Nishijima:2024aa}
-8. Nishijima,S., Stankevic,E., Aasmets,O., Schmidt,T.S.B., Nagata,N.,
-Keller,M.I., Ferretti,P., Juel,H.B., Fullam,A., Robbani,S.M., \emph{et
-al.} (2024) Fecal microbial load is a major determinant of gut
-microbiome variation and a confounder for disease associations.
-\emph{Cell},
-\href{https://doi.org/10.1016/j.cell.2024.10.022}{10.1016/j.cell.2024.10.022}.
-
 \bibitem[\citeproctext]{ref-Thorsen:2016aa}
-9. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
+4. Thorsen,J., Brejnrod,A., Mortensen,M., Rasmussen,M.A., Stokholm,J.,
 Al-Soud,W.A., Sørensen,S., Bisgaard,H. and Waage,J. (2016)
@@
 \bibitem[\citeproctext]{ref-Quinn:2018aa}
-10. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
+5. Quinn,T.P., Crowley,T.M. and Richardson,M.F. (2018)
 \href{https://doi.org/10.1186/s12859-018-2261-8}{Benchmarking
@@
 \bibitem[\citeproctext]{ref-Nearing:2022aa}
-11. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
+6. Nearing,J.T., Douglas,G.M., Hayes,M.G., MacDonald,J., Desai,D.K.,
 Allward,N., Jones,C.M.A., Wright,R.J., Dhanani,A.S., Comeau,A.M.,
@@
 \bibitem[\citeproctext]{ref-Ge:2021aa}
-12. Ge,X., Chen,Y.E., Song,D., McDermott,M., Woyshner,K.,
+7. Ge,X., Chen,Y.E., Song,D., McDermott,M., Woyshner,K.,
 Manousopoulou,A., Wang,N., Li,W., Wang,L.D. and Li,J.J. (2021)
@@
 \bibitem[\citeproctext]{ref-Li:2022aa}
-13. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+8. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
 \href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
@@
 \bibitem[\citeproctext]{ref-Schurch:2016aa}
-14. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
+9. Schurch,N.J., Schofield,P., Gierliński,M., Cole,C., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G.G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-storey2003positive}
-15. Storey,J.D. (2003) The positive false discovery rate: A bayesian
+10. Storey,J.D. (2003) The positive false discovery rate: A bayesian
 interpretation and the q-value. \emph{The annals of statistics},
@@
 \bibitem[\citeproctext]{ref-Cui:2003aa}
-16. Cui,X. and Churchill,G.A. (2003)
+11. Cui,X. and Churchill,G.A. (2003)
 \href{https://www.ncbi.nlm.nih.gov/pubmed/12702200}{Statistical tests
@@
 \bibitem[\citeproctext]{ref-Zhang:2009aa}
-17. Zhang,S. and Cao,J. (2009)
+12. Zhang,S. and Cao,J. (2009)
 \href{https://doi.org/10.1186/1471-2105-10-402}{A close examination of
@@
 \bibitem[\citeproctext]{ref-Ebrahimpoor:2021aa}
-18. Ebrahimpoor,M. and Goeman,J.J. (2021)
+13. Ebrahimpoor,M. and Goeman,J.J. (2021)
 \href{https://doi.org/10.1093/bib/bbab053}{Inflated false discovery rate
@@
 
+\bibitem[\citeproctext]{ref-simmons2011false}
+14. Simmons,J.P., Nelson,L.D. and Simonsohn,U. (2011) False-positive
+psychology: Undisclosed flexibility in data collection and analysis
+allows presenting anything as significant. \emph{Psychological science},
+\textbf{22}, 1359--1366.
+
 \bibitem[\citeproctext]{ref-Wu2021}
-19. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
+15. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
 \href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
@@
 
+\bibitem[\citeproctext]{ref-Nishijima:2024aa}
+16. Nishijima,S., Stankevic,E., Aasmets,O., Schmidt,T.S.B., Nagata,N.,
+Keller,M.I., Ferretti,P., Juel,H.B., Fullam,A., Robbani,S.M., \emph{et
+al.} (2024) Fecal microbial load is a major determinant of gut
+microbiome variation and a confounder for disease associations.
+\emph{Cell},
+\href{https://doi.org/10.1016/j.cell.2024.10.022}{10.1016/j.cell.2024.10.022}.
+
+\bibitem[\citeproctext]{ref-Bullard:2010}
+17. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
+\href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
+methods for normalization and differential expression in m{RNA-seq}
+experiments}. \emph{BMC Bioinformatics}, \textbf{11}, 94.
+
+\bibitem[\citeproctext]{ref-Jiang:2011aa}
+18. Jiang,L., Schlesinger,F., Davis,C.A., Zhang,Y., Li,R., Salit,M.,
+Gingeras,T.R. and Oliver,B. (2011)
+\href{https://doi.org/10.1101/gr.121095.111}{Synthetic spike-in
+standards for RNA-seq experiments}. \emph{Genome Res}, \textbf{21},
+1543--51.
+
+\bibitem[\citeproctext]{ref-Loven:2012aa}
+19. Lovén,J., Orlando,D.A., Sigova,A.A., Lin,C.Y., Rahl,P.B.,
+Burge,C.B., Levens,D.L., Lee,T.I. and Young,R.A. (2012)
+\href{https://doi.org/10.1016/j.cell.2012.10.012}{Revisiting global gene
+expression analysis}. \emph{Cell}, \textbf{151}, 476--82.
+
+\bibitem[\citeproctext]{ref-Vandeputte:2017aa}
+20. Vandeputte,D., Kathagen,G., D'hoe,K., Vieira-Silva,S.,
+Valles-Colomer,M., Sabino,J., Wang,J., Tito,R.Y., De Commer,L.,
+Darzi,Y., \emph{et al.} (2017)
+\href{https://doi.org/10.1038/nature24460}{Quantitative microbiome
+profiling links gut community variation to microbial load}.
+\emph{Nature}, \textbf{551}, 507--511.
+
+\bibitem[\citeproctext]{ref-Props:2017aa}
+21. Props,R., Kerckhof,F.-M., Rubbens,P., De Vrieze,J., Hernandez
+Sanabria,E., Waegeman,W., Monsieurs,P., Hammes,F. and Boon,N. (2017)
+\href{https://doi.org/10.1038/ismej.2016.117}{Absolute quantification of
+microbial taxon abundances}. \emph{ISME J}, \textbf{11}, 584--587.
+
 \bibitem[\citeproctext]{ref-Nixon2024B}
-20. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
+22. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
 normalization: Incorporating scale uncertainty in microbiome and gene
@@
 \bibitem[\citeproctext]{ref-Lovell:2015}
-21. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
+23. Lovell,D., Pawlowsky-Glahn,V., Egozcue,J.J., Marguerat,S. and
 Bähler,J. (2015)
@@
 \bibitem[\citeproctext]{ref-Nie:2012aa}
-22. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
+24. Nie,Z., Hu,G., Wei,G., Cui,K., Yamane,A., Resch,W., Wang,R.,
 Green,D.R., Tessarollo,L., Casellas,R., \emph{et al.} (2012)
@@
 \bibitem[\citeproctext]{ref-Scott:2010}
-23. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
+25. Scott,M., Gunderson,C.W., Mateescu,E.M., Zhang,Z. and Hwa,T. (2010)
 \href{https://doi.org/10.1126/science.1192588}{Interdependence of cell
@@
 \bibitem[\citeproctext]{ref-Yoshikawa:2011aa}
-24. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
+26. Yoshikawa,K., Tanaka,T., Ida,Y., Furusawa,C., Hirasawa,T. and
 Shimizu,H. (2011) \href{https://doi.org/10.1002/yea.1843}{Comprehensive
@@
 \bibitem[\citeproctext]{ref-Lin:2018aa}
-25. Lin,J. and Amir,A. (2018)
+27. Lin,J. and Amir,A. (2018)
 \href{https://doi.org/10.1038/s41467-018-06714-z}{Homeostasis of protein
@@
 \bibitem[\citeproctext]{ref-Zozaya:2010}
-26. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
+28. Zozaya-Hinchliffe,M., Lillis,R., Martin,D.H. and Ferris,M.J. (2010)
 \href{https://doi.org/10.1128/JCM.00851-09}{Quantitative PCR assessments
@@
 \bibitem[\citeproctext]{ref-Ravel:2010}
-27. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
+29. Ravel,J., Gajer,P., Abdo,Z., Schneider,G.M., Koenig,S.S.K.,
 McCulle,S.L., Karlebach,S., Gorle,R., Russell,J., Tacket,C.O., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hummelen:2010}
-28. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
+30. Hummelen,R., Fernandes,A.D., Macklaim,J.M., Dickson,R.J.,
 Changalucha,J., Gloor,G.B. and Reid,G. (2010)
@@
 
-\bibitem[\citeproctext]{ref-Bullard:2010}
-29. Bullard,J.H., Purdom,E., Hansen,K.D. and Dudoit,S. (2010)
-\href{https://doi.org/10.1186/1471-2105-11-94}{Evaluation of statistical
-methods for normalization and differential expression in m{RNA-seq}
-experiments}. \emph{BMC Bioinformatics}, \textbf{11}, 94.
-
 \bibitem[\citeproctext]{ref-Dillies:2013}
-30. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
+31. Dillies,M.-A., Rau,A., Aubert,J., Hennequet-Antier,C.,
 Jeanmougin,M., Servant,N., Keime,C., Marot,G., Castel,D., Estelle,J.,
@@
 \bibitem[\citeproctext]{ref-Weiss:2017aa}
-31. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
+32. Weiss,S., Xu,Z.Z., Peddada,S., Amir,A., Bittinger,K., Gonzalez,A.,
 Lozupone,C., Zaneveld,J.R., Vázquez-Baeza,Y., Birmingham,A., \emph{et
@@
 \bibitem[\citeproctext]{ref-Hughes:2005tu}
-32. Hughes,J.B. and Hellmann,J.J. (2005)
+33. Hughes,J.B. and Hellmann,J.J. (2005)
 \href{https://doi.org/10.1016/S0076-6879(05)97017-1}{The application of
@@
 \bibitem[\citeproctext]{ref-Mortazavi:2008}
-33. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
+34. Mortazavi,A., Williams,B.A., McCue,K., Schaeffer,L. and Wold,B.
 (2008) \href{https://doi.org/10.1038/nmeth.1226}{Mapping and quantifying
@@
 \bibitem[\citeproctext]{ref-wagner:tpm}
-34. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
+35. Wagner,G.P., Kin,K. and Lynch,V.J. (2012)
 \href{https://doi.org/10.1007/s12064-012-0162-3}{Measurement of mRNA
@@
 \bibitem[\citeproctext]{ref-Robinson:2010a}
-35. Robinson,M.D. and Oshlack,A. (2010)
+36. Robinson,M.D. and Oshlack,A. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-3-r25}{A scaling normalization
@@
 
+\bibitem[\citeproctext]{ref-Vandesompele:2002aa}
+37. Vandesompele,J., De Preter,K., Pattyn,F., Poppe,B., Van Roy,N., De
+Paepe,A. and Speleman,F. (2002)
+\href{https://www.ncbi.nlm.nih.gov/pubmed/12184808}{Accurate
+normalization of real-time quantitative RT-PCR data by geometric
+averaging of multiple internal control genes}. \emph{Genome Biol},
+\textbf{3}, RESEARCH0034.
+
 \bibitem[\citeproctext]{ref-aitchison1982}
-36. Aitchison,J. (1982) The statistical analysis of compositional data.
+38. Aitchison,J. (1982) The statistical analysis of compositional data.
 \emph{Journal of the Royal Statistical Society: Series B
@@
 \bibitem[\citeproctext]{ref-Anders:2010}
-37. Anders,S. and Huber,W. (2010)
+39. Anders,S. and Huber,W. (2010)
 \href{https://doi.org/10.1186/gb-2010-11-10-r106}{Differential
@@
 \bibitem[\citeproctext]{ref-Aitchison:1986}
-38. Aitchison,J. (1986) The statistical analysis of compositional data
+40. Aitchison,J. (1986) The statistical analysis of compositional data
 Chapman \& Hall, London, England.
@@
 \bibitem[\citeproctext]{ref-fernandes:2013}
-39. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
-(2013) \href{https://doi.org/10.1371/journal.pone.0067019}{ANOVA-like
-differential expression (ALDEx) analysis for mixed population RNA-seq}.
-\emph{PLoS One}, \textbf{8}, e67019.
+41. Fernandes,A.D., Macklaim,J.M., Linn,T.G., Reid,G. and Gloor,G.B.
+(2013) \href{https://doi.org/10.1371/journal.pone.0067019}{{ANOVA}-like
+differential expression (ALDEx) analysis for mixed population
+{RNA}-seq}. \emph{PLoS One}, \textbf{8}, e67019.
 
 \bibitem[\citeproctext]{ref-Yerke:2024aa}
-40. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
+42. Yerke,A., Fry Brumit,D. and Fodor,A.A. (2024)
 \href{https://doi.org/10.1186/s40168-023-01747-z}{Proportion-based
@@
 
+\bibitem[\citeproctext]{ref-Dos-Santos:2024aa}
+43. Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and Gloor,G.B.
+(2024) \href{https://doi.org/10.1186/s40168-024-01992-w}{Vaginal
+metatranscriptome meta-analysis reveals functional {BV} subgroups and
+novel colonisation strategies}. \emph{Microbiome}, \textbf{12}, 271.
+
 \bibitem[\citeproctext]{ref-Love:2014aa}
-41. Love,M.I., Huber,W. and Anders,S. (2014)
+44. Love,M.I., Huber,W. and Anders,S. (2014)
 \href{https://doi.org/10.1186/s13059-014-0550-8}{Moderated estimation of
@@
 \bibitem[\citeproctext]{ref-Paulson:2013aa}
-42. Paulson,J.N., Stine,O.C., Bravo,H.C. and Pop,M. (2013)
+45. Paulson,J.N., Stine,O.C., Bravo,H.C. and Pop,M. (2013)
 \href{https://doi.org/10.1038/nmeth.2658}{Differential abundance
@@
 \bibitem[\citeproctext]{ref-fernandes:2014}
-43. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
+46. Fernandes,A.D., Reid,J.N., Macklaim,J.M., McMurrough,T.A.,
 Edgell,D.R. and Gloor,G.B. (2014)
@@
 \bibitem[\citeproctext]{ref-Gierlinski:2015aa}
-44. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
+47. Gierliński,M., Cole,C., Schofield,P., Schurch,N.J., Sherstnev,A.,
 Singh,V., Wrobel,N., Gharbi,K., Simpson,G., Owen-Hughes,T., \emph{et
@@
 \bibitem[\citeproctext]{ref-benjamini:1995}
-45. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
+48. Benjamini,Y. and Hochberg,Y. (1995) Controlling the false discovery
 rate: A practical and powerful approach to multiple testing.
@@
 \bibitem[\citeproctext]{ref-efron2008FDR}
-46. Efron,B. (2008) Microarrays, empirical bayes and the two-groups
+49. Efron,B. (2008) Microarrays, empirical bayes and the two-groups
 model. \emph{Statist. Sci.}, \textbf{23}, 1--22.
@@
 \bibitem[\citeproctext]{ref-gloor:effect}
-47. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
+50. Gloor,G., Macklaim,J. and Fernandes,A. (2016)
 \href{https://doi.org/10.1080/10618600.2015.1131161}{Displaying
@@
 
-\bibitem[\citeproctext]{ref-Rocha:2020aa}
-48. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
-(2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
-expression analysis in bacteria by RT-qPCR}. \emph{Methods Mol Biol},
-\textbf{2065}, 119--137.
-
-\bibitem[\citeproctext]{ref-Taniguchi:2010aa}
-49. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
-Emili,A. and Xie,X.S. (2010)
-\href{https://doi.org/10.1126/science.1188308}{Quantifying e. Coli
-proteome and transcriptome with single-molecule sensitivity in single
-cells}. \emph{Science}, \textbf{329}, 533--8.
+\bibitem[\citeproctext]{ref-witten2007comparison}
+51. Witten,D. and Tibshirani,R. (2007) A comparison of fold-change and
+the t-statistic for microarray data analysis. \emph{Analysis},
+\textbf{1776}, 58--85.
+
+\bibitem[\citeproctext]{ref-Liu:2022aa}
+52. Liu,X., Zhao,J., Xue,L., Zhao,T., Ding,W., Han,Y. and Ye,H. (2022)
+\href{https://doi.org/10.1186/s12864-022-08465-0}{A comparison of
+transcriptome analysis methods with reference genome}. \emph{BMC
+Genomics}, \textbf{23}, 232.
+
+\bibitem[\citeproctext]{ref-Riaz:2017aa}
+53. Riaz,N., Havel,J.J., Makarov,V., Desrichard,A., Urba,W.J.,
+Sims,J.S., Hodi,F.S., Martín-Algarra,S., Mandal,R., Sharfman,W.H.,
+\emph{et al.} (2017)
+\href{https://doi.org/10.1016/j.cell.2017.09.028}{Tumor and
+microenvironment evolution during immunotherapy with nivolumab}.
+\emph{Cell}, \textbf{171}, 934--949.e16.
+
+\bibitem[\citeproctext]{ref-Gerard:2020aa}
+54. Gerard,D. (2020)
+\href{https://doi.org/10.1186/s12859-020-3450-9}{Data-based RNA-seq
+simulations by binomial thinning}. \emph{BMC Bioinformatics},
+\textbf{21}, 206.
 
 \bibitem[\citeproctext]{ref-dosSantos:2024}
-50. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
+55. Dos Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and
 Gloor,G.B. (2024) Vaginal metatranscriptome meta-analysis reveals
@@
 
-\bibitem[\citeproctext]{ref-macklaim:2013}
-51. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
-and Gloor,G.B. (2013)
-\href{https://doi.org/10.1186/2049-2618-1-12}{Comparative meta-{RNA}-seq
-of the vaginal microbiota and differential expression by lactobacillus
-iners in health and dysbiosis}. \emph{Microbiome}, \textbf{1}, 12.
+\bibitem[\citeproctext]{ref-Macklaim:2018aa}
+56. Macklaim,J.M. and Gloor,G.B. (2018)
+\href{https://doi.org/10.1007/978-1-4939-8728-3/_13}{From {RNA}-seq to
+biological inference: Using compositional data analysis in
+meta-transcriptomics}. \emph{Methods Mol Biol}, \textbf{1849}, 193--213.
 
 \bibitem[\citeproctext]{ref-Denge00262-18}
-52. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
+57. Deng,Z.-L., Gottschick,C., Bhuju,S., Masur,C., Abels,C. and
 Wagner-Döbler,I. (2018)
@@
 
+\bibitem[\citeproctext]{ref-Zhang:2020ab}
+58. Zhang,Y., Parmigiani,G. and Johnson,W.E. (2020)
+\href{https://doi.org/10.1093/nargab/lqaa078}{ComBat-seq: Batch effect
+adjustment for RNA-seq count data}. \emph{NAR Genom Bioinform},
+\textbf{2}, lqaa078.
+
 \bibitem[\citeproctext]{ref-Fettweis:2019aa}
-53. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
+59. Fettweis,J.M., Serrano,M.G., Brooks,J.P., Edwards,D.J., Girerd,P.H.,
 Parikh,H.I., Huang,B., Arodz,T.J., Edupuganti,L., Glascock,A.L.,
@@
 
+\bibitem[\citeproctext]{ref-macklaim:2013}
+60. Macklaim,J.M., Fernandes,A.D., Di Bella,J.M., Hammond,J.-A., Reid,G.
+and Gloor,G.B. (2013)
+\href{https://doi.org/10.1186/2049-2618-1-12}{Comparative meta-{RNA}-seq
+of the vaginal microbiota and differential expression by
+\emph{{L}actobacillus iners} in health and dysbiosis}.
+\emph{Microbiome}, \textbf{1}, 12.
+
 \bibitem[\citeproctext]{ref-Okuda:2008}
-54. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
+61. Okuda,S., Yamada,T., Hamajima,M., Itoh,M., Katayama,T., Bork,P.,
 Goto,S. and Kanehisa,M. (2008)
@@
 
+\bibitem[\citeproctext]{ref-Rocha:2020aa}
+62. Rocha,D.J.P.G., Castro,T.L.P., Aguiar,E.R.G.R. and Pacheco,L.G.C.
+(2020) \href{https://doi.org/10.1007/978-1-4939-9833-3_10}{Gene
+expression analysis in bacteria by RT-qPCR}. \emph{Methods Mol Biol},
+\textbf{2065}, 119--137.
+
+\bibitem[\citeproctext]{ref-Taniguchi:2010aa}
+63. Taniguchi,Y., Choi,P.J., Li,G.-W., Chen,H., Babu,M., Hearn,J.,
+Emili,A. and Xie,X.S. (2010)
+\href{https://doi.org/10.1126/science.1188308}{Quantifying e. Coli
+proteome and transcriptome with single-molecule sensitivity in single
+cells}. \emph{Science}, \textbf{329}, 533--8.
+
 \bibitem[\citeproctext]{ref-yeast-absolute}
-55. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
+64. Marguerat,S., Schmidt,A., Codlin,S., Chen,W., Aebersold,R. and
 Bähler,J. (2012)
@@
 
+\bibitem[\citeproctext]{ref-McGovern:2023}
+65. McGovern,K.C., Nixon,M.P. and Silverman,J.D. (2023)
+\href{https://doi.org/10.1371/journal.pcbi.1011659}{Addressing erroneous
+scale assumptions in microbe and gene set enrichment analysis}.
+\emph{PLoS Comput Biol}, \textbf{19}, e1011659.
+
+\bibitem[\citeproctext]{ref-Zhang:2021ab}
+66. Zhang,Y., Thompson,K.N., Huttenhower,C. and Franzosa,E.A. (2021)
+\href{https://doi.org/10.1093/bioinformatics/btab327}{Statistical
+approaches for differential expression analysis in metatranscriptomics}.
+\emph{Bioinformatics}, \textbf{37}, i34--i41.
+
 \bibitem[\citeproctext]{ref-Soneson:2013}
-56. Soneson,C. and Delorenzi,M. (2013)
+67. Soneson,C. and Delorenzi,M. (2013)
 \href{https://doi.org/10.1186/1471-2105-14-91}{A comparison of methods
@@
 \bibitem[\citeproctext]{ref-McMurdie:2014a}
-57. McMurdie,P.J. and Holmes,S. (2014)
+68. McMurdie,P.J. and Holmes,S. (2014)
 \href{https://doi.org/10.1371/journal.pcbi.1003531}{Waste not, want not:
@@
 \bibitem[\citeproctext]{ref-hawinkel2017}
-58. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
+69. Hawinkel,S., Mattiello,F., Bijnens,L. and Thas,O. (2018)
 \href{http://dx.doi.org/10.1093/bib/bbx104}{A broken promise :
@@
 
+\bibitem[\citeproctext]{ref-Zhang:2021aa}
+70. Zhang,Y., Thompson,K.N., Branck,T., Yan,Y., Nguyen,L.H.,
+Franzosa,E.A. and Huttenhower,C. (2021)
+\href{https://doi.org/10.1146/annurev-biodatasci-031121-103035}{Metatranscriptomics
+for the human microbiome and microbial community functional profiling}.
+\emph{Annu Rev Biomed Data Sci}, \textbf{4}, 279--311.
+
 \bibitem[\citeproctext]{ref-Brooks:2024aa}
-59. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
+71. Brooks,T.G., Lahens,N.F., Mrčela,A. and Grant,G.R. (2024)
 \href{https://doi.org/10.1038/s41576-023-00679-6}{Challenges and best
@@
 \bibitem[\citeproctext]{ref-gustafson2015bayesian}
-60. Gustafson,P. (2015) Bayesian inference for partially identified
+72. Gustafson,P. (2015) Bayesian inference for partially identified
 models: Exploring the limits of limited data CRC Press.
@@
 \bibitem[\citeproctext]{ref-Thellin:1999aa}
-61. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
+73. Thellin,O., Zorzi,W., Lakaye,B., De Borman,B., Coumans,B.,
 Hennen,G., Grisar,T., Igout,A. and Heinen,E. (1999)
@@
 \bibitem[\citeproctext]{ref-SEQCux2fMAQC-III-Consortium:2014aa}
-62. SEQC/MAQC-III Consortium (2014)
+74. SEQC/MAQC-III Consortium (2014)
 \href{https://doi.org/10.1038/nbt.2957}{A comprehensive assessment of
@@
 \bibitem[\citeproctext]{ref-Song:2023aa}
-63. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
+75. Song,H., Ling,W., Zhao,N., Plantinga,A.M., Broedlow,C.A.,
 Klatt,N.R., Hensley-McBain,T. and Wu,M.C. (2023)

### 2025-04-14 information-supplement.Rmd
Penultimate before cleaning

@@
 t.out <- matrix(data=NA, nrow=12, ncol=7)
-colnames(t.out) <- c("h1", "h2", "H1",H2","mh", 'mH','prior')
+colnames(t.out) <- c("h1", "h2", "H1", "H2", "mh",'mH','prior')
 prior[1] <- 0.001

### 2025-04-14 supplement_figs.Rmd
Penultimate before cleaning

@@
 ```{r echo=F, warning=F, message=F}
+library(seqgendiff)
 library(ALDEx2)
 data(selex)
+library(DESeq2)
+library(edgeR)
 ```
-```{r echo=FALSE, include=TRUE, out.width="0.8\\linewidth",  fig.align="center", fig.cap=c("The aldex.senAnalysis() function was used to generate a dataset with \\(\\gamma\\) values of 1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7 and 1.
+```{r echo=FALSE, include=TRUE, out.width="0.8\\linewidth",  fig.align="center", fig.cap=c("The aldex.senAnalysis() function was used with the yeast transcriptome dataset to generate outputs with \\(\\gamma\\) values of 1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7 and 1.
 The plotGamma() function was used to plot the result.
@@
 ")}
+
+####
+# Data generated by yst-sens.R
+####
+
 knitr::include_graphics("analysis/scaleSim-yeast.pdf")
@@
 
- One root cause of the large number of significant parts is the very low dispersion of transcripts.
-Figure 1 shows a graphical output from the texttt{aldex.scaleSim()} function with the yeast transcriptome dataset described in the text [@Gierlinski:2015aa;@Schurch:2016aa].
-This allows us to examine which transcripts are sensitive to even minimal amounts of scale uncertainty using \(\gamma = (1e-3, 0.1, 0.2, 0.3, 0.4,0.5, 0.7, 1) \).
-Here it is obvious that even a negligible amount of scale uncertainty removes over half of the transcripts that were formerly significantly different, and all of these had very small effect sizes.
+Supplementary Figure 1 shows that  root cause of the large number of significant parts in this dataset is the very low dispersion of transcripts.
+Here we see  a graphical output from the *aldex.scaleSim()* function with the yeast transcriptome dataset described in the text [@Gierlinski:2015aa;@Schurch:2016aa].
+This allows us to examine which transcripts are sensitive to even minimal amounts of scale uncertainty using \(\gamma = (1e-3, 0.1, 0.2, 0.3, 0.4, 0.5, 0.7, 1) \).
+Adding scale uncertainty increases the minimum dispersion in the analysis.
+Here it is obvious that even a negligible amount of scale uncertainty \(\gamma=0.1\) removes over half of the transcripts that were formerly significantly different, and all of these had very small effect sizes.
 
-We recommend a minimum scale uncertainty of 0.5, and suggest that the texttt{aldex.senAnalysis()} function be run on all analyses.
-The individual \texttt{aldex()} outputs can be accessed as sequential entries in the list output, or the analysis as a whole can be plotted with the \texttt{plotGamma} function.
+We recommend a minimum scale uncertainty of between 0.2 and 0.5, and suggest that the *aldex.senAnalysis()* function be run on all datasets to identify transcripts that are sensitive to small \(\gamma\) values.
+The individual *aldex()* outputs can be accessed as sequential entries in the list output, or the analysis as a whole can be plotted with the *plotGamma* function.
+
+```{r, brca-sens, echo=FALSE, include=TRUE, fig.dim=c(8,8),  fig.align="center", out.width="0.8\\linewidth",  fig.align="center", fig.cap=c('Adding scale to the BRCA dataset reduces false positives in the same way as the PD-1 dataset.')}
+
+source('code/brca.R')
+
+par(mfrow=c(2,2))
+plot(means$coef[ald.row], means$FDR[ald.row], ylim=c(0,0.6), type='b',lwd=3,
+  xlab="modeled minimum difference", ylab="mean FDR", col='blue')
+title("A: FDR", adj=0, line= 0.8) 
+points(means$coef[ald2.row], means$FDR[ald2.row],  type='b', lwd=3, 
+  col='grey')
+points(means$coef[ald5.row], means$FDR[ald5.row],  type='b', lwd=3, 
+  col='grey50')
+points(means$coef[des.row], means$FDR[des.row],  type='b', lwd=3, col='red')
+points(means$coef[des5.row], means$FDR[des5.row],  type='b', lwd=3,
+  col='darkred')
+legend(0,0.6, legend=c('A 0', 'A 0.2', 'A 0.5'), pch=19,
+  col=c('blue', 'grey', 'grey50'), bty="n", cex=1.2) 
+legend(0.25,0.6, legend=c('D 0', 'D 0.5'), pch=19,
+  col=c('red', 'darkred'), bty="n", cex=1.2) 
+rect(xleft=0, xright=0.55,  ybottom=0.4, ytop=0.6)
+
+plot(means$coef[ald.row], means$SEN[ald.row], ylim=c(0,1), lwd=3, type='b',
+  xlab="modeled minimum difference", ylab="mean Sensitivity", col='blue')
+title("B: sensitivity", adj=0, line= 0.8) 
+points(means$coef[ald2.row], means$SEN[ald2.row],  lwd=3, type='b')
+points(means$coef[ald2.row], means$SEN[ald2.row],  lwd=3, type='b', col='grey')
+points(means$coef[ald5.row], means$SEN[ald5.row],  lwd=3, type='b', col='grey50')
+points(means$coef[des.row], means$SEN[des.row],  lwd=3, type='b', col='red')
+points(means$coef[des5.row], means$SEN[des5.row], lwd=3,  type='b', col='darkred')
+
+addmin <- min(brca.data.out[[10]]$ald0$we.eBH[brca.data.out[[10]]$ald0$we.eBH > 0])/10
+
+aldex.plot(brca.data.out[[10]]$ald0, type='volcano')
+title("C: Volcano", adj=0, line= 0.8) 
+points(brca.data.out[[10]]$ald0$diff.btw[which(brca.data.out[[10]]$ald2$we.eBH < 0.05)], 
+  -log10(brca.data.out[[10]]$ald0$we.eBH[which(brca.data.out[[10]]$ald2$we.eBH < 0.05)]+ addmin), cex=0.6, pch=19, 
+  col='orange')
+points(brca.data.out[[10]]$ald0$diff.btw[which(brca.data.out[[10]]$ald5$we.eBH < 0.05)], 
+  -log10(brca.data.out[[10]]$ald0$we.eBH[which(brca.data.out[[10]]$ald5$we.eBH < 0.05)]+ addmin), col='blue', 
+  cex=0.6, pch=19)
+
+abline(v=c(-0.5,0.5), lty=2, col='darkgrey')
+abline(v=c(-1.5,1.5), lty=2, col='darkgrey')
+
+gam0 <- expression(paste(gamma,"=0"))
+gam2 <- expression(paste(gamma,"=0.2"))
+gam5 <- expression(paste(gamma,"=0.5"))
+
+legend(-6, 120, legend=c("NS", gam0, gam2, gam5), pch=19, col=c("black", "red", "orange", "blue"), cex=1.2)
+
+aldex.plot(brca.data.out[[10]]$ald0, type='MW')
+title("D: effect", adj=0, line= 0.8) 
+points(brca.data.out[[10]]$ald0$diff.win[which(brca.data.out[[10]]$ald2$we.eBH < 0.05)], 
+  (brca.data.out[[10]]$ald0$diff.btw[which(brca.data.out[[10]]$ald2$we.eBH < 0.05)]), cex=0.6, pch=19, 
+  col='orange')
+points(brca.data.out[[10]]$ald0$diff.win[which(brca.data.out[[10]]$ald5$we.eBH < 0.05)], 
+  (brca.data.out[[10]]$ald0$diff.btw[which(brca.data.out[[10]]$ald5$we.eBH < 0.05)]), col='blue', 
+  cex=0.6, pch=19)
+
+abline(v=c(-0.5,0.5), lty=2, col='darkgrey')
+abline(v=c(-1.5,1.5), lty=2, col='darkgrey')
+
+```
+
+<!-- # this is no longer useful
 
 ```{r, echo=F, warning=F, message=F,comment=F, result=F} 
-library(DESeq2)
-library(ALDEx2)
 #ALDEx2
@@
 ```
-
 Figure 2 compares the significant transcripts using effect and volcano plots for the first two \(\gamma\) parameters.
@@
 ```
+companion to Li et al 2022 
+https://rpubs.com/LiYumei/806213
+Running this vs. the test data with ALDEx2 gives ~same Wilcoxon results 
+they convert to cpm, I used the actual data
+adding scale removed almost all significant features but the base result is 
+similar
+cor is ~0.97
+-->
+
+Supplementary Figure 2 shows an  analysis of synthetic data to identify the FDR and sensitivity with a standard of truth in modeled data.
+This second dataset is the BRCA1 dataset from Li et al.
+[@Li:2022aa] which is  a transcriptome analysis of breast cancer and matched tissue controls from The Cancer Genome Atlas collection [@CGA:2013aa] with the count tables collected from the Li et al.
+supplementary dataset.
+Transcripts that were differentially abundant at different gamma values are plotted.
+The first panel in the top row shows FDR at different \(\gamma\) (for ALDEx2) and fold-change cutoffs (for DESeq2) and the second panel in the top row does the same for sensitivity.
+It is noteworthy that in this dataset the FDR for DESeq2 is still very high and that ALDEx2 behaves similarly to the PD-1 dataset.
+The sensitivity for both tools is much higher and adding a small amount of scale uncertainty strongly controls the FDR while having minimal impact on sensitivity.
+The panels in the  bottom row show volcano and effect plots of the outputs for ALDEx2 and recapitulate the results of Figure 3 in the text.
+
+```{r brca_data, echo=FALSE, include=TRUE, fig.dim=c(8,5),  fig.align="center", out.width="1\\linewidth",  fig.align="center", fig.cap=c('Adding scale reduces positive identifications  in biological replicate experiments; data from the Li et al.
+2022.
+The sensitivity analysis in the plot on the left shows that many transcripts are excluded, even when the gamma value is very low.
+The colors are the same as in Supplementary Figure 1.
+The effect plot on the left shows that those transcripts with low difference between and low dispersion are not identified when scale is included (blue) vs.
+when scale is not included (blue or red).
+The middle plot shows that those transcripts where the difference in dispersion between the scaled and unscaled analysis is largest are most affected by adding scale.
+This plot and the next, color transcripts identified as signficant (FDR less than 0.05) when scale is included (blue) vs.
+when scale is not included (blue and red).
+The right-hand histogram shows that the centre of the data is not on 0 fold-change between groups, and so a full scale model would likely be useful.')}
+
+##### BRCA data from Li 2022
+# an example of a bio-replicated dataset
+
+# brca <- read.table('data-Li2022/permuted 
+# datasets_TCGA/TCGA-BRCA.normal-tumor.pair.rawCount.tsv', header=T, row.names=1, 
+# sep='\t')
+
+# brca.conds <- as.vector(unlist(read.table('data-Li2022/permuted 
+# datasets_TCGA/TCGA-BRCA.conditions.tsv', sep='\t')))
+
+# filter using edgeR 
+# from 60483 to 21813 features
+# library(edgeR)
+# 
+#  y <- DGEList(counts=brca, group=factor(brca.conds))
+#  keep <- filterByExpr(y)
+#  y <- y[keep,keep.lib.sizes=FALSE]
+#  brca.data <- y$counts
+#  
+# s.l.o.w.
+# brca.sens <- aldex.senAnalysis(x, gamma=c(0,0.1,0.2,0.3,0.5,0.7))
+
+# save(brca.sens, file='data-Li2022/brca.sens.Rmd')
+load(file='data-Li2022/brca.sens.Rmd')
+
+sig.orig <- which(brca.sens$gamma_0$we.eBH < 0.05)
+sig.scale <- which(brca.sens$gamma_0.5$we.eBH < 0.05)
+
+par(mfrow=c(1,3))
+
+# which change diff.win
+plot(brca.sens$gamma_0$diff.win, (brca.sens$gamma_0.5$diff.win - brca.sens$gamma_0$diff.win), pch=19, cex=0.4, col=rgb(0,0,0,0.3),xlab="base dispersion", ylab="scale - base dispersion")
+
+points(brca.sens$gamma_0$diff.win[sig.orig], (brca.sens$gamma_0.5$diff.win - brca.sens$gamma_0$diff.win)[sig.orig], pch=19, cex=0.4, col=rgb(1,0,0,0.7))
+
+points(brca.sens$gamma_0$diff.win[sig.scale], (brca.sens$gamma_0.5$diff.win-brca.sens$gamma_0$diff.win)[sig.scale], pch=19, cex=0.4, col=rgb(0,0,1,1))
+
+# which change diff.btw
+plot(brca.sens$gamma_0$diff.btw, (brca.sens$gamma_0.5$diff.btw-brca.sens$gamma_0$diff.btw), pch=19, cex=0.4, col=rgb(0,0,0,0.3), xlab="base logFC", ylab="scale - base logFC")
+
+points(brca.sens$gamma_0$diff.btw[sig.orig], (brca.sens$gamma_0.5$diff.btw-brca.sens$gamma_0$diff.btw)[sig.orig], pch=19, cex=0.4, col=rgb(1,0,0,0.7))
+
+points(brca.sens$gamma_0$diff.btw[sig.scale], (brca.sens$gamma_0.5$diff.btw-brca.sens$gamma_0$diff.btw)[sig.scale], pch=19, cex=0.4, col=rgb(0,0,1,1))
+
+hist(brca.sens$gamma_0$diff.btw, breaks=99, xlab="logFC", main="Histogram of LFC")
+abline(v=0, col='red')
+par(mfrow=c(1,1))
+
+```
+
+The first panel in Supplementary Figure 3 shows that  the lowest dispersion transcripts are differentially affected by adding scale uncertainty.
+The graph in this first panel supports the concept that there is not a defined fold-change cutoff when \(\gamma=0.5\), but rather it is a dynamic between fold-change and dispersion and is consistent with the result seen in Figure 3 in the main text in a different dataset.
+The second panel shows that the log-fold change is only affected minimally by adding scale.
+The final panel shows that there is a slight scale mis-specification, in that the median difference between groups is not centered exactly at 0, and so this dataset may benefit from a full (informed) scale model.
+This analysis was again conducted using the BRCA1 dataset.
+
+```{r sensitivity.plot, echo=FALSE, include=TRUE, fig.dim=c(8,5),  fig.align="center", out.width="1\\linewidth",  fig.align="center", fig.cap=c('Adding scale uncertainty reduces positive identifications  in biological replicate experiments; data from the BRCA dataset obtained from Li et al.
+2022.
+The sensitivity analysis  plot  shows that many transcripts are excluded, even when the gamma value is very low.
+This is an indication that these transcripts are extremely sensitive to normalization assumptions and are likely to be FP.
+The colors are the same as in Supplementary Figure 1.')}
+plotGamma(brca.sens, blackWhite=T)[[2]]
+
+```
+
+Supplementary Figure 4 shows a sensitivity analysis of the BRCA1 dataset.
+In the when \(\gamma=1e^{-3}\) there are `r length(which(brca.sens$gamma_0$we.eBH < 0.05))` significant transcripts (72% of all 21813 transcripts!), which drops to `r length(which(brca.sens$gamma_0.1$we.eBH < 0.05))` when \(\gamma=0.1\) and `r length(which(brca.sens$gamma_0.2$we.eBH < 0.05))` when \(\gamma=0.2\), `r length(which(brca.sens$gamma_0.3$we.eBH < 0.05))` when \(\gamma=0.3\), `r length(which(brca.sens$gamma_0.5$we.eBH < 0.05))` when \(\gamma=0.5\) and finally `r length(which(brca.sens$gamma_0.7$we.eBH < 0.05))` when \(\gamma=0.7.\).
+Here there is a sharp drop in significant transcripts betwee \(\gamma=0.1, 0.3\) and then relative stability between 0.3 and 0.5, suggesting that these latter \(\gamma\) values are likely to be appropriate.
+Note that in biologically-replicated data adding a small amount of scale uncertainty has less of an effect at small uncertainty values than in technically replicated data.
 \newpage
 
-Figure 3 shows how dispersion behaves in linear and log space.
+<!--
+Supplementary Figure 3 shows how dispersion behaves in linear and log space.
 The variance or dispersion always increases with increasing raw (or normalized) read count but decreases when measured on the log-ratio transformed data [@fernandes:2013;@Love:2014aa], reaching a minimum at some mid-point of the distribution.
@@
 [-@nixon2023scale], it is reasonable to conclude that some of these results may be due to the choice of normalization.
+-->
+
+```{r disp, eval=T, echo=F, warning=F, message=F,comment=F, fig.cap="Adding scale uncertainty changes the dispersion distribution.
+Panel A shows a plot of the expected value for relative abundance vs the expected value for the pooled dispersion as output by \\texttt{aldex.effect}.
+The dashed  horizontal lines show the median value for the features with a rAbundance between -0.5 and 0.5, and the light colored lines are lowess lines of fit through the center of mass of the data.
+Panel B plots the dispersion difference between  \\(\\gamma = 1\\) and \\(\\gamma = 0\\); note the non-linear relationship that highlights  the rotation that is evident in Panel A.
+The colored lines indicate the lowess line of fit through the centre of mass of the plot for the various  populations of points.
+The thick grey line is the total population and shows the difference \\(\\Delta\\), the red line is the population of significant transcripts (\\*) with \\(\\gamma = 0\\), the dashed orange line is the population of significant transcripts with a difference threshold (T) of about \\(\\pm 2^{1.4}\\)-fold change, the dashed blue line is the population of significant transcripts with \\(\\gamma = 0.5\\), and the dashed cyan line is the significant population with \\(\\gamma = 1\\).
+\\(\\Delta\\): Difference, *: significant, T: thresholded.
+"}
+
+load(file="analysis/yst.all.Rda") 
+load(file="analysis/yst.s.all.Rda") 
+load(file="analysis/yst.1.all.Rda") # DESeq2 
+load('analysis/res.Rda')
+
+# with gamma = 0.5 
+load('analysis/x.s.all.Rda') 
+sig.des <- which(res@listData$padj < 0.05) 
+sig.ald <- which(yst.all$we.eBH < 0.05) 
+sig.s.ald <- which(yst.s.all$we.eBH < 0.05)
+
+sig.all <- yst.all$we.eBH < 0.05 
+sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4 
+sig.s <- yst.s.all$we.eBH < 0.05 
+sig.1 <- yst.1.all$we.eBH < 0.05
+
+cuts <- yst.all$rab.all > -0.5 & yst.all$rab.all < 5
+
+par(mfrow=c(1,2)) 
+mn.mid <- median(yst.all$diff.win[cuts]) 
+mn.mid.s <- median(yst.s.all$diff.win[cuts]) 
+mn.mid.1 <- median(yst.1.all$diff.win[cuts])
+
+no.scale <- data.frame(yst.all$rab.all, yst.all$diff.win) 
+half.scale <- data.frame(yst.s.all$rab.all, yst.s.all$diff.win) 
+full.scale <- data.frame(yst.1.all$rab.all, yst.1.all$diff.win)
+
+# we change dispersion a lot for the bulk of the features 
+plot(yst.all$rab.all, yst.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,0,0.3), xlab='rAbundance', ylab='Dispersion') 
+lines(lowess(no.scale, f=0.1), col='grey', lwd=3) 
+points(yst.s.all$rab.all, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3)) 
+lines(lowess(half.scale, f=0.1), col='pink', lwd=3) 
+points(yst.1.all$rab.all, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3)) 
+lines(lowess(full.scale, f=0.1), col='royalblue1', lwd=3) 
+abline(h=mn.mid, lty=2) 
+abline(h=mn.mid.s, lty=2, col='red') 
+abline(h=mn.mid.1, lty=2, col='blue') 
+title(main='A', line=-1.2, adj=0.9)
+
+g0 <- expression(paste(gamma, " = 0")) 
+g5 <- expression(paste(gamma, " = 0.5")) 
+g1 <- expression(paste(gamma, " = 1")) 
+legend(1,4, legend=c(g0, g5, g1), pch=19, col=c("black", "red", "blue"))
+
+# plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5, 
+# col=rgb(0,0,1,0.3), ylim=c(0,4.5), xlab='g=0 Dispersion', ylab='g=0.5 | g=1 
+# Dispersion') points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5, 
+# col=rgb(1,0,0,0.3)) abline(0,1, lty=2) title(main='B', line=-1.2, adj=0.5) 
+# legend(0,4, legend=c( "0.5", "1"), pch=19, col=c("red", "blue"))
+
+diff.df <- data.frame(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win) 
+sig.df <- data.frame(yst.all$rab.all[sig.all], yst.1.all$diff.win[sig.all]- yst.all$diff.win[sig.all]) 
+sig.t.df <- data.frame(yst.all$rab.al[sig.t], yst.1.all$diff.win[sig.t] - yst.all$diff.win[sig.t]) 
+sig.s.df <- data.frame(yst.all$rab.all[sig.s], yst.1.all$diff.win[sig.s] - yst.all$diff.win[sig.s]) 
+sig.1.df <- data.frame(yst.all$rab.all[sig.1], yst.1.all$diff.win[sig.1]- yst.all$diff.win[sig.1])
+
+ylab1 <- expression(paste("Dispersion: ", gamma, "1 - ", gamma, "0"))
+
+plot(yst.all$rab.all, yst.1.all$diff.win- yst.all$diff.win, xlab='rAbundance', ylab=ylab1, pch=19, cex=0.5, xlim=c(-11,10), ylim=c(.1,1.5), col=rgb(0,0,0,0.8)) 
+title(main='B', line=-1.2, adj=0.9) 
+lines(lowess(diff.df, f=0.1), col='grey70', lwd=6) 
+lines(lowess(sig.df, f=0.1), col='red', lwd=3) 
+lines(lowess(sig.t.df, f=0.5), col='orange', lwd=3, lty=2) 
+lines(lowess(sig.s.df, f=0.5), col='royalblue1', lwd=3, lty=2) 
+lines(lowess(sig.1.df, f=0.7), col='cyan', lwd=3, lty=2)
+
+d0 <- expression(paste(Delta)) 
+dstar <- expression(paste(Delta,"*")) 
+dstarT <- expression(paste(Delta,"*T")) 
+dg5 <- expression(paste(Delta,"*", gamma, "0.5")) 
+dg1 <- expression(paste(Delta,"*", gamma, "1"))
+
+legend(-12,1.6, legend=c( d0, dstar, dstarT,dg5, dg1), pch="-", col=c("grey", "red", "orange", "royalblue1", "cyan"))
+
+```
+
+The effect on dispersion with increasing  amounts of scale uncertainty in the technical replication dataset (yeast) are shown in Supplementary Figure 5A, where we can see that the  dispersion increases as uncertainty is added.
+Note that the dispersion in the unscaled analysis in Figure 5A reaches a minimum near the mid-point of the distribution, and also does so when the analysis is conducted with DESeq2 (Supplementary Figure 5).
+This shows more clearly that dispersion of many transcripts is  almost negligible in the absence of scale uncertainty.
+This plot makes the counter-intuitive suggestion that the variance in expression of the majority of genes with moderate expression is more predictable  than highly-expressed genes or of housekeeping genes [@Rocha:2020aa].
+This is at odds with the known biology of cells where single cell counting of highly-expressed transcripts shows that they have little intrinsic variation [@Taniguchi:2010aa;@Scott:2010].
+
+Adding scale uncertainty by setting \(\gamma=0.5\) , or \(\gamma = 1.0\), increases the minimum dispersion as shown in Figure 5A by the red and blue data points, and by the colored lines of fit through the centre of mass of the data.
+Less obvious is that the additional dispersion is  not applied equally to all points.
+Figure 5B shows a plot of the difference between the \(\gamma= 0\) and \(\gamma= 1\) data  and here we can see that  scale uncertainty is preferentially increasing the dispersion of the mid-expressed transcripts that formerly had negligible dispersion; examine the grey line of best fit (overlaid by the red line) for the trend.
+Panel B also shows the trend of the expression-dispersion relationship for transcripts that are classed as statistically significant.
+The red line shows the trendline with no added scale uncertainty, and this trendline exactly overlays with the grey trendline for the bulk of transcripts.
+The orange trendline indicates those transcripts that are both statistically significant and that have a thresholded expression level of \( \pm 1.4\), and the dark blue and cyan lines show the statistically significant trendline for \(\gamma=0.5\) , or \(1.0\).
 
@@
 
-Figure 4 shows that the informed models with 5% or 14% difference in location between groups and \(\gamma=0.5\) provide nearly the same output for q-values, effect sizes and difference between groups (black).
+Supplementary Figure 6 shows that the informed models with 5% or 14% difference in location between groups and \(\gamma=0.5\) provide nearly the same output for q-values, effect sizes and difference between groups (black).
 The line of identity is given as the grey diagonal.

### 2025-04-14 code/brca.R

@@ -1,0 +1,183 @@
+if(file.exists("analysis/thin_sim_brca.out.Rda")){
+  load("analysis/thin_sim_brca.out.Rda") # file is data.out, list
+} else {
+  brca <- read.table('data-Li2022/permuted datasets_TCGA/TCGA-BRCA.normal-tumor.pair.rawCount.tsv', header=T, row.names=1, 
+ sep='\t')
+ 
+ brca.conds <- as.vector(unlist(read.table('data-Li2022/permuted datasets_TCGA/TCGA-BRCA.conditions.tsv', sep='\t')))
+	conditions <- data.frame(brca.conds)
+
+
+  library(ALDEx2, warn.conflicts=F)
+	library(seqgendiff, warn.conflicts=F)
+  library(edgeR, warn.conflicts=F)
+  library(DESeq2, warn.conflicts=F)
+
+	###
+	# edgeR functions
+	 
+	y <- DGEList(counts=brca, group=factor(brca.conds))
+	keep <- filterByExpr(y)
+	y <- y[keep,keep.lib.sizes=FALSE]
+	
+	# make the filtered base dataset
+	brca.data <- y$counts
+
+   brca.data.out <- list()
+   
+   # do 10 replicates and keep outputs
+   for(i in 1:10){
+   # this adds rnorm noise to 5% of the transcripts
+   # setting alpha=1 gives no difference between if features are
+   # approximately gaussian
+   thin.brca <- thin_2group(brca.data, prop_null=0.95, alpha=0,
+     signal_fun = stats::rnorm, signal_params = list(mean = 0, sd = 2))
+   
+   x <- aldex(thin.brca$mat, conditions=as.vector(thin.brca$designmat), gamma=1e-3)
+   x.2 <- aldex(thin.brca$mat, conditions=as.vector(thin.brca$designmat), gamma=0.2)
+   x.5 <- aldex(thin.brca$mat, conditions=as.vector(thin.brca$designmat), gamma=0.5)
+   
+   conds <- as.vector(thin.brca$designmat)
+   #DESeq2 functions
+   dds.th  <- DESeqDataSetFromMatrix(countData = thin.brca$mat,
+                                 colData = data.frame(conds),
[…]

### 2025-04-14 code/fix_authors.sh

@@ -1,6 +1,6 @@
 #!/bin/zsh
 
-awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin D. Silverman$^{2,3,4,5}$")};1' go3.tex > jnk.tex
+awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pistner Nixon$^2$ \\and Justin D. Silverman$^{3,4,5}$")};1' go3.tex > jnk.tex
 
 awk '{sub(/\\date{}/,"\\date{\\today}")};1' jnk.tex > jnk2.tex
 
@@ -13,10 +13,11 @@ pdflatex go3-author.tex -o go3-author.pdf
 rm jnk.tex
 rm jnk2.tex
 rm go3.pdf
-rm go3.aux
-rm go3.out
-rm go3.toc
-rm go3.log
+rm go3.syncte*
+#rm go3.aux
+#rm go3.out
+#rm go3.toc
+#rm go3.log
 rm go3-author.aux
 rm go3-author.log
 rm go3-author.blg

### 2025-04-14 code/immuno.R

@@ -1,0 +1,321 @@
+#####
+# R code for the modeled immuno dataset from Li et al
+# This dataset gave many false positives when shuffled
+# new in this analysis is including 5% true positives in
+# the shuffled datasets to examine both FP and TP identification
+# TP data are added using the seqgendiff R package
+#####
+immuno<-read.table(file="data-Li2022//GSE91061_raw_counts_GRCh38.p13_NCBI.tsv", header = T, skip=35, sep='\t', row.names = 1)
+
+m <- read.table(file="data-Li2022//metadata.txt", header=F, row.names=1, sep='\t')
+
+
+### this is butt ugly but it's super easy to follow ...
+# if the output file for the shuffling and modeling exists
+# skip making a new one and load the data instead
+###
+
+if(file.exists("analysis/thin_sim_data.out.Rda")){
+  load("analysis/thin_sim_data.out.Rda") # file is data.out, list
+} else {
+  library(ALDEx2, warn.conflicts=F)
+	library(seqgendiff, warn.conflicts=F)
+  library(edgeR, warn.conflicts=F)
+  library(DESeq2, warn.conflicts=F)
+
+	###
+	# edgeR functions
+	conditions <- rep("Pre", 109)
+	conditions[grep("_On",m)] <- "On"
+	immuno.conds <- data.frame(conditions)
+	 
+	y <- DGEList(counts=immuno, group=factor(conditions))
+	keep <- filterByExpr(y)
+	y <- y[keep,keep.lib.sizes=FALSE]
+	
+	# make the filtered base dataset
+	immuno.data <- y$counts
+
+   data.out <- list()
+   
+   # do 10 replicates and keep outputs
+   for(i in 1:10){
+   # this adds rnorm noise to 5% of the transcripts
+   # setting alpha=1 gives no difference between if features are
+   # approximately gaussian
[…]

### 2025-04-14 checking-housekeeping-yeast.R
resubmitted

@@ -1,28 +1,0 @@
-# note: need to manually pull changed genes and then get the equivalent in KEGG search
-
-up.path <-  read.table('up.path.txt', row.names=1, sep='\t')
-dn.path <-  read.table('dn.path.txt', row.names=1, sep='\t')
-
-both.path <- intersect(rownames(up.path), rownames(dn.path))
-up.only.path <- setdiff(rownames(up.path), rownames(dn.path))
-dn.only.path <- setdiff(rownames(dn.path), rownames(up.path))
-all.path <- union(rownames(up.path), rownames(dn.path))
-
-all.path.df <- as.data.frame(matrix(data=NA, ncol=3, nrow=length(all.path))
-
-colnames(all.path.df) <- c('pathway', 'N.up', 'N.dn')
- 
-rownames(all.path.df) <- all.path
-
-rownames(all.path.df) %in% up.only.path
-
-all.path.df[rownames(all.path.df) %in% up.only.path,1:2] <- up.path[up.only.path,1:2]
-all.path.df[rownames(all.path.df) %in% dn.only.path,c(1,3)] <- dn.path[dn.only.path,c(1,2)]
-all.path.df[rownames(all.path.df) %in% both.path,c(1,3)] <- dn.path[both.path,c(1,2)]
-all.path.df[rownames(all.path.df) %in% both.path,2] <- up.path[both.path,2]
-
-up.path[up.only.path,1]
-
-ribo <- c('YBL027W','YBL072C','YBR146W','YBR181C','YBR189W','YDL075W','YDL082W','YDL136W','YDR064W','YDR237W','YDR382W','YDR418W','YDR447C','YEL054C','YER102W','YER131W','YGL068W','YGL123W','YGR085C','YHL004W','YHL015W','YIL069C','YIL133C','YJL177W','YKL180W','YKR094C','YLL045C','YLR048W','YLR333C','YLR441C','YML026C','YML063W','YMR142C','YMR242C','YNL069C','YNL178W','YOL040C','YOR096W','YOR312C','YPL131W','YPR102C')
-
-ribo.unch <- c('YBR084C-A','YBR251W','YDL081C','YDL202W','YER050C','YER117W','YGL103W','YGR027C','YGR118W','YGR220C','YHR147C','YLR287C-A','YML024W','YML025C','YMR143W','YMR194W','YMR286W','YNL306W','YOR167C','YPL173W','YPL249C-A','YPR132W')

### 2025-04-15 response.Rmd
cleaned up version

@@
 ---
+
+Dear Editor and reviewers
+
+Thank you all for thoughtful and helpful comments.
+In response to your queries we have added a new Figure 1 that more clearly articulates (we hope) the need for scale uncertainty.
+Secondly, we  added in new analyses using semi-synthetic datasets to add in a standard of truth.
+We did this using some of the datasets of, but in a way that was independent of Li et al [@Li:2022aa] and observed essentially the same result.
+In doing this analysis it became clear that the addition of scale uncertainty was much better at controlling FP than  fold-change cutoffs and we came to understand why this is so and have included this in the revision.
+
+We have also tried to clean up language and clarify sources of data, etc as requested.
+
+We have included line numbers in many of the responses and have included line numbers in our resubmission for cross-reference.
+
+Finally, we have included a file with the differences (diff.pdf) marked up in the package so that the changes can be identified.
+
+We believe  that this revised document is now much stronger because of the reviews and hope it meets approval
+
+The repository for this version is at: https://github.com/ggloor/scale-sim-bio/ and is now public
+
+Best regards
+
+Greg Gloor, on behalf of the authors
 

### 2025-04-15 response.tex
cleaned up version

@@
 
+Dear Editor and reviewers
+
+Thank you all for thoughtful and helpful comments. In response to your
+queries we have added a new Figure 1 that more clearly articulates (we
+hope) the need for scale uncertainty. Secondly, we added in new analyses
+using semi-synthetic datasets to add in a standard of truth. We did this
+using some of the datasets of, but in a way that was independent of Li
+et al (1) and observed essentially the same result. In doing this
+analysis it became clear that the addition of scale uncertainty was much
+better at controlling FP than fold-change cutoffs and we came to
+understand why this is so and have included this in the revision.
+
+We have also tried to clean up language and clarify sources of data, etc
+as requested.
+
+We have included line numbers in many of the responses and have included
+line numbers in our resubmission for cross-reference.
+
+Finally, we have included a file with the differences (diff.pdf) marked
+up in the package so that the changes can be identified.
+
+We believe that this revised document is now much stronger because of
+the reviews and hope it meets approval
+
+The repository for this version is at:
+\url{https://github.com/ggloor/scale-sim-bio/} and is now public
+
+Best regards
+
+Greg Gloor, on behalf of the authors
+
 Reviewer: 1
@@
     tightly target that analytic audience. Companion articles by by
-    McGovern et al. (1) discussing the use in gene set enrichment and by
-    Nixon et al. (2) targeting the microbiome community are published in
+    McGovern et al. (2) discussing the use in gene set enrichment and by
+    Nixon et al. (3) targeting the microbiome community are published in
     PLoS Comp. Bio. and under second revision at Genome Biol.
@@
     metatranscriptomics rather than systems-level metatranscriptomics as
-    shown in our report and in (3). We would need to redo all of the
+    shown in our report and in (4). We would need to redo all of the
     analysis in the cited paper in order to make this comparison.
@@
     We thank the reviewer for this suggestion while the original paper
-    by Nixon et al. (4) included extensive simulation and theoretic
+    by Nixon et al. (5) included extensive simulation and theoretic
     justification we recognize that this is not yet published and a
@@
     between groups should be close to 0 (no difference) to avoid FP and
-    FN (5). (lines 358-365, 398-410)
+    FN (6). (lines 358-365, 398-410)
   \end{itemize}
@@
   while the scale approach is generalizable and could be applied to the
-  types of data given in the review paper (the actual method is in (6)),
+  types of data given in the review paper (the actual method is in (7)),
   but that approach is designed to ask a different type of question than
@@
 \item
-  The companion articles by by McGovern et al. (1) discussing the use in
-  gene set enrichment and by Nixon et al. (2) targeting the microbiome
+  The companion articles by by McGovern et al. (2) discussing the use in
+  gene set enrichment and by Nixon et al. (3) targeting the microbiome
   community are published in PLoS Comp. Bio. and under second revision
   at Genome Biol. respectively. The companion article introducing the
-  theory and the initial modeling by Nixon et al. (4) is still under
+  theory and the initial modeling by Nixon et al. (5) is still under
   review at Annals of applied Statistics which is a frustration for the
@@
 \begin{CSLReferences}{1}{1}
+\bibitem[\citeproctext]{ref-Li:2022aa}
+1. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
+\href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
+positives by popular differential expression methods when analyzing
+human population samples}. \emph{Genome Biol}, \textbf{23}, 79.
+
 \bibitem[\citeproctext]{ref-McGovern:2023}
-1. McGovern,K.C., Nixon,M.P. and Silverman,J.D. (2023)
+2. McGovern,K.C., Nixon,M.P. and Silverman,J.D. (2023)
 \href{https://doi.org/10.1371/journal.pcbi.1011659}{Addressing erroneous
@@
 \bibitem[\citeproctext]{ref-Nixon2024B}
-2. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
+3. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
 normalization: Incorporating scale uncertainty in microbiome and gene
@@
 \bibitem[\citeproctext]{ref-Dos-Santos:2024aa}
-3. Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and Gloor,G.B.
+4. Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and Gloor,G.B.
 (2024) \href{https://doi.org/10.1186/s40168-024-01992-w}{Vaginal
@@
 \bibitem[\citeproctext]{ref-nixon2024scale}
-4. Nixon,M.P., McGovern,K.C., Letourneau,J., David,L.A., Lazar,N.A.,
+5. Nixon,M.P., McGovern,K.C., Letourneau,J., David,L.A., Lazar,N.A.,
 Mukherjee,S. and Silverman,J.D. (2024)
@@
 \bibitem[\citeproctext]{ref-Wu2021}
-5. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
+6. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
 \href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
@@
 \bibitem[\citeproctext]{ref-Zhang:2021ab}
-6. Zhang,Y., Thompson,K.N., Huttenhower,C. and Franzosa,E.A. (2021)
+7. Zhang,Y., Thompson,K.N., Huttenhower,C. and Franzosa,E.A. (2021)
 \href{https://doi.org/10.1093/bioinformatics/btab327}{Statistical

### 2025-04-15 go3-author.tex
cleaned up version

@@
 
-\begin{abstract}
-In high-throughput sequencing (HTS) studies, sample-to-sample variation
-in sequencing depth is driven by technical factors, and not by variation
-in the scale (size) of the biological system. Typically a statistical
-normalization removes unwanted technical variation in the data or the
-parameters of the model to enable differential abundance analyses. We
-recently showed that all normalizations make implicit assumptions about
-the unmeasured system scale and that errors in these assumptions can
-dramatically increase false positive and false negative rates. We
-demonstrated that these errors can be mitigated by accounting for
-uncertainty using a \emph{scale model}, which we integrated into the
-ALDEx2 R package. This article provides new insights focusing on the
-application to transcriptomic analysis. We provide transcriptomic case
-studies demonstrating how scale models, rather than traditional
-normalizations, can reduce false positive and false negative rates in
-practice while enhancing the transparency and reproducibility of
-analyses. These scale models replace the need for dual cutoff approaches
-often used to address the disconnect between practical and statistical
-significance. We demonstrate the utility of scale models built based on
-known housekeeping genes in complex metatranscriptomic datasets. Thus
-this work provides guidance on how to incorporate scale into
-transcriptomic data sets.
-\end{abstract}
+$^1$corresponding author: ggloor@uwo.ca
 
@@
 
-\section{Implementation}\label{implementation}
+\section{Methods}\label{methods}
 
@@
 
-\singlespacing
+Acknowledgements: JDS and MPN were supported in part by NIH
+1R01GM148972-01. Some aspects of this was supported by an NSERC grant to
+GBG.
+
+Availability of Code: All code is available at
+\url{https://github.com/ggloor/scale-sim-bio}
+
+Author Contributions: Conceptualization, Methodology, Investigation,
+Formal Analysis, Software GBG, MPN, JDS; Visualization GBG, MPN; Writing
+Original Draft GBG; Writing Review and Editing GBG, MPN, JDS;
+Supervision, Funding JDS. \singlespacing
 

### 2025-04-15 go3.Rmd
cleaned up version

@@
     affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania State University, Department of Medicine, Pennsylvania State University, Institute for Computational and Data Science, Pennsylvania State University, Department of Statistics, Pennsylvania State University
+
 cite-method: citeproc 
@@
 
-abstract: In high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system.
+Abstract: In high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system.
 Typically a statistical normalization removes unwanted technical variation in the data or the parameters of the model to enable differential abundance analyses.
@@
 
-# Implementation
+# Methods
 
@@
 
+Acknowledgements: JDS and MPN were supported in part by NIH 1R01GM148972-01. Some aspects of this was supported by an NSERC grant to GBG.
+
+Availability of Code: All code is available at https://github.com/ggloor/scale-sim-bio
+
+Author Contributions: Conceptualization, Methodology, Investigation, Formal Analysis, Software GBG, MPN, JDS; Visualization GBG, MPN; Writing Original Draft GBG; Writing Review and Editing GBG, MPN, JDS; Supervision, Funding JDS.
 \singlespacing

### 2025-04-15 go3.tex
cleaned up version

@@
 \maketitle
-\begin{abstract}
-In high-throughput sequencing (HTS) studies, sample-to-sample variation
-in sequencing depth is driven by technical factors, and not by variation
-in the scale (size) of the biological system. Typically a statistical
-normalization removes unwanted technical variation in the data or the
-parameters of the model to enable differential abundance analyses. We
-recently showed that all normalizations make implicit assumptions about
-the unmeasured system scale and that errors in these assumptions can
-dramatically increase false positive and false negative rates. We
-demonstrated that these errors can be mitigated by accounting for
-uncertainty using a \emph{scale model}, which we integrated into the
-ALDEx2 R package. This article provides new insights focusing on the
-application to transcriptomic analysis. We provide transcriptomic case
-studies demonstrating how scale models, rather than traditional
-normalizations, can reduce false positive and false negative rates in
-practice while enhancing the transparency and reproducibility of
-analyses. These scale models replace the need for dual cutoff approaches
-often used to address the disconnect between practical and statistical
-significance. We demonstrate the utility of scale models built based on
-known housekeeping genes in complex metatranscriptomic datasets. Thus
-this work provides guidance on how to incorporate scale into
-transcriptomic data sets.
-\end{abstract}
 

### 2025-04-15 supplement_figs.Rmd
cleaned up version

@@
   pdf_document
+header-includes:
+  \renewcommand{\figurename}{Supplementary Figure}
 ---

### 2025-04-15 code/fix_authors.sh

@@ -4,7 +4,7 @@ awk '{sub(/true \\and true \\and true/,"Gregory B. Gloor$^1$ \\and Michelle Pist
 
 awk '{sub(/\\date{}/,"\\date{\\today}")};1' jnk.tex > jnk2.tex
 
-awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$Department of Biochemistry, University of Western Ontario\n\$^2\$Department of Population Health Sciences, Geisinger, Danville, PA\n\$^3\$Department of Medicine, Pennsylvania State University\n\$^4\$Institute for Computational and Data Science, Pennsylvania State University\n\$^5\$Department of Statistics, Pennsylvania State University\n")};1' jnk2.tex > go3-author.tex
+awk '{sub(/\\maketitle/,"\\maketitle\n\n\$^1\$Department of Biochemistry, University of Western Ontario\n\$^2\$Department of Population Health Sciences, Geisinger, Danville, PA\n\$^3\$Department of Medicine, Pennsylvania State University\n\$^4\$Institute for Computational and Data Science, Pennsylvania State University\n\$^5\$Department of Statistics, Pennsylvania State University\n\n\$^1\$corresponding author: ggloor@uwo.ca")};1' jnk2.tex > go3-author.tex
 
 pdflatex go3-author.tex -o go3-author.pdf
 bibtex go3-author

### 2025-06-12 go3.Rmd
revised submission

@@
   - \usepackage[left]{lineno}
-
-package: ALDEx2
-
-Abstract: In high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system.
+abstract: In high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system.
 Typically a statistical normalization removes unwanted technical variation in the data or the parameters of the model to enable differential abundance analyses.
@@
 Thus this work provides guidance on how to incorporate scale into transcriptomic data sets.
+
+package: ALDEx2
 

### 2025-06-12 code/immuno.R

@@ -5,7 +5,7 @@
 # the shuffled datasets to examine both FP and TP identification
 # TP data are added using the seqgendiff R package
 #####
-immuno<-read.table(file="data-Li2022//GSE91061_raw_counts_GRCh38.p13_NCBI.tsv", header = T, skip=35, sep='\t', row.names = 1)
+immuno<-read.table(file="data-Li2022/GSE91061_raw_counts_GRCh38.p13_NCBI.tsv", header = T, skip=35, sep='\t', row.names = 1)
 
 m <- read.table(file="data-Li2022//metadata.txt", header=F, row.names=1, sep='\t')
 

### 2025-07-08 go3-author-2.tex
final version

--- 2025-06-12 go3.tex
@@
 data with ALDEx2}
-\author{true \and true \and true}
-\date{}
+\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{3,4,5}$}
+\date{\today}
 
@@
 \maketitle
+
+$^1$Department of Biochemistry, University of Western Ontario
+$^2$Department of Population Health Sciences, Geisinger, Danville, PA
+$^3$Department of Medicine, Pennsylvania State University
+$^4$Institute for Computational and Data Science, Pennsylvania State University
+$^5$Department of Statistics, Pennsylvania State University
+
+$^1$corresponding author: ggloor@uwo.ca
 \begin{abstract}
